### 2022-05-19 production/prod_analysis.Rmd
preparacion bbdd elsoc y comunas
- datos individuales
- datos contextuales
- codebooks

---
title: "Preparación de datos"
date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
      number_sections: yes
    code_folding: hide
    number_sections: yes
editor_options:
  chunk_output_type: console
---

# Setup

```{r setup}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE, 
	fig.height = 18, 
	fig.width = 11, 
	cache = TRUE
)
options(scipen=9999) # desactivar notacion cientifica
remove(list = ls()) #limpieza del entorno de trabajo
```

```{css, echo=FALSE}
# /* expandir el espacio del documento*/
body .main-container {
      max-width: 1600px !important;
      width: 1600px !important;
    }
#/* expandir width de la TOC*/
div.tocify {
    width: 22% !important;
    max-width: 331px !important;
    max-height: 102% !important;
}
```

# Análisis descriptivo

```{r tab-desc, results="asis", warning=FALSE}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               car,         # recodificar variables
               sjlabelled,  # etiquetado de variables
               sjmisc,      # descriptivos y frecuencias
               sjPlot,      # tablas, plots y descriptivos
               knitr,
               kableExtra,
               summarytools,
               lavaan
               )
load(file = here::here("input/data-proc/study_1.RData"))

# Css
st_css()
st_options(lang = "en",
           footnote = NA,
           bootstrap.css = F,
           custom.css = here::here("input/css/dfsummary.css"),
           dfSummary.custom.1 = NA
           )

# Df variables nivel 1
df<- dfSummary(df_study1 %>% select(in_simpat:edad),
               plain.ascii = FALSE,
               style = "grid",
               tmp.img.dir = "/tmp",
               graph.magnif = TRUE,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = T, # plot
               valid.col = T, # n valido
               col.widths = c(700,10,10,10,10)
               )
df$Variable <- NULL # delete variable column
view(df,
     file = here::here("output/tables/desc01.html")
     ) # Ver tabla en un archivo HTML
webshot::webshot(url =here::here("output/tables/desc01.html"),
        file =here::here("output/tables/desc01.png")) # Remover archivo HTML
print(df, method = "render")
```

### 2022-05-19 production/prod_prep.Rmd
preparacion bbdd elsoc y comunas
- datos individuales
- datos contextuales
- codebooks

---
title: "Preparación de datos"
date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
      number_sections: yes
    code_folding: hide
    number_sections: yes
editor_options:
  chunk_output_type: console
---

# Setup

```{r setup}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
options(scipen=9999) # desactivar notacion cientifica
remove(list = ls()) #limpieza del entorno de trabajo
```

```{css, echo=FALSE}
# /* expandir el espacio del documento*/
body .main-container {
      max-width: 1600px !important;
      width: 1600px !important;
    }
#/* expandir width de la TOC*/
div.tocify {
    width: 22% !important;
    max-width: 331px !important;
    max-height: 102% !important;
}
```

# Download data
```{r download data, eval=FALSE, include=FALSE}
#download and save data

file.exists(here::here("input/data-original/elsoc_long.Rdata"))

load(url("https://dataverse.harvard.edu/api/access/datafile/6160173")) #ELSOC
save(elsoc_long_2016_2021,file = here::here("input/data-original/elsoc_long.Rdata"))
cit17<- read.table(url("https://dataverse.harvard.edu/api/access/datafile/5216836"), 
                   header = T, sep = "\t", fill = TRUE)
save(cit17,file = here::here("input/data-original/cit17.Rdata"))
cit18<- read.table(url("https://dataverse.harvard.edu/api/access/datafile/5216941"), 
                   header = T, sep = "\t", fill = TRUE)
save(cit18,file = here::here("input/data-original/cit18.Rdata"))
cit19<- read.table(url("https://dataverse.harvard.edu/api/access/datafile/5216990"), 
                   header = T, sep = "\t", fill = TRUE)
save(cit19,file = here::here("input/data-original/cit19.Rdata"))

# Chilean CENSUS data (2017)____________________________________________________
download.file(url = "https://github.com/jciturras/data-chile/raw/main/data/censo-2017/censo-2017.RData",
              destfile = here::here("input/data-original/censo17.Rdata"))

# Chilean Socioeconomic Household Survey (CASEN)_______________________________

# Year 2017_______________________________________________
download.file(url = "https://github.com/jciturras/data-chile/raw/main/data/casen-2017/casen2017.RData",
              destfile = here::here("input/data-original/casen17.Rdata"))
# Year 2020_______________________________________________
# Download the original data from:
# http://observatorio.ministeriodesarrollosocial.gob.cl/storage/docs/casen/2020/Casen_en_Pandemia_2020_STATA.dta.zip 
# note: we changed the original name of the file -> casen20.dta 
casen20 <- 
  haven::read_dta(file = here::here("input/data-original/casen20.dta"))
# save the original data in R format 
save(casen20,file = here::here("input/data-original/casen20.Rdata"))
```

## Individual data

```{r}
remove(list = ls()) # clean workspace
# .rs.restartR() #restart R session
if (!require("pacman")) install.packages("pacman") # install pacman
# load libraries
pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here)
load(here::here("input/data-original/elsoc_long.Rdata"))

elsoc_long_2016_2021[elsoc_long_2016_2021 ==-999] <- NA
elsoc_long_2016_2021[elsoc_long_2016_2021 ==-888] <- NA
elsoc_long_2016_2021[elsoc_long_2016_2021 ==-777] <- NA
elsoc_long_2016_2021[elsoc_long_2016_2021 ==-666] <- NA

# Attitudes toward inmigrants__________________________________________________
elsoc_long_2016_2021 <- 
elsoc_long_2016_2021 %>% 
  #create new variables
  mutate(in_simpat=r09,
         in_simili=r10,
         in_ansied=r11,
         in_valfam=r12_01,
         in_valami=r12_02,
         in_amsimb=r12_03,
         in_amreal=r12_04,
         in_acpier=r12_05,
         in_acadop=r12_06,
         in_amigcl=r12_07) %>% 
  sjlabelled::drop_labels(., drop.na = FALSE)

#
sjmisc::frq(elsoc_long_2016_2021$in_simpat)
elsoc_long_2016_2021$in_simpat <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_simpat, 
                      label = "Degree of sympathy for [PER/HAI/VEN] living in Chile") 
elsoc_long_2016_2021$in_simpat <-
sjlabelled::set_labels(x = elsoc_long_2016_2021$in_simpat,
                       labels = c("Very little or not at all", "A little",
                                  "Somewhat","Quite a lot","A lot"))  
sjmisc::frq(elsoc_long_2016_2021$in_simpat)

#
sjmisc::frq(elsoc_long_2016_2021$in_simili)
elsoc_long_2016_2021$in_simili <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_simili, 
                      label = "Degree of similarity between Chileans and [PER/HAI/VEN].") 
elsoc_long_2016_2021$in_simili <-
sjlabelled::set_labels(x = elsoc_long_2016_2021$in_simili,
                       labels = c("Not similar", "Slightly similar", "Somewhat similar",
                                  "Quite similar", "Very similar"))
sjmisc::frq(elsoc_long_2016_2021$in_simili)

#
sjmisc::frq(elsoc_long_2016_2021$in_ansied)
elsoc_long_2016_2021$in_ansied <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_ansied, 
                      label = "Interaction anxiety with [PER/HAI/VEN].") 
elsoc_long_2016_2021$in_ansied <-
sjlabelled::set_labels(x = elsoc_long_2016_2021$in_ansied,
                       labels = c("Very uncomfortable", "Uncomfortable", 
                                  "Neither comfortable nor uncomfortable",
                                  "Comfortable", "Very comfortable"))
sjmisc::frq(elsoc_long_2016_2021$in_ansied)

for (i in c("in_valfam","in_valami","in_amsimb","in_amreal",
            "in_acpier","in_acadop","in_amigcl")) {
elsoc_long_2016_2021[[i]] <-
sjlabelled::set_labels(x = elsoc_long_2016_2021[[i]],
                       labels = c("Strongly disagree","Disagree",
                                  "Neither disagree nor agree",
                                  "Agree", "Strongly agree"))
}

sjmisc::frq(elsoc_long_2016_2021$in_valfam)
elsoc_long_2016_2021$in_valfam <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_valfam, 
                      label = "Degree of agreement: My family values my friends [PER/HAI/VEN].") 
sjmisc::frq(elsoc_long_2016_2021$in_valfam)

sjmisc::frq(elsoc_long_2016_2021$in_valami)
elsoc_long_2016_2021$in_valami <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_valami, 
                      label = "Degree of agreement: My friends value my friends [PER/HAI/VEN].") 
sjmisc::frq(elsoc_long_2016_2021$in_valami)

sjmisc::frq(elsoc_long_2016_2021$in_amsimb)
elsoc_long_2016_2021$in_amsimb <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_amsimb, 
                      label = "Degree of agreement: Chile loses its identity with the arrival of [PER/HAI/VEN].") 
sjmisc::frq(elsoc_long_2016_2021$in_amsimb)

sjmisc::frq(elsoc_long_2016_2021$in_amreal)
elsoc_long_2016_2021$in_amreal <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_amreal, 
                      label = "Degree of agreement: With the arrival of [PER/HAI/VEN] unemployment rises.") 
sjmisc::frq(elsoc_long_2016_2021$in_amreal)

sjmisc::frq(elsoc_long_2016_2021$in_acpier)
elsoc_long_2016_2021$in_acpier <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_acpier, 
                      label = "Degree of agreement: [PER/HAI/VEN] keep their customs") 
sjmisc::frq(elsoc_long_2016_2021$in_acpier)

sjmisc::frq(elsoc_long_2016_2021$in_acadop)
elsoc_long_2016_2021$in_acadop <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_acadop, 
                      label = "Degree of agreement: [PER/HAI/VEN] adopt Chilean customs") 
sjmisc::frq(elsoc_long_2016_2021$in_acadop)

sjmisc::frq(elsoc_long_2016_2021$in_amigcl)
elsoc_long_2016_2021$in_amigcl <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$in_amigcl, 
                      label = "Degree of agreement: [PER/HAI/VEN] have Chilean friends") 
sjmisc::frq(elsoc_long_2016_2021$in_amigcl)

# Socioeconomic_________________________________________________________________

# Education_______________________________________
elsoc_long_2016_2021$educ <- 
  car::recode(elsoc_long_2016_2021$m01,
              "c(1,2,3)=1;c(4,5)=2;c(6,7)=3;c(8,9,10)=4; c(-888,-999)=NA")
elsoc_long_2016_2021$educ <-
  factor(elsoc_long_2016_2021$educ,
         labels = c("Primary","High school","Vocational","Universitary"))

elsoc_long_2016_2021$educ <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$educ,
                      label = "Education")
sjmisc::frq(elsoc_long_2016_2021$educ)

#Recoding of education to years based on casen 2017.
elsoc_long_2016_2021$educ_n<- as.numeric(
  car::recode(elsoc_long_2016_2021$m01, 
              "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
               7=14.8;8=14.9;9=16.9;10=19.07;c(-888,-999)=NA", 
              as.numeric = T))

elsoc_long_2016_2021$educ_n <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$educ_n,
                      label = "Education in years")

class(elsoc_long_2016_2021$educ_n)
sjmisc::frq(elsoc_long_2016_2021$educ_n)

# Household income_________________________________________

#Impute midpoint of income ranges
elsoc_long_2016_2021$m30_rec <-
as.numeric(car::recode(elsoc_long_2016_2021$m30,
           "1=110000;2=251000;3=305000;4=355000;5=400000;
            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))
#Recode DK/DA of Income to NA
elsoc_long_2016_2021$m29_rec <-
  as.numeric(car::recode(elsoc_long_2016_2021$m29,"c(-888,-999)=NA"))

#replace NA of income with new imputed variable
elsoc_long_2016_2021$m29_imp <- 
  ifelse(test = !is.na(elsoc_long_2016_2021$m29_rec),
         yes =  elsoc_long_2016_2021$m29_rec,
         no =  elsoc_long_2016_2021$m30_rec)

# deflate at each year's prices
elsoc_long_2016_2021$deflactor <-
  with(elsoc_long_2016_2021, case_when(
    ola == 2016 ~ 113.88 / 123.82,
    ola == 2017 ~ 116.46 / 123.82,
    ola == 2018 ~ 119.45 / 123.82,
    ola == 2019 ~ 123.82 / 123.82
  ))

# N Household:
elsoc_long_2016_2021 <-
  elsoc_long_2016_2021 %>%
  mutate(n_hogar =
           dplyr::case_when(ola == 1 ~ nhogar1,
                            ola == 2 ~ m46_nhogar,
                            ola == 3 ~ m54,
                            ola == 4 ~ m54))
sjmisc::frq(elsoc_long_2016_2021$n_hogar)

#Recode DK/DA to NA
elsoc_long_2016_2021$n_hogar_r<-
  car::recode(elsoc_long_2016_2021$n_hogar,"c(-888,-999)=NA")

# Per capita household income:
elsoc_long_2016_2021$ing_pc <- 
  (elsoc_long_2016_2021$m29_imp/elsoc_long_2016_2021$n_hogar_r)

elsoc_long_2016_2021$ing_pc <-
sjlabelled::set_label(x = elsoc_long_2016_2021$ing_pc,
                      label = "Household income per capita")

sjmisc::descr(elsoc_long_2016_2021$ing_pc)

# Compute income quintiles
elsoc_long_2016_2021 <- elsoc_long_2016_2021 %>% 
  group_by(ola) %>% 
  mutate(quintil = ntile(-desc(ing_pc), 5)) %>% 
  ungroup()

elsoc_long_2016_2021$quintil <- 
  factor(elsoc_long_2016_2021$quintil,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

elsoc_long_2016_2021$quintil <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$quintil,
                      label = "Houshold income quintile per capita")

sjmisc::frq(elsoc_long_2016_2021$quintil)

#include new quintile category with missing cases
elsoc_long_2016_2021$quintil1<-
  car::recode(elsoc_long_2016_2021$quintil, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

elsoc_long_2016_2021$quintil1 <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$quintil1,
                      label = "Houshold income quintile per capita (NA)") 
sjmisc::frq(elsoc_long_2016_2021$quintil1)
#Age Ranges
elsoc_long_2016_2021$edad <- 
  factor(car::recode(elsoc_long_2016_2021$m0_edad, 
                     "18:29=1;30:49=2;50:64=3;65:150=4"),
         labels = c('18-29', '30-49', '50-64', '65 or more'))
elsoc_long_2016_2021$edad <-
  sjlabelled::set_label(elsoc_long_2016_2021$edad, 
                        label = c("Age groups")) 
# Other controls______________________________________________________________
# Subjective social status: individual ____________________________

elsoc_long_2016_2021$ess <- 
as.numeric(elsoc_long_2016_2021$d01_01)

sjmisc::frq(elsoc_long_2016_2021$ess)
elsoc_long_2016_2021$ess <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$ess,
                      label = "Subjective Social Status: individual")

sjmisc::frq(elsoc_long_2016_2021$ess)
# sjPlot::plot_frq(elsoc_long_2016_2021$ess)
# Subjective social status: migrants (w04 and w05)____________________________
elsoc_long_2016_2021$ess_inm <- 
as.numeric(elsoc_long_2016_2021$d01_04)

sjmisc::frq(elsoc_long_2016_2021$ess_inm)
elsoc_long_2016_2021$ess_inm <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$ess_inm,
                      label = "Subjective Social Status: inmigrants")

sjmisc::frq(elsoc_long_2016_2021$ess_inm)
# sjPlot::plot_frq(elsoc_long_2016_2021$ess_inm)

# Nationality of the interviewee____________________________

elsoc_long_2016_2021$nation <- 
elsoc_long_2016_2021$m45
sjmisc::frq(elsoc_long_2016_2021$nation)

elsoc_long_2016_2021$nation <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$nation, 
                      label = "Nationality") 
elsoc_long_2016_2021$nation <-
sjlabelled::set_labels(x = elsoc_long_2016_2021$nation,
                       labels = c("Chilean","Peruvian","Argentina","Colombian",
                                  "Bolivian", "Haitian","Spanish","Other"))
sjmisc::frq(elsoc_long_2016_2021$nation)

# Known: Immigrant [PER/HAI/VEN]____________________________

# sjmisc::find_var(elsoc_long_2016_2021,"Conocidos")
sjmisc::frq(elsoc_long_2016_2021$r05_01)
elsoc_long_2016_2021$know_inm <- 
elsoc_long_2016_2021$r05_01

elsoc_long_2016_2021$know_inm <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$know_inm, 
                      label = "Known: Immigrants") 
elsoc_long_2016_2021$know_inm <-
sjlabelled::set_labels(x = elsoc_long_2016_2021$know_inm,
                       labels = c("None","Few","Some","Quite a few","Many"))

# Political position ____________________________
elsoc_long_2016_2021$pos_id <-
factor(
  car::recode(
    elsoc_long_2016_2021$c15,
    "c(11,12,-888,-999)='Does not identify';c(0,1,2,3,4)='Left';
     c(5)='Center';c(6,7,8,9,10)='Right'"
  ),
  levels = c('Left', 'Center', 'Right', 'Does not identify')
)

elsoc_long_2016_2021$pos_id <- 
sjlabelled::set_label(x = elsoc_long_2016_2021$pos_id, 
                      label = "Political identification")

sjmisc::frq(elsoc_long_2016_2021$pos_id)

# Select data______________________________________________________________

df_study1 <- 
elsoc_long_2016_2021 %>% 
  select(idencuesta,ola,region,region_cod,comuna,comuna_cod,
         starts_with("in_"),
         educ,educ_n,ing_pc,quintil,quintil1,
         ess,ess_inm,
         nation,know_inm,pos_id,edad)

# sjPlot::view_df(df_study1,show.frq = T,show.values = T,show.na = T)

# save data
save(df_study1,file = here::here("input/data-proc/study_1.RData"))

# save codebook
sjPlot::view_df(df_study1,
        show.frq = T, 
        show.prc = T, 
        show.na = T,
        file = here::here("output/df_study1.html"))
```

## Contextual data

Notas de Trabajo de campo:

* CENSO 2017, día 19 de abril de 2017. (pegar a 2016)
* CASEN 2017, 2 Noviembre 2017 – 4 Febrero 2018. (pegar a 2017)
* CASEN 2020, 31 de octubre 2020 – 04 de febrero 2021. (pegar a 2020)

* Faltarían datos 2019 para algunas comunas

* Crear variable de cambio en la tasa de migrantes a nivel comunal 
  * Censo 2017 - CASEN 2020

![](https://radiografia-cambio-social-2016-2021.netlify.app/reporte-elsoc_files/figure-html/amen1-wave-1.png)

```{r}
remove(list = ls()) # clean workspace
# .rs.restartR() #restart R session
if (!require("pacman")) install.packages("pacman") # install pacman
# load libraries
pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here)
load(here::here("input/data-original/cit17.Rdata"))
load(here::here("input/data-original/cit18.Rdata"))
load(here::here("input/data-original/cit19.Rdata"))
load(here::here("input/data-original/casen17.Rdata"))
load(here::here("input/data-original/casen20.Rdata"))
load(here::here("input/data-original/censo17.Rdata"))
load(here::here("input/data-original/elsoc_long.Rdata"))

df_ine <- 
readxl::read_excel("input/data-original/data_comunas_ine_dem.xlsx")
df_ine$comuna_lc <- tolower(df_ine$comuna)

df_pro_ine <- 
readxl::read_excel("input/data-original/estimaciones-y-proyecciones-2002-2035-comunas.xlsx")
names(df_pro_ine)
df_proye <- 
df_pro_ine %>% 
  select(comuna_cod=Comuna,comuna_nom=`Nombre Comuna`,
         "sexo"=Sexo_1_Hombre_2_Mujer,
         "edad"=Edad,
         pob_17=`Poblacion 2017`,
         pob_18=`Poblacion 2018`,
         pob_19=`Poblacion 2019`,
         pob_20=`Poblacion 2020`)
names(df_proye)

# calculate population proyections 
df_proye_comunas <- 
df_proye %>% 
  group_by(comuna_cod, comuna_nom) %>% 
  summarise(pob_17=sum(pob_17),
            pob_18=sum(pob_18),
            pob_19=sum(pob_19),
            pob_20=sum(pob_20))

df_proye_comunas <-
df_proye_comunas %>% 
  mutate(pob_total_17=sum(df_proye_comunas$pob_17),
         pob_total_18=sum(df_proye_comunas$pob_18),
         pob_total_19=sum(df_proye_comunas$pob_19),
         pob_total_20=sum(df_proye_comunas$pob_20))
names(df_proye_comunas)
# ELSOC long data ____________________________________________________________
df_elsoc <- 
elsoc_long_2016_2021 %>% 
  select(comuna,comuna_cod) %>%
  mutate(comuna_lc=tolower(comuna)) %>% 
  group_by(comuna,comuna_lc,comuna_cod) %>% 
  summarise(n_elsoc=n())

df_a <- full_join(df_elsoc,df_ine, by = "comuna_lc", suffix = c("_elsoc","_ine"))
summary(df_a)

# ELSOC CIT - territorial data _________________________________________________
df_17 <- 
cit17 %>% 
  select(idencuesta,inmigrante:centroam)

df_18 <- 
cit18 %>% 
  select(idencuesta,inmigrante:centroam)

df_19 <- 
cit19 %>% 
  select(idencuesta,inmigrante:centroam)
df_19[df_19==-995] <- NA

# create data set
df_cit_a <- 
elsoc_long_2016_2021 %>% 
  select(comuna,comuna_cod,idencuesta) %>% 
  left_join(x = .,y = df_19, by ="idencuesta") %>% 
  select(-idencuesta)
#check for duplicates
df_cit_b <- 
  df_cit_a %>% distinct() %>% arrange(comuna_cod, inmigrante)

#generate final dataset
df_cit <- 
  df_cit_b %>% 
  group_by(comuna, comuna_cod) %>% 
  summarise(inm_cit17=sum(inmigrante,na.rm = T))

sjmisc::frq(df_cit$inm_cit17)

# CASEN nationality ___________________________________________________________
# 2017_________________________________________________________
df_17 <- 
casen2017 %>% 
  select(comuna,starts_with("r1b"), starts_with("exp"))
# sjmisc::frq(df_17$r1b)

#dummy for inmigrant 
df_17$r1b[df_17$r1b==9] <- NA
# sjmisc::frq(df_17$r1b)
df_17$inm <- ifelse(test = (df_17$r1b==3),yes = 1,no = 0)
# sjmisc::frq(df_17$inm)

# calculate the mean of the dummy by comuna
df_17_comuna <- 
df_17 %>% 
  mutate(comuna_n_17 =sjlabelled::as_character(comuna)) %>% 
  group_by(comuna_n_17, "comuna_cod"=comuna) %>% 
  summarise(
    inm_2017_cas = mean(inm,na.rm=T)*100, #non weigthed
    inm_2017_cas_w = questionr::wtd.mean(inm,weights = expc)*100 #weigthed
    ) %>% 
  arrange(comuna_cod)

summary(df_17_comuna$inm_2017_cas)  #check
summary(df_17_comuna$inm_2017_cas_w)#check
# 2020_________________________________________________________
df_20 <- 
casen20 %>% 
  select(comuna,
         starts_with("r1b"),starts_with("exp"))

#dummy for inmigrant 
df_20$r1b[df_20$r1b==9] <- NA
# sjmisc::frq(df_20$r1b)
df_20$inm <- ifelse(test = (df_20$r1b==3),yes = 1,no = 0)
# sjmisc::frq(df_20$inm)

# calculate the mean of the dummy by comuna
df_20_comuna <- 
df_20 %>% 
  mutate(comuna_n_20 =sjlabelled::as_character(comuna)) %>% 
  group_by(comuna_n_20, "comuna_cod"=comuna) %>% 
  summarise(
    inm_2020_cas = mean(inm,na.rm=T)*100, #non weigthed
    inm_2020_cas_w = questionr::wtd.mean(inm,weights = expc)*100 #weigthed
    ) %>% 
  arrange(comuna_cod)

summary(df_20_comuna$inm_2020_cas)  #check
summary(df_20_comuna$inm_2020_cas_w)#check

df_b <- full_join(df_a, df_17_comuna, by = "comuna_cod")
df_c <- full_join(df_b, df_20_comuna, by = "comuna_cod")

# CENSUS Data_________________________________________________________________
df_censo <- 
censo2017 %>% 
  select(P12, COMUNA)
# sjmisc::frq(df_censo$P12)
# (0-997)
# 998 No aplica
# 999 Missing
# create dummy for inmigrant
df_censo$inm_censo17 <- car::recode(df_censo$P12,recodes = "1:2=0;3:8=1;98:99=NA")
# sjmisc::frq(df_censo$inm_censo17)

# create % of inmigrant by comuna__________________________

df_inm_comunas <- 
df_censo %>% 
  group_by(comuna_cod=COMUNA,) %>% 
  summarise(inm_propcenso17 = mean(inm_censo17, na.rm=T)*100,
            inm_totcenso17=sum(inm_censo17, na.rm=T))

#Merge with CENSUS data (2017)
df_d <- left_join(df_c, df_inm_comunas,  by = "comuna_cod")

#Merge with Population proyection data (2017 - 2020)
df_e <- left_join(df_d,df_proye_comunas, by = "comuna_cod")

#Merge CIT data for 2019 (Census)
df_f <- left_join(df_e,df_cit, by = "comuna_cod")

#Calculate proportion of inmigrant based on proyections and CIT data
df_f <- 
df_f %>% 
  mutate(inm_dem_proye18=(n_2018/pob_18)*100,
         inm_dem_proye19=(n_2019/pob_19)*100,
         inm_dem_proye20=(n_2020/pob_20)*100,
         #CIT data / total pop by comune 2017 CENSUS
         inm_cit_proye17=(inm_cit17/inm_totcenso17)*100)

# Select and organize variables________________________________________________
df_study1_comunas <- 
df_f %>%
  mutate(comuna=sjlabelled::as_character(comuna_cod)) %>% 
  dplyr::select(comuna,
                comuna_cod,
                inm_prop_cas17_w=inm_2017_cas_w,
                inm_tota_cen17=inm_totcenso17,
                inm_prop_cen17=inm_propcenso17,
                inm_tota_cit17=inm_cit17,
                inm_prop_citproy17=inm_cit_proye17,
                inm_prop_demproy18=inm_dem_proye18,
                inm_prop_demproy19=inm_dem_proye19,
                inm_prop_demproy20=inm_dem_proye20,
                inm_prop_cas20_w=inm_2020_cas_w) %>% 
  mutate(dif_inm = inm_prop_cas20_w-inm_prop_cas17_w)
df_study1_comunas$comuna_lc <- NULL
# names(df_study1_comunas)

# Label data____________________________________________________________________
sjlabelled::set_label(df_study1_comunas) <- 
c("Comuna",
  "Codigo Comuna",
  "Proporcion de poblacion inmigrante  (ponderado) - CASEN 2017 - Comunal",
  "Número de poblacion inmigrante - CENSO 2017 - Comunal",
  "Proporcion de poblacion inmigrante - CENSO 2017 - Comunal",
  "Número de poblacion inmigrante - CIT 2017 - Comunal (usando zona censal)",
  "Proporcion de poblacion inmigrante (CIT / DEM)- CIT 2017 y INE/DEM 2017 - Comunal",
  "Proporcion de poblacion inmigrante proyectada -  INE/DEM 2018 - Comunal",
  "Proporcion de poblacion inmigrante proyectada -  INE/DEM 2019 - Comunal",
  "Proporcion de poblacion inmigrante proyectada -  INE/DEM 2020 - Comunal",
  "Proporcion de poblacion inmigrante (ponderado) - CASEN 2020 - Comunal",
  "Cambio en la proporción de población inmigrante entre 2017 y 2020 - CASEN")

# save data____________________________________________________________________
save(df_study1_comunas,file = here::here("input/data-proc/study_1_comunas.RData"))

# save codebook
sjPlot::view_df(df_study1_comunas,
        show.frq = T, 
        show.prc = T, 
        show.na = T,
        file = here::here("output/df_study1_comunas.html"))
```

# Merge individual & contextual

```{r}
remove(list = ls()) # clean workspace
# .rs.restartR() #restart R session
if (!require("pacman")) install.packages("pacman") # install pacman
# load libraries
pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here)

load(file = here::here("input/data-proc/study_1.RData"))
load(file = here::here("input/data-proc/study_1_comunas.RData"))

df_study1_ind_comunas <- 
 dplyr::left_join(x = df_study1,
                  y = df_study1_comunas, 
                  by = "comuna_cod") %>% 
 dplyr::select(idencuesta:edad,
               inm_prop_cas17_w,
               inm_prop_cas20_w,
               dif_inm) %>% 
 dplyr::rename(comuna=comuna.x)

# save data____________________________________________________________________
save(df_study1_ind_comunas,file = here::here("input/data-proc/df_study1_ind_comunas.RData"))

# save codebook
sjPlot::view_df(df_study1_ind_comunas,
        show.frq = T, 
        show.prc = T, 
        show.na = T,
        file = here::here("output/df_study1_ind_comunas.html"))
```

### 2022-05-24 production/prod_prep.Rmd
ajustes bbdd
- individual y comunal

@@
             19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))
+
+#Impute midpoint of income ranges (2021)
+elsoc_long_2016_2021$m30b_rec <-
+as.numeric(car::recode(elsoc_long_2016_2021$m30b,
+           "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))
+
+sjmisc::frq(elsoc_long_2016_2021$m30b_rec)
+
 #Recode DK/DA of Income to NA
@@
          no =  elsoc_long_2016_2021$m30_rec)
+summary(elsoc_long_2016_2021$m29_imp)
+
+elsoc_long_2016_2021$m29_imp <- 
+  ifelse(test = is.na(elsoc_long_2016_2021$m29_imp),
+         yes =  elsoc_long_2016_2021$m30b_rec,
+         no =  elsoc_long_2016_2021$m29_imp)
+summary(elsoc_long_2016_2021$m29_imp)
 
@@
                             ola == 3 ~ m54,
-                            ola == 4 ~ m54))
+                            ola == 4 ~ m54,
+                            ola == 5 ~ m54))
 sjmisc::frq(elsoc_long_2016_2021$n_hogar)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$quintil,
-                      label = "Houshold income quintile per capita")
+                      label = "Household income quintile per capita")
 
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$quintil1,
-                      label = "Houshold income quintile per capita (NA)") 
+                      label = "Household income quintile per capita (NA)") 
 sjmisc::frq(elsoc_long_2016_2021$quintil1)
@@
                        labels = c("None","Few","Some","Quite a few","Many"))
+sjmisc::frq(elsoc_long_2016_2021$know_inm)
+
+# Level of contact: Migrant friends_________________________
+sjmisc::frq(elsoc_long_2016_2021$r05_02)
+elsoc_long_2016_2021$frien_inm <- 
+elsoc_long_2016_2021$r05_02
+
+elsoc_long_2016_2021$frien_inm <- 
+sjlabelled::set_label(x = elsoc_long_2016_2021$frien_inm, 
+                      label = "Level of contact: Inmigrant friends")
+
+elsoc_long_2016_2021$frien_inm <-
+sjlabelled::set_labels(x = elsoc_long_2016_2021$frien_inm,
+                       labels = c("None","Few","Some","Quite a few","Many"))
+sjmisc::frq(elsoc_long_2016_2021$frien_inm)
 
@@
 
-# Select data______________________________________________________________
+# CONTACT_______________________________________________________________
+
+# NATION to wave 1 (original)___________________
+df_nation <- elsoc_long_2016_2021 %>%  
+  select(idencuesta,nation_fix=nation,ola,muestra) %>% 
+  filter(ola %in% c(1,3)) %>% #keep wave 1 and 3 
+  group_by(idencuesta) %>% 
+  summarise(nation_fix=max(nation_fix,na.rm = F)) #keep the maximum value 
+sjmisc::frq(df_nation$nation_fix) # 2 NA
+
+elsoc <- elsoc_long_2016_2021 #dummy dataset _________________________________
+
+# merge the data with fixed variable:
+elsoc <- left_join(x = elsoc,y = df_nation,by ="idencuesta") 
+# sjmisc::frq(elsoc$nation_fix)
+
+table(elsoc$nation==elsoc$nation_fix) #how many cases are different ?
+elsoc$check_nation <- ifelse(elsoc$nation==elsoc$nation_fix,
+                             yes = "IGUALES",
+                             no = "DISTINTOS")
+# sjmisc::frq(elsoc$check_nation) #18 cases
+
+# edit, if value is different from original variable, set the value of wave 1
+elsoc$nation_t1 <- if_else(condition = elsoc$check_nation=="DISTINTOS",
+                           true = elsoc$nation,
+                           false = elsoc$nation_fix,
+                           missing = elsoc$nation_fix)
+
+table(elsoc$nation==elsoc$nation_t1)
+sjmisc::frq(elsoc$nation_t1)
+elsoc$nation <- NULL; elsoc$nation_fix <- NULL; elsoc$check_nation <- NULL
+
+elsoc$nation_t1 <- set_label(elsoc$nation_t1, label = "Nationality")
+
+# KNOW inmigrants _____________________________
+sjPlot::tab_xtab(elsoc$know_inm,
+                 elsoc$ola,show.cell.prc = T,show.na = T)
+sjmisc::frq(elsoc$know_inm)
+elsoc$know_inm_bi <- car::recode(elsoc$know_inm,
+                                                "1=0;2:5=1")
+sjmisc::frq(elsoc$know_inm_bi)
+
+know_inm_bi <- 
+elsoc %>% group_by(idencuesta) %>%
+  summarize(
+            know_inm_t1t4 = know_inm_bi[ola==4] - know_inm_bi[ola==1]
+            )
+
+elsoc <- left_join(x = elsoc,y = know_inm_bi, by = "idencuesta")
+elsoc$know_inm_t1t4 <- set_label(elsoc$know_inm_t1t4,
+                                  label = "Know immigrants (diff. t4 - t1)")
+# sjmisc::frq(elsoc$know_inm_t1t4)
+# sjPlot::tab_xtab(elsoc$know_inm_t1t4,elsoc$ola)
+
+# FRIENDS inmigrants _____________________________
+sjPlot::tab_xtab(elsoc$frien_inm,
+                 elsoc$ola,show.cell.prc = T,show.na = T)
+sjmisc::frq(elsoc$frien_inm)
+elsoc$frien_inm_bi <- car::recode(elsoc$frien_inm,
+                                                "1=0;2:5=1")
+sjmisc::frq(elsoc$frien_inm_bi)
+
+frien_inm_bi <- 
+elsoc %>% group_by(idencuesta) %>%
+  summarize(
+            frien_inm_t1t4 = frien_inm_bi[ola==4] - frien_inm_bi[ola==1]
+            )
+
+elsoc <- left_join(x = elsoc,y = frien_inm_bi, by = "idencuesta")
+
+elsoc$frien_inm_t1t4 <- set_label(elsoc$frien_inm_t1t4,
+                                  label = "Have immigrant friends (diff. t4 - t1)")
+# sjmisc::frq(elsoc$frien_inm_bi)
+# sjmisc::frq(elsoc$frien_inm_t1t4)
+# sjPlot::tab_xtab(elsoc$frien_inm_t1t4,elsoc$ola)
+
+# Select variables______________________________________________________________
+elsoc_long_2016_2021<- elsoc
 
@@
   select(idencuesta,ola,region,region_cod,comuna,comuna_cod,
+         muestra,
+         cuestion_mig,
          starts_with("in_"),
          educ,educ_n,ing_pc,quintil,quintil1,
-         ess,ess_inm,
-         nation,know_inm,pos_id,edad)
+         ess,
+         # ess_inm,
+         nation_t1,
+         know_inm,
+         know_inm_bi,
+         know_inm_t1t4,
+         frien_inm,
+         frien_inm_bi,
+         frien_inm_t1t4,
+         pos_id,
+         edad)  %>% 
+  filter(muestra == 1) #keep original sample (vs refresh)
+
+# sjmisc::frq(df_study1$ola)
+sjPlot::tab_xtab(df_study1$muestra,df_study1$ola)
+
+df_study1_2 <- 
+elsoc_long_2016_2021 %>% 
+  select(idencuesta,ola,region,region_cod,comuna,comuna_cod,
+         muestra,
+         cuestion_mig,
+         starts_with("in_"),
+         educ,
+         educ_n,
+         ing_pc,
+         quintil,
+         quintil1,
+         ess,
+         nation_t1,
+         know_inm_t1t4,
+         frien_inm_t1t4,
+         pos_id,
+         edad
+         ) %>% 
+  filter(muestra == 1) %>% #keep original sample (vs refresh)
+  na.omit()
+
+sjPlot::view_df(df_study1_2,show.frq = T,show.values = T,show.na = T,show.prc = T)
+sjPlot::tab_xtab(var.row = df_study1_2$muestra,
+                 var.col = df_study1_2$ola)
+
+# sjmisc::frq(df_study1_2$ola)
+# sjmisc::frq(df_study1_2$muestra)
+# sjmisc::frq(df_study1_2$cuestion_mig)
+# sjmisc::frq(df_study1_2$idencuesta,show.na = F) %>% data.frame() %>% View()
+
+# Dataset fixed at t5_________________________________________________________
+# keep idencuesta for the t5
+id <- distinct(df_study1_2 %>% select(idencuesta,ola) %>% filter(ola==5))
+# filter data by the idencuesta of t5
+df_study1_3 <- 
+  df_study1_2 %>%  filter(idencuesta %in% id$idencuesta)
+
+# sjmisc::frq(df_study1_2$comuna)
+# sjPlot::view_df(df_study1_3,show.frq = T,show.values = T,show.na = T,show.prc = T)
+# 
+# sjPlot::tab_xtab(var.row = df_study1_3$cuestion_mig,
+#                  var.col = df_study1_3$ola)
+# 
+# sjPlot::tab_xtab(var.row = df_study1_3$muestra,
+#                  var.col = df_study1_3$ola)
 
@@
 
-# save data
+# SAVE DATA____________________________________________________________________
 save(df_study1,file = here::here("input/data-proc/study_1.RData"))
+save(df_study1_2,file = here::here("input/data-proc/study_1_2.RData"))
+save(df_study1_3,file = here::here("input/data-proc/study_1_3.RData"))
 
-# save codebook
+# # save codebook
 sjPlot::view_df(df_study1,
@@
         file = here::here("output/df_study1.html"))
+
+sjPlot::view_df(df_study1_2,
+        show.frq = T,
+        show.prc = T,
+        show.na = T,
+        file = here::here("output/df_study1_2.html"))
+
+sjPlot::view_df(df_study1_3,
+        show.frq = T,
+        show.prc = T,
+        show.na = T,
+        file = here::here("output/df_study1_3.html"))
 ```
@@
 load(file = here::here("input/data-proc/study_1.RData"))
+load(file = here::here("input/data-proc/study_1_2.RData"))
+load(file = here::here("input/data-proc/study_1_3.RData"))
 load(file = here::here("input/data-proc/study_1_comunas.RData"))
@@
 
+df_study1_2_ind_comunas <- 
+ dplyr::left_join(x = df_study1_2,
+                  y = df_study1_comunas, 
+                  by = "comuna_cod") %>% 
+ dplyr::select(idencuesta:edad,
+               inm_prop_cas17_w,
+               inm_prop_cas20_w,
+               dif_inm) %>% 
+ dplyr::rename(comuna=comuna.x)
+
+df_study1_3_ind_comunas <- 
+ dplyr::left_join(x = df_study1_3,
+                  y = df_study1_comunas, 
+                  by = "comuna_cod") %>% 
+ dplyr::select(idencuesta:edad,
+               inm_prop_cas17_w,
+               inm_prop_cas20_w,
+               dif_inm) %>% 
+ dplyr::rename(comuna=comuna.x)
+
 # save data____________________________________________________________________
-save(df_study1_ind_comunas,file = here::here("input/data-proc/df_study1_ind_comunas.RData"))
+save(df_study1_ind_comunas,
+     file = here::here("input/data-proc/df_study1_ind_comunas.RData"))
+save(df_study1_2_ind_comunas,
+     file = here::here("input/data-proc/df_study1_2_ind_comunas.RData"))
+save(df_study1_3_ind_comunas,
+     file = here::here("input/data-proc/df_study1_3_ind_comunas.RData"))
 
@@
         file = here::here("output/df_study1_ind_comunas.html"))
+
+sjPlot::view_df(df_study1_2_ind_comunas,
+        show.frq = T, 
+        show.prc = T, 
+        show.na = T,
+        file = here::here("output/df_study1_2_ind_comunas.html"))
+
+sjPlot::view_df(df_study1_3_ind_comunas,
+        show.frq = T, 
+        show.prc = T, 
+        show.na = T,
+        file = here::here("output/df_study1_3_ind_comunas.html"))
 ```

### 2022-05-25 production/prod_analysis.Rmd
base long full y restringida

@@
 pacman::p_load(dplyr,       # Manipulacion de datos
-               car,         # recodificar variables
-               sjlabelled,  # etiquetado de variables
-               sjmisc,      # descriptivos y frecuencias
-               sjPlot,      # tablas, plots y descriptivos
                knitr,
                kableExtra,
-               summarytools,
-               lavaan
+               summarytools
                )
-load(file = here::here("input/data-proc/study_1.RData"))
+load(file = here::here("input/data-proc/df_study1_long.RData"))
+load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
 
@@
 # Df variables nivel 1
-df<- dfSummary(df_study1 %>% select(in_simpat:edad),
+df<- dfSummary(df_study1_long %>% select(in_simpat:edad),
                plain.ascii = FALSE,
@@
 print(df, method = "render")
+
+# Df variables nivel __________________________________________________________
+df<- dfSummary(df_study1_long_t5 %>% select(in_simpat:edad),
+               plain.ascii = FALSE,
+               style = "grid",
+               tmp.img.dir = "/tmp",
+               graph.magnif = TRUE,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = T, # plot
+               valid.col = T, # n valido
+               col.widths = c(700,10,10,10,10)
+               )
+df$Variable <- NULL # delete variable column
+view(df,
+     file = here::here("output/tables/desc02.html")
+     ) # Ver tabla en un archivo HTML
+webshot::webshot(url =here::here("output/tables/desc02.html"),
+        file =here::here("output/tables/desc02.png")) # Remover archivo HTML
+print(df, method = "render")
 ```
+
+```{r}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4
+               )
+load(file = here::here("input/data-proc/df_study1_long.RData"))
+load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
+```

### 2022-05-25 production/prod_prep.Rmd
base long full y restringida

@@
 #Recoding of education to years based on casen 2017.
-elsoc_long_2016_2021$educ_n<- as.numeric(
+elsoc_long_2016_2021$educyear<- as.numeric(
   car::recode(elsoc_long_2016_2021$m01, 
@@
 
-elsoc_long_2016_2021$educ_n <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$educ_n,
+elsoc_long_2016_2021$educyear <- 
+sjlabelled::set_label(x = elsoc_long_2016_2021$educyear,
                       label = "Education in years")
 
-class(elsoc_long_2016_2021$educ_n)
-sjmisc::frq(elsoc_long_2016_2021$educ_n)
+class(elsoc_long_2016_2021$educyear)
+sjmisc::frq(elsoc_long_2016_2021$educyear)
 
@@
                                   "Bolivian", "Haitian","Spanish","Other"))
+
 sjmisc::frq(elsoc_long_2016_2021$nation)
@@
 
+elsoc_long_2016_2021$pos_id <- factor(elsoc_long_2016_2021$pos_id,
+                                      levels = levels(elsoc_long_2016_2021$pos_id))
+
 elsoc_long_2016_2021$pos_id <- 
@@
 sjmisc::frq(elsoc$nation_t1)
+
 elsoc$nation <- NULL; elsoc$nation_fix <- NULL; elsoc$check_nation <- NULL
@@
 elsoc$nation_t1 <- set_label(elsoc$nation_t1, label = "Nationality")
-
+elsoc$nation_t1 <- set_labels(elsoc$nation_t1,
+                              labels =c("Chilean","Peruvian",
+                                        "Argentina","Colombian",
+                                        "Bolivian", "Haitian","Spanish","Other"))
+elsoc$nation_t1 <- factor(elsoc$nation_t1,
+                          levels = 1:8,
+                          labels = c("Chilean","Peruvian",
+                                     "Argentina","Colombian",
+                                     "Bolivian", "Haitian","Spanish","Other"))
+sjmisc::frq(elsoc$nation_t1)
 # KNOW inmigrants _____________________________
@@
 elsoc %>% group_by(idencuesta) %>%
-  summarize(
+  summarize(know_inm_bi_t4=know_inm_bi[ola==4],
+            know_inm_bi_t1=know_inm_bi[ola==1],
             know_inm_t1t4 = know_inm_bi[ola==4] - know_inm_bi[ola==1]
@@
 elsoc <- left_join(x = elsoc,y = know_inm_bi, by = "idencuesta")
+sjmisc::frq(elsoc$know_inm_t1t4)
 elsoc$know_inm_t1t4 <- set_label(elsoc$know_inm_t1t4,
@@
 elsoc %>% group_by(idencuesta) %>%
-  summarize(
+  summarize(frien_inm_bi_t4=frien_inm_bi[ola==4],
+            frien_inm_bi_t1=frien_inm_bi[ola==1],
             frien_inm_t1t4 = frien_inm_bi[ola==4] - frien_inm_bi[ola==1]
@@
 elsoc <- left_join(x = elsoc,y = frien_inm_bi, by = "idencuesta")
+summary(elsoc$frien_inm_t1t4)
+# if NA in difference, impute data from t4
+elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
+                                true = elsoc$frien_inm_bi_t4,
+                                false = elsoc$frien_inm_t1t4 )
+summary(elsoc$frien_inm_t1t4)
+# if NA in difference, impute data from t1
+elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
+                                true = elsoc$frien_inm_bi_t1,
+                                false = elsoc$frien_inm_t1t4)
+summary(elsoc$frien_inm_t1t4)
 
-elsoc$frien_inm_t1t4 <- set_label(elsoc$frien_inm_t1t4,
+elsoc$frien_inm_t1t4 <- 
+  set_label(elsoc$frien_inm_t1t4,
             label = "Have immigrant friends (diff. t4 - t1)")
-# sjmisc::frq(elsoc$frien_inm_bi)
-# sjmisc::frq(elsoc$frien_inm_t1t4)
+
+sjmisc::frq(elsoc$frien_inm_t1t4)
 # sjPlot::tab_xtab(elsoc$frien_inm_t1t4,elsoc$ola)
@@
 elsoc_long_2016_2021 %>% 
-  select(idencuesta,ola,region,region_cod,comuna,comuna_cod,
+  filter(muestra == 1) %>% #keep original sample (vs refresh)
+  select(idencuesta,ola,region,region_cod,comuna,comunacod=comuna_cod,
          muestra,
@@
          starts_with("in_"),
-         educ,educ_n,ing_pc,quintil,quintil1,
+         educ,educyear,
+         # ing_pc,
+         # quintil,
+         quintil1,
          ess,
@@
          nation_t1,
-         know_inm,
-         know_inm_bi,
+         # know_inm,
+         # know_inm_bi,
+         # frien_inm_bi_t4,
+         # frien_inm_bi_t1,
          know_inm_t1t4,
-         frien_inm,
-         frien_inm_bi,
+         # frien_inm,
+         # frien_inm_bi,
          frien_inm_t1t4,
          pos_id,
-         edad)  %>% 
-  filter(muestra == 1) #keep original sample (vs refresh)
+         edad)
 
-# sjmisc::frq(df_study1$ola)
-sjPlot::tab_xtab(df_study1$muestra,df_study1$ola)
+# Reshape long to wide
+df_study1_wide <- df_study1 %>% 
+  tidyr::pivot_wider(id_cols = c("idencuesta","muestra"),
+                     names_from = "ola",
+                     # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
+                     values_from = names(select(df_study1,comuna,comunacod,cuestion_mig,in_simpat:edad))
+                     )
 
-df_study1_2 <- 
-elsoc_long_2016_2021 %>% 
-  select(idencuesta,ola,region,region_cod,comuna,comuna_cod,
-         muestra,
-         cuestion_mig,
-         starts_with("in_"),
-         educ,
-         educ_n,
-         ing_pc,
-         quintil,
-         quintil1,
-         ess,
-         nation_t1,
-         know_inm_t1t4,
-         frien_inm_t1t4,
-         pos_id,
-         edad
-         ) %>% 
-  filter(muestra == 1) %>% #keep original sample (vs refresh)
-  na.omit()
+# fix data to w01 values
+df_study1_wide$edad_2 <-df_study1_wide$edad_1 #age
+df_study1_wide$edad_3 <-df_study1_wide$edad_1
+df_study1_wide$edad_4 <-df_study1_wide$edad_1
+df_study1_wide$edad_5 <-df_study1_wide$edad_1
 
-sjPlot::view_df(df_study1_2,show.frq = T,show.values = T,show.na = T,show.prc = T)
-sjPlot::tab_xtab(var.row = df_study1_2$muestra,
-                 var.col = df_study1_2$ola)
+df_study1_wide$educ_2 <-df_study1_wide$educ_1 #education
+df_study1_wide$educ_3 <-df_study1_wide$educ_1
+df_study1_wide$educ_4 <-df_study1_wide$educ_1
+df_study1_wide$educ_5 <-df_study1_wide$educ_1
 
-# sjmisc::frq(df_study1_2$ola)
-# sjmisc::frq(df_study1_2$muestra)
-# sjmisc::frq(df_study1_2$cuestion_mig)
-# sjmisc::frq(df_study1_2$idencuesta,show.na = F) %>% data.frame() %>% View()
+df_study1_wide$educyear_2 <-df_study1_wide$educyear_1 #education years
+df_study1_wide$educyear_3 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_4 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_5 <-df_study1_wide$educyear_1
 
-# Dataset fixed at t5_________________________________________________________
-# keep idencuesta for the t5
-id <- distinct(df_study1_2 %>% select(idencuesta,ola) %>% filter(ola==5))
+df_study1_wide$quintil1_2 <-df_study1_wide$quintil1_1 #quintiles
+df_study1_wide$quintil1_3 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_4 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_5 <-df_study1_wide$quintil1_1
+
+df_study1_wide$ess_2 <-df_study1_wide$ess_1 # subjective status
+df_study1_wide$ess_3 <-df_study1_wide$ess_1
+df_study1_wide$ess_4 <-df_study1_wide$ess_1
+df_study1_wide$ess_5 <-df_study1_wide$ess_1
+
+df_study1_wide$pos_id_2 <-df_study1_wide$pos_id_1 # political position
+df_study1_wide$pos_id_3 <-df_study1_wide$pos_id_1
+df_study1_wide$pos_id_4 <-df_study1_wide$pos_id_1
+df_study1_wide$pos_id_5 <-df_study1_wide$pos_id_1
+
+df_study1_wide$nation_t1_2 <-df_study1_wide$nation_t1_1 # nation
+df_study1_wide$nation_t1_3 <-df_study1_wide$nation_t1_1
+df_study1_wide$nation_t1_4 <-df_study1_wide$nation_t1_1
+df_study1_wide$nation_t1_5 <-df_study1_wide$nation_t1_1
+
+df_study1_wide$know_inm_t1t4_2 <-df_study1_wide$know_inm_t1t4_1 # know immigrants
+df_study1_wide$know_inm_t1t4_3 <-df_study1_wide$know_inm_t1t4_1
+df_study1_wide$know_inm_t1t4_4 <-df_study1_wide$know_inm_t1t4_1
+df_study1_wide$know_inm_t1t4_5 <-df_study1_wide$know_inm_t1t4_1
+
+df_study1_wide$frien_inm_t1t4_2 <- df_study1_wide$frien_inm_t1t4_1 # immigrant friends
+df_study1_wide$frien_inm_t1t4_3 <- df_study1_wide$frien_inm_t1t4_1
+df_study1_wide$frien_inm_t1t4_4 <- df_study1_wide$frien_inm_t1t4_1
+df_study1_wide$frien_inm_t1t4_5 <- df_study1_wide$frien_inm_t1t4_1
+
+df_study1_wide$comuna_2 <-df_study1_wide$comuna_1 #comuna
+df_study1_wide$comuna_3 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_4 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_5 <-df_study1_wide$comuna_1
+
+df_study1_wide$comunacod_2 <-df_study1_wide$comunacod_1 #comuna
+df_study1_wide$comunacod_3 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_4 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_5 <-df_study1_wide$comunacod_1
+
+dim(df_study1_wide)
+
+sjPlot::view_df(df_study1_wide,
+                show.frq = T,show.values = T,show.na = T,show.prc = T, show.type = T)
+
+# reshape from long to wide
+pacman::p_load(datasets,data.table)
+df_study1_long <- data.table::melt.data.table(data.table::setDT(df_study1_wide),
+              id.vars = c("idencuesta","muestra"),
+              variable.name = c("ola"),
+              measure = patterns("^comuna_","^comunacod_","^cuestion_mig_",
+                                 "^in_simpat_","^in_simili_","^in_ansied_","^in_valfam_",
+                                 "^in_valami_","^in_amsimb_","^in_amreal_","^in_acpier_",
+                                 "^in_acadop_","^in_amigcl_",
+                                 "^educ_","^educyear_","^quintil1_","^ess_","^nation_t1_",
+                                 "^know_inm_t1t4_","^frien_inm_t1t4_","^pos_id_",
+                                 "^edad_"
+                                 ),
+              value.name = c("comuna","comunacod","cuestion_mig",
+                             "in_simpat","in_simili","in_ansied","in_valfam",
+                             "in_valami","in_amsimb","in_amreal",
+                             "in_acpier","in_acadop","in_amigcl",
+                             "educ","educyear","quintil1","ess","nation_t1",
+                             "know_inm_t1t4","frien_inm_t1t4","pos_id",
+                             "edad"),
+              na.rm = F,value.factor = TRUE
+              )
+
+names(df_study1_long) #check names of long dataset
+dim(df_study1_long) #check dimensions of the dataframe
+# filter the sataset for the waves 1 to 5
+df_study1_long <-
+df_study1_long %>% 
+  filter(ola %in% c(1,2,3,4,5)) %>% 
+  mutate(ola=factor(ola,levels = 1:5,labels = 1:5))
+dim(df_study1_long) #check, now is OK
+
+df_study1_long <- 
+set_label(x = df_study1_long,
+          label = get_label(select(df_study1,names(df_study1_long))))
+
+sjPlot::view_df(df_study1_long,
+                show.frq = T,show.values = T,show.na = T,show.prc = T,
+                show.type = T)
+#______________________________________________________________________________
+# obtain the idencuesta for wave 5
+ids <- 
+  df_study1 %>% 
+  select(idencuesta,ola) %>% 
+  filter(ola==5) %>% 
+  sjmisc::frq(idencuesta,show.na = F) %>% as.data.frame()
+
 # filter data by the idencuesta of t5
-df_study1_3 <- 
-  df_study1_2 %>%  filter(idencuesta %in% id$idencuesta)
+df_study1_long_t5 <- 
+  df_study1_long %>%
+  filter(idencuesta %in% ids$val)
 
-# sjmisc::frq(df_study1_2$comuna)
-# sjPlot::view_df(df_study1_3,show.frq = T,show.values = T,show.na = T,show.prc = T)
-# 
-# sjPlot::tab_xtab(var.row = df_study1_3$cuestion_mig,
-#                  var.col = df_study1_3$ola)
-# 
-# sjPlot::tab_xtab(var.row = df_study1_3$muestra,
-#                  var.col = df_study1_3$ola)
-
-# sjPlot::view_df(df_study1,show.frq = T,show.values = T,show.na = T)
+names(df_study1_long_t5)
+dim(df_study1_long_t5)
+sjmisc::frq(df_study1_long_t5$ola)
 
 # SAVE DATA____________________________________________________________________
-save(df_study1,file = here::here("input/data-proc/study_1.RData"))
-save(df_study1_2,file = here::here("input/data-proc/study_1_2.RData"))
-save(df_study1_3,file = here::here("input/data-proc/study_1_3.RData"))
+save(df_study1_long,file = here::here("input/data-proc/df_study1_long.RData"))
+save(df_study1_long_t5,file = here::here("input/data-proc/df_study1_long_t5.RData"))
 
-# # save codebook
-sjPlot::view_df(df_study1,
+# save codebook
+sjPlot::view_df(df_study1_long,
         show.frq = T,
@@
         show.na = T,
-        file = here::here("output/df_study1.html"))
+        file = here::here("output/df_study1_long.html"))
 
-sjPlot::view_df(df_study1_2,
+sjPlot::view_df(df_study1_long_t5,
         show.frq = T,
@@
         show.na = T,
-        file = here::here("output/df_study1_2.html"))
-
-sjPlot::view_df(df_study1_3,
-        show.frq = T,
-        show.prc = T,
-        show.na = T,
-        file = here::here("output/df_study1_3.html"))
+        file = here::here("output/df_study1_long_t5.html"))
 ```
@@
   dplyr::select(comuna,
-                comuna_cod,
+                comunacod=comuna_cod,
                 inm_prop_cas17_w=inm_2017_cas_w,
@@
 
-load(file = here::here("input/data-proc/study_1.RData"))
-load(file = here::here("input/data-proc/study_1_2.RData"))
-load(file = here::here("input/data-proc/study_1_3.RData"))
+load(file = here::here("input/data-proc/df_study1_long.RData"))
+load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
 load(file = here::here("input/data-proc/study_1_comunas.RData"))
 
-df_study1_ind_comunas <- 
- dplyr::left_join(x = df_study1,
+df_study1_long_comunas <- 
+ dplyr::left_join(x = df_study1_long,
                   y = df_study1_comunas, 
-                  by = "comuna_cod") %>% 
+                  by = "comunacod") %>% 
  dplyr::select(idencuesta:edad,
@@
 
-df_study1_2_ind_comunas <- 
- dplyr::left_join(x = df_study1_2,
+df_study_long_t5_comunas <- 
+ dplyr::left_join(x = df_study1_long_t5,
                   y = df_study1_comunas, 
-                  by = "comuna_cod") %>% 
- dplyr::select(idencuesta:edad,
-               inm_prop_cas17_w,
-               inm_prop_cas20_w,
-               dif_inm) %>% 
- dplyr::rename(comuna=comuna.x)
-
-df_study1_3_ind_comunas <- 
- dplyr::left_join(x = df_study1_3,
-                  y = df_study1_comunas, 
-                  by = "comuna_cod") %>% 
+                  by = "comunacod") %>% 
  dplyr::select(idencuesta:edad,
@@
 # save data____________________________________________________________________
-save(df_study1_ind_comunas,
-     file = here::here("input/data-proc/df_study1_ind_comunas.RData"))
-save(df_study1_2_ind_comunas,
-     file = here::here("input/data-proc/df_study1_2_ind_comunas.RData"))
-save(df_study1_3_ind_comunas,
-     file = here::here("input/data-proc/df_study1_3_ind_comunas.RData"))
+save(df_study1_long_comunas,
+     file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+save(df_study_long_t5_comunas,
+     file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
 
 # save codebook
-sjPlot::view_df(df_study1_ind_comunas,
+sjPlot::view_df(df_study1_long_comunas,
         show.frq = T, 
@@
         show.na = T,
-        file = here::here("output/df_study1_ind_comunas.html"))
+        file = here::here("output/df_study1_long_comunas.html"))
 
-sjPlot::view_df(df_study1_2_ind_comunas,
+sjPlot::view_df(df_study_long_t5_comunas,
         show.frq = T, 
@@
         show.na = T,
-        file = here::here("output/df_study1_2_ind_comunas.html"))
-
-sjPlot::view_df(df_study1_3_ind_comunas,
-        show.frq = T, 
-        show.prc = T, 
-        show.na = T,
-        file = here::here("output/df_study1_3_ind_comunas.html"))
+        file = here::here("output/df_study_long_t5_comunas.html"))
 ```

### 2022-06-13 production/prod_analysis-fix-t5.Rmd
analysis

---
title: "Preparación de datos"
date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
      number_sections: yes
    code_folding: hide
    number_sections: yes
editor_options:
  chunk_output_type: console
---
# Setup

```{r setup}
knitr::opts_chunk$set(
	echo = FALSE,
	results = "asis",
	message = FALSE,
	warning = FALSE, 
	fig.height = 18, 
	fig.width = 11, 
	cache = TRUE
)
options(scipen=9999) # desactivar notacion cientifica
remove(list = ls()) #limpieza del entorno de trabajo
```

```{css, echo=FALSE}
# /* expandir el espacio del documento*/
body .main-container {
      max-width: 1600px !important;
      width: 1600px !important;
    }
#/* expandir width de la TOC*/
div.tocify {
    width: 22% !important;
    max-width: 331px !important;
    max-height: 102% !important;
}
```

# Descriptive analysis

## Univariate

```{r tab-desc, results="asis", warning=FALSE}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# Css
st_css()
st_options(lang = "en",
           footnote = NA,
           bootstrap.css = F,
           custom.css = here::here("input/css/dfsummary.css"),
           dfSummary.custom.1 = NA
           )

# Df variables nivel 1
# df<- dfSummary(df_study1_long %>% select(in_simpat:edad),
#                plain.ascii = FALSE,
#                style = "grid",
#                tmp.img.dir = "/tmp",
#                graph.magnif = TRUE,
#                headings = F,  # encabezado
#                varnumbers = F, # num variable
#                labels.col = T, # etiquetas
#                na.col = F,    # missing
#                graph.col = T, # plot
#                valid.col = T, # n valido
#                col.widths = c(700,10,10,10,10)
#                )
# df$Variable <- NULL # delete variable column
# view(df,
#      file = here::here("output/tables/desc01.html")
#      ) # Ver tabla en un archivo HTML
# webshot::webshot(url =here::here("output/tables/desc01.html"),
#         file =here::here("output/tables/desc01.png")) # Remover archivo HTML
# # print(df, method = "render")

# Df variables nivel 1 __________________________________________________________
df_study1 <- df_study_long_t5_comunas %>% select(in_simpat:edad,
                                                 inm_prop_cas17_w,
                                                 inm_prop_cas20_w,dif_inm) 
lab_study1 <- sjlabelled::get_label(df_study1) 
df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)

df<- dfSummary(x = df_study1,
               plain.ascii = FALSE,
               style = "grid",
               tmp.img.dir = "/tmp",
               graph.magnif = TRUE,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = T, # plot
               valid.col = T, # n valido
               col.widths = c(700,10,10,10,10,10)
               )
# df$Variable <- NULL # delete variable column
view(df,
     file = here::here("output/tables/desc02.html")
     ) # Ver tabla en un archivo HTML
webshot::webshot(url =here::here("output/tables/desc02.html"),
        file =here::here("output/tables/desc02.png")) # Remover archivo HTML
# print(df, method = "render")
```

## Correlation

```{r cor-plot, fig.dim=c(10,6)}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               sjPlot,
               sjmisc,
               summarytools,
               lavaan
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# Full sample_______________________________________________________________________
mat_full<- 
df_study_long_t5_comunas %>%
  # filter(ola==1) %>%
  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
         in_simili,in_amsimb,in_acpier, in_acadop, # identity
         in_amreal # threat
         ) %>% 
  psych::mixedCor()

cor_full <- as.matrix(mat_full$rho)
diag(cor_full) = NA #set diagonal values to NA
# Set Row names of the matrix
rownames(cor_full) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
                     "C. Family value of immigrant friends [PER/HAI/VEN]",
                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
                     "F. Similarity between Chileans and [PER/HAI/VEN]",
                     "G. Loose of identity because of immigrants ",
                     "H. Immigrant keeping their customs",
                     "I. Adoption of Chilean customs",
                     "J. Unemployment increases")
#set Column names of the matrix
colnames(cor_full) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
#Plot the matrix using corrplot
corrplot::corrplot(cor_full,
  method = "color",
  addCoef.col = "#000390",
  type = "upper",
  tl.col = "black",
  col=colorRampPalette(c("white","#0068DC"))(12),
  bg = "white",
  na.label = "-",
  title = "\n Full sample")

# Wave 1_______________________________________________________________________
mat_1<- 
df_study_long_t5_comunas %>%
  filter(ola==1) %>%
  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
         in_simili,in_amsimb,in_acpier, in_acadop, # identity
         in_amreal # threat
         ) %>% 
  psych::mixedCor()

cor_1 <- as.matrix(mat_1$rho)
diag(cor_1) = NA #set diagonal values to NA
# Set Row names of the matrix
rownames(cor_1) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
                     "C. Family value of immigrant friends [PER/HAI/VEN]",
                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
                     "F. Similarity between Chileans and [PER/HAI/VEN]",
                     "G. Loose of identity because of immigrants ",
                     "H. Immigrant keeping their customs",
                     "I. Adoption of Chilean customs",
                     "J. Unemployment increases")
#set Column names of the matrix
colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
#Plot the matrix using corrplot
corrplot::corrplot(cor_1,
  method = "color",
  addCoef.col = "#000390",
  type = "upper",
  tl.col = "black",
  col=colorRampPalette(c("white","#0068DC"))(12),
  bg = "white",
  na.label = "-",
  title = "\n Wave 1")

# Wave 2_______________________________________________________________________
mat_2<-   
df_study_long_t5_comunas %>%
  filter(ola==2) %>% 
  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
         in_simili,in_amsimb,in_acpier, in_acadop, # identity
         in_amreal # threat
         ) %>% 
  psych::mixedCor()

cor_2 <- as.matrix(mat_2$rho)
diag(cor_2) = NA #set diagonal values to NA
# Set Row names of the matrix
rownames(cor_2) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
                     "C. Family value of immigrant friends [PER/HAI/VEN]",
                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
                     "F. Similarity between Chileans and [PER/HAI/VEN]",
                     "G. Loose of identity because of immigrants ",
                     "H. Immigrant keeping their customs",
                     "I. Adoption of Chilean customs",
                     "J. Unemployment increases")
#set Column names of the matrix
colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
#Plot the matrix using corrplot
corrplot::corrplot(cor_2,
  method = "color",
  addCoef.col = "#000390",
  type = "upper",
  tl.col = "black",
  col=colorRampPalette(c("white","#0068DC"))(12),
  bg = "white",
  na.label = "-",
  title = "\n Wave 2")

# Wave 3_______________________________________________________________________
mat_3 <- 
df_study_long_t5_comunas %>%
  filter(ola==3) %>% 
  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
         in_simili,in_amsimb,in_acpier, in_acadop, # identity
         in_amreal # threat
         ) %>% 
  psych::mixedCor()

cor_3 <- as.matrix(mat_3$rho)
diag(cor_3) = NA #set diagonal values to NA
# Set Row names of the matrix
rownames(cor_3) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
                     "C. Family value of immigrant friends [PER/HAI/VEN]",
                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
                     "F. Similarity between Chileans and [PER/HAI/VEN]",
                     "G. Loose of identity because of immigrants ",
                     "H. Immigrant keeping their customs",
                     "I. Adoption of Chilean customs",
                     "J. Unemployment increases")
#set Column names of the matrix
colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
#Plot the matrix using corrplot
corrplot::corrplot(cor_3,
  method = "color",
  addCoef.col = "#000390",
  type = "upper",
  tl.col = "black",
  col=colorRampPalette(c("white","#0068DC"))(12),
  bg = "white",
  na.label = "-",
  title = "\n Wave 3")

# Wave 4_______________________________________________________________________
mat_4 <- 
df_study_long_t5_comunas %>%
  filter(ola==4) %>% 
  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
         in_simili,in_amsimb,in_acpier, in_acadop, # identity
         in_amreal # threat
         ) %>% 
  psych::mixedCor()

cor_4 <- as.matrix(mat_4$rho)
diag(cor_4) = NA #set diagonal values to NA
# Set Row names of the matrix
rownames(cor_4) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
                     "C. Family value of immigrant friends [PER/HAI/VEN]",
                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
                     "F. Similarity between Chileans and [PER/HAI/VEN]",
                     "G. Loose of identity because of immigrants ",
                     "H. Immigrant keeping their customs",
                     "I. Adoption of Chilean customs",
                     "J. Unemployment increases")
#set Column names of the matrix
colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
#Plot the matrix using corrplot
corrplot::corrplot(cor_4,
  method = "color",
  addCoef.col = "#000390",
  type = "upper",
  tl.col = "black",
  col=colorRampPalette(c("white","#0068DC"))(12),
  bg = "white",
  na.label = "-",
  title = "\n Wave 4")

# Wave 5_______________________________________________________________________
mat_5 <- 
df_study_long_t5_comunas %>%
  filter(ola==5) %>% 
  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
         in_simili,in_amsimb,in_acpier, in_acadop, # identity
         in_amreal # threat
         ) %>% 
  psych::mixedCor()

cor_5 <- as.matrix(mat_5$rho)
diag(cor_5) = NA #set diagonal values to NA
# Set Row names of the matrix
rownames(cor_5) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
                     "C. Family value of immigrant friends [PER/HAI/VEN]",
                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
                     "F. Similarity between Chileans and [PER/HAI/VEN]",
                     "G. Loose of identity because of immigrants ",
                     "H. Immigrant keeping their customs",
                     "I. Adoption of Chilean customs",
                     "J. Unemployment increases")
#set Column names of the matrix
colnames(cor_5) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
#Plot the matrix using corrplot
corrplot::corrplot(cor_5,
  method = "color",
  addCoef.col = "#000390",
  type = "upper",
  tl.col = "black",
  col=colorRampPalette(c("white","#0068DC"))(12),
  bg = "white",
  na.label = "-",
  title = "\n Wave 5")
```

# Models

## A. Anxiety

```{r anxiety}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )
# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# A. Anxiety (in_ansied)
anx.null <- lmer(formula(paste0("in_ansied~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
anx1 <- lmer(formula(paste0("in_ansied~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
anx2.ed <- lmer(formula(paste0("in_ansied~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
anx2.in <- lmer(formula(paste0("in_ansied~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
anx3 <- lmer(formula(paste0("in_ansied~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
anx4.kno <- lmer(formula(paste0("in_ansied~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
anx4.fri <- lmer(formula(paste0("in_ansied~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(anx1,anx2.ed,anx2.in,anx3,anx4.fri,anx4.kno),
          omit.coef = omit)
```

## B. Simpathy

```{r simpathy}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )
# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# B. Simpathy (in_simpat)
simp.null <- lmer(formula(paste0("in_simpat~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
simp1 <- lmer(formula(paste0("in_simpat~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simp2.ed <- lmer(formula(paste0("in_simpat~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simp2.in <- lmer(formula(paste0("in_simpat~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simp3 <- lmer(formula(paste0("in_simpat~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simp4.kno <- lmer(formula(paste0("in_simpat~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simp4.fri <- lmer(formula(paste0("in_simpat~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(simp1,simp2.ed,simp2.in,simp3,simp4.fri,simp4.kno),
          omit.coef = omit)
```

## C. Family value migrant friends

```{r family-value}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )
# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# C. Family value migrant friends (in_simpat)
valfam.null <- lmer(formula(paste0("in_valfam~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
valfam1 <- lmer(formula(paste0("in_valfam~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valfam2.ed <- lmer(formula(paste0("in_valfam~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valfam2.in <- lmer(formula(paste0("in_valfam~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valfam3 <- lmer(formula(paste0("in_valfam~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valfam4.kno <- lmer(formula(paste0("in_valfam~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valfam4.fri <- lmer(formula(paste0("in_valfam~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(valfam1,valfam2.ed,valfam2.in,valfam3,valfam4.fri,valfam4.kno),
          omit.coef = omit)
```

## D. Friend value migrant friends

```{r friends-value}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )
# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# D. Friend value migrant friends(in_simpat)
valami.null <- lmer(formula(paste0("in_valami~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
valami1 <- lmer(formula(paste0("in_valami~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valami2.ed <- lmer(formula(paste0("in_valami~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valami2.in <- lmer(formula(paste0("in_valami~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valami3 <- lmer(formula(paste0("in_valami~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valami4.kno <- lmer(formula(paste0("in_valami~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
valami4.fri <- lmer(formula(paste0("in_valami~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(valami1,valami2.ed,valami2.in,valami3,valami4.fri,valami4.kno),
          omit.coef = omit)
```

## E. Immigrants have Chilean friends

```{r imig-clfriends}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )
# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# E. Immigrants have Chilean friends 
amicl.null <- lmer(formula(paste0("in_amigcl~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)

amicl1 <- lmer(formula(paste0("in_amigcl~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
amicl2.ed <- lmer(formula(paste0("in_amigcl~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
amicl2.in <- lmer(formula(paste0("in_amigcl~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
amicl3 <- lmer(formula(paste0("in_amigcl~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
amicl4.kno <- lmer(formula(paste0("in_amigcl~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
amicl4.fri <- lmer(formula(paste0("in_amigcl~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(amicl1,amicl2.ed,amicl2.in,amicl3,amicl4.fri,amicl4.kno),
          omit.coef = omit)
```

## Conviviality (average)

```{r conviviality}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )

# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# E. Immigrants have Chilean friends 
conv.null <- lmer(formula(paste0("conviviality~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
conv1 <- lmer(formula(paste0("conviviality~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
conv2.ed <- lmer(formula(paste0("conviviality~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
conv2.in <- lmer(formula(paste0("conviviality~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
conv3 <- lmer(formula(paste0("conviviality~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
conv4.kno <- lmer(formula(paste0("conviviality~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
conv4.fri <- lmer(formula(paste0("conviviality~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(conv1,conv2.ed,conv2.in,conv3,conv4.fri,conv4.kno),
          omit.coef = omit)
```

## F. Similarity

```{r similarity}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )

# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# F. Similarity  
simi.null <- lmer(formula(paste0("in_simili~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)

simi1 <- lmer(formula(paste0("in_simili~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simi2.ed <- lmer(formula(paste0("in_simili~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simi2.in <- lmer(formula(paste0("in_simili~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simi3 <- lmer(formula(paste0("in_simili~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simi4.kno <- lmer(formula(paste0("in_simili~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
simi4.fri <- lmer(formula(paste0("in_simili~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(simi1,simi2.ed,simi2.in,simi3,simi4.fri,simi4.kno),
          omit.coef = omit)
```

## G. Loose identity

```{r loose-id}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )

# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

## G. Loose identity
loos.null <- lmer(formula(paste0("in_amsimb~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)

loos1 <- lmer(formula(paste0("in_amsimb~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
loos2.ed <- lmer(formula(paste0("in_amsimb~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
loos2.in <- lmer(formula(paste0("in_amsimb~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
loos3 <- lmer(formula(paste0("in_amsimb~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
loos4.kno <- lmer(formula(paste0("in_amsimb~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
loos4.fri <- lmer(formula(paste0("in_amsimb~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(loos1,loos2.ed,loos2.in,loos3,loos4.fri,loos4.kno),
          omit.coef = omit)
```

## H. Immigrants keep customs

```{r keep-custom}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola))

# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# H. Immigrants keep customs  
keep.null <- lmer(formula(paste0("in_acpier~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)

keep1 <- lmer(formula(paste0("in_acpier~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
keep2.ed <- lmer(formula(paste0("in_acpier~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
keep2.in <- lmer(formula(paste0("in_acpier~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
keep3 <- lmer(formula(paste0("in_acpier~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
keep4.kno <- lmer(formula(paste0("in_acpier~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
keep4.fri <- lmer(formula(paste0("in_acpier~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(keep1,keep2.ed,keep2.in,keep3,keep4.fri,keep4.kno),
          omit.coef = omit)
```

## I. Adopt chilean customs

```{r}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola))

# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# I. Adpot chilean customs
adopt.null <- lmer(formula(paste0("in_acpier~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
adopt1 <- lmer(formula(paste0("in_acpier~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
adopt2.ed <- lmer(formula(paste0("in_acpier~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
adopt2.in <- lmer(formula(paste0("in_acpier~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
adopt3 <- lmer(formula(paste0("in_acpier~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
adopt4.kno <- lmer(formula(paste0("in_acpier~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
adopt4.fri <- lmer(formula(paste0("in_acpier~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(adopt1,adopt2.ed,adopt2.in,adopt3,adopt4.fri,adopt4.kno),
          omit.coef = omit)
```

## Identity (average)

```{r identity}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )

# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# Identity
ident.null <- lmer(formula(paste0("identity~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
ident1 <- lmer(formula(paste0("identity~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
ident2.ed <- lmer(formula(paste0("identity~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
ident2.in <- lmer(formula(paste0("identity~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
ident3 <- lmer(formula(paste0("identity~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
ident4.kno <- lmer(formula(paste0("identity~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
ident4.fri <- lmer(formula(paste0("identity~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(ident1,ident2.ed,ident2.in,ident3,ident4.fri,ident4.kno),
          omit.coef = omit)
```

## J. Unemployment increases

```{r threat}
rm(list=ls())
if (!require("pacman")) install.packages("pacman") # instalar pacman
                            # cargar librerias
pacman::p_load(dplyr,       # Manipulacion de datos
               knitr,
               kableExtra,
               summarytools,
               lme4,
               texreg
               )
load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# generate analytical sample
df_study1 <- 
  df_study_long_t5_comunas %>%
  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
  na.omit() %>% 
  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
         threat = as.numeric(in_amreal),
         ola=as.numeric(ola)
         )

# Hipotesis
h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"

# J. Unemployment increases
threat.null <- lmer(formula(paste0("threat~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
threat1 <- lmer(formula(paste0("threat~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
threat2.ed <- lmer(formula(paste0("threat~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
threat2.in <- lmer(formula(paste0("threat~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
threat3 <- lmer(formula(paste0("threat~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
threat4.kno <- lmer(formula(paste0("threat~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
threat4.fri <- lmer(formula(paste0("threat~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)

omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
texreg::knitreg(list(threat1,threat2.ed,threat2.in,threat3,threat4.fri,threat4.kno),
          omit.coef = omit)
```

### 2022-06-13 production/prod_prep.Rmd
analysis

@@
 
+#reverse education, reference level is the highest level
+elsoc_long_2016_2021$educ <- forcats::fct_rev(elsoc_long_2016_2021$educ)
+
 elsoc_long_2016_2021$educ <- 
@@
 
+#reverse quintile, reference level is the highest quintile
+elsoc_long_2016_2021$quintil <- forcats::fct_rev(elsoc_long_2016_2021$quintil)
+
 elsoc_long_2016_2021$quintil <- 
@@
 
+elsoc_long_2016_2021$quintil1 <- factor(elsoc_long_2016_2021$quintil1, c("Q5","Q4","Q3","Q2","Q1","QNA"))
+
 elsoc_long_2016_2021$quintil1 <- 
@@
 sjmisc::frq(elsoc_long_2016_2021$quintil1)
+
 #Age Ranges
@@
 as.numeric(elsoc_long_2016_2021$d01_01)
-
 sjmisc::frq(elsoc_long_2016_2021$ess)
+elsoc_long_2016_2021$ess <- sjmisc::rec(elsoc_long_2016_2021$ess,rec = "rev")
+sjmisc::frq(elsoc_long_2016_2021$ess)
+
 elsoc_long_2016_2021$ess <- 
 sjlabelled::set_label(x = elsoc_long_2016_2021$ess,
-                      label = "Subjective Social Status: individual")
+                      label = "Subjective Social Status: individual (reverse)")
 
@@
 
-elsoc$nation_t1 <- set_label(elsoc$nation_t1, label = "Nationality")
 elsoc$nation_t1 <- set_labels(elsoc$nation_t1,
@@
                                      "Bolivian", "Haitian","Spanish","Other"))
+elsoc$nation_t1 <- set_label(elsoc$nation_t1, label = "Nationality")
 sjmisc::frq(elsoc$nation_t1)
@@
 sjmisc::frq(elsoc$know_inm_t1t4)
+
+elsoc$know_inm_t1t4 <- factor(elsoc$know_inm_t1t4,levels = c(-1,0,1),
+                              labels = c("No longer know","Still know","Now know"))
+
 elsoc$know_inm_t1t4 <- set_label(elsoc$know_inm_t1t4,
                                   label = "Know immigrants (diff. t4 - t1)")
-# sjmisc::frq(elsoc$know_inm_t1t4)
+sjmisc::frq(elsoc$know_inm_t1t4)
 # sjPlot::tab_xtab(elsoc$know_inm_t1t4,elsoc$ola)
@@
                                 false = elsoc$frien_inm_t1t4)
-summary(elsoc$frien_inm_t1t4)
+sjmisc::frq(elsoc$frien_inm_t1t4)
+
+elsoc$frien_inm_t1t4 <- factor(elsoc$frien_inm_t1t4,
+                               levels = c(-1,0,1),
+                               labels = c("No longer have","Still have","Now have"))
 
@@
 casen2017 %>% 
-  select(comuna,starts_with("r1b"), starts_with("exp"))
+  select(comuna,starts_with("r1"), starts_with("exp"))
+sjmisc::frq(df_17$r1a)
+sjmisc::frq(df_17$r1a_cod)
+sjmisc::frq(df_17$r1b)
+sjmisc::frq(df_17$r1b_p_cod)
+
+# Latam & caribean 401:588
+# recent migrant groups (2020) 
+  # Venezuela: 513
+  # Peru: 509
+  # Haiti: 413
+  # Colombia: 505
+  # Bolivia: 502
+
+# Al igual que la tendencia reflejada en las estimaciones de 2018 y 2019, la mayoría (poco más de 79% del total) de las personas extranjeras residentes en el país al 31 de diciembre de 2020 proviene de Venezuela (30,7%), Perú (16,3%), Haití (12,5%), Colombia (11,4%) y Bolivia (8,5%).
+
+#dummy for inmigrant 
+df_17$pais <- 
+car::recode(df_17$r1b_p_cod,
+            recodes = "c(513,509,413,505,502)=1;c(888,999)=NA;else=0",
+            as.factor = T)
+sjmisc::frq(df_17$pais)
+
+# df_17$r1b[df_17$r1b==9] <- NA
 # sjmisc::frq(df_17$r1b)
 
-#dummy for inmigrant 
-df_17$r1b[df_17$r1b==9] <- NA
-# sjmisc::frq(df_17$r1b)
-df_17$inm <- ifelse(test = (df_17$r1b==3),yes = 1,no = 0)
-# sjmisc::frq(df_17$inm)
+df_17$inm <- ifelse(test = (df_17$pais==1),yes = 1,no = 0)
+
+sjmisc::frq(df_17$inm)
 
@@
   select(comuna,
-         starts_with("r1b"),starts_with("exp"))
-
+         starts_with("r1"),starts_with("exp"))
+
+sjmisc::frq(df_20$r1b_p_cod)
 #dummy for inmigrant 
-df_20$r1b[df_20$r1b==9] <- NA
+df_20$pais <- 
+car::recode(df_20$r1b_p_cod,
+            recodes = "c(513,509,413,505,502)=1;c(888,999)=NA;else=0",
+            as.factor = T)
+sjmisc::frq(df_20$pais)
+
+# df_20$r1b[df_20$r1b==9] <- NA
 # sjmisc::frq(df_20$r1b)
-df_20$inm <- ifelse(test = (df_20$r1b==3),yes = 1,no = 0)
+df_20$inm <- ifelse(test = (df_20$pais==1),yes = 1,no = 0)
 # sjmisc::frq(df_20$inm)
@@
   "Codigo Comuna",
-  "Proporcion de poblacion inmigrante  (ponderado) - CASEN 2017 - Comunal",
-  "Número de poblacion inmigrante - CENSO 2017 - Comunal",
-  "Proporcion de poblacion inmigrante - CENSO 2017 - Comunal",
-  "Número de poblacion inmigrante - CIT 2017 - Comunal (usando zona censal)",
-  "Proporcion de poblacion inmigrante (CIT / DEM)- CIT 2017 y INE/DEM 2017 - Comunal",
-  "Proporcion de poblacion inmigrante proyectada -  INE/DEM 2018 - Comunal",
-  "Proporcion de poblacion inmigrante proyectada -  INE/DEM 2019 - Comunal",
-  "Proporcion de poblacion inmigrante proyectada -  INE/DEM 2020 - Comunal",
-  "Proporcion de poblacion inmigrante (ponderado) - CASEN 2020 - Comunal",
-  "Cambio en la proporción de población inmigrante entre 2017 y 2020 - CASEN")
+  "Proportion of immigrant population (weighted) - CASEN 2017 - Municipality",
+  "Number of immigrant population - CENSUS 2017 - Municipality",
+  "Proportion of immigrant population - CENSUS 2017 - Municipality",
+  "Number of immigrant population - CIT 2017 - Municipality (using census tract)",
+  "Proportion of immigrant population (CIT / DEM)- CIT 2017 and INE/DEM 2017 - Communal",
+  "Proportion of projected immigrant population - INE/DEM 2018 - Communal",
+  "Proportion of projected immigrant population -  INE/DEM 2019 - Comunal",
+  "Proportion of projected immigrant population -  INE/DEM 2020 - Comunal",
+  "Proportion of immigrant population (weighted) - CASEN 2020 - Municipality",
+  "Change in the proportion of the immigrant population between 2017 and 2020 - CASEN")
 

### 2022-06-15 production/prod_analysis-fix-t5.Rmd
update ola cateogorica y % migrant

@@
 ---
-title: "Preparación de datos"
+title: "Analysis"
 date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
@@
 body .main-container {
-      max-width: 1600px !important;
-      width: 1600px !important;
+      max-width: 2000px !important;
+      width: 2000px !important;
     }
@@
 load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
 # Css
@@
 
-# Df variables nivel 1
-# df<- dfSummary(df_study1_long %>% select(in_simpat:edad),
-#                plain.ascii = FALSE,
-#                style = "grid",
-#                tmp.img.dir = "/tmp",
-#                graph.magnif = TRUE,
-#                headings = F,  # encabezado
-#                varnumbers = F, # num variable
-#                labels.col = T, # etiquetas
-#                na.col = F,    # missing
-#                graph.col = T, # plot
-#                valid.col = T, # n valido
-#                col.widths = c(700,10,10,10,10)
-#                )
-# df$Variable <- NULL # delete variable column
-# view(df,
-#      file = here::here("output/tables/desc01.html")
-#      ) # Ver tabla en un archivo HTML
-# webshot::webshot(url =here::here("output/tables/desc01.html"),
-#         file =here::here("output/tables/desc01.png")) # Remover archivo HTML
-# # print(df, method = "render")
-
 # Df variables nivel 1 __________________________________________________________
-df_study1 <- df_study_long_t5_comunas %>% select(in_simpat:edad,
-                                                 inm_prop_cas17_w,
-                                                 inm_prop_cas20_w,dif_inm) 
+df_study1 <- 
+  df_study_long_t5_comunas %>% 
+  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) 
 lab_study1 <- sjlabelled::get_label(df_study1) 
@@
 view(df,
-     file = here::here("output/tables/desc02.html")
+     file = here::here("output/tables/desc01.html")
      ) # Ver tabla en un archivo HTML
-webshot::webshot(url =here::here("output/tables/desc02.html"),
-        file =here::here("output/tables/desc02.png")) # Remover archivo HTML
+webshot::webshot(url =here::here("output/tables/desc01.html"),
+        file =here::here("output/tables/desc01.png")) # Remover archivo HTML
 # print(df, method = "render")
@@
 
-## A. Anxiety
-
-```{r anxiety}
+```{r, results='asis'}
 rm(list=ls())
@@
   df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  select(idencuesta,ola,comuna,cuestion_mig,
+         in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
   na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# A. Anxiety (in_ansied)
-anx.null <- lmer(formula(paste0("in_ansied~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-anx1 <- lmer(formula(paste0("in_ansied~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-anx2.ed <- lmer(formula(paste0("in_ansied~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-anx2.in <- lmer(formula(paste0("in_ansied~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-anx3 <- lmer(formula(paste0("in_ansied~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-anx4.kno <- lmer(formula(paste0("in_ansied~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-anx4.fri <- lmer(formula(paste0("in_ansied~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(anx1,anx2.ed,anx2.in,anx3,anx4.fri,anx4.kno),
-          omit.coef = omit)
-```
-
-## B. Simpathy
-
-```{r simpathy}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# B. Simpathy (in_simpat)
-simp.null <- lmer(formula(paste0("in_simpat~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-simp1 <- lmer(formula(paste0("in_simpat~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simp2.ed <- lmer(formula(paste0("in_simpat~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simp2.in <- lmer(formula(paste0("in_simpat~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simp3 <- lmer(formula(paste0("in_simpat~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simp4.kno <- lmer(formula(paste0("in_simpat~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simp4.fri <- lmer(formula(paste0("in_simpat~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(simp1,simp2.ed,simp2.in,simp3,simp4.fri,simp4.kno),
-          omit.coef = omit)
-```
-
-## C. Family value migrant friends
-
-```{r family-value}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# C. Family value migrant friends (in_simpat)
-valfam.null <- lmer(formula(paste0("in_valfam~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-valfam1 <- lmer(formula(paste0("in_valfam~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valfam2.ed <- lmer(formula(paste0("in_valfam~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valfam2.in <- lmer(formula(paste0("in_valfam~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valfam3 <- lmer(formula(paste0("in_valfam~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valfam4.kno <- lmer(formula(paste0("in_valfam~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valfam4.fri <- lmer(formula(paste0("in_valfam~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(valfam1,valfam2.ed,valfam2.in,valfam3,valfam4.fri,valfam4.kno),
-          omit.coef = omit)
-```
-
-## D. Friend value migrant friends
-
-```{r friends-value}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# D. Friend value migrant friends(in_simpat)
-valami.null <- lmer(formula(paste0("in_valami~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-valami1 <- lmer(formula(paste0("in_valami~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valami2.ed <- lmer(formula(paste0("in_valami~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valami2.in <- lmer(formula(paste0("in_valami~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valami3 <- lmer(formula(paste0("in_valami~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valami4.kno <- lmer(formula(paste0("in_valami~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-valami4.fri <- lmer(formula(paste0("in_valami~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(valami1,valami2.ed,valami2.in,valami3,valami4.fri,valami4.kno),
-          omit.coef = omit)
-```
-
-## E. Immigrants have Chilean friends
-
-```{r imig-clfriends}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# E. Immigrants have Chilean friends 
-amicl.null <- lmer(formula(paste0("in_amigcl~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-
-amicl1 <- lmer(formula(paste0("in_amigcl~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-amicl2.ed <- lmer(formula(paste0("in_amigcl~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-amicl2.in <- lmer(formula(paste0("in_amigcl~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-amicl3 <- lmer(formula(paste0("in_amigcl~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-amicl4.kno <- lmer(formula(paste0("in_amigcl~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-amicl4.fri <- lmer(formula(paste0("in_amigcl~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(amicl1,amicl2.ed,amicl2.in,amicl3,amicl4.fri,amicl4.kno),
-          omit.coef = omit)
-```
-
-## Conviviality (average)
-
-```{r conviviality}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
+  mutate(
+    # conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
+         # identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
+         ola_f = ola,
+         ola=as.numeric(ola),
+         -in_amreal
          )
 
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
+# Hypothesis___________________________________________________________________
+h1.cat <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
 
-# E. Immigrants have Chilean friends 
-conv.null <- lmer(formula(paste0("conviviality~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-conv1 <- lmer(formula(paste0("conviviality~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-conv2.ed <- lmer(formula(paste0("conviviality~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-conv2.in <- lmer(formula(paste0("conviviality~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-conv3 <- lmer(formula(paste0("conviviality~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-conv4.kno <- lmer(formula(paste0("conviviality~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-conv4.fri <- lmer(formula(paste0("conviviality~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+hipotesis <- c(h1.cat,h1,h2.educ,h2.incom,h3,h4.know,h4.friend)
+# Set the independent variables (IVs)
+ivs<- df_study1 %>% select(starts_with("in_"),"conviviality","identity") %>% names()
 
+# Create the object for each model
+models <- list()
+for (i in ivs) {
+models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
+}
+
+# Estimate the multilevel models
+fits <- list()
+for (i in ivs) {
+  for (z in models[[i]][1:length(hipotesis)]) {
+    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
+  }
+}
+fit1 <- fits
+# change model names within list
+names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
+
+# Tables_______________________________________________________________________
 omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(conv1,conv2.ed,conv2.in,conv3,conv4.fri,conv4.kno),
-          omit.coef = omit)
+reorderc <- c(1:4,19,5:18,20:31) # the depends on the number of coeficients (!)
+# detail for subsections:
+# https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
+
+cat(paste0("\n## ","A. Anxiety","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_ansied", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","B. Simpathy ","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_simpat", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","C. Family value migrant friends","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_valfam", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","D. Friend value migrant friends","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_valami", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","E. Immigrants have Chilean friends","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_amigcl", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","Conviviality (average)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+#______________________________________________________________________________
+
+cat(paste0("\n## ","F. Similarity","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_simili", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","G. Loose identity","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_amsimb", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","H. Immigrants keep customs","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_acpier", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","I. Adopt chilean customs","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_acadop", names(fit1))]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","Identity (average)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+#______________________________________________________________________________
+cat(paste0("\n## ","J. Unemployment increases (Threat)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
+        omit.coef = omit,reorder.coef = reorderc)
 ```
-
-## F. Similarity
-
-```{r similarity}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# F. Similarity  
-simi.null <- lmer(formula(paste0("in_simili~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-
-simi1 <- lmer(formula(paste0("in_simili~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simi2.ed <- lmer(formula(paste0("in_simili~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simi2.in <- lmer(formula(paste0("in_simili~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simi3 <- lmer(formula(paste0("in_simili~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simi4.kno <- lmer(formula(paste0("in_simili~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-simi4.fri <- lmer(formula(paste0("in_simili~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(simi1,simi2.ed,simi2.in,simi3,simi4.fri,simi4.kno),
-          omit.coef = omit)
-```
-
-## G. Loose identity
-
-```{r loose-id}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-## G. Loose identity
-loos.null <- lmer(formula(paste0("in_amsimb~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-
-loos1 <- lmer(formula(paste0("in_amsimb~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-loos2.ed <- lmer(formula(paste0("in_amsimb~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-loos2.in <- lmer(formula(paste0("in_amsimb~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-loos3 <- lmer(formula(paste0("in_amsimb~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-loos4.kno <- lmer(formula(paste0("in_amsimb~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-loos4.fri <- lmer(formula(paste0("in_amsimb~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(loos1,loos2.ed,loos2.in,loos3,loos4.fri,loos4.kno),
-          omit.coef = omit)
-```
-
-## H. Immigrants keep customs
-
-```{r keep-custom}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola))
-
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# H. Immigrants keep customs  
-keep.null <- lmer(formula(paste0("in_acpier~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-
-keep1 <- lmer(formula(paste0("in_acpier~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-keep2.ed <- lmer(formula(paste0("in_acpier~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-keep2.in <- lmer(formula(paste0("in_acpier~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-keep3 <- lmer(formula(paste0("in_acpier~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-keep4.kno <- lmer(formula(paste0("in_acpier~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-keep4.fri <- lmer(formula(paste0("in_acpier~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(keep1,keep2.ed,keep2.in,keep3,keep4.fri,keep4.kno),
-          omit.coef = omit)
-```
-
-## I. Adopt chilean customs
-
-```{r}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola))
-
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# I. Adpot chilean customs
-adopt.null <- lmer(formula(paste0("in_acpier~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-adopt1 <- lmer(formula(paste0("in_acpier~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-adopt2.ed <- lmer(formula(paste0("in_acpier~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-adopt2.in <- lmer(formula(paste0("in_acpier~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-adopt3 <- lmer(formula(paste0("in_acpier~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-adopt4.kno <- lmer(formula(paste0("in_acpier~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-adopt4.fri <- lmer(formula(paste0("in_acpier~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(adopt1,adopt2.ed,adopt2.in,adopt3,adopt4.fri,adopt4.kno),
-          omit.coef = omit)
-```
-
-## Identity (average)
-
-```{r identity}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# Identity
-ident.null <- lmer(formula(paste0("identity~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-ident1 <- lmer(formula(paste0("identity~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-ident2.ed <- lmer(formula(paste0("identity~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-ident2.in <- lmer(formula(paste0("identity~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-ident3 <- lmer(formula(paste0("identity~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-ident4.kno <- lmer(formula(paste0("identity~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-ident4.fri <- lmer(formula(paste0("identity~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(ident1,ident2.ed,ident2.in,ident3,ident4.fri,ident4.kno),
-          omit.coef = omit)
-```
-
-## J. Unemployment increases
-
-```{r threat}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,in_simpat:edad,inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  mutate(conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         threat = as.numeric(in_amreal),
-         ola=as.numeric(ola)
-         )
-
-# Hipotesis
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+dif_inm+nation_t1+cuestion_mig"
-
-# J. Unemployment increases
-threat.null <- lmer(formula(paste0("threat~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-threat1 <- lmer(formula(paste0("threat~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-threat2.ed <- lmer(formula(paste0("threat~",h2.educ,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-threat2.in <- lmer(formula(paste0("threat~",h2.incom,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-threat3 <- lmer(formula(paste0("threat~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-threat4.kno <- lmer(formula(paste0("threat~",h4.know,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-threat4.fri <- lmer(formula(paste0("threat~",h4.friend,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(threat1,threat2.ed,threat2.in,threat3,threat4.fri,threat4.kno),
-          omit.coef = omit)
-```

### 2022-06-15 production/prod_prep.Rmd
update ola cateogorica y % migrant

@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_simpat, 
-                      label = "Degree of sympathy for [PER/HAI/VEN] living in Chile") 
+                      label = "Sympathy for [PER/HAI/VEN] living in Chile") 
 elsoc_long_2016_2021$in_simpat <-
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_simili, 
-                      label = "Degree of similarity between Chileans and [PER/HAI/VEN].") 
+                      label = "Similarity between Chileans and [PER/HAI/VEN]") 
 elsoc_long_2016_2021$in_simili <-
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_valfam, 
-                      label = "Degree of agreement: My family values my friends [PER/HAI/VEN].") 
+                      label = "Family value of immigrant friends [PER/HAI/VEN]") 
 sjmisc::frq(elsoc_long_2016_2021$in_valfam)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_valami, 
-                      label = "Degree of agreement: My friends value my friends [PER/HAI/VEN].") 
+                      label = "Friends value of immigrant friends [PER/HAI/VEN]") 
 sjmisc::frq(elsoc_long_2016_2021$in_valami)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_amsimb, 
-                      label = "Degree of agreement: Chile loses its identity with the arrival of [PER/HAI/VEN].") 
+                      label = "Loose of identity because of immigrants") 
 sjmisc::frq(elsoc_long_2016_2021$in_amsimb)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_amreal, 
-                      label = "Degree of agreement: With the arrival of [PER/HAI/VEN] unemployment rises.") 
+                      label = "Unemployment increases") 
 sjmisc::frq(elsoc_long_2016_2021$in_amreal)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_acpier, 
-                      label = "Degree of agreement: [PER/HAI/VEN] keep their customs") 
+                      label = "Immigrant keeping their customs") 
 sjmisc::frq(elsoc_long_2016_2021$in_acpier)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_acadop, 
-                      label = "Degree of agreement: [PER/HAI/VEN] adopt Chilean customs") 
+                      label = "Adoption of Chilean customs") 
 sjmisc::frq(elsoc_long_2016_2021$in_acadop)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_amigcl, 
-                      label = "Degree of agreement: [PER/HAI/VEN] have Chilean friends") 
+                      label = "Immigrants [PER/HAI/VEN] have Chilean friends") 
 sjmisc::frq(elsoc_long_2016_2021$in_amigcl)
+
+# Average variables___________________________________________________________
+
+elsoc_long_2016_2021 <- elsoc_long_2016_2021 %>% 
+  mutate(conviviality=as.numeric((in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5),
+         identity= as.numeric((in_simili+in_amsimb+in_acpier+in_acadop)/4),
+         threat = as.numeric(in_amreal))
+
+sjmisc::frq(elsoc_long_2016_2021$conviviality)
+elsoc_long_2016_2021$conviviality <- 
+sjlabelled::set_label(x = elsoc_long_2016_2021$conviviality, 
+                      label = "Conviviality")
+sjmisc::frq(elsoc_long_2016_2021$conviviality)
+
+sjmisc::frq(elsoc_long_2016_2021$identity)
+elsoc_long_2016_2021$identity <- 
+sjlabelled::set_label(x = elsoc_long_2016_2021$identity, 
+                      label = "Identity")
+sjmisc::frq(elsoc_long_2016_2021$identity)
 
@@
          starts_with("in_"),
+         conviviality,
+         identity,
          educ,educyear,
@@
                                  "^in_acadop_","^in_amigcl_",
+                                 "^conviviality","^identity",
                                  "^educ_","^educyear_","^quintil1_","^ess_","^nation_t1_",
@@
                              "in_acpier","in_acadop","in_amigcl",
+                             "conviviality","identity",
                              "educ","educyear","quintil1","ess","nation_t1",

### 2022-06-17 production/prod_analysis-fix-t5.Rmd
ajustes labels descriptives

@@
                kableExtra,
-               summarytools
+               summarytools,
+               sjlabelled
                )
@@
          conviviality, identity,
-         in_amreal,educ,quintil1:edad,
+         in_amreal,
+         educ,quintil1:edad,
          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm)
+
 lab_study1 <- sjlabelled::get_label(df_study1)
+
+df_study1$in_ansied <- factor(df_study1$in_ansied,1:5,get_labels(df_study1$in_ansied))
+df_study1$in_simpat <- factor(df_study1$in_simpat,1:5,get_labels(df_study1$in_simpat))
+df_study1$in_valfam <- factor(df_study1$in_valfam,1:5,get_labels(df_study1$in_valfam))
+df_study1$in_valami <- factor(df_study1$in_valami,1:5,get_labels(df_study1$in_valami))
+df_study1$in_amigcl <- factor(df_study1$in_amigcl,1:5,get_labels(df_study1$in_amigcl))
+
+df_study1$in_simili <- factor(df_study1$in_simili,1:5,get_labels(df_study1$in_simili))
+df_study1$in_amsimb <- factor(df_study1$in_amsimb,1:5,get_labels(df_study1$in_amsimb))
+df_study1$in_acpier <- factor(df_study1$in_acpier,1:5,get_labels(df_study1$in_acpier))
+df_study1$in_acadop <- factor(df_study1$in_acadop,1:5,get_labels(df_study1$in_acadop))
+
+df_study1$in_amreal <- factor(df_study1$in_amreal,1:5,get_labels(df_study1$in_amreal))
+
 df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)

### 2022-06-17 production/prod_analysis-fix-t5.Rmd
plots interacciones

@@
 
-# Df variables nivel 1 __________________________________________________________
+# Descriptive table __________________________________________________________
 df_study1 <- 
@@
          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) 
-
+# extract variable labels
 lab_study1 <- sjlabelled::get_label(df_study1) 
-
-df_study1$in_ansied <- factor(df_study1$in_ansied,1:5,get_labels(df_study1$in_ansied))
-df_study1$in_simpat <- factor(df_study1$in_simpat,1:5,get_labels(df_study1$in_simpat))
-df_study1$in_valfam <- factor(df_study1$in_valfam,1:5,get_labels(df_study1$in_valfam))
-df_study1$in_valami <- factor(df_study1$in_valami,1:5,get_labels(df_study1$in_valami))
-df_study1$in_amigcl <- factor(df_study1$in_amigcl,1:5,get_labels(df_study1$in_amigcl))
-
-df_study1$in_simili <- factor(df_study1$in_simili,1:5,get_labels(df_study1$in_simili))
-df_study1$in_amsimb <- factor(df_study1$in_amsimb,1:5,get_labels(df_study1$in_amsimb))
-df_study1$in_acpier <- factor(df_study1$in_acpier,1:5,get_labels(df_study1$in_acpier))
-df_study1$in_acadop <- factor(df_study1$in_acadop,1:5,get_labels(df_study1$in_acadop))
-
-df_study1$in_amreal <- factor(df_study1$in_amreal,1:5,get_labels(df_study1$in_amreal))
-
+# convert items to factor
+for (i in names(select(df_study1,starts_with("in_")))) {
+  df_study1[[i]] <- factor(df_study1[[i]],1:5,get_labels(df_study1[[i]]))
+}
+# set variable labels
 df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
@@
 
-```{r, results='asis'}
+```{r, results='asis',fig.dim=c(10,6)}
 rm(list=ls())
@@
                lme4,
-               texreg
+               texreg,
+               sjPlot
                )
@@
 
+plot_model(model = fit1$conviviality3,
+           type = "int")
+
+plot_model(model = fit1$conviviality6,
+           type = "int")
+
+plot_model(model = fit1$conviviality7,
+           type = "int")
+
 #______________________________________________________________________________
@@
 #______________________________________________________________________________
+
 cat(paste0("\n## ","J. Unemployment increases (Threat)","\n"))
@@
         omit.coef = omit,reorder.coef = reorderc)
+
+plot_model(model = fit1$in_amreal3,
+           type = "int")
+
+plot_model(model = fit1$in_amreal5,
+           type = "int")
+
+plot_model(model = fit1$in_amreal6,
+           type = "int")
 ```

### 2022-06-22 production/prod_analysis-fix-t5.Rmd
modelos dummy educacion y 3-way-interac

@@
 
-```{r, results='asis',fig.dim=c(10,6)}
+## Education categorical {-}
+```{r average-educ, results='asis',fig.dim=c(10,6)}
 rm(list=ls())
@@
                texreg,
-               sjPlot
+               sjPlot,
+               ggplot2
                )
 load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+
+old <- theme_set(theme_bw());theme_set(old)
+theme_replace(legend.position="bottom")
 
@@
          ola=as.numeric(ola),
+         primary = car::recode(educ,recodes = "'Primary'=1;else=0",as.factor = T),
          -in_amreal
@@
 h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.fri.time <- "ola*educ*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
 
-hipotesis <- c(h1.cat,h1,h2.educ,h2.incom,h3,h4.know,h4.friend)
+hipotesis <- c(h1.cat,h1,h2.educ,h2.incom,h3,h4.know,h4.friend
+               # h.ed.know.time,
+               # h.ed.fri.time
+               )
 # Set the independent variables (IVs)
@@
 omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-reorderc <- c(1:4,19,5:18,20:31) # the depends on the number of coeficients (!)
+reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
 # detail for subsections:
 # https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
+
+cat(paste0("\n## ","Conviviality (avg)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+        omit.coef = omit,
+        reorder.coef = reorderc
+        )
+
+plot_model(model = fit1$conviviality3,
+           type = "int")
+
+plot_model(model = fit1$conviviality6,
+           type = "int")
+
+plot_model(model = fit1$conviviality7,
+           type = "int")
+
+cat(paste0("\n ","<br>","\n"))
+
+cat(paste0("\n## ","Identity (avg)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+cat(paste0("\n## ","J. Unemployment increases (Threat)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
+        omit.coef = omit,reorder.coef = reorderc)
+
+plot_model(model = fit1$in_amreal3,
+           type = "int")
+
+plot_model(model = fit1$in_amreal5,
+           type = "int")
+
+plot_model(model = fit1$in_amreal6,
+           type = "int")
+```
+
+## Education dummy {-}
+
+```{r average-primary, fig.dim=c(10,6), results='asis'}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg,
+               sjPlot
+               )
+load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+
+old <- theme_set(theme_bw());theme_set(old)
+theme_replace(legend.position="bottom")
+
+# generate analytical sample
+df_study1 <- 
+  df_study_long_t5_comunas %>%
+  select(idencuesta,ola,comuna,cuestion_mig,
+         in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>% 
+  mutate(
+         ola_f = ola,
+         ola=as.numeric(ola),
+         primary = car::recode(educ,recodes = "'Primary'=1;else=0",as.factor = T),
+         -in_amreal
+         )
+
+# Hypothesis___________________________________________________________________
+h1.cat <- "ola_f+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h1 <- "ola+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h2.primary <- "ola*primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h2.incom <- "ola+primary+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h3 <- "ola*dif_inm+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h4.know <- "ola+dif_inm+primary+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h4.friend <- "ola+dif_inm+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.know.time <- "ola*primary*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.fri.time <- "ola*primary*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+hipotesis <- c(h1.cat,h1,h2.primary,h2.incom,h3,h4.know,h4.friend,
+               h.ed.know.time,
+               h.ed.fri.time
+               )
+# Set the independent variables (IVs)
+ivs<- df_study1 %>% select("in_amreal","conviviality","identity") %>% names()
+
+# Create the object for each model
+models <- list()
+for (i in ivs) {
+models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
+}
+
+# Estimate the multilevel models
+fits <- list()
+for (i in ivs) {
+  for (z in models[[i]][1:length(hipotesis)]) {
+    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
+  }
+}
+fit1 <- fits
+# change model names within list
+names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
+
+# Tables_______________________________________________________________________
+omit <- 
+  paste0(
+    "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+    )
+reorderc <- c(1:4,19,5:18,20:43) # the depends on the number of coeficients (!)
+# detail for subsections:
+# https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
+
+cat(paste0("\n## ","Conviviality (avg)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+        omit.coef = omit,
+        single.row = F,
+        reorder.coef = reorderc
+        )
+
+cat(paste0("\n## ","Identity (avg)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
+        omit.coef = omit,
+        reorder.coef = reorderc
+        )
+
+cat(paste0("\n## ","Unemployment increases (Threat)","\n"))
+
+knitreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
+        omit.coef = omit,
+        reorder.coef = reorderc
+        ) 
+```
+
+```{r by-item, fig.dim=c(10,6), results='asis'}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg,
+               sjPlot
+               )
+load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+
+old <- theme_set(theme_bw());theme_set(old)
+theme_replace(legend.position="bottom")
+
+# generate analytical sample
+df_study1 <- 
+  df_study_long_t5_comunas %>%
+  select(idencuesta,ola,comuna,cuestion_mig,
+         in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>% 
+  mutate(
+         ola_f = ola,
+         ola=as.numeric(ola),
+         primary = car::recode(educ,recodes = "'Primary'=1;else=0",as.factor = T),
+         -in_amreal
+         )
+
+# Hypothesis___________________________________________________________________
+h1.cat <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.fri.time <- "ola*educ*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+hipotesis <- c(h1.cat,h1,h2.educ,h2.incom,h3,h4.know,h4.friend
+               # h.ed.know.time,
+               # h.ed.fri.time
+               )
+# Set the independent variables (IVs)
+ivs<- df_study1 %>% select(
+  starts_with("in_")
+  # "conviviality",
+  # "identity"
+  ) %>% names()
+
+# Create the object for each model
+models <- list()
+for (i in ivs) {
+models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
+}
+
+# Estimate the multilevel models
+fits <- list()
+for (i in ivs) {
+  for (z in models[[i]][1:length(hipotesis)]) {
+    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
+  }
+}
+fit1 <- fits
+# change model names within list
+names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
+
+# Tables_______________________________________________________________________
+omit <- 
+  paste0(
+    "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+    )
+reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
+# detail for subsections:
+# https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
+
+cat(paste0("\n# ","Annexes","\n"))
 
@@
 knitreg(l = fit1[names(fit1)[grepl("in_ansied", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
+        omit.coef = omit,reorder.coef = reorderc
+        )
 cat(paste0("\n## ","B. Simpathy ","\n"))
@@
         omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","Conviviality (average)","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-plot_model(model = fit1$conviviality3,
-           type = "int")
-
-plot_model(model = fit1$conviviality6,
-           type = "int")
-
-plot_model(model = fit1$conviviality7,
-           type = "int")
 
@@
         omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","Identity (average)","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-#______________________________________________________________________________
-
-cat(paste0("\n## ","J. Unemployment increases (Threat)","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-plot_model(model = fit1$in_amreal3,
-           type = "int")
-
-plot_model(model = fit1$in_amreal5,
-           type = "int")
-
-plot_model(model = fit1$in_amreal6,
-           type = "int")
 ```

### 2022-06-22 production/prod_prep.Rmd
modelos dummy educacion y 3-way-interac

@@
             know_inm_bi_t1=know_inm_bi[ola==1],
-            know_inm_t1t4 = know_inm_bi[ola==4] - know_inm_bi[ola==1]
+            # know_inm_t1t4 = know_inm_bi[ola==4] - know_inm_bi[ola==1],
+            know_inm_t1t4 = paste0(know_inm_bi[ola==1],know_inm_bi[ola==4])
             )
+sjmisc::frq(know_inm_bi$know_inm_t1t4)
+know_inm_bi$know_inm_t1t4 <- 
+  car::recode(know_inm_bi$know_inm_t1t4,
+              recodes = "c('0NA','NA0','00')='Stable, do not know';c('1NA','NA1','11')='Stable, know';'01'='Now know';'10'='No longer know'",as.factor = T)
+
+know_inm_bi$know_inm_t1t4 <- 
+  factor(know_inm_bi$know_inm_t1t4,
+         levels = c("Stable, do not know","Stable, know","Now know","No longer know"))
+sjmisc::frq(know_inm_bi$know_inm_t1t4)
 
@@
 
-elsoc$know_inm_t1t4 <- factor(elsoc$know_inm_t1t4,levels = c(-1,0,1),
-                              labels = c("No longer know","Still know","Now know"))
-
 elsoc$know_inm_t1t4 <- set_label(elsoc$know_inm_t1t4,
                                   label = "Know immigrants (diff. t4 - t1)")
+
 sjmisc::frq(elsoc$know_inm_t1t4)
@@
             frien_inm_bi_t1=frien_inm_bi[ola==1],
-            frien_inm_t1t4 = frien_inm_bi[ola==4] - frien_inm_bi[ola==1]
+            # frien_inm_t1t4 = frien_inm_bi[ola==4] - frien_inm_bi[ola==1],
+            frien_inm_t1t4 = paste0(frien_inm_bi[ola==1],frien_inm_bi[ola==4])
             )
 
+sjmisc::frq(frien_inm_bi$frien_inm_t1t4)
+frien_inm_bi$frien_inm_t1t4 <-
+  car::recode(frien_inm_bi$frien_inm_t1t4,
+              recodes = "c('0NA','NA0','00')='Stable, do not have friends';c('1NA','NA1','11')='Stable, have friends';'01'='Now have friends';'10'='No longer have friends';'NANA'=NA",as.factor = T)
+sjmisc::frq(frien_inm_bi$frien_inm_t1t4)
+
+frien_inm_bi$frien_inm_t1t4 <- 
+  factor(frien_inm_bi$frien_inm_t1t4,
+         levels = c("Stable, do not have friends","Stable, have friends",
+                    "Now have friends","No longer have friends"))
+sjmisc::frq(frien_inm_bi$frien_inm_t1t4)
+
 elsoc <- left_join(x = elsoc,y = frien_inm_bi, by = "idencuesta")
-summary(elsoc$frien_inm_t1t4)
-# if NA in difference, impute data from t4
-elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
-                                true = elsoc$frien_inm_bi_t4,
-                                false = elsoc$frien_inm_t1t4 )
-summary(elsoc$frien_inm_t1t4)
-# if NA in difference, impute data from t1
-elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
-                                true = elsoc$frien_inm_bi_t1,
-                                false = elsoc$frien_inm_t1t4)
 sjmisc::frq(elsoc$frien_inm_t1t4)
-
-elsoc$frien_inm_t1t4 <- factor(elsoc$frien_inm_t1t4,
-                               levels = c(-1,0,1),
-                               labels = c("No longer have","Still have","Now have"))
-
+# View(elsoc[c("idencuesta","frien_inm_bi_t4","frien_inm_bi_t1","frien_inm_t1t4")])
+
+# # if NA in difference, impute data from t4
+# elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
+#                                 true = elsoc$frien_inm_bi_t4,
+#                                 false = elsoc$frien_inm_t1t4)
+# summary(elsoc$frien_inm_t1t4)
+# # if NA in difference, impute data from t1
+# elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
+#                                 true = elsoc$frien_inm_bi_t1,
+#                                 false = elsoc$frien_inm_t1t4)
+sjmisc::frq(elsoc$frien_inm_t1t4)
+
+# elsoc$frien_inm_t1t4 <- factor(elsoc$frien_inm_t1t4,
+#                                levels = c(-1,0,1),
+#                                labels = c("No longer have",
+#                                           "Constant",
+#                                           "Now have"))
+
+# ref dejamos los 
 elsoc$frien_inm_t1t4 <- 
@@
 sjmisc::frq(elsoc$frien_inm_t1t4)
-# sjPlot::tab_xtab(elsoc$frien_inm_t1t4,elsoc$ola)
-
+sjPlot::tab_xtab(elsoc$frien_inm_bi,elsoc$know_inm_bi)
 # Select variables______________________________________________________________

### 2022-07-15 production/prod_analysis-fix-t5.Rmd
tablas descriptivas individual y municipal

@@
 df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
+df_study1$in_amreal <- sjlabelled::set_label(df_study1$in_amreal,label = "Threat")
 
@@
                valid.col = T, # n valido
-               col.widths = c(700,10,10,10,10,10)
+               # col.widths = c(700,10,10,10,10,10)
                )
@@
         file =here::here("output/tables/desc01.png")) # Remover archivo HTML
+# print(df, method = "render")
+
+individual <- 
+  df_study1 %>% 
+  select(conviviality, identity,
+         in_amreal,
+         educ,quintil1,ess,know_inm_t1t4,frien_inm_t1t4)
+
+df_individual<- dfSummary(x = individual,
+               plain.ascii = FALSE,
+               style = "grid",
+               tmp.img.dir = "/tmp",
+               graph.magnif = TRUE,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = T, # plot
+               valid.col = F,# n valido
+               # max.distinct.values = 15,
+               max.string.width = 50,
+               # col.widths = c(700,10,10,10,10)
+               )
+df_individual$Variable <- NULL # delete variable column
+df_individual <- df_individual %>% rename("Variable"=Label)
+
+view(df_individual,
+     file = here::here("output/tables/ds_individual.html")
+     ) # Ver tabla en un archivo HTML
+webshot::webshot(url =here::here("output/tables/ds_individual.html"),
+        file =here::here("output/tables/ds_individual.png")) # Remover archivo HTML
+# print(df, method = "render")
+
+comunal <- 
+  df_study1 %>% 
+  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl, #conviviality
+         in_simili,in_amsimb,in_acpier,in_acadop, #identity
+         conviviality, identity, 
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  select(inm_prop_cas17_w,inm_prop_cas20_w,dif_inm)
+
+df_municipal<- dfSummary(x = comunal,
+               plain.ascii = FALSE,
+               style = "grid",
+               tmp.img.dir = "/tmp",
+               graph.magnif = TRUE,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = T, # plot
+               valid.col = F,# n valido
+               # max.distinct.values = 15,
+               # max.string.width = 50,
+               # col.widths = c(700,10,10,10,10)
+               )
+
+df_municipal$Variable <- NULL # delete variable column
+df_municipal$`Freqs (% of Valid)` <- NULL # delete variable column
+df_municipal <- df_municipal %>% rename("Variable"=Label)
+
+view(df_municipal,
+     file = here::here("output/tables/ds_municipal.html")
+     ) # Ver tabla en un archivo HTML
+webshot::webshot(url =here::here("output/tables/ds_municipal.html"),
+        file =here::here("output/tables/ds_municipal.png")) # Remover archivo HTML
 # print(df, method = "render")
@@
   title = "\n Wave 5")
+```
+
+```{r items-alpha}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               sjPlot,
+               sjmisc,
+               summarytools,
+               lavaan
+               )
+load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+# Descriptive table __________________________________________________________
+df_study1 <- 
+  df_study_long_t5_comunas %>% 
+  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm)
+
+df_study1 <- df_study1 %>% na.omit()
+
+alpha_convi <- 
+  df_study1 %>% 
+  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl) 
+sjPlot::tab_corr(alpha_convi)
+sjPlot::tab_itemscale(alpha_convi)
+
+alpha_ident <- 
+  df_study1 %>% 
+  select(in_simili,in_amsimb,in_acpier,in_acadop)
+
+sjmisc::frq(alpha_ident)
+sjPlot::tab_corr(alpha_ident)
+sjPlot::tab_itemscale(alpha_ident)
 ```

### 2022-07-21 production/prod_analysis-fix-t5.Rmd
ajuste item loose ident
- descriptivos
- modelos

@@
                summarytools,
-               lavaan
+               lavaan,
+               psych
                )
@@
   df_study1 %>% 
-  select(in_simili,in_amsimb,in_acpier,in_acadop)
+  select(in_simili,
+         in_amsimb, #reverse item
+         in_acpier,in_acadop)
 

### 2022-07-21 production/prod_prep.Rmd
ajuste item loose ident
- descriptivos
- modelos

@@
 sjmisc::frq(elsoc_long_2016_2021$in_amsimb) 
+# Revert item
+elsoc_long_2016_2021$in_amsimb<- 
+  sjmisc::rec(elsoc_long_2016_2021$in_amsimb,rec = "rev")
+sjmisc::frq(elsoc_long_2016_2021$in_amsimb) # check reverted item
 

### 2022-07-30 production/prod_analysis-fix-t5.Rmd
ajustes plots y tablas

@@
 df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
-df_study1$in_amreal <- sjlabelled::set_label(df_study1$in_amreal,label = "Threat")
+df_study1$in_amreal <- sjlabelled::set_label(df_study1$in_amreal,label = "Threat (Unemployment increases)")
 
@@
 
+```{r likert}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+# cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               sjlabelled,
+               ggplot2,
+               ggrepel,
+               cowplot,
+               sjPlot,
+               ggpubr
+)
+load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+
+# Descriptive  __________________________________________________________
+df_study1 <- 
+  df_study_long_t5_comunas %>% 
+  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm)
+
+# para modelar estaba invertido:
+#original: fraseo negativo: mayor acuerdo es "negativo" porque se pierde identidad
+#recodificado: inversion de valores 5 es "menos perdida". Es positivo.
+#ecodificado a sentido original para plot: 5 es "mayor perdida". Es negativo
+sjmisc::frq(df_study1$in_amsimb)
+sjmisc::frq(sjmisc::rec(df_study1$in_amsimb,rec = "rev"))
+df_study1$in_amsimb <- sjmisc::rec(df_study1$in_amsimb,rec = "rev")
+# extract variable labels
+lab_study1 <- sjlabelled::get_label(df_study1)
+# # convert items to factor
+# for (i in names(select(df_study1,starts_with("in_")))) {
+#   df_study1[[i]] <- factor(df_study1[[i]],1:5,get_labels(df_study1[[i]]))
+# }
+# set variable labels
+df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
+df_study1$in_amreal <- sjlabelled::set_label(df_study1$in_amreal,label = "Threat (Unemployment increases)")
+df_study1$in_amsimb <- sjlabelled::set_label(df_study1$in_amsimb,label = "Loose of identity because of migrants")
+
+df_study1 <- 
+  df_study1 %>% 
+  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         in_amreal)
+# view_df(df_study1)
+
+# se invierten todos los items para el plot, NO SU SENTIDO  para modelamiento
+for (i in names(df_study1)) {
+  df_study1[[i]] <- sjmisc::rec(x = df_study1[[i]],rec = "rev")  
+}
+
+  # Declare ggplot2 features
+  ggplot2::theme_set(ggplot2::theme(panel.background = ggplot2::element_rect(fill = "gray85",
+                                                                             colour = "gray85"),
+                                    # panel.border = ggplot2::element_blank(),
+                                    axis.text.y = ggplot2::element_text(size = 13,
+                                                                        hjust = 1),
+                                    title = ggplot2::element_text(size = 18,
+                                                                  face = "bold"),
+                                    legend.text = ggplot2::element_text(size = 12),
+                                    plot.caption = ggplot2::element_text(size = 10,
+                                                                         face = "plain",
+                                                                         hjust = 1
+                                                                         )))
+#
+
+#Plot 1: Conviality-----------------------------------------------------------
+p1<-
+  dplyr::select(df_study1,in_ansied,in_simpat,in_valfam,in_valami,in_amigcl) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      title = c("Conviviality"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE) +
+  ggplot2::theme(legend.position = "none")
+
+#Plot 2: Identity-------------------------------------------------------------
+p2<-
+  dplyr::select(df_study1,in_simili,in_amsimb,in_acpier,in_acadop) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      title = c("Identity"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE)  +
+  ggplot2::theme(legend.position = "none")
+
+#Plot 3: Threat---------------------------------------------------------------
+p3<-
+  dplyr::select(df_study1,in_amreal) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      # title = c("Threat"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE)  +
+  ggplot2::theme(legend.position = "none")
+
+#Legend: Extract legend for the final plot using ggpubr
+leg<- ggpubr::as_ggplot(
+  ggpubr::get_legend(
+    dplyr::select(df_study1,in_amreal) %>%
+      sjPlot::plot_likert(geom.colors = "PuBu",
+                          catcount = 4,
+                          cat.neutral = 3) +
+      ggplot2::theme(legend.position = "bottom") +
+      ggplot2::scale_fill_manual(
+        values = c("#b3b3b3ff", "#f1eef6ff", "#bdc9e1ff","#74a9cfff","#0570b0ff"),
+        breaks=c("neutral",4,3,2,1),
+        labels=c(
+          "Neither comfortable nor uncomfortable\nSomewhat\nSomewhat similar\nNeither agree nor disagree\n",
+          "Very uncomfortable\nVery little or not at all\nNot similar\nStrongly disagree",
+          "Uncomfortable\nA little\nSlightly similar\nDisagree",
+          "Comfortable\nQuite a lot\nQuite similar\nAgree",
+          "Very comfortable\nA lot\nVery similar\nTotally agree")
+      )
+  )
+)
+
+# plot individuales-----------------------------------------------------------
+colors<- c("#b3b3b3ff","#f1eef6ff","#bdc9e1ff","#74a9cfff","#0570b0ff")
+p_ansied<-
+  dplyr::select(df_study1,in_ansied) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      title = c("Convivality"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE) + 
+  ggplot2::theme(legend.position = "bottom") +
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c("neutral", 4, 3, 2, 1),
+    labels = c(
+      "Neither comfortable nor uncomfortable",      
+      "Very uncomfortable",
+      "Uncomfortable",
+      "Comfortable",
+      "Very comfortable"
+    )
+  )
+
+p_simpat<-
+  dplyr::select(df_study1,in_simpat) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      # title = c("Threat"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE)  +
+  ggplot2::theme(legend.position = "bottom") +
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c("neutral", 4, 3, 2, 1),
+    labels = c(
+      "Somewhat ",      
+      "Very little or not at all",
+      "A little",
+      "Quite a lot",
+      "A lot"
+    )
+  )
+
+p_convi<-
+  dplyr::select(df_study1,in_valfam,in_valami,in_amigcl) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      # title = c("Threat"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE)  +
+  ggplot2::theme(legend.position = "bottom")+
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c("neutral", 4, 3, 2, 1),
+    labels = c(
+      "Disagree Neither disagree nor agree",      
+      "Strongly disagree",
+      "Disagree",
+      "Agree",
+      "Strongly agree"
+    )
+  )
+
+# Plot individual: IDENTITY -----------------------------------------------
+p_simili<-
+  dplyr::select(df_study1,in_simili) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      title = c("Identity"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE)  +
+  ggplot2::theme(legend.position = "bottom")+
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c("neutral", 4, 3, 2, 1),
+    labels = c(
+      "Somewhat similar",      
+      "Not similiar",
+      "Slightly similar",
+      "Quite similar",
+      "Very similar"
+    )
+  )
+
+p_identi<-
+  dplyr::select(df_study1,in_amsimb,in_acpier,in_acadop,
+                in_amsimb:in_amreal) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      # title = c("Threat"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE)  +
+  ggplot2::theme(legend.position = "bottom")+
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c("neutral", 4, 3, 2, 1),
+    labels = c(
+      "Disagree Neither disagree nor agree",      
+      "Strongly disagree",
+      "Disagree",
+      "Agree",
+      "Strongly agree"
+    )
+  )
+
+# Plot individual: THREAT -----------------------------------------------
+p_threat<-
+    dplyr::select(df_study1,in_amreal) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      title = c("Threat"),
+                      geom.size = 0.8,
+                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
+                      catcount = 4,
+                      cat.neutral = 3,
+                      grid.range  =  c (1.2 , 1.2),
+                      values  =  "sum.outside",
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE)  +
+  ggplot2::theme(legend.position = "bottom") +
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c("neutral", 4, 3, 2, 1),
+    labels = c(
+      "Disagree Neither disagree nor agree",      
+      "Strongly disagree",
+      "Disagree",
+      "Agree",
+      "Strongly agree"
+    )
+  )
+
+# create plot-------------------------------------------------------------------
+if (!require("pacman")) install.packages("pacman")
+pacman::p_load(cowplot)
+s <- 0.35
+m <- 0.6
+l <- 0.7
+#Final plot
+likert1<-   
+  cowplot::add_sub(
+    cowplot::plot_grid(p_ansied,p_simpat,p_convi,p_simili,p_identi,p_threat, 
+                       align = "hv",
+                       axis = "a",
+                       nrow = 6, #4 plots
+                       ncol = 1, #1 column
+                       rel_heights = c(s,s,m,s,l,s) #fig size within plot
+    ),
+    label = paste0("Source: Own elaboration based on ELSOC Survey",
+                   " (n=",dim(na.omit(df_study1))[1],")"
+    ),        # Caption: source and sample size
+    size = 12,  # Caption font size
+    hjust = -0.25# Caption align
+  ) %>%
+  cowplot::ggdraw()  
+# likert1
+ggplot2::ggsave(filename = "plot_likert2.png",
+                plot = likert1,
+                path = here::here("output/images"),
+                device = "png",
+                units = "cm",
+                scale = 3.5,
+                width = 10,height = 10)
+pacman::p_load(cowplot)
+
+#Final plot
+likert1<-   
+  cowplot::add_sub(
+    cowplot::plot_grid(p_a,p_b,p_c,p_d, #plot 1,2,3 and legend (bottom)
+                       align = "hv",
+                       axis = "a",
+                       nrow = 4, #4 plots
+                       ncol = 1, #1 column
+                       rel_heights = c(0.25,0.25,0.25,0.90) #fig size within plot
+    ),
+    label = paste0("Source: Own elaboration based on ELSOC Survey",
+                   " (n=",dim(na.omit(df_study1))[1],")"
+    ),        # Caption: source and sample size
+    size = 12,  # Caption font size
+    hjust = -0.25# Caption align
+  ) %>%
+  cowplot::ggdraw()  
+likert1
+ggplot2::ggsave(filename = "plot_likert.png",
+                plot = likert1,
+                path = here::here("output/images"),
+                device = "png",
+                units = "cm",
+                scale = 3,
+                width = 10,height = 8)
+```
+
 ## Correlation
@@
   # filter(ola==1) %>%
+    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>% 
   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
@@
 
+# dim(mat_full)
+
 cor_full <- as.matrix(mat_full$rho)
@@
 # Set Row names of the matrix
-rownames(cor_full) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
-                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/HAI/VEN]",
-                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
-                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/HAI/VEN]",
-                     "G. Loose of identity because of immigrants ",
-                     "H. Immigrant keeping their customs",
+rownames(cor_full) <- c("A. Interaction anxiety with [PER/VEN]",
+                     "B. Sympathy for [PER/VEN] living in Chile",
+                     "C. Family value of migrant friends [PER/VEN]",
+                     "D. Friends value of migrant friends [PER/VEN]",
+                     "E. Migrants [PER/VEN] have Chilean friends",
+                     "F. Similarity between Chileans and [PER/VEN]",
+                     "G. Loose of identity because of migrants (reverse) ",
+                     "H. Migrant keeping their customs",
                      "I. Adoption of Chilean customs",
@@
 
+png(file=here::here("output/images/plot_corr.png"),width=900, height=700)
+corrplot::corrplot(cor_full,
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  # title = "\n Full sample"
+  )
+dev.off()
+
 # Wave 1_______________________________________________________________________
@@
   filter(ola==1) %>%
+    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>%   
   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
@@
 # Set Row names of the matrix
-rownames(cor_1) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
-                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/HAI/VEN]",
-                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
-                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/HAI/VEN]",
-                     "G. Loose of identity because of immigrants ",
+rownames(cor_1) <- c("A. Interaction anxiety with [PER/VEN]",
+                     "B. Sympathy for [PER/VEN] living in Chile",
+                     "C. Family value of immigrant friends [PER/VEN]",
+                     "D. Friends value of immigrant friends [PER/VEN]",
+                     "E. Immigrants [PER/VEN] have Chilean friends",
+                     "F. Similarity between Chileans and [PER/VEN]",
+                     "G. Loose of identity because of immigrants (reverse) ",
                      "H. Immigrant keeping their customs",
@@
   filter(ola==2) %>% 
+    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>%   
   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
@@
 # Set Row names of the matrix
-rownames(cor_2) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
-                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/HAI/VEN]",
-                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
-                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/HAI/VEN]",
-                     "G. Loose of identity because of immigrants ",
+rownames(cor_2) <- c("A. Interaction anxiety with [PER/VEN]",
+                     "B. Sympathy for [PER/VEN] living in Chile",
+                     "C. Family value of immigrant friends [PER/VEN]",
+                     "D. Friends value of immigrant friends [PER/VEN]",
+                     "E. Immigrants [PER/VEN] have Chilean friends",
+                     "F. Similarity between Chileans and [PER/VEN]",
+                     "G. Loose of identity because of immigrants (reverse) ",
                      "H. Immigrant keeping their customs",
@@
   filter(ola==3) %>% 
+    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>%   
   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
@@
 # Set Row names of the matrix
-rownames(cor_3) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
-                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/HAI/VEN]",
-                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
-                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/HAI/VEN]",
-                     "G. Loose of identity because of immigrants ",
+rownames(cor_3) <- c("A. Interaction anxiety with [PER/VEN]",
+                     "B. Sympathy for [PER/VEN] living in Chile",
+                     "C. Family value of immigrant friends [PER/VEN]",
+                     "D. Friends value of immigrant friends [PER/VEN]",
+                     "E. Immigrants [PER/VEN] have Chilean friends",
+                     "F. Similarity between Chileans and [PER/VEN]",
+                     "G. Loose of identity because of immigrants (reverse) ",
                      "H. Immigrant keeping their customs",
@@
   filter(ola==4) %>% 
+    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>%   
   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
@@
 # Set Row names of the matrix
-rownames(cor_4) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
-                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/HAI/VEN]",
-                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
-                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/HAI/VEN]",
-                     "G. Loose of identity because of immigrants ",
+rownames(cor_4) <- c("A. Interaction anxiety with [PER/VEN]",
+                     "B. Sympathy for [PER/VEN] living in Chile",
+                     "C. Family value of immigrant friends [PER/VEN]",
+                     "D. Friends value of immigrant friends [PER/VEN]",
+                     "E. Immigrants [PER/VEN] have Chilean friends",
+                     "F. Similarity between Chileans and [PER/VEN]",
+                     "G. Loose of identity because of immigrants (reverse) ",
                      "H. Immigrant keeping their customs",
@@
   filter(ola==5) %>% 
+    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>%   
   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
@@
 # Set Row names of the matrix
-rownames(cor_5) <- c("A. Interaction anxiety with [PER/HAI/VEN]",
-                     "B. Sympathy for [PER/HAI/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/HAI/VEN]",
-                     "D. Friends value of immigrant friends [PER/HAI/VEN]",
-                     "E. Immigrants [PER/HAI/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/HAI/VEN]",
-                     "G. Loose of identity because of immigrants ",
+rownames(cor_5) <- c("A. Interaction anxiety with [PER/VEN]",
+                     "B. Sympathy for [PER/VEN] living in Chile",
+                     "C. Family value of immigrant friends [PER/VEN]",
+                     "D. Friends value of immigrant friends [PER/VEN]",
+                     "E. Immigrants [PER/VEN] have Chilean friends",
+                     "F. Similarity between Chileans and [PER/VEN]",
+                     "G. Loose of identity because of immigrants (reverse) ",
                      "H. Immigrant keeping their customs",
@@
 
-sjmisc::frq(alpha_ident)
+# sjmisc::frq(alpha_ident)
 sjPlot::tab_corr(alpha_ident)
@@
 # Hypothesis___________________________________________________________________
-h1.cat <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+m1.cat <- "ola_f+pos_id+edad+nation_t1+cuestion_mig"
+m2 <- "ola_f+educ+quintil1+ess+pos_id+edad+nation_t1+cuestion_mig"
+m3 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+nation_t1+cuestion_mig"
+m4 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+inm_prop_cas20_w+dif_inm+pos_id+edad+nation_t1+cuestion_mig"
+
+# h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+m5 <- "educ*ola+quintil1*ola+ess+know_inm_t1t4*ola+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm*ola+nation_t1+cuestion_mig"
+
 h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
@@
 
-hipotesis <- c(h1.cat,h1,h2.educ,h2.incom,h3,h4.know,h4.friend
+hipotesis <- c(m1.cat,m2,m3,m4
                # h.ed.know.time,
@@
 omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
+# reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
 # detail for subsections:
@@
 
+coef_names <- c("Wave 2017 (ref: Wave 2016)","Wave 2018", "Wave 2019","Wave 2020",
+                "Technical (ref: Universitary)","High school","Primary",
+                "Quintile 4 (ref: Quintile 5)", "Quintile 3","Quintile 2","Quintile 1",
+                "Subjective Social Status",
+                "Stable, know (ref: Stable, do not know)","Now know","No longer know",
+                "Stable, have friends (ref: Stable, do not have friends)","Now have friends","No longer have friends",
+                "Propotion of migrants",
+                "Change in the proportion of migrants")
+
 cat(paste0("\n## ","Conviviality (avg)","\n"))
 
-knitreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+caption_conv <- "Multilevel linear regression models for Conviviality"
+htmlreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+        omit.coef = omit,single.row = T,
+        caption = caption_conv,
+        caption.above = T,
+        custom.coef.names=coef_names,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        file = here::here("output/tables/tab_conv.html")
+        # reorder.coef = reorderc
+        )
+webshot::webshot(url = here::here("output/tables/tab_conv.html"),
+                 file = here::here("output/tables/tab_conv.png"))
+
+plot_model(fit1$conviviality1,type = "pred",terms = " ola_f")
+
+cat(paste0("\n ","<br>","\n"))
+
+cat(paste0("\n## ","Identity (avg)","\n"))
+
+caption_id <- "Multilevel linear regression models for Identity"
+htmlreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
         omit.coef = omit,
-        reorder.coef = reorderc
-        )
-
-plot_model(model = fit1$conviviality3,
-           type = "int")
-
-plot_model(model = fit1$conviviality6,
-           type = "int")
-
-plot_model(model = fit1$conviviality7,
-           type = "int")
-
-cat(paste0("\n ","<br>","\n"))
-
-cat(paste0("\n## ","Identity (avg)","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
-        omit.coef = omit,reorder.coef = reorderc)
+        caption = caption_id,
+        caption.above = T,
+        single.row = T,
+        custom.coef.names=coef_names,
+        # reorder.coef = reorderc
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        file = here::here("output/tables/tab_iden.html")        
+        )
+webshot::webshot(url = here::here("output/tables/tab_iden.html"),
+                 file = here::here("output/tables/tab_iden.png"))
+
+plot_model(fit1$identity1,type = "pred",terms = "ola_f")
 
@@
 
-knitreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-plot_model(model = fit1$in_amreal3,
-           type = "int")
-
-plot_model(model = fit1$in_amreal5,
-           type = "int")
-
-plot_model(model = fit1$in_amreal6,
-           type = "int")
+caption_th <- "Multilevel linear regression models for Threat"
+htmlreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
+        omit.coef = omit,single.row = T,
+        caption = caption_th,
+        caption.above = T,
+        custom.coef.names=coef_names,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        file = here::here("output/tables/tab_thr.html")         
+        # reorder.coef = reorderc
+        ) 
+webshot::webshot(url = here::here("output/tables/tab_thr.html"),
+                 file = here::here("output/tables/tab_thr.png"))
+
+plot_model(fit1$in_amreal1,type = "pred",terms = " ola_f")
+
+cat(paste0("\n## ","Interactions","\n"))  #------------------------------------#
+
+hipotesis_int <- c(m5
+               # h.ed.know.time,
+               # h.ed.fri.time
+               )
+
+# Create the object for each model
+models_int <- list()
+for (i in ivs) {
+models_int[[i]] <- paste0(i,"~",hipotesis_int,"+(1|idencuesta)+(1|comuna)")  
+}
+
+# Estimate the multilevel models
+fit_int <- list()
+for (i in ivs) {
+  for (z in models_int[[i]][1:length(hipotesis_int)]) {
+    fit_int[[z]] <- lme4::lmer(formula = z, data = df_study1)
+  }
+}
+
+names(fit_int) <- paste0(rep(ivs,each=length(hipotesis_int)),1:length(hipotesis_int))
+
+omit_int <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)|(ess)|(inm_prop_cas20_w)"
+coef_int <- c("Technical (ref: Universitary) ","High school","Primary","Wave",
+              "Quintile 4 (ref: Quintile 5)", "Quintile 3","Quintile 2","Quintile 1",
+              # "Subjective Social Status",
+              "Stable, know (ref: Stable, do not know)","Now know","No longer know",
+              "Stable, have friends (ref: Stable, do not have friends)","Now have friends","No longer have friends",
+              # "Propotion of migrants",
+              "Change in the proportion of migrants",
+              #----------------------------------------------------------------#
+              "Technical x Wave","High school x Wave","Primary x Wave",
+              "Quintile 4 x Wave", "Quintile 3 x Wave","Quintile 2 x Wave","Quintile 1 x Wave",
+              "Stable, know  x Wave","Now know x Wave","No longer know x Wave",
+              "Stable, have friends x Wave","Now have friends x Wave","No longer have friends x Wave",
+              "Change in the proportion of migrants  x Wave"              
+              )
+
+caption_int <- "Multilevel linear regression models with interactions"
+htmlreg(l = fit_int[c("conviviality1","identity1","in_amreal1")],
+        omit.coef = omit_int,
+        single.row = T,
+        custom.coef.names=coef_int,
+        reorder.coef = c(4,1:3,5:29),
+        groups = list(
+          # "Education (ref: Universitary)"=2:4,
+                      # "Household Income Quintile (ref: Quintile 5)"=5:8,
+                      # "Know migrant (ref: Stable, do not know)" =9:11,
+                      # "Friendship with migrants (ref: Stable, do not have friends)" =12:14,
+                      "Education x Wave" =16:18,
+                      "Quintile x Wave" =19:22,
+                      "Know migrant x Wave" =23:25,
+                      "Migrant friend x Wave" =26:28,
+                      "Migrant Propulation x Wave" =29
+                      ),
+        caption = caption_int,
+        caption.above = T,
+        custom.model.names = c("Model 1\n(Conviviality)","Model 2\n(Identity)","Model 3\n(Threat)"),
+        file = here::here("output/tables/tab_int.html")
+        )
+webshot::webshot(url = here::here("output/tables/tab_int.html"),
+                 file = here::here("output/tables/tab_int.png"))
 ```
@@
 
-```{r average-primary, fig.dim=c(10,6), results='asis'}
+```{r average-primary, eval=FALSE, fig.dim=c(10,6), include=FALSE, results='asis'}
 rm(list=ls())

### 2022-07-30 production/prod_prep.Rmd
ajustes plots y tablas

--- 2022-06-22 production/prod_prep.Rmd
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_simpat, 
-                      label = "Sympathy for [PER/HAI/VEN] living in Chile") 
+                      label = "Sympathy for [PER/VEN] living in Chile") 
 elsoc_long_2016_2021$in_simpat <-
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_simili, 
-                      label = "Similarity between Chileans and [PER/HAI/VEN]") 
+                      label = "Similarity between Chileans and [PER/VEN]") 
 elsoc_long_2016_2021$in_simili <-
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_ansied, 
-                      label = "Interaction anxiety with [PER/HAI/VEN].") 
+                      label = "Interaction anxiety with [PER/VEN].") 
 elsoc_long_2016_2021$in_ansied <-
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_valfam, 
-                      label = "Family value of immigrant friends [PER/HAI/VEN]") 
+                      label = "Family value of migrant friends [PER/VEN]") 
 sjmisc::frq(elsoc_long_2016_2021$in_valfam)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_valami, 
-                      label = "Friends value of immigrant friends [PER/HAI/VEN]") 
+                      label = "Friends value of migrant friends [PER/VEN]") 
 sjmisc::frq(elsoc_long_2016_2021$in_valami)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_amsimb, 
-                      label = "Loose of identity because of immigrants") 
+                      label = "Loose of identity because of migrants (reverse)")
 sjmisc::frq(elsoc_long_2016_2021$in_amsimb) 
+# Revert item
+elsoc_long_2016_2021$in_amsimb<- 
+  sjmisc::rec(elsoc_long_2016_2021$in_amsimb,rec = "rev")
+sjmisc::frq(elsoc_long_2016_2021$in_amsimb) # check reverted item
 
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_acpier, 
-                      label = "Immigrant keeping their customs") 
+                      label = "Migrant keeping their customs") 
 sjmisc::frq(elsoc_long_2016_2021$in_acpier)
@@
 sjlabelled::set_label(x = elsoc_long_2016_2021$in_amigcl, 
-                      label = "Immigrants [PER/HAI/VEN] have Chilean friends") 
+                      label = "migrants [PER/VEN] have Chilean friends") 
 sjmisc::frq(elsoc_long_2016_2021$in_amigcl)
@@
 
-# Known: Immigrant [PER/HAI/VEN]____________________________
+# Known: Immigrant [PER/VEN]____________________________
 
@@
 elsoc$know_inm_t1t4 <- set_label(elsoc$know_inm_t1t4,
-                                  label = "Know immigrants (diff. t4 - t1)")
+                                  label = "Know migrants (diff. t4 - t1)")
 
@@
   set_label(elsoc$frien_inm_t1t4,
-            label = "Have immigrant friends (diff. t4 - t1)")
+            label = "Have migrant friends (diff. t4 - t1)")
 
@@
   "Codigo Comuna",
-  "Proportion of immigrant population (weighted) - CASEN 2017 - Municipality",
-  "Number of immigrant population - CENSUS 2017 - Municipality",
-  "Proportion of immigrant population - CENSUS 2017 - Municipality",
-  "Number of immigrant population - CIT 2017 - Municipality (using census tract)",
-  "Proportion of immigrant population (CIT / DEM)- CIT 2017 and INE/DEM 2017 - Communal",
-  "Proportion of projected immigrant population - INE/DEM 2018 - Communal",
-  "Proportion of projected immigrant population -  INE/DEM 2019 - Comunal",
-  "Proportion of projected immigrant population -  INE/DEM 2020 - Comunal",
-  "Proportion of immigrant population (weighted) - CASEN 2020 - Municipality",
-  "Change in the proportion of the immigrant population between 2017 and 2020 - CASEN")
+  "Proportion of migrant population (weighted) - CASEN 2017 - Municipality",
+  "Number of migrant population - CENSUS 2017 - Municipality",
+  "Proportion of migrant population - CENSUS 2017 - Municipality",
+  "Number of migrant population - CIT 2017 - Municipality (using census tract)",
+  "Proportion of migrant population (CIT / DEM)- CIT 2017 and INE/DEM 2017 - Communal",
+  "Proportion of projected migrant population - INE/DEM 2018 - Communal",
+  "Proportion of projected migrant population -  INE/DEM 2019 - Comunal",
+  "Proportion of projected migrant population -  INE/DEM 2020 - Comunal",
+  "Proportion of migrant population (weighted) - CASEN 2020 - Municipality",
+  "Change in the proportion of the migrant population between 2017 and 2020 - CASEN")
 

### 2022-08-02 production/prod_analysis-fix-t5.Rmd
submission

@@
 webshot::webshot(url =here::here("output/tables/ds_individual.html"),
-        file =here::here("output/tables/ds_individual.png")) # Remover archivo HTML
+        file =here::here("output/tables/table2.jpg")) # Remover archivo HTML
 # print(df, method = "render")
@@
 webshot::webshot(url =here::here("output/tables/ds_municipal.html"),
-        file =here::here("output/tables/ds_municipal.png")) # Remover archivo HTML
+        file =here::here("output/tables/table3.jpg")) # Remover archivo HTML
 # print(df, method = "render")
@@
                                                                              colour = "gray85"),
-                                    # panel.border = ggplot2::element_blank(),
+                                    panel.border = ggplot2::element_blank(),
                                     axis.text.y = ggplot2::element_text(size = 13,
@@
                                                                          hjust = 1
-                                                                         )))
+                                                                         )
+                                    )
+                     )
 #
@@
     )
-  )
+  ) +
+  ggplot2::geom_label(label.padding = 1, label.r = 1,label.size = 0.1)
+
+p_ansied[["layers"]]
 
@@
     labels = c(
-      "Disagree Neither disagree nor agree",      
+      "Neither disagree nor agree",      
       "Strongly disagree",
@@
   dplyr::select(df_study1,in_amsimb,in_acpier,in_acadop,
-                in_amsimb:in_amreal) %>%
+                in_amsimb) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
@@
 #Final plot
+title <- cowplot::ggdraw() + 
+  cowplot::draw_label(
+    "Figure 1: Frequencies for Conviviality, Identity and Threat items",
+    fontface = 'bold',
+    size = 15,
+    x = 0,
+    hjust = 0
+  )
+
 likert1<-   
   cowplot::add_sub(
-    cowplot::plot_grid(p_ansied,p_simpat,p_convi,p_simili,p_identi,p_threat, 
+    cowplot::plot_grid(title,p_ansied,p_simpat,p_convi,p_simili,p_identi,p_threat, 
                        align = "hv",
@@
                        ncol = 1, #1 column   
-                       rel_heights = c(s,s,m,s,l,s) #fig size within plot
+                       rel_heights = c(0.1,s,s,m,s,m,s) #fig size within plot
     ),
@@
 # likert1
-ggplot2::ggsave(filename = "plot_likert2.png",
+ggplot2::ggsave(filename = "figure1.jpg",
                 plot = likert1,
                 path = here::here("output/images"),
-                device = "png",
+                device = "jpg",
                 units = "cm",
@@
 likert1
-ggplot2::ggsave(filename = "plot_likert.png",
+ggplot2::ggsave(filename = "figure1.jpg",
                 plot = likert1,
                 path = here::here("output/images"),
-                device = "png",
+                device = "jpg",
                 units = "cm",
@@
 
-png(file=here::here("output/images/plot_corr.png"),width=900, height=700)
+png(file=here::here("output/images/figure2.jpg"),width=900, height=700)
 corrplot::corrplot(cor_full,
+  title = "\nFigure 2: Polychoric correlation matrix for Conviviality, Identity and Threat items",
   method = "color",

### 2022-08-02 production/prod_prep.Rmd
submission

@@
   factor(elsoc_long_2016_2021$educ,
-         labels = c("Primary","High school","Vocational","Universitary"))
+         labels = c("Primary","High school","Technical","Universitary"))
 

### 2022-08-02 production/prod_analysis-fix-t5.Rmd
submission

@@
 pacman::p_load(cowplot)
-s <- 0.35
+s <- 0.4
 m <- 0.6
-l <- 0.7
+l <- 0.8
 #Final plot
@@
                        axis = "a",
-                       nrow = 6, #4 plots
+                       nrow = 7, #4 plots
                        ncol = 1, #1 column   
-                       rel_heights = c(0.1,s,s,m,s,m,s) #fig size within plot
+                       rel_heights = c(s,s,0.35,m,s,m,s) #fig size within plot
     ),
@@
                 width = 10,height = 10)
-pacman::p_load(cowplot)
 

### 2022-10-24 production/prod_analysis-fix-t5.Rmd
ajustes likertplot

@@
          educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) 
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm,sexo,edad) 
 # extract variable labels
@@
 view(df,
-     file = here::here("output/tables/desc01.html")
+     # file = here::here("output/tables/desc01.html")
      ) # Ver tabla en un archivo HTML
@@
          in_amreal,
-         educ,quintil1,ess,know_inm_t1t4,frien_inm_t1t4)
+         educ,quintil1,ess,know_inm_t1t4,frien_inm_t1t4,edad,sexo)
 
@@
 df_study1$in_amsimb <- sjlabelled::set_label(df_study1$in_amsimb,label = "Loose of identity because of migrants")
+df_study1$in_amigcl <- sjlabelled::set_label(df_study1$in_amigcl,label = "Migrants [PER/VEN] have Chilean friends")
 
@@
   # Declare ggplot2 features
-  ggplot2::theme_set(ggplot2::theme(panel.background = ggplot2::element_rect(fill = "gray85",
+  ggplot2::theme_set(
+    ggplot2::theme(
+      panel.background = ggplot2::element_rect(fill = "gray85",
                                                colour = "gray85"),
@@
                                            face = "plain",
-                                                                         hjust = 1
-                                                                         )
+                                           hjust = 1),
+      axis.ticks.x=element_blank(),
+      axis.text.x=element_blank(),
     )
@@
 
-#Plot 1: Conviality-----------------------------------------------------------
-p1<-
-  dplyr::select(df_study1,in_ansied,in_simpat,in_valfam,in_valami,in_amigcl) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      title = c("Conviviality"),
-                      geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE) +
-  ggplot2::theme(legend.position = "none")
-
-#Plot 2: Identity-------------------------------------------------------------
-p2<-
-  dplyr::select(df_study1,in_simili,in_amsimb,in_acpier,in_acadop) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      title = c("Identity"),
-                      geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "none")
-
-#Plot 3: Threat---------------------------------------------------------------
-p3<-
-  dplyr::select(df_study1,in_amreal) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
-                      geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "none")
-
-#Legend: Extract legend for the final plot using ggpubr
-leg<- ggpubr::as_ggplot(
-  ggpubr::get_legend(
-    dplyr::select(df_study1,in_amreal) %>%
-      sjPlot::plot_likert(geom.colors = "PuBu",
-                          catcount = 4,
-                          cat.neutral = 3) +
-      ggplot2::theme(legend.position = "bottom") +
-      ggplot2::scale_fill_manual(
-        values = c("#b3b3b3ff", "#f1eef6ff", "#bdc9e1ff","#74a9cfff","#0570b0ff"),
-        breaks=c("neutral",4,3,2,1),
-        labels=c(
-          "Neither comfortable nor uncomfortable\nSomewhat\nSomewhat similar\nNeither agree nor disagree\n",
-          "Very uncomfortable\nVery little or not at all\nNot similar\nStrongly disagree",
-          "Uncomfortable\nA little\nSlightly similar\nDisagree",
-          "Comfortable\nQuite a lot\nQuite similar\nAgree",
-          "Very comfortable\nA lot\nVery similar\nTotally agree")
-      )
-  )
-)
-
 # plot individuales-----------------------------------------------------------
-colors<- c("#b3b3b3ff","#f1eef6ff","#bdc9e1ff","#74a9cfff","#0570b0ff")
+
+# Plot individual: CONVIVIALITY -----------------------------------------------
+
+# colors<- c("#b3b3b3ff","#f1eef6ff","#bdc9e1ff","#74a9cfff","#0570b0ff")
+colors<- c("#f1eef6ff","#bdc9e1ff","#b3b3b3ff","#74a9cfff","#0570b0ff")
+range <- c(0.55, 1)
+
 p_ansied<-
@@
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
                       reverse.colors = T,
                       reverse.scale = F,
-                      show.n = FALSE) + 
+                      show.n = FALSE,
+                      intercept.line.color = "white") + 
   ggplot2::theme(legend.position = "bottom") +
+  ggplot2::geom_hline(yintercept = 0.32,colour="black",size=1) +
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
+    breaks = c(5, 4, 3, 2, 1),    
     labels = c(
-      "Neither comfortable nor uncomfortable",      
       "Very uncomfortable",
       "Uncomfortable",
+      "Neither comfortable nor uncomfortable",      
       "Comfortable",
@@
   ggplot2::geom_label(label.padding = 1, label.r = 1,label.size = 0.1)
-
-p_ansied[["layers"]]
 
@@
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
-                      geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
                       reverse.colors = T,
                       reverse.scale = F,
-                      show.n = FALSE)  +
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
   ggplot2::theme(legend.position = "bottom") +
+  ggplot2::geom_hline(yintercept = 0.331,colour="black",size=1) + 
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
+    breaks = c(5, 4, 3, 2, 1),    
     labels = c(
-      "Somewhat ",      
       "Very little or not at all",
       "A little",
+      "Somewhat ",
       "Quite a lot",
@@
 
-p_convi<-
-  dplyr::select(df_study1,in_valfam,in_valami,in_amigcl) %>%
+p_valfam<-
+  dplyr::select(df_study1,
+                in_valfam) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
                       reverse.colors = T,
                       reverse.scale = F,
-                      show.n = FALSE)  +
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
   ggplot2::theme(legend.position = "bottom")+
+  ggplot2::geom_hline(yintercept = 0.415,colour="black",size=1) + 
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
+    breaks = c(5, 4, 3, 2, 1),
+  ) +
+  theme(legend.position = "none")
+
+p_valami<-
+  dplyr::select(df_study1,
+                in_valami) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      geom.size = 0.8,
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom")+
+  ggplot2::geom_hline(yintercept = 0.419,colour="black",size=1)  + 
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c(5, 4, 3, 2, 1),
+  ) +
+  theme(legend.position = "none")
+
+p_amigcl<-
+  dplyr::select(df_study1,
+                in_amigcl) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      geom.size = 0.8,
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE, 
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom")+
+  ggplot2::geom_hline(yintercept = 0.122,colour="black",size=1) + 
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c(5, 4, 3, 2, 1),    
     labels = c(
-      "Neither disagree nor agree",      
       "Strongly disagree",
       "Disagree",
+      "Neither disagree nor agree",
       "Agree",
@@
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom")+
+  ggplot2::geom_hline(yintercept = 0.278,colour="black",size=1) +  
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c(5, 4, 3, 2, 1),
+    labels = c(
+      "Not similiar",
+      "Slightly similar",
+      "Somewhat similar",      
+      "Quite similar",
+      "Very similar"
+    )
+  )
+
+p_amsimb<-
+  dplyr::select(df_study1,in_amsimb) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      geom.size = 0.8,
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = T,
+                      reverse.scale = F,
+                      show.n = FALSE,
+                      )  +
+  ggplot2::geom_hline(yintercept = 0.145,colour="black",size=1) + 
+  ggplot2::theme(legend.position = "bottom")+
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c(5, 4, 3, 2, 1),
+  ) +
+  theme(legend.position = "none")
+
+p_acpier<-
+  dplyr::select(df_study1,in_acpier) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      geom.size = 0.8,
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
                       reverse.colors = T,
@@
                       show.n = FALSE)  +
+  ggplot2::geom_hline(yintercept = 0.125,colour="black",size=1) + 
   ggplot2::theme(legend.position = "bottom")+
@@
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
-    labels = c(
-      "Somewhat similar",      
-      "Not similiar",
-      "Slightly similar",
-      "Quite similar",
-      "Very similar"
-    )
-  )
-
-p_identi<-
-  dplyr::select(df_study1,in_amsimb,in_acpier,in_acadop,
-                in_amsimb) %>%
+    breaks = c(5, 4, 3, 2, 1),
+  ) +
+  theme(legend.position = "none")
+
+p_acadop<-
+  dplyr::select(df_study1,in_acadop) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
+                      catcount = 5,
+                      grid.range  =  c (0.6, 1),
+                      show.prc.sign = T,
                       reverse.colors = T,
@@
                       show.n = FALSE)  +
+  ggplot2::geom_hline(yintercept = 0.158,colour="black",size=1) + 
   ggplot2::theme(legend.position = "bottom")+
@@
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
+    breaks = c(5, 4, 3, 2, 1),
     labels = c(
-      "Neither disagree nor agree",      
       "Strongly disagree",
       "Disagree",
+      "Neither disagree nor agree",         
       "Agree",
@@
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
+                      catcount = 5,
+                      grid.range  =  c(0.6 , 1),
+                      show.prc.sign = T,
                       reverse.colors = T,
@@
   ggplot2::theme(legend.position = "bottom") +
+  ggplot2::geom_hline(yintercept = 0.12,colour="black",size=1) + 
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
+    breaks = c(5, 4, 3, 2, 1),
     labels = c(
-      "Neither disagree nor agree",      
       "Strongly disagree",
       "Disagree",
+      "Neither disagree nor agree",      
       "Agree",
@@
 pacman::p_load(cowplot)
-s <- 0.4
-m <- 0.6
-l <- 0.8
+t <- 0.1
+s <- 0.45
+m <- 0.4
+xs <- 0.25
 #Final plot
@@
 
+caption <- 
+  stringr::str_wrap(
+    paste0("Note: Own elaboration based on ELSOC Survey ",
+           "(n=",dim(na.omit(df_study1))[1],"). ",
+           "Distribution of responses for each item regarding the variables Conviviality, Identity and Threat.",
+           " The vertical black line is the threshold between negative (left) and positive (right) attitudes towards migration. ",
+          "\n[PER/VEN] represent Peruvians and Venezuelans, the different target groups included in each statement."),
+    width = 100)
+
 likert1<-   
   cowplot::add_sub(
-    cowplot::plot_grid(title,p_ansied,p_simpat,p_convi,p_simili,p_identi,p_threat, 
+    cowplot::plot_grid(title,
+                       p_ansied,
+                       p_simpat,
+                       p_valfam,p_valami,p_amigcl,
+                       p_simili,
+                       p_amsimb, p_acpier,p_acadop,
+                       p_threat, 
                        align = "hv",
                        axis = "a",
-                       nrow = 7, #4 plots
+                       nrow = 11, #10 plots
                        ncol = 1, #1 column   
-                       rel_heights = c(s,s,0.35,m,s,m,s) #fig size within plot
+                       rel_heights = c(t,
+                                       s,m,xs,xs,m,
+                                       s,xs,xs,m,
+                                       s) #fig size within plot
     ),
-    label = paste0("Source: Own elaboration based on ELSOC Survey",
-                   " (n=",dim(na.omit(df_study1))[1],")"
-    ),        # Caption: source and sample size
-    size = 12,  # Caption font size
-    hjust = -0.25# Caption align  
+    label = caption, # Caption: source and sample size
+    size = 11,  # Caption font size
+    hjust = 0# Caption align 
   ) %>% 
   cowplot::ggdraw()
-# likert1
-ggplot2::ggsave(filename = "figure1.jpg",
-                plot = likert1,
-                path = here::here("output/images"),
-                device = "jpg",
-                units = "cm",
-                scale = 3.5,
-                width = 10,height = 10)
-
-#Final plot
-likert1<-   
-  cowplot::add_sub(
-    cowplot::plot_grid(p_a,p_b,p_c,p_d, #plot 1,2,3 and legend (bottom)
-                       align = "hv",
-                       axis = "a",
-                       nrow = 4, #4 plots
-                       ncol = 1, #1 column
-                       rel_heights = c(0.25,0.25,0.25,0.90) #fig size within plot
-    ),
-    label = paste0("Source: Own elaboration based on ELSOC Survey",
-                   " (n=",dim(na.omit(df_study1))[1],")"
-    ),        # Caption: source and sample size
-    size = 12,  # Caption font size
-    hjust = -0.25# Caption align
-  ) %>%
-  cowplot::ggdraw()  
+jpeg(here::here("output/images/figure1.jpg"),
+    # units = "cm",
+    # res = 100,
+    width = 1100,
+    height = 1100)
 likert1
-ggplot2::ggsave(filename = "figure1.jpg",
-                plot = likert1,
-                path = here::here("output/images"),
-                device = "jpg",
-                units = "cm",
-                scale = 3,
-                width = 10,height = 8)
+dev.off() 
 ```
@@
 # Full sample_______________________________________________________________________
-mat_full<- 
+df_corr <- 
 df_study_long_t5_comunas %>%
@@
          in_amreal # threat
-         ) %>% 
-  psych::mixedCor()
-
-# dim(mat_full)
+         )
+
+mat_full<- df_corr %>%  psych::mixedCor()
 
@@
 
-png(file=here::here("output/images/figure2.jpg"),width=900, height=700)
+png(file=here::here("output/images/figure2.jpg"),width=1000, height=800)
 corrplot::corrplot(cor_full,
@@
   )
+  text(-3, 1,
+       adj = 0,
+         paste0("Note: Authors calculation based on ELSOC Survey ",
+                "(n=",dim(df_corr)[1],").",
+                "Polychoric correlations for each item for Conviviality, Identity and Threat.",
+                "\n[PER/VEN] represent Peruvians and Venezuelans, the different target groups included in each statement."),   
+       cex = 0.79)
 dev.off()
 
-# Wave 1_______________________________________________________________________
-mat_1<- 
-df_study_long_t5_comunas %>%
-  filter(ola==1) %>%
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
-         ) %>% 
-  psych::mixedCor()
-
-cor_1 <- as.matrix(mat_1$rho)
-diag(cor_1) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_1) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
-#set Column names of the matrix
-colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_1,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 1")
-
-# Wave 2_______________________________________________________________________
-mat_2<-   
-df_study_long_t5_comunas %>%
-  filter(ola==2) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
-         ) %>% 
-  psych::mixedCor()
-
-cor_2 <- as.matrix(mat_2$rho)
-diag(cor_2) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_2) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
-#set Column names of the matrix
-colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_2,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 2")
-
-# Wave 3_______________________________________________________________________
-mat_3 <- 
-df_study_long_t5_comunas %>%
-  filter(ola==3) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
-         ) %>% 
-  psych::mixedCor()
-
-cor_3 <- as.matrix(mat_3$rho)
-diag(cor_3) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_3) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
-#set Column names of the matrix
-colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_3,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 3")
-
-# Wave 4_______________________________________________________________________
-mat_4 <- 
-df_study_long_t5_comunas %>%
-  filter(ola==4) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
-         ) %>% 
-  psych::mixedCor()
-
-cor_4 <- as.matrix(mat_4$rho)
-diag(cor_4) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_4) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
-#set Column names of the matrix
-colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_4,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 4")
-
-# Wave 5_______________________________________________________________________
-mat_5 <- 
-df_study_long_t5_comunas %>%
-  filter(ola==5) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
-         ) %>% 
-  psych::mixedCor()
-
-cor_5 <- as.matrix(mat_5$rho)
-diag(cor_5) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_5) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
-#set Column names of the matrix
-colnames(cor_5) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_5,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 5")
+# # Wave 1-----------------------------------------------
+# mat_1<- 
+# df_study_long_t5_comunas %>%
+#   filter(ola==1) %>%
+#     select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+#          in_simili,in_amsimb,in_acpier,in_acadop,
+#          conviviality, identity,
+#          in_amreal,
+#          educ,quintil1:edad,
+#          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+#   na.omit() %>%   
+#   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
+#          in_simili,in_amsimb,in_acpier, in_acadop, # identity
+#          in_amreal # threat
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# cor_1 <- as.matrix(mat_1$rho)
+# diag(cor_1) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_1) <- c("A. Interaction anxiety with [PER/VEN]",
+#                      "B. Sympathy for [PER/VEN] living in Chile",
+#                      "C. Family value of immigrant friends [PER/VEN]",
+#                      "D. Friends value of immigrant friends [PER/VEN]",
+#                      "E. Immigrants [PER/VEN] have Chilean friends",
+#                      "F. Similarity between Chileans and [PER/VEN]",
+#                      "G. Loose of identity because of immigrants (reverse) ",
+#                      "H. Immigrant keeping their customs",
+#                      "I. Adoption of Chilean customs",
+#                      "J. Unemployment increases")
+# #set Column names of the matrix
+# colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_1,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 1")
+# 
+# # Wave 2-----------------------------------------------
+# mat_2<-   
+# df_study_long_t5_comunas %>%
+#   filter(ola==2) %>% 
+#     select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+#          in_simili,in_amsimb,in_acpier,in_acadop,
+#          conviviality, identity,
+#          in_amreal,
+#          educ,quintil1:edad,
+#          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+#   na.omit() %>%   
+#   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
+#          in_simili,in_amsimb,in_acpier, in_acadop, # identity
+#          in_amreal # threat
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# cor_2 <- as.matrix(mat_2$rho)
+# diag(cor_2) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_2) <- c("A. Interaction anxiety with [PER/VEN]",
+#                      "B. Sympathy for [PER/VEN] living in Chile",
+#                      "C. Family value of immigrant friends [PER/VEN]",
+#                      "D. Friends value of immigrant friends [PER/VEN]",
+#                      "E. Immigrants [PER/VEN] have Chilean friends",
+#                      "F. Similarity between Chileans and [PER/VEN]",
+#                      "G. Loose of identity because of immigrants (reverse) ",
+#                      "H. Immigrant keeping their customs",
+#                      "I. Adoption of Chilean customs",
+#                      "J. Unemployment increases")
+# #set Column names of the matrix
+# colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_2,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 2")
+# 
+# 
+# # Wave 3-----------------------------------------------
+# mat_3 <- 
+# df_study_long_t5_comunas %>%
+#   filter(ola==3) %>% 
+#     select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+#          in_simili,in_amsimb,in_acpier,in_acadop,
+#          conviviality, identity,
+#          in_amreal,
+#          educ,quintil1:edad,
+#          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+#   na.omit() %>%   
+#   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
+#          in_simili,in_amsimb,in_acpier, in_acadop, # identity
+#          in_amreal # threat
+#          ) %>% 
+#   psych::mixedCor()  
+# 
+# cor_3 <- as.matrix(mat_3$rho)
+# diag(cor_3) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_3) <- c("A. Interaction anxiety with [PER/VEN]",
+#                      "B. Sympathy for [PER/VEN] living in Chile",
+#                      "C. Family value of immigrant friends [PER/VEN]",
+#                      "D. Friends value of immigrant friends [PER/VEN]",
+#                      "E. Immigrants [PER/VEN] have Chilean friends",
+#                      "F. Similarity between Chileans and [PER/VEN]",
+#                      "G. Loose of identity because of immigrants (reverse) ",
+#                      "H. Immigrant keeping their customs",
+#                      "I. Adoption of Chilean customs",
+#                      "J. Unemployment increases")
+# #set Column names of the matrix
+# colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_3,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 3")
+# 
+# # Wave 4-----------------------------------------------
+# mat_4 <- 
+# df_study_long_t5_comunas %>%
+#   filter(ola==4) %>% 
+#     select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+#          in_simili,in_amsimb,in_acpier,in_acadop,
+#          conviviality, identity,
+#          in_amreal,
+#          educ,quintil1:edad,
+#          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+#   na.omit() %>%   
+#   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
+#          in_simili,in_amsimb,in_acpier, in_acadop, # identity
+#          in_amreal # threat
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# 
+# cor_4 <- as.matrix(mat_4$rho)
+# diag(cor_4) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_4) <- c("A. Interaction anxiety with [PER/VEN]",
+#                      "B. Sympathy for [PER/VEN] living in Chile",
+#                      "C. Family value of immigrant friends [PER/VEN]",
+#                      "D. Friends value of immigrant friends [PER/VEN]",
+#                      "E. Immigrants [PER/VEN] have Chilean friends",
+#                      "F. Similarity between Chileans and [PER/VEN]",
+#                      "G. Loose of identity because of immigrants (reverse) ",
+#                      "H. Immigrant keeping their customs",
+#                      "I. Adoption of Chilean customs",
+#                      "J. Unemployment increases")
+# #set Column names of the matrix
+# colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_4,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 4")
+# 
+# 
+# # Wave 5-----------------------------------------------
+# mat_5 <- 
+# df_study_long_t5_comunas %>%
+#   filter(ola==5) %>% 
+#     select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+#          in_simili,in_amsimb,in_acpier,in_acadop,
+#          conviviality, identity,
+#          in_amreal,
+#          educ,quintil1:edad,
+#          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+#   na.omit() %>%   
+#   select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
+#          in_simili,in_amsimb,in_acpier, in_acadop, # identity
+#          in_amreal # threat
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# cor_5 <- as.matrix(mat_5$rho)
+# diag(cor_5) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_5) <- c("A. Interaction anxiety with [PER/VEN]",
+#                      "B. Sympathy for [PER/VEN] living in Chile",
+#                      "C. Family value of immigrant friends [PER/VEN]",
+#                      "D. Friends value of immigrant friends [PER/VEN]",
+#                      "E. Immigrants [PER/VEN] have Chilean friends",
+#                      "F. Similarity between Chileans and [PER/VEN]",
+#                      "G. Loose of identity because of immigrants (reverse) ",
+#                      "H. Immigrant keeping their customs",
+#                      "I. Adoption of Chilean customs",
+#                      "J. Unemployment increases")
+# #set Column names of the matrix
+# colnames(cor_5) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_5,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 5")
 ```
@@
 ## Education categorical {-}
-```{r average-educ, results='asis',fig.dim=c(10,6)}
+```{r models-educ, results='asis',fig.dim=c(10,6)}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg,
+               sjPlot,
+               ggplot2,
+               broom
+               )
+load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+
+old <- theme_set(theme_bw());theme_set(old)
+theme_replace(legend.position="bottom")
+
+# generate analytical sample
+df_study1 <- 
+  df_study_long_t5_comunas %>%
+  select(idencuesta,ola,comuna,cuestion_mig,
+         in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,educ,quintil1:edad,sexo,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  na.omit() %>% 
+  mutate(
+    # conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
+         # identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
+         ola_f = ola,
+         ola=as.numeric(ola),
+         primary = car::recode(educ,recodes = "'Primary'=1;else=0",as.factor = T),
+         -in_amreal
+         )
+
+# Hypothesis___________________________________________________________________
+m1.cat <- "ola_f+pos_id+edad+nation_t1+cuestion_mig"
+m2 <- "ola_f+educ+quintil1+ess+pos_id+edad+sexo+nation_t1+cuestion_mig"
+m3 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+sexo+nation_t1+cuestion_mig"
+m4 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+inm_prop_cas20_w+dif_inm+pos_id+edad+sexo+nation_t1+cuestion_mig"
+
+# h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+# h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+m5 <- "educ*ola+quintil1*ola+ess+know_inm_t1t4*ola+frien_inm_t1t4*ola+pos_id+edad+sexo+inm_prop_cas20_w+dif_inm*ola+nation_t1+cuestion_mig"
+
+h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+sexo+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.fri.time <- "ola*educ*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+sexo+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+
+hipotesis <- c(m1.cat,m2,m3,m4
+               # h.ed.know.time,
+               # h.ed.fri.time
+               )
+# Set the independent variables (IVs)
+ivs<- df_study1 %>% select(starts_with("in_"),"conviviality","identity") %>% names()
+
+# Create the object for each model
+models <- list()
+for (i in ivs) {
+models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
+}
+
+# Estimate the multilevel models
+fits <- list()
+for (i in ivs) {
+  for (z in models[[i]][1:length(hipotesis)]) {
+    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
+  }
+}
+fit1 <- fits
+# change model names within list
+names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
+
+# Tables_______________________________________________________________________
+omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+# reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
+# detail for subsections:
+# https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
+
+coef_names <- c("Wave 2017 ","Wave 2018", "Wave 2019","Wave 2020",
+                "Technical ","High school","Primary",
+                "Quintile 4 ", "Quintile 3","Quintile 2","Quintile 1",
+                "Subjective Social Status",
+                "Woman (ref: Man)",
+                "Stable, know ","Now know","No longer know",
+                "Stable, have friends ","Now have friends","No longer have friends",
+                "Propotion of migrants",
+                "Change in the proportion of migrants")
+
+var_groups  <- list(
+    "Wave (ref: Wave 2016)" = 1:4,
+    "Education (ref: Universitary)" = 5:7,
+    "Household Income (ref: Quintile 5)" = 8:11,
+    "Know migrant (ref: Stable, do not know)" = 14:16,
+    "Migrant friend (ref: Stable, do not have friends)" = 17:19,
+    "Municipality characteristics" = 20:21
+  )
+
+gof_names = c(
+  "AIC",
+  "BIC",
+  "Log Likelihood",
+  "L1: Num. obs.",
+  "L2: Num. Individual",
+  "L3: Num. Municipality",
+  "Var: Individual (Intercept)",
+  "Var: Municipality (Intercept)",
+  "Var: Residual"
+)
+order_gof = c(2,3,4,1,5,6,7,8,9,10)
+
+custom_note = paste("%stars.
+<br>  Note: Standard error in parenthesis.
+Model 1 shows the estimation for the average change between waves.
+<br> Models 2 y 3 show the estimations of the random intercept multilevel models of the individual social status and changes in contact and friendship with migrants.
+<br> The estimation considers three levels of nesting: Observations (L1) within Individuals (L2) which are also nested within Municipalities (L3)")
+
+cat(paste0("\n## ","Conviviality (avg)","\n"))
+
+# generate likelihood ratio test
+models_conv<- fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]]
+lrt_conv <- as.data.frame(anova(models_conv[[1]],models_conv[[2]],
+                              models_conv[[3]],models_conv[[4]]))
+lrt_conv$report <-
+  paste0(
+    round(lrt_conv$Chisq, 1),
+    "(",
+    lrt_conv$Df,
+    ")",
+    gtools::stars.pval(lrt_conv$`Pr(>Chisq)`)
+  )
+
+lrt_conv$report[[1]] <- ""
+
+screenreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+        omit.coef = omit,
+        single.row = T,
+        # caption = caption_conv,
+        caption.above = T,
+        groups = var_groups,
+        custom.coef.names=coef_names,
+        # custom.gof.rows = list("Likelihood-ratio Test" = lrt_conv$report),
+        # custom.gof.names = gof_names,
+        # reorder.gof = order_gof,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        # file = here::here("output/tables/tab_conv.html"),
+        custom.note = custom_note
+        )
+
+caption_conv <- "Multilevel linear regression models for Conviviality"
+htmlreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+        omit.coef = omit,single.row = T,
+        caption = caption_conv,
+        caption.above = T,
+        groups = var_groups,
+        custom.coef.names=coef_names,
+        custom.gof.rows = list("Likelihood-ratio Test" = lrt_conv$report),
+        custom.gof.names = gof_names,
+        reorder.gof = order_gof,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        file = here::here("output/tables/tab_conv.html"),
+        custom.note = custom_note
+        # reorder.coef = reorderc
+        )
+webshot::webshot(url = here::here("output/tables/tab_conv.html"),
+                 file = here::here("output/tables/tab_conv.png"))
+
+# plot_model(fit1$conviviality1,type = "pred",terms = " ola_f")
+
+cat(paste0("\n ","<br>","\n"))
+
+cat(paste0("\n## ","Identity (avg)","\n"))
+
+# generate likelihood ratio test
+models_ident<- fit1[names(fit1)[grepl("identity", names(fit1),"\n")]]
+lrt_ident <- as.data.frame(anova(models_ident[[1]],models_ident[[2]],
+                              models_ident[[3]],models_ident[[4]]))
+lrt_ident$report <-
+  paste0(
+    round(lrt_ident$Chisq, 1),
+    "(",
+    lrt_ident$Df,
+    ")",
+    gtools::stars.pval(lrt_ident$`Pr(>Chisq)`)
+  )
+
+lrt_ident$report[[1]] <- ""
+
+caption_id <- "Multilevel linear regression models for Identity"
+
+screenreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
+        # omit.coef = omit,
+        single.row = T,
+        # caption = caption_conv,
+        caption.above = T,
+        # groups = var_groups,
+        # custom.coef.names=coef_names,
+        # custom.gof.rows = list("Likelihood-ratio Test" = lrt_conv$report),
+        # custom.gof.names = gof_names,
+        # reorder.gof = order_gof,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        # file = here::here("output/tables/tab_conv.html"),
+        custom.note = custom_note
+        # reorder.coef = reorderc
+        )
+
+htmlreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
+        omit.coef = omit,
+        caption = caption_id,
+        caption.above = T,
+        single.row = T,
+        groups = var_groups,
+        custom.coef.names=coef_names,
+        custom.gof.rows = list("Likelihood-ratio Test" = lrt_ident$report),
+        custom.gof.names = gof_names,
+        reorder.gof = order_gof,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        file = here::here("output/tables/tab_iden.html"),
+        custom.note = custom_note
+        )
+webshot::webshot(url = here::here("output/tables/tab_iden.html"),
+                 file = here::here("output/tables/tab_iden.png"))
+
+# plot_model(fit1$identity1,type = "pred",terms = "ola_f")
+
+cat(paste0("\n## ","J. Unemployment increases (Threat)","\n"))
+
+# generate likelihood ratio test
+models_thre<- fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]]
+lrt_thre <- as.data.frame(anova(models_thre[[1]],models_thre[[2]],
+                              models_thre[[3]],models_thre[[4]]))
+lrt_thre$report <-
+  paste0(
+    round(lrt_thre$Chisq, 1),
+    "(",
+    lrt_thre$Df,
+    ")",
+    gtools::stars.pval(lrt_thre$`Pr(>Chisq)`)
+  )
+
+lrt_thre$report[[1]] <- ""
+
+caption_th <- "Multilevel linear regression models for Threat"
+screenreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
+        # omit.coef = omit,
+        single.row = T,
+        # caption = caption_conv,
+        caption.above = T,
+        # groups = var_groups,
+        # custom.coef.names=coef_names,
+        # custom.gof.rows = list("Likelihood-ratio Test" = lrt_conv$report),
+        # custom.gof.names = gof_names,
+        # reorder.gof = order_gof,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        # file = here::here("output/tables/tab_conv.html"),
+        custom.note = custom_note
+        # reorder.coef = reorderc
+        )
+
+htmlreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
+        omit.coef = omit,single.row = T,
+        caption = caption_th,
+        caption.above = T,
+        groups = var_groups,
+        custom.coef.names=coef_names,
+        custom.gof.rows = list("Likelihood-ratio Test" = lrt_thre$report),
+        custom.gof.names = gof_names,
+        reorder.gof = order_gof,
+        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
+        file = here::here("output/tables/tab_thr.html"),
+        custom.note = custom_note        
+        # reorder.coef = reorderc
+        ) 
+webshot::webshot(url = here::here("output/tables/tab_thr.html"),
+                 file = here::here("output/tables/tab_thr.png"))
+
+# plot_model(fit1$in_amreal1,type = "pred",terms = " ola_f")
+```
+
+```{r interactions-educ, results='asis',fig.dim=c(10,6)}
 rm(list=ls())
@@
          conviviality, identity,
-         in_amreal,educ,quintil1:edad,
+         in_amreal,educ,quintil1:edad,sexo,
          inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
@@
 # Hypothesis___________________________________________________________________
-m1.cat <- "ola_f+pos_id+edad+nation_t1+cuestion_mig"
-m2 <- "ola_f+educ+quintil1+ess+pos_id+edad+nation_t1+cuestion_mig"
-m3 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+nation_t1+cuestion_mig"
-m4 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+inm_prop_cas20_w+dif_inm+pos_id+edad+nation_t1+cuestion_mig"
-
-# h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-m5 <- "educ*ola+quintil1*ola+ess+know_inm_t1t4*ola+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm*ola+nation_t1+cuestion_mig"
-
-h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h.ed.fri.time <- "ola*educ*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-
-hipotesis <- c(m1.cat,m2,m3,m4
-               # h.ed.know.time,
-               # h.ed.fri.time
-               )
-# Set the independent variables (IVs)
-ivs<- df_study1 %>% select(starts_with("in_"),"conviviality","identity") %>% names()
-
-# Create the object for each model
-models <- list()
-for (i in ivs) {
-models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
-}
-
-# Estimate the multilevel models
-fits <- list()
-for (i in ivs) {
-  for (z in models[[i]][1:length(hipotesis)]) {
-    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
-  }
-}
-fit1 <- fits
-# change model names within list
-names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
+m5 <- "educ*ola+quintil1*ola+ess+know_inm_t1t4*ola+frien_inm_t1t4*ola+pos_id+edad+sexo+inm_prop_cas20_w+dif_inm*ola+nation_t1+cuestion_mig"
+h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+sexo+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
+h.ed.fri.time <- "ola*educ*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+sexo+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
 
 # Tables_______________________________________________________________________
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+omit <- "(pos_id)|(edad)|(sexo)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
 # reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
@@
 
-coef_names <- c("Wave 2017 (ref: Wave 2016)","Wave 2018", "Wave 2019","Wave 2020",
-                "Technical (ref: Universitary)","High school","Primary",
-                "Quintile 4 (ref: Quintile 5)", "Quintile 3","Quintile 2","Quintile 1",
+coef_names <- c("Wave 2017 ","Wave 2018", "Wave 2019","Wave 2020",
+                "Technical ","High school","Primary",
+                "Quintile 4 ", "Quintile 3","Quintile 2","Quintile 1",
                 "Subjective Social Status",
-                "Stable, know (ref: Stable, do not know)","Now know","No longer know",
-                "Stable, have friends (ref: Stable, do not have friends)","Now have friends","No longer have friends",
+                "Stable, know ","Now know","No longer know",
+                "Stable, have friends ","Now have friends","No longer have friends",
                 "Propotion of migrants",
@@
 
-cat(paste0("\n## ","Conviviality (avg)","\n"))
-
-caption_conv <- "Multilevel linear regression models for Conviviality"
-htmlreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
-        omit.coef = omit,single.row = T,
-        caption = caption_conv,
-        caption.above = T,
-        custom.coef.names=coef_names,
-        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
-        file = here::here("output/tables/tab_conv.html")
-        # reorder.coef = reorderc
-        )
-webshot::webshot(url = here::here("output/tables/tab_conv.html"),
-                 file = here::here("output/tables/tab_conv.png"))
-
-plot_model(fit1$conviviality1,type = "pred",terms = " ola_f")
-
-cat(paste0("\n ","<br>","\n"))
-
-cat(paste0("\n## ","Identity (avg)","\n"))
-
-caption_id <- "Multilevel linear regression models for Identity"
-htmlreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
-        omit.coef = omit,
-        caption = caption_id,
-        caption.above = T,
-        single.row = T,
-        custom.coef.names=coef_names,
-        # reorder.coef = reorderc
-        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
-        file = here::here("output/tables/tab_iden.html")        
-        )
-webshot::webshot(url = here::here("output/tables/tab_iden.html"),
-                 file = here::here("output/tables/tab_iden.png"))
-
-plot_model(fit1$identity1,type = "pred",terms = "ola_f")
-
-cat(paste0("\n## ","J. Unemployment increases (Threat)","\n"))
-
-caption_th <- "Multilevel linear regression models for Threat"
-htmlreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
-        omit.coef = omit,single.row = T,
-        caption = caption_th,
-        caption.above = T,
-        custom.coef.names=coef_names,
-        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
-        file = here::here("output/tables/tab_thr.html")         
-        # reorder.coef = reorderc
-        ) 
-webshot::webshot(url = here::here("output/tables/tab_thr.html"),
-                 file = here::here("output/tables/tab_thr.png"))
-
-plot_model(fit1$in_amreal1,type = "pred",terms = " ola_f")
-
-cat(paste0("\n## ","Interactions","\n"))  #------------------------------------#
+var_groups  <- list(
+    "Wave (ref: Wave 2016)" = 1:4,
+    "Education (ref: Universitary)" = 5:7,
+    "Household Income (ref: Quintile 5)" = 8:11,
+    "Know migrant (ref: Stable, do not know)" = 13:15,
+    "Migrant friend (ref: Stable, do not have friends)" = 16:18,
+    "Municipality characteristics" = 19:20
+  )
+
+gof_names = c(
+  "AIC",
+  "BIC",
+  "Log Likelihood",
+  "L1: Num. obs.",
+  "L2: Num. Individual",
+  "L3: Num. Municipality",
+  "Var: Individual (Intercept)",
+  "Var: Individual Wave",
+  "Cov: Individual (Intercept) Wave",
+  "Var: Municipality (Intercept)",
+  "Var: Residual"
+)
+order_gof = c(2,3,4,1,5,6,7,8,9,10)
+
+custom_note = paste("%stars.
+<br>  Note: Standard error in parenthesis.
+Model 1 to 3 shows the results of the interaction effects <br> of time with socioeconomic status, know migrants, has migrant friend and change on migrant population.
+The time is specified as a random slope")
+
+cat(paste0("\n## ","Interactions","\n"))  # Interactions ----------------------
 
@@
 
-# Create the object for each model
-models_int <- list()
-for (i in ivs) {
-models_int[[i]] <- paste0(i,"~",hipotesis_int,"+(1|idencuesta)+(1|comuna)")  
-}
-
-# Estimate the multilevel models
-fit_int <- list()
-for (i in ivs) {
-  for (z in models_int[[i]][1:length(hipotesis_int)]) {
-    fit_int[[z]] <- lme4::lmer(formula = z, data = df_study1)
-  }
-}
-
-names(fit_int) <- paste0(rep(ivs,each=length(hipotesis_int)),1:length(hipotesis_int))
-
-omit_int <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)|(ess)|(inm_prop_cas20_w)"
+ivs<- df_study1 %>%
+  select("in_amreal","conviviality","identity") %>% names()
+
+# Generate Table elements-------------------------------------------------------
+omit_int <- "(pos_id)|(edad)|(sexo)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)|(ess)|(inm_prop_cas20_w)"
 coef_int <- c("Technical (ref: Universitary) ","High school","Primary","Wave",
@@
 caption_int <- "Multilevel linear regression models with interactions"
+
+# Table: random-slope status & ties migrants by municipality--------------------
+# Load the previously estimated models
+load(here::here("output/tables/models_rs_muni.RData"))
+screenreg(l = fit_int[c("conviviality1","identity1","in_amreal1")],
+        omit.coef = omit_int,
+        single.row = T,
+             # caption = caption_int,
+        caption.above = T,
+        custom.model.names = c("Model 1\n(Conviviality)",
+                               "Model 2\n(Identity)",
+                               "Model 3\n(Threat)"),
+        # file = here::here("output/tables/tab_rs_muni.txt")
+        )
+
 htmlreg(l = fit_int[c("conviviality1","identity1","in_amreal1")],
+        omit.coef = omit_int,
+        single.row = T,
+        # custom.coef.names=coef_int,
+        # reorder.coef = c(4,1:3,5:29),
+        # groups = list(
+        #   # "Education (ref: Universitary)"=2:4,
+        #               # "Household Income Quintile (ref: Quintile 5)"=5:8,
+        #               # "Know migrant (ref: Stable, do not know)" =9:11,
+        #               # "Friendship with migrants (ref: Stable, do not have friends)" =12:14,
+        #               "Education x Wave" =16:18,
+        #               "Quintile x Wave" =19:22,
+        #               "Know migrant x Wave" =23:25,
+        #               "Migrant friend x Wave" =26:28,
+        #               "Migrant Propulation x Wave" =29
+        #               ),
+        caption = caption_int,
+        caption.above = T,
+        custom.model.names = c("Model 1\n(Conviviality)","Model 2\n(Identity)","Model 3\n(Threat)"),
+        custom.gof.names = gof_names,        
+        file = here::here("output/tables/tab_int.html")
+        )
+webshot::webshot(url = here::here("output/tables/tab_int.html"),
+                 file = here::here("output/tables/tab_int.png"))
+
+# Create the object for each model 
+# (random slope wave | subject)-------------------------------------------------
+models_int <- list()
+for (i in ivs) {
+models_int[[i]] <- 
+  paste0(i,"~",
+         hipotesis_int,
+         "+(1+ola|idencuesta)+(1|comuna)")  
+}
+
+# Estimate the multilevel models
+tictoc::tic()
+fit_int <- list()
+for (i in ivs) {
+  for (z in models_int[[i]][1:length(hipotesis_int)]) {
+    fit_int[[z]] <- lme4::lmer(formula = z, data = df_study1)
+  }
+}
+tictoc::toc()
+names(fit_int) <- paste0(rep(ivs,each=length(hipotesis_int)),1:length(hipotesis_int))
+models_rs_individual <- fit_int 
+# save: random slope wave within subject ----------------
+save(models_rs_individual,file = here::here("output/tables/models_rs_individual.RData"))
+
+# Load the previously estimated models 
+load(here::here("output/tables/models_rs_individual.RData"))
+
+#Table: random-slope of year within subject-------------------------
+screenreg(l = models_rs_individual[c("conviviality1","identity1","in_amreal1")],
+        omit.coef = omit_int,
+        single.row = T,
+        caption = caption_int,
+        caption.above = T,
+        custom.model.names = c("Model 1\n(Conviviality)",
+                               "Model 2\n(Identity)",
+                               "Model 3\n(Threat)"),
+        # file = here::here("output/tables/tab_rs_wave_individual.txt")        
+        )
+
+htmlreg(l = models_rs_individual[c("conviviality1","identity1","in_amreal1")],
         omit.coef = omit_int,
@@
         groups = list(
-          # "Education (ref: Universitary)"=2:4,
-                      # "Household Income Quintile (ref: Quintile 5)"=5:8,
-                      # "Know migrant (ref: Stable, do not know)" =9:11,
-                      # "Friendship with migrants (ref: Stable, do not have friends)" =12:14,
                       "Education x Wave" =16:18,
@@
         custom.model.names = c("Model 1\n(Conviviality)","Model 2\n(Identity)","Model 3\n(Threat)"),
-        file = here::here("output/tables/tab_int.html")
+        custom.gof.names = gof_names,
+        custom.note = custom_note,
+        file = here::here("output/tables/tab_rs_wave_individual.html")
         )
@@
                  file = here::here("output/tables/tab_int.png"))
+
 ```

### 2022-10-24 production/prod_prep.Rmd
ajustes likertplot

@@
 # Chilean CENSUS data (2017)____________________________________________________
+ options(timeout=100*100)
 download.file(url = "https://github.com/jciturras/data-chile/raw/main/data/censo-2017/censo-2017.RData",
@@
   mutate(in_simpat=r09,
+         sexo=m0_sexo,
          in_simili=r10,
@@
          muestra,
+         sexo,
          cuestion_mig,
@@
                      # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
-                     values_from = names(select(df_study1,comuna,comunacod,cuestion_mig,in_simpat:edad))
+                     values_from = names(select(df_study1,comuna,comunacod,cuestion_mig,in_simpat:edad,sexo))
                      )
@@
 df_study1_wide$edad_5 <-df_study1_wide$edad_1
+
+df_study1_wide$sexo_2 <-df_study1_wide$sexo_1 #sex
+df_study1_wide$sexo_3 <-df_study1_wide$sexo_1
+df_study1_wide$sexo_4 <-df_study1_wide$sexo_1
+df_study1_wide$sexo_5 <-df_study1_wide$sexo_1
 
@@
                                  "^know_inm_t1t4_","^frien_inm_t1t4_","^pos_id_",
-                                 "^edad_"
+                                 "^edad_","^sexo_"
                                  ),
@@
                              "know_inm_t1t4","frien_inm_t1t4","pos_id",
-                             "edad"),
+                             "edad","sexo"),
               na.rm = F,value.factor = TRUE
@@
                   by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,
+ dplyr::select(idencuesta:edad,sexo,
                inm_prop_cas17_w,

### 2023-03-17 production/prod_analysis-fix-t5.Rmd
ajustes nota emol

@@
 # Descriptive analysis
+
+```{r}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               sjlabelled,ggplot2
+               )
+load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+
+# Descriptive table __________________________________________________________
+df_study1 <- 
+  df_study_long_t5_comunas %>% 
+  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,ola,idencuesta,
+         in_simili,in_amsimb,in_acpier,in_acadop,
+         conviviality, identity,
+         in_amreal,
+         educ,quintil1:edad,
+         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm,sexo,edad,) 
+# extract variable labels
+lab_study1 <- sjlabelled::get_label(df_study1) 
+# convert items to factor
+for (i in names(select(df_study1,starts_with("in_")))) {
+  df_study1[[i]] <- factor(df_study1[[i]],1:5,get_labels(df_study1[[i]]))
+}
+# set variable labels
+df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
+df_study1$in_amreal <- sjlabelled::set_label(df_study1$in_amreal,label = "Threat (Unemployment increases)")
+
+sjmisc::frq(df_study1$in_amreal)
+sjmisc::frq(as.numeric(df_study1$in_amreal))
+sjmisc::frq(df_study1$educ)
+
+levels(df_study1$ola) <- c("2016","2017","2018","2019","2021")  
+levels(df_study1$educ) <- c("Universitaria","Técnica","Media","Básica")
+
+#Paso 1
+#dropeamos el nivel -999 de la variable:
+df_study1$in_amreal1 <- car::recode(df_study1$in_amreal,
+                                    "c('Strongly disagree','Disagree')='Desacuerdo';
+                                    'Neither disagree nor agree'='Ni de acuerdo, ni desacuerdo';
+                                    c('Agree','Strongly agree')='Deacuerdo'")
+
+df_study1$in_amreal1 <- factor(df_study1$in_amreal1,
+                               labels =  c('Deacuerdo',
+                                          'Ni de acuerdo, ni desacuerdo',
+                                          'Desacuerdo'))
+
+a1 <- data.frame(table(haven::as_factor(df_study1$in_amreal1),
+                       haven::as_factor(df_study1$ola),
+                       haven::as_factor(df_study1$idencuesta)))
+
+a1 <- a1 %>% filter(Var2 %in% c(2016,2018,2021))
+
+a1 <- a1 %>% dplyr::filter(Freq>0)
+names(a1) <- c("g_civico","ola","folio","Freq")
+a2 <- a1 %>% group_by(ola) %>% transmute(porcentaje=Freq/sum(Freq)) 
+a2$ola <- NULL
+datos.d.1<- cbind(a1,a2)
+datos.d.1 <- na.omit(datos.d.1)
+#Paso 2
+b1 <- data.frame(table(haven::as_factor(df_study1$in_amreal1),
+                       haven::as_factor(df_study1$ola)))
+
+b1 <-b1 %>% filter(Var2 %in% c(2016,2018,2021))
+
+names(b1) <- c("g_civico","ola","Freq")
+b1 <- b1 %>% dplyr::filter(Freq>0)
+b2 <- b1 %>% 
+  group_by(ola) %>% 
+  transmute(porcentaje=Freq/sum(Freq)) %>% 
+  mutate(folio = 1)
+b2$ola <- NULL
+etiquetas.d.1<- cbind(b1,b2)
+etiquetas.d.1 <- na.omit(etiquetas.d.1)
+dim<- sjmisc::frq(datos.d.1$folio) %>% data.frame() %>% filter(frq==2) %>% dim()
+#-----------------------------------------------------------------------------#
+threat <- ggplot(datos.d.1, aes(x = ola, fill = g_civico, stratum = g_civico,
+                             alluvium = folio, y = porcentaje)) +
+  ggalluvial::geom_flow(alpha = .66) + 
+  ggalluvial::geom_stratum(linetype = 0) +
+  scale_y_continuous(labels = scales::percent) + 
+  # scale_x_discrete(labels=c(2019,,2))+  
+  ylab(label = NULL) +
+  xlab(label = NULL) + 
+  theme(legend.position = 'right',
+        legend.title = element_blank()) +
+  scale_fill_viridis_d(begin = 0, end = .85, direction = -1, option = 'viridis') +
+  geom_text(data = etiquetas.d.1, 
+            aes(label = ifelse(porcentaje > 0.03 , scales::percent(porcentaje, accuracy = .1),"")),
+            position = position_stack(vjust = .5),
+            show.legend = FALSE,
+            size = 4,
+            color = rep('white'))+
+  ggtitle("Grado de acuerdo: Con llegada de migrantes aumenta el desempleo") 
+  labs(caption =  paste0("Fuente: Elaboración propia en base a ELSOC (n = ",
+                         dim[1],")"))
+threat
+
+#---------------------------------------------------------------------------#
+
+#Paso 1
+#dropeamos el nivel -999 de la variable:
+df_study1$in_amigcl1 <- car::recode(df_study1$in_amigcl,
+                                    "c('Strongly disagree','Disagree')='Desacuerdo';
+                                    'Neither disagree nor agree'='Ni de acuerdo, ni desacuerdo';
+                                    c('Agree','Strongly agree')='Deacuerdo'")
+
+df_study1$in_amigcl1 <- factor(df_study1$in_amigcl1,
+                               labels =  c('Deacuerdo',
+                                          'Ni de acuerdo, ni desacuerdo',
+                                          'Desacuerdo'))
+
+a1 <- data.frame(table(haven::as_factor(df_study1$in_amigcl1),
+                       haven::as_factor(df_study1$ola),
+                       haven::as_factor(df_study1$idencuesta)))
+
+a1 <- a1 %>% filter(Var2 %in% c(2016,2018,2021))
+
+a1 <- a1 %>% dplyr::filter(Freq>0)
+names(a1) <- c("g_civico","ola","folio","Freq")
+a2 <- a1 %>% group_by(ola) %>% transmute(porcentaje=Freq/sum(Freq)) 
+a2$ola <- NULL
+datos.d.1<- cbind(a1,a2)
+datos.d.1 <- na.omit(datos.d.1)
+#Paso 2
+b1 <- data.frame(table(haven::as_factor(df_study1$in_amigcl1),
+                       haven::as_factor(df_study1$ola)))
+
+b1 <-b1 %>% filter(Var2 %in% c(2016,2018,2021))
+
+names(b1) <- c("g_civico","ola","Freq")
+b1 <- b1 %>% dplyr::filter(Freq>0)
+b2 <- b1 %>% 
+  group_by(ola) %>% 
+  transmute(porcentaje=Freq/sum(Freq)) %>% 
+  mutate(folio = 1)
+b2$ola <- NULL
+etiquetas.d.1<- cbind(b1,b2)
+etiquetas.d.1 <- na.omit(etiquetas.d.1)
+dim<- sjmisc::frq(datos.d.1$folio) %>% data.frame() %>% filter(frq==2) %>% dim()
+#-----------------------------------------------------------------------------#
+amigos <- ggplot(datos.d.1, aes(x = ola, fill = g_civico, stratum = g_civico,
+                             alluvium = folio, y = porcentaje)) +
+  ggalluvial::geom_flow(alpha = .66) + 
+  ggalluvial::geom_stratum(linetype = 0) +
+  scale_y_continuous(labels = scales::percent) + 
+  # scale_x_discrete(labels=c(2019,,2))+  
+  ylab(label = NULL) +
+  xlab(label = NULL) + 
+  theme(legend.position = 'right',
+        legend.title = element_blank()) +
+  scale_fill_viridis_d(begin = 0, end = .85, direction = -1, option = 'viridis') +
+  geom_text(data = etiquetas.d.1, 
+            aes(label = ifelse(porcentaje > 0.03 , scales::percent(porcentaje, accuracy = .1),"")),
+            position = position_stack(vjust = .5),
+            show.legend = FALSE,
+            size = 4,
+            color = rep('white'))+
+  ggtitle("Grado de acuerdo: Que migrantes tengan amigos chilenos") 
+  labs(caption =  paste0("Fuente: Elaboración propia en base a ELSOC (n = ",
+                         dim[1],")"))
+amigos
+
+df_study1$in_amreal1 <- car::recode(df_study1$in_amreal,
+                                    "c('Strongly disagree','Disagree','Neither disagree nor agree')='Desacuerdo';
+                                    c('Agree','Strongly agree')='De acuerdo'")
+
+df_study1$in_amreal1 <- factor(df_study1$in_amreal1,
+                               labels =  c('De acuerdo: desempleo aumenta',
+                                          'Desacuerdo'))
+
+df_study1$in_simpat1 <- 
+  car::recode(as_numeric(df_study1$in_simpat),
+              "1:2='Ninguna o poca simpatía';3:5='Simpatía'")
+
+df_study1$in_simili1 <- 
+  car::recode(as_numeric(df_study1$in_simili),
+              "1:3='Ninguna o poca similitud';4:5='Similaridad'")
+
+a1 <- 
+prop.table(table(df_study1$in_amreal1,df_study1$ola),margin = 2) %>% 
+  as.data.frame() %>% 
+  filter(Var1 == "De acuerdo: desempleo aumenta")  
+a2 <- 
+prop.table(table(df_study1$in_simpat1,df_study1$ola),margin = 2) %>% 
+  as.data.frame() %>% 
+  filter(Var1 == "Ninguna o poca simpatía")
+
+a3 <- 
+prop.table(table(df_study1$in_simili1,df_study1$ola),margin = 2) %>% 
+  as.data.frame() %>% 
+  filter(Var1 == "Ninguna o poca similitud")
+
+df<- rbind(a1,a2,a3)
+df$Var1 <- 
+  factor(df$Var1,
+         levels =c("Ninguna o poca simpatía","Ninguna o poca similitud",
+                   "De acuerdo: desempleo aumenta") )
+
+ggplot(data=df, aes(x=Var2, y=Freq, color = Var1,
+                    group=Var1,
+                    label = paste0(round((Freq)*100,2),'%'))) +
+  geom_line()+
+  geom_point() +
+  geom_text(hjust=0, vjust=0) + 
+  scale_x_discrete("Año") +
+  scale_y_continuous("Porcentaje") +  
+  theme_minimal() +
+  theme(legend.position="bottom",
+        legend.title=element_blank())
+ggsave(filename = 'figure_emol.png',device = 'png',
+       plot = last_plot(),path = "output/images",
+       units = "cm",width = 10,height = 10,scale = 2)
+
+ggsave(filename = 'figure_emol.pdf',device = 'pdf',
+       plot = last_plot(),path = "output/images",
+       units = "cm",width = 10,height = 10,scale = 2)
+```
+
+```{r grafico-linea-e.3}
+e.3 <- ggplot(subset.e.3,aes(y = porcentaje, x = ola, color = variable, group = variable,
+               label = as.character(scales::percent(porcentaje, accuracy = .1)))) +
+    geom_line(size = 1) +
+    geom_point(size = 1.8) +
+    scale_y_continuous(labels = scales::percent,
+                       limits = c(0,1)) +
+    ylab(label = NULL) +
+    xlab(label = NULL) +
+    scale_color_viridis_d(begin = .2, end = .7, direction = 1, option = 'viridis') +
+    facet_wrap(.~zona)+
+    geom_text(vjust = -0.8,
+    posilinetion = position_dodge(width = .9),
+    size= 1.75) +
+    theme(legend.position = 'top',
+          legend.title = element_blank()) +
+    ggtitle('Cambio de frecuencias en el grado de confianza "Nada" en algunas instituciones')
+
+e.3
+```
 
@@
 dev.off()
+
+pdf(here::here("output/images/figure1.pdf"),
+    # units = "cm",
+    # res = 100,
+    width = 15,
+    height = 15)
+likert1
+dev.off() 
 ```
@@
 png(file=here::here("output/images/figure2.jpg"),width=1000, height=800)
+corrplot::corrplot(cor_full,
+  title = "\nFigure 2: Polychoric correlation matrix for Conviviality, Identity and Threat items",
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  # title = "\n Full sample"
+  )
+  text(-3, 1,
+       adj = 0,
+         paste0("Note: Authors calculation based on ELSOC Survey ",
+                "(n=",dim(df_corr)[1],").",
+                "Polychoric correlations for each item for Conviviality, Identity and Threat.",
+                "\n[PER/VEN] represent Peruvians and Venezuelans, the different target groups included in each statement."),   
+       cex = 0.79)
+dev.off()
+
+pdf(file=here::here("output/images/figure2.pdf"),width=13, height=11)
 corrplot::corrplot(cor_full,

### 2023-09-28 production/prod_analysis.Rmd
re-estimacion con education distributive justice
- commit inicial desde fork frontiers
- nueva estimacion regresion
- solo promedio escolaridad como var nivel 2

--- 2022-05-19 production/prod_analysis.Rmd
@@
 	echo = FALSE,
+	results = "asis",
 	message = FALSE,
@@
 
-# Análisis descriptivo
+# Descriptive analysis
+
+## Univariate
 
@@
 pacman::p_load(dplyr,       # Manipulacion de datos
-               car,         # recodificar variables
-               sjlabelled,  # etiquetado de variables
-               sjmisc,      # descriptivos y frecuencias
-               sjPlot,      # tablas, plots y descriptivos
                knitr,
                kableExtra,
-               summarytools,
-               lavaan
+               summarytools
                )
-load(file = here::here("input/data-proc/study_1.RData"))
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
 
@@
 # Df variables nivel 1
-df<- dfSummary(df_study1 %>% select(in_simpat:edad),
+# df<- dfSummary(df_study1_long %>% select(in_simpat:edad),
+#                plain.ascii = FALSE,
+#                style = "grid",
+#                tmp.img.dir = "/tmp",
+#                graph.magnif = TRUE,
+#                headings = F,  # encabezado
+#                varnumbers = F, # num variable
+#                labels.col = T, # etiquetas
+#                na.col = F,    # missing
+#                graph.col = T, # plot
+#                valid.col = T, # n valido
+#                col.widths = c(700,10,10,10,10)
+#                )
+# df$Variable <- NULL # delete variable column
+# view(df,
+#      file = here::here("output/tables/desc01.html")
+#      ) # Ver tabla en un archivo HTML
+# webshot::webshot(url =here::here("output/tables/desc01.html"),
+#         file =here::here("output/tables/desc01.png")) # Remover archivo HTML
+# # print(df, method = "render")
+
+# Df variables nivel 1 __________________________________________________________
+df_study1 <- df_study1_long_comunas %>% select(just_educ:merit_talent,
+                                               ahead_family:sexo) 
+lab_study1 <- sjlabelled::get_label(df_study1) 
+df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
+
+df<- dfSummary(x = df_study1,
                plain.ascii = FALSE,
@@
                valid.col = T, # n valido
-               col.widths = c(700,10,10,10,10)
+               col.widths = c(700,10,10,10,10,10)
                )
-df$Variable <- NULL # delete variable column
+# df$Variable <- NULL # delete variable column
 view(df,
-     file = here::here("output/tables/desc01.html")
+     file = here::here("output/tables/desc02.html")
      ) # Ver tabla en un archivo HTML
-webshot::webshot(url =here::here("output/tables/desc01.html"),
-        file =here::here("output/tables/desc01.png")) # Remover archivo HTML
-print(df, method = "render")
+webshot::webshot(url =here::here("output/tables/desc02.html"),
+        file =here::here("output/tables/desc02.png")) # Remover archivo HTML
+# print(df, method = "render")
 ```
+
+## Correlation
+
+```{r cor-plot, fig.dim=c(10,6)}
+rm(list=ls())
+if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               sjPlot,
+               sjmisc,
+               summarytools,
+               lavaan
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# Full sample_______________________________________________________________________
+mat_full<- 
+df_study1_long_comunas %>%
+  # filter(ola==1) %>%
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+         ) %>% 
+  psych::mixedCor()
+
+cor_full <- as.matrix(mat_full$rho)
+diag(cor_full) = NA #set diagonal values to NA
+# Set Row names of the matrix
+rownames(cor_full) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
+#set Column names of the matrix
+colnames(cor_full) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+#Plot the matrix using corrplot
+corrplot::corrplot(cor_full,
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  title = "\n Full sample")
+
+# Wave 1_______________________________________________________________________
+mat_1<- 
+df_study1_long_comunas %>%
+  filter(ola==1) %>%
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+         ) %>% 
+  psych::mixedCor()
+
+cor_1 <- as.matrix(mat_1$rho)
+diag(cor_1) = NA #set diagonal values to NA
+# Set Row names of the matrix
+rownames(cor_1) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
+#set Column names of the matrix
+colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+#Plot the matrix using corrplot
+corrplot::corrplot(cor_1,
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  title = "\n Wave 1")
+
+# Wave 2_______________________________________________________________________
+mat_2<-   
+df_study1_long_comunas %>%
+  filter(ola==2) %>% 
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+         ) %>% 
+  psych::mixedCor()
+
+cor_2 <- as.matrix(mat_2$rho)
+diag(cor_2) = NA #set diagonal values to NA
+# Set Row names of the matrix
+rownames(cor_2) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
+#set Column names of the matrix
+colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+#Plot the matrix using corrplot
+corrplot::corrplot(cor_2,
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  title = "\n Wave 2")
+
+# Wave 3_______________________________________________________________________
+mat_3 <- 
+df_study1_long_comunas %>%
+  filter(ola==3) %>% 
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+         ) %>% 
+  psych::mixedCor()
+
+cor_3 <- as.matrix(mat_3$rho)
+diag(cor_3) = NA #set diagonal values to NA
+# Set Row names of the matrix
+rownames(cor_3) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
+#set Column names of the matrix
+colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+#Plot the matrix using corrplot
+corrplot::corrplot(cor_3,
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  title = "\n Wave 3")
+
+# Wave 4_______________________________________________________________________
+mat_4 <- 
+df_study1_long_comunas %>%
+  filter(ola==4) %>% 
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+         ) %>% 
+  psych::mixedCor()
+
+cor_4 <- as.matrix(mat_4$rho)
+diag(cor_4) = NA #set diagonal values to NA
+# Set Row names of the matrix
+rownames(cor_4) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
+#set Column names of the matrix
+colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+#Plot the matrix using corrplot
+corrplot::corrplot(cor_4,
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  title = "\n Wave 4")
+
+# Wave 6_______________________________________________________________________
+mat_6 <- 
+df_study1_long_comunas %>%
+  filter(ola==6) %>% 
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+         ) %>% 
+  psych::mixedCor()
+
+cor_6 <- as.matrix(mat_6$rho)
+diag(cor_6) = NA #set diagonal values to NA
+# Set Row names of the matrix
+rownames(cor_6) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
+#set Column names of the matrix
+colnames(cor_6) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+#Plot the matrix using corrplot
+corrplot::corrplot(cor_6,
+  method = "color",
+  addCoef.col = "#000390",
+  type = "upper",
+  tl.col = "black",
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
+  na.label = "-",
+  title = "\n Wave 6")
+```
+
+# Models
+
+## A. Education distribution justice
+
+```{r anxiety}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+#load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:esc_cit17) %>% 
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1 <- "ola"
+h2 <- "ola+merit_effort+merit_talent"
+h3 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork"
+h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+esc_cit17"
+h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+esc_cit17+educ+quintil1+ess+pos_id+edad+sexo"
+
+# A. Anxiety (in_ansied)
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(educ.null, educ1, educ2, educ3, educ4, educ5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: Familia",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: trabajo duro",
+                                      "Promedio escolaridad comuna",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"
+                              )
+          #omit.coef = omit
+          )
+```

### 2023-09-28 production/prod_prep.Rmd
re-estimacion con education distributive justice
- commit inicial desde fork frontiers
- nueva estimacion regresion
- solo promedio escolaridad como var nivel 2

@@
 
-load(url("https://dataverse.harvard.edu/api/access/datafile/6160173")) #ELSOC
-save(elsoc_long_2016_2021,file = here::here("input/data-original/elsoc_long.Rdata"))
+load(url("https://dataverse.harvard.edu/api/access/datafile/7245118")) #ELSOC
+save(elsoc_long_2016_2022.2,file = here::here("input/data-original/elsoc_long.Rdata"))
+
 cit17<- read.table(url("https://dataverse.harvard.edu/api/access/datafile/5216836"), 
@@
 
-elsoc_long_2016_2021[elsoc_long_2016_2021 ==-999] <- NA
-elsoc_long_2016_2021[elsoc_long_2016_2021 ==-888] <- NA
-elsoc_long_2016_2021[elsoc_long_2016_2021 ==-777] <- NA
-elsoc_long_2016_2021[elsoc_long_2016_2021 ==-666] <- NA
+elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-999] <- NA
+elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-888] <- NA
+elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-777] <- NA
+elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-666] <- NA
 
 # Attitudes toward inmigrants__________________________________________________
-elsoc_long_2016_2021 <- 
-elsoc_long_2016_2021 %>% 
+elsoc_long_2016_2022.2 <- 
+elsoc_long_2016_2022.2 %>% 
   #create new variables
-  mutate(in_simpat=r09,
+  mutate(just_educ=d02_02,
          sexo=m0_sexo,
-         in_simili=r10,
-         in_ansied=r11,
-         in_valfam=r12_01,
-         in_valami=r12_02,
-         in_amsimb=r12_03,
-         in_amreal=r12_04,
-         in_acpier=r12_05,
-         in_acadop=r12_06,
-         in_amigcl=r12_07) %>% 
+         merit_effort=c18_09,
+         merit_talent=c18_10,
+         ahead_family=r12_02,
+         ahead_educ=r12_03,
+         ahead_ambition=r12_04,
+         ahead_hardwork=r12_05) %>% 
   sjlabelled::drop_labels(., drop.na = FALSE)
@@
 #
-sjmisc::frq(elsoc_long_2016_2021$in_simpat)
-elsoc_long_2016_2021$in_simpat <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_simpat, 
-                      label = "Sympathy for [PER/VEN] living in Chile") 
-elsoc_long_2016_2021$in_simpat <-
-sjlabelled::set_labels(x = elsoc_long_2016_2021$in_simpat,
-                       labels = c("Very little or not at all", "A little",
-                                  "Somewhat","Quite a lot","A lot"))  
-sjmisc::frq(elsoc_long_2016_2021$in_simpat)
+sjmisc::frq(elsoc_long_2016_2022.2$just_educ)
+elsoc_long_2016_2022.2$just_educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$just_educ, 
+                      label = "Education distributive justice")
 
 #
-sjmisc::frq(elsoc_long_2016_2021$in_simili)
-elsoc_long_2016_2021$in_simili <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_simili, 
-                      label = "Similarity between Chileans and [PER/VEN]") 
-elsoc_long_2016_2021$in_simili <-
-sjlabelled::set_labels(x = elsoc_long_2016_2021$in_simili,
-                       labels = c("Not similar", "Slightly similar", "Somewhat similar",
-                                  "Quite similar", "Very similar"))
-sjmisc::frq(elsoc_long_2016_2021$in_simili)
+sjmisc::frq(elsoc_long_2016_2022.2$merit_effort)
+elsoc_long_2016_2022.2$merit_effort <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$merit_effort, 
+                      label = "People are rewarded for their efforts")
 
 #
-sjmisc::frq(elsoc_long_2016_2021$in_ansied)
-elsoc_long_2016_2021$in_ansied <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_ansied, 
-                      label = "Interaction anxiety with [PER/VEN].") 
-elsoc_long_2016_2021$in_ansied <-
-sjlabelled::set_labels(x = elsoc_long_2016_2021$in_ansied,
-                       labels = c("Very uncomfortable", "Uncomfortable", 
-                                  "Neither comfortable nor uncomfortable",
-                                  "Comfortable", "Very comfortable"))
-sjmisc::frq(elsoc_long_2016_2021$in_ansied)
+sjmisc::frq(elsoc_long_2016_2022.2$merit_talent)
+elsoc_long_2016_2022.2$merit_talent <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$merit_talent, 
+                      label = "People are rewarded for their intelligence")
 
-for (i in c("in_valfam","in_valami","in_amsimb","in_amreal",
-            "in_acpier","in_acadop","in_amigcl")) {
-elsoc_long_2016_2021[[i]] <-
-sjlabelled::set_labels(x = elsoc_long_2016_2021[[i]],
+#
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_family)
+elsoc_long_2016_2022.2$ahead_family <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_family, 
+                      label = "Coming from a wealthy family")
+
+#
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_educ)
+elsoc_long_2016_2022.2$ahead_educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_educ, 
+                      label = "Having a good level of education")
+
+#
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_ambition)
+elsoc_long_2016_2022.2$ahead_ambition <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_ambition, 
+                      label = "Having ambition")
+
+#
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_hardwork)
+elsoc_long_2016_2022.2$ahead_hardwork <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_hardwork, 
+                      label = "Hard work")
+
+for (i in c("just_educ","merit_effort","merit_talent",
+            "ahead_family","ahead_educ","ahead_ambition", "ahead_hardwork")) {
+elsoc_long_2016_2022.2[[i]] <-
+sjlabelled::set_labels(x = elsoc_long_2016_2022.2[[i]],
                        labels = c("Strongly disagree","Disagree",
@@
 
-sjmisc::frq(elsoc_long_2016_2021$in_valfam)
-elsoc_long_2016_2021$in_valfam <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_valfam, 
-                      label = "Family value of migrant friends [PER/VEN]") 
-sjmisc::frq(elsoc_long_2016_2021$in_valfam)
-
-sjmisc::frq(elsoc_long_2016_2021$in_valami)
-elsoc_long_2016_2021$in_valami <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_valami, 
-                      label = "Friends value of migrant friends [PER/VEN]") 
-sjmisc::frq(elsoc_long_2016_2021$in_valami)
-
-sjmisc::frq(elsoc_long_2016_2021$in_amsimb)
-elsoc_long_2016_2021$in_amsimb <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_amsimb, 
-                      label = "Loose of identity because of migrants (reverse)")
-sjmisc::frq(elsoc_long_2016_2021$in_amsimb) 
-# Revert item
-elsoc_long_2016_2021$in_amsimb<- 
-  sjmisc::rec(elsoc_long_2016_2021$in_amsimb,rec = "rev")
-sjmisc::frq(elsoc_long_2016_2021$in_amsimb) # check reverted item
-
-sjmisc::frq(elsoc_long_2016_2021$in_amreal)
-elsoc_long_2016_2021$in_amreal <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_amreal, 
-                      label = "Unemployment increases") 
-sjmisc::frq(elsoc_long_2016_2021$in_amreal)
-
-sjmisc::frq(elsoc_long_2016_2021$in_acpier)
-elsoc_long_2016_2021$in_acpier <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_acpier, 
-                      label = "Migrant keeping their customs") 
-sjmisc::frq(elsoc_long_2016_2021$in_acpier)
-
-sjmisc::frq(elsoc_long_2016_2021$in_acadop)
-elsoc_long_2016_2021$in_acadop <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_acadop, 
-                      label = "Adoption of Chilean customs") 
-sjmisc::frq(elsoc_long_2016_2021$in_acadop)
-
-sjmisc::frq(elsoc_long_2016_2021$in_amigcl)
-elsoc_long_2016_2021$in_amigcl <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$in_amigcl, 
-                      label = "migrants [PER/VEN] have Chilean friends") 
-sjmisc::frq(elsoc_long_2016_2021$in_amigcl)
-
-# Average variables___________________________________________________________
-
-elsoc_long_2016_2021 <- elsoc_long_2016_2021 %>% 
-  mutate(conviviality=as.numeric((in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5),
-         identity= as.numeric((in_simili+in_amsimb+in_acpier+in_acadop)/4),
-         threat = as.numeric(in_amreal))
-
-sjmisc::frq(elsoc_long_2016_2021$conviviality)
-elsoc_long_2016_2021$conviviality <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$conviviality, 
-                      label = "Conviviality")
-sjmisc::frq(elsoc_long_2016_2021$conviviality)
-
-sjmisc::frq(elsoc_long_2016_2021$identity)
-elsoc_long_2016_2021$identity <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$identity, 
-                      label = "Identity")
-sjmisc::frq(elsoc_long_2016_2021$identity)
-
+sjmisc::frq(elsoc_long_2016_2022.2$just_educ)
+sjmisc::frq(elsoc_long_2016_2022.2$merit_effort)
+sjmisc::frq(elsoc_long_2016_2022.2$merit_talent)
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_family)
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_educ)
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_ambition)
+sjmisc::frq(elsoc_long_2016_2022.2$ahead_hardwork)
 # Socioeconomic_________________________________________________________________
@@
 # Education_______________________________________
-elsoc_long_2016_2021$educ <- 
-  car::recode(elsoc_long_2016_2021$m01,
+elsoc_long_2016_2022.2$educ <- 
+  car::recode(elsoc_long_2016_2022.2$m01,
               "c(1,2,3)=1;c(4,5)=2;c(6,7)=3;c(8,9,10)=4; c(-888,-999)=NA")
-elsoc_long_2016_2021$educ <-
-  factor(elsoc_long_2016_2021$educ,
+elsoc_long_2016_2022.2$educ <-
+  factor(elsoc_long_2016_2022.2$educ,
          labels = c("Primary","High school","Technical","Universitary"))
@@
 #reverse education, reference level is the highest level
-elsoc_long_2016_2021$educ <- forcats::fct_rev(elsoc_long_2016_2021$educ)
+#elsoc_long_2016_2022.2$educ <- forcats::fct_rev(elsoc_long_2016_2022.2$educ)
 
-elsoc_long_2016_2021$educ <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$educ,
+elsoc_long_2016_2022.2$educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$educ,
                       label = "Education")
-sjmisc::frq(elsoc_long_2016_2021$educ)
+sjmisc::frq(elsoc_long_2016_2022.2$educ)
 
 #Recoding of education to years based on casen 2017.
-elsoc_long_2016_2021$educyear<- as.numeric(
-  car::recode(elsoc_long_2016_2021$m01, 
+elsoc_long_2016_2022.2$educyear<- as.numeric(
+  car::recode(elsoc_long_2016_2022.2$m01, 
               "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
@@
 
-elsoc_long_2016_2021$educyear <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$educyear,
+elsoc_long_2016_2022.2$educyear <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$educyear,
                       label = "Education in years")
 
-class(elsoc_long_2016_2021$educyear)
-sjmisc::frq(elsoc_long_2016_2021$educyear)
+class(elsoc_long_2016_2022.2$educyear)
+sjmisc::frq(elsoc_long_2016_2022.2$educyear)
 
@@
 #Impute midpoint of income ranges
-elsoc_long_2016_2021$m30_rec <-
-as.numeric(car::recode(elsoc_long_2016_2021$m30,
+elsoc_long_2016_2022.2$m30_rec <-
+as.numeric(car::recode(elsoc_long_2016_2022.2$m30,
            "1=110000;2=251000;3=305000;4=355000;5=400000;
@@
 #Impute midpoint of income ranges (2021)
-elsoc_long_2016_2021$m30b_rec <-
-as.numeric(car::recode(elsoc_long_2016_2021$m30b,
+elsoc_long_2016_2022.2$m30b_rec <-
+as.numeric(car::recode(elsoc_long_2016_2022.2$m30b,
            "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))
 
-sjmisc::frq(elsoc_long_2016_2021$m30b_rec)
+sjmisc::frq(elsoc_long_2016_2022.2$m30b_rec)
 
 #Recode DK/DA of Income to NA
-elsoc_long_2016_2021$m29_rec <-
-  as.numeric(car::recode(elsoc_long_2016_2021$m29,"c(-888,-999)=NA"))
+elsoc_long_2016_2022.2$m29_rec <-
+  as.numeric(car::recode(elsoc_long_2016_2022.2$m29,"c(-888,-999)=NA"))
 
 #replace NA of income with new imputed variable
-elsoc_long_2016_2021$m29_imp <- 
-  ifelse(test = !is.na(elsoc_long_2016_2021$m29_rec),
-         yes =  elsoc_long_2016_2021$m29_rec,
-         no =  elsoc_long_2016_2021$m30_rec)
-summary(elsoc_long_2016_2021$m29_imp)
+elsoc_long_2016_2022.2$m29_imp <- 
+  ifelse(test = !is.na(elsoc_long_2016_2022.2$m29_rec),
+         yes =  elsoc_long_2016_2022.2$m29_rec,
+         no =  elsoc_long_2016_2022.2$m30_rec)
+summary(elsoc_long_2016_2022.2$m29_imp)
 
-elsoc_long_2016_2021$m29_imp <- 
-  ifelse(test = is.na(elsoc_long_2016_2021$m29_imp),
-         yes =  elsoc_long_2016_2021$m30b_rec,
-         no =  elsoc_long_2016_2021$m29_imp)
-summary(elsoc_long_2016_2021$m29_imp)
+elsoc_long_2016_2022.2$m29_imp <- 
+  ifelse(test = is.na(elsoc_long_2016_2022.2$m29_imp),
+         yes =  elsoc_long_2016_2022.2$m30b_rec,
+         no =  elsoc_long_2016_2022.2$m29_imp)
+summary(elsoc_long_2016_2022.2$m29_imp)
 
 # deflate at each year's prices
-elsoc_long_2016_2021$deflactor <-
-  with(elsoc_long_2016_2021, case_when(
+elsoc_long_2016_2022.2$deflactor <-
+  with(elsoc_long_2016_2022.2, case_when(
     ola == 2016 ~ 113.88 / 123.82,
@@
 # N Household:
-elsoc_long_2016_2021 <-
-  elsoc_long_2016_2021 %>%
+elsoc_long_2016_2022.2 <-
+  elsoc_long_2016_2022.2 %>%
   mutate(n_hogar =
@@
                             ola == 5 ~ m54))
-sjmisc::frq(elsoc_long_2016_2021$n_hogar)
+sjmisc::frq(elsoc_long_2016_2022.2$n_hogar)
 
 #Recode DK/DA to NA
-elsoc_long_2016_2021$n_hogar_r<-
-  car::recode(elsoc_long_2016_2021$n_hogar,"c(-888,-999)=NA")
+elsoc_long_2016_2022.2$n_hogar_r<-
+  car::recode(elsoc_long_2016_2022.2$n_hogar,"c(-888,-999)=NA")
 
 # Per capita household income:
-elsoc_long_2016_2021$ing_pc <- 
-  (elsoc_long_2016_2021$m29_imp/elsoc_long_2016_2021$n_hogar_r)
+elsoc_long_2016_2022.2$ing_pc <- 
+  (elsoc_long_2016_2022.2$m29_imp/elsoc_long_2016_2022.2$n_hogar_r)
 
-elsoc_long_2016_2021$ing_pc <-
-sjlabelled::set_label(x = elsoc_long_2016_2021$ing_pc,
+elsoc_long_2016_2022.2$ing_pc <-
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$ing_pc,
                       label = "Household income per capita")
 
-sjmisc::descr(elsoc_long_2016_2021$ing_pc)
+sjmisc::descr(elsoc_long_2016_2022.2$ing_pc)
 
 # Compute income quintiles
-elsoc_long_2016_2021 <- elsoc_long_2016_2021 %>% 
+elsoc_long_2016_2022.2 <- elsoc_long_2016_2022.2 %>% 
   group_by(ola) %>% 
@@
 
-elsoc_long_2016_2021$quintil <- 
-  factor(elsoc_long_2016_2021$quintil,
+elsoc_long_2016_2022.2$quintil <- 
+  factor(elsoc_long_2016_2022.2$quintil,
          levels = c(1, 2, 3, 4, 5),
@@
 #reverse quintile, reference level is the highest quintile
-elsoc_long_2016_2021$quintil <- forcats::fct_rev(elsoc_long_2016_2021$quintil)
+#elsoc_long_2016_2022.2$quintil <- forcats::fct_rev(elsoc_long_2016_2022.2$quintil)
 
-elsoc_long_2016_2021$quintil <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$quintil,
+elsoc_long_2016_2022.2$quintil <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$quintil,
                       label = "Household income quintile per capita")
 
-sjmisc::frq(elsoc_long_2016_2021$quintil)
+sjmisc::frq(elsoc_long_2016_2022.2$quintil)
 
 #include new quintile category with missing cases
-elsoc_long_2016_2021$quintil1<-
-  car::recode(elsoc_long_2016_2021$quintil, 
+elsoc_long_2016_2022.2$quintil1<-
+  car::recode(elsoc_long_2016_2022.2$quintil, 
               "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")
 
-elsoc_long_2016_2021$quintil1 <- factor(elsoc_long_2016_2021$quintil1, c("Q5","Q4","Q3","Q2","Q1","QNA"))
+#elsoc_long_2016_2022.2$quintil1 <- factor(elsoc_long_2016_2022.2$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))
 
-elsoc_long_2016_2021$quintil1 <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$quintil1,
+elsoc_long_2016_2022.2$quintil1 <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$quintil1,
                       label = "Household income quintile per capita (NA)") 
-sjmisc::frq(elsoc_long_2016_2021$quintil1)
+sjmisc::frq(elsoc_long_2016_2022.2$quintil1)
 
 #Age Ranges
-elsoc_long_2016_2021$edad <- 
-  factor(car::recode(elsoc_long_2016_2021$m0_edad, 
+elsoc_long_2016_2022.2$edad <- 
+  factor(car::recode(elsoc_long_2016_2022.2$m0_edad, 
                      "18:29=1;30:49=2;50:64=3;65:150=4"),
          labels = c('18-29', '30-49', '50-64', '65 or more'))
-elsoc_long_2016_2021$edad <-
-  sjlabelled::set_label(elsoc_long_2016_2021$edad, 
+elsoc_long_2016_2022.2$edad <-
+  sjlabelled::set_label(elsoc_long_2016_2022.2$edad, 
                         label = c("Age groups")) 
@@
 
-elsoc_long_2016_2021$ess <- 
-as.numeric(elsoc_long_2016_2021$d01_01)
-sjmisc::frq(elsoc_long_2016_2021$ess)
-elsoc_long_2016_2021$ess <- sjmisc::rec(elsoc_long_2016_2021$ess,rec = "rev")
-sjmisc::frq(elsoc_long_2016_2021$ess)
+elsoc_long_2016_2022.2$ess <- 
+as.numeric(elsoc_long_2016_2022.2$d01_01)
+sjmisc::frq(elsoc_long_2016_2022.2$ess)
+#elsoc_long_2016_2022.2$ess <- sjmisc::rec(elsoc_long_2016_2022.2$ess,rec = "rev")
+#sjmisc::frq(elsoc_long_2016_2022.2$ess)
 
-elsoc_long_2016_2021$ess <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$ess,
-                      label = "Subjective Social Status: individual (reverse)")
+elsoc_long_2016_2022.2$ess <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$ess,
+                      label = "Subjective Social Status: individual")
 
-sjmisc::frq(elsoc_long_2016_2021$ess)
-# sjPlot::plot_frq(elsoc_long_2016_2021$ess)
-# Subjective social status: migrants (w04 and w05)____________________________
-elsoc_long_2016_2021$ess_inm <- 
-as.numeric(elsoc_long_2016_2021$d01_04)
-
-sjmisc::frq(elsoc_long_2016_2021$ess_inm)
-elsoc_long_2016_2021$ess_inm <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$ess_inm,
-                      label = "Subjective Social Status: inmigrants")
-
-sjmisc::frq(elsoc_long_2016_2021$ess_inm)
-# sjPlot::plot_frq(elsoc_long_2016_2021$ess_inm)
-
-# Nationality of the interviewee____________________________
-
-elsoc_long_2016_2021$nation <- 
-elsoc_long_2016_2021$m45
-sjmisc::frq(elsoc_long_2016_2021$nation)
-
-elsoc_long_2016_2021$nation <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$nation, 
-                      label = "Nationality") 
-elsoc_long_2016_2021$nation <-
-sjlabelled::set_labels(x = elsoc_long_2016_2021$nation,
-                       labels = c("Chilean","Peruvian","Argentina","Colombian",
-                                  "Bolivian", "Haitian","Spanish","Other"))
-
-sjmisc::frq(elsoc_long_2016_2021$nation)
-
-# Known: Immigrant [PER/VEN]____________________________
-
-# sjmisc::find_var(elsoc_long_2016_2021,"Conocidos")
-sjmisc::frq(elsoc_long_2016_2021$r05_01)
-elsoc_long_2016_2021$know_inm <- 
-elsoc_long_2016_2021$r05_01
-
-elsoc_long_2016_2021$know_inm <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$know_inm, 
-                      label = "Known: Immigrants") 
-elsoc_long_2016_2021$know_inm <-
-sjlabelled::set_labels(x = elsoc_long_2016_2021$know_inm,
-                       labels = c("None","Few","Some","Quite a few","Many"))
-sjmisc::frq(elsoc_long_2016_2021$know_inm)
-
-# Level of contact: Migrant friends_________________________
-sjmisc::frq(elsoc_long_2016_2021$r05_02)
-elsoc_long_2016_2021$frien_inm <- 
-elsoc_long_2016_2021$r05_02
-
-elsoc_long_2016_2021$frien_inm <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$frien_inm, 
-                      label = "Level of contact: Inmigrant friends")
-
-elsoc_long_2016_2021$frien_inm <-
-sjlabelled::set_labels(x = elsoc_long_2016_2021$frien_inm,
-                       labels = c("None","Few","Some","Quite a few","Many"))
-sjmisc::frq(elsoc_long_2016_2021$frien_inm)
+sjmisc::frq(elsoc_long_2016_2022.2$ess)
+# sjPlot::plot_frq(elsoc_long_2016_2022.2$ess)
 
 # Political position ____________________________
-elsoc_long_2016_2021$pos_id <-
+elsoc_long_2016_2022.2$pos_id <-
 factor(
   car::recode(
-    elsoc_long_2016_2021$c15,
+    elsoc_long_2016_2022.2$c15,
     "c(11,12,-888,-999)='Does not identify';c(0,1,2,3,4)='Left';
@@
 
-elsoc_long_2016_2021$pos_id <- factor(elsoc_long_2016_2021$pos_id,
-                                      levels = levels(elsoc_long_2016_2021$pos_id))
+elsoc_long_2016_2022.2$pos_id <- factor(elsoc_long_2016_2022.2$pos_id,
+                                      levels = levels(elsoc_long_2016_2022.2$pos_id))
 
-elsoc_long_2016_2021$pos_id <- 
-sjlabelled::set_label(x = elsoc_long_2016_2021$pos_id, 
+elsoc_long_2016_2022.2$pos_id <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$pos_id, 
                       label = "Political identification")
 
-sjmisc::frq(elsoc_long_2016_2021$pos_id)
+sjmisc::frq(elsoc_long_2016_2022.2$pos_id)
 
-# CONTACT_______________________________________________________________
-
-# NATION to wave 1 (original)___________________
-df_nation <- elsoc_long_2016_2021 %>%  
-  select(idencuesta,nation_fix=nation,ola,muestra) %>% 
-  filter(ola %in% c(1,3)) %>% #keep wave 1 and 3 
-  group_by(idencuesta) %>% 
-  summarise(nation_fix=max(nation_fix,na.rm = F)) #keep the maximum value 
-sjmisc::frq(df_nation$nation_fix) # 2 NA
-
-elsoc <- elsoc_long_2016_2021 #dummy dataset _________________________________
-
-# merge the data with fixed variable:
-elsoc <- left_join(x = elsoc,y = df_nation,by ="idencuesta") 
-# sjmisc::frq(elsoc$nation_fix)
-
-table(elsoc$nation==elsoc$nation_fix) #how many cases are different ?
-elsoc$check_nation <- ifelse(elsoc$nation==elsoc$nation_fix,
-                             yes = "IGUALES",
-                             no = "DISTINTOS")
-# sjmisc::frq(elsoc$check_nation) #18 cases
-
-# edit, if value is different from original variable, set the value of wave 1
-elsoc$nation_t1 <- if_else(condition = elsoc$check_nation=="DISTINTOS",
-                           true = elsoc$nation,
-                           false = elsoc$nation_fix,
-                           missing = elsoc$nation_fix)
-
-table(elsoc$nation==elsoc$nation_t1)
-sjmisc::frq(elsoc$nation_t1)
-
-elsoc$nation <- NULL; elsoc$nation_fix <- NULL; elsoc$check_nation <- NULL
-
-elsoc$nation_t1 <- set_labels(elsoc$nation_t1,
-                              labels =c("Chilean","Peruvian",
-                                        "Argentina","Colombian",
-                                        "Bolivian", "Haitian","Spanish","Other"))
-elsoc$nation_t1 <- factor(elsoc$nation_t1,
-                          levels = 1:8,
-                          labels = c("Chilean","Peruvian",
-                                     "Argentina","Colombian",
-                                     "Bolivian", "Haitian","Spanish","Other"))
-elsoc$nation_t1 <- set_label(elsoc$nation_t1, label = "Nationality")
-sjmisc::frq(elsoc$nation_t1)
-# KNOW inmigrants _____________________________
-sjPlot::tab_xtab(elsoc$know_inm,
-                 elsoc$ola,show.cell.prc = T,show.na = T)
-sjmisc::frq(elsoc$know_inm)
-elsoc$know_inm_bi <- car::recode(elsoc$know_inm,
-                                                "1=0;2:5=1")
-sjmisc::frq(elsoc$know_inm_bi)
-
-know_inm_bi <- 
-elsoc %>% group_by(idencuesta) %>%
-  summarize(know_inm_bi_t4=know_inm_bi[ola==4],
-            know_inm_bi_t1=know_inm_bi[ola==1],
-            # know_inm_t1t4 = know_inm_bi[ola==4] - know_inm_bi[ola==1],
-            know_inm_t1t4 = paste0(know_inm_bi[ola==1],know_inm_bi[ola==4])
-            )
-sjmisc::frq(know_inm_bi$know_inm_t1t4)
-know_inm_bi$know_inm_t1t4 <- 
-  car::recode(know_inm_bi$know_inm_t1t4,
-              recodes = "c('0NA','NA0','00')='Stable, do not know';c('1NA','NA1','11')='Stable, know';'01'='Now know';'10'='No longer know'",as.factor = T)
-
-know_inm_bi$know_inm_t1t4 <- 
-  factor(know_inm_bi$know_inm_t1t4,
-         levels = c("Stable, do not know","Stable, know","Now know","No longer know"))
-sjmisc::frq(know_inm_bi$know_inm_t1t4)
-
-elsoc <- left_join(x = elsoc,y = know_inm_bi, by = "idencuesta")
-sjmisc::frq(elsoc$know_inm_t1t4)
-
-elsoc$know_inm_t1t4 <- set_label(elsoc$know_inm_t1t4,
-                                  label = "Know migrants (diff. t4 - t1)")
-
-sjmisc::frq(elsoc$know_inm_t1t4)
-# sjPlot::tab_xtab(elsoc$know_inm_t1t4,elsoc$ola)
-
-# FRIENDS inmigrants _____________________________
-sjPlot::tab_xtab(elsoc$frien_inm,
-                 elsoc$ola,show.cell.prc = T,show.na = T)
-sjmisc::frq(elsoc$frien_inm)
-elsoc$frien_inm_bi <- car::recode(elsoc$frien_inm,
-                                                "1=0;2:5=1")
-sjmisc::frq(elsoc$frien_inm_bi)
-
-frien_inm_bi <- 
-elsoc %>% group_by(idencuesta) %>%
-  summarize(frien_inm_bi_t4=frien_inm_bi[ola==4],
-            frien_inm_bi_t1=frien_inm_bi[ola==1],
-            # frien_inm_t1t4 = frien_inm_bi[ola==4] - frien_inm_bi[ola==1],
-            frien_inm_t1t4 = paste0(frien_inm_bi[ola==1],frien_inm_bi[ola==4])
-            )
-
-sjmisc::frq(frien_inm_bi$frien_inm_t1t4)
-frien_inm_bi$frien_inm_t1t4 <-
-  car::recode(frien_inm_bi$frien_inm_t1t4,
-              recodes = "c('0NA','NA0','00')='Stable, do not have friends';c('1NA','NA1','11')='Stable, have friends';'01'='Now have friends';'10'='No longer have friends';'NANA'=NA",as.factor = T)
-sjmisc::frq(frien_inm_bi$frien_inm_t1t4)
-
-frien_inm_bi$frien_inm_t1t4 <- 
-  factor(frien_inm_bi$frien_inm_t1t4,
-         levels = c("Stable, do not have friends","Stable, have friends",
-                    "Now have friends","No longer have friends"))
-sjmisc::frq(frien_inm_bi$frien_inm_t1t4)
-
-elsoc <- left_join(x = elsoc,y = frien_inm_bi, by = "idencuesta")
-sjmisc::frq(elsoc$frien_inm_t1t4)
-# View(elsoc[c("idencuesta","frien_inm_bi_t4","frien_inm_bi_t1","frien_inm_t1t4")])
-
-# # if NA in difference, impute data from t4
-# elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
-#                                 true = elsoc$frien_inm_bi_t4,
-#                                 false = elsoc$frien_inm_t1t4)
-# summary(elsoc$frien_inm_t1t4)
-# # if NA in difference, impute data from t1
-# elsoc$frien_inm_t1t4 <- if_else(condition = is.na(elsoc$frien_inm_t1t4),
-#                                 true = elsoc$frien_inm_bi_t1,
-#                                 false = elsoc$frien_inm_t1t4)
-sjmisc::frq(elsoc$frien_inm_t1t4)
-
-# elsoc$frien_inm_t1t4 <- factor(elsoc$frien_inm_t1t4,
-#                                levels = c(-1,0,1),
-#                                labels = c("No longer have",
-#                                           "Constant",
-#                                           "Now have"))
-
-# ref dejamos los 
-elsoc$frien_inm_t1t4 <- 
-  set_label(elsoc$frien_inm_t1t4,
-            label = "Have migrant friends (diff. t4 - t1)")
-
-sjmisc::frq(elsoc$frien_inm_t1t4)
-sjPlot::tab_xtab(elsoc$frien_inm_bi,elsoc$know_inm_bi)
 # Select variables______________________________________________________________
-elsoc_long_2016_2021<- elsoc
 
 df_study1 <- 
-elsoc_long_2016_2021 %>% 
+elsoc_long_2016_2022.2 %>% 
   filter(muestra == 1) %>% #keep original sample (vs refresh)
@@
          sexo,
-         cuestion_mig,
-         starts_with("in_"),
-         conviviality,
-         identity,
+         # cuestion_mig,
+         # starts_with("in_"),
+         just_educ,
+         merit_effort,
+         merit_talent,
+         ahead_family,
+         ahead_educ,
+         ahead_ambition,
+         ahead_hardwork,
          educ,educyear,
@@
          # ess_inm,
-         nation_t1,
+         # nation_t1,
          # know_inm,
@@
          # frien_inm_bi_t1,
-         know_inm_t1t4,
+         # know_inm_t1t4,
          # frien_inm,
          # frien_inm_bi,
-         frien_inm_t1t4,
+         # frien_inm_t1t4,
          pos_id,
@@
                      # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
-                     values_from = names(select(df_study1,comuna,comunacod,cuestion_mig,in_simpat:edad,sexo))
+                     values_from = names(select(df_study1,comuna,comunacod,just_educ:edad,sexo))
                      )
@@
 df_study1_wide$edad_5 <-df_study1_wide$edad_1
+df_study1_wide$edad_6 <-df_study1_wide$edad_1
 
@@
 df_study1_wide$sexo_5 <-df_study1_wide$sexo_1
+df_study1_wide$sexo_6 <-df_study1_wide$sexo_1
 
@@
 df_study1_wide$educ_5 <-df_study1_wide$educ_1
+df_study1_wide$educ_6 <-df_study1_wide$educ_1
 
@@
 df_study1_wide$educyear_5 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_6 <-df_study1_wide$educyear_1
 
@@
 df_study1_wide$quintil1_5 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_6 <-df_study1_wide$quintil1_1
 
@@
 df_study1_wide$ess_5 <-df_study1_wide$ess_1
+df_study1_wide$ess_6 <-df_study1_wide$ess_1
 
@@
 df_study1_wide$pos_id_5 <-df_study1_wide$pos_id_1
-
-df_study1_wide$nation_t1_2 <-df_study1_wide$nation_t1_1 # nation
-df_study1_wide$nation_t1_3 <-df_study1_wide$nation_t1_1
-df_study1_wide$nation_t1_4 <-df_study1_wide$nation_t1_1
-df_study1_wide$nation_t1_5 <-df_study1_wide$nation_t1_1
-
-df_study1_wide$know_inm_t1t4_2 <-df_study1_wide$know_inm_t1t4_1 # know immigrants
-df_study1_wide$know_inm_t1t4_3 <-df_study1_wide$know_inm_t1t4_1
-df_study1_wide$know_inm_t1t4_4 <-df_study1_wide$know_inm_t1t4_1
-df_study1_wide$know_inm_t1t4_5 <-df_study1_wide$know_inm_t1t4_1
-
-df_study1_wide$frien_inm_t1t4_2 <- df_study1_wide$frien_inm_t1t4_1 # immigrant friends
-df_study1_wide$frien_inm_t1t4_3 <- df_study1_wide$frien_inm_t1t4_1
-df_study1_wide$frien_inm_t1t4_4 <- df_study1_wide$frien_inm_t1t4_1
-df_study1_wide$frien_inm_t1t4_5 <- df_study1_wide$frien_inm_t1t4_1
+df_study1_wide$pos_id_6 <-df_study1_wide$pos_id_1
 
@@
 df_study1_wide$comuna_5 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_6 <-df_study1_wide$comuna_1
 
@@
 df_study1_wide$comunacod_5 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_6 <-df_study1_wide$comunacod_1
 
@@
               variable.name = c("ola"),
-              measure = patterns("^comuna_","^comunacod_","^cuestion_mig_",
-                                 "^in_simpat_","^in_simili_","^in_ansied_","^in_valfam_",
-                                 "^in_valami_","^in_amsimb_","^in_amreal_","^in_acpier_",
-                                 "^in_acadop_","^in_amigcl_",
-                                 "^conviviality","^identity",
-                                 "^educ_","^educyear_","^quintil1_","^ess_","^nation_t1_",
-                                 "^know_inm_t1t4_","^frien_inm_t1t4_","^pos_id_",
+              measure = patterns("^comuna_","^comunacod_",
+                                 "^just_educ_","^merit_effort_","^merit_talent_",
+                                 "^ahead_family_","^ahead_educ_","^ahead_ambition_","^ahead_hardwork_",
+                                 "^educ_","^educyear_","^quintil1_","^ess_",
+                                 "^pos_id_",
                                  "^edad_","^sexo_"
                                  ),
-              value.name = c("comuna","comunacod","cuestion_mig",
-                             "in_simpat","in_simili","in_ansied","in_valfam",
-                             "in_valami","in_amsimb","in_amreal",
-                             "in_acpier","in_acadop","in_amigcl",
-                             "conviviality","identity",
-                             "educ","educyear","quintil1","ess","nation_t1",
-                             "know_inm_t1t4","frien_inm_t1t4","pos_id",
+              value.name = c("comuna","comunacod",
+                             "just_educ","merit_effort","merit_talent",
+                             "ahead_family","ahead_educ","ahead_ambition","ahead_hardwork",
+                             "educ","educyear","quintil1","ess",
+                             "pos_id",
                              "edad","sexo"),
@@
 df_study1_long %>% 
-  filter(ola %in% c(1,2,3,4,5)) %>% 
-  mutate(ola=factor(ola,levels = 1:5,labels = 1:5))
+  filter(ola %in% c(1,2,3,4,6)) %>% 
+  mutate(ola=factor(ola,levels = 1:6,labels = 1:6))
 dim(df_study1_long) #check, now is OK
@@
 #______________________________________________________________________________
-# obtain the idencuesta for wave 5
+# obtain the idencuesta for wave 6
 ids <- 
@@
   select(idencuesta,ola) %>% 
-  filter(ola==5) %>% 
+  filter(ola==6) %>% 
   sjmisc::frq(idencuesta,show.na = F) %>% as.data.frame()
@@
 # filter data by the idencuesta of t5
-df_study1_long_t5 <- 
+df_study1_long_t6 <- 
   df_study1_long %>%
@@
 
-names(df_study1_long_t5)
-dim(df_study1_long_t5)
-sjmisc::frq(df_study1_long_t5$ola)
+names(df_study1_long_t6)
+dim(df_study1_long_t6)
+sjmisc::frq(df_study1_long_t6$ola)
 
@@
 save(df_study1_long,file = here::here("input/data-proc/df_study1_long.RData"))
-save(df_study1_long_t5,file = here::here("input/data-proc/df_study1_long_t5.RData"))
+save(df_study1_long_t6,file = here::here("input/data-proc/df_study1_long_t6.RData"))
 
@@
 
-sjPlot::view_df(df_study1_long_t5,
+sjPlot::view_df(df_study1_long_t6,
         show.frq = T,
@@
         show.na = T,
-        file = here::here("output/df_study1_long_t5.html"))
+        file = here::here("output/df_study1_long_t6.html"))
 ```
@@
 * Faltarían datos 2019 para algunas comunas
-
-* Crear variable de cambio en la tasa de migrantes a nivel comunal 
-  * Censo 2017 - CASEN 2020
 
@@
 df_elsoc <- 
-elsoc_long_2016_2021 %>% 
+elsoc_long_2016_2022.2 %>% 
   select(comuna,comuna_cod) %>%
@@
 cit17 %>% 
-  select(idencuesta,inmigrante:centroam)
+  select(idencuesta,escolaridad)
 
@@
 cit18 %>% 
-  select(idencuesta,inmigrante:centroam)
+  select(idencuesta,escolaridad)
 
@@
 cit19 %>% 
-  select(idencuesta,inmigrante:centroam)
+  select(idencuesta,escolaridad)
 df_19[df_19==-995] <- NA
@@
 df_cit_a <- 
-elsoc_long_2016_2021 %>% 
+elsoc_long_2016_2022.2 %>% 
   select(comuna,comuna_cod,idencuesta) %>% 
@@
 df_cit_b <- 
-  df_cit_a %>% distinct() %>% arrange(comuna_cod, inmigrante)
+  df_cit_a %>% distinct() %>% arrange(comuna_cod, escolaridad)
 
@@
   group_by(comuna, comuna_cod) %>% 
-  summarise(inm_cit17=sum(inmigrante,na.rm = T))
+  summarise(esc_cit17=mean(escolaridad,na.rm = T))
 
-sjmisc::frq(df_cit$inm_cit17)
-
-# CASEN nationality ___________________________________________________________
-# 2017_________________________________________________________
-df_17 <- 
-casen2017 %>% 
-  select(comuna,starts_with("r1"), starts_with("exp"))
-sjmisc::frq(df_17$r1a)
-sjmisc::frq(df_17$r1a_cod)
-sjmisc::frq(df_17$r1b)
-sjmisc::frq(df_17$r1b_p_cod)
-
-# Latam & caribean 401:588
-# recent migrant groups (2020) 
-  # Venezuela: 513
-  # Peru: 509
-  # Haiti: 413
-  # Colombia: 505
-  # Bolivia: 502
-
-# Al igual que la tendencia reflejada en las estimaciones de 2018 y 2019, la mayoría (poco más de 79% del total) de las personas extranjeras residentes en el país al 31 de diciembre de 2020 proviene de Venezuela (30,7%), Perú (16,3%), Haití (12,5%), Colombia (11,4%) y Bolivia (8,5%).
-
-#dummy for inmigrant 
-df_17$pais <- 
-car::recode(df_17$r1b_p_cod,
-            recodes = "c(513,509,413,505,502)=1;c(888,999)=NA;else=0",
-            as.factor = T)
-sjmisc::frq(df_17$pais)
-
-# df_17$r1b[df_17$r1b==9] <- NA
-# sjmisc::frq(df_17$r1b)
-
-df_17$inm <- ifelse(test = (df_17$pais==1),yes = 1,no = 0)
-
-sjmisc::frq(df_17$inm)
-
-# calculate the mean of the dummy by comuna
-df_17_comuna <- 
-df_17 %>% 
-  mutate(comuna_n_17 =sjlabelled::as_character(comuna)) %>% 
-  group_by(comuna_n_17, "comuna_cod"=comuna) %>% 
-  summarise(
-    inm_2017_cas = mean(inm,na.rm=T)*100, #non weigthed
-    inm_2017_cas_w = questionr::wtd.mean(inm,weights = expc)*100 #weigthed
-    ) %>% 
-  arrange(comuna_cod)
-
-summary(df_17_comuna$inm_2017_cas)  #check
-summary(df_17_comuna$inm_2017_cas_w)#check
-# 2020_________________________________________________________
-df_20 <- 
-casen20 %>% 
-  select(comuna,
-         starts_with("r1"),starts_with("exp"))
-
-sjmisc::frq(df_20$r1b_p_cod)
-#dummy for inmigrant 
-df_20$pais <- 
-car::recode(df_20$r1b_p_cod,
-            recodes = "c(513,509,413,505,502)=1;c(888,999)=NA;else=0",
-            as.factor = T)
-sjmisc::frq(df_20$pais)
-
-# df_20$r1b[df_20$r1b==9] <- NA
-# sjmisc::frq(df_20$r1b)
-df_20$inm <- ifelse(test = (df_20$pais==1),yes = 1,no = 0)
-# sjmisc::frq(df_20$inm)
-
-# calculate the mean of the dummy by comuna
-df_20_comuna <- 
-df_20 %>% 
-  mutate(comuna_n_20 =sjlabelled::as_character(comuna)) %>% 
-  group_by(comuna_n_20, "comuna_cod"=comuna) %>% 
-  summarise(
-    inm_2020_cas = mean(inm,na.rm=T)*100, #non weigthed
-    inm_2020_cas_w = questionr::wtd.mean(inm,weights = expc)*100 #weigthed
-    ) %>% 
-  arrange(comuna_cod)
-
-summary(df_20_comuna$inm_2020_cas)  #check
-summary(df_20_comuna$inm_2020_cas_w)#check
-
-df_b <- full_join(df_a, df_17_comuna, by = "comuna_cod")
-df_c <- full_join(df_b, df_20_comuna, by = "comuna_cod")
-
-# CENSUS Data_________________________________________________________________
-df_censo <- 
-censo2017 %>% 
-  select(P12, COMUNA)
-# sjmisc::frq(df_censo$P12)
-# (0-997)
-# 998 No aplica
-# 999 Missing
-# create dummy for inmigrant
-df_censo$inm_censo17 <- car::recode(df_censo$P12,recodes = "1:2=0;3:8=1;98:99=NA")
-# sjmisc::frq(df_censo$inm_censo17)
-
-# create % of inmigrant by comuna__________________________
-
-df_inm_comunas <- 
-df_censo %>% 
-  group_by(comuna_cod=COMUNA,) %>% 
-  summarise(inm_propcenso17 = mean(inm_censo17, na.rm=T)*100,
-            inm_totcenso17=sum(inm_censo17, na.rm=T))
-
-#Merge with CENSUS data (2017)
-df_d <- left_join(df_c, df_inm_comunas,  by = "comuna_cod")
-
-#Merge with Population proyection data (2017 - 2020)
-df_e <- left_join(df_d,df_proye_comunas, by = "comuna_cod")
-
-#Merge CIT data for 2019 (Census)
-df_f <- left_join(df_e,df_cit, by = "comuna_cod")
-
-#Calculate proportion of inmigrant based on proyections and CIT data
-df_f <- 
-df_f %>% 
-  mutate(inm_dem_proye18=(n_2018/pob_18)*100,
-         inm_dem_proye19=(n_2019/pob_19)*100,
-         inm_dem_proye20=(n_2020/pob_20)*100,
-         #CIT data / total pop by comune 2017 CENSUS
-         inm_cit_proye17=(inm_cit17/inm_totcenso17)*100)
+sjmisc::frq(df_cit$esc_cit17)
 
@@
 df_study1_comunas <- 
-df_f %>%
+df_cit %>%
   mutate(comuna=sjlabelled::as_character(comuna_cod)) %>% 
@@
                 comunacod=comuna_cod,
-                inm_prop_cas17_w=inm_2017_cas_w,
-                inm_tota_cen17=inm_totcenso17,
-                inm_prop_cen17=inm_propcenso17,
-                inm_tota_cit17=inm_cit17,
-                inm_prop_citproy17=inm_cit_proye17,
-                inm_prop_demproy18=inm_dem_proye18,
-                inm_prop_demproy19=inm_dem_proye19,
-                inm_prop_demproy20=inm_dem_proye20,
-                inm_prop_cas20_w=inm_2020_cas_w) %>% 
-  mutate(dif_inm = inm_prop_cas20_w-inm_prop_cas17_w)
-df_study1_comunas$comuna_lc <- NULL
+                esc_cit17
+                )
 # names(df_study1_comunas)
@@
   "Codigo Comuna",
-  "Proportion of migrant population (weighted) - CASEN 2017 - Municipality",
-  "Number of migrant population - CENSUS 2017 - Municipality",
-  "Proportion of migrant population - CENSUS 2017 - Municipality",
-  "Number of migrant population - CIT 2017 - Municipality (using census tract)",
-  "Proportion of migrant population (CIT / DEM)- CIT 2017 and INE/DEM 2017 - Communal",
-  "Proportion of projected migrant population - INE/DEM 2018 - Communal",
-  "Proportion of projected migrant population -  INE/DEM 2019 - Comunal",
-  "Proportion of projected migrant population -  INE/DEM 2020 - Comunal",
-  "Proportion of migrant population (weighted) - CASEN 2020 - Municipality",
-  "Change in the proportion of the migrant population between 2017 and 2020 - CASEN")
+  "Escolarity average - CIT 2017 - Municipality (using census tract)"
+)
 
@@
  dplyr::select(idencuesta:edad,sexo,
-               inm_prop_cas17_w,
-               inm_prop_cas20_w,
-               dif_inm) %>% 
+               esc_cit17) %>% 
  dplyr::rename(comuna=comuna.x)

### 2023-11-14 production/prod_prep-gini.R
data-casen
datos casen por comuna 2015, 2017, 2022
incluye:
- Gini comunal,
- Diferencia Gini 2022-2015, 2022-2017, 2017-2015

@@ -1,0 +1,143 @@
+if (!require("pacman"))
+  install.packages("pacman")
+pacman::p_load(dplyr, knitr, kableExtra, haven, acid, DescTools)
+
+cut <- read.csv(file = here::here(("input/data-original/cut.csv")))
+
+# Casen 2015 --------------------------------------------------------------
+# casen2015 <- read_dta(file = "input/data/casen2015.dta")
+# save(casen2015,file = "input/data/casen2015.RData")
+load(file = here::here("input/data-original/casen2015.RData"))
+
+df15 <-
+  casen2015 %>%
+  select(
+    comuna,
+    starts_with("ytrabajo"),
+    starts_with("ytot"),
+    starts_with("yaut"),
+    starts_with("ypc"),
+    expr,
+    expc
+  )
+df15 <- df15 %>% left_join(cut[, c("COMUNA", "COMUNA_15R")], by = c("comuna" =
+                                                                      "COMUNA_15R"))
+
+gini15 <-
+  df15 %>%
+  group_by(COMUNA) %>%
+  summarise("gini_ypch" = as.numeric(acid::weighted.gini(x = ypch, w =
+                                                           expc)$bcwGini)) %>%
+  mutate(year = 2015)
+
+# Casen 2017 --------------------------------------------------------------
+# casen2017 <- read_dta(file = "input/data/casen2017.dta")
+# save(casen2017,file = "input/data/casen2017.RData")
+load(file = here::here(("input/data-original/casen2017.RData")))
+
+
+df17 <-
+  casen2017 %>%
+  select(
+    folio:zona,
+    starts_with("ytrabajo"),
+    starts_with("ytot"),
+    starts_with("yaut"),
+    starts_with("ypc"),
+    expr,
+    expc,
+    COMUNA = comuna
+  )
+
+
+gini17 <-
+  df17 %>%
+  group_by(COMUNA) %>%
+  summarise("gini_ypch" = as.numeric(acid::weighted.gini(x = ypch, w =
+                                                           expc)$bcwGini)) %>%
+  mutate(year = 2017)
+names(gini17)
+
+# Casen 2022 --------------------------------------------------------------
[…]

### 2023-11-15 production/prod_analysis.Rmd
update analisis
* Incluye var dep justicia pensiones, salud y percepción desigualdad
* incluye gini comunal como indep
* descriptivos ingresos long

@@
 
-```{r anxiety}
+```{r distribution-educ}
 #rm(list=ls())
@@
                )
-#load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
 
@@
   df_study1_long_comunas %>%
-  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:esc_cit17) %>% 
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022) %>% 
   na.omit() %>% 
@@
 h3 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork"
-h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+esc_cit17"
-h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+esc_cit17+educ+quintil1+ess+pos_id+edad+sexo"
+h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+ginilag22_17+gini_ypch_2022"
+h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+ginilag22_17+gini_ypch_2022+educ+quintil1+ess+pos_id+edad+sexo"
 
-# A. Anxiety (in_ansied)
+# A. Education distributive justice
 educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
@@
                                       "Goahead: trabajo duro",
-                                      "Promedio escolaridad comuna",
+                                      "Cambio gini 2022-2017",
+                                      "Gini comunal 2022",
                                       "Ed. media <br> <i>(Ref. Ed. básica)</i>",
@@
 ```
+
+## B. Pension distribution justice
+
+```{r distribution-pension}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_pension:merit_talent, ahead_family:sexo, ginilag22_17,gini_ypch_2022) %>% 
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+
+# B. Pension distributive justice
+pension.null <- lmer(formula(paste0("just_pension~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+pension1 <- lmer(formula(paste0("just_pension~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension2 <- lmer(formula(paste0("just_pension~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension3 <- lmer(formula(paste0("just_pension~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension4 <- lmer(formula(paste0("just_pension~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension5 <- lmer(formula(paste0("just_pension~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(pension.null, pension1, pension2, pension3, pension4, pension5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: Familia",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: trabajo duro",
+                                      "Cambio gini 2022-2017",
+                                      "Gini comunal 2022",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"
+                              )
+          #omit.coef = omit
+          )
+```
+
+## C. Health distribution justice
+
+```{r distribution-health}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_salud:merit_talent, ahead_family:sexo, ginilag22_17,gini_ypch_2022) %>% 
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+
+# C. Health distributive justice
+salud.null <- lmer(formula(paste0("just_salud~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+salud1 <- lmer(formula(paste0("just_salud~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud2 <- lmer(formula(paste0("just_salud~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud3 <- lmer(formula(paste0("just_salud~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud4 <- lmer(formula(paste0("just_salud~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud5 <- lmer(formula(paste0("just_salud~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(salud.null, salud1, salud2, salud3, salud4, salud5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: Familia",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: trabajo duro",
+                                      "Cambio gini 2022-2017",
+                                      "Gini comunal 2022",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"
+                              )
+          #omit.coef = omit
+          )
+```
+
+## D. Income differences are too large
+
+```{r inequality}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study2_long_comunas.RData"))
+
+# generate analytical sample
+df_study2 <- 
+  df_study2_long_comunas %>%
+  select(idencuesta,ola,comuna,des_perc:merit_talent, ahead_family:sexo, ginilag22_17,gini_ypch_2022) %>% 
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+
+# C. Health distributive justice
+des.null <- lmer(formula(paste0("des_perc~","1 + (1|idencuesta)+(1|comuna)")),data = df_study2)
+des1 <- lmer(formula(paste0("des_perc~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des2 <- lmer(formula(paste0("des_perc~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des3 <- lmer(formula(paste0("des_perc~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des4 <- lmer(formula(paste0("des_perc~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des5 <- lmer(formula(paste0("des_perc~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(des.null, des1, des2, des3, des4, des5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "Ola 2021",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: Familia",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: trabajo duro",
+                                      "Cambio gini 2022-2017",
+                                      "Gini comunal 2022",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"
+                              )
+          #omit.coef = omit
+          )
+```

### 2023-11-15 production/prod_prep-ingresos.Rmd
update analisis
* Incluye var dep justicia pensiones, salud y percepción desigualdad
* incluye gini comunal como indep
* descriptivos ingresos long

---
title: "prod_prep_ingresos"
date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
      number_sections: yes
    code_folding: hide
    number_sections: yes
editor_options:
  chunk_output_type: console
---

# Setup

```{r setup}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
options(scipen=9999) # desactivar notacion cientifica
remove(list = ls()) #limpieza del entorno de trabajo
```

```{css, echo=FALSE}
# /* expandir el espacio del documento*/
body .main-container {
      max-width: 1600px !important;
      width: 1600px !important;
    }
#/* expandir width de la TOC*/
div.tocify {
    width: 22% !important;
    max-width: 331px !important;
    max-height: 102% !important;
}
```

# Download data
```{r download data, eval=FALSE, include=FALSE}
#download and save data

file.exists(here::here("input/data-original/elsoc_wide.Rdata"))

load(url("https://dataverse.harvard.edu/api/access/datafile/7245115")) #ELSOC
save(elsoc_wide_2016_2022.2,file = here::here("input/data-original/elsoc_wide.Rdata"))
```

## Individual data

```{r}
remove(list = ls()) # clean workspace
# .rs.restartR() #restart R session
if (!require("pacman")) install.packages("pacman") # install pacman
# load libraries
pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here, summarytools)
load(here::here("input/data-original/elsoc_wide.Rdata"))

elsoc_wide_2016_2022.2[elsoc_wide_2016_2022.2 ==-999] <- NA
elsoc_wide_2016_2022.2[elsoc_wide_2016_2022.2 ==-888] <- NA
elsoc_wide_2016_2022.2[elsoc_wide_2016_2022.2 ==-777] <- NA
elsoc_wide_2016_2022.2[elsoc_wide_2016_2022.2 ==-666] <- NA

# ingresos__________________________________________________
ingresos <- 
elsoc_wide_2016_2022.2 %>% 
  #create new variables
  select(m30_w01,
         m30_w02,
         m30_w03,
         m30_w04,
         m30_w06,
         m30b_w05,
         m30b_w06,
         m29_w01,
         m29_w02,
         m29_w03,
         m29_w04,
         m29_w05,
         m29_w06,
         ) %>% 
  sjlabelled::drop_labels(., drop.na = FALSE)
```

```{r echo=FALSE}
print(dfSummary(ingresos, headings = FALSE, graph.col = FALSE), method = "render")
```

```{r}
# Household income_________________________________________

#Impute midpoint of income ranges
elsoc_wide_2016_2022.2$m30_rec_w01 <-
as.numeric(car::recode(elsoc_wide_2016_2022.2$m30_w01,
           "1=110000;2=251000;3=305000;4=355000;5=400000;
            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m30_rec_w02 <-
as.numeric(car::recode(elsoc_wide_2016_2022.2$m30_w02,
           "1=110000;2=251000;3=305000;4=355000;5=400000;
            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m30_rec_w03 <-
as.numeric(car::recode(elsoc_wide_2016_2022.2$m30_w03,
           "1=110000;2=251000;3=305000;4=355000;5=400000;
            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m30_rec_w04 <-
as.numeric(car::recode(elsoc_wide_2016_2022.2$m30_w04,
           "1=110000;2=251000;3=305000;4=355000;5=400000;
            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m30_rec_w06 <-
as.numeric(car::recode(elsoc_wide_2016_2022.2$m30_w06,
           "1=110000;2=251000;3=305000;4=355000;5=400000;
            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))

#Impute midpoint of income ranges (2021)
elsoc_wide_2016_2022.2$m30b_rec_w05 <-
as.numeric(car::recode(elsoc_wide_2016_2022.2$m30b_w05,
           "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m30b_rec_w06 <-
as.numeric(car::recode(elsoc_wide_2016_2022.2$m30b_w06,
           "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))

sjmisc::frq(elsoc_wide_2016_2022.2$m30b_rec_w05)

#Recode DK/DA of Income to NA
elsoc_wide_2016_2022.2$m29_rec_w01 <-
  as.numeric(car::recode(elsoc_wide_2016_2022.2$m29_w01,"c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m29_rec_w02 <-
  as.numeric(car::recode(elsoc_wide_2016_2022.2$m29_w02,"c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m29_rec_w03 <-
  as.numeric(car::recode(elsoc_wide_2016_2022.2$m29_w03,"c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m29_rec_w04 <-
  as.numeric(car::recode(elsoc_wide_2016_2022.2$m29_w04,"c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m29_rec_w05 <-
  as.numeric(car::recode(elsoc_wide_2016_2022.2$m29_w05,"c(-888,-999)=NA"))

elsoc_wide_2016_2022.2$m29_rec_w06 <-
  as.numeric(car::recode(elsoc_wide_2016_2022.2$m29_w06,"c(-888,-999)=NA"))

#replace NA of income with new imputed variable
elsoc_wide_2016_2022.2$m29_imp_w01 <- 
  ifelse(test = !is.na(elsoc_wide_2016_2022.2$m29_rec_w01),
         yes =  elsoc_wide_2016_2022.2$m29_rec_w01,
         no =  elsoc_wide_2016_2022.2$m30_rec_w01)
summary(elsoc_wide_2016_2022.2$m29_imp_w01)

elsoc_wide_2016_2022.2$m29_imp_w02 <- 
  ifelse(test = !is.na(elsoc_wide_2016_2022.2$m29_rec_w02),
         yes =  elsoc_wide_2016_2022.2$m29_rec_w02,
         no =  elsoc_wide_2016_2022.2$m30_rec_w02)
summary(elsoc_wide_2016_2022.2$m29_imp_w02)

elsoc_wide_2016_2022.2$m29_imp_w03 <- 
  ifelse(test = !is.na(elsoc_wide_2016_2022.2$m29_rec_w03),
         yes =  elsoc_wide_2016_2022.2$m29_rec_w03,
         no =  elsoc_wide_2016_2022.2$m30_rec_w03)
summary(elsoc_wide_2016_2022.2$m29_imp_w03)

elsoc_wide_2016_2022.2$m29_imp_w04 <- 
  ifelse(test = !is.na(elsoc_wide_2016_2022.2$m29_rec_w04),
         yes =  elsoc_wide_2016_2022.2$m29_rec_w04,
         no =  elsoc_wide_2016_2022.2$m30_rec_w04)
summary(elsoc_wide_2016_2022.2$m29_imp_w04)

elsoc_wide_2016_2022.2$m29_imp_w05 <- 
  ifelse(test = !is.na(elsoc_wide_2016_2022.2$m29_rec_w05),
         yes =  elsoc_wide_2016_2022.2$m29_rec_w05,
         no =  elsoc_wide_2016_2022.2$m30b_rec_w05)
summary(elsoc_wide_2016_2022.2$m29_imp_w05)

elsoc_wide_2016_2022.2$m29_imp_w06 <- 
  ifelse(test = !is.na(elsoc_wide_2016_2022.2$m29_rec_w06),
         yes =  elsoc_wide_2016_2022.2$m29_rec_w06,
         no =  elsoc_wide_2016_2022.2$m30_rec_w06)
summary(elsoc_wide_2016_2022.2$m29_imp_w06)

###########

elsoc_wide_2016_2022.2$m29_imp_w05 <- 
  ifelse(test = is.na(elsoc_wide_2016_2022.2$m29_imp_w05),
         yes =  elsoc_wide_2016_2022.2$m30b_rec_w05,
         no =  elsoc_wide_2016_2022.2$m29_imp_w05)
summary(elsoc_wide_2016_2022.2$m29_imp_w05)

elsoc_wide_2016_2022.2$m29_imp_w06 <- 
  ifelse(test = is.na(elsoc_wide_2016_2022.2$m29_imp_w06),
         yes =  elsoc_wide_2016_2022.2$m30b_rec_w06,
         no =  elsoc_wide_2016_2022.2$m29_imp_w06)
summary(elsoc_wide_2016_2022.2$m29_imp_w06)

# deflate at each year's prices
#elsoc_wide_2016_2022.2$deflactor <-
#  with(elsoc_wide_2016_2022.2, case_when(
#    ola == 2016 ~ 113.88 / 123.82,
#    ola == 2017 ~ 116.46 / 123.82,
#    ola == 2018 ~ 119.45 / 123.82,
#    ola == 2019 ~ 123.82 / 123.82
#  ))

# Per capita household income:
elsoc_wide_2016_2022.2$ing_pc_w01 <- 
  (elsoc_wide_2016_2022.2$m29_imp_w01/elsoc_wide_2016_2022.2$nhogar1_w01)

elsoc_wide_2016_2022.2$ing_pc_w02 <- 
  (elsoc_wide_2016_2022.2$m29_imp_w02/elsoc_wide_2016_2022.2$m46_nhogar_w02)

elsoc_wide_2016_2022.2$ing_pc_w03 <- 
  (elsoc_wide_2016_2022.2$m29_imp_w03/elsoc_wide_2016_2022.2$m54_w03)

elsoc_wide_2016_2022.2$ing_pc_w04 <- 
  (elsoc_wide_2016_2022.2$m29_imp_w04/elsoc_wide_2016_2022.2$m54_w04)

elsoc_wide_2016_2022.2$ing_pc_w05 <- 
  (elsoc_wide_2016_2022.2$m29_imp_w05/elsoc_wide_2016_2022.2$m54_w05)

elsoc_wide_2016_2022.2$ing_pc_w06 <- 
  (elsoc_wide_2016_2022.2$m29_imp_w06/elsoc_wide_2016_2022.2$m54_w05)

elsoc_wide_2016_2022.2$ing_pc_w01 <-
sjlabelled::set_label(x = elsoc_wide_2016_2022.2$ing_pc_w01,
                      label = "Household income per capita 2016")

sjmisc::descr(elsoc_wide_2016_2022.2$ing_pc_w01)

# Compute income quintiles
elsoc_wide_2016_2022.2 <- elsoc_wide_2016_2022.2 %>% 
  mutate(quintil_w01 = ntile(-desc(ing_pc_w01), 5))

elsoc_wide_2016_2022.2 <- elsoc_wide_2016_2022.2 %>% 
  mutate(quintil_w02 = ntile(-desc(ing_pc_w02), 5))

elsoc_wide_2016_2022.2 <- elsoc_wide_2016_2022.2 %>% 
  mutate(quintil_w03 = ntile(-desc(ing_pc_w03), 5))

elsoc_wide_2016_2022.2 <- elsoc_wide_2016_2022.2 %>% 
  mutate(quintil_w04 = ntile(-desc(ing_pc_w04), 5))

elsoc_wide_2016_2022.2 <- elsoc_wide_2016_2022.2 %>% 
  mutate(quintil_w05 = ntile(-desc(ing_pc_w05), 5))

elsoc_wide_2016_2022.2 <- elsoc_wide_2016_2022.2 %>% 
  mutate(quintil_w06 = ntile(-desc(ing_pc_w06), 5))

elsoc_wide_2016_2022.2$quintil_w01 <- 
  factor(elsoc_wide_2016_2022.2$quintil_w01,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

elsoc_wide_2016_2022.2$quintil_w02 <- 
  factor(elsoc_wide_2016_2022.2$quintil_w02,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

elsoc_wide_2016_2022.2$quintil_w03 <- 
  factor(elsoc_wide_2016_2022.2$quintil_w03,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

elsoc_wide_2016_2022.2$quintil_w04 <- 
  factor(elsoc_wide_2016_2022.2$quintil_w04,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

elsoc_wide_2016_2022.2$quintil_w05 <- 
  factor(elsoc_wide_2016_2022.2$quintil_w05,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

elsoc_wide_2016_2022.2$quintil_w06 <- 
  factor(elsoc_wide_2016_2022.2$quintil_w06,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

#reverse quintile, reference level is the highest quintile
#elsoc_wide_2016_2022.2$quintil <- forcats::fct_rev(elsoc_wide_2016_2022.2$quintil)

elsoc_wide_2016_2022.2$quintil_w01 <- 
sjlabelled::set_label(x = elsoc_wide_2016_2022.2$quintil_w01,
                      label = "Household income quintile per capita")

sjmisc::frq(elsoc_wide_2016_2022.2$quintil_w01)

#include new quintile category with missing cases
elsoc_wide_2016_2022.2$quintil1_w01<-
  car::recode(elsoc_wide_2016_2022.2$quintil_w01, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

#include new quintile category with missing cases
elsoc_wide_2016_2022.2$quintil1_w02<-
  car::recode(elsoc_wide_2016_2022.2$quintil_w02, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

#include new quintile category with missing cases
elsoc_wide_2016_2022.2$quintil1_w03<-
  car::recode(elsoc_wide_2016_2022.2$quintil_w03, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

#include new quintile category with missing cases
elsoc_wide_2016_2022.2$quintil1_w04<-
  car::recode(elsoc_wide_2016_2022.2$quintil_w04, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

#include new quintile category with missing cases
elsoc_wide_2016_2022.2$quintil1_w05<-
  car::recode(elsoc_wide_2016_2022.2$quintil_w05, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

#include new quintile category with missing cases
elsoc_wide_2016_2022.2$quintil1_w06<-
  car::recode(elsoc_wide_2016_2022.2$quintil_w06, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

#elsoc_wide_2016_2022.2$quintil1 <- factor(elsoc_wide_2016_2022.2$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))

elsoc_wide_2016_2022.2$quintil1_w01 <- 
sjlabelled::set_label(x = elsoc_wide_2016_2022.2$quintil1_w01,
                      label = "Household income quintile per capita (NA)") 
sjmisc::frq(elsoc_wide_2016_2022.2$quintil1_w01)

quintiles <- elsoc_wide_2016_2022.2 %>% select(quintil_w01, quintil_w02, quintil_w03, quintil_w04, quintil_w05, quintil_w06)
quintiles$na_count <- rowSums(is.na(quintiles))
round(prop.table(table(quintiles$na_count))*100,2)
```

```{r echo=FALSE}
print(dfSummary(quintiles, headings = FALSE, graph.col = FALSE), method = "render")
```

### 2023-11-15 production/prod_prep.Rmd
update analisis
* Incluye var dep justicia pensiones, salud y percepción desigualdad
* incluye gini comunal como indep
* descriptivos ingresos long

@@
   mutate(just_educ=d02_02,
+         just_pension=d02_01,
+         just_salud=d02_03,
+         des_perc=c18_11,
          sexo=m0_sexo,
@@
                       label = "Education distributive justice")
+
+#
+elsoc_long_2016_2022.2$just_pension <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$just_pension, 
+                      label = "Pension distributive justice")
+
+#
+elsoc_long_2016_2022.2$just_salud <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$just_salud, 
+                      label = "Health distributive justice")
+
+#
+elsoc_long_2016_2022.2$des_perc <- 
+sjlabelled::set_label(x = elsoc_long_2016_2022.2$des_perc, 
+                      label = "Income differences are too large")
 
@@
 
-for (i in c("just_educ","merit_effort","merit_talent",
+for (i in c("just_educ", "just_pension", "just_salud", "des_perc",
+            "merit_effort","merit_talent",
             "ahead_family","ahead_educ","ahead_ambition", "ahead_hardwork")) {
@@
          just_educ,
+         just_pension,
+         just_salud,
+         des_perc,
          merit_effort,
@@
               measure = patterns("^comuna_","^comunacod_",
-                                 "^just_educ_","^merit_effort_","^merit_talent_",
-                                 "^ahead_family_","^ahead_educ_","^ahead_ambition_","^ahead_hardwork_",
+                                 "^just_educ_", "^just_pension_", "^just_salud_", "^des_perc_",
+                                 "^merit_effort_","^merit_talent_",
+                                 "^ahead_family_", "^ahead_educ_", "^ahead_ambition_", "^ahead_hardwork_",
                                  "^educ_","^educyear_","^quintil1_","^ess_",
@@
               value.name = c("comuna","comunacod",
-                             "just_educ","merit_effort","merit_talent",
+                             "just_educ", "just_pension","just_salud", "des_perc",
+                             "merit_effort","merit_talent",
                              "ahead_family","ahead_educ","ahead_ambition","ahead_hardwork",
@@
 dim(df_study1_long) #check dimensions of the dataframe
-# filter the sataset for the waves 1 to 5
+# Original dataset with 6 waves
+df_study2_long <- df_study1_long
+
+# filter the dataset for the waves 1 to 4 and 6
 df_study1_long <-
@@
 save(df_study1_long_t6,file = here::here("input/data-proc/df_study1_long_t6.RData"))
-
+save(df_study2_long,file = here::here("input/data-proc/df_study2_long.RData"))
 # save codebook
@@
 load(here::here("input/data-original/cit19.Rdata"))
-load(here::here("input/data-original/casen17.Rdata"))
-load(here::here("input/data-original/casen20.Rdata"))
+load(here::here("input/data-original/casen2017.Rdata"))
+load(here::here("input/data-original/casen2022.Rdata"))
 load(here::here("input/data-original/censo17.Rdata"))
@@
 load(file = here::here("input/data-proc/df_study1_long.RData"))
+load(file = here::here("input/data-proc/df_study2_long.RData"))
 load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
 load(file = here::here("input/data-proc/study_1_comunas.RData"))
+load(file = here::here("input/data-proc/df_gini_comunas_wide.RData"))
+load(file = here::here("input/data-proc/df_gini_comunas_long.RData"))
+df_gini_comunas_wide <-  df_gini_comunas_wide %>% dplyr::rename(comunacod=COMUNA_COD) 
+df_gini_comunas_wide$comunacod <- as.numeric(df_gini_comunas_wide$comunacod)
 
@@
  dplyr::rename(comuna=comuna.x)
+
+df_study1_long_comunas <- 
+ dplyr::left_join(x = df_study1_long_comunas,
+                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini")), 
+                  by = c("comunacod"))
+
+df_study2_long_comunas <- 
+ dplyr::left_join(x = df_study2_long,
+                  y = df_study1_comunas, 
+                  by = "comunacod") %>% 
+ dplyr::select(idencuesta:edad,sexo,
+               esc_cit17) %>% 
+ dplyr::rename(comuna=comuna.x)
+
+df_study2_long_comunas <- 
+ dplyr::left_join(x = df_study2_long_comunas,
+                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini")), 
+                  by = c("comunacod"))
 
@@
      file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+save(df_study2_long_comunas,
+     file = here::here("input/data-proc/df_study2_long_comunas.RData"))
 save(df_study_long_t5_comunas,

### 2023-12-07 production/prod_prep-mergedata.Rmd
merge new data

---
title: "prod_prep_mergedata"
date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
      number_sections: yes
    code_folding: hide
    number_sections: yes
editor_options:
  chunk_output_type: console
---

# Setup

```{r setup}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
options(scipen=9999) # desactivar notacion cientifica
remove(list = ls()) #limpieza del entorno de trabajo
```

```{css, echo=FALSE}
# /* expandir el espacio del documento*/
body .main-container {
      max-width: 1600px !important;
      width: 1600px !important;
    }
#/* expandir width de la TOC*/
div.tocify {
    width: 22% !important;
    max-width: 331px !important;
    max-height: 102% !important;
}
```

# Merge individual & contextual

```{r}
# .rs.restartR() #restart R session
if (!require("pacman")) install.packages("pacman") # install pacman
# load libraries
pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here)

load(file = here::here("input/data-proc/df_study1_long.RData"))
load(file = here::here("input/data-proc/df_study2_long.RData"))
load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
load(file = here::here("input/data-proc/study_1_comunas.RData"))
load(file = here::here("input/data-proc/df_casen_comunas_wide.RData")) # contienen las variables de Gini comunal y Proporcion de Universitarios comunal.
load(file = here::here("input/data-proc/df_casen_comunas_long.RData")) # contienen las variables de Gini comunal y Proporcion de Universitarios comunal.
load(file = here::here("input/data-proc/df_matricula_comunas_wide.RData")) # contienen las variables de Proporción de Matrícula en establecimientos Particular Pagados
load(file = here::here("input/data-proc/df_matricula_comunas_long.RData")) # contienen las variables de Proporción de Matrícula en establecimientos Particular Pagados
load(file = here::here("input/data-proc/df_simce_comunas_wide.RData")) # contiene las variables de Simce a nivel comunal promedio y la varianza dentro de la comuna
load(file = here::here("input/data-proc/df_simce_comunas_long.RData")) # contiene las variables de Simce a nivel comunal promedio y la varianza dentro de la comuna

## Juntar Gini y prop universitarios por comuna

df_gini_comunas_wide <-  df_gini_comunas_wide %>% dplyr::rename(comunacod=COMUNA_COD) 
df_gini_comunas_wide$comunacod <- as.numeric(df_gini_comunas_wide$comunacod)

df_study1_long_comunas <- 
 dplyr::left_join(x = df_study1_long,
                  y = df_study1_comunas, 
                  by = "comunacod") %>% 
 dplyr::select(idencuesta:edad,sexo,
               esc_cit17) %>% 
 dplyr::rename(comuna=comuna.x)

df_study1_long_comunas <- 
 dplyr::left_join(x = df_study1_long_comunas,
                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini"), starts_with("prop_univ"), starts_with("univlag")), 
                  by = c("comunacod"))

df_study2_long_comunas <- 
 dplyr::left_join(x = df_study2_long,
                  y = df_study1_comunas, 
                  by = "comunacod") %>% 
 dplyr::select(idencuesta:edad,sexo,
               esc_cit17) %>% 
 dplyr::rename(comuna=comuna.x)

df_study2_long_comunas <- 
 dplyr::left_join(x = df_study2_long_comunas,
                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini"), starts_with("prop_univ"), starts_with("univlag")), 
                  by = c("comunacod"))

df_study_long_t5_comunas <- 
 dplyr::left_join(x = df_study1_long_t5,
                  y = df_study1_comunas, 
                  by = "comunacod") %>% 
 dplyr::select(idencuesta:edad,sexo,
               esc_cit17) %>% 
 dplyr::rename(comuna=comuna.x)

## Juntar matricula privada por comuna

df_study1_long_comunas <- 
 dplyr::left_join(x = df_study1_long_comunas,
                  y = select(df_matricula_comunas_wide, comunacod=cod_com_rbd_16,starts_with("total_mat_pagado"), starts_with("prop_pagado"), pagadolag22_16), 
                  by = c("comunacod"))

df_study2_long_comunas <- 
 dplyr::left_join(x = df_study2_long_comunas,
                  y = select(df_matricula_comunas_wide, comunacod=cod_com_rbd_16,starts_with("total_mat_pagado"), starts_with("prop_pagado"), pagadolag22_16),
                  by = c("comunacod"))

## Juntar promedio simce y sd por comuna

df_study1_long_comunas <- 
 dplyr::left_join(x = df_study1_long_comunas,
                  y = select(df_comuna_rbd_wide, comunacod=COMUNA,starts_with("prom_simce"), starts_with("var_simce")), 
                  by = c("comunacod"))

df_study2_long_comunas <- 
 dplyr::left_join(x = df_study2_long_comunas,
                  y = select(df_comuna_rbd_wide, comunacod=COMUNA,starts_with("gini"), starts_with("prop_univ"), starts_with("univlag")), 
                  by = c("comunacod"))

# save data____________________________________________________________________
save(df_study1_long_comunas,
     file = here::here("input/data-proc/df_study1_long_comunas.RData"))
save(df_study2_long_comunas,
     file = here::here("input/data-proc/df_study2_long_comunas.RData"))
save(df_study_long_t5_comunas,
     file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))

# save codebook
sjPlot::view_df(df_study1_long_comunas,
        show.frq = T, 
        show.prc = T, 
        show.na = T,
        file = here::here("output/df_study1_long_comunas.html"))

sjPlot::view_df(df_study_long_t5_comunas,
        show.frq = T, 
        show.prc = T, 
        show.na = T,
        file = here::here("output/df_study_long_t5_comunas.html"))
```

### 2023-12-07 production/prod_prep.Rmd
merge new data

@@
 ```
-
-# Merge individual & contextual
-
-```{r}
-remove(list = ls()) # clean workspace
-# .rs.restartR() #restart R session
-if (!require("pacman")) install.packages("pacman") # install pacman
-# load libraries
-pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here)
-
-load(file = here::here("input/data-proc/df_study1_long.RData"))
-load(file = here::here("input/data-proc/df_study2_long.RData"))
-load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
-load(file = here::here("input/data-proc/study_1_comunas.RData"))
-load(file = here::here("input/data-proc/df_gini_comunas_wide.RData"))
-load(file = here::here("input/data-proc/df_gini_comunas_long.RData"))
-df_gini_comunas_wide <-  df_gini_comunas_wide %>% dplyr::rename(comunacod=COMUNA_COD) 
-df_gini_comunas_wide$comunacod <- as.numeric(df_gini_comunas_wide$comunacod)
-
-df_study1_long_comunas <- 
- dplyr::left_join(x = df_study1_long,
-                  y = df_study1_comunas, 
-                  by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo,
-               esc_cit17) %>% 
- dplyr::rename(comuna=comuna.x)
-
-df_study1_long_comunas <- 
- dplyr::left_join(x = df_study1_long_comunas,
-                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini")), 
-                  by = c("comunacod"))
-
-df_study2_long_comunas <- 
- dplyr::left_join(x = df_study2_long,
-                  y = df_study1_comunas, 
-                  by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo,
-               esc_cit17) %>% 
- dplyr::rename(comuna=comuna.x)
-
-df_study2_long_comunas <- 
- dplyr::left_join(x = df_study2_long_comunas,
-                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini")), 
-                  by = c("comunacod"))
-
-df_study_long_t5_comunas <- 
- dplyr::left_join(x = df_study1_long_t5,
-                  y = df_study1_comunas, 
-                  by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo,
-               esc_cit17) %>% 
- dplyr::rename(comuna=comuna.x)
-
-# save data____________________________________________________________________
-save(df_study1_long_comunas,
-     file = here::here("input/data-proc/df_study1_long_comunas.RData"))
-save(df_study2_long_comunas,
-     file = here::here("input/data-proc/df_study2_long_comunas.RData"))
-save(df_study_long_t5_comunas,
-     file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# save codebook
-sjPlot::view_df(df_study1_long_comunas,
-        show.frq = T, 
-        show.prc = T, 
-        show.na = T,
-        file = here::here("output/df_study1_long_comunas.html"))
-
-sjPlot::view_df(df_study_long_t5_comunas,
-        show.frq = T, 
-        show.prc = T, 
-        show.na = T,
-        file = here::here("output/df_study_long_t5_comunas.html"))
-```

### 2023-12-07 production/prod_prep-matricula.R

@@ -168,5 +168,5 @@ df_matricula_comunas_long<-
   select(cod_com_rbd_15,cod_com_rbd_16,year,everything()) %>%
   arrange(cod_com_rbd_15)
 
-save(df_matricula_comunas_wide, file = here::here("input/data-proc/df_matricula_comunas_long.RData"))
+save(df_matricula_comunas_long, file = here::here("input/data-proc/df_matricula_comunas_long.RData"))
 

### 2023-12-11 production/prod_analysis.Rmd
update context variables

@@
 ```{r tab-desc, results="asis", warning=FALSE}
-rm(list=ls())
 if (!require("pacman")) install.packages("pacman") # instalar pacman
@@
   df_study1_long_comunas %>%
-  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022) %>% 
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
   na.omit() %>% 
@@
 h3 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork"
-h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+ginilag22_17+gini_ypch_2022"
-h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+ginilag22_17+gini_ypch_2022+educ+quintil1+ess+pos_id+edad+sexo"
+h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc"
+h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna"
+h6 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna+educ+quintil1+ess+pos_id+edad+sexo"
 
@@
 educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
 
 #omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::htmlreg(list(educ.null, educ1, educ2, educ3, educ4, educ5),
+texreg::htmlreg(list(educ.null, educ1, educ2, educ3, educ4, educ5, educ6),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
@@
                                       "Goahead: trabajo duro",
+                                      "Gini comunal 2022",
                                       "Cambio gini 2022-2017",
-                                      "Gini comunal 2022",
+                                      "Proporción matrícula privada",
+                                      "Total matrícula privada",
+                                      "Varianza Simce comuna",
+                                      "Promedio Simce comuna",
                                       "Ed. media <br> <i>(Ref. Ed. básica)</i>",
@@
 
+## Contextual analysis
+
+```{r distribution-educ-context}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1 <- "gini_ypch"
+h2 <- "gini_gc"
+h3 <- "prop_pagado"
+h4 <- "total_mat_pagado"
+h5 <- "var_simce_comuna"
+h6 <- "prom_simce_comuna"
+h7 <- "gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(educ.null, educ1, educ2, educ3, educ4, educ5, educ6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "Gini comunal 2022",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Total matrícula privada",
+                                      "Varianza Simce comuna",
+                                      "Promedio Simce comuna"
+                              )
+          #omit.coef = omit
+          )
+```
+
 ## B. Pension distribution justice
 
-```{r distribution-pension}
+```{r distribution-pension, eval=FALSE}
 #rm(list=ls())
@@
 
-```{r distribution-health}
+```{r distribution-health, eval=FALSE}
 #rm(list=ls())
@@
 
-```{r inequality}
+```{r inequality, eval=FALSE}
 #rm(list=ls())

### 2023-12-11 production/prod_prep-mergedata.Rmd
update context variables

@@
 
+df_study1_long_comunas <- df_study1_long_comunas %>% mutate(year= case_when(ola==1~2017,
+                                                                            ola==2~2017,
+                                                                            ola==3~2017,
+                                                                            ola==4~2022,
+                                                                            ola==6~2022))
+
+df_study1_long_comunas <- 
+ dplyr::left_join(x = df_study1_long_comunas,
+                  y = select(df_gini_comunas_long, comunacod=COMUNA_COD,year, starts_with("gini"),prop_univ), 
+                  by = c("comunacod", "year"))
+
 df_study1_long_comunas <- 
@@
 
+df_study1_long_comunas <- df_study1_long_comunas %>% mutate(year= case_when(ola==1~2016,
+                                                                            ola==2~2017,
+                                                                            ola==3~2018,
+                                                                            ola==4~2019,
+                                                                            ola==6~2022))
+
+df_study1_long_comunas <- 
+ dplyr::left_join(x = df_study1_long_comunas,
+                  y = select(df_matricula_comunas_long, comunacod=cod_com_rbd_15,year, total_mat_pagado,prop_pagado), 
+                  by = c("comunacod", "year"))
+
 df_study1_long_comunas <- 
@@
 ## Juntar promedio simce y sd por comuna
+
+df_study1_long_comunas <- 
+ dplyr::left_join(x = df_study1_long_comunas,
+                  y = select(df_simce_comunas_long, comunacod=cod_com_15,year, prom_simce_comuna,var_simce_comuna), 
+                  by = c("comunacod", "year"))
 

### 2023-12-11 production/prod_analysis.Rmd
update alternativas y readme
* update analisis alternativos
* update link readme

--- 2023-09-28 production/prod_analysis.Rmd
@@
   df_study1_long_comunas %>%
-  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:esc_cit17) %>% 
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
   na.omit() %>% 
@@
 h3 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork"
-h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+esc_cit17"
-h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+esc_cit17+educ+quintil1+ess+pos_id+edad+sexo"
+h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc"
+h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna"
+h6 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna+educ+quintil1+ess+pos_id+edad+sexo"
 
-# A. Anxiety (in_ansied)
+# A. Education distributive justice
 educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
@@
                                       "Goahead: trabajo duro",
-                                      "Promedio escolaridad comuna",
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Total matrícula privada",
+                                      "Varianza Simce comuna",
+                                      "Promedio Simce comuna",
                                       "Ed. media <br> <i>(Ref. Ed. básica)</i>",
@@
 ```
+
+## Contextual analysis
+
+```{r distribution-educ-context}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1b <- "gini_ypch"
+h2b <- "gini_gc"
+h3b <- "prop_pagado"
+h4b <- "total_mat_pagado"
+h5b <- "var_simce_comuna"
+h6b <- "prom_simce_comuna"
+h7b <- "gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(educ.null, educ1, educ2, educ3, educ4, educ5, educ6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Total matrícula privada",
+                                      "Varianza Simce comuna",
+                                      "Promedio Simce comuna"
+                              )
+          #omit.coef = omit
+          )
+```
+
+## B. Pension distribution justice
+
+```{r distribution-pension}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_pension:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+
+# B. Pension distributive justice
+pension.null <- lmer(formula(paste0("just_pension~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+pension1 <- lmer(formula(paste0("just_pension~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension2 <- lmer(formula(paste0("just_pension~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension3 <- lmer(formula(paste0("just_pension~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension4 <- lmer(formula(paste0("just_pension~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension5 <- lmer(formula(paste0("just_pension~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+pension6 <- lmer(formula(paste0("just_pension~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(pension.null, pension1, pension2, pension3, pension4, pension5, pension6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: Familia",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: trabajo duro",
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Total matrícula privada",
+                                      "Varianza Simce comuna",
+                                      "Promedio Simce comuna",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"
+                              )
+          #omit.coef = omit
+          )
+```
+
+## C. Health distribution justice
+
+```{r distribution-health}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_salud:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+
+# C. Health distributive justice
+salud.null <- lmer(formula(paste0("just_salud~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+salud1 <- lmer(formula(paste0("just_salud~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud2 <- lmer(formula(paste0("just_salud~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud3 <- lmer(formula(paste0("just_salud~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud4 <- lmer(formula(paste0("just_salud~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud5 <- lmer(formula(paste0("just_salud~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+salud6 <- lmer(formula(paste0("just_salud~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(salud.null, salud1, salud2, salud3, salud4, salud5, salud6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: Familia",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: trabajo duro",
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Total matrícula privada",
+                                      "Varianza Simce comuna",
+                                      "Promedio Simce comuna",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"
+                              )
+          #omit.coef = omit
+          )
+```
+
+## D. Income differences are too large
+
+```{r inequality}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study2_long_comunas.RData"))
+
+# generate analytical sample
+df_study2 <- 
+  df_study2_long_comunas %>%
+  select(idencuesta,ola,comuna,des_perc:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+
+# C. Health distributive justice
+des.null <- lmer(formula(paste0("des_perc~","1 + (1|idencuesta)+(1|comuna)")),data = df_study2)
+des1 <- lmer(formula(paste0("des_perc~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des2 <- lmer(formula(paste0("des_perc~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des3 <- lmer(formula(paste0("des_perc~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des4 <- lmer(formula(paste0("des_perc~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des5 <- lmer(formula(paste0("des_perc~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+des6 <- lmer(formula(paste0("des_perc~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study2)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(des.null, des1, des2, des3, des4, des5, des6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+           #                           "Ola 2021",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: Familia",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: trabajo duro",
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Total matrícula privada",
+                                      "Varianza Simce comuna",
+                                      "Promedio Simce comuna",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"
+                              )
+          #omit.coef = omit
+          )
+```

### 2023-12-11 production/prod_prep-mergedata.Rmd
update alternativas y readme
* update analisis alternativos
* update link readme

@@
 
+df_study2_long_comunas <- df_study2_long_comunas %>% mutate(year= case_when(ola==1~2017,
+                                                                            ola==2~2017,
+                                                                            ola==3~2017,
+                                                                            ola==4~2022,
+                                                                            ola==6~2022))
+
+df_study2_long_comunas <- 
+ dplyr::left_join(x = df_study2_long_comunas,
+                  y = select(df_gini_comunas_long, comunacod=COMUNA_COD,year, starts_with("gini"),prop_univ), 
+                  by = c("comunacod", "year"))
+
 df_study2_long_comunas <- 
@@
 
+df_study2_long_comunas <- df_study2_long_comunas %>% mutate(year= case_when(ola==1~2016,
+                                                                            ola==2~2017,
+                                                                            ola==3~2018,
+                                                                            ola==4~2019,
+                                                                            ola==5~2021,
+                                                                            ola==6~2022))
+
 df_study1_long_comunas <- 
@@
                   by = c("comunacod"))
+
+df_study2_long_comunas <- 
+ dplyr::left_join(x = df_study2_long_comunas,
+                  y = select(df_matricula_comunas_long, comunacod=cod_com_rbd_15,year, total_mat_pagado,prop_pagado), 
+                  by = c("comunacod", "year"))
 
@@
                   by = c("comunacod"))
+
+df_study2_long_comunas <- 
+ dplyr::left_join(x = df_study2_long_comunas,
+                  y = select(df_simce_comunas_long, comunacod=cod_com_15,year, prom_simce_comuna,var_simce_comuna), 
+                  by = c("comunacod", "year"))
 

### 2023-12-19 production/prod_analysis.Rmd
add contextuales

@@
   df_study1_long_comunas %>%
-  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna, univlag22_17) %>%
   na.omit() %>% 
@@
 h2 <- "ola+merit_effort+merit_talent"
-h3 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork"
-h4 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc"
-h5 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna"
-h6 <- "ola+merit_effort+merit_talent+ahead_family+ahead_educ+ahead_ambition+ahead_hardwork+gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna+educ+quintil1+ess+pos_id+edad+sexo"
+h3 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo"
+h4 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17"
+h5 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ"
+h6 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+univlag22_17"
+h7 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ+univlag22_17"
+h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+var_simce_comuna+prom_simce_comuna"
 
@@
 educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::knitreg(list(educ1, educ2, educ3, educ4, educ5, educ6, educ7,educ8),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+#                custom.coef.names = c("Intercepto",
+#                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+#                                      "ola 2018",
+#                                      "ola 2019",
+##                                      "ola 2022",
+#                                      "Merito: esfuerzo",
+#                                      "Merito: Inteligencia",
+##                                      "Goahead: Familia",
+#                                      "Goahead: educacion",
+#                                      "Goahead: ambicion",
+#                                      "Goahead: trabajo duro",
+#                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+#                                      "Ed. técnica",
+#                                      "Ed. universitaria",
+#                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+#                                      "Quintil 3",
+#                                      "Quintil 4",
+#                                      "Quintil 5",
+#                                      "Quintil NA",
+#                                      "Estatus social subjetivo",
+#                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+#                                      "Pos. política: Derecha",
+#                                      "Pos. política: No se identifica",
+#                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+#                                      "Edad 50-64",
+#                                      "Edad 65 o más",
+#                                      "Mujer <br> <i>(Ref. Hombre)</i>",
+#                                      "Gini comunal",
+#                                      "Cambio gini 2022-2017",
+#                                      "Proporción educ universitaria",
+#                                      "Proporción matrícula privada",
+#                                      "Total matrícula privada",
+#                                      "Varianza Simce comuna",
+#                                      "Promedio Simce comuna"
 
-#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::htmlreg(list(educ.null, educ1, educ2, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
-                custom.coef.names = c("Intercepto",
-                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
-                                      "ola 2018",
-                                      "ola 2019",
-                                      "ola 2022",
-                                      "Merito: esfuerzo",
-                                      "Merito: Inteligencia",
-                                      "Goahead: Familia",
-                                      "Goahead: educacion",
-                                      "Goahead: ambicion",
-                                      "Goahead: trabajo duro",
-                                      "Gini comunal",
-                                      "Cambio gini 2022-2017",
-                                      "Proporción matrícula privada",
-                                      "Total matrícula privada",
-                                      "Varianza Simce comuna",
-                                      "Promedio Simce comuna",
-                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
-                                      "Ed. técnica",
-                                      "Ed. universitaria",
-                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
-                                      "Quintil 3",
-                                      "Quintil 4",
-                                      "Quintil 5",
-                                      "Quintil NA",
-                                      "Estatus social subjetivo",
-                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
-                                      "Pos. política: Derecha",
-                                      "Pos. política: No se identifica",
-                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Edad 50-64",
-                                      "Edad 65 o más",
-                                      "Mujer <br> <i>(Ref. Hombre)</i>"
                               )
           #omit.coef = omit
-          )
+
 ```

### 2023-12-19 production/prod_analysis.Rmd
Update prod_analysis.Rmd

@@
   df_study1_long_comunas %>%
-  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna, univlag22_17) %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado,pagadolag22_16, total_mat_pagado, var_simce_comuna, prom_simce_comuna, univlag22_17) %>%
   na.omit() %>% 
@@
 h7 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ+univlag22_17"
-h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+var_simce_comuna+prom_simce_comuna"
+h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna"
+h9 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+var_simce_comuna"
+h10 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna+var_simce_comuna"
+h11 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado"
+h12 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+pagadolag22_16"
+h13 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16"
+h13 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16"
+
+h13.1 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16+prom_simce_comuna+var_simce_comuna+prop_univ+univlag22_17+gini_ypch+ginilag22_17"
 
@@
 educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ9 <- lmer(formula(paste0("just_educ~",h9,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ10 <- lmer(formula(paste0("just_educ~",h10,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ11 <- lmer(formula(paste0("just_educ~",h11,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ12 <- lmer(formula(paste0("just_educ~",h12,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ13 <- lmer(formula(paste0("just_educ~",h13,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ13.1 <- lmer(formula(paste0("just_educ~",h13.1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
 #omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(educ1, educ2, educ3, educ4, educ5, educ6, educ7,educ8),
+texreg::knitreg(list(educ1, educ2, educ3, educ4, educ5, educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13,educ13.1),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 

### 2024-01-09 production/prod_analysis.Rmd
add interacciones

@@
 h13 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16"
-h13 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16"
 
@@
 #omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(educ1, educ2, educ3, educ4, educ5, educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13,educ13.1),
+texreg::htmlreg(list(educ1, educ2, educ3, educ4, educ5, educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13,educ13.1),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
@@
 
-## Contextual analysis
+## A2. Education distribution justice (export tables)
+
+```{r}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study1_long_comunas %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1 <- "ola"
+h2 <- "ola+merit_effort+merit_talent"
+h3 <- "ola+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
+h4 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
+h5 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo"
+
+h6 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17"
+h7 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ+univlag22_17"
+h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17"
+h9 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadlag22_16"
+h10 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+h11 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+
+hfull <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ9 <- lmer(formula(paste0("just_educ~",h9,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ10 <- lmer(formula(paste0("just_educ~",h10,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ11 <- lmer(formula(paste0("just_educ~",h11,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ12 <- lmer(formula(paste0("just_educ~",hfull,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+# Table 1: individual
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(educ1, educ2, educ3, educ4, educ5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: trabajo duro",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: Familia",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"),
+                file = "output/tables/table1-individuales.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table1-individuales.html" ,file ="output/tables/table1-individuales.png")
+
+```
+
+```{r}
+# table 2: contextuales
+omit <- "(ola)|(merit_effort)|(merit_talent)|(ahead_hardwork)|(ahead_educ)|(ahead_ambition)|(ahead_family)|(educ)|(quintil1)|(ess)|(pos_id)|(edad)|(sexo)|(Intercept)"
+texreg::htmlreg(list(educ6, educ7,educ8,educ9,educ10,educ11,educ12),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                omit.coef = omit,
+                custom.coef.names = c(
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción educ universitaria",
+                                      "Cambio prop. universitaria 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Cambio matrícula privada 2022-2016",
+                                      "Promedio Simce comuna",
+                                      "Varianza Simce comuna"),
+                file = "output/tables/table2-contextual.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table2-contextual.html" ,file ="output/tables/table2-contextual.png")
+
+```
+
+## Interacciones
 
@@
   df_study1_long_comunas %>%
-  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, total_mat_pagado, var_simce_comuna, prom_simce_comuna) %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17) %>%
   na.omit() %>% 
-  mutate(ola=as.factor(ola),
+  mutate(ola=as.numeric(ola),
          sexo=as.factor(sexo)
@@
 # Hipotesis
-h1b <- "gini_ypch"
-h2b <- "gini_gc"
-h3b <- "prop_pagado"
-h4b <- "total_mat_pagado"
-h5b <- "var_simce_comuna"
-h6b <- "prom_simce_comuna"
-h7b <- "gini_ypch+gini_gc+prop_pagado+total_mat_pagado+var_simce_comuna+prom_simce_comuna"
+h1b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+h2b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*merit_effort"
+h3b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*merit_talent"
+h4b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_hardwork"
+h5b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_educ"
+h6b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_ambition"
+h7b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_family"
 
@@
 
-#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::htmlreg(list(educ.null, educ1, educ2, educ3, educ4, educ5, educ6),
+# table 3: interacciones
+texreg::htmlreg(list(educ1, educ2,educ3,educ4,educ5,educ6,educ7),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
                 custom.coef.names = c("Intercepto",
+                                      "ola",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: trabajo duro",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: Familia",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>",
                                       "Gini comunal",
                                       "Cambio gini 2022-2017",
+                                      "Proporción educ universitaria",
+                                      "Cambio prop. universitaria 2022-2017",
                                       "Proporción matrícula privada",
-                                      "Total matrícula privada",
+                                      "Cambio matrícula privada 2022-2016",
+                                      "Promedio Simce comuna",
                                       "Varianza Simce comuna",
-                                      "Promedio Simce comuna"
-                              )
+                                      "ola*merit:esfuerzo",
+                                      "ola*merit:talento",
+                                      "ola*goahead:hardwork",
+                                      "ola*goahead:educ",
+                                      "ola*goahead:ambition",
+                                      "ola*goahead:family"),
+                file = "output/tables/table3-interac.html")
           #omit.coef = omit
-          )
+
+webshot2::webshot(url ="output/tables/table3-interac.html" ,file ="output/tables/table3-interac.png")
 ```

### 2024-01-09 production/prod_prep-mergedata.Rmd
add interacciones

@@
 
+## Rescale simce
+summary(df_study1_long_comunas$prom_simce_comuna)
+summary(df_study1_long_comunas$var_simce_comuna)
+
+df_study1_long_comunas$prom_simce_comuna_rescaled <- df_study1_long_comunas$prom_simce_comuna / 10
+df_study1_long_comunas$var_simce_comuna_rescaled <- df_study1_long_comunas$var_simce_comuna / 100
+# Verificar el rescalado
+summary(df_study1_long_comunas$prom_simce_comuna_rescaled)
+summary(df_study1_long_comunas$var_simce_comuna_rescaled)
 # save data____________________________________________________________________

### 2024-01-10 production/prod_prep-casen.R
income casen comunas 2015 2022

@@ -30,7 +30,8 @@ gini15 <-
   df15 %>%
   group_by(COMUNA) %>%
   summarise("gini_ypch" = as.numeric(acid::weighted.gini(x = ypch, w =expc)$bcwGini),
-            "prop_univ"= collapse::fmean(x = univ,w=expc,na.rm=TRUE)) %>%
+            "prop_univ"= collapse::fmean(x = univ,w=expc,na.rm=TRUE),
+            "inc_mean"= mean(ypch,na.rm = T)) %>%
   mutate(year = 2015)
 
 # Casen 2017 --------------------------------------------------------------
@@ -60,7 +61,8 @@ gini17 <-
   group_by(COMUNA) %>%
   summarise("gini_ypch" = as.numeric(acid::weighted.gini(x = ypch, w =
                                                            expc)$bcwGini),
-            "prop_univ"= collapse::fmean(x = univ,w=expc,na.rm=TRUE)
+            "prop_univ"= collapse::fmean(x = univ,w=expc,na.rm=TRUE),
+            "inc_mean"= mean(ypch,na.rm = T)
             ) %>%
   mutate(year = 2017)
 names(gini17)
@@ -101,7 +103,8 @@ gini22 <-
   group_by(COMUNA) %>%
   summarise("gini_ypch" = as.numeric(acid::weighted.gini(x = ypch, w =
                                                            expc)$bcwGini),
-            "prop_univ"= collapse::fmean(x = univ,w=expc,na.rm=TRUE)) %>%
+            "prop_univ"= collapse::fmean(x = univ,w=expc,na.rm=TRUE),
+            "inc_mean"= mean(ypch,na.rm = T)) %>%
   mutate(year = 2022)
 
 
@@ -134,7 +137,7 @@ df_gini_comunas_wide <-gini15 %>%
   full_join(gini17, by = "COMUNA", suffix = c("_2015", "_2017")) %>%
   full_join(gini22, by = "COMUNA", suffix = c("_2022", "")) %>%
   select(everything(), starts_with("gini_ypch"), -starts_with("year")) %>%
-  rename(COMUNA_COD = COMUNA, gini_ypch_2022 = gini_ypch,prop_univ_2022 = prop_univ) %>%
[…]

### 2024-01-10 production/prod_prep-casen.R
variables lagged

@@ -161,10 +161,12 @@ df_gini_comunas_wide <-gini15 %>%
     univlag22_17 = prop_univ_2022 - prop_univ_2017,
     univlag17_15 = prop_univ_2017 - prop_univ_2015,
 # income comunal
-incomelag22_15 = inc_mean_2022 - inc_mean_2015,
-incomelag22_17 = inc_mean_2022 - inc_mean_2017,
-incomelag17_15 = inc_mean_2017 - inc_mean_2015
-)
+  incomelag22_15 = inc_mean_2022 - inc_mean_2015,
+  incomelag22_17 = inc_mean_2022 - inc_mean_2017,
+  incomelag17_15 = inc_mean_2017 - inc_mean_2015,
+  logincomelag22_15 = log(inc_mean_2022) - log(inc_mean_2015),
+  logincomelag22_17 = log(inc_mean_2022) - log(inc_mean_2017),
+  logincomelag17_15 = log(inc_mean_2017) - log(inc_mean_2015))
 
 summary(df_gini_comunas_wide)
 

### 2024-01-10 production/prod_prep-casen.R
Update prod_prep-casen.R

@@ -127,7 +127,9 @@ df_gini_comunas_long <-
     year,
     everything(),
     -X
-  )
+  ) %>%
+  mutate(loginc_mean=inc_mean)
+summary(df_gini_comunas_long)
 
 #Center within cluster (comuna) GINI
 df_gini_comunas_long$gini_gc <-  MLMusingR::group_center(as.numeric(df_gini_comunas_long$gini_ypch), grp = df_gini_comunas_long$COMUNA_COD)

### 2024-01-26 production/prod_analysis.Rmd
reorder contextual effects

@@
   df_study1_long_comunas %>%
-  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17) %>%
+  select(idencuesta,ola,comuna,just_educ:merit_talent, ahead_family:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17, loginc_mean, logincomelag22_17) %>%
   na.omit() %>% 
@@
 h11 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+h12 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+loginc_mean+logincomelag22_17"
 
-hfull <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+hfull <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+loginc_mean+logincomelag22_17"
 
@@
 educ11 <- lmer(formula(paste0("just_educ~",h11,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ12 <- lmer(formula(paste0("just_educ~",hfull,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ12 <- lmer(formula(paste0("just_educ~",h12,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ13 <- lmer(formula(paste0("just_educ~",hfull,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
 
@@
 omit <- "(ola)|(merit_effort)|(merit_talent)|(ahead_hardwork)|(ahead_educ)|(ahead_ambition)|(ahead_family)|(educ)|(quintil1)|(ess)|(pos_id)|(edad)|(sexo)|(Intercept)"
-texreg::htmlreg(list(educ6, educ7,educ8,educ9,educ10,educ11,educ12),
+texreg::htmlreg(list(educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
@@
                                       "Promedio Simce comuna",
-                                      "Varianza Simce comuna"),
+                                      "Varianza Simce comuna",
+                                      "Ingreso comunal",
+                                      "Cambio ingreso comunal 2022-2017"),
                 file = "output/tables/table2-contextual.html")
@@
   na.omit() %>% 
-  mutate(ola=as.numeric(ola),
+  mutate(ola=as.factor(ola),
          sexo=as.factor(sexo)
@@
 h2b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*merit_effort"
-h3b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*merit_talent"
+h3b <- "merit_effort+merit_talent+ola+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+merit_talent*ola"
 h4b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_hardwork"
@@
 # table 3: interacciones
-texreg::htmlreg(list(educ1, educ2,educ3,educ4,educ5,educ6,educ7),
+texreg::knitreg(list(educ1, educ2,educ3,educ4,educ5,educ6,educ7),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
                 custom.coef.names = c("Intercepto",
-                                      "ola",
+                                      "2017",
+                                      "2018",
+                                      "2019",
+                                      "2022",
                                       "Merito: esfuerzo",
@@
 webshot2::webshot(url ="output/tables/table3-interac.html" ,file ="output/tables/table3-interac.png")
+
+## graph interact
+#sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
+#predit<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
+#predit$data[c(1,2,4,6,8),]
+#predit$data[c(1,2,4,6,8),] %>% ggplot2::ggplot() %>%
 ```

### 2024-01-26 production/prod_prep-mergedata.Rmd
reorder contextual effects

@@
  dplyr::left_join(x = df_study1_long_comunas,
-                  y = select(df_gini_comunas_long, comunacod=COMUNA_COD,year, starts_with("gini"),prop_univ), 
+                  y = select(df_gini_comunas_long, comunacod=COMUNA_COD,year, starts_with("gini"), prop_univ, starts_with("inc"), starts_with("log")), 
                   by = c("comunacod", "year"))
@@
  dplyr::left_join(x = df_study1_long_comunas,
-                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini"), starts_with("prop_univ"), starts_with("univlag")), 
+                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini"), starts_with("prop_univ"), starts_with("univlag"), starts_with("inc"), starts_with("log")), 
                   by = c("comunacod"))

### 2024-03-04 production/prod_analysis.Rmd
update graphs

@@
                kableExtra,
-               summarytools
+               summarytools,
+               ggplot2,
+               sjPlot,
+               tidyverse
                )
@@
 # Df variables nivel 1 __________________________________________________________
-df_study1 <- df_study1_long_comunas %>% select(just_educ:merit_talent,
+df_study1 <- df_study1_long_comunas %>% select(ola,
+                                               just_educ:merit_talent,
                                                ahead_family:sexo) 
@@
 # print(df, method = "render")
+```
+
+```{r bivariate-plots}
+merit_plot <- sjPlot::plot_stackfrq(select(df_study1, ahead_family, ahead_ambition, ahead_educ, ahead_hardwork, merit_talent, merit_effort),
+                      expand.grid =TRUE,
+                      geom.colors = 'Spectral',
+                      show.total = FALSE, 
+                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),6),
+                      legend.labels = c("Strongly disagree", "Disagree", "Neither agree
+nor disagree", "Agree", "Strongly agree"),
+                      axis.labels = c("Ahead family", "Ahead ambition", "Ahead education", "Ahead hardwork", "Merit talent", "Merit effort")
+                      ) +
+  theme_light() +
+  theme(legend.position="bottom", text = element_text(size = 14))
+merit_plot
+
+ggsave(merit_plot, file="output/graphs/merit_plot.png", height = 7, width = 8)
+```
+
+justification education vs year
+
+```{r}
+just_educ_year = df_study1 %>%# filter(!is.na(esfuerzo_soc)) %>% 
+  group_by(ola) %>%
+  summarise(education = mean(just_educ, na.rm = T))
+
+categorias_x <- c(2016, 2017, 2018, 2019, 2022)
+
+educ_year_plot <- just_educ_year %>% ggplot(aes(ola, education)) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4)+
+  ylab("Justification of inequality") +
+  xlab("Wave") +
+  scale_x_discrete(labels = categorias_x) +
+  ylim(1,3)
+educ_year_plot
+
+ggsave(educ_year_plot, file="output/graphs/educ_year_plot.png", width = 7, height = 7)
+```
+
+Bivariate descriptive by year
+
+probar ponderadores longitudinales
+
+```{r}
+educ_effort_year = df_study1 %>%
+  group_by(ola, merit_effort) %>%
+  summarise(education = mean(just_educ, na.rm = T)) %>% 
+  ungroup()
+
+educ_talent_year = df_study1 %>%
+  group_by(ola, merit_talent) %>%
+  summarise(education = mean(just_educ, na.rm = T)) %>% 
+  ungroup()
+
+categorias_x <- c("Strongly disagree", "Disagree", "Neither agree 
+nor disagree", "Agree", "Strongly agree")
+categorias_f <- c(2016, 2017, 2018, 2019, 2022)
+
+effort_educ_year_plot <- educ_effort_year %>% 
+  ggplot(aes(factor(merit_effort), education, color = factor(ola), shape = factor(ola))) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4) +
+  ylab("Justification of inequality") +
+  xlab("Merit: effort") +
+  ylim(1, 3) +
+  scale_x_discrete(labels = categorias_x) +
+  scale_color_discrete(labels = categorias_f, name="Wave") +
+  scale_shape_discrete(labels = categorias_f, name="Wave")
+
+effort_educ_year_plot
+
+talent_educ_year_plot <- educ_talent_year %>% 
+  ggplot(aes(factor(merit_talent), education, color = factor(ola), shape = factor(ola))) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4) +
+  ylab("Justification of inequality") +
+  xlab("Merit: talent") +
+  ylim(1, 3) +
+  scale_x_discrete(labels = categorias_x) +
+  scale_color_discrete(labels = categorias_f, name="Wave") +
+  scale_shape_discrete(labels = categorias_f, name="Wave")
+
+talent_educ_year_plot
+
+ggsave(effort_educ_year_plot, file="output/graphs/effort_educ_year.png", height = 7, width = 7)
+ggsave(talent_educ_year_plot, file="output/graphs/talent_educ_year.png", height = 7, width = 7)
+
+educ_effort_year <- educ_effort_year %>% pivot_longer("merit_effort", 
+                              names_to="merit", 
+                              values_to="value")
+
+educ_talent_year <- educ_talent_year %>% pivot_longer("merit_talent", 
+                              names_to="merit", 
+                              values_to="value")
+
+educ_merit_year <- rbind(educ_effort_year, educ_talent_year)
+
+custom_labeller <- as_labeller(function(variable, value) {
+  return(categorias_f[value])
+})
+categorias_z <- c("Effort", "Talent")
+merit_educ_year <- educ_merit_year %>% 
+  ggplot(aes(factor(value), education)) +
+  theme_bw(base_size = 14) +
+  geom_point(aes(color=merit, shape=merit), size=3) +
+  ylab("Justification of inequality") +
+  xlab("Merit") +
+  ylim(1, 3) +
+  facet_wrap(~factor(ola), scales = "free", labeller = custom_labeller) +
+  scale_color_discrete(labels = categorias_z, name="Merit") +
+  scale_shape_discrete(labels = categorias_z, name="Merit")
+merit_educ_year
+
+ggsave(merit_educ_year, file="output/graphs/merit_educ_year.png", height = 7, width = 7)
 ```
@@
                lme4,
-               texreg
+               texreg,
+               ggplot2
                )
@@
 #sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
-#predit<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
-#predit$data[c(1,2,4,6,8),]
-#predit$data[c(1,2,4,6,8),] %>% ggplot2::ggplot() %>%
+predict<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
+predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
+predict_data <- predict_data %>% as.data.frame()
+
+predict_data$value <- c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6")
+
+predict_data$value <- factor(predict_data$value,
+                   levels = c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6"),
+                   labels = c("wave 1", 
+                              "wave 2",
+                              "wave 3",
+                              "wave 4",
+                              "wave 6"))
+
+interact_effect <- predict_data %>% ggplot(aes(value, coef1)) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4)+
+  geom_errorbar(aes(ymin=lb, ymax=ub),
+                  width=0.2) +
+  ylab("Coef") +
+  xlab("")
+
+ggsave(interact_effect, file="output/graphs/interact_effect.png", width = 7, height = 7)
 ```

### 2024-03-13 production/prod_analysis-fix-t5.Rmd
modificaciones gráficos + add percepcion desigualdad

--- 2022-08-02 production/prod_analysis-fix-t5.Rmd
@@
 body .main-container {
-      max-width: 2000px !important;
-      width: 2000px !important;
+      max-width: 1600px !important;
+      width: 1600px !important;
     }
@@
                summarytools,
-               sjlabelled
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+               ggplot2,
+               sjPlot,
+               tidyverse,
+               ggalluvial,
+               survey,
+               shadowtext
+               )
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
+
 # Css
@@
 
-# Descriptive table __________________________________________________________
-df_study1 <- 
-  df_study_long_t5_comunas %>% 
-  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) 
-# extract variable labels
+# Df variables nivel 1
+# df<- dfSummary(df_study1_long %>% select(in_simpat:edad),
+#                plain.ascii = FALSE,
+#                style = "grid",
+#                tmp.img.dir = "/tmp",
+#                graph.magnif = TRUE,
+#                headings = F,  # encabezado
+#                varnumbers = F, # num variable
+#                labels.col = T, # etiquetas
+#                na.col = F,    # missing
+#                graph.col = T, # plot
+#                valid.col = T, # n valido
+#                col.widths = c(700,10,10,10,10)
+#                )
+# df$Variable <- NULL # delete variable column
+# view(df,
+#      file = here::here("output/tables/desc01.html")
+#      ) # Ver tabla en un archivo HTML
+# webshot::webshot(url =here::here("output/tables/desc01.html"),
+#         file =here::here("output/tables/desc01.png")) # Remover archivo HTML
+# # print(df, method = "render")
+
+# Df variables nivel 1 _________________________________________________________
+df_study1 <- df_study_long_t6_comunas %>% dplyr::select(ola,
+                                               just_educ, merit_effort, merit_talent,
+                                               ahead_family:educ, quintil1:sexo)
+
+df_study1$ola <- factor(df_study1$ola,
+         labels = c("2016","2017","2018","2019", "2022"))
+df_study1$ola <- 
+sjlabelled::set_label(x = df_study1$ola,
+                      label = "Wave")
+
+df_study1$sexo <- factor(df_study1$sexo,
+         labels = c("Male", "Female"))
+df_study1$sexo <- 
+sjlabelled::set_label(x = df_study1$sexo,
+                      label = "Gender")
+
+for (i in c("just_educ",
+            "merit_effort","merit_talent")) {
+  df_study1[[i]] <- factor(df_study1[[i]],
+                            levels = 1:5,
+                            labels = c("Strongly disagree","Disagree",
+                                       "Neither disagree nor agree",
+                                       "Agree", "Strongly agree"))
+}
+
+for (i in c("ahead_family","ahead_educ",
+            "ahead_ambition", "ahead_hardwork")) {
+  df_study1[[i]] <- factor(df_study1[[i]],
+                            levels = 1:5,
+                           labels = c("Not at all important",
+                                  "A little important",
+                                  "Somewhat important",
+                                  "Pretty important",
+                                  "Very important"))
+}
+
+df_study1$just_educ <- 
+sjlabelled::set_label(x = df_study1$just_educ, 
+                      label = "Education distributive justice")
+
+#
+df_study1$merit_effort <- 
+sjlabelled::set_label(x = df_study1$merit_effort, 
+                      label = "People are rewarded for their efforts")
+
+#
+df_study1$merit_talent <- 
+sjlabelled::set_label(x = df_study1$merit_talent, 
+                      label = "People are rewarded for their intelligence")
+
+#
+df_study1$ahead_family <- 
+sjlabelled::set_label(x = df_study1$ahead_family, 
+                      label = "Coming from a wealthy family")
+
+#
+df_study1$ahead_educ <- 
+sjlabelled::set_label(x = df_study1$ahead_educ, 
+                      label = "Having a good level of education")
+
+#
+df_study1$ahead_ambition <- 
+sjlabelled::set_label(x = df_study1$ahead_ambition, 
+                      label = "Having ambition")
+
+#
+df_study1$ahead_hardwork <- 
+sjlabelled::set_label(x = df_study1$ahead_hardwork, 
+                      label = "Hard work")
+
 lab_study1 <- sjlabelled::get_label(df_study1) 
-# convert items to factor
-for (i in names(select(df_study1,starts_with("in_")))) {
-  df_study1[[i]] <- factor(df_study1[[i]],1:5,get_labels(df_study1[[i]]))
-}
-# set variable labels
 df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
-df_study1$in_amreal <- sjlabelled::set_label(df_study1$in_amreal,label = "Threat (Unemployment increases)")
-
-df<- dfSummary(x = df_study1,
+
+df<- dfSummary(x = dplyr::select(df_study1, ola, just_educ, merit_effort, merit_talent, ahead_hardwork, ahead_educ, ahead_ambition, ahead_family, educ, quintil1, ess, pos_id, edad, sexo),
                plain.ascii = FALSE,
@@
                graph.col = T, # plot
-               valid.col = T, # n valido
-               # col.widths = c(700,10,10,10,10,10)
-               )
-# df$Variable <- NULL # delete variable column
-view(df,
-     file = here::here("output/tables/desc01.html")
+               valid.col = F, # n valido
+               col.widths = c(700,10,10,10)
+               )
+df$Variable <- NULL # delete variable column
+summarytools::view(df,
+     file = here::here("output/tables/desc02.html")
      ) # Ver tabla en un archivo HTML
-webshot::webshot(url =here::here("output/tables/desc01.html"),
-        file =here::here("output/tables/desc01.png")) # Remover archivo HTML
+webshot::webshot(url =here::here("output/tables/desc02.html"),
+        file =here::here("output/tables/desc02.png")) # Remover archivo HTML
 # print(df, method = "render")
-
-individual <- 
-  df_study1 %>% 
-  select(conviviality, identity,
-         in_amreal,
-         educ,quintil1,ess,know_inm_t1t4,frien_inm_t1t4)
-
-df_individual<- dfSummary(x = individual,
-               plain.ascii = FALSE,
-               style = "grid",
-               tmp.img.dir = "/tmp",
-               graph.magnif = TRUE,
-               headings = F,  # encabezado
-               varnumbers = F, # num variable
-               labels.col = T, # etiquetas
-               na.col = F,    # missing
-               graph.col = T, # plot
-               valid.col = F,# n valido
-               # max.distinct.values = 15,
-               max.string.width = 50,
-               # col.widths = c(700,10,10,10,10)
-               )
-df_individual$Variable <- NULL # delete variable column
-df_individual <- df_individual %>% rename("Variable"=Label)
-
-view(df_individual,
-     file = here::here("output/tables/ds_individual.html")
-     ) # Ver tabla en un archivo HTML
-webshot::webshot(url =here::here("output/tables/ds_individual.html"),
-        file =here::here("output/tables/table2.jpg")) # Remover archivo HTML
-# print(df, method = "render")
-
-comunal <- 
-  df_study1 %>% 
-  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl, #conviviality
-         in_simili,in_amsimb,in_acpier,in_acadop, #identity
-         conviviality, identity, 
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  select(inm_prop_cas17_w,inm_prop_cas20_w,dif_inm)
-
-df_municipal<- dfSummary(x = comunal,
-               plain.ascii = FALSE,
-               style = "grid",
-               tmp.img.dir = "/tmp",
-               graph.magnif = TRUE,
-               headings = F,  # encabezado
-               varnumbers = F, # num variable
-               labels.col = T, # etiquetas
-               na.col = F,    # missing
-               graph.col = T, # plot
-               valid.col = F,# n valido
-               # max.distinct.values = 15,
-               # max.string.width = 50,
-               # col.widths = c(700,10,10,10,10)
-               )
-
-df_municipal$Variable <- NULL # delete variable column
-df_municipal$`Freqs (% of Valid)` <- NULL # delete variable column
-df_municipal <- df_municipal %>% rename("Variable"=Label)
-
-view(df_municipal,
-     file = here::here("output/tables/ds_municipal.html")
-     ) # Ver tabla en un archivo HTML
-webshot::webshot(url =here::here("output/tables/ds_municipal.html"),
-        file =here::here("output/tables/table3.jpg")) # Remover archivo HTML
-# print(df, method = "render")
-```
-
-```{r likert}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-# cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               sjlabelled,
-               ggplot2,
-               ggrepel,
-               cowplot,
-               sjPlot,
-               ggpubr
-)
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-# Descriptive  __________________________________________________________
-df_study1 <- 
-  df_study_long_t5_comunas %>% 
-  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm)
-
-# para modelar estaba invertido:
-#original: fraseo negativo: mayor acuerdo es "negativo" porque se pierde identidad
-#recodificado: inversion de valores 5 es "menos perdida". Es positivo.
-#ecodificado a sentido original para plot: 5 es "mayor perdida". Es negativo
-sjmisc::frq(df_study1$in_amsimb)
-sjmisc::frq(sjmisc::rec(df_study1$in_amsimb,rec = "rev"))
-df_study1$in_amsimb <- sjmisc::rec(df_study1$in_amsimb,rec = "rev")
-# extract variable labels
-lab_study1 <- sjlabelled::get_label(df_study1)
-# # convert items to factor
-# for (i in names(select(df_study1,starts_with("in_")))) {
-#   df_study1[[i]] <- factor(df_study1[[i]],1:5,get_labels(df_study1[[i]]))
-# }
-# set variable labels
-df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
-df_study1$in_amreal <- sjlabelled::set_label(df_study1$in_amreal,label = "Threat (Unemployment increases)")
-df_study1$in_amsimb <- sjlabelled::set_label(df_study1$in_amsimb,label = "Loose of identity because of migrants")
-
-df_study1 <- 
-  df_study1 %>% 
-  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         in_amreal)
-# view_df(df_study1)
-
-# se invierten todos los items para el plot, NO SU SENTIDO  para modelamiento
-for (i in names(df_study1)) {
-  df_study1[[i]] <- sjmisc::rec(x = df_study1[[i]],rec = "rev")  
-}
-
-  # Declare ggplot2 features
-  ggplot2::theme_set(ggplot2::theme(panel.background = ggplot2::element_rect(fill = "gray85",
-                                                                             colour = "gray85"),
-                                    panel.border = ggplot2::element_blank(),
-                                    axis.text.y = ggplot2::element_text(size = 13,
-                                                                        hjust = 1),
-                                    title = ggplot2::element_text(size = 18,
-                                                                  face = "bold"),
-                                    legend.text = ggplot2::element_text(size = 12),
-                                    plot.caption = ggplot2::element_text(size = 10,
-                                                                         face = "plain",
-                                                                         hjust = 1
-                                                                         )
-                                    )
-                     )
-#
-
-#Plot 1: Conviality-----------------------------------------------------------
-p1<-
-  dplyr::select(df_study1,in_ansied,in_simpat,in_valfam,in_valami,in_amigcl) %>%
+```
+
+```{r}
+df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+                                               just_educ, merit_effort, merit_talent,
+                                               ahead_family:educ, quintil1:sexo)
+df_study1 <- df_study1[complete.cases(df_study1$segmento), ]
+
+df_study1$just_educ <- factor(df_study1$just_educ,
+         labels = c("Strongly disagree", "Disagree", "Neither agree
+nor disagree", "Agree", "Strongly agree"))
+
+elsoc_diseno <- svydesign(ids = ~segmento, #muestreo por conglomerado a nivel de manzanas (segmento)
+                          strata = ~estrato, #muestreo estratificado a nivel ciudad (estato)
+                          weights = ~ponderador_long_total, #ponderador longitudinal
+                          nest = TRUE,
+                          data = df_study1)
+
+#Paso 1
+datos.justeduc <- data.frame((svytable(~just_educ + ola + idencuesta, elsoc_diseno, round = F))) %>% dplyr::filter(Freq>0)  %>% group_by(ola) %>% mutate(porcentaje=Freq/sum(Freq)) %>% na.omit()
+
+#Paso 2
+etiquetas.justeduc <- data.frame((svytable(~just_educ + ola, elsoc_diseno, round = F))) %>% group_by(ola) %>% mutate(porcentaje=Freq/sum(Freq)) %>% na.omit() %>% 
+  mutate(idencuesta = 1)
+```
+
+```{r alluvial}
+alluvial_justeduc <- ggplot(datos.justeduc, aes(x = ola, fill = just_educ, stratum = just_educ,
+                             alluvium = idencuesta, y = porcentaje))+
+    ggalluvial::geom_flow(alpha = .66) + 
+    ggalluvial::geom_stratum(linetype = 0) +
+    scale_y_continuous(labels = scales::percent) + 
+    ylab(label = NULL) +
+    xlab(label = NULL) + 
+    theme(legend.position = 'top',
+          legend.title = element_blank(),
+          plot.caption = element_text(hjust = 0)) +
+    scale_fill_viridis_d(begin = 0, end = .95, direction = -1, option = 'viridis') +
+    geom_shadowtext(data = etiquetas.justeduc, 
+              aes(label = ifelse(porcentaje > 0.03 , scales::percent(porcentaje, accuracy = .1),"")),
+              position = position_stack(vjust = .5),
+              show.legend = FALSE,
+              size = 4,
+              color = rep('white'),
+              bg.colour='grey30')+
+  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022))
+
+alluvial_justeduc
+
+ggsave(alluvial_justeduc, file="output/graphs/alluvial_dep.png", height = 7, width = 8)
+```
+
+```{r bivariate-plots}
+merit_plot <- df_study1 %>% 
+  select(ahead_family, ahead_ambition, ahead_educ, ahead_hardwork, merit_talent, merit_effort) %>% 
+  sjPlot::plot_stackfrq(weight.by = df_study1$ponderador_long_total,
+                        expand.grid =TRUE,
+                      geom.colors = 'Spectral',
+                      show.total = FALSE, 
+                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),6),
+                      legend.labels = c("Strongly disagree", "Disagree", "Neither agree
+nor disagree", "Agree", "Strongly agree"),
+                      axis.labels = c("Ahead family", "Ahead ambition", "Ahead education", "Ahead hardwork", "Merit talent", "Merit effort")
+                      ) +
+  theme_light() +
+  theme(legend.position="bottom", text = element_text(size = 14))
+merit_plot
+
+ggsave(merit_plot, file="output/graphs/merit_plot.png", height = 7, width = 8)
+```
+
+```{r}
+colors<- c("#f1eef6ff","#bdc9e1ff","#b3b3b3ff","#74a9cfff","#0570b0ff")
+range <- c(.9, .8)
+
+p_effort<-
+  dplyr::select(df_study1,
+                merit_effort) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      title = c("Conviviality"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
                       reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE) +
-  ggplot2::theme(legend.position = "none")
-
-#Plot 2: Identity-------------------------------------------------------------
-p2<-
-  dplyr::select(df_study1,in_simili,in_amsimb,in_acpier,in_acadop) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      title = c("Identity"),
-                      geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "none")
-
-#Plot 3: Threat---------------------------------------------------------------
-p3<-
-  dplyr::select(df_study1,in_amreal) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
-                      geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "none")
-
-#Legend: Extract legend for the final plot using ggpubr
-leg<- ggpubr::as_ggplot(
-  ggpubr::get_legend(
-    dplyr::select(df_study1,in_amreal) %>%
-      sjPlot::plot_likert(geom.colors = "PuBu",
-                          catcount = 4,
-                          cat.neutral = 3) +
-      ggplot2::theme(legend.position = "bottom") +
-      ggplot2::scale_fill_manual(
-        values = c("#b3b3b3ff", "#f1eef6ff", "#bdc9e1ff","#74a9cfff","#0570b0ff"),
-        breaks=c("neutral",4,3,2,1),
-        labels=c(
-          "Neither comfortable nor uncomfortable\nSomewhat\nSomewhat similar\nNeither agree nor disagree\n",
-          "Very uncomfortable\nVery little or not at all\nNot similar\nStrongly disagree",
-          "Uncomfortable\nA little\nSlightly similar\nDisagree",
-          "Comfortable\nQuite a lot\nQuite similar\nAgree",
-          "Very comfortable\nA lot\nVery similar\nTotally agree")
-      )
-  )
-)
-
-# plot individuales-----------------------------------------------------------
-colors<- c("#b3b3b3ff","#f1eef6ff","#bdc9e1ff","#74a9cfff","#0570b0ff")
-p_ansied<-
-  dplyr::select(df_study1,in_ansied) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      title = c("Convivality"),
-                      geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE) + 
-  ggplot2::theme(legend.position = "bottom") +
+                      reverse.scale = T,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
-    labels = c(
-      "Neither comfortable nor uncomfortable",      
-      "Very uncomfortable",
-      "Uncomfortable",
-      "Comfortable",
-      "Very comfortable"
-    )
+    breaks = c(1,2,3,4,5),
   ) +
-  ggplot2::geom_label(label.padding = 1, label.r = 1,label.size = 0.1)
-
-p_ansied[["layers"]]
-
-p_simpat<-
-  dplyr::select(df_study1,in_simpat) %>%
+  theme(legend.position = "none")
+
+p_talent<-
+  dplyr::select(df_study1,
+                merit_talent) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "bottom") +
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = F,
+                      reverse.scale = T,
+                      show.n = FALSE, 
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
+    breaks = c(1,2,3,4,5), 
     labels = c(
-      "Somewhat ",      
-      "Very little or not at all",
-      "A little",
-      "Quite a lot",
-      "A lot"
-    )
-  )
-
-p_convi<-
-  dplyr::select(df_study1,in_valfam,in_valami,in_amigcl) %>%
+      "Strongly disagree",
+      "Disagree",
+      "Neither disagree nor agree",
+      "Agree",
+      "Strongly agree"
+    )
+  )
+
+p_hardwork<-
+  dplyr::select(df_study1,
+                ahead_hardwork) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "bottom")+
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = F,
+                      reverse.scale = T,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
-    labels = c(
-      "Neither disagree nor agree",      
-      "Strongly disagree",
-      "Disagree",
-      "Agree",
-      "Strongly agree"
-    )
-  )
-
-# Plot individual: IDENTITY -----------------------------------------------
-p_simili<-
-  dplyr::select(df_study1,in_simili) %>%
+    breaks = c(1,2,3,4,5),
+  ) +
+  theme(legend.position = "none")
+
+p_educ<-
+  dplyr::select(df_study1,
+                ahead_educ) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      title = c("Identity"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "bottom")+
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = F,
+                      reverse.scale = T,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
-    labels = c(
-      "Somewhat similar",      
-      "Not similiar",
-      "Slightly similar",
-      "Quite similar",
-      "Very similar"
-    )
-  )
-
-p_identi<-
-  dplyr::select(df_study1,in_amsimb,in_acpier,in_acadop,
-                in_amsimb) %>%
+    breaks = c(1,2,3,4,5),
+  ) +
+  theme(legend.position = "none")
+
+p_ambition<-
+  dplyr::select(df_study1,
+                ahead_ambition) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      # title = c("Threat"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "bottom")+
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = F,
+                      reverse.scale = T,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
-    labels = c(
-      "Neither disagree nor agree",      
-      "Strongly disagree",
-      "Disagree",
-      "Agree",
-      "Strongly agree"
-    )
-  )
-
-# Plot individual: THREAT -----------------------------------------------
-p_threat<-
-    dplyr::select(df_study1,in_amreal) %>%
+    breaks = c(1,2,3,4,5),
+  ) +
+  theme(legend.position = "none")
+
+p_family<-
+  dplyr::select(df_study1,
+                ahead_family) %>%
   sjPlot::plot_likert(geom.colors = "PuBu",
-                      title = c("Threat"),
                       geom.size = 0.8,
-                      # axis.labels = c("Effort", "Talent", "Rich Parents", "Contacts"),
-                      catcount = 4,
-                      cat.neutral = 3,
-                      grid.range  =  c (1.2 , 1.2),
-                      values  =  "sum.outside",
-                      reverse.colors = T,
-                      reverse.scale = F,
-                      show.n = FALSE)  +
-  ggplot2::theme(legend.position = "bottom") +
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = F,
+                      reverse.scale = T,
+                      show.n = FALSE, 
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
   ggplot2::scale_fill_manual(
     values = colors,
-    breaks = c("neutral", 4, 3, 2, 1),
+    breaks = c(1,2,3,4,5),   
     labels = c(
-      "Neither disagree nor agree",      
-      "Strongly disagree",
-      "Disagree",
-      "Agree",
-      "Strongly agree"
-    )
-  )
-
-# create plot-------------------------------------------------------------------
-if (!require("pacman")) install.packages("pacman")
+      "Not at all important",
+      "A little important",
+      "Somewhat 
+important",
+      "Pretty important",
+      "Very important"
+    )
+  )
+```
+##### Create plot
+```{r}
 pacman::p_load(cowplot)
-s <- 0.4
-m <- 0.6
-l <- 0.8
-#Final plot
+t <- 0.1
+s <- 0.45
+m <- 0.4
+xs <- 0.25
+
+caption <- 
+  stringr::str_wrap(
+    paste0("Note: Own elaboration based on ELSOC Survey ",
+           "(n=",dim(na.omit(df_study1))[1],"). "),
+    width = 100)
+
 title <- cowplot::ggdraw() + 
   cowplot::draw_label(
-    "Figure 1: Frequencies for Conviviality, Identity and Threat items",
+    "Frequencies for meritocracy",
     fontface = 'bold',
@@
   cowplot::add_sub(
-    cowplot::plot_grid(title,p_ansied,p_simpat,p_convi,p_simili,p_identi,p_threat, 
+    cowplot::plot_grid(
+                       p_effort,
+                       p_talent,
+                       p_hardwork,
+                       p_educ,
+                       p_ambition,
+                       p_family,
                        align = "hv",
                        axis = "a",
-                       nrow = 7, #4 plots
+                       nrow = 7, #6 plots
                        ncol = 1, #1 column   
-                       rel_heights = c(s,s,0.35,m,s,m,s) #fig size within plot
+                       rel_heights = c(
+                                       s,s,s,s,
+                                       s,
+                                       s) #fig size within plot
     ),
-    label = paste0("Source: Own elaboration based on ELSOC Survey",
-                   " (n=",dim(na.omit(df_study1))[1],")"
-    ),        # Caption: source and sample size
-    size = 12,  # Caption font size
-    hjust = -0.25# Caption align  
+    label = "", # Caption: source and sample size
+    size = 11,  # Caption font size
+    hjust = 0# Caption align 
   ) %>% 
   cowplot::ggdraw()
-# likert1
-ggplot2::ggsave(filename = "figure1.jpg",
-                plot = likert1,
-                path = here::here("output/images"),
-                device = "jpg",
-                units = "cm",
-                scale = 3.5,
-                width = 10,height = 10)
-
-#Final plot
-likert1<-   
-  cowplot::add_sub(
-    cowplot::plot_grid(p_a,p_b,p_c,p_d, #plot 1,2,3 and legend (bottom)
-                       align = "hv",
-                       axis = "a",
-                       nrow = 4, #4 plots
-                       ncol = 1, #1 column
-                       rel_heights = c(0.25,0.25,0.25,0.90) #fig size within plot
-    ),
-    label = paste0("Source: Own elaboration based on ELSOC Survey",
-                   " (n=",dim(na.omit(df_study1))[1],")"
-    ),        # Caption: source and sample size
-    size = 12,  # Caption font size
-    hjust = -0.25# Caption align
-  ) %>%
-  cowplot::ggdraw()  
+jpeg(here::here("output/graphs/figure1.jpg"),
+    # units = "cm",
+    # res = 100,
+    width = 800,
+    height = 800)
 likert1
-ggplot2::ggsave(filename = "figure1.jpg",
-                plot = likert1,
-                path = here::here("output/images"),
-                device = "jpg",
-                units = "cm",
-                scale = 3,
-                width = 10,height = 8)
+dev.off()
+```
+
+justification education vs year
+
+```{r}
+just_educ_year = df_study1 %>%
+  group_by(ola) %>%
+  summarise(education = mean(just_educ, na.rm = T))
+
+categorias_x <- c(2016, 2017, 2018, 2019, 2022)
+
+educ_year_plot <- just_educ_year_survey %>% ggplot(aes(ola, education)) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4)+
+  ylab("Justification of inequality") +
+  xlab("Wave") +
+  scale_x_discrete(labels = categorias_x) +
+  ylim(1,3)
+educ_year_plot
+
+ggsave(educ_year_plot, file="output/graphs/educ_year_plot.png", width = 7, height = 7)
+```
+
+Bivariate descriptive by year
+
+probar ponderadores longitudinales
+
+```{r}
+educ_effort_year = elsoc_diseno %>% filter(!is.na(merit_effort)) %>% 
+  group_by(ola, merit_effort) %>%
+  summarise(education = survey_mean(just_educ, vartype = "ci", na.rm = T)) %>% 
+  ungroup()
+
+educ_talent_year = elsoc_diseno %>% filter(!is.na(merit_talent)) %>% 
+  group_by(ola, merit_talent) %>%
+  summarise(education = survey_mean(just_educ, vartype = "ci", na.rm = T)) %>% 
+  ungroup()
+
+categorias_x <- c("Strongly disagree", "Disagree", "Neither agree 
+nor disagree", "Agree", "Strongly agree")
+categorias_f <- c(2016, 2017, 2018, 2019, 2022)
+
+effort_educ_year_plot <- educ_effort_year %>% 
+  ggplot(aes(factor(merit_effort), education, color = factor(ola), shape = factor(ola))) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4)+ 
+  ylab("Justification of inequality") +
+  xlab("Merit: effort") +
+  ylim(1, 3) +
+  scale_x_discrete(labels = categorias_x) +
+  scale_color_discrete(labels = categorias_f, name="Wave") +
+  scale_shape_discrete(labels = categorias_f, name="Wave")
+
+effort_educ_year_plot
+
+talent_educ_year_plot <- educ_talent_year %>% 
+  ggplot(aes(factor(merit_talent), education, color = factor(ola), shape = factor(ola))) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4) +
+  ylab("Justification of inequality") +
+  xlab("Merit: talent") +
+  ylim(1, 3) +
+  scale_x_discrete(labels = categorias_x) +
+  scale_color_discrete(labels = categorias_f, name="Wave") +
+  scale_shape_discrete(labels = categorias_f, name="Wave")
+
+talent_educ_year_plot
+
+ggsave(effort_educ_year_plot, file="output/graphs/effort_educ_year.png", height = 7, width = 7)
+ggsave(talent_educ_year_plot, file="output/graphs/talent_educ_year.png", height = 7, width = 7)
+
+educ_effort_year <- educ_effort_year %>% pivot_longer("merit_effort", 
+                              names_to="merit", 
+                              values_to="value")
+
+educ_talent_year <- educ_talent_year %>% pivot_longer("merit_talent", 
+                              names_to="merit", 
+                              values_to="value")
+
+educ_merit_year <- rbind(educ_effort_year, educ_talent_year)
+
+custom_labeller <- as_labeller(function(variable, value) {
+  return(categorias_f[value])
+})
+categorias_z <- c("Effort", "Talent")
+merit_educ_year <- educ_merit_year %>% 
+  ggplot(aes(factor(value), education)) +
+  theme_bw(base_size = 14) +
+  geom_point(aes(color=merit, shape=merit), size=3) +
+  ylab("Justification of inequality") +
+  xlab("Merit") +
+  ylim(1, 3) +
+  facet_wrap(~factor(ola), scales = "free", labeller = custom_labeller) +
+  scale_color_discrete(labels = categorias_z, name="Merit") +
+  scale_shape_discrete(labels = categorias_z, name="Merit")
+merit_educ_year
+
+ggsave(merit_educ_year, file="output/graphs/merit_educ_year.png", height = 7, width = 7)
 ```
@@
                summarytools,
-               lavaan,
-               psych
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+               lavaan
+               )
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
 
@@
 mat_full<- 
-df_study_long_t5_comunas %>%
+df_study_long_t6_comunas %>%
   # filter(ola==1) %>%
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>% 
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
          ) %>% 
   psych::mixedCor()
-
-# dim(mat_full)
 
@@
 # Set Row names of the matrix
-rownames(cor_full) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of migrant friends [PER/VEN]",
-                     "D. Friends value of migrant friends [PER/VEN]",
-                     "E. Migrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of migrants (reverse) ",
-                     "H. Migrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
+rownames(cor_full) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
 #set Column names of the matrix
-colnames(cor_full) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+colnames(cor_full) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
 #Plot the matrix using corrplot
+png("output/graphs/corr.png",width=600,height=300, pointsize = 10)
 corrplot::corrplot(cor_full,
@@
   bg = "white",
-  na.label = "-",
-  title = "\n Full sample")
-
-png(file=here::here("output/images/figure2.jpg"),width=900, height=700)
-corrplot::corrplot(cor_full,
-  title = "\nFigure 2: Polychoric correlation matrix for Conviviality, Identity and Threat items",
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  # title = "\n Full sample"
-  )
+  na.label = "-")
 dev.off()
 
+ggsave(cor, file="output/graphs/cor.png")
 # Wave 1_______________________________________________________________________
 mat_1<- 
-df_study_long_t5_comunas %>%
+df_study1_long_comunas %>%
   filter(ola==1) %>%
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
          ) %>% 
@@
 # Set Row names of the matrix
-rownames(cor_1) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
+rownames(cor_1) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
 #set Column names of the matrix
-colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
 #Plot the matrix using corrplot
@@
 mat_2<-   
-df_study_long_t5_comunas %>%
+df_study1_long_comunas %>%
   filter(ola==2) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
          ) %>% 
@@
 # Set Row names of the matrix
-rownames(cor_2) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
+rownames(cor_2) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
 #set Column names of the matrix
-colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
 #Plot the matrix using corrplot
@@
 mat_3 <- 
-df_study_long_t5_comunas %>%
+df_study1_long_comunas %>%
   filter(ola==3) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
          ) %>% 
@@
 # Set Row names of the matrix
-rownames(cor_3) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
+rownames(cor_3) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
 #set Column names of the matrix
-colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
 #Plot the matrix using corrplot
@@
 mat_4 <- 
-df_study_long_t5_comunas %>%
+df_study1_long_comunas %>%
   filter(ola==4) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
          ) %>% 
@@
 # Set Row names of the matrix
-rownames(cor_4) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
+rownames(cor_4) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
 #set Column names of the matrix
-colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
 #Plot the matrix using corrplot
@@
 
-# Wave 5_______________________________________________________________________
-mat_5 <- 
-df_study_long_t5_comunas %>%
-  filter(ola==5) %>% 
-    select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
-  na.omit() %>%   
-  select(in_ansied,in_simpat,in_valfam, in_valami,in_amigcl, # conviviality
-         in_simili,in_amsimb,in_acpier, in_acadop, # identity
-         in_amreal # threat
+# Wave 6_______________________________________________________________________
+mat_6 <- 
+df_study1_long_comunas %>%
+  filter(ola==6) %>% 
+  select(just_educ, # distributive justice
+         merit_effort, merit_talent, # meritocracy
+         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
          ) %>% 
@@
 
-cor_5 <- as.matrix(mat_5$rho)
-diag(cor_5) = NA #set diagonal values to NA
+cor_6 <- as.matrix(mat_6$rho)
+diag(cor_6) = NA #set diagonal values to NA
 # Set Row names of the matrix
-rownames(cor_5) <- c("A. Interaction anxiety with [PER/VEN]",
-                     "B. Sympathy for [PER/VEN] living in Chile",
-                     "C. Family value of immigrant friends [PER/VEN]",
-                     "D. Friends value of immigrant friends [PER/VEN]",
-                     "E. Immigrants [PER/VEN] have Chilean friends",
-                     "F. Similarity between Chileans and [PER/VEN]",
-                     "G. Loose of identity because of immigrants (reverse) ",
-                     "H. Immigrant keeping their customs",
-                     "I. Adoption of Chilean customs",
-                     "J. Unemployment increases")
+rownames(cor_6) <- c("A. Education distributive justice",
+                     "B. People are rewarded for their efforts",
+                     "C. People are rewarded for their intelligence",
+                     "D. Go ahead: wealthy family",
+                     "E. Go ahead: Having ambition",
+                     "F. Go ahead: Good level of education",
+                     "G. Go ahead: Hard work")
 #set Column names of the matrix
-colnames(cor_5) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)","(H)","(I)","(J)")
+colnames(cor_6) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
 #Plot the matrix using corrplot
-corrplot::corrplot(cor_5,
+corrplot::corrplot(cor_6,
   method = "color",
@@
   na.label = "-",
-  title = "\n Wave 5")
-```
-
-```{r items-alpha}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
+  title = "\n Wave 6")
+```
+
+# Models
+
+## A. Education distribution justice
+
+```{r distribution-educ}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
                             # cargar librerias
@@
                kableExtra,
-               sjPlot,
-               sjmisc,
                summarytools,
-               lavaan
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-# Descriptive table __________________________________________________________
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
+
+# generate analytical sample
 df_study1 <- 
-  df_study_long_t5_comunas %>% 
-  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,
-         educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm)
-
-df_study1 <- df_study1 %>% na.omit()
-
-alpha_convi <- 
-  df_study1 %>% 
-  select(in_ansied,in_simpat,in_valfam,in_valami,in_amigcl) 
-sjPlot::tab_corr(alpha_convi)
-sjPlot::tab_itemscale(alpha_convi)
-
-alpha_ident <- 
-  df_study1 %>% 
-  select(in_simili,
-         in_amsimb, #reverse item
-         in_acpier,in_acadop)
-
-# sjmisc::frq(alpha_ident)
-sjPlot::tab_corr(alpha_ident)
-sjPlot::tab_itemscale(alpha_ident)
-```
-
-# Models
-
-## Education categorical {-}
-```{r average-educ, results='asis',fig.dim=c(10,6)}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total, just_educ, merit_effort, merit_talent, ahead_family:educ, quintil1:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado,pagadolag22_16, total_mat_pagado, var_simce_comuna, prom_simce_comuna, univlag22_17) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1 <- "ola"
+h2 <- "ola+merit_effort+merit_talent"
+h3 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo"
+h4 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17"
+h5 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ"
+h6 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+univlag22_17"
+h7 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ+univlag22_17"
+h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna"
+h9 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+var_simce_comuna"
+h10 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna+var_simce_comuna"
+h11 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado"
+h12 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+pagadolag22_16"
+h13 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16"
+
+h13.1 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16+prom_simce_comuna+var_simce_comuna+prop_univ+univlag22_17+gini_ypch+ginilag22_17"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ9 <- lmer(formula(paste0("just_educ~",h9,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ10 <- lmer(formula(paste0("just_educ~",h10,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ11 <- lmer(formula(paste0("just_educ~",h11,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ12 <- lmer(formula(paste0("just_educ~",h12,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ13 <- lmer(formula(paste0("just_educ~",h13,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ13.1 <- lmer(formula(paste0("just_educ~",h13.1,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::knitreg(list(educ1, educ2, educ3, educ4, educ5, educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13,educ13.1),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+#                custom.coef.names = c("Intercepto",
+#                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+#                                      "ola 2018",
+#                                      "ola 2019",
+##                                      "ola 2022",
+#                                      "Merito: esfuerzo",
+#                                      "Merito: Inteligencia",
+##                                      "Goahead: Familia",
+#                                      "Goahead: educacion",
+#                                      "Goahead: ambicion",
+#                                      "Goahead: trabajo duro",
+#                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+#                                      "Ed. técnica",
+#                                      "Ed. universitaria",
+#                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+#                                      "Quintil 3",
+#                                      "Quintil 4",
+#                                      "Quintil 5",
+#                                      "Quintil NA",
+#                                      "Estatus social subjetivo",
+#                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+#                                      "Pos. política: Derecha",
+#                                      "Pos. política: No se identifica",
+#                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+#                                      "Edad 50-64",
+#                                      "Edad 65 o más",
+#                                      "Mujer <br> <i>(Ref. Hombre)</i>",
+#                                      "Gini comunal",
+#                                      "Cambio gini 2022-2017",
+#                                      "Proporción educ universitaria",
+#                                      "Proporción matrícula privada",
+#                                      "Total matrícula privada",
+#                                      "Varianza Simce comuna",
+#                                      "Promedio Simce comuna"
+
+                              )
+          #omit.coef = omit
+
+```
+
+## A2. Education distribution justice (export tables)
+
+```{r}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study_t6_indep_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna,just_educ, merit_effort, merit_talent, ahead_family:educ,quintil1:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17, loginc_mean, logincomelag22_17) %>%
+  na.omit() %>% 
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1 <- "ola"
+h2 <- "ola+merit_effort+merit_talent"
+h3 <- "ola+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
+h4 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
+h5 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo"
+
+h6 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17"
+h7 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ+univlag22_17"
+h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17"
+h9 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadlag22_16"
+h10 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+h11 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+h12 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+loginc_mean+logincomelag22_17"
+
+hfull <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+loginc_mean+logincomelag22_17"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ9 <- lmer(formula(paste0("just_educ~",h9,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ10 <- lmer(formula(paste0("just_educ~",h10,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ11 <- lmer(formula(paste0("just_educ~",h11,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ12 <- lmer(formula(paste0("just_educ~",h12,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ13 <- lmer(formula(paste0("just_educ~",hfull,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+# Table 1: individual
+
+#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
+texreg::htmlreg(list(educ1, educ2, educ3, educ4, educ5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                custom.coef.names = c("Intercepto",
+                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
+                                      "ola 2018",
+                                      "ola 2019",
+                                      "ola 2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: trabajo duro",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: Familia",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>"),
+                file = "output/tables/table1-individuales.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table1-individuales.html" ,file ="output/tables/table1-individuales.png")
+
+```
+
+```{r}
+# table 2: contextuales
+omit <- "(ola)|(merit_effort)|(merit_talent)|(ahead_hardwork)|(ahead_educ)|(ahead_ambition)|(ahead_family)|(educ)|(quintil1)|(ess)|(pos_id)|(edad)|(sexo)|(Intercept)"
+texreg::htmlreg(list(educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
+                omit.coef = omit,
+                custom.coef.names = c(
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción educ universitaria",
+                                      "Cambio prop. universitaria 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Cambio matrícula privada 2022-2016",
+                                      "Promedio Simce comuna",
+                                      "Varianza Simce comuna",
+                                      "Ingreso comunal",
+                                      "Cambio ingreso comunal 2022-2017"),
+                file = "output/tables/table2-contextual.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table2-contextual.html" ,file ="output/tables/table2-contextual.png")
+
+```
+
+## Interacciones
+
+```{r distribution-educ-context}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
                             # cargar librerias
@@
                texreg,
-               sjPlot,
                ggplot2
                )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-old <- theme_set(theme_bw());theme_set(old)
-theme_replace(legend.position="bottom")
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
 
@@
 df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,
-         in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna,just_educ, merit_effort, merit_talent, ahead_family:educ,quintil1:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17, loginc_mean, logincomelag22_17) %>%
   na.omit() %>% 
-  mutate(
-    # conviviality=(in_ansied+in_simpat+in_valfam+in_valami+in_amigcl)/5,
-         # identity= (in_simili+in_amsimb+in_acpier+in_acadop)/4,
-         ola_f = ola,
-         ola=as.numeric(ola),
-         primary = car::recode(educ,recodes = "'Primary'=1;else=0",as.factor = T),
-         -in_amreal
-         )
-
-# Hypothesis___________________________________________________________________
-m1.cat <- "ola_f+pos_id+edad+nation_t1+cuestion_mig"
-m2 <- "ola_f+educ+quintil1+ess+pos_id+edad+nation_t1+cuestion_mig"
-m3 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+nation_t1+cuestion_mig"
-m4 <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+inm_prop_cas20_w+dif_inm+pos_id+edad+nation_t1+cuestion_mig"
-
-# h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-# h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-m5 <- "educ*ola+quintil1*ola+ess+know_inm_t1t4*ola+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm*ola+nation_t1+cuestion_mig"
-
-h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h.ed.fri.time <- "ola*educ*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-
-hipotesis <- c(m1.cat,m2,m3,m4
-               # h.ed.know.time,
-               # h.ed.fri.time
-               )
-# Set the independent variables (IVs)
-ivs<- df_study1 %>% select(starts_with("in_"),"conviviality","identity") %>% names()
-
-# Create the object for each model
-models <- list()
-for (i in ivs) {
-models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
-}
-
-# Estimate the multilevel models
-fits <- list()
-for (i in ivs) {
-  for (z in models[[i]][1:length(hipotesis)]) {
-    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
-  }
-}
-fit1 <- fits
-# change model names within list
-names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
-
-# Tables_______________________________________________________________________
-omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-# reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
-# detail for subsections:
-# https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
-
-coef_names <- c("Wave 2017 (ref: Wave 2016)","Wave 2018", "Wave 2019","Wave 2020",
-                "Technical (ref: Universitary)","High school","Primary",
-                "Quintile 4 (ref: Quintile 5)", "Quintile 3","Quintile 2","Quintile 1",
-                "Subjective Social Status",
-                "Stable, know (ref: Stable, do not know)","Now know","No longer know",
-                "Stable, have friends (ref: Stable, do not have friends)","Now have friends","No longer have friends",
-                "Propotion of migrants",
-                "Change in the proportion of migrants")
-
-cat(paste0("\n## ","Conviviality (avg)","\n"))
-
-caption_conv <- "Multilevel linear regression models for Conviviality"
-htmlreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
-        omit.coef = omit,single.row = T,
-        caption = caption_conv,
-        caption.above = T,
-        custom.coef.names=coef_names,
-        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
-        file = here::here("output/tables/tab_conv.html")
-        # reorder.coef = reorderc
-        )
-webshot::webshot(url = here::here("output/tables/tab_conv.html"),
-                 file = here::here("output/tables/tab_conv.png"))
-
-plot_model(fit1$conviviality1,type = "pred",terms = " ola_f")
-
-cat(paste0("\n ","<br>","\n"))
-
-cat(paste0("\n## ","Identity (avg)","\n"))
-
-caption_id <- "Multilevel linear regression models for Identity"
-htmlreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
+  mutate(ola=as.factor(ola),
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
+h2b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*merit_effort"
+h3b <- "merit_effort+merit_talent+ola+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+merit_talent*ola"
+h4b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_hardwork"
+h5b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_educ"
+h6b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_ambition"
+h7b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_family"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
+
+# table 3: interacciones
+omit <- "(educ)|(quintil1)|(ess)|(pos_id)|(edad)|(sexo)|(Intercept)|(gini_ypch)|(ginilag22_17)|(prop_univ)|(univlag22_17)|(prop_pagado)|(pagadlag22_16)|(prom_simce_comuna_rescaled)|(var_simce_comuna_rescaled)"
+texreg::htmlreg(list(educ1, educ2,educ3,educ4,educ5,educ6,educ7),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05",
+                custom.coef.names = c("Intercepto",
+                                      "2017",
+                                      "2018",
+                                      "2019",
+                                      "2022",
+                                      "Merito: esfuerzo",
+                                      "Merito: Inteligencia",
+                                      "Goahead: trabajo duro",
+                                      "Goahead: educacion",
+                                      "Goahead: ambicion",
+                                      "Goahead: Familia",
+                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
+                                      "Ed. técnica",
+                                      "Ed. universitaria",
+                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
+                                      "Quintil 3",
+                                      "Quintil 4",
+                                      "Quintil 5",
+                                      "Quintil NA",
+                                      "Estatus social subjetivo",
+                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
+                                      "Pos. política: Derecha",
+                                      "Pos. política: No se identifica",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Mujer <br> <i>(Ref. Hombre)</i>",
+                                      "Gini comunal",
+                                      "Cambio gini 2022-2017",
+                                      "Proporción educ universitaria",
+                                      "Cambio prop. universitaria 2022-2017",
+                                      "Proporción matrícula privada",
+                                      "Cambio matrícula privada 2022-2016",
+                                      "Promedio Simce comuna",
+                                      "Varianza Simce comuna",
+                                      "2017*merit:esfuerzo",
+                                      "2018*merit:esfuerzo",
+                                      "2019*merit:esfuerzo",
+                                      "2022*merit:esfuerzo",
+                                      "2017*merit:talento",
+                                      "2018*merit:talento",
+                                      "2019*merit:talento",
+                                      "2022*merit:talento",
+                                      "2017*goahead:hardwork",
+                                      "2018*goahead:hardwork",
+                                      "2019*goahead:hardwork",
+                                      "2022*goahead:hardwork",
+                                      "2017*goahead:educ",
+                                      "2018*goahead:educ",
+                                      "2019*goahead:educ",
+                                      "2022*goahead:educ",
+                                      "2017*goahead:ambition",
+                                      "2018*goahead:ambition",
+                                      "2019*goahead:ambition",
+                                      "2022*goahead:ambition",
+                                      "2017*goahead:family",
+                                      "2018*goahead:family",
+                                      "2019*goahead:family",
+                                      "2022*goahead:family"),
+                file = "output/tables/table3-interac.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table3-interac.html" ,file ="output/tables/table3-interac.png")
+
+## graph interact
+#sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
+predict<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
+predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
+predict_data <- predict_data %>% as.data.frame()
+
+predict_data$value <- c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6")
+
+predict_data$value <- factor(predict_data$value,
+                   levels = c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6"),
+                   labels = c("wave 1", 
+                              "wave 2",
+                              "wave 3",
+                              "wave 4",
+                              "wave 6"))
+
+interact_effect <- predict_data %>% ggplot(aes(value, coef1)) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4)+
+  geom_errorbar(aes(ymin=lb, ymax=ub),
+                  width=0.2) +
+  ylab("Coef") +
+  xlab("")
+interact_effect
+
+ggsave(interact_effect, file="output/graphs/interact_effect.png", width = 7, height = 7)
+```
+
+## Ponderadores
+
+```{r}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
+                            # cargar librerias
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, ahead_family:ahead_hardwork,education=educ,quintil1:sexo) %>% 
+  na.omit() %>% 
+  mutate(ola = as.factor(ola),
+         ola_num = as.numeric(ola),
+         ola_2=as.numeric(ola)^2,
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h0 <- "des_perc"
+h1 <- "des_perc+merit_effort+merit_talent"
+h2 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
+h3 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo"
+h4 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola"
+h5 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola_num+ola_2"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+
+# Table 1: individual
+
+omit <- "(education)|(sexo)|(ess)"
+texreg::htmlreg(list(educ0, educ1, educ2, educ3, educ4, educ5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Models 4, 5, and 6 are controlled by educational level, gender and subjective social status with no significant effects.", 
                 omit.coef = omit,
-        caption = caption_id,
-        caption.above = T,
-        single.row = T,
-        custom.coef.names=coef_names,
-        # reorder.coef = reorderc
-        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
-        file = here::here("output/tables/tab_iden.html")        
-        )
-webshot::webshot(url = here::here("output/tables/tab_iden.html"),
-                 file = here::here("output/tables/tab_iden.png"))
-
-plot_model(fit1$identity1,type = "pred",terms = "ola_f")
-
-cat(paste0("\n## ","J. Unemployment increases (Threat)","\n"))
-
-caption_th <- "Multilevel linear regression models for Threat"
-htmlreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
-        omit.coef = omit,single.row = T,
-        caption = caption_th,
-        caption.above = T,
-        custom.coef.names=coef_names,
-        custom.model.names = c("Model 1","Model 2","Model 3","Model 4"),
-        file = here::here("output/tables/tab_thr.html")         
-        # reorder.coef = reorderc
-        ) 
-webshot::webshot(url = here::here("output/tables/tab_thr.html"),
-                 file = here::here("output/tables/tab_thr.png"))
-
-plot_model(fit1$in_amreal1,type = "pred",terms = " ola_f")
-
-cat(paste0("\n## ","Interactions","\n"))  #------------------------------------#
-
-hipotesis_int <- c(m5
-               # h.ed.know.time,
-               # h.ed.fri.time
-               )
-
-# Create the object for each model
-models_int <- list()
-for (i in ivs) {
-models_int[[i]] <- paste0(i,"~",hipotesis_int,"+(1|idencuesta)+(1|comuna)")  
-}
-
-# Estimate the multilevel models
-fit_int <- list()
-for (i in ivs) {
-  for (z in models_int[[i]][1:length(hipotesis_int)]) {
-    fit_int[[z]] <- lme4::lmer(formula = z, data = df_study1)
-  }
-}
-
-names(fit_int) <- paste0(rep(ivs,each=length(hipotesis_int)),1:length(hipotesis_int))
-
-omit_int <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)|(ess)|(inm_prop_cas20_w)"
-coef_int <- c("Technical (ref: Universitary) ","High school","Primary","Wave",
-              "Quintile 4 (ref: Quintile 5)", "Quintile 3","Quintile 2","Quintile 1",
-              # "Subjective Social Status",
-              "Stable, know (ref: Stable, do not know)","Now know","No longer know",
-              "Stable, have friends (ref: Stable, do not have friends)","Now have friends","No longer have friends",
-              # "Propotion of migrants",
-              "Change in the proportion of migrants",
-              #----------------------------------------------------------------#
-              "Technical x Wave","High school x Wave","Primary x Wave",
-              "Quintile 4 x Wave", "Quintile 3 x Wave","Quintile 2 x Wave","Quintile 1 x Wave",
-              "Stable, know  x Wave","Now know x Wave","No longer know x Wave",
-              "Stable, have friends x Wave","Now have friends x Wave","No longer have friends x Wave",
-              "Change in the proportion of migrants  x Wave"              
-              )
-
-caption_int <- "Multilevel linear regression models with interactions"
-htmlreg(l = fit_int[c("conviviality1","identity1","in_amreal1")],
-        omit.coef = omit_int,
-        single.row = T,
-        custom.coef.names=coef_int,
-        reorder.coef = c(4,1:3,5:29),
-        groups = list(
-          # "Education (ref: Universitary)"=2:4,
-                      # "Household Income Quintile (ref: Quintile 5)"=5:8,
-                      # "Know migrant (ref: Stable, do not know)" =9:11,
-                      # "Friendship with migrants (ref: Stable, do not have friends)" =12:14,
-                      "Education x Wave" =16:18,
-                      "Quintile x Wave" =19:22,
-                      "Know migrant x Wave" =23:25,
-                      "Migrant friend x Wave" =26:28,
-                      "Migrant Propulation x Wave" =29
-                      ),
-        caption = caption_int,
-        caption.above = T,
-        custom.model.names = c("Model 1\n(Conviviality)","Model 2\n(Identity)","Model 3\n(Threat)"),
-        file = here::here("output/tables/tab_int.html")
-        )
-webshot::webshot(url = here::here("output/tables/tab_int.html"),
-                 file = here::here("output/tables/tab_int.png"))
-```
-
-## Education dummy {-}
-
-```{r average-primary, eval=FALSE, fig.dim=c(10,6), include=FALSE, results='asis'}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
+                custom.coef.names = c("Intercept",
+                                      "Inequality perception",
+                                      "Merit: effort",
+                                      "Merit: Talent",
+                                      "Getahead: Hardwork",
+                                      "Getahead: Education",
+                                      "Getahead: Ambition",
+                                      "Getahead: Family",
+                                      "Quintile 2 <br> <i>(Ref. Quintile 1)</i>",
+                                      "Quintile 3",
+                                      "Quintile 4",
+                                      "Quintile 5",
+                                      "Quintile NA",
+                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
+                                      "Pol. pos: Right",
+                                      "Pol. pos: Does not identify",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Wave 2017 <br> <i>(Ref. Wave 2016)</i>",
+                                      "Wave 2018",
+                                      "Wave 2019",
+                                      "Wave 2022",
+                                      "Wave",
+                                      "Wave^2"),
+                file = "output/tables/table1-individuales-pond-v2.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table1-individuales-pond-v2.html" ,file ="output/tables/table1-individuales-pond-v2.png")
+
+```
+
+## Interacciones
+
+```{r distribution-educ-context}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
                             # cargar librerias
@@
                texreg,
-               sjPlot
-               )
-load(file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
-
-old <- theme_set(theme_bw());theme_set(old)
-theme_replace(legend.position="bottom")
+               ggplot2
+               )
+load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
 
@@
 df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,
-         in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, merit_effort, merit_talent, ahead_family:ahead_hardwork,education=educ,quintil1:sexo) %>% 
   na.omit() %>% 
-  mutate(
-         ola_f = ola,
-         ola=as.numeric(ola),
-         primary = car::recode(educ,recodes = "'Primary'=1;else=0",as.factor = T),
-         -in_amreal
-         )
-
-# Hypothesis___________________________________________________________________
-h1.cat <- "ola_f+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h1 <- "ola+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h2.primary <- "ola*primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+primary+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+primary+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+primary+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h.ed.know.time <- "ola*primary*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h.ed.fri.time <- "ola*primary*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-hipotesis <- c(h1.cat,h1,h2.primary,h2.incom,h3,h4.know,h4.friend,
-               h.ed.know.time,
-               h.ed.fri.time
-               )
-# Set the independent variables (IVs)
-ivs<- df_study1 %>% select("in_amreal","conviviality","identity") %>% names()
-
-# Create the object for each model
-models <- list()
-for (i in ivs) {
-models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
-}
-
-# Estimate the multilevel models
-fits <- list()
-for (i in ivs) {
-  for (z in models[[i]][1:length(hipotesis)]) {
-    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
-  }
-}
-fit1 <- fits
-# change model names within list
-names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
-
-# Tables_______________________________________________________________________
-omit <- 
-  paste0(
-    "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-    )
-reorderc <- c(1:4,19,5:18,20:43) # the depends on the number of coeficients (!)
-# detail for subsections:
-# https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
-
-cat(paste0("\n## ","Conviviality (avg)","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("conviviality", names(fit1),"\n")]],
+  mutate(ola = as.factor(ola),
+         ola_2=as.numeric(ola)^2,
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola"
+h2b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*merit_effort"
+h3b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*merit_talent"
+h4b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_hardwork"
+h5b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_educ"
+h6b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_ambition"
+h7b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_family"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+
+# table 3: interacciones
+omit <- "(edad)|(education)|(sexo)|(ess)|(quintil1)|(pos_id)"
+texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6,educ7),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
+All the models are controlled by educational level, age, gender, subjective social status, political position, and income quintile",
                 omit.coef = omit,
-        single.row = F,
-        reorder.coef = reorderc
-        )
-
-cat(paste0("\n## ","Identity (avg)","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("identity", names(fit1),"\n")]],
-        omit.coef = omit,
-        reorder.coef = reorderc
-        )
-
-cat(paste0("\n## ","Unemployment increases (Threat)","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_amreal", names(fit1),"\n")]],
-        omit.coef = omit,
-        reorder.coef = reorderc
-        ) 
-```
-
-```{r by-item, fig.dim=c(10,6), results='asis'}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
+                custom.coef.names = c("Intercept",
+                                      "Merit: effort",
+                                      "Merit: Talent",
+                                      "Getahead: Hardwork",
+                                      "Getahead: Education",
+                                      "Getahead: Ambition",
+                                      "Getahead: Family",
+                                      "Wave 2017 <br> <i>(Ref. Wave 2016)</i>",
+                                      "Wave 2018",
+                                      "Wave 2019",
+                                      "Wave 2022",
+                                      "2017*merit:effort",
+                                      "2018*merit:effort",
+                                      "2019*merit:effort",
+                                      "2022*merit:effort",
+                                      "2017*merit:talent",
+                                      "2018*merit:talent",
+                                      "2019*merit:talent",
+                                      "2022*merit:talent",
+                                      "2017*getahead:hardwork",
+                                      "2018*getahead:hardwork",
+                                      "2019*getahead:hardwork",
+                                      "2022*getahead:hardwork",
+                                      "2017*getahead:education",
+                                      "2018*getahead:education",
+                                      "2019*getahead:education",
+                                      "2022*getahead:education",
+                                      "2017*getahead:ambition",
+                                      "2018*getahead:ambition",
+                                      "2019*getahead:ambition",
+                                      "2022*getahead:ambition",
+                                      "2017*getahead:family",
+                                      "2018*getahead:family",
+                                      "2019*getahead:family",
+                                      "2022*getahead:family"),
+                file = "output/tables/table3-interac_pond.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table3-interac_pond.html" ,file ="output/tables/table3-interac-pond.png")
+
+## graph interact
+#sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
+predict<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
+predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
+predict_data <- predict_data %>% as.data.frame()
+
+predict_data$value <- c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6")
+
+predict_data$value <- factor(predict_data$value,
+                   levels = c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6"),
+                   labels = c("wave 1", 
+                              "wave 2",
+                              "wave 3",
+                              "wave 4",
+                              "wave 6"))
+
+interact_effect <- predict_data %>% ggplot(aes(value, coef1)) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4)+
+  geom_errorbar(aes(ymin=lb, ymax=ub),
+                  width=0.2) +
+  ylab("Coef") +
+  xlab("")
+interact_effect
+
+ggsave(interact_effect, file="output/graphs/interact_effect-pond.png", width = 7, height = 7)
+```
+
+```{r distribution-educ-context}
+#rm(list=ls())
+#if (!require("pacman")) install.packages("pacman") # instalar pacman
                             # cargar librerias
@@
 df_study1 <- 
-  df_study_long_t5_comunas %>%
-  select(idencuesta,ola,comuna,cuestion_mig,
-         in_ansied,in_simpat,in_valfam,in_valami,in_amigcl,
-         in_simili,in_amsimb,in_acpier,in_acadop,
-         conviviality, identity,
-         in_amreal,educ,quintil1:edad,
-         inm_prop_cas17_w,inm_prop_cas20_w,dif_inm) %>% 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, merit_effort, merit_talent, ahead_family:ahead_hardwork,education=educ,quintil1:sexo) %>% 
   na.omit() %>% 
-  mutate(
-         ola_f = ola,
-         ola=as.numeric(ola),
-         primary = car::recode(educ,recodes = "'Primary'=1;else=0",as.factor = T),
-         -in_amreal
-         )
-
-# Hypothesis___________________________________________________________________
-h1.cat <- "ola_f+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h1 <- "ola+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h2.educ <- "ola*educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h2.incom <- "ola+educ+quintil1*ola+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h3 <- "ola*dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h4.know <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4*ola+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h4.friend <- "ola+dif_inm+educ+quintil1+ess+know_inm_t1t4+frien_inm_t1t4*ola+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h.ed.know.time <- "ola*educ*know_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-h.ed.fri.time <- "ola*educ*frien_inm_t1t4+quintil1+ess+know_inm_t1t4+frien_inm_t1t4+pos_id+edad+inm_prop_cas20_w+dif_inm+nation_t1+cuestion_mig"
-hipotesis <- c(h1.cat,h1,h2.educ,h2.incom,h3,h4.know,h4.friend
-               # h.ed.know.time,
-               # h.ed.fri.time
-               )
-# Set the independent variables (IVs)
-ivs<- df_study1 %>% select(
-  starts_with("in_")
-  # "conviviality",
-  # "identity"
-  ) %>% names()
-
-# Create the object for each model
-models <- list()
-for (i in ivs) {
-models[[i]] <- paste0(i,"~",hipotesis,"+(1|idencuesta)+(1|comuna)")  
-}
-
-# Estimate the multilevel models
-fits <- list()
-for (i in ivs) {
-  for (z in models[[i]][1:length(hipotesis)]) {
-    fits[[z]] <- lme4::lmer(formula = z, data = df_study1)
-  }
-}
-fit1 <- fits
-# change model names within list
-names(fit1) <- paste0(rep(ivs,each=length(hipotesis)),1:length(hipotesis))
-
-# Tables_______________________________________________________________________
-omit <- 
-  paste0(
-    "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-    )
-reorderc <- c(1:4,21,5:20,22:35) # the depends on the number of coeficients (!)
-# detail for subsections:
-# https://stackoverflow.com/questions/69338331/r-markdown-create-section-headers-in-r-and-print-r-output-in-same-section
-
-cat(paste0("\n# ","Annexes","\n"))
-
-cat(paste0("\n## ","A. Anxiety","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_ansied", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc
-        )
-cat(paste0("\n## ","B. Simpathy ","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_simpat", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","C. Family value migrant friends","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_valfam", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","D. Friend value migrant friends","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_valami", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","E. Immigrants have Chilean friends","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_amigcl", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-#______________________________________________________________________________
-
-cat(paste0("\n## ","F. Similarity","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_simili", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","G. Loose identity","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_amsimb", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","H. Immigrants keep customs","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_acpier", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-
-cat(paste0("\n## ","I. Adopt chilean customs","\n"))
-
-knitreg(l = fit1[names(fit1)[grepl("in_acadop", names(fit1))]],
-        omit.coef = omit,reorder.coef = reorderc)
-```
+  mutate(ola = as.numeric(ola),
+         ola_2=as.numeric(ola)^2,
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h1b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2"
+h2b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*merit_effort"
+h3b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*merit_talent"
+h4b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_hardwork"
+h5b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_educ"
+h6b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_ambition"
+h7b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_family"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+
+# table 3: interacciones
+omit <- "(edad)|(education)|(sexo)|(ess)|(quintil1)|(pos_id)"
+texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6,educ7),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
+All the models are controlled by educational level, age, gender, subjective social status, political position, and income quintile",
+                omit.coef = omit,
+                custom.coef.names = c("Intercept",
+                                      "Merit: effort",
+                                      "Merit: Talent",
+                                      "Getahead: Hardwork",
+                                      "Getahead: Education",
+                                      "Getahead: Ambition",
+                                      "Getahead: Family",
+                                      "Wave",
+                                      "Wave^2",
+                                      "wave*merit:effort",
+                                      "wave*merit:talent",
+                                      "Wave*getahead:hardwork",
+                                      "Wave*getahead:education",
+                                      "Wave*getahead:ambition",
+                                      "Wave*getahead:family"),
+                file = "output/tables/table3-interac_pond-cuadratic.html")
+          #omit.coef = omit
+
+webshot2::webshot(url ="output/tables/table3-interac_pond-cuadratic.html" ,file ="output/tables/table3-interac-pond-cuadratic.png")
+
+## graph interact
+#sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
+predict<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
+predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
+predict_data <- predict_data %>% as.data.frame()
+
+predict_data$value <- c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6")
+
+predict_data$value <- factor(predict_data$value,
+                   levels = c("ola1", 
+                             "ola2",
+                             "ola3", 
+                             "ola4",
+                             "ola6"),
+                   labels = c("wave 1", 
+                              "wave 2",
+                              "wave 3",
+                              "wave 4",
+                              "wave 6"))
+
+interact_effect <- predict_data %>% ggplot(aes(value, coef1)) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4)+
+  geom_errorbar(aes(ymin=lb, ymax=ub),
+                  width=0.2) +
+  ylab("Coef") +
+  xlab("")
+interact_effect
+
+ggsave(interact_effect, file="output/graphs/interact_effect-pond.png", width = 7, height = 7)
+```

### 2024-03-13 production/prod_prep-mergedata.Rmd
modificaciones gráficos + add percepcion desigualdad

@@
 load(file = here::here("input/data-proc/df_study2_long.RData"))
-load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
+load(file = here::here("input/data-proc/df_study1_long_t6.RData"))
 load(file = here::here("input/data-proc/study_1_comunas.RData"))
@@
 
-df_study_long_t5_comunas <- 
- dplyr::left_join(x = df_study1_long_t5,
+df_study_long_t6_comunas <- 
+ dplyr::left_join(x = df_study1_long_t6,
                   y = df_study1_comunas, 
@@
  dplyr::rename(comuna=comuna.x)
+
+df_study_long_t6_comunas <- df_study_long_t6_comunas %>% mutate(year= case_when(ola==1~2017,
+                                                                            ola==2~2017,
+                                                                            ola==3~2017,
+                                                                            ola==4~2022,
+                                                                            ola==6~2022))
+
+df_study_long_t6_comunas <- 
+ dplyr::left_join(x = df_study_long_t6_comunas,
+                  y = select(df_gini_comunas_long, comunacod=COMUNA_COD,year, starts_with("gini"), prop_univ, starts_with("inc"), starts_with("log")), 
+                  by = c("comunacod", "year"))
+
+df_study_long_t6_comunas <- 
+ dplyr::left_join(x = df_study_long_t6_comunas,
+                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini"), starts_with("prop_univ"), starts_with("univlag"), starts_with("inc"), starts_with("log")), 
+                  by = c("comunacod"))
 
@@
 df_study2_long_comunas <- df_study2_long_comunas %>% mutate(year= case_when(ola==1~2016,
+                                                                            ola==2~2017,
+                                                                            ola==3~2018,
+                                                                            ola==4~2019,
+                                                                            ola==5~2021,
+                                                                            ola==6~2022))
+
+df_study_long_t6_comunas <- df_study_long_t6_comunas %>% mutate(year= case_when(ola==1~2016,
                                                                             ola==2~2017,
@@
 
+df_study_long_t6_comunas <- 
+ dplyr::left_join(x = df_study_long_t6_comunas,
+                  y = select(df_matricula_comunas_long, comunacod=cod_com_rbd_15,year, total_mat_pagado,prop_pagado), 
+                  by = c("comunacod", "year"))
+
+df_study_long_t6_comunas <- 
+ dplyr::left_join(x = df_study_long_t6_comunas,
+                  y = select(df_matricula_comunas_wide, comunacod=cod_com_rbd_16,starts_with("total_mat_pagado"), starts_with("prop_pagado"), pagadolag22_16), 
+                  by = c("comunacod"))
+
 ## Juntar promedio simce y sd por comuna
@@
 
+df_study_long_t6_comunas <- 
+ dplyr::left_join(x = df_study_long_t6_comunas,
+                  y = select(df_simce_comunas_long, comunacod=cod_com_15,year, prom_simce_comuna,var_simce_comuna), 
+                  by = c("comunacod", "year"))
+
+df_study_long_t6_comunas <- 
+ dplyr::left_join(x = df_study_long_t6_comunas,
+                  y = select(df_comuna_rbd_wide, comunacod=COMUNA,starts_with("prom_simce"), starts_with("var_simce")), 
+                  by = c("comunacod"))
+
 ## Rescale simce
@@
 df_study1_long_comunas$var_simce_comuna_rescaled <- df_study1_long_comunas$var_simce_comuna / 100
+df_study_long_t6_comunas$prom_simce_comuna_rescaled <- df_study_long_t6_comunas$prom_simce_comuna / 10
+df_study_long_t6_comunas$var_simce_comuna_rescaled <- df_study_long_t6_comunas$var_simce_comuna / 100
 # Verificar el rescalado
@@
      file = here::here("input/data-proc/df_study2_long_comunas.RData"))
-save(df_study_long_t5_comunas,
-     file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+save(df_study_long_t6_comunas,
+     file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
 
@@
 
-sjPlot::view_df(df_study_long_t5_comunas,
+sjPlot::view_df(df_study_long_t6_comunas,
         show.frq = T, 

### 2024-03-13 production/prod_prep.Rmd
modificaciones gráficos + add percepcion desigualdad

@@
          muestra,
+         ponderador_long_total, segmento, estrato,
          sexo,
@@
                      # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
-                     values_from = names(select(df_study1,comuna,comunacod,just_educ:edad,sexo))
+                     values_from = names(select(df_study1,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:edad,sexo))
                      )
@@
               variable.name = c("ola"),
-              measure = patterns("^comuna_","^comunacod_",
+              measure = patterns("^comuna_","^comunacod_","^ponderador_long_total", "^segmento", "^estrato",
                                  "^just_educ_", "^just_pension_", "^just_salud_", "^des_perc_",
@@
                                  ),
-              value.name = c("comuna","comunacod",
+              value.name = c("comuna","comunacod","ponderador_long_total","segmento","estrato",
                              "just_educ", "just_pension","just_salud", "des_perc",
@@
 load(here::here("input/data-original/casen2022.Rdata"))
-load(here::here("input/data-original/censo17.Rdata"))
+#load(here::here("input/data-original/censo17.Rdata"))
 load(here::here("input/data-original/elsoc_long.Rdata"))
@@
   select(-idencuesta)
+
+df_cit_a <- df_cit_a %>%
+  mutate(comuna = ifelse(test = comuna=="Ñunoa",
+                         yes = "Nhunhoa",
+                         no =  comuna))
+
+df_cit_a <- df_cit_a %>%
+  mutate(comuna = ifelse(test = comuna=="Vicuna",
+                         yes = "Vicunha",
+                         no =  comuna))
+
+df_cit_a <- df_cit_a %>%
+  mutate(comuna = ifelse(test = comuna=="Vina del mar",
+                         yes = "Vinha del mar",
+                         no =  comuna))
+
+df_cit_a <- df_cit_a %>%
+  mutate(comuna = ifelse(test = comuna=="Donihue",
+                         yes = "Donhihue",
+                         no =  comuna))
+
+df_cit_a <- df_cit_a %>%
+  mutate(comuna = ifelse(test = comuna=="Penaflor",
+                         yes = "Penhaflor",
+                         no =  comuna))
+df_cit_a <- df_cit_a %>%
+  mutate(comuna = ifelse(test = comuna=="Penalolen",
+                         yes = "Penhalolen",
+                         no =  comuna))
+
 #check for duplicates

### 2024-03-13 production/prod_analysis-fix-t5.Rmd
reorder models + graphs

@@
 df_study1 <- df_study_long_t6_comunas %>% dplyr::select(ola,
-                                               just_educ, merit_effort, merit_talent,
-                                               ahead_family:educ, quintil1:sexo)
+                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               educ, quintil1:sexo)
 
@@
 
-for (i in c("just_educ",
+for (i in c("just_educ", "des_perc",
             "merit_effort","merit_talent")) {
@@
 
-for (i in c("ahead_family","ahead_educ",
-            "ahead_ambition", "ahead_hardwork")) {
-  df_study1[[i]] <- factor(df_study1[[i]],
-                            levels = 1:5,
-                           labels = c("Not at all important",
-                                  "A little important",
-                                  "Somewhat important",
-                                  "Pretty important",
-                                  "Very important"))
-}
-
 df_study1$just_educ <- 
 sjlabelled::set_label(x = df_study1$just_educ, 
-                      label = "Education distributive justice")
+                      label = "Justification of educational inequality")
+
+#
+df_study1$des_perc <- 
+sjlabelled::set_label(x = df_study1$des_perc, 
+                      label = "Income differences are too large")
 
@@
 sjlabelled::set_label(x = df_study1$merit_talent, 
-                      label = "People are rewarded for their intelligence")
-
-#
-df_study1$ahead_family <- 
-sjlabelled::set_label(x = df_study1$ahead_family, 
-                      label = "Coming from a wealthy family")
-
-#
-df_study1$ahead_educ <- 
-sjlabelled::set_label(x = df_study1$ahead_educ, 
-                      label = "Having a good level of education")
-
-#
-df_study1$ahead_ambition <- 
-sjlabelled::set_label(x = df_study1$ahead_ambition, 
-                      label = "Having ambition")
-
-#
-df_study1$ahead_hardwork <- 
-sjlabelled::set_label(x = df_study1$ahead_hardwork, 
-                      label = "Hard work")
+                      label = "People are rewarded for their intelligence and skills")
 
@@
 
-df<- dfSummary(x = dplyr::select(df_study1, ola, just_educ, merit_effort, merit_talent, ahead_hardwork, ahead_educ, ahead_ambition, ahead_family, educ, quintil1, ess, pos_id, edad, sexo),
+df<- dfSummary(x = dplyr::select(df_study1, ola, just_educ, des_perc, merit_effort, merit_talent, educ, quintil1, ess, pos_id, edad, sexo),
                plain.ascii = FALSE,
@@
 df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, merit_effort, merit_talent,
-                                               ahead_family:educ, quintil1:sexo)
+                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               educ, quintil1:sexo)
 df_study1 <- df_study1[complete.cases(df_study1$segmento), ]
@@
 ```{r bivariate-plots}
+df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               educ, quintil1:sexo)
 merit_plot <- df_study1 %>% 
-  select(ahead_family, ahead_ambition, ahead_educ, ahead_hardwork, merit_talent, merit_effort) %>% 
+  select(merit_talent, merit_effort, des_perc, just_educ) %>% 
   sjPlot::plot_stackfrq(weight.by = df_study1$ponderador_long_total,
                         expand.grid =TRUE,
-                      geom.colors = 'Spectral',
+                      geom.colors = "PuBu",
                       show.total = FALSE, 
-                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),6),
+                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),4),
                       legend.labels = c("Strongly disagree", "Disagree", "Neither agree
 nor disagree", "Agree", "Strongly agree"),
-                      axis.labels = c("Ahead family", "Ahead ambition", "Ahead education", "Ahead hardwork", "Merit talent", "Merit effort")
+                      axis.labels = c("Merit talent", "Merit effort", "Inequality perception", "Justification of educational inequality")
                       ) +
@@
 colors<- c("#f1eef6ff","#bdc9e1ff","#b3b3b3ff","#74a9cfff","#0570b0ff")
-range <- c(.9, .8)
+range <- c(1, 1)
+
+p_just_educ<-
+  dplyr::select(df_study1,
+                just_educ) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      geom.size = 0.8,
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = F,
+                      reverse.scale = T,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c(1,2,3,4,5),
+  ) +
+  theme(legend.position = "none")
+
+p_des_perc<-
+  dplyr::select(df_study1,
+                des_perc) %>%
+  sjPlot::plot_likert(geom.colors = "PuBu",
+                      geom.size = 0.8,
+                      catcount = 5,
+                      grid.range  =  range,
+                      show.prc.sign = T,
+                      reverse.colors = F,
+                      reverse.scale = T,
+                      show.n = FALSE,
+                      intercept.line.color = "white")  +
+  ggplot2::theme(legend.position = "bottom",
+    text = ggplot2::element_text(size = 16))+
+  ggplot2::scale_fill_manual(
+    values = colors,
+    breaks = c(1,2,3,4,5),
+  ) +
+  theme(legend.position = "none")
 
@@
   )
-
-p_hardwork<-
-  dplyr::select(df_study1,
-                ahead_hardwork) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_educ<-
-  dplyr::select(df_study1,
-                ahead_educ) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_ambition<-
-  dplyr::select(df_study1,
-                ahead_ambition) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_family<-
-  dplyr::select(df_study1,
-                ahead_family) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE, 
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),   
-    labels = c(
-      "Not at all important",
-      "A little important",
-      "Somewhat 
-important",
-      "Pretty important",
-      "Very important"
-    )
-  )
-```
+```
+
 ##### Create plot
@@
   cowplot::draw_label(
-    "Frequencies for meritocracy",
+    "Frequencies",
     fontface = 'bold',
@@
   cowplot::add_sub(
-    cowplot::plot_grid(
+    cowplot::plot_grid(p_just_educ,
+                       p_des_perc,
                        p_effort,
                        p_talent,
-                       p_hardwork,
-                       p_educ,
-                       p_ambition,
-                       p_family,
                        align = "hv",
@@
   select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+         des_perc,
+         merit_effort, merit_talent # meritocracy
          ) %>% 
@@
 # Set Row names of the matrix
-rownames(cor_full) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
+rownames(cor_full) <- c("A. Justification of educational inequality",
+                        "B. Inequality perception",
+                     "C. People are rewarded for their efforts",
+                     "D. People are rewarded for their intelligence")
 #set Column names of the matrix
-colnames(cor_full) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+colnames(cor_full) <-c("(A)", "(B)","(C)","(D)")
 #Plot the matrix using corrplot
@@
 dev.off()
-
-ggsave(cor, file="output/graphs/cor.png")
 # Wave 1_______________________________________________________________________
@@
 
-## A. Education distribution justice
+## A. Justification of educational inequality
 
 ```{r distribution-educ}
-#rm(list=ls())
+rm(list=ls())
 #if (!require("pacman")) install.packages("pacman") # instalar pacman
@@
   df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total, just_educ, merit_effort, merit_talent, ahead_family:educ, quintil1:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado,pagadolag22_16, total_mat_pagado, var_simce_comuna, prom_simce_comuna, univlag22_17) %>%
+  select(idencuesta,ola,comuna, ponderador_long_total, just_educ, des_perc, merit_effort, merit_talent, educ, quintil1:sexo7) %>% 
   na.omit() %>% 
@@
   df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, ahead_family:ahead_hardwork,education=educ,quintil1:sexo) %>% 
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
   na.omit() %>% 
@@
 # Hipotesis
-h0 <- "des_perc"
-h1 <- "des_perc+merit_effort+merit_talent"
-h2 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
-h3 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo"
-h4 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola"
-h5 <- "des_perc+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola_num+ola_2"
+h0 <- "ola"
+h1 <- "ola_num+ola_2"
+h2 <- "ola_num+ola_2+des_perc"
+h3 <- "ola_num+ola_2+des_perc+merit_talent"
+h4 <- "ola_num+ola_2+des_perc+merit_talent+merit_effort"
+h5 <- "ola_num+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo"
 
 # A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
@@
 
-omit <- "(education)|(sexo)|(ess)"
+omit <- "(education)|(ess)"
 texreg::htmlreg(list(educ0, educ1, educ2, educ3, educ4, educ5),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Models 4, 5, and 6 are controlled by educational level, gender and subjective social status with no significant effects.", 
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Model 6 is controlled by educational level, and subjective social status with no significant effects.", 
                 omit.coef = omit,
                 custom.coef.names = c("Intercept",
+                                      "Wave 2017 <br> <i>(Ref. Wave 2016)</i>",
+                                      "Wave 2018",
+                                      "Wave 2019",
+                                      "Wave 2022",
+                                      "Wave",
+                                      "Wave^2",
                                       "Inequality perception",
+                                      "Merit: Talent",
                                       "Merit: effort",
-                                      "Merit: Talent",
-                                      "Getahead: Hardwork",
-                                      "Getahead: Education",
-                                      "Getahead: Ambition",
-                                      "Getahead: Family",
                                       "Quintile 2 <br> <i>(Ref. Quintile 1)</i>",
@@
                                       "Edad 65 o más",
-                                      "Wave 2017 <br> <i>(Ref. Wave 2016)</i>",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2"),
+                                      "Sexo <br> <i>(Ref. Male)</i>"),
                 file = "output/tables/table1-individuales-pond-v2.html")
@@
                )
-load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
 
@@
   df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, merit_effort, merit_talent, ahead_family:ahead_hardwork,education=educ,quintil1:sexo) %>% 
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
   na.omit() %>% 
@@
 # Hipotesis
-h1b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola"
-h2b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*merit_effort"
-h3b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*merit_talent"
-h4b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_hardwork"
-h5b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_educ"
-h6b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_ambition"
-h7b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola*ahead_family"
+h1b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo"
+h2b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*des_perc"
+h3b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_talent"
+h4b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_effort"
+h5b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*pos_id"
+h6b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*edad"
 
 # A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
 # table 3: interacciones
-omit <- "(edad)|(education)|(sexo)|(ess)|(quintil1)|(pos_id)"
-texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6,educ7),
+omit <- "(education)|(ess)|(quintil1)|(sexo)"
+texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                 custom.coef.names = c("Intercept",
-                                      "Merit: effort",
-                                      "Merit: Talent",
-                                      "Getahead: Hardwork",
-                                      "Getahead: Education",
-                                      "Getahead: Ambition",
-                                      "Getahead: Family",
                                       "Wave 2017 <br> <i>(Ref. Wave 2016)</i>",
@@
                                       "Wave 2022",
+                                      "Inequality perception",
+                                      "Merit: Talent",
+                                      "Merit: effort",
+                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
+                                      "Pol. pos: Right",
+                                      "Pol. pos: Does not identify",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "2017*inequality perception",
+                                      "2018*inequality perception",
+                                      "2019*inequality perception",
+                                      "2022*inequality perception",
+                                      "2017*merit:talent",
+                                      "2018*merit:talent",
+                                      "2019*merit:talent",
+                                      "2022*merit:talent",
                                       "2017*merit:effort",
@@
                                       "2022*merit:effort",
-                                      "2017*merit:talent",
-                                      "2018*merit:talent",
-                                      "2019*merit:talent",
-                                      "2022*merit:talent",
-                                      "2017*getahead:hardwork",
-                                      "2018*getahead:hardwork",
-                                      "2019*getahead:hardwork",
-                                      "2022*getahead:hardwork",
-                                      "2017*getahead:education",
-                                      "2018*getahead:education",
-                                      "2019*getahead:education",
-                                      "2022*getahead:education",
-                                      "2017*getahead:ambition",
-                                      "2018*getahead:ambition",
-                                      "2019*getahead:ambition",
-                                      "2022*getahead:ambition",
-                                      "2017*getahead:family",
-                                      "2018*getahead:family",
-                                      "2019*getahead:family",
-                                      "2022*getahead:family"),
+                                      "2017*Center",
+                                      "2018*Center",
+                                      "2019*Center",
+                                      "2022*Center",
+                                      "2017*Right",
+                                      "2018*Right",
+                                      "2019*Right",
+                                      "2022*Right",
+                                      "2017*Not identify",
+                                      "2018*Not identify",
+                                      "2019*Not identify",
+                                      "2022*Not identify",
+                                      "2017*age30-49 ",
+                                      "2018*age30-49 ",
+                                      "2019*age30-49 ",
+                                      "2022*age30-49 ",
+                                      "2017*age50-64 ",
+                                      "2018*age50-64 ",
+                                      "2019*age50-64 ",
+                                      "2022*age50-64 ",
+                                      "2017*age65 or more",
+                                      "2018*age65 or more",
+                                      "2019*age65 or more",
+                                      "2022*age65 or more"
+                                      ),
                 file = "output/tables/table3-interac_pond.html")
@@
 # Hipotesis
-h1b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2"
-h2b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*merit_effort"
-h3b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*merit_talent"
-h4b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_hardwork"
-h5b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_educ"
-h6b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_ambition"
-h7b <- "merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+education+quintil1+ess+pos_id+edad+sexo+ola+ola_2+ola*ahead_family"
+h1b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo"
+h2b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*des_perc"
+h3b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_talent"
+h4b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_effort"
+h5b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*pos_id"
+h6b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*edad"
 
 # A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
 # table 3: interacciones
-omit <- "(edad)|(education)|(sexo)|(ess)|(quintil1)|(pos_id)"
-texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6,educ7),
+omit <- "(education)|(ess)|(quintil1)|(sexo)"
+texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                 custom.coef.names = c("Intercept",
-                                      "Merit: effort",
-                                      "Merit: Talent",
-                                      "Getahead: Hardwork",
-                                      "Getahead: Education",
-                                      "Getahead: Ambition",
-                                      "Getahead: Family",
                                       "Wave",
                                       "Wave^2",
-                                      "wave*merit:effort",
-                                      "wave*merit:talent",
-                                      "Wave*getahead:hardwork",
-                                      "Wave*getahead:education",
-                                      "Wave*getahead:ambition",
-                                      "Wave*getahead:family"),
-                file = "output/tables/table3-interac_pond-cuadratic.html")
+                                      "Inequality perception",
+                                      "Merit: Talent",
+                                      "Merit: effort",
+                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
+                                      "Pol. pos: Right",
+                                      "Pol. pos: Does not identify",
+                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Edad 50-64",
+                                      "Edad 65 o más",
+                                      "Wave*inequality perception",
+                                      "Wave*merit:talent",
+                                      "Wave*merit:effort",
+                                      "Wave*Center",
+                                      "Wave*Right",
+                                      "Wave*Not identify",
+                                      "Wave*age30-49 ",
+                                      "Wave*age50-64 ",
+                                      "Wave*age65 or more"
+                                      ),
+                file = "output/tables/table3-interac_pond-numeric.html")
           #omit.coef = omit
 
-webshot2::webshot(url ="output/tables/table3-interac_pond-cuadratic.html" ,file ="output/tables/table3-interac-pond-cuadratic.png")
+webshot2::webshot(url ="output/tables/table3-interac_pond-numeric.html" ,file ="output/tables/table3-interac-pond-numeric.png")
 

### 2024-03-19 production/prod_analysis-fix-t5.Rmd
update modelos aleatorios
add modelos aleatorios tiempo y var indep interacciones
Change alluvial color
Change variables by year plot

@@
                survey,
-               shadowtext
+               shadowtext,
+               srvyr
                )
@@
 df_study1 <- df_study_long_t6_comunas %>% dplyr::select(ola,
-                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               just_educ, des_perc, merit_talent, merit_effort,
                                                educ, quintil1:sexo)
@@
 
-df<- dfSummary(x = dplyr::select(df_study1, ola, just_educ, des_perc, merit_effort, merit_talent, educ, quintil1, ess, pos_id, edad, sexo),
+df<- dfSummary(x = dplyr::select(df_study1, ola, just_educ, des_perc, merit_talent, merit_effort, educ, quintil1, ess, pos_id, edad, sexo),
                plain.ascii = FALSE,
@@
 ```{r}
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
 df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
@@
 df_study1$just_educ <- factor(df_study1$just_educ,
-         labels = c("Strongly disagree", "Disagree", "Neither agree
-nor disagree", "Agree", "Strongly agree"))
+         labels = c("Strongly 
+disagree", "Disagree", "Neither agree
+nor disagree", "Agree", "Strongly 
+agree"))
 
@@
   mutate(idencuesta = 1)
+datos.justeduc$just_educ <- factor(datos.justeduc$just_educ, levels = rev(levels(datos.justeduc$just_educ)))
+etiquetas.justeduc$just_educ <- factor(etiquetas.justeduc$just_educ, levels = rev(levels(etiquetas.justeduc$just_educ)))
 ```
@@
     xlab(label = NULL) + 
-    theme(legend.position = 'top',
+    theme(#legend.position = 'top',
           legend.title = element_blank(),
           plot.caption = element_text(hjust = 0)) +
-    scale_fill_viridis_d(begin = 0, end = .95, direction = -1, option = 'viridis') +
+    scale_fill_manual(values = colors) +
     geom_shadowtext(data = etiquetas.justeduc, 
-              aes(label = ifelse(porcentaje > 0.03 , scales::percent(porcentaje, accuracy = .1),"")),
+              aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
               position = position_stack(vjust = .5),
@@
 merit_plot <- df_study1 %>% 
-  select(merit_talent, merit_effort, des_perc, just_educ) %>% 
+  select(merit_effort, merit_talent, des_perc, just_educ) %>% 
   sjPlot::plot_stackfrq(weight.by = df_study1$ponderador_long_total,
@@
 nor disagree", "Agree", "Strongly agree"),
-                      axis.labels = c("Merit talent", "Merit effort", "Inequality perception", "Justification of educational inequality")
+                      axis.labels = c("Merit effort", "Merit talent", "Inequality 
+perception", "Justification of 
+educational inequality")
                       ) +
@@
 
-ggsave(merit_plot, file="output/graphs/merit_plot.png", height = 7, width = 8)
+ggsave(merit_plot, file="output/graphs/merit_plot.png", height = 6, width = 10)
 ```
@@
 ```{r}
-colors<- c("#f1eef6ff","#bdc9e1ff","#b3b3b3ff","#74a9cfff","#0570b0ff")
+
 range <- c(1, 1)
@@
 ```{r}
-educ_effort_year = elsoc_diseno %>% filter(!is.na(merit_effort)) %>% 
-  group_by(ola, merit_effort) %>%
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
+df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               educ, quintil1:sexo)
+df_study1 <- df_study1[complete.cases(df_study1$segmento), ]
+
+elsoc_diseno <- as_survey_design(df_study1,
+                                 ids = segmento, #muestreo por conglomerado a nivel de manzanas (segmento)
+                          strata = estrato, #muestreo estratificado a nivel ciudad (estato)
+                          weights = ponderador_long_total, #ponderador longitudinal
+                          nest = TRUE
+                          )
+
+just_years_mean = elsoc_diseno %>% #filter(!is.na(ola)) %>% 
+  group_by(ola) %>%
   summarise(education = survey_mean(just_educ, vartype = "ci", na.rm = T)) %>% 
@@
 
-educ_talent_year = elsoc_diseno %>% filter(!is.na(merit_talent)) %>% 
-  group_by(ola, merit_talent) %>%
-  summarise(education = survey_mean(just_educ, vartype = "ci", na.rm = T)) %>% 
+just_years <- just_years_mean %>% 
+  pivot_longer(cols = c("education"), 
+               names_to = "variable",
+               values_to = "value") %>% 
+  cbind(type=c("Dependent")) %>% 
+  select(ola, variable, value, type, value_upp=education_upp, value_low=education_low)
+
+ineq_years = elsoc_diseno %>% #filter(!is.na(ola)) %>% 
+  group_by(ola) %>%
+  summarise(inequality = survey_mean(des_perc, vartype = "ci", na.rm = T),
+            ) %>% 
   ungroup()
 
-categorias_x <- c("Strongly disagree", "Disagree", "Neither agree 
-nor disagree", "Agree", "Strongly agree")
-categorias_f <- c(2016, 2017, 2018, 2019, 2022)
-
-effort_educ_year_plot <- educ_effort_year %>% 
-  ggplot(aes(factor(merit_effort), education, color = factor(ola), shape = factor(ola))) +
-  theme_bw(base_size = 14) +
-  geom_point(size = 4)+ 
-  ylab("Justification of inequality") +
-  xlab("Merit: effort") +
-  ylim(1, 3) +
-  scale_x_discrete(labels = categorias_x) +
-  scale_color_discrete(labels = categorias_f, name="Wave") +
-  scale_shape_discrete(labels = categorias_f, name="Wave")
-
-effort_educ_year_plot
-
-talent_educ_year_plot <- educ_talent_year %>% 
-  ggplot(aes(factor(merit_talent), education, color = factor(ola), shape = factor(ola))) +
+ineq_years <- ineq_years %>% 
+  pivot_longer(cols = c(
+                        "inequality"), 
+               names_to = "variable",
+               values_to = "value") %>% 
+  cbind(type=c("Independent")) %>% 
+  select(ola, variable, value, type, value_upp=inequality_upp, value_low=inequality_low)
+
+talent_years = elsoc_diseno %>% #filter(!is.na(ola)) %>% 
+  group_by(ola) %>%
+  summarise(
+            talent = survey_mean(merit_talent, vartype = "ci", na.rm = T)) %>% 
+  ungroup()
+
+talent_years <- talent_years %>% 
+  pivot_longer(cols = c(
+                        "talent"), 
+               names_to = "variable",
+               values_to = "value") %>% 
+  cbind(type=c("Independent")) %>% 
+  select(ola, variable, value, type, value_upp=talent_upp, value_low=talent_low)
+
+effort_years = elsoc_diseno %>% #filter(!is.na(ola)) %>% 
+  group_by(ola) %>%
+  summarise(
+            effort = survey_mean(merit_effort, vartype = "ci", na.rm = T)) %>% 
+  ungroup()
+
+effort_years <- effort_years %>% 
+  pivot_longer(cols = c(
+                        "effort"), 
+               names_to = "variable",
+               values_to = "value") %>% 
+  cbind(type=c("Independent")) %>% 
+  select(ola, variable, value, type, value_upp=effort_upp, value_low=effort_low)
+
+vars_years <- rbind(just_years, ineq_years, talent_years, effort_years)
+
+categorias_x <- c(2016, 2017, 2018, 2019, 2022)
+vars_years$variable <- factor(vars_years$variable, levels = c("education", "inequality", "talent", "effort"), labels=  c("Justification
+educational
+inequality
+(dependent)", "Inequality
+perception", "Merit: Talent", "Merit: Effort"),
+ordered = TRUE)
+
+categorias_z <- c("Justification
+educational
+inequality
+(dependent)", "Inequality
+perception
+(independent)", "Merit: Talent
+(independent)", "Merit: Effort
+(independent)")
+
+## Crear el gráfico principal
+years_plot <- ggplot(vars_years, aes(ola, value, color = variable, shape = variable)) +
   theme_bw(base_size = 14) +
   geom_point(size = 4) +
-  ylab("Justification of inequality") +
-  xlab("Merit: talent") +
-  ylim(1, 3) +
+  geom_line(aes(group = variable)) +
+  geom_errorbar(aes(ymin = value_low, ymax = value_upp), width = 0.2) +
+  ylab("Mean") +
+  xlab("Wave") +
+  ylim(1, 5) +
   scale_x_discrete(labels = categorias_x) +
-  scale_color_discrete(labels = categorias_f, name="Wave") +
-  scale_shape_discrete(labels = categorias_f, name="Wave")
-
-talent_educ_year_plot
-
-ggsave(effort_educ_year_plot, file="output/graphs/effort_educ_year.png", height = 7, width = 7)
-ggsave(talent_educ_year_plot, file="output/graphs/talent_educ_year.png", height = 7, width = 7)
-
-educ_effort_year <- educ_effort_year %>% pivot_longer("merit_effort", 
-                              names_to="merit", 
-                              values_to="value")
-
-educ_talent_year <- educ_talent_year %>% pivot_longer("merit_talent", 
-                              names_to="merit", 
-                              values_to="value")
-
-educ_merit_year <- rbind(educ_effort_year, educ_talent_year)
-
-custom_labeller <- as_labeller(function(variable, value) {
-  return(categorias_f[value])
-})
-categorias_z <- c("Effort", "Talent")
-merit_educ_year <- educ_merit_year %>% 
-  ggplot(aes(factor(value), education)) +
-  theme_bw(base_size = 14) +
-  geom_point(aes(color=merit, shape=merit), size=3) +
-  ylab("Justification of inequality") +
-  xlab("Merit") +
-  ylim(1, 3) +
-  facet_wrap(~factor(ola), scales = "free", labeller = custom_labeller) +
-  scale_color_discrete(labels = categorias_z, name="Merit") +
-  scale_shape_discrete(labels = categorias_z, name="Merit")
-merit_educ_year
-
-ggsave(merit_educ_year, file="output/graphs/merit_educ_year.png", height = 7, width = 7)
+  scale_color_discrete(labels = categorias_z, name = "Variable") +
+  scale_shape_manual(labels = categorias_z, name = "Variable", values = c("square", "circle", "triangle", "diamond")) +
+  theme(legend.position = "none") +  # Desactivar la leyenda por ahora
+  coord_cartesian(xlim = c(1, 6)) +  # Ampliar el eje x
+  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022, ""))  # Personalizar etiquetas del eje x
+
+# Obtener los datos para etiquetar las líneas
+label_data <- vars_years %>%
+  group_by(variable) %>%
+  filter(row_number() == n()) %>%
+  ungroup() %>%
+  mutate(label_x = 5.5, label_y = value)
+
+# Añadir etiquetas personalizadas
+years_plot <- years_plot +
+  geom_point(data = label_data, aes(label_x - 0.05, label_y, color = variable, shape = variable), size = 4) +  # Capa de puntos para los símbolos de forma
+  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 0.05, nudge_y = 0, size = 4)  # Etiquetas
+
+years_plot
+
+ggsave("output/graphs/years_plot.png", plot = years_plot, height = 7, width = 7, dpi = 300)
+
+```
+
+```{r}
+# Calcular cuantiles para las variables
+
+boxplot <- df_study_long_t6_comunas %>% select(just_educ, des_perc, merit_talent, merit_effort, ola) %>% as.data.frame()
+
+boxplot <- boxplot %>% 
+  pivot_longer(cols = c("just_educ", 
+                        "des_perc", 
+                        "merit_talent", 
+                        "merit_effort"), 
+               names_to = "variable",
+               values_to = "value") %>% na.omit()
+
+boxplot$variable <- factor(boxplot$variable,
+                   levels = c("just_educ", 
+                        "des_perc", 
+                        "merit_talent", 
+                        "merit_effort"),
+                   labels = c("Justification
+educational
+inequality", "Inequality
+perception", "Merit: Talent", "Merit: Effort"))
+
+boxplot_plot <- boxplot %>%
+  ggplot(aes(x = as.factor(ola), y = value, fill=variable, color=variable)) + 
+  geom_boxplot() +
+  scale_y_continuous(name = "Value", limits = c(1, 5)) +
+  scale_x_discrete(name = "Wave", labels=categorias_x) +
+  labs(fill = "Variable", color = "Variable") +
+  theme_bw() +
+  theme(axis.text.x = element_text(margin = margin(b = 15)))
+boxplot_plot
+
+ggsave(boxplot_plot, file="output/graphs/boxplot_plot.png", height = 7, width = 7)
 ```
@@
 h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+des_perc"
-h3 <- "ola_num+ola_2+des_perc+merit_talent"
-h4 <- "ola_num+ola_2+des_perc+merit_talent+merit_effort"
-h5 <- "ola_num+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo"
+h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
+h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
+h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
+h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
+h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
 
@@
 educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
@@
 omit <- "(education)|(ess)"
-texreg::htmlreg(list(educ0, educ1, educ2, educ3, educ4, educ5),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Model 6 is controlled by educational level, and subjective social status with no significant effects.", 
+texreg::htmlreg(list(educ0, educ1, educ2, educ3, educ4, educ5, educ6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Model 3 to 7 are controlled by educational level, and subjective social status with no significant effects.", 
                 omit.coef = omit,
@@
                                       "Wave^2",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: effort",
                                       "Quintile 2 <br> <i>(Ref. Quintile 1)</i>",
@@
                                       "Quintile NA",
+                                      "Age 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Age 50-64",
+                                      "Age 65 o más",
+                                      "Gender <br> <i>(Ref. Male)</i>",
                                       "Pol. pos: Center <br> <i>(Ref. Left)</i>",
@@
                                       "Pol. pos: Does not identify",
-                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Edad 50-64",
-                                      "Edad 65 o más",
-                                      "Sexo <br> <i>(Ref. Male)</i>"),
+                                      "Inequality perception",
+                                      "Merit: Talent",
+                                      "Merit: effort" ),
                 file = "output/tables/table1-individuales-pond-v2.html")
@@
   na.omit() %>% 
-  mutate(ola = as.factor(ola),
+  mutate(ola = as.numeric(ola),
          ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
+         sexo=as.factor(sexo),
+         edad=as.numeric(edad)
+         #pos_id=as.numeric(pos_id)
+         )
+
 # Hipotesis
-h1b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo"
-h2b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*des_perc"
-h3b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_talent"
-h4b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_effort"
-h5b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*pos_id"
-h6b <- "ola+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*edad"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*des_perc"
+h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*merit_talent"
+h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*merit_effort"
+h5b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*edad"
+h6b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*sexo"
+
+# A. justification educational inequality
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+I(ola_2)|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+I(ola_2)+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+I(ola_2)+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+I(ola_2)+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+I(ola_2)+edad|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(ola+I(ola_2)+sexo|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
 # table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(sexo)"
-texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-All the models are controlled by educational level, age, gender, subjective social status, political position, and income quintile",
+omit <- "(education)|(ess)|(quintil1)|(pos_id)"
+texreg::htmlreg(list(educ1,educ2,educ3,educ4,educ5,educ6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. All the models are controlled by educational level, subjective social status, political position, and income quintile",
                 omit.coef = omit,
                 custom.coef.names = c("Intercept",
-                                      "Wave 2017 <br> <i>(Ref. Wave 2016)</i>",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
+                                      "Wave",
+                                      "Wave^2",
+                                      "Age",
+                                      "Gender <br> <i>(Ref. Male)</i>",
                                       "Inequality perception",
@@
                                       "Merit: effort",
-                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
-                                      "Pol. pos: Right",
-                                      "Pol. pos: Does not identify",
-                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Edad 50-64",
-                                      "Edad 65 o más",
-                                      "2017*inequality perception",
-                                      "2018*inequality perception",
-                                      "2019*inequality perception",
-                                      "2022*inequality perception",
-                                      "2017*merit:talent",
-                                      "2018*merit:talent",
-                                      "2019*merit:talent",
-                                      "2022*merit:talent",
-                                      "2017*merit:effort",
-                                      "2018*merit:effort",
-                                      "2019*merit:effort",
-                                      "2022*merit:effort",
-                                      "2017*Center",
-                                      "2018*Center",
-                                      "2019*Center",
-                                      "2022*Center",
-                                      "2017*Right",
-                                      "2018*Right",
-                                      "2019*Right",
-                                      "2022*Right",
-                                      "2017*Not identify",
-                                      "2018*Not identify",
-                                      "2019*Not identify",
-                                      "2022*Not identify",
-                                      "2017*age30-49 ",
-                                      "2018*age30-49 ",
-                                      "2019*age30-49 ",
-                                      "2022*age30-49 ",
-                                      "2017*age50-64 ",
-                                      "2018*age50-64 ",
-                                      "2019*age50-64 ",
-                                      "2022*age50-64 ",
-                                      "2017*age65 or more",
-                                      "2018*age65 or more",
-                                      "2019*age65 or more",
-                                      "2022*age65 or more"
+                                      "Wave^2 * Inequality perception",
+                                      "Wave^2 * Merit: talent",
+                                      "Wave^2 * Merit: effort",
+                                      "Wave^2 * Age",
+                                      "Wave^2 * Gender"
                                       ),
@@
 ```
-
-```{r distribution-educ-context}
-#rm(list=ls())
-#if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg,
-               ggplot2
-               )
-load(file = here::here("input/data-proc/df_study1_long_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h1b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo"
-h2b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*des_perc"
-h3b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_talent"
-h4b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*merit_effort"
-h5b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*pos_id"
-h6b <- "ola+ola_2+des_perc+merit_talent+merit_effort+education+quintil1+ess+pos_id+edad+sexo+ola*edad"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(sexo)"
-texreg::htmlreg(list(educ2,educ3,educ4,educ5,educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-All the models are controlled by educational level, age, gender, subjective social status, political position, and income quintile",
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: effort",
-                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
-                                      "Pol. pos: Right",
-                                      "Pol. pos: Does not identify",
-                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Edad 50-64",
-                                      "Edad 65 o más",
-                                      "Wave*inequality perception",
-                                      "Wave*merit:talent",
-                                      "Wave*merit:effort",
-                                      "Wave*Center",
-                                      "Wave*Right",
-                                      "Wave*Not identify",
-                                      "Wave*age30-49 ",
-                                      "Wave*age50-64 ",
-                                      "Wave*age65 or more"
-                                      ),
-                file = "output/tables/table3-interac_pond-numeric.html")
-          #omit.coef = omit
-
-webshot2::webshot(url ="output/tables/table3-interac_pond-numeric.html" ,file ="output/tables/table3-interac-pond-numeric.png")
-
-## graph interact
-#sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
-predict<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
-predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
-predict_data <- predict_data %>% as.data.frame()
-
-predict_data$value <- c("ola1", 
-                             "ola2",
-                             "ola3", 
-                             "ola4",
-                             "ola6")
-
-predict_data$value <- factor(predict_data$value,
-                   levels = c("ola1", 
-                             "ola2",
-                             "ola3", 
-                             "ola4",
-                             "ola6"),
-                   labels = c("wave 1", 
-                              "wave 2",
-                              "wave 3",
-                              "wave 4",
-                              "wave 6"))
-
-interact_effect <- predict_data %>% ggplot(aes(value, coef1)) +
-  theme_bw(base_size = 14) +
-  geom_point(size = 4)+
-  geom_errorbar(aes(ymin=lb, ymax=ub),
-                  width=0.2) +
-  ylab("Coef") +
-  xlab("")
-interact_effect
-
-ggsave(interact_effect, file="output/graphs/interact_effect-pond.png", width = 7, height = 7)
-```

### 2024-03-20 production/prod_analysis-fix-t5.Rmd
update interacciones aleatorias

@@
          ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo),
-         edad=as.numeric(edad)
+         sexo=as.factor(sexo)
+ #        edad=as.numeric(edad)
          #pos_id=as.numeric(pos_id)
@@
 h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*des_perc"
+h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+des_perc*ola_2"
 h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*merit_talent"
-h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*merit_effort"
-h5b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*edad"
+h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola_2"
+h5b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*pos_id"
 h6b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*sexo"
+h7b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*quintil1"
+h8b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*edad"
 
 # A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+I(ola_2)|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+I(ola_2)+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+I(ola_2)+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+I(ola_2)+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+I(ola_2)+edad|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(ola+I(ola_2)+sexo|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+ola_2+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+ola_2+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+ola_2+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ8 <- lmer(formula(paste0("just_educ~",h8b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
 # table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(pos_id)"
-texreg::htmlreg(list(educ1,educ2,educ3,educ4,educ5,educ6),
+omit <- "(education)|(ess)"
+texreg::htmlreg(list(educ1,educ2,educ3,educ4,educ5,educ6,educ7,educ8),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. All the models are controlled by educational level, subjective social status, political position, and income quintile",
@@
                                       "Wave^2", 
-                                      "Age",
+                                      "Quintile 2 <br> <i>(Ref. Quintile 1)</i>",
+                                      "Quintile 3",
+                                      "Quintile 4",
+                                      "Quintile 5",
+                                      "Quintile NA",
+                                      "Age 30-49 <br> <i>(Ref. 18-29)</i>",
+                                      "Age 50-64",
+                                      "Age 65 o más",
                                       "Gender <br> <i>(Ref. Male)</i>",
+                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
+                                      "Pol. pos: Right",
+                                      "Pol. pos: Does not identify",
                                       "Inequality perception",
@@
                                       "Wave^2 * Merit: effort",
-                                      "Wave^2 * Age",
-                                      "Wave^2 * Gender"
+                                      "Wave^2 * Center",
+                                      "Wave^2 * Right",
+                                      "Wave^2 * Does not identify",
+                                      "Wave^2 * gender",
+                                      "Wave^2 * Quintile 2",
+                                      "Wave^2 * Quintile 3",
+                                      "Wave^2 * Quintile 4",
+                                      "Wave^2 * Quintile 5",
+                                      "Wave^2 * Quintile NA",
+                                      "Wave^2 * Age 30-49",
+                                      "Wave^2 * Age 50-64",
+                                      "Wave^2 * Age 65 or more"
                                       ),
@@
 webshot2::webshot(url ="output/tables/table3-interac_pond.html" ,file ="output/tables/table3-interac-pond.png")
-
-## graph interact
-#sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
-predict<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
+```
+
+```{r }
+fit <- lmer(just_educ ~ ola + ola_2 + (ola + ola_2 | idencuesta), data = df_study1)
+df_study1$pred_inter_pend_aleatorio <- predict(fit)
+
+# Predictions that ignore random effects:
+df_study1$fitted <- predict(fit, re.form = ~0)
+
+# Plot just fixed effects:
+plot<-df_study1 %>% 
+  ggplot(aes(y = fitted, x = ola_2)) +
+  geom_line() +
+  theme_light()
+
+ggsave(plot, file="output/graphs/fixed_effects.png")
+
+predict<-interplot::interplot(m = educ1,var1 = "ola",point = T)
 predict_data <- predict$data[c(1, 2, 4, 6, 8), ]

### 2024-03-21 production/prod_analysis-fix-t5.Rmd
change orden interactions and captions graphs

@@
                                                just_educ, des_perc, merit_effort, merit_talent,
-                                               educ, quintil1:sexo)
+                                               educ, quintil1:sexo) %>% na.omit()
 df_study1 <- df_study1[complete.cases(df_study1$segmento), ]
@@
 ```{r alluvial}
+colors<- c("#f1eef6ff","#bdc9e1ff","#b3b3b3ff","#74a9cfff","#0570b0ff")
 alluvial_justeduc <- ggplot(datos.justeduc, aes(x = ola, fill = just_educ, stratum = just_educ,
@@
           legend.title = element_blank(),
-          plot.caption = element_text(hjust = 0)) +
+          plot.caption = element_text(hjust = 1)) +
     scale_fill_manual(values = colors) +
@@
               bg.colour='grey30')+
-  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022))
+  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022)) +
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
 
@@
 
-ggsave(alluvial_justeduc, file="output/graphs/alluvial_dep.png", height = 7, width = 8)
+ggsave(alluvial_justeduc, file="output/graphs/alluvial_dep.png", height = 8, width = 7)
 ```
@@
                                                just_educ, des_perc, merit_effort, merit_talent,
-                                               educ, quintil1:sexo)
+                                               educ, quintil1:sexo) %>% na.omit()
 merit_plot <- df_study1 %>% 
@@
                       axis.labels = c("Merit effort", "Merit talent", "Inequality 
-perception", "Justification of 
+perception", "Justification 
+of 
 educational inequality")
                       ) +
-  theme_light() +
-  theme(legend.position="bottom", text = element_text(size = 14))
+  theme_light(base_size = 16) +
+  theme(legend.position="bottom", text = element_text(size = 14),
+        plot.caption = element_text(hjust = 1)) +
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
 merit_plot
@@
 ## Crear el gráfico principal
-years_plot <- ggplot(vars_years, aes(ola, value, color = variable, shape = variable)) +
+years_plot <- ggplot(vars_years, aes(ola, value, color = variable, shape = variable, linetype = variable)) +
   theme_bw(base_size = 14) +
@@
   scale_x_discrete(labels = categorias_x) +
-  scale_color_discrete(labels = categorias_z, name = "Variable") +
+  scale_color_manual(labels = categorias_z, name = "Variable", values =  c("#bdc9e1ff","#0570b0ff","#74a9cfff","#b3b3b3ff")) +
   scale_shape_manual(labels = categorias_z, name = "Variable", values = c("square", "circle", "triangle", "diamond")) +
-  theme(legend.position = "none") +  # Desactivar la leyenda por ahora
+  scale_linetype_manual(labels = categorias_z, name = "Variable", 
+                        values = c("solid", "dashed", "dotted", "dotdash")) +  # Definir diferentes tipos de línea
+   theme(legend.position = "none") +
   coord_cartesian(xlim = c(1, 6)) +  # Ampliar el eje x
@@
   geom_point(data = label_data, aes(label_x - 0.05, label_y, color = variable, shape = variable), size = 4) +  # Capa de puntos para los símbolos de forma
-  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 0.05, nudge_y = 0, size = 4)  # Etiquetas
+  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 0.05, nudge_y = 0, size = 4)  +
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
 
@@
 
-ggsave("output/graphs/years_plot.png", plot = years_plot, height = 7, width = 7, dpi = 300)
+ggsave("output/graphs/years_plot.png", plot = years_plot, height = 6, width = 7, dpi = 300)
 
@@
 load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
+df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               educ, quintil1:sexo) %>% na.omit()
 # Full sample_______________________________________________________________________
 mat_full<- 
-df_study_long_t6_comunas %>%
+df_study1 %>%
   # filter(ola==1) %>%
@@
   na.label = "-")
+# Agregar caption
+grid::grid.text("Source: own elaboration with pooled data 
+from ELSOC 2016-2022 (n = 8042)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
 dev.off()
@@
 texreg::htmlreg(list(educ0, educ1, educ2, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. Model 3 to 7 are controlled by educational level, and subjective social status with no significant effects.", 
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. <br>
+Note: Model 3 to 7 are controlled by educational level, and subjective social status with no significant effects.", 
                 omit.coef = omit,
@@
                                       "Merit: Talent",
-                                      "Merit: effort" ),
+                                      "Merit: Effort" ),
+                custom.gof.names = c("AIC",
+                                     "BIC",
+                                     "Log Likelihood",
+                                     "Num. obs.",
+                                     "Num. groups: Individuals",
+                                     "Var: Individuals (Intercept)",
+                                     "Var: Residual"),
                 file = "output/tables/table1-individuales-pond-v2.html")
@@
 # Hipotesis
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+des_perc*ola_2"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*merit_talent"
-h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola_2"
-h5b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*pos_id"
-h6b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*sexo"
-h7b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*quintil1"
-h8b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*edad"
+#h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+merit_effort+merit_effort*ola+merit_effort*ola_2"
+h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+merit_talent+merit_talent*ola+merit_talent*ola_2"
+h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+des_perc*ola+des_perc*ola_2"
+#h5b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*pos_id"
+#h6b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*sexo"
+#h7b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*quintil1"
+#h8b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*edad"
 
 # A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+ola_2+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+#educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+ola_2+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+ola_2+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+ola_2+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ8 <- lmer(formula(paste0("just_educ~",h8b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+ola_2+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+#educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+#educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+#educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+#educ8 <- lmer(formula(paste0("just_educ~",h8b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
 # table 3: interacciones
-omit <- "(education)|(ess)"
-texreg::htmlreg(list(educ1,educ2,educ3,educ4,educ5,educ6,educ7,educ8),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. All the models are controlled by educational level, subjective social status, political position, and income quintile",
+omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)"
+texreg::htmlreg(list(educ2,educ3,educ4),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. <br>
+Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender and political position",
                 omit.coef = omit,
@@
                                       "Wave^2",
-                                      "Quintile 2 <br> <i>(Ref. Quintile 1)</i>",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile NA",
-                                      "Age 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Age 50-64",
-                                      "Age 65 o más",
-                                      "Gender <br> <i>(Ref. Male)</i>",
-                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
-                                      "Pol. pos: Right",
-                                      "Pol. pos: Does not identify",
+                                      "Merit: Effort",
+                                      "Wave * Merit: Effort",
+                                      "Wave^2 * Merit: Effort",
+                                      "Merit: Talent",
+                                      "Wave * Merit: Talent",
+                                      "Wave^2 * Merit: Talent",
                                       "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: effort",
-                                      "Wave^2 * Inequality perception",
-                                      "Wave^2 * Merit: talent",
-                                      "Wave^2 * Merit: effort",
-                                      "Wave^2 * Center",
-                                      "Wave^2 * Right",
-                                      "Wave^2 * Does not identify",
-                                      "Wave^2 * gender",
-                                      "Wave^2 * Quintile 2",
-                                      "Wave^2 * Quintile 3",
-                                      "Wave^2 * Quintile 4",
-                                      "Wave^2 * Quintile 5",
-                                      "Wave^2 * Quintile NA",
-                                      "Wave^2 * Age 30-49",
-                                      "Wave^2 * Age 50-64",
-                                      "Wave^2 * Age 65 or more"
+                                      "Wave * Inequality perception",
+                                      "Wave^2 * Inequality perception"
+                                      ),
+                custom.gof.names = c("AIC", "BIC", "Log Likelihood",
+                                     "Num. obs.",
+                                     "Num. groups: Individuals",
+                                     "Var: Individuals (Intercept)",
+                                     "Var: Individuals Wave",
+                                     "Var: Individuals Wave^2",
+                                     "Var: Individuals Merit: Effort",
+                                     "Cov: Individuals (Intercept) Wave",
+                                     "Cov: Individuals (Intercept) Wave^2",
+                                     "Cov: Individuals (Intercept) Merit: Effort",
+                                     "Cov: Individuals Wave Wave^2",
+                                     "Cov: Individuals Wave Merit: Effort",
+                                     "Cov: Individuals Wave^2 Merit: Effort",
+                                     "Var: Residual",
+                                     "Var: Individuals Merit: Talent",
+                                     "Cov: Individuals (Intercept) Merit: Talent",
+                                     "Cov: Individuals Wave Merit: Talent",
+                                     "Cov: Individuals Wave^2 Merit: Talent",
+                                     "Var: Individuals Inequality perception",
+                                     "Cov: Individuals (Intercept) Inequality perception",
+                                     "Cov: Individuals Wave Inequality perception",
+                                     "Cov: Individuals Wave^2 Inequality perception"
                                      ),

### 2024-03-25 Title-Page.Rmd
render paper

---
title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic [^*]"
linestretch: '1.5'
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    number_sections: false
geometry: margin=0.78in
header-includes:
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage[figurename=Fig.]{caption}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
fontsize: 12pt

---

Juan Carlos Castillo, Department of Sociology, Universidad de Chile, Santiago de Chile. *

ORCID: 0000-0003-1265-7854

Julio Iturra, Bremen International Graduate School of Social Sciences, Bremen, Germany.

ORCID: 0000-0003-1993-3479

Kevin Carrasco, Center for Social Conflict and Cohesion Studies

ORCID: 0000-0002-9348-8361

Word count:

[^*]: Corresponding author (juancastillov@uchile.cl - [jc-castillo.com](https://juancarloscastillo.github.io/jc-castillo/) -  Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile). The authors recognize the support of ANID/FONDECYT grant 1210847 and of the Center for Conflict and Social Cohesion Studies (COES)  grant ANID/FONDAP 1513009.

### 2024-03-25 paper.Rmd
render paper

---
title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
# params:
# lang: en-GB
author: 
- name: Juan Carlos Castillo
  affiliation: Department of Sociology, Universidad de Chile
  email: juancastillov@uchile.cl
  number: 1
- name: Julio Iturra
  affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
  number: 2
- name: Kevin Carrasco
  affiliation: Center for Social Conflict and Cohesion Studies
  number: 3
abstract: |
  .
  \newline
  \newline

   **Keywords**: meritocracy, social inequality, inequality justification
output:
  bookdown::pdf_document2:
    template: null
    number_sections: false    
    toc: false
    keep_tex: false
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
  bookdown::word_document2:
    number_sections: false
    reference_docx: input/template.docx
    toc: false
  bookdown::html_document2:
    number_sections: false
linkcolor: blue
bibliography:
  - input/bib/merit-edjust.bib
csl: input/bib/apa7-singlespace.csl
editor_options:
  chunk_output_type: console
geometry: margin=1in
header-includes:
  - \usepackage{times}           # Times New Roman
  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
always_allow_html: true # for word output
fontsize: 10pt
papersize: a4
---

```{r eval=FALSE, include=FALSE}
 # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
 # clean #in the yml
 rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
 rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
 rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
```

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H", fig.align = 'center')
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

<!---
\pagebreak

# Acknowledgments

The authors recognize the support of ANID/FONDECYT grant 1160921 and of  the Center for Conflict and Social Cohesion Studies (COES)  grant ANID/FONDAP 1513009.

\pagebreak

--->

# Introduction

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023]. Furthermore, empirical evidence from social science research has revealed the exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021]. Despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].

In the present paper we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities. Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background. In recent years there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022]. This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008]. Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and on to what extent this phenomenon is associated with the perception of meritocracy and it variations during the COVID pandemic.

The Chilean case is particularly relevant when it comes to the study of inequalities in education. The country is characterized by a high level of economic inequality associated with the privatization and commodification in several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016]. The neoliberal reforms implemented in this country during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joikomujicaCuasimercadoEducativoChile2019]. Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities. Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramosarellanoEducacionPromesaIncumplida2022]. Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that advanced free higher education reforms and diminish the role of private actors in the educational system.

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021]. A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. Within this framework, we point out that periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that could be associated with a challenge of meritocratic ideals. Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.

# Inequality beliefs and the justification of economic inequality

The justification of economic inequalities is a controversial field of study. As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable. Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p. 3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021]. Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.

The field of study of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016]. From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in identifying a field of studies comprising perceptions, beliefs, and justifications in areas such as inequality of opportunities, poverty, and distributive justice. Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p. 30]. However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality. In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p. 2]. This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978]. Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.

Theoretically, @liebigSociologyJustice2016 establishes three fundamental distinctions in justice attitudes in the field of sociological justice research. First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement. Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities). Finally, they can be _outcome-related attitudes_ related primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes). Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated to concepts that reflect distributive principles [@kluegel_social_1995;@sonhing_merit_2011]

An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986]. The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021]. In contrast to this evidence, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003,@vandertoornTwentyYearsSystem2014]. Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020]. As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jostTheorySystemJustification2020].

Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022]. Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as just in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987, @Wegener1990]. Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018]. In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality. For instance, it has been established that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011]. Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022]. Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.

# The justification of educational inequality

Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences. A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services. For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services. Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes. Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.

The study of inequality justification in the sphere of education is a rather specific and unattended field. Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities. This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area. This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).

In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are ISSP (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the ESS 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects. The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.

A number of studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies. Some studies emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016]. Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. However, most of the studies in this area go beyond status differences, focusing on how subjective factors such as perceptions and beliefs - among them meritocracy - are related to different levels of justification.

The perception of economic inequality is one of the main factors associated with the justification of educational inequality. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being perceived [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the undersetimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]

Besides focusing on the relationship between perception and justification of educational inequality, another group of studies have addressed how the justification of inequality is associated with the support of different distributive principles. In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification  [@jost_psychology_2003]) is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school'). In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.

_Justificacion, inequality and meritocracy in the educational domain_

The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to Justificacion, inequality and meritocracy in the educational domainlegitimizing social inequality [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background (Goldin, 2000).

Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs. Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities as well as status differences [@McCoy2007]. This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023]. Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes. However, empirical findings also reveal nuances in this relationship. Research has identified contextual factors that moderate the link between meritocratic perceptions and acceptance of educational inequality. For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023]. This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.

_This study_

The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification. As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations inequality justification in this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (as the pandemic) impacts on public concerns about equality and redistribution [@breznau_welfare_2021]. Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, as education, could have been impacted in this period. Consistently, our first hypothesis is:

$H_1$: The justification of educational inequalities decreases after the pandemic

Our second argument is based on the evidence that relates rational interests with justifications and preferences [@hing_failure_2019], expecting that:

$H_2$: Those with lower status characteristics show lower justification of educational inequalities

Thirdly, we address inequality perception. As referred to above, the justification of inequality requires, first and foremost, that such inequality is known or perceived by individuals. As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:

$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.

Our fourth and last hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:

$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.

Finally, and in line with hypothesis 1 and 4, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic have affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged. As a consequence, the justification of inequality could show a decrease over time:

$H_5$:[interaction] The positive association between meritocracy and inequality justification mitigates along time

#  Data, variables & methods

## Data

The main data source is the Chilean Longitudinal Social Survey 2016–2022. ELSOC has been designed to evaluate the way in which individuals think, feel, and behave yearly regarding a set of social issues related to conflict and social cohesion in Chile. The sampling design is probabilistic, stratified, clustered, and multistage. It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities comprising a total of 2,927 participants aged between 18 and 75 years on wave 1. It is representative of people in the north and south of the country. In addition, the sample represents 77% of the country's total population and 93% of the urban population, with a response rate of 62.4% [@elsoc_estudio_2022].

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The questionnaire was administered face-to-face, but in the 2021 wave, it was conducted entirely over the phone, and the dependent variables of this research were excluded. In 2018, wave 3 included a refreshment sample in order to counter survey attrition. The same sampling strategy as that of wave 1 was implemented for selecting new cases. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in this article because we wanted to analyze a longer trend. Thus, only cases from the original sample are employed in the analytical sample. Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants. The attrition in subsequent waves was 15.5% in wave 2 (N = 2,473), 9.9% in wave 3 (N = 2,229), 3.4% in wave 4 (N = 2,153), 19.2% in wave 5 (N = 1,739), and 0.6% in wave 6 (N = 1728). In broader terms, the accumulated attrition between Wave 1 and Wave 6 is 41%. The analysis used longitudinal weights to avoid bias due to the study’s sampling design. Longitudinal weights allow us to control for attrition biases arising from systematic patterns of non-participation in the survey after the first wave, and they also allow us to incorporate known information about the target population in the construction of the weights. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.

## Variables

The dependent variable of this study is the justification of educational inequality. This variable is measured from the phrase "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish). Respondents rated their preferences using a Likert scale of 5 categories from “strongly disagree” to “strongly agree.” The change in this variable over the years is as follows: 2016 mean=1.87, sd=0.88; 2017 mean=1.80, sd=0.91; 2018 mean=1.82, sd=0.92; 2019 mean=1.81, sd=0.84; 2022 mean=2.11, sd=0.93.

The main independent variables correspond to three items divided into two groups: one related to the perception of inequality (i.e., in Chile, the income differences are too large) and the other related to the perception of meritocracy (i.e., in Chile, people are rewarded for their effort). Both groups were measured using a Likert scale of 5 categories from “strongly disagree” to “strongly agree.”

The educational level of the respondents, the household's income quintile per capita, their subjective social status, their political identification, and their age and gender were included as control variables. \@ref(fig:descriptives) shows the items used, their response categories, and their frequency.

```{r descriptives, echo=FALSE, fig.cap = "Independent variables", fig.align='center'}
 knitr::include_graphics("output/tables/desc02.png")
```

Figure 1 shows the distribution of the dependent variable of Justification of educational inequality and both groups of independent variables for all the years analyzed. On the one hand, Figure 1 shows that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education. Likewise, there is a large group of respondents (91.7%) who agree or strongly agree that income differences in Chile are too large. On the other hand, there is a significant degree of disagreement (disagree + strongly disagree) about whether people are rewarded for their talent (47.6%)  and for their efforts (57%).

```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
 knitr::include_graphics("output/graphs/figure1.png")
```

```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
 knitr::include_graphics("output/graphs/corr.png")
```

Figure 2 presents a correlation matrix of the main variables analyzed, using data from all survey waves. In this matrix, the correlations vary between high and moderate values, which indicates a certain level of association between the variables. First, the dependent variable has a moderate and negative association with the perception of inequality and a moderate and positive association with both meritocracy perception variables. Secondly, the perception of inequality has a moderate and negative association with meritocracy perception variables. Finally, the two meritocracy variables present a high and positive association with each other.

## Methods

Given the hierarchical structure of the data (observations nested in surveys nested in municipalities), we applied a longitudinal multilevel strategy [@singer_applied_2009]. Longitudinal multilevel models are suited to account for the shared variance among units in the data to estimate standard errors better. Since individuals share variance within themselves over time, if the error structure is not considered, it would be as if they were considered different individuals. Multilevel models allow a solution in regression estimation by adding a random term that represents the variance associated with the nesting of the data (random effects). The linear multilevel models are estimated using the R library “lme4” [@singer_applied_2009, p.4].

```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
  #RUN THIS BEFORE THE ANALYSIS
  pacman::p_load(knitr)
  table_format = if(is_html_output()) {
    "html"
  } else if(is_latex_output()) {
    "latex"
  }

  table_format2 = if(is_html_output()) {
    T
  } else if(is_latex_output()) {
    F
  }
```

# Results

```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
 knitr::include_graphics("output/graphs/alluvial_dep.png")
```

Figure 3 illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022. Each year represents stacked percentual frequencies, and the flows in between represent the within-subjects’ change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010]. For instance, from the 32.2% who strongly disagree with inequality justification in 2016, about half of them keeps responding the same in 2016, whereas the other half shifts their opinion to other response categories. In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave. This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to a half when compared with previous years.

```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
 knitr::include_graphics("output/graphs/years_plot.png")
```

Figure 4 shows the average changes in the main variables considered for this study. Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception. Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases. As the merit variables is concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated to talent.

## Multivariate analysis

Table 2 shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves in order to observe whether there are intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01). Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease of inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.

Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality. On the contrary, the gender variable has a negative effect, expressing that women in average justify less inequality in education when compared to males. Educational level and subjective social status are not shown in this table as they do not exhibit significant effects, but the complete estimation of the models is presented in the appendix. Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing. Inequality perception is added in Model 4, showing a negative association with justification of educational inequality as hypothetisized that remains stable when controlling for the rest of the variables. Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<0.01), however, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<0.01), but it has a larger weight than the perception of talent.

```{r multilevel, echo=FALSE, fig.cap = "Multilevel longitudinal models for the justification of inequality in education", fig.align='center'}
 knitr::include_graphics("output/tables/table1-individuales-pond-v2.png")
```

```{r interact, echo=FALSE, fig.cap = "Interaction effects for the justification of economic inequality", fig.align='center'}
 knitr::include_graphics("output/tables/table3-interac-pond.png")
```

In this final part of the analyses we contrast longitudinal hypotheses regarding changes in the relationships between meritocracy and justification of economic inequality over time. In hypothesis 5 we proposed that the association between perception of meritocracy and inequality justification mitigates along time, as meritocratic ideals could have weaken due to critical situations associated to the COVID pandemic. We test this hypothesis through interaction effects, displayed in Table 2. Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for). Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort. As observed, only the effort variable show a significant interaction with the time variable (0.01, p<0.05), meaning that, in average, the association between perception of meritocracy based on effort and the justification of inequality increases in 0.01 points in every measurement point.

# References

# Appendix

```{r multilevel-full, echo=FALSE, fig.cap = "Multilevel effects of justification of economic inequality", fig.align='center'}
 knitr::include_graphics("output/tables/table1-individuales-pond-appendix.png")
```

### 2024-03-25 production/prod_analysis-fix-t5.Rmd
render paper

@@
 omit <- "(education)|(ess)"
-texreg::htmlreg(list(educ0, educ1, educ2, educ3, educ4, educ5, educ6),
+texreg::htmlreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. <br>
-Note: Model 3 to 7 are controlled by educational level, and subjective social status with no significant effects.", 
+Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
                 omit.coef = omit,
@@
                                       "Wave^2",
+#                                      "High School <br> <i>(Ref. Primary)</i>",
+#                                      "Technical",
+#                                      "Universitary",
                                       "Quintile 2 <br> <i>(Ref. Quintile 1)</i>",
@@
                                       "Quintile NA",
+#                                      "Subjective Social Status",
                                       "Age 30-49 <br> <i>(Ref. 18-29)</i>",
@@
 # Hipotesis
-#h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+merit_effort+merit_effort*ola+merit_effort*ola_2"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+merit_talent+merit_talent*ola+merit_talent*ola_2"
-h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+des_perc*ola+des_perc*ola_2"
+h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
+#h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola+merit_effort*ola_2"
+h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+
+h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
+#h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+des_perc*ola+des_perc*ola_2"
 #h5b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*pos_id"
@@
 # A. justification educational inequality
-#educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+ola_2+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+ola_2+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+ola_2+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+#educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+ola_2+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+#educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+ola_2+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 #educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
@@
 # table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)"
-texreg::htmlreg(list(educ2,educ3,educ4),
+omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
+texreg::htmlreg(list(educ6,educ3,educ1),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. <br>
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender and political position",
+Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, <br> political position and inequality perception",
                 omit.coef = omit,
@@
                                       "Wave^2",
+                                      "Merit: Talent",
                                       "Merit: Effort",
-                                      "Wave * Merit: Effort",
-                                      "Wave^2 * Merit: Effort",
-                                      "Merit: Talent",
                                       "Wave * Merit: Talent",
-                                      "Wave^2 * Merit: Talent",
-                                      "Inequality perception",
-                                      "Wave * Inequality perception",
-                                      "Wave^2 * Inequality perception"
+                                      "Wave * Merit: Effort"
                                       ),
@@
                                      "Var: Individuals (Intercept)",
+                                     "Var: Residual",
                                      "Var: Individuals Wave",
-                                     "Var: Individuals Wave^2",
-                                     "Var: Individuals Merit: Effort",
+                                     "Var: Individuals Merit: Talent",
                                      "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Wave^2",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Wave^2",
-                                     "Cov: Individuals Wave Merit: Effort",
-                                     "Cov: Individuals Wave^2 Merit: Effort",
-                                     "Var: Residual",
-                                     "Var: Individuals Merit: Talent",
                                      "Cov: Individuals (Intercept) Merit: Talent",
                                      "Cov: Individuals Wave Merit: Talent",
-                                     "Cov: Individuals Wave^2 Merit: Talent",
-                                     "Var: Individuals Inequality perception",
-                                     "Cov: Individuals (Intercept) Inequality perception",
-                                     "Cov: Individuals Wave Inequality perception",
-                                     "Cov: Individuals Wave^2 Inequality perception"
-                                     ),
+                                     "Var: Individuals Merit: Effort",
+                                     "Cov: Individuals (Intercept) Merit: Effort",
+                                     "Cov: Individuals Wave Merit: Effort"),
                 file = "output/tables/table3-interac_pond.html")

### 2024-03-25 paper.Rmd
add conclusion

@@
 The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-In contrast to this evidence, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003,@vandertoornTwentyYearsSystem2014].
+In contrast to this evidence, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
 Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jostTheorySystemJustification2020].
+As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
 
 Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as just in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987, @Wegener1990].
+Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as just in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
 Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
@@
 
+# Discussion and conclusions
+
+Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
+In general, our results showed mixed evidence with regard to our hypothesis.
+Whereas we find support for the association between perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
+We argued that  in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
+To this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is an striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
+As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises, but a more direct and rational explanation of this phenomenon is difficult to sustain.
+As in the Chilean case the pandemic period occurred along with a constitucional process, disentangling different forces driving public opinion it becomes difficult and deserves more cualitative and quantitative studies.
+Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that lead to the deligitimation of this assembly.
+This could have drive preferences in a conservative direction, which was actually reflected in the election of a right-wing second assembly after a failed first process [@palanza_chile_2023;@sazo_chile_2023].
+
+Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
+In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on that rational interests of the better-off would be consistent with a larger inequality justification.
+Besides status, we also considered inequality perception, as the literature on attitudes towards inequality has argued that a greater justification of inequality its associated to its perceived magnitude ($H_3$).
+Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
+
+The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
+At a first look the results are consistent with such hypothesis, however, there are some nuances attending to the different aspects of meritocracy considered here, namely effort and talent.
+However, conceptually, meritocracy centralizes people's efforts and talent as the main factors for achieving status.
+In this regard, our results are favorable for the effort dimension but are not confirmed for the talent dimension.
+In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
+This could be explained as that talent could be more associated to luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
+
+Finally, as our first hypothesis supported, our expectations were that the relationship between meritocratic perceptions and justification would tend to diminish over time.
+However, longitudinal evidence shows that the perception of meritocracy in relation to effort strengthens its relation over time with respect to the justification of inequality in education, which offers evidence against the temporal change hypothesis ($H_5$).
+In other words, we know that people who perceive that effort in Chile is rewarded are also more likely to justify that those with higher incomes have better access to education for their children.
+Additionally, we see the passage of time, including the health crisis generated by the COVID-19 pandemic.
+In this regard, an alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort-, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
+
+Among the limitations of our studies, we can mention at least three.
+First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
+In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of an individual or educational institutions.
+Second, as has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
+In this sense, the available measurements restrict us to the use of single indicators for effort and talent, and further research should improve the measurement quality in this regard.
+Finally, there are limitations involving the longitudinal dimension of the data used.
+In this regard, the ELSOC had to be modified in order to maintain its stability over time during the pandemic, which implied that the educational justice indicator was excluded in the 2021 measurement.
+This implies that the temporal change is not exactly one year in each measurement, with the largest temporal gap being between 2019 and 2022.
+Consequently, the conclusions regarding the temporal change have to be considered carefully, and further data waves could give us more information in this matter.
+
+The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
+In this regard, thanks to the survey's sampling strategy, it is possible to make contextual estimates at the municipality level.
+Future research could also include individual-level data with contextual information from administrative data sources, allowing for instance the consideration of contextual socioeconomic inequality as wellas its dynamics over time.
+Furthermore, this dataset could also be used in the future for comparing inequality justification in policy areas besides education, such as in health and in pensions.
+Future research may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes and their impacts for solidarity and social cohesion.
+
 # References

### 2024-03-25 paper.Rmd
Co-authored-by: Julio César Iturra Sanhueza <julioiturrasanhueza@gmail.com> Co-authored-by: Kevin Carrasco <Kevin-carrasco@users.noreply.github.com>

@@
 As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed the exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
+Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
+Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
+In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
 
-In the present paper we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities.
 Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
+In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
 This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and on to what extent this phenomenon is associated with the perception of meritocracy and it variations during the COVID pandemic.
+Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
 
 The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification in several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented in this country during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joikomujicaCuasimercadoEducativoChile2019].
+The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
+The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joikomujicaCuasimercadoEducativoChile2019].
 Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
 Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramosarellanoEducacionPromesaIncumplida2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that advanced free higher education reforms and diminish the role of private actors in the educational system.
+Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
 
@@
 A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that could be associated with a challenge of meritocratic ideals.
+Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
 Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
@@
 
-The field of study of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in identifying a field of studies comprising perceptions, beliefs, and justifications in areas such as inequality of opportunities, poverty, and distributive justice.
+The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
+From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
 Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
@@
 Finally, they can be _outcome-related attitudes_ related primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated to concepts that reflect distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
+Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
 
@@
 The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-In contrast to this evidence, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
+Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
 Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
@@
 Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as just in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
+Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
 Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
 In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, it has been established that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
+For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
 Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
@@
 
-A number of studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some studies emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
+Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
+Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
 Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-However, most of the studies in this area go beyond status differences, focusing on how subjective factors such as perceptions and beliefs - among them meritocracy - are related to different levels of justification.
+However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
 
-The perception of economic inequality is one of the main factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being perceived [@sonhing_merit_2011;@castillo_multidimensional_2023].
+The perception of economic inequality is one of the factors associated with the justification of educational inequality.
+As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
 For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
@@
 
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies have addressed how the justification of inequality is associated with the support of different distributive principles.
+Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
 In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003]) is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
+They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
 In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
@@
 However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to Justificacion, inequality and meritocracy in the educational domainlegitimizing social inequality [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background (Goldin, 2000).
+This would contribute to justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
 
 Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities as well as status differences [@McCoy2007].
+Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
 This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
@@
 However, empirical findings also reveal nuances in this relationship.
-Research has identified contextual factors that moderate the link between meritocratic perceptions and acceptance of educational inequality.
 For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
@@
 The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (as the pandemic) impacts on public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, as education, could have been impacted in this period.
+As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
+Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
+Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
 Consistently, our first hypothesis is:
@@
 Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is known or perceived by individuals.
+As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
 As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
@@
 
-Our fourth and last hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
+Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
 
@@
 Finally, and in line with hypothesis 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic have affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
+As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
 As a consequence, the justification of inequality could show a decrease over time:
 
-$H_5$:[interaction] The positive association between meritocracy and inequality justification mitigates along time
+$H_5$: The positive association between meritocracy and inequality justification mitigates along time
 
@@
 The main data source is the Chilean Longitudinal Social Survey 2016–2022.
-ELSOC has been designed to evaluate the way in which individuals think, feel, and behave yearly regarding a set of social issues related to conflict and social cohesion in Chile.
+The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
 The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities comprising a total of 2,927 participants aged between 18 and 75 years on wave 1.
-It is representative of people in the north and south of the country.
-In addition, the sample represents 77% of the country's total population and 93% of the urban population, with a response rate of 62.4% [@elsoc_estudio_2022].
+It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
+Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
+In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
 
 The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The questionnaire was administered face-to-face, but in the 2021 wave, it was conducted entirely over the phone, and the dependent variables of this research were excluded.
-In 2018, wave 3 included a refreshment sample in order to counter survey attrition.
+The waves 2016,2017,2018,2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
+However, in 2021 a reduced version was conducted using computer-assisted telephone interviews (CATI).
+In addition, wave 3 included a refreshment sample to counter survey attrition.
 The same sampling strategy as that of wave 1 was implemented for selecting new cases.
 As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in this article because we wanted to analyze a longer trend.
-Thus, only cases from the original sample are employed in the analytical sample.
+The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
 Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-The attrition in subsequent waves was 15.5% in wave 2 (N = 2,473), 9.9% in wave 3 (N = 2,229), 3.4% in wave 4 (N = 2,153), 19.2% in wave 5 (N = 1,739), and 0.6% in wave 6 (N = 1728).
-In broader terms, the accumulated attrition between Wave 1 and Wave 6 is 41%.
-The analysis used longitudinal weights to avoid bias due to the study’s sampling design.
-Longitudinal weights allow us to control for attrition biases arising from systematic patterns of non-participation in the survey after the first wave, and they also allow us to incorporate known information about the target population in the construction of the weights.
+In broader terms, the accumulated attrition between Wave 1 and Wave 6 is 41%, achieving a final sample of N = 1728.
+The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
 For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
@@
 
-The dependent variable of this study is the justification of educational inequality.
-This variable is measured from the phrase "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Respondents rated their preferences using a Likert scale of 5 categories from “strongly disagree” to “strongly agree.” The change in this variable over the years is as follows: 2016 mean=1.87, sd=0.88; 2017 mean=1.80, sd=0.91; 2018 mean=1.82, sd=0.92; 2019 mean=1.81, sd=0.84; 2022 mean=2.11, sd=0.93.
+The main variable of this study is the justification of educational inequality.
+This item is measured using the following statement:"It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
+Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree.” (5)
 
-The main independent variables correspond to three items divided into two groups: one related to the perception of inequality (i.e., in Chile, the income differences are too large) and the other related to the perception of meritocracy (i.e., in Chile, people are rewarded for their effort).
-Both groups were measured using a Likert scale of 5 categories from “strongly disagree” to “strongly agree.”
+The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
+In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
+All these variables were also measured using a Likert scale from "strongly disagree" (1) to "strongly agree" (5).
+Then,  educational level and household income quintiles are included as objective status measures, as well as subjective social status.
+Finally, political identification on a left-right scale, age and gender were included as control variables.
+\@ref(tab:descriptives) shows the items used, their response categories, and their frequency.
 
-The educational level of the respondents, the household's income quintile per capita, their subjective social status, their political identification, and their age and gender were included as control variables.
-\@ref(fig:descriptives) shows the items used, their response categories, and their frequency.
-
-```{r descriptives, echo=FALSE, fig.cap = "Independent variables", fig.align='center'}
- knitr::include_graphics("output/tables/desc02.png")
+```{r descriptives, echo=FALSE, results='asis'}
+kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
+      format="pipe",
+             caption="Independent variables",
+             col.names = "",
+             row.names = F)
 ```
 
-Figure 1 shows the distribution of the dependent variable of Justification of educational inequality and both groups of independent variables for all the years analyzed.
-On the one hand, Figure 1 shows that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, there is a large group of respondents (91.7%) who agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is a significant degree of disagreement (disagree + strongly disagree) about whether people are rewarded for their talent (47.6%)  and for their efforts (57%).
+Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for the dependent and independent variables.
+We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
+Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
+On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
 
@@
 
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/corr.png")
-```
-
-Figure 2 presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between high and moderate values, which indicates a certain level of association between the variables.
-First, the dependent variable has a moderate and negative association with the perception of inequality and a moderate and positive association with both meritocracy perception variables.
-Secondly, the perception of inequality has a moderate and negative association with meritocracy perception variables.
-Finally, the two meritocracy variables present a high and positive association with each other.
-
 ## Methods
 
-Given the hierarchical structure of the data (observations nested in surveys nested in municipalities), we applied a longitudinal multilevel strategy [@singer_applied_2009].
+Given the hierarchical structure of the data (observations nested in survey waves), we applied a longitudinal multilevel strategy [@singer_applied_2009].
 Longitudinal multilevel models are suited to account for the shared variance among units in the data to estimate standard errors better.
@@
 Multilevel models allow a solution in regression estimation by adding a random term that represents the variance associated with the nesting of the data (random effects).
-The linear multilevel models are estimated using the R library “lme4” [@singer_applied_2009, p.4].
+The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
 
@@
 
-Figure 3 illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
+Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
 Each year represents stacked percentual frequencies, and the flows in between represent the within-subjects’ change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, from the 32.2% who strongly disagree with inequality justification in 2016, about half of them keeps responding the same in 2016, whereas the other half shifts their opinion to other response categories.
+For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
 In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
 Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to a half when compared with previous years.
+This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
 
@@
 
-Figure 4 shows the average changes in the main variables considered for this study.
+Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
 Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
 Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables is concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated to talent.
+As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
+
+```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
+ knitr::include_graphics("output/graphs/corr.png")
+```
+
+Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
+In this matrix, the correlations vary between low and moderate values.
+Justification of economic inequality depicts a moderate and negative association with the perception of inequality and a moderate and positive association with both meritocracy perception variables.
+Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables.
+Finally, the two meritocracy variables present a high and positive association with each other.
 
@@
 
-Table 2 shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves in order to observe whether there are intertemporal changes in the dependent variable.
+Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
+Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
 Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
@@
 Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease of inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
+On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
 
 Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women in average justify less inequality in education when compared to males.
+On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
 Educational level and subjective social status are not shown in this table as they do not exhibit significant effects, but the complete estimation of the models is presented in the appendix.
 Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing.
-Inequality perception is added in Model 4, showing a negative association with justification of educational inequality as hypothetisized that remains stable when controlling for the rest of the variables.
+Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized that remains stable when controlling for the rest of the variables.
 Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
@@
 
-```{r multilevel, echo=FALSE, fig.cap = "Multilevel longitudinal models for the justification of inequality in education", fig.align='center'}
- knitr::include_graphics("output/tables/table1-individuales-pond-v2.png")
+```{r multilevel, echo=FALSE, results='asis'}
+kable(data.frame(image='![](output/tables/table1-individuales-pond-v2.png)'),    # generar tabla
+      format="pipe",
+             caption="Multilevel longitudinal models for the justification of inequality in education",
+             col.names = "",
+             row.names = F)
 ```
 
-```{r interact, echo=FALSE, fig.cap = "Interaction effects for the justification of economic inequality", fig.align='center'}
- knitr::include_graphics("output/tables/table3-interac-pond.png")
+```{r interact, echo=FALSE, results='asis'}
+kable(data.frame(image='![](output/tables/table3-interac-pond.png)'),    # generar tabla
+      format="pipe",
+             caption="Interaction effects for the justification of economic inequality",
+             col.names = "",
+             row.names = F)
 ```
 
-In this final part of the analyses we contrast longitudinal hypotheses regarding changes in the relationships between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between perception of meritocracy and inequality justification mitigates along time, as meritocratic ideals could have weaken due to critical situations associated to the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table 2.
+In this final part of the analyses, we contrast longitudinal hypotheses regarding changes in the relationships between meritocracy and justification of economic inequality over time.
+In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
+We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
 Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
 Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable show a significant interaction with the time variable (0.01, p<0.05), meaning that, in average, the association between perception of meritocracy based on effort and the justification of inequality increases in 0.01 points in every measurement point.
+As observed, only the effort variable shows a significant interaction with the time variable (0.01, p<0.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
 
@@
 Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence with regard to our hypothesis.
-Whereas we find support for the association between perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
+In general, our results showed mixed evidence about our hypothesis.
+Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
 We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-To this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is an striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
+In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
 As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises, but a more direct and rational explanation of this phenomenon is difficult to sustain.
-As in the Chilean case the pandemic period occurred along with a constitucional process, disentangling different forces driving public opinion it becomes difficult and deserves more cualitative and quantitative studies.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that lead to the deligitimation of this assembly.
-This could have drive preferences in a conservative direction, which was actually reflected in the election of a right-wing second assembly after a failed first process [@palanza_chile_2023;@sazo_chile_2023].
+As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves more qualitative and quantitative studies.
+Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
+This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after a failed first process [@palanza_chile_2023;@sazo_chile_2023].
 
 Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on that rational interests of the better-off would be consistent with a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes towards inequality has argued that a greater justification of inequality its associated to its perceived magnitude ($H_3$).
+In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off would be consistent with a larger inequality justification.
+Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
 Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
@@
 The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At a first look the results are consistent with such hypothesis, however, there are some nuances attending to the different aspects of meritocracy considered here, namely effort and talent.
-However, conceptually, meritocracy centralizes people's efforts and talent as the main factors for achieving status.
+At first look, the results are consistent with the hypothesis, however, some nuances are attending to the different aspects of meritocracy considered here, namely effort and talent.
+Nevertheless, conceptually, meritocracy centralizes people's efforts and talent as the main factors for achieving status.
 In this regard, our results are favorable for the effort dimension but are not confirmed for the talent dimension.
 In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated to luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
+This could be explained as that talent could be more associated with luck in terms of random assignment, and in this term would not be enough reason (as it is effort) to justify educational inequality.
 
-Finally, as our first hypothesis supported, our expectations were that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy in relation to effort strengthens its relation over time with respect to the justification of inequality in education, which offers evidence against the temporal change hypothesis ($H_5$).
+Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
+However, longitudinal evidence shows that the perception of meritocracy in relation to effort strengthens its relation over time concerning the justification of inequality in education, which offers evidence against the temporal change hypothesis ($H_5$).
 In other words, we know that people who perceive that effort in Chile is rewarded are also more likely to justify that those with higher incomes have better access to education for their children.
@@
 First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of an individual or educational institutions.
+In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
 Second, as has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
@@
 Finally, there are limitations involving the longitudinal dimension of the data used.
-In this regard, the ELSOC had to be modified in order to maintain its stability over time during the pandemic, which implied that the educational justice indicator was excluded in the 2021 measurement.
+In this regard, the ELSOC had to be modified in order to maintain its stability over time during the pandemic, which implied that the educational justice indicator was excluded from the 2021 measurement.
 This implies that the temporal change is not exactly one year in each measurement, with the largest temporal gap being between 2019 and 2022.
-Consequently, the conclusions regarding the temporal change have to be considered carefully, and further data waves could give us more information in this matter.
+Consequently, the conclusions regarding the temporal change have to be considered carefully, and further data waves could give us more information on this matter.
 
@@
 In this regard, thanks to the survey's sampling strategy, it is possible to make contextual estimates at the municipality level.
-Future research could also include individual-level data with contextual information from administrative data sources, allowing for instance the consideration of contextual socioeconomic inequality as wellas its dynamics over time.
-Furthermore, this dataset could also be used in the future for comparing inequality justification in policy areas besides education, such as in health and in pensions.
-Future research may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes and their impacts for solidarity and social cohesion.
+Future research could also include individual-level data with contextual information from administrative data sources, allowing for instance the consideration of contextual socioeconomic inequality as well as its dynamics over time.
+Furthermore, this dataset could also be used in the future for comparing inequality justification in policy areas besides education, such as in health and pensions.
+Future research may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes and their impacts on solidarity and social cohesion.
 

### 2024-03-25 paper.Rmd
second collaborative review on vsc

@@
 On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table as they do not exhibit significant effects, but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized that remains stable when controlling for the rest of the variables.
+Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
+Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
+Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
+
 Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<0.01), however, when controlling for the perception that effort is rewarded, this effect is no longer significant.
+In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<0.01).
+However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
 In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<0.01), but it has a larger weight than the perception of talent.
@@
 
-In this final part of the analyses, we contrast longitudinal hypotheses regarding changes in the relationships between meritocracy and justification of economic inequality over time.
+In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
 In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
@@
 Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypothesis.
+In general, our results showed mixed evidence about our hypotheses.
 Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
@@
 In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises, but a more direct and rational explanation of this phenomenon is difficult to sustain.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves more qualitative and quantitative studies.
+As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
+As in the Chilean case the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
+A closer look to the political environment during the last year in the country could give additional hints.
 Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after a failed first process [@palanza_chile_2023;@sazo_chile_2023].
+This could have driven preferences in a conservative direction, which was actually reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
 
 Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off would be consistent with a larger inequality justification.
+In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
 Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
@@
 The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis, however, some nuances are attending to the different aspects of meritocracy considered here, namely effort and talent.
-Nevertheless, conceptually, meritocracy centralizes people's efforts and talent as the main factors for achieving status.
-In this regard, our results are favorable for the effort dimension but are not confirmed for the talent dimension.
+At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
+Our results are favorable for the effort dimension but not confirmed for the talent dimension.
 In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this term would not be enough reason (as it is effort) to justify educational inequality.
+This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
 
 Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy in relation to effort strengthens its relation over time concerning the justification of inequality in education, which offers evidence against the temporal change hypothesis ($H_5$).
-In other words, we know that people who perceive that effort in Chile is rewarded are also more likely to justify that those with higher incomes have better access to education for their children.
-Additionally, we see the passage of time, including the health crisis generated by the COVID-19 pandemic.
-In this regard, an alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort-, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
+However, longitudinal evidence shows that the perception of meritocracy linked effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
+In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
+An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort-, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
+Again, this certainly needs further studies that include additional specific conceptual and methodological approach to this phenomenon.
 
-Among the limitations of our studies, we can mention at least three.
+Among the limitations of our study, we can mention at least three.
 First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
 In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restrict us to the use of single indicators for effort and talent, and further research should improve the measurement quality in this regard.
+Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
+In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
 Finally, there are limitations involving the longitudinal dimension of the data used.
-In this regard, the ELSOC had to be modified in order to maintain its stability over time during the pandemic, which implied that the educational justice indicator was excluded from the 2021 measurement.
-This implies that the temporal change is not exactly one year in each measurement, with the largest temporal gap being between 2019 and 2022.
-Consequently, the conclusions regarding the temporal change have to be considered carefully, and further data waves could give us more information on this matter.
+Given limitations for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately the educational justice indicator was excluded in that survey.
+Consequently, the results about temporal change have to be considered carefully, and further data waves could give us more information in this matter.
 
 The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the survey's sampling strategy, it is possible to make contextual estimates at the municipality level.
-Future research could also include individual-level data with contextual information from administrative data sources, allowing for instance the consideration of contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could also be used in the future for comparing inequality justification in policy areas besides education, such as in health and pensions.
-Future research may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes and their impacts on solidarity and social cohesion.
+In this regard, thanks to the survey's sampling strategy, it would be possible to make contextual estimates at the municipality level in future studies.
+For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
+Furthermore, this dataset could  be used for comparing inequality justification in policy areas besides education, such as in health and pensions.
+Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
 

### 2024-03-25 paper.Rmd
last render

@@
   email: juancastillov@uchile.cl
-  number: 1
+  number: 1, 3 
 - name: Julio Iturra
@@
 abstract: | 
+  Education is considered a key tool for social mobility and equality of opportunities.
+However, despite the widespread appreciation of education, disparities in educational outcomes persist over time, linked to variables such as families socioeconomic status, gender, ethnic origin, and migratory background, among others.
+While it is often assumed that there is a consensus on equal access to educational opportunities, empirical research on distributive justice teaches us that this assumption is far from reality.
+A significant body of evidence in this area indicates a substantial degree of acceptance and even justification of social inequalities, with education being no exception.
+In this regard, meritocracy emerges as one of the central concepts in the educational context related to justifying inequalities.
+As meritocracy emphasizes individual factors such as effort and talent in individual performance, the role of opportunities and social background is usually sidelined, leading to justifying inequalities of outcomes that perpetuate systemic disadvantages for marginalized groups.
+This research focuses on the changes in the justification of inequalities in access to education in Chile and how this justification is associated with the perception of meritocracy.
+The Chilean case offers an interesting context for this study due to its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
+The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic).
+This change could be associated with a challenge of meritocratic ideals.
+Furthermore, we suggest that the association between a low meritocratic perception and less inequality justification would be stronger among traditionally vulnerable groups, such as individuals from lower social classes, less educated, and women.
+We analyze the ELSOC representative panel survey (2016 – 2022, 6 waves, N = 2,927), a study that includes various social, political, and economic topics during the last years.
+Estimating the statistical longitudinal models (multilevel random effects) shows mixed evidence about our hypotheses.
+Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
+In addition, individuals in more advantaged situations and with greater resources tend to defend their interests and justify greater education inequality.
+Throughout the years, a larger perceived economic inequality has motivated a lower justification of inequality in education.
   .
@@
 
-   **Keywords**: meritocracy, social inequality, inequality justification
+   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
 output:
@@
     template: null
-    number_sections: false    
+    number_sections: true    
     toc: false
@@
 always_allow_html: true # for word output
-fontsize: 10pt
+fontsize: 12pt
 papersize: a4
@@
 
-<!---
 \pagebreak
@@
 
-The authors recognize the support of ANID/FONDECYT grant 1160921 and of  the Center for Conflict and Social Cohesion Studies (COES)  grant ANID/FONDAP 1513009.
+The authors recognize the support of ANID/FONDECYT grant 1160921; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046
 
 \pagebreak
-
---->
 
@@
 Finally, political identification on a left-right scale, age and gender were included as control variables.
-\@ref(tab:descriptives) shows the items used, their response categories, and their frequency.
+Table \@ref(tab:descriptives) shows the items used, their response categories, and their frequency.
 
 ```{r descriptives, echo=FALSE, results='asis'}
+pacman::p_load(kableExtra)
 kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
@@
 ```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/figure1.png")
+ knitr::include_graphics("output/graphs/merit_plot.png")
 ```
@@
 ```{r multilevel, echo=FALSE, results='asis'}
-kable(data.frame(image='![](output/tables/table1-individuales-pond-v2.png)'),    # generar tabla
-      format="pipe",
-             caption="Multilevel longitudinal models for the justification of inequality in education",
-             col.names = "",
-             row.names = F)
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
+
+# generate analytical sample
+df_study1 <- 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
+  na.omit() %>% 
+  mutate(ola = as.factor(ola),
+         ola_num = as.numeric(ola),
+         ola_2=as.numeric(ola)^2,
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h0 <- "ola"
+h1 <- "ola_num+ola_2"
+h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
+h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
+h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
+h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
+h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+
+# Table 1: individual
+
+omit <- "(education)|(ess)"
+texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
+Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
+                omit.coef = omit,
+                custom.coef.names = c("Intercept",
+                                      "Wave 2017",
+                                      "Wave 2018",
+                                      "Wave 2019",
+                                      "Wave 2022",
+                                      "Wave",
+                                      "Wave^2",
+#                                      "High School <br> <i>(Ref. Primary)</i>",
+#                                      "Technical",
+#                                      "Universitary",
+                                      "Quintile 2",
+                                      "Quintile 3",
+                                      "Quintile 4",
+                                      "Quintile 5",
+                                      "Quintile NA",
+#                                      "Subjective Social Status",
+                                      "Age 30-49",
+                                      "Age 50-64",
+                                      "Age 65 or more",
+                                      "Gender (Ref. Male)",
+                                      "Center",
+                                      "Right",
+                                      "Does not identify",
+                                      "Inequality perception",
+                                      "Merit: Talent",
+                                      "Merit: Effort" ),
+                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
+                  "Quintile (Ref.= Quintile 1)" = 8:12,
+                  "Age (Ref.= 18-29)" = 13:15,
+                  "Pol. pos (Ref.= Left)" = 17:19
+                  ),
+                include.loglik = FALSE,
+                include.aic = FALSE,
+                custom.gof.names = c("BIC",
+                                     "Num. obs.",
+                                     "Num. groups: Individuals",
+                                     "Var: Individuals (Intercept)",
+                                     "Var: Residual"),
+               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
+               center = TRUE,
+               caption.above = TRUE,
+               scalebox = 0.765,
+               float.pos = "!ht", 
+               booktabs = FALSE, 
+               use.packages = FALSE, 
+               include.rs = FALSE, 
+               include.rmse = FALSE,
+               include.ci = FALSE)
+
 ```
@@
 ```{r interact, echo=FALSE, results='asis'}
-kable(data.frame(image='![](output/tables/table3-interac-pond.png)'),    # generar tabla
-      format="pipe",
-             caption="Interaction effects for the justification of economic inequality",
-             col.names = "",
-             row.names = F)
+df_study1 <- 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
+  na.omit() %>% 
+  mutate(ola = as.numeric(ola),
+         ola_2=as.numeric(ola)^2,
+         sexo=as.factor(sexo)
+ #        edad=as.numeric(edad)
+         #pos_id=as.numeric(pos_id)
+         )
+
+# Hipotesis
+
+h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
+h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
+
+# A. justification educational inequality
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+
+# table 3: interacciones
+omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
+texreg::texreg(list(educ6,educ3,educ1),
+                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
+Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, <br> political position and inequality perception",
+                omit.coef = omit,
+                include.loglik = FALSE,
+                include.aic = FALSE,
+                custom.coef.names = c("Intercept",
+                                      "Wave",
+                                      "Wave^2",
+                                      "Merit: Talent",
+                                      "Merit: Effort",
+                                      "Wave * Merit: Talent",
+                                      "Wave * Merit: Effort"
+                                      ),
+                custom.gof.names = c("BIC",
+                                     "Num. obs.",
+                                     "Num. groups: Individuals",
+                                     "Var: Individuals (Intercept)",
+                                     "Var: Residual",
+                                     "Var: Individuals Wave",
+                                     "Var: Individuals Merit: Talent",
+                                     "Cov: Individuals (Intercept) Wave",
+                                     "Cov: Individuals (Intercept) Merit: Talent",
+                                     "Cov: Individuals Wave Merit: Talent",
+                                     "Var: Individuals Merit: Effort",
+                                     "Cov: Individuals (Intercept) Merit: Effort",
+                                     "Cov: Individuals Wave Merit: Effort"),  
+               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
+               center = TRUE,
+               caption.above = TRUE,
+               scalebox = 0.765,
+               float.pos = "!ht", 
+               booktabs = TRUE, 
+               use.packages = FALSE, 
+               include.rs = FALSE, 
+               include.rmse = FALSE,
+               include.ci = FALSE,
+               threeparttable = TRUE)
 ```
@@
 
-# Appendix
+::: {#refs}
+:::
 
-```{r multilevel-full, echo=FALSE, fig.cap = "Multilevel effects of justification of economic inequality", fig.align='center'}
- knitr::include_graphics("output/tables/table1-individuales-pond-appendix.png")
+# (APPENDIX) Appendix {.unnumbered}
+
+`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
+
+```{r multilevel-full, echo=FALSE, results='asis'}
+df_study1 <- 
+  df_study_long_t6_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
+  na.omit() %>% 
+  mutate(ola = as.factor(ola),
+         ola_num = as.numeric(ola),
+         ola_2=as.numeric(ola)^2,
+         sexo=as.factor(sexo)
+         )
+# Hipotesis
+h0 <- "ola"
+h1 <- "ola_num+ola_2"
+h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
+h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
+h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
+h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
+h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+
+# A. Education distributive justice
+educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+
+# Table 1: individual
+
+omit <- "(education)|(ess)"
+texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
+#                omit.coef = omit,
+                custom.coef.names = c("Intercept",
+                                      "Wave 2017",
+                                      "Wave 2018",
+                                      "Wave 2019",
+                                      "Wave 2022",
+                                      "Wave",
+                                      "Wave^2",
+                                      "High School",
+                                      "Technical",
+                                      "Universitary",
+                                      "Quintile 2",
+                                      "Quintile 3",
+                                      "Quintile 4",
+                                      "Quintile 5",
+                                      "Quintile NA",
+                                      "Subjective Social Status",
+                                      "Age 30-49",
+                                      "Age 50-64",
+                                      "Age 65 or more",
+                                      "Gender (Ref. Male)",
+                                      "Center",
+                                      "Right",
+                                      "Does not identify",
+                                      "Inequality perception",
+                                      "Merit: Talent",
+                                      "Merit: Effort" ),
+                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
+                              "Education (Ref.=Primary)"= 8:10,
+                  "Quintile (Ref.= Quintile 1)" = 11:15,
+                  "Age (Ref.= 18-29)" = 17:19,
+                  "Pol. pos (Ref.= Left)" = 21:23
+                  ),
+                include.loglik = FALSE,
+                include.aic = FALSE,
+                custom.gof.names = c("BIC",
+                                     "Num. obs.",
+                                     "Num. groups: Individuals",
+                                     "Var: Individuals (Intercept)",
+                                     "Var: Residual"),
+               caption = paste("(\\#tab:multilevel-full)","Multilevel effects of justification of economic inequality"),
+               center = TRUE,
+               caption.above = TRUE,
+               scalebox = 0.73,
+               float.pos = "!ht", 
+               booktabs = FALSE, 
+               use.packages = FALSE, 
+               include.rs = FALSE, 
+               include.rmse = FALSE,
+               include.ci = FALSE)
+
 ```

### 2024-03-25 production/prod_analysis-fix-t5.Rmd
last render

@@
 
-df<- dfSummary(x = dplyr::select(df_study1, ola, just_educ, des_perc, merit_talent, merit_effort, educ, quintil1, ess, pos_id, edad, sexo),
+df<- dfSummary(x = dplyr::select(df_study1, merit_talent, merit_effort, des_perc, educ, quintil1, ess, edad, sexo, pos_id),
                plain.ascii = FALSE,

### 2024-03-25 paper.Rmd
arreglo footnote tab3

@@
                 custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, <br> political position and inequality perception",
+Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
                 omit.coef = omit,

### 2024-03-26 paper.Rmd
correcciones varias

@@
   Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread appreciation of education, disparities in educational outcomes persist over time, linked to variables such as families socioeconomic status, gender, ethnic origin, and migratory background, among others.
-While it is often assumed that there is a consensus on equal access to educational opportunities, empirical research on distributive justice teaches us that this assumption is far from reality.
-A significant body of evidence in this area indicates a substantial degree of acceptance and even justification of social inequalities, with education being no exception.
-In this regard, meritocracy emerges as one of the central concepts in the educational context related to justifying inequalities.
-As meritocracy emphasizes individual factors such as effort and talent in individual performance, the role of opportunities and social background is usually sidelined, leading to justifying inequalities of outcomes that perpetuate systemic disadvantages for marginalized groups.
-This research focuses on the changes in the justification of inequalities in access to education in Chile and how this justification is associated with the perception of meritocracy.
-The Chilean case offers an interesting context for this study due to its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic).
-This change could be associated with a challenge of meritocratic ideals.
-Furthermore, we suggest that the association between a low meritocratic perception and less inequality justification would be stronger among traditionally vulnerable groups, such as individuals from lower social classes, less educated, and women.
-We analyze the ELSOC representative panel survey (2016 – 2022, 6 waves, N = 2,927), a study that includes various social, political, and economic topics during the last years.
-Estimating the statistical longitudinal models (multilevel random effects) shows mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-In addition, individuals in more advantaged situations and with greater resources tend to defend their interests and justify greater education inequality.
-Throughout the years, a larger perceived economic inequality has motivated a lower justification of inequality in education.
+However, despite the widespread value of education, disparities in educational outcomes persist over time.
+To what extent are such educational disparities justified in society?
+And, what are the main factors driving this kind of justification?
+This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
+The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
+The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
+For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
+Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
+The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
+
   .
@@
 ```
-
-\pagebreak
-
-# Acknowledgments
-
-The authors recognize the support of ANID/FONDECYT grant 1160921; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046
 
@@
 
+\pagebreak
+
+_Acknowledgments_
+
+The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046
+
 # References

### 2024-03-26 paper.Rmd
typos, labels in proofread

@@
 
-Theoretically, @liebigSociologyJustice2016 establishes three fundamental distinctions in justice attitudes in the field of sociological justice research.
+ Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
 First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
 Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ related primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
+Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
 Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
@@
 Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are ISSP (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the ESS 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
+Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
 Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
@@
 Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the undersetimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
+Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
 
@@
 The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
+The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
 Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
 However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
+This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
 
@@
 
-Finally, and in line with hypothesis 1 and 4, we explore interactions between meritocracy and inequality justification over time.
+Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
 As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
@@
 The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016,2017,2018,2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, in 2021 a reduced version was conducted using computer-assisted telephone interviews (CATI).
+The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
+However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
 In addition, wave 3 included a refreshment sample to counter survey attrition.
@@
 Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between Wave 1 and Wave 6 is 41%, achieving a final sample of N = 1728.
+In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
 The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
@@
 Then,  educational level and household income quintiles are included as objective status measures, as well as subjective social status.
-Finally, political identification on a left-right scale, age and gender were included as control variables.
+Finally, political identification on a left-right scale, age, and gender were included as control variables.
 Table \@ref(tab:descriptives) shows the items used, their response categories, and their frequency.
@@
 Given the hierarchical structure of the data (observations nested in survey waves), we applied a longitudinal multilevel strategy [@singer_applied_2009].
-Longitudinal multilevel models are suited to account for the shared variance among units in the data to estimate standard errors better.
+In this matter, longitudinal multilevel models are suited to account for the shared variance among units in the data to estimate standard errors better as observations are nested within respondents.
 Since individuals share variance within themselves over time, if the error structure is not considered, it would be as if they were considered different individuals.
-Multilevel models allow a solution in regression estimation by adding a random term that represents the variance associated with the nesting of the data (random effects).
-The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
+Multilevel models allow a solution in regression estimation by adding a random term that represents the variance associated with the nested structure of the data (random effects).
+Consequently, the linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
 
@@
 Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between represent the within-subjects’ change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
+Each year represents stacked percentual frequencies, and the flows in between represent the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
 For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
@@
 In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality and a moderate and positive association with both meritocracy perception variables.
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables.
-Finally, the two meritocracy variables present a high and positive association with each other.
+Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
+Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
+Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
 
@@
 Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<0.01).
+In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
 However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<0.01), but it has a larger weight than the perception of talent.
+In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
 
@@
                                       "Quintile 5",
-                                      "Quintile NA",
+                                      "Quintile (No information)",
 #                                      "Subjective Social Status",
@@
                 groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Quintile (Ref.= Quintile 1)" = 8:12,
+                  "Household Income (Ref.= Quintile 1)" = 8:12,
                   "Age (Ref.= 18-29)" = 13:15,
@@
 Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable (0.01, p<0.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
+As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
 
@@
 As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
+As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
 A closer look to the political environment during the last year in the country could give additional hints.
@@
 In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort-, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approach to this phenomenon.
+An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
+Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
 
@@
 Finally, there are limitations involving the longitudinal dimension of the data used.
-Given limitations for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately the educational justice indicator was excluded in that survey.
-Consequently, the results about temporal change have to be considered carefully, and further data waves could give us more information in this matter.
+Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
+Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
 
 The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the survey's sampling strategy, it would be possible to make contextual estimates at the municipality level in future studies.
+In this regard, thanks to the sampling strategy the survey, it would be possible to make contextual estimates at the municipality level in future studies.
 For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could  be used for comparing inequality justification in policy areas besides education, such as in health and pensions.
+Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
 Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
@@
 
+\newpage 
 # (APPENDIX) Appendix {.unnumbered}
@@
                                       "Quintile 5",
-                                      "Quintile NA",
+                                      "Quintile (No information)",
                                       "Subjective Social Status",
@@
                               "Education (Ref.=Primary)"= 8:10,
-                  "Quintile (Ref.= Quintile 1)" = 11:15,
+                  "Household Income (Ref.= Quintile 1)" = 11:15,
                   "Age (Ref.= 18-29)" = 17:19,
@@
                                      "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel effects of justification of economic inequality"),
+               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
                center = TRUE,
                caption.above = TRUE,
-               scalebox = 0.73,
+               scalebox = 0.60,
                float.pos = "!ht", 

### 2024-03-26 paper.Rmd
varias revisiones

@@
 
-# Inequality beliefs and the justification of economic inequality
+# Inequality, beliefs and justification
 
@@
 For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
+The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.
 
@@
 
-The main variable of this study is the justification of educational inequality.
-This item is measured using the following statement:"It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree.” (5)
-
+The dependent variable of this study is the justification of educational inequality.
+This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
+Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
 The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
 In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-All these variables were also measured using a Likert scale from "strongly disagree" (1) to "strongly agree" (5).
-Then,  educational level and household income quintiles are included as objective status measures, as well as subjective social status.
+Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
+We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
+Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
+On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
+
+```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
+ knitr::include_graphics("output/graphs/merit_plot.png")
+```
+
+For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
 Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the items used, their response categories, and their frequency.
+Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
 
@@
 
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for the dependent and independent variables.
-We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
-
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/merit_plot.png")
-```
-
 ## Methods
 
-Given the hierarchical structure of the data (observations nested in survey waves), we applied a longitudinal multilevel strategy [@singer_applied_2009].
-In this matter, longitudinal multilevel models are suited to account for the shared variance among units in the data to estimate standard errors better as observations are nested within respondents.
-Since individuals share variance within themselves over time, if the error structure is not considered, it would be as if they were considered different individuals.
-Multilevel models allow a solution in regression estimation by adding a random term that represents the variance associated with the nested structure of the data (random effects).
-Consequently, the linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
+Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models [@singer_applied_2009].
+Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
+The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
 
@@
 Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between represent the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
+Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
 For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
@@
 Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was actually reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
+This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
 
@@
 Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
+However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
 In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
@@
 The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy the survey, it would be possible to make contextual estimates at the municipality level in future studies.
+In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
 For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.

### 2024-03-26 paper.tex
varias revisiones

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  12pt,
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic},
%   colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{times}
\hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}
\usepackage{caption}
\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic}
    \usepackage[]{authblk}
    \setlength{\affilsep}{2em}
    \renewcommand*{\Affilfont}{\normalsize\normalfont}
    \renewcommand\Authsep{, }
    \renewcommand\Authand{ \& }
    \renewcommand\Authands{, \& }
                                        \author[1, 3]{Juan Carlos Castillo}
                                                            \affil[]{Department of Sociology, Universidad de Chile \thanks{\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}}}
                                                                                \author[2]{Julio Iturra}
                                                            \affil[]{Bremen International Graduate School of Social Sciences, Universität Bremen}
                                                                                \author[3]{Kevin Carrasco}
                                                            \affil[]{Center for Social Conflict and Cohesion Studies}
                                            \date{}

\providecommand{\keywords}[1]
{
  \small
  \textbf{Keywords:} #1
}

\begin{document}
\maketitle
\begin{abstract}
Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. To what extent are such educational disparities justified in society? And, what are the main factors driving this kind of justification? This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy. The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.

.
\newline
\newline

\textbf{Keywords}: meritocracy, social inequality, inequality justification, COVID-19
\end{abstract}

\setstretch{1.5}
\pagebreak

\section{Introduction}\label{introduction}

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA Report points out, ``Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries'' (\citeproc{ref-oecd_pisa_2023}{OECD, 2023}). Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 (\citeproc{ref-darmodyImpactsCOVID19Control2021}{Darmody et al., 2021}). Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education (\citeproc{ref-leeFairnessPerceptionsEducational2023}{Lee \& Stacey, 2023}). In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.

Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background. In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed (\citeproc{ref-hadjar_meritokratie_2008}{Hadjar, 2008}; \citeproc{ref-mcnamee_meritocracy_2004}{McNamee \& Miller, 2004}; \citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.

The Chilean case is particularly relevant when it comes to the study of inequalities in education. The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system (\citeproc{ref-bellei_estudio_2013}{Bellei, 2013}; \citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}). The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation (\citeproc{ref-joikomujicaCuasimercadoEducativoChile2019}{Joiko Mujica, 2019}). Previous research in Chile has emphasized the concept of an ``educational market'' (\citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}), characterized by selection, competition, and the provision of goods and services associated with payment capacities. Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences (\citeproc{ref-ramosarellanoEducacionPromesaIncumplida2022}{Ramos Arellano, 2022}). Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}; \citeproc{ref-hankivsky_introduction_2022}{Hankivsky et al., 2022}). A pertinent aspect of this discourse is articulated in the works of Mijs (\citeproc{ref-mijs_stratified_2016}{2016}, \citeproc{ref-mijs_visualizing_2018}{2018}), showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals. Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.

\section{Inequality, beliefs and justification}\label{inequality-beliefs-and-justification}

The justification of economic inequalities is a controversial field of study. As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable. Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as ``the defensibility of unequal relations between people'' (\citeproc{ref-barry_theories_1989}{1989, p. 3}), whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification (\citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}; \citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-trump_income_2018}{Trump, 2018}). Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.

The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research (\citeproc{ref-Resh2016}{Resh \& Sabbagh, 2016}). From a sociological point of view, the works of Kluegel \& Smith (\citeproc{ref-kluegelBeliefsStratification1981}{1981}) can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice. Theoretically, the authors understand the concept of ``beliefs'' in a ``broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action.'' (\citeproc{ref-kluegelBeliefsStratification1981}{Kluegel \& Smith, 1981, p. 30}). However, the literature has strained the concept of ``beliefs'' by more specific distinctions around the individual and structural dimensions of economic inequality. In this regard, Wegener (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{1987}) has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987, p. 2}). This approach established that status positions influence how people perceive things \emph{are} and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust (\citeproc{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}). Additionally, Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}) has suggested more detailed conceptual distinctions about how people perceive economic inequality \emph{is}, how they believe or prefer it \emph{should be}, and the process of evaluating such differences as fair.

Theoretically, Liebig \& Sauer (\citeproc{ref-liebigSociologyJustice2016}{2016}) establish three fundamental distinctions in justice attitudes in the field of sociological justice research. First, \emph{order-related attitudes} refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement. Second, \emph{procedural justice attitudes} account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities). Finally, they can be \emph{outcome-related attitudes} which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes). Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles (\citeproc{ref-kluegel_social_1995}{Kluegel et al., 1995}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011})

An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{Janmaat, 2013}; \citeproc{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \& Smith, 1986}). The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries (\citeproc{ref-Evans1992}{Evans et al., 1992}; \citeproc{ref-Evans2017}{Evans \& Kelley, 2017}) and greater justification of salary gaps (\citeproc{ref-hadler_why_2005}{Hadler, 2005}; \citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}). Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}; \citeproc{ref-vandertoorn_twenty_2014}{van der Toorn \& Jost, 2014}). Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies (\citeproc{ref-trump_when_2020}{Trump, 2020}). As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position (\citeproc{ref-jost_theory_2020}{Jost, 2020}).

Another factor that has been linked to inequality beliefs is the perception of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-kelleyLegitimationInequalityOccupational1993}{Kelley \& Evans, 1993}; \citeproc{ref-osberg_fair_2006}{Osberg \& Smeeding, 2006}). Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987}, \citeproc{ref-Wegener1990}{1990}). Thus, both the role of social status and economic inequality perceptions have found support in correlational (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012}; \citeproc{ref-schroder_income_2017}{Schröder, 2017}), as well as in experimental studies that have manipulated inequality perception (\citeproc{ref-iturra_percepcion_2023}{Iturra et al., 2023}; \citeproc{ref-trump_income_2018}{Trump, 2018}). In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality. For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification (\citeproc{ref-Castillo2011}{Castillo, 2011}). Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences (\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De Carvalho Galvão, 2022}). Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.

\section{The justification of educational inequality}\label{the-justification-of-educational-inequality}

Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences. A strand of this research has focused on specific policy areas such as health, pensions, and education, attending to the degree of justification of market-based access to social services. For instance, Lindh (\citeproc{ref-lindhPublicOpinionMarkets2015}{2015}) found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services. Similarly, Busemeyer \& Iversen (\citeproc{ref-busemeyer_welfare_2020}{2020}) shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes. Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.

The study of inequality justification in the sphere of education is a rather specific and unattended field. Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities. This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area. This paradox is phrased by Conforti (\citeproc{ref-confortiLegitimationInequalityAmerican1992}{1992}) as ``while equality is embraced as a value, inequality is accepted as a reality'' (p.230).

In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are the International Social Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that people with higher incomes can buy better education for their children than people with lower incomes?}), and the European Social Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I have had a fair chance of achieving the level of education I was seeking}). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects. The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.

Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies. Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about---and more supportive of proposals to close---wealth-based achievement gaps than Black-White or Hispanic-White gaps (\citeproc{ref-valantPoliticsAchievementGaps2016}{Valant \& Newark, 2016}). Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification (\citeproc{ref-stoilovaFairnessEducationalOpportunities2023}{Stoilova \& Ilieva-Trichkova, 2023}). However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.

The perception of economic inequality is one of the factors associated with the justification of educational inequality. As pointed out by Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}), justifications are usually confused under the labels of ``beliefs'' and ``attitudes,'' and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011}). For instance, a recent study by Day \& Norton (\citeproc{ref-dayPerceivedIdealInequality2023}{2023}) on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman, 2018})

Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles. In an experiment by Igliozzi et al. (\citeproc{ref-igliozziFairShareEffects2024}{2024}), they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}) is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance (`Students who place within the highest 50\% of testing scores from the previous year can attend the magnet school'). In the same line, Lee \& Stacey (\citeproc{ref-leeFairnessPerceptionsEducational2023}{2023}) show that performance assessments (in what they call ``neoliberal orientations'') predicted people's fair perceptions of educational inequality.

\emph{Justificacion, inequality and meritocracy in the educational domain}

The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by the sociologist Michael Young in his satirical work ``The Rise of the Meritocracy'' (\citeproc{ref-young_rise_1962}{1962}), has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities (\citeproc{ref-bell_politics_2020}{Bell, 2020}). This would contribute to the justification of inequality in the educational domain (\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}), as individual ability is often inherited, and educational attainment is influenced by family background (\citeproc{ref-goldin_meritocracy_2000}{Goldin, 2000}).

Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs. Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences (\citeproc{ref-McCoy2007}{McCoy \& Major, 2007}). This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes. However, empirical findings also reveal nuances in this relationship. For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities (\citeproc{ref-dayPerceivedIdealInequality2023}{Day \& Norton, 2023}). This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.

\emph{This study}

The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification. As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}). Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period. Consistently, our first hypothesis is:

\(H_1\): The justification of educational inequalities decreases after the pandemic

Our second argument is based on the evidence that relates rational interests with justifications and preferences (\citeproc{ref-hing_failure_2019}{Hing et al., 2019}), expecting that:

\(H_2\): Those with lower status characteristics show lower justification of educational inequalities

Thirdly, we address inequality perception. As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized. As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:

\(H_3\): A larger inequality perception is negatively associated with the justification of educational inequalities.

Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:

\(H_4\): Higher meritocratic perceptions are positively related to the justification of educational inequalities.

Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time. As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged. As a consequence, the justification of inequality could show a decrease over time:

\(H_5\): The positive association between meritocracy and inequality justification mitigates along time

\section{Data, variables \& methods}\label{data-variables-methods}

\subsection{Data}\label{data}

The main data source is the Chilean Longitudinal Social Survey 2016--2022. The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile. The sampling design is probabilistic, stratified, clustered, and multistage. It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities. Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1. In addition, the sample represents 77\% of the total national population and 93\% of the urban areas (\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}).

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI). However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021. In addition, wave 3 included a refreshment sample to counter survey attrition. The same sampling strategy as that of wave 1 was implemented for selecting new cases. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends. Regarding the original sample, the response rate was 62.4\% in wave 1, achieving N = 2,927 participants. In broader terms, the accumulated attrition between wave 1 and wave 6 is 41\%, achieving a final sample of N = 1728. The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit \url{https://coes.cl/encuesta-panel/}.

\subsection{Variables}\label{variables}

The dependent variable of this study is the justification of educational inequality. This construct is measured using the following statement: ``It is just that high-income people have a better education for their children than people with lower incomes'' (``Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos'' in Spanish). Here, the respondents declared their preferences on a Likert scale from ``strongly disagree'' (1) to ``strongly agree'' (5). The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort (``In Chile, people are rewarded for their efforts''), and the other related to talent (``In Chile, people are rewarded for their intelligence and skills''). In addition, we included an indicator for economic inequality perception (``In Chile, the income differences are too large''). Figure \ref{fig:frecuencies} shows the average distribution throughout the years for these variables. We observe that there is a large group (88.2\%) of respondents who disagree or strongly disagree that people with higher incomes can access better education. Likewise, a large group of respondents (91.7\%) agree or strongly agree that income differences in Chile are too large. On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6\%) and for their efforts (57\%).

\begin{figure}[H]

{\centering \includegraphics[width=0.85\linewidth]{output/graphs/merit_plot}

}

\caption{Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:frecuencies}
\end{figure}

For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status. Finally, political identification on a left-right scale, age, and gender were included as control variables. Table \ref{tab:descriptives} shows the whole set of independent variables, the corresponding items, response categories, and frequencies.

\begin{longtable}[]{@{}l@{}}
\caption{\label{tab:descriptives}Independent variables}\tabularnewline
\toprule\noalign{}
\endfirsthead
\endhead
\bottomrule\noalign{}
\endlastfoot
\includegraphics{output/tables/desc02.png} \\
\end{longtable}

\subsection{Methods}\label{methods}

Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009}). Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors. The linear multilevel models are estimated using the R library ``lme4'' (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009, p. 4}).

\section{Results}\label{results}

\begin{figure}[H]

{\centering \includegraphics[width=0.85\linewidth]{output/graphs/alluvial_dep}

}

\caption{Change in the justification of educational inequality over time (2016-2022)}\label{fig:alluvial}
\end{figure}

Figure \ref{fig:alluvial} illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022. Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data (\citeproc{ref-rosvall_mapping_2010}{Rosvall \& Bergstrom, 2010}). For instance, of the 32.2\% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories. In general, the large majority - between 80 and 90\% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave. This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.

\begin{figure}[H]

{\centering \includegraphics[width=0.85\linewidth]{output/graphs/years_plot}

}

\caption{Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years}\label{fig:means}
\end{figure}

Figure \ref{fig:means} shows the average changes in the main variables considered for this study. Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception. Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases. As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.

\begin{figure}[H]

{\centering \includegraphics[width=0.85\linewidth]{output/graphs/corr}

}

\caption{Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:correlation}
\end{figure}

Figure \ref{fig:correlation} presents a correlation matrix of the main variables analyzed, using data from all survey waves. In this matrix, the correlations vary between low and moderate values. Justification of economic inequality depicts a moderate and negative association with the perception of inequality (\emph{r}=-0.29, p\textless.01) and a moderate and positive association with both meritocracy perception variables (\emph{r}=0.18, p\textless.01; \emph{r}=0.15, p\textless.01). Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (\emph{r}=-0.15, p\textless.01; \emph{r}=-0.10, p\textless.01). Finally, the two meritocracy variables present a high and positive association with each other (\emph{r}=0.74, p\textless.01).

\subsection{Multivariate analysis}\label{multivariate-analysis}

Table \ref{tab:multilevel} shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 (\(\beta\)=-0.15, p\textless.001), 2018 (\(\beta\)=-0.06, p\textless.05), and 2019 (\(\beta\)=-0.07, p\textless.01). Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality (\(\beta\)=0.21, p\textless.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.

Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality. On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males. Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix. Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented. Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.

Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 (\(\beta\)=0.04, p\textless.01). However, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality (\(\beta\)=0.05, p\textless.01), but it has a larger weight than the perception of talent.

\begin{table}[!ht]
\caption{\label{tab:multilevel} Multilevel longitudinal models for the justification of inequality in education}
\begin{center}
\scalebox{0.765}{
\begin{tabular}{l c c c c c c}
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\
\hline
Intercept                           & $1.86^{***}$  & $1.98^{***}$  & $1.72^{***}$  & $2.22^{***}$  & $2.11^{***}$  & $2.04^{***}$  \\
                                    & $(0.02)$      & $(0.04)$      & $(0.09)$      & $(0.10)$      & $(0.10)$      & $(0.10)$      \\
Wave (Ref.= Wave 2016)              &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Wave 2017                     & $-0.15^{***}$ &               &               &               &               &               \\
                                    & $(0.02)$      &               &               &               &               &               \\
\quad Wave 2018                     & $-0.06^{*}$   &               &               &               &               &               \\
                                    & $(0.03)$      &               &               &               &               &               \\
\quad Wave 2019                     & $-0.07^{**}$  &               &               &               &               &               \\
                                    & $(0.03)$      &               &               &               &               &               \\
\quad Wave 2022                     & $0.21^{***}$  &               &               &               &               &               \\
                                    & $(0.02)$      &               &               &               &               &               \\
Wave                                &               & $-0.17^{***}$ & $-0.17^{***}$ & $-0.13^{***}$ & $-0.14^{***}$ & $-0.14^{***}$ \\
                                    &               & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      \\
Wave$^2$                            &               & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  \\
                                    &               & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
Household Income (Ref.= Quintile 1) &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Quintile 2                    &               &               & $0.07$        & $0.07$        & $0.07$        & $0.07$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Quintile 3                    &               &               & $0.08$        & $0.08$        & $0.08$        & $0.08$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Quintile 4                    &               &               & $0.12^{*}$    & $0.13^{*}$    & $0.13^{*}$    & $0.13^{*}$    \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
\quad Quintile 5                    &               &               & $0.19^{**}$   & $0.19^{**}$   & $0.19^{***}$  & $0.20^{***}$  \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
\quad Quintile (No information)     &               &               & $0.17^{*}$    & $0.17^{*}$    & $0.18^{*}$    & $0.18^{*}$    \\
                                    &               &               & $(0.08)$      & $(0.08)$      & $(0.08)$      & $(0.08)$      \\
Age (Ref.= 18-29)                   &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Age 30-49                     &               &               & $0.02$        & $0.02$        & $0.02$        & $0.02$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Age 50-64                     &               &               & $0.07$        & $0.08$        & $0.08$        & $0.07$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Age 65 or more                &               &               & $0.19^{**}$   & $0.20^{**}$   & $0.19^{**}$   & $0.18^{**}$   \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
Gender (Ref. Male)                  &               &               & $-0.08^{*}$   & $-0.08^{*}$   & $-0.08^{*}$   & $-0.07^{*}$   \\
                                    &               &               & $(0.03)$      & $(0.03)$      & $(0.03)$      & $(0.03)$      \\
Pol. pos (Ref.= Left)               &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Center                        &               &               & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Right                         &               &               & $0.29^{***}$  & $0.28^{***}$  & $0.27^{***}$  & $0.27^{***}$  \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Does not identify             &               &               & $0.15^{***}$  & $0.15^{***}$  & $0.14^{***}$  & $0.14^{***}$  \\
                                    &               &               & $(0.04)$      & $(0.04)$      & $(0.04)$      & $(0.04)$      \\
Inequality perception               &               &               &               & $-0.12^{***}$ & $-0.12^{***}$ & $-0.12^{***}$ \\
                                    &               &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
Merit: Talent                       &               &               &               &               & $0.04^{***}$  & $0.01$        \\
                                    &               &               &               &               & $(0.01)$      & $(0.01)$      \\
Merit: Effort                       &               &               &               &               &               & $0.05^{***}$  \\
                                    &               &               &               &               &               & $(0.01)$      \\
\hline
BIC                                 & $30911.29$    & $30907.80$    & $31053.70$    & $30942.30$    & $30940.78$    & $30936.16$    \\
Num. obs.                           & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        \\
Num. groups: Individuals            & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        \\
Var: Individuals (Intercept)        & $0.19$        & $0.19$        & $0.18$        & $0.17$        & $0.17$        & $0.16$        \\
Var: Residual                       & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.46$        \\
\hline
\multicolumn{7}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.
Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.}}
\end{tabular}
}
\label{table:coefficients}
\end{center}
\end{table}

\begin{table}[!ht]
\caption{\label{tab:interact} Interaction effects for the justification of economic inequality}
\begin{center}
\scalebox{0.765}{
\begin{threeparttable}
\begin{tabular}{l c c c}
\toprule
 & Model 1 & Model 2 & Model 3 \\
\midrule
Intercept                                  & $2.04^{***}$  & $1.99^{***}$  & $2.17^{***}$  \\
                                           & $(0.10)$      & $(0.11)$      & $(0.11)$      \\
Wave                                       & $-0.14^{***}$ & $-0.11^{***}$ & $-0.17^{***}$ \\
                                           & $(0.02)$      & $(0.03)$      & $(0.03)$      \\
Wave$^2$                                   & $0.03^{***}$  & $0.02^{***}$  & $0.03^{***}$  \\
                                           & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
Merit: Talent                              & $0.01$        & $0.03$        & $0.01$        \\
                                           & $(0.01)$      & $(0.02)$      & $(0.01)$      \\
Merit: Effort                              & $0.05^{***}$  & $0.06^{***}$  & $0.02$        \\
                                           & $(0.01)$      & $(0.01)$      & $(0.02)$      \\
Wave * Merit: Talent                       &               & $-0.00$       &               \\
                                           &               & $(0.01)$      &               \\
Wave * Merit: Effort                       &               &               & $0.01^{*}$    \\
                                           &               &               & $(0.01)$      \\
\midrule
BIC                                        & $30936.16$    & $29763.63$    & $29921.16$    \\
Num. obs.                                  & $8042$        & $8042$        & $8042$        \\
Num. groups: Individuals                   & $1689$        & $1689$        & $1689$        \\
Var: Individuals (Intercept)               & $0.16$        & $1.13$        & $1.03$        \\
Var: Residual                              & $0.46$        & $0.32$        & $0.33$        \\
Var: Individuals Wave                      & $$            & $0.02$        & $0.02$        \\
Var: Individuals Merit: Talent             & $$            & $0.11$        & $$            \\
Cov: Individuals (Intercept) Wave          & $$            & $-0.07$       & $-0.07$       \\
Cov: Individuals (Intercept) Merit: Talent & $$            & $-0.30$       & $$            \\
Cov: Individuals Wave Merit: Talent        & $$            & $0.00$        & $$            \\
Var: Individuals Merit: Effort             & $$            & $$            & $0.10$        \\
Cov: Individuals (Intercept) Merit: Effort & $$            & $$            & $-0.27$       \\
Cov: Individuals Wave Merit: Effort        & $$            & $$            & $0.01$        \\
\bottomrule
\end{tabular}
\begin{tablenotes}[flushleft]
\scriptsize{\item *** p < 0.001; ** p < 0.01; * p < 0.05. Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception}
\end{tablenotes}
\end{threeparttable}
}
\label{table:coefficients}
\end{center}
\end{table}

In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time. In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic. We test this hypothesis through interaction effects, displayed in Table \ref{tab:interact}. Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for). Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort. As observed, only the effort variable shows a significant interaction with the time variable (\(\beta\)=0.01, p\textless.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.

\section{Discussion and conclusions}\label{discussion-and-conclusions}

Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic. In general, our results showed mixed evidence about our hypotheses. Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal. We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens. In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic. As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises. As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies. A closer look to the political environment during the last year in the country could give additional hints. Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly. This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process (\citeproc{ref-palanza_chile_2023}{Palanza \& Sotomayor, 2023}; \citeproc{ref-sazo_chile_2023}{Sazo, 2023}).

Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality (\(H_2\)). In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification. Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude (\(H_3\)). Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.

The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education (\(H_4\)). At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent. Our results are favorable for the effort dimension but not confirmed for the talent dimension. In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent. This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.

Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time. However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis (\(H_5\)). In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era. An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers. Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.

Among the limitations of our study, we can mention at least three. First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies. In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions. Second, as it has been established by recent research (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}), the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital. In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard. Finally, there are limitations involving the longitudinal dimension of the data used. Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey. Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.

The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates. In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies. For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time. Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions. Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.

\pagebreak

\emph{Acknowledgments}

The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022\_046

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-barry_theories_1989}
Barry, B. (1989). \emph{Theories of justice}. Univ. of California Pr.

\bibitem[\citeproctext]{ref-batruch_belief_2023}
Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F. (2023). Belief in {School Meritocracy} and the {Legitimization} of {Social} and {Income Inequality}. \emph{Social Psychological and Personality Science}, \emph{14}(5), 621--635. \url{https://doi.org/10.1177/19485506221111017}

\bibitem[\citeproctext]{ref-bell_politics_2020}
Bell, E. (2020). The {Politics} of {Designing Tuition-Free College}: {How Socially Constructed Target Populations Influence Policy Support}. \emph{The Journal of Higher Education}, \emph{91}(6), 888--926. \url{https://doi.org/10.1080/00221546.2019.1706015}

\bibitem[\citeproctext]{ref-bellei_estudio_2013}
Bellei, C. (2013). El estudio de la segregaci{ó}n socioecon{ó}mica y acad{é}mica de la educaci{ó}n chilena. \emph{Estudios Pedag{ó}gicos (Valdivia)}, \emph{39}(1), 325--345. \url{https://doi.org/10.4067/S0718-07052013000100019}

\bibitem[\citeproctext]{ref-breznau_welfare_2021}
Breznau, N. (2021). The welfare state and risk perceptions: The {Novel Coronavirus Pandemic} and public concern in 70 countries. \emph{European Societies}, \emph{23}(sup1), S33--S46. \url{https://doi.org/10.1080/14616696.2020.1793215}

\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
Busemeyer, M. R., \& Iversen, T. (2020). The {Welfare State} with {Private Alternatives}: {The Transformation} of {Popular Support} for {Social Insurance}. \emph{The Journal of Politics}, \emph{82}(2), 671--686. \url{https://doi.org/10.1086/706980}

\bibitem[\citeproctext]{ref-Castillo2011}
Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice Research}, \emph{24}(4), 314--340. \url{https://doi.org/10.1007/s11211-011-0144-5}

\bibitem[\citeproctext]{ref-Castillo2012a_justice}
Castillo, J. C. (2012). Is {Inequality Becoming Just}? {Changes} in {Public Opinion} about {Economic Distribution} in {Chile}. \emph{Bulletin of Latin American Research}, \emph{31}(1), 1--18. \url{https://doi.org/10.1111/j.1470-9856.2011.00605.x}

\bibitem[\citeproctext]{ref-castillo_perception_2022}
Castillo, J. C., García-Castro, J.-D., \& Venegas, M. (2022). Perception of economic inequality: Concepts, associated factors and prospects of a burgeoning research agenda. \emph{International Journal of Social Psychology}, \emph{37}(1), 180--207. \url{https://doi.org/10.1080/02134748.2021.2009275}

\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F. (2023). A {Multidimensional Approach} for {Measuring Meritocratic Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP Social Inequality Survey}. \emph{International Journal of Sociology}, 1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}

\bibitem[\citeproctext]{ref-confortiLegitimationInequalityAmerican1992}
Conforti, J. M. (1992). The legitimation of inequality in {American} education. \emph{The Urban Review}, \emph{24}(4), 227--238. \url{https://doi.org/10.1007/BF01108357}

\bibitem[\citeproctext]{ref-corvalanMercadoEscolarLibertad2016}
Corvalán, J., Carrasco, A., \& García-Huidobro, J. E. (Eds.). (2016). \emph{Mercado escolar: {Libertad}, diversidad y desigualdad} (1st ed.). Ediciones UC. \url{https://doi.org/10.2307/j.ctv14rmrhn}

\bibitem[\citeproctext]{ref-darmodyImpactsCOVID19Control2021}
Darmody, M., Smyth, E., \& Russell, H. (2021). Impacts of the {COVID-19 Control Measures} on {Widening Educational Inequalities}. \emph{YOUNG}, \emph{29}(4), 366--380. \url{https://doi.org/10.1177/11033088211027412}

\bibitem[\citeproctext]{ref-dayPerceivedIdealInequality2023}
Day, M. V., \& Norton, M. I. (2023). Perceived and {Ideal Inequality} in {University Endowments} in the {United States}. \emph{Personality and Social Psychology Bulletin}, \emph{49}(8), 1151--1165. \url{https://doi.org/10.1177/01461672221083766}

\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
ELSOC, S. T. (2022). \emph{Estudio {Longitudinal Social} de {Chile}}. Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}

\bibitem[\citeproctext]{ref-Evans2017}
Evans, M. D. R., \& Kelley, J. (2017). Communism, {Capitalism}, and {Images} of {Class}: {Effects} of {Reference Groups}, {Reality}, and {Regime} in 43 {Nations} and 110,000 {Individuals}, 1987-2009. \emph{Cross-Cultural Research}, \emph{51}(514), 315--359. \url{https://doi.org/10.1177/1069397116677963}

\bibitem[\citeproctext]{ref-Evans1992}
Evans, M. D. R., Kelley, J., \& Kolosi, T. (1992). Images of {Class}: {Public Perceptions} in {Hungary} and {Australia}. \emph{American Sociological Review}, \emph{57}(4), 461. \url{https://doi.org/10.2307/2096095}

\bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
García-Sánchez, E., \& De Carvalho Galvão, S. (2022). Las creencias que justifican la desigualdad moderan la relaci{ó}n entre el estatus socioecon{ó}mico y el apoyo a la redistribuci{ó}n. \emph{Revista Internacional de Sociolog{í}a}, \emph{80}(3), e210. \url{https://doi.org/10.3989/ris.2022.80.3.21.29}

\bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality. \emph{Economics \& Politics}, \emph{30}(1), 27--54. \url{https://doi.org/10.1111/ecpo.12103}

\bibitem[\citeproctext]{ref-goldin_meritocracy_2000}
Goldin, C. (2000). Meritocracy and {Economic Inequality}. {Edited} by {Kenneth Arrow}, {Samuel Bowles}, and {Steven Durlauf} ( {Princeton}, {Princeton University Press}, 2000) 348 pp. \emph{Journal of Interdisciplinary History}, \emph{31}(3), 431a--433. \url{https://doi.org/10.1162/jinh.2000.31.3.431a}

\bibitem[\citeproctext]{ref-hadjar_meritokratie_2008}
Hadjar, A. (2008). \emph{{Meritokratie als Legitimationsprinzip}}. VS Verlag f{ü}r Sozialwissenschaften.

\bibitem[\citeproctext]{ref-hadler_why_2005}
Hadler, M. (2005). Why {Do People Accept Different Income Ratios}?: {A Multi-level Comparison} of {Thirty Countries}. \emph{Acta Sociologica}, \emph{48}(2), 131--154. \url{https://doi.org/10.1177/0001699305053768}

\bibitem[\citeproctext]{ref-hankivsky_introduction_2022}
Hankivsky, O., Morrow, M., \& Varcoe, C. (2022). {INTRODUCTION Women}'s {Health} in {Canada}: {Critical Intersectional Perspectives} on {Theory} and {Policy}. In M. Morrow, O. Hankivsky, \& C. Varcoe (Eds.), \emph{Women's {Health} in {Canada}} (pp. 1--10). University of Toronto Press. \url{https://doi.org/10.3138/9781442623958-003}

\bibitem[\citeproctext]{ref-hing_failure_2019}
Hing, L. S. S., Wilson, A. E., Gourevitch, P., English, J., \& Sin, P. (2019). Failure to {Respond} to {Rising Income Inequality}: {Processes That Legitimize Growing Disparities}. \emph{Daedalus}, \emph{148}(3), 105--135. \url{https://doi.org/10.1162/daed_a_01752}

\bibitem[\citeproctext]{ref-igliozziFairShareEffects2024}
Igliozzi, D., Granot, Y., \& Ottati, V. (2024). A fair share: {Effects} of disparity, allocation strategy and system justification on perceptions of policy support in the education domain. \emph{European Journal of Social Psychology}, ejsp.3040. \url{https://doi.org/10.1002/ejsp.3040}

\bibitem[\citeproctext]{ref-iturra_percepcion_2023}
Iturra, J., Castillo, J. C., Rufs Orellana, C., \& Maldonado, L. (2023). La percepci{ó}n de desigualdad econ{ó}mica y su influencia sobre la justificaci{ó}n de las diferencias de ingreso leg{í}timas. \emph{Estudios Sociol{ó}gicos de El Colegio de M{é}xico}, \emph{41}(122), 309--352. \url{https://doi.org/10.24201/es.2023v41n122.2260}

\bibitem[\citeproctext]{ref-janmaatSubjectiveInequalityReview2013}
Janmaat, J. G. (2013). Subjective inequality: {A} review of international comparative studies on people's views about inequality. \emph{Archives Europeennes de Sociologie}, \emph{54}(3), 357--389. \url{https://doi.org/10.1017/S0003975613000209}

\bibitem[\citeproctext]{ref-jassoJusticeEarningsNew1978}
Jasso, G. (1978). On the {Justice} of {Earnings}: {A New Specification} of the {Justice Evaluation Function}. \emph{American Journal of Sociology}, \emph{83}(6), 1398--1419. \url{https://www.jstor.org/stable/2778110}

\bibitem[\citeproctext]{ref-joikomujicaCuasimercadoEducativoChile2019}
Joiko Mujica, S. (2019). {El cuasi-mercado educativo en Chile: desarrollo y consecuencias}. \emph{Revista Electr{ó}nica Di{á}logos Educativos; Vol. 12 N{ú}m. 23 (2012); 148-174}.

\bibitem[\citeproctext]{ref-jost_theory_2020}
Jost, J. (2020). \emph{A theory of system justification}. Harvard University Press.

\bibitem[\citeproctext]{ref-jost_psychology_2003}
Jost, J., \& Hunyady, O. (2003). The psychology of system justification and the palliative function of ideology. \emph{European Review of Social Psychology}, \emph{13}(1), 111--153. \url{https://doi.org/10.1080/10463280240000046}

\bibitem[\citeproctext]{ref-kelleyLegitimationInequalityOccupational1993}
Kelley, J., \& Evans, M. D. R. (1993). The legitimation of inequality: {Occupational} earnings in nine nations. \emph{American Journal of Sociology}, \emph{99}(1), 75--125. \url{https://www.jstor.org/stable/2781956}

\bibitem[\citeproctext]{ref-kelley_legitimate_2021}
Kelley, J., \& Evans, M. D. R. (2021). Legitimate earnings inequality and national welfare commitment: {Correspondence} between economic institutions and the pay 80,000+ people in 30 nations think legitimate for ordinary jobs and for elite jobs. \emph{Social Science Research}, \emph{94}, 102446. \url{https://doi.org/10.1016/j.ssresearch.2020.102446}

\bibitem[\citeproctext]{ref-kluegel_social_1995}
Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995). \emph{Social justice and political change: Public opinion in capitalist and post-communist states}. A. de Gruyter.

\bibitem[\citeproctext]{ref-kluegelBeliefsStratification1981}
Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}. \emph{Annual Review of Sociology}, 29--56.

\bibitem[\citeproctext]{ref-kluegelBeliefsInequalityAmericans1986}
Kluegel, J. R., \& Smith, E. R. (1986). \emph{Beliefs about {Inequality}: {Americans}' {Views} of {What Is} and {What Ought} to {Be}} (1st ed.). Routledge.

\bibitem[\citeproctext]{ref-leeFairnessPerceptionsEducational2023}
Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational inequality: The effects of self-interest and neoliberal orientations. \emph{The Australian Educational Researcher}. \url{https://doi.org/10.1007/s13384-023-00636-6}

\bibitem[\citeproctext]{ref-liebigSociologyJustice2016}
Liebig, S., \& Sauer, C. (2016). Sociology of {Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social Justice Theory} and {Research}} (pp. 37--59). Springer New York.

\bibitem[\citeproctext]{ref-lindhPublicOpinionMarkets2015}
Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes} towards {Market Distribution} of {Social Services} -- {A Comparison} of 17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7), 887--910. \url{https://doi.org/10.1111/spol.12105}

\bibitem[\citeproctext]{ref-McCoy2007}
McCoy, S. K., \& Major, B. (2007). Priming meritocracy and the psychological justification of inequality. \emph{Journal of Experimental Social Psychology}, \emph{43}(3), 341--351. \url{https://doi.org/10.1016/j.jesp.2006.04.009}

\bibitem[\citeproctext]{ref-mcnamee_meritocracy_2004}
McNamee, S., \& Miller, R. (2004). \emph{The meritocracy myth} (2004th ed.). Rowman \& Littlefield.

\bibitem[\citeproctext]{ref-mijs_stratified_2016}
Mijs, J. (2016). Stratified {Failure}: {Educational Stratification} and {Students}' {Attributions} of {Their Mathematics Performance} in 24 {Countries}. \emph{Sociology of Education}, \emph{89}(2), 137--153. \url{https://doi.org/10.1177/0038040716636434}

\bibitem[\citeproctext]{ref-mijs_visualizing_2018}
Mijs, J. (2018). Visualizing {Belief} in {Meritocracy}, 1930--2010. \emph{Socius}, \emph{4}, 2378023118811805. \url{https://doi.org/10.1177/2378023118811805}

\bibitem[\citeproctext]{ref-mijs_paradox_2021}
Mijs, J. (2021). The paradox of inequality: Income inequality and belief in meritocracy go hand in hand. \emph{Socio-Economic Review}, \emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}

\bibitem[\citeproctext]{ref-mijs_belief_2022}
Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022). Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of {Subjective Social Mobility}. \emph{Social Psychology Quarterly}, \emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}

\bibitem[\citeproctext]{ref-oecd_pisa_2023}
OECD. (2023). \emph{{PISA} 2022 {Results} ({Volume I}): {The State} of {Learning} and {Equity} in {Education}}. OECD. \url{https://doi.org/10.1787/53f23881-en}

\bibitem[\citeproctext]{ref-osberg_fair_2006}
Osberg, L., \& Smeeding, T. (2006). {``{Fair}''} {Inequality}? {Attitudes} toward {Pay Differentials}: {The United States} in {Comparative Perspective}. \emph{American Sociological Review}, \emph{71}(3), 450--473. \url{https://doi.org/10.1177/000312240607100305}

\bibitem[\citeproctext]{ref-palanza_chile_2023}
Palanza, V., \& Sotomayor, P. (2023). Chile's failed constitutional intent: {Polarization}, fragmentation, haste and delegitimization. \emph{Global Constitutionalism}, 1--10. \url{https://doi.org/10.1017/S204538172300028X}

\bibitem[\citeproctext]{ref-ramosarellanoEducacionPromesaIncumplida2022}
Ramos Arellano, M. (Ed.). (2022). \emph{Educaci{ó}n: {La} promesa incumplida: Esfuerzo, miedos y esperanzas de familias chilenas en el mercado escolar} (Primera edici{ó}n). Catalonia : CIAE, Centro de Investigaci{ó}n Avanzada en Educaci{ó}n, Universidad de Chile.

\bibitem[\citeproctext]{ref-Resh2016}
Resh, N., \& Sabbagh, C. (2016). Justice and education. In \emph{Handbook of Social Justice Theory and Research}. \url{https://doi.org/10.1007/978-1-4939-3216-0_19}

\bibitem[\citeproctext]{ref-rosvall_mapping_2010}
Rosvall, M., \& Bergstrom, C. T. (2010). Mapping {Change} in {Large Networks}. \emph{PLoS ONE}, \emph{5}(1), e8694. \url{https://doi.org/10.1371/journal.pone.0008694}

\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the common good?} (First edition). {Farrar, Straus and Giroux}.

\bibitem[\citeproctext]{ref-sazo_chile_2023}
Sazo, D. (2023). Chile 2022: {From} great expectations to rising pessimism. \emph{Revista de Ciencia Pol{í}tica (Santiago)}, \emph{ahead}. \url{https://doi.org/10.4067/s0718-090x2023005000118}

\bibitem[\citeproctext]{ref-schroder_income_2017}
Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for {Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47. \url{https://doi.org/10.1007/s11211-016-0276-8}

\bibitem[\citeproctext]{ref-singer_applied_2009}
Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data analysis: Modeling change and event occurence}. Oxford University Press, Incorporated.

\bibitem[\citeproctext]{ref-sonhing_merit_2011}
Son Hing, L. S., Ramona, D., Zanna, M. P., Garcia, D. M., Gee, S. S., \& Orazietti, K. (2011). The merit of meritocracy. \emph{Journal of Personality and Social Psychology}, \emph{101}(3), 433--450. \url{https://doi.org/10.1037/a0024618}

\bibitem[\citeproctext]{ref-stoilovaFairnessEducationalOpportunities2023}
Stoilova, R., \& Ilieva-Trichkova, P. (2023). Fairness of educational opportunities and income distribution: Gender-sensitive analysis in a {European} comparative perspective. \emph{International Journal of Sociology and Social Policy}, \emph{43}(1/2), 272--291. \url{https://doi.org/10.1108/IJSSP-02-2022-0065}

\bibitem[\citeproctext]{ref-trump_income_2018}
Trump, K.-S. (2018). Income {Inequality Influences Perceptions} of {Legitimate Income Differences}. \emph{British Journal of Political Science}, \emph{48}(4), 929--952. \url{https://doi.org/10.1017/S0007123416000326}

\bibitem[\citeproctext]{ref-trump_when_2020}
Trump, K.-S. (2020). When and why is economic inequality seen as fair. \emph{Current Opinion in Behavioral Sciences}, \emph{34}, 46--51. \url{https://doi.org/10.1016/j.cobeha.2019.12.001}

\bibitem[\citeproctext]{ref-valantPoliticsAchievementGaps2016}
Valant, J., \& Newark, D. A. (2016). The {Politics} of {Achievement Gaps}: {U}.{S}. {Public Opinion} on {Race-Based} and {Wealth-Based Differences} in {Test Scores}. \emph{Educational Researcher}, \emph{45}(6), 331--346. \url{https://doi.org/10.3102/0013189X16658447}

\bibitem[\citeproctext]{ref-vandertoorn_twenty_2014}
van der Toorn, J., \& Jost, J. (2014). Twenty years of system justification theory: {Introduction} to the special issue on {``{Ideology} and system justification processes.''} \emph{Group Processes \& Intergroup Relations}, \emph{17}(4), 413--419. \url{https://doi.org/10.1177/1368430214531509}

\bibitem[\citeproctext]{ref-wegenerIllusionDistributiveJustice1987}
Wegener, B. (1987). The {Illusion} of {Distributive Justice}. \emph{European Sociological Review}, \emph{3}(1), 1--13. \url{https://www.jstor.org/stable/522527}

\bibitem[\citeproctext]{ref-Wegener1990}
Wegener, B. (1990). Equity, relative deprivation, and the value consensus paradox. \emph{Social Justice Research}, \emph{4}(1), 65--86. \url{https://doi.org/10.1007/BF01048536}

\bibitem[\citeproctext]{ref-young_rise_1962}
Young, M. (1962). \emph{The rise of the meritocracy}. Penguin Books.

\end{CSLReferences}

\newpage

\appendix

\appendix \section{Appendix}

\begin{table}[!ht]
\caption{\label{tab:multilevel-full} Multilevel model for the justification of economic inequality}
\begin{center}
\scalebox{0.6}{
\begin{tabular}{l c c c c c c}
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\
\hline
Intercept                           & $1.86^{***}$  & $1.98^{***}$  & $1.72^{***}$  & $2.22^{***}$  & $2.11^{***}$  & $2.04^{***}$  \\
                                    & $(0.02)$      & $(0.04)$      & $(0.09)$      & $(0.10)$      & $(0.10)$      & $(0.10)$      \\
Wave (Ref.= Wave 2016)              &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Wave 2017                     & $-0.15^{***}$ &               &               &               &               &               \\
                                    & $(0.02)$      &               &               &               &               &               \\
\quad Wave 2018                     & $-0.06^{*}$   &               &               &               &               &               \\
                                    & $(0.03)$      &               &               &               &               &               \\
\quad Wave 2019                     & $-0.07^{**}$  &               &               &               &               &               \\
                                    & $(0.03)$      &               &               &               &               &               \\
\quad Wave 2022                     & $0.21^{***}$  &               &               &               &               &               \\
                                    & $(0.02)$      &               &               &               &               &               \\
Wave                                &               & $-0.17^{***}$ & $-0.17^{***}$ & $-0.13^{***}$ & $-0.14^{***}$ & $-0.14^{***}$ \\
                                    &               & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      \\
Wave$^2$                            &               & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  \\
                                    &               & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
Education (Ref.=Primary)            &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad High School                   &               &               & $0.02$        & $0.03$        & $0.04$        & $0.05$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Technical                     &               &               & $-0.01$       & $-0.00$       & $0.01$        & $0.01$        \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
\quad Universitary                  &               &               & $-0.09$       & $-0.05$       & $-0.04$       & $-0.04$       \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
Household Income (Ref.= Quintile 1) &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Quintile 2                    &               &               & $0.07$        & $0.07$        & $0.07$        & $0.07$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Quintile 3                    &               &               & $0.08$        & $0.08$        & $0.08$        & $0.08$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Quintile 4                    &               &               & $0.12^{*}$    & $0.13^{*}$    & $0.13^{*}$    & $0.13^{*}$    \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
\quad Quintile 5                    &               &               & $0.19^{**}$   & $0.19^{**}$   & $0.19^{***}$  & $0.20^{***}$  \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
\quad Quintile (No information)     &               &               & $0.17^{*}$    & $0.17^{*}$    & $0.18^{*}$    & $0.18^{*}$    \\
                                    &               &               & $(0.08)$      & $(0.08)$      & $(0.08)$      & $(0.08)$      \\
Subjective Social Status            &               &               & $0.01$        & $0.01$        & $0.00$        & $0.00$        \\
                                    &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
Age (Ref.= 18-29)                   &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Age 30-49                     &               &               & $0.02$        & $0.02$        & $0.02$        & $0.02$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Age 50-64                     &               &               & $0.07$        & $0.08$        & $0.08$        & $0.07$        \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Age 65 or more                &               &               & $0.19^{**}$   & $0.20^{**}$   & $0.19^{**}$   & $0.18^{**}$   \\
                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
Gender (Ref. Male)                  &               &               & $-0.08^{*}$   & $-0.08^{*}$   & $-0.08^{*}$   & $-0.07^{*}$   \\
                                    &               &               & $(0.03)$      & $(0.03)$      & $(0.03)$      & $(0.03)$      \\
Pol. pos (Ref.= Left)               &               &               &               &               &               &               \\
                                    &               &               &               &               &               &               \\
\quad Center                        &               &               & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Right                         &               &               & $0.29^{***}$  & $0.28^{***}$  & $0.27^{***}$  & $0.27^{***}$  \\
                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
\quad Does not identify             &               &               & $0.15^{***}$  & $0.15^{***}$  & $0.14^{***}$  & $0.14^{***}$  \\
                                    &               &               & $(0.04)$      & $(0.04)$      & $(0.04)$      & $(0.04)$      \\
Inequality perception               &               &               &               & $-0.12^{***}$ & $-0.12^{***}$ & $-0.12^{***}$ \\
                                    &               &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
Merit: Talent                       &               &               &               &               & $0.04^{***}$  & $0.01$        \\
                                    &               &               &               &               & $(0.01)$      & $(0.01)$      \\
Merit: Effort                       &               &               &               &               &               & $0.05^{***}$  \\
                                    &               &               &               &               &               & $(0.01)$      \\
\hline
BIC                                 & $30911.29$    & $30907.80$    & $31053.70$    & $30942.30$    & $30940.78$    & $30936.16$    \\
Num. obs.                           & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        \\
Num. groups: Individuals            & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        \\
Var: Individuals (Intercept)        & $0.19$        & $0.19$        & $0.18$        & $0.17$        & $0.17$        & $0.16$        \\
Var: Residual                       & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.46$        \\
\hline
\multicolumn{7}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.}}
\end{tabular}
}
\label{table:coefficients}
\end{center}
\end{table}

\end{document}

### 2024-03-26 paper.Rmd
ajustes

@@
 # lang: en-GB
-author: 
-- name: Juan Carlos Castillo
-  affiliation: Department of Sociology, Universidad de Chile
-  email: juancastillov@uchile.cl
-  number: 1, 3 
-- name: Julio Iturra
-  affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-  number: 2
-- name: Kevin Carrasco
-  affiliation: Center for Social Conflict and Cohesion Studies
-  number: 3
+# author: 
+# - name: Juan Carlos Castillo
+#   affiliation: Department of Sociology, Universidad de Chile
+#   email: juancastillov@uchile.cl
+#   number: 1, 3 
+# - name: Julio Iturra
+#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
+#   number: 2
+# - name: Kevin Carrasco
+#   affiliation: Center for Social Conflict and Cohesion Studies
+#   number: 3
 abstract: | 
@@
     number_sections: false
-    reference_docx: input/template.docx
+    # reference_docx: input/template.docx
     toc: false
@@
 ```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H", fig.align = 'center')
+ knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
+                       # , fig.align = 'center'
+                       )
  # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
@@
 The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joikomujicaCuasimercadoEducativoChile2019].
+The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
 Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramosarellanoEducacionPromesaIncumplida2022].
+Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
 Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
@@
 
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@hing_failure_2019], expecting that:
+Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
 
@@
 
-_Acknowledgments_
+<!-- _Acknowledgments_ -->
 
-The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046
+<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
 

### 2024-03-26 Title-Page.Rmd
update title page

@@
 
-Word count:
+Word count: 7663
 
 [^*]: Corresponding author (juancastillov@uchile.cl - [jc-castillo.com](https://juancarloscastillo.github.io/jc-castillo/) -  Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile).
-The authors recognize the support of ANID/FONDECYT grant 1210847 and of the Center for Conflict and Social Cohesion Studies (COES)  grant ANID/FONDAP 1513009.
+The authors recognize the support of the National Research and Development Agency (ANID) through the grant from the National Fund for Scientific and Technological Development (FONDECYT) N° 1210847; from the Center for Conflict and Social Cohesion Studies (COES) grant Financing Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046

### 2024-03-26 paper.Rmd
update title page

@@
 We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
+Model 1 is shown as a baseline model, it is the same as Model 6 in Table \@ref(tab:multilevel) but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
 Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.

### 2024-03-26 Title-page.tex
up tex files

--- 2024-03-26 paper.tex
@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
 %
@@
   12pt,
-  a4paper,
 ]{article}
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{setspace}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic},
-%   colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
+\usepackage[margin=0.78in]{geometry}
 \usepackage{graphicx}
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{times}
-\hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}
 \usepackage{caption}
 \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
+\usepackage[figurename=Fig.]{caption}
 \usepackage{graphicx}
@@
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic },
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
-\title{Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic}
-    \usepackage[]{authblk}
-    \setlength{\affilsep}{2em}
-    \renewcommand*{\Affilfont}{\normalsize\normalfont}
-    \renewcommand\Authsep{, }
-    \renewcommand\Authand{ \& }
-    \renewcommand\Authands{, \& }
-                                        \author[1, 3]{Juan Carlos Castillo}
-                                                            \affil[]{Department of Sociology, Universidad de Chile \thanks{\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}}}
-                                                                                \author[2]{Julio Iturra}
-                                                            \affil[]{Bremen International Graduate School of Social Sciences, Universität Bremen}
-                                                                                \author[3]{Kevin Carrasco}
-                                                            \affil[]{Center for Social Conflict and Cohesion Studies}
-                                            \date{}
-
-\providecommand{\keywords}[1]
-{
-  \small
-  \textbf{Keywords:} #1
-}
+\title{Changes in the justification of educational inequalities: The
+role of perceptions of inequality and meritocracy during the COVID
+pandemic \footnote{Corresponding author
+  (\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}
+  -
+  \href{https://juancarloscastillo.github.io/jc-castillo/}{jc-castillo.com}
+  - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile). The authors
+  recognize the support of the National Research and Development Agency
+  (ANID) through the grant from the National Fund for Scientific and
+  Technological Development (FONDECYT) N° 1210847; from the Center for
+  Conflict and Social Cohesion Studies (COES) grant Financing Fund for
+  Research Centers in Priority Areas (FONDAP) N° 1523A0005; and
+  Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS)
+  NCS2022\_046}}
+\author{}
+\date{\vspace{-2.5em}}
 
@@
 \maketitle
-\begin{abstract}
-Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-.
-\newline
-\newline
-
-\textbf{Keywords}: meritocracy, social inequality, inequality justification, COVID-19
-\end{abstract}
 
 \setstretch{1.5}
-\pagebreak
+Juan Carlos Castillo, Department of Sociology, Universidad de Chile,
+Santiago de Chile. *
 
-\section{Introduction}\label{introduction}
+ORCID: 0000-0003-1265-7854
 
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, ``Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries'' (\citeproc{ref-oecd_pisa_2023}{OECD, 2023}).
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 (\citeproc{ref-darmodyImpactsCOVID19Control2021}{Darmody et al., 2021}).
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education (\citeproc{ref-leeFairnessPerceptionsEducational2023}{Lee \& Stacey, 2023}).
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
+Julio Iturra, Bremen International Graduate School of Social Sciences,
+Bremen, Germany.
 
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed (\citeproc{ref-hadjar_meritokratie_2008}{Hadjar, 2008}; \citeproc{ref-mcnamee_meritocracy_2004}{McNamee \& Miller, 2004}; \citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}).
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
+ORCID: 0000-0003-1993-3479
 
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system (\citeproc{ref-bellei_estudio_2013}{Bellei, 2013}; \citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}).
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation (\citeproc{ref-joikomujicaCuasimercadoEducativoChile2019}{Joiko Mujica, 2019}).
-Previous research in Chile has emphasized the concept of an ``educational market'' (\citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}), characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences (\citeproc{ref-ramosarellanoEducacionPromesaIncumplida2022}{Ramos Arellano, 2022}).
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
+Kevin Carrasco, Center for Social Conflict and Cohesion Studies
 
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}; \citeproc{ref-hankivsky_introduction_2022}{Hankivsky et al., 2022}).
-A pertinent aspect of this discourse is articulated in the works of Mijs (\citeproc{ref-mijs_stratified_2016}{2016}, \citeproc{ref-mijs_visualizing_2018}{2018}), showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
+ORCID: 0000-0002-9348-8361
 
-\section{Inequality, beliefs and justification}\label{inequality-beliefs-and-justification}
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as ``the defensibility of unequal relations between people'' (\citeproc{ref-barry_theories_1989}{1989, p.
-3}), whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification (\citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}; \citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-trump_income_2018}{Trump, 2018}).
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research (\citeproc{ref-Resh2016}{Resh \& Sabbagh, 2016}).
-From a sociological point of view, the works of Kluegel \& Smith (\citeproc{ref-kluegelBeliefsStratification1981}{1981}) can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of ``beliefs'' in a ``broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action.'' (\citeproc{ref-kluegelBeliefsStratification1981}{Kluegel \& Smith, 1981, p.
-30}).
-However, the literature has strained the concept of ``beliefs'' by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, Wegener (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{1987}) has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987, p.
-2}).
-This approach established that status positions influence how people perceive things \emph{are} and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust (\citeproc{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}).
-Additionally, Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}) has suggested more detailed conceptual distinctions about how people perceive economic inequality \emph{is}, how they believe or prefer it \emph{should be}, and the process of evaluating such differences as fair.
-
-Theoretically, Liebig \& Sauer (\citeproc{ref-liebigSociologyJustice2016}{2016}) establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, \emph{order-related attitudes} refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, \emph{procedural justice attitudes} account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be \emph{outcome-related attitudes} which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles (\citeproc{ref-kluegel_social_1995}{Kluegel et al., 1995}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011})
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{Janmaat, 2013}; \citeproc{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \& Smith, 1986}).
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries (\citeproc{ref-Evans1992}{Evans et al., 1992}; \citeproc{ref-Evans2017}{Evans \& Kelley, 2017}) and greater justification of salary gaps (\citeproc{ref-hadler_why_2005}{Hadler, 2005}; \citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}).
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}; \citeproc{ref-vandertoorn_twenty_2014}{van der Toorn \& Jost, 2014}).
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies (\citeproc{ref-trump_when_2020}{Trump, 2020}).
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position (\citeproc{ref-jost_theory_2020}{Jost, 2020}).
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-kelleyLegitimationInequalityOccupational1993}{Kelley \& Evans, 1993}; \citeproc{ref-osberg_fair_2006}{Osberg \& Smeeding, 2006}).
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987}, \citeproc{ref-Wegener1990}{1990}).
-Thus, both the role of social status and economic inequality perceptions have found support in correlational (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012}; \citeproc{ref-schroder_income_2017}{Schröder, 2017}), as well as in experimental studies that have manipulated inequality perception (\citeproc{ref-iturra_percepcion_2023}{Iturra et al., 2023}; \citeproc{ref-trump_income_2018}{Trump, 2018}).
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification (\citeproc{ref-Castillo2011}{Castillo, 2011}).
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences (\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De Carvalho Galvão, 2022}).
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-\section{The justification of educational inequality}\label{the-justification-of-educational-inequality}
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education, attending to the degree of justification of market-based access to social services.
-For instance, Lindh (\citeproc{ref-lindhPublicOpinionMarkets2015}{2015}) found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, Busemeyer \& Iversen (\citeproc{ref-busemeyer_welfare_2020}{2020}) shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by Conforti (\citeproc{ref-confortiLegitimationInequalityAmerican1992}{1992}) as ``while equality is embraced as a value, inequality is accepted as a reality'' (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that people with higher incomes can buy better education for their children than people with lower incomes?}), and the European Social Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I have had a fair chance of achieving the level of education I was seeking}).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about---and more supportive of proposals to close---wealth-based achievement gaps than Black-White or Hispanic-White gaps (\citeproc{ref-valantPoliticsAchievementGaps2016}{Valant \& Newark, 2016}).
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification (\citeproc{ref-stoilovaFairnessEducationalOpportunities2023}{Stoilova \& Ilieva-Trichkova, 2023}).
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}), justifications are usually confused under the labels of ``beliefs'' and ``attitudes,'' and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011}).
-For instance, a recent study by Day \& Norton (\citeproc{ref-dayPerceivedIdealInequality2023}{2023}) on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman, 2018})
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by Igliozzi et al.
-(\citeproc{ref-igliozziFairShareEffects2024}{2024}), they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}) is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance (`Students who place within the highest 50\% of testing scores from the previous year can attend the magnet school').
-In the same line, Lee \& Stacey (\citeproc{ref-leeFairnessPerceptionsEducational2023}{2023}) show that performance assessments (in what they call ``neoliberal orientations'') predicted people's fair perceptions of educational inequality.
-
-\emph{Justificacion, inequality and meritocracy in the educational domain}
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work ``The Rise of the Meritocracy'' (\citeproc{ref-young_rise_1962}{1962}), has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities (\citeproc{ref-bell_politics_2020}{Bell, 2020}).
-This would contribute to the justification of inequality in the educational domain (\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}), as individual ability is often inherited, and educational attainment is influenced by family background (\citeproc{ref-goldin_meritocracy_2000}{Goldin, 2000}).
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences (\citeproc{ref-McCoy2007}{McCoy \& Major, 2007}).
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}).
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities (\citeproc{ref-dayPerceivedIdealInequality2023}{Day \& Norton, 2023}).
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-\emph{This study}
-
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}).
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-\(H_1\): The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences (\citeproc{ref-hing_failure_2019}{Hing et al., 2019}), expecting that:
-
-\(H_2\): Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-\(H_3\): A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-\(H_4\): Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-\(H_5\): The positive association between meritocracy and inequality justification mitigates along time
-
-\section{Data, variables \& methods}\label{data-variables-methods}
-
-\subsection{Data}\label{data}
-
-The main data source is the Chilean Longitudinal Social Survey 2016--2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77\% of the total national population and 93\% of the urban areas (\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}).
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4\% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41\%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit \url{https://coes.cl/encuesta-panel/}.
-
-\subsection{Variables}\label{variables}
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: ``It is just that high-income people have a better education for their children than people with lower incomes'' (``Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos'' in Spanish).
-Here, the respondents declared their preferences on a Likert scale from ``strongly disagree'' (1) to ``strongly agree'' (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort (``In Chile, people are rewarded for their efforts''), and the other related to talent (``In Chile, people are rewarded for their intelligence and skills'').
-In addition, we included an indicator for economic inequality perception (``In Chile, the income differences are too large'').
-Figure \ref{fig:frecuencies} shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2\%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7\%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6\%) and for their efforts (57\%).
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/merit_plot}
-
-}
-
-\caption{Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:frecuencies}
-\end{figure}
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \ref{tab:descriptives} shows the whole set of independent variables, the corresponding items, response categories, and frequencies.
-
-\begin{longtable}[]{@{}l@{}}
-\caption{\label{tab:descriptives}Independent variables}\tabularnewline
-\toprule\noalign{}
-\endfirsthead
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-\includegraphics{output/tables/desc02.png} \\
-\end{longtable}
-
-\subsection{Methods}\label{methods}
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009}).
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library ``lme4'' (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009, p.
-4}).
-
-\section{Results}\label{results}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/alluvial_dep}
-
-}
-
-\caption{Change in the justification of educational inequality over time (2016-2022)}\label{fig:alluvial}
-\end{figure}
-
-Figure \ref{fig:alluvial} illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data (\citeproc{ref-rosvall_mapping_2010}{Rosvall \& Bergstrom, 2010}).
-For instance, of the 32.2\% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90\% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/years_plot}
-
-}
-
-\caption{Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years}\label{fig:means}
-\end{figure}
-
-Figure \ref{fig:means} shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/corr}
-
-}
-
-\caption{Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:correlation}
-\end{figure}
-
-Figure \ref{fig:correlation} presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (\emph{r}=-0.29, p\textless.01) and a moderate and positive association with both meritocracy perception variables (\emph{r}=0.18, p\textless.01; \emph{r}=0.15, p\textless.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (\emph{r}=-0.15, p\textless.01; \emph{r}=-0.10, p\textless.01).
-Finally, the two meritocracy variables present a high and positive association with each other (\emph{r}=0.74, p\textless.01).
-
-\subsection{Multivariate analysis}\label{multivariate-analysis}
-
-Table \ref{tab:multilevel} shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 (\(\beta\)=-0.15, p\textless.001), 2018 (\(\beta\)=-0.06, p\textless.05), and 2019 (\(\beta\)=-0.07, p\textless.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality (\(\beta\)=0.21, p\textless.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 (\(\beta\)=0.04, p\textless.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality (\(\beta\)=0.05, p\textless.01), but it has a larger weight than the perception of talent.
-
-\begin{table}[!ht]
-\caption{\label{tab:multilevel} Multilevel longitudinal models for the justification of inequality in education}
-\begin{center}
-\scalebox{0.765}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\
-\hline
-Intercept                           & $1.86^{***}$  & $1.98^{***}$  & $1.72^{***}$  & $2.22^{***}$  & $2.11^{***}$  & $2.04^{***}$  \\
-                                    & $(0.02)$      & $(0.04)$      & $(0.09)$      & $(0.10)$      & $(0.10)$      & $(0.10)$      \\
-Wave (Ref.= Wave 2016)              &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Wave 2017                     & $-0.15^{***}$ &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-\quad Wave 2018                     & $-0.06^{*}$   &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2019                     & $-0.07^{**}$  &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2022                     & $0.21^{***}$  &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-Wave                                &               & $-0.17^{***}$ & $-0.17^{***}$ & $-0.13^{***}$ & $-0.14^{***}$ & $-0.14^{***}$ \\
-                                    &               & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      \\
-Wave$^2$                            &               & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  \\
-                                    &               & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
-Household Income (Ref.= Quintile 1) &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Quintile 2                    &               &               & $0.07$        & $0.07$        & $0.07$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 3                    &               &               & $0.08$        & $0.08$        & $0.08$        & $0.08$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 4                    &               &               & $0.12^{*}$    & $0.13^{*}$    & $0.13^{*}$    & $0.13^{*}$    \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile 5                    &               &               & $0.19^{**}$   & $0.19^{**}$   & $0.19^{***}$  & $0.20^{***}$  \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile (No information)     &               &               & $0.17^{*}$    & $0.17^{*}$    & $0.18^{*}$    & $0.18^{*}$    \\
-                                    &               &               & $(0.08)$      & $(0.08)$      & $(0.08)$      & $(0.08)$      \\
-Age (Ref.= 18-29)                   &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Age 30-49                     &               &               & $0.02$        & $0.02$        & $0.02$        & $0.02$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 50-64                     &               &               & $0.07$        & $0.08$        & $0.08$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 65 or more                &               &               & $0.19^{**}$   & $0.20^{**}$   & $0.19^{**}$   & $0.18^{**}$   \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-Gender (Ref. Male)                  &               &               & $-0.08^{*}$   & $-0.08^{*}$   & $-0.08^{*}$   & $-0.07^{*}$   \\
-                                    &               &               & $(0.03)$      & $(0.03)$      & $(0.03)$      & $(0.03)$      \\
-Pol. pos (Ref.= Left)               &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Center                        &               &               & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Right                         &               &               & $0.29^{***}$  & $0.28^{***}$  & $0.27^{***}$  & $0.27^{***}$  \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Does not identify             &               &               & $0.15^{***}$  & $0.15^{***}$  & $0.14^{***}$  & $0.14^{***}$  \\
-                                    &               &               & $(0.04)$      & $(0.04)$      & $(0.04)$      & $(0.04)$      \\
-Inequality perception               &               &               &               & $-0.12^{***}$ & $-0.12^{***}$ & $-0.12^{***}$ \\
-                                    &               &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-Merit: Talent                       &               &               &               &               & $0.04^{***}$  & $0.01$        \\
-                                    &               &               &               &               & $(0.01)$      & $(0.01)$      \\
-Merit: Effort                       &               &               &               &               &               & $0.05^{***}$  \\
-                                    &               &               &               &               &               & $(0.01)$      \\
-\hline
-BIC                                 & $30911.29$    & $30907.80$    & $31053.70$    & $30942.30$    & $30940.78$    & $30936.16$    \\
-Num. obs.                           & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        \\
-Num. groups: Individuals            & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        \\
-Var: Individuals (Intercept)        & $0.19$        & $0.19$        & $0.18$        & $0.17$        & $0.17$        & $0.16$        \\
-Var: Residual                       & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.46$        \\
-\hline
-\multicolumn{7}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.}}
-\end{tabular}
-}
-\label{table:coefficients}
-\end{center}
-\end{table}
-
-\begin{table}[!ht]
-\caption{\label{tab:interact} Interaction effects for the justification of economic inequality}
-\begin{center}
-\scalebox{0.765}{
-\begin{threeparttable}
-\begin{tabular}{l c c c}
-\toprule
- & Model 1 & Model 2 & Model 3 \\
-\midrule
-Intercept                                  & $2.04^{***}$  & $1.99^{***}$  & $2.17^{***}$  \\
-                                           & $(0.10)$      & $(0.11)$      & $(0.11)$      \\
-Wave                                       & $-0.14^{***}$ & $-0.11^{***}$ & $-0.17^{***}$ \\
-                                           & $(0.02)$      & $(0.03)$      & $(0.03)$      \\
-Wave$^2$                                   & $0.03^{***}$  & $0.02^{***}$  & $0.03^{***}$  \\
-                                           & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
-Merit: Talent                              & $0.01$        & $0.03$        & $0.01$        \\
-                                           & $(0.01)$      & $(0.02)$      & $(0.01)$      \\
-Merit: Effort                              & $0.05^{***}$  & $0.06^{***}$  & $0.02$        \\
-                                           & $(0.01)$      & $(0.01)$      & $(0.02)$      \\
-Wave * Merit: Talent                       &               & $-0.00$       &               \\
-                                           &               & $(0.01)$      &               \\
-Wave * Merit: Effort                       &               &               & $0.01^{*}$    \\
-                                           &               &               & $(0.01)$      \\
-\midrule
-BIC                                        & $30936.16$    & $29763.63$    & $29921.16$    \\
-Num. obs.                                  & $8042$        & $8042$        & $8042$        \\
-Num. groups: Individuals                   & $1689$        & $1689$        & $1689$        \\
-Var: Individuals (Intercept)               & $0.16$        & $1.13$        & $1.03$        \\
-Var: Residual                              & $0.46$        & $0.32$        & $0.33$        \\
-Var: Individuals Wave                      & $$            & $0.02$        & $0.02$        \\
-Var: Individuals Merit: Talent             & $$            & $0.11$        & $$            \\
-Cov: Individuals (Intercept) Wave          & $$            & $-0.07$       & $-0.07$       \\
-Cov: Individuals (Intercept) Merit: Talent & $$            & $-0.30$       & $$            \\
-Cov: Individuals Wave Merit: Talent        & $$            & $0.00$        & $$            \\
-Var: Individuals Merit: Effort             & $$            & $$            & $0.10$        \\
-Cov: Individuals (Intercept) Merit: Effort & $$            & $$            & $-0.27$       \\
-Cov: Individuals Wave Merit: Effort        & $$            & $$            & $0.01$        \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}[flushleft]
-\scriptsize{\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception}
-\end{tablenotes}
-\end{threeparttable}
-}
-\label{table:coefficients}
-\end{center}
-\end{table}
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \ref{tab:interact}.
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable (\(\beta\)=0.01, p\textless.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-\section{Discussion and conclusions}\label{discussion-and-conclusions}
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process (\citeproc{ref-palanza_chile_2023}{Palanza \& Sotomayor, 2023}; \citeproc{ref-sazo_chile_2023}{Sazo, 2023}).
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality (\(H_2\)).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude (\(H_3\)).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education (\(H_4\)).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis (\(H_5\)).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
-
-Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}), the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-\emph{Acknowledgments}
-
-The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022\_046
-
-\section{References}\label{references}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-barry_theories_1989}
-Barry, B. (1989). \emph{Theories of justice}. Univ. of California Pr.
-
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
-Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
-(2023).
-Belief in {School Meritocracy} and the {Legitimization} of {Social} and {Income Inequality}.
-\emph{Social Psychological and Personality Science}, \emph{14}(5), 621--635.
-\url{https://doi.org/10.1177/19485506221111017}
-
-\bibitem[\citeproctext]{ref-bell_politics_2020}
-Bell, E.
-(2020).
-The {Politics} of {Designing Tuition-Free College}: {How Socially Constructed Target Populations Influence Policy Support}.
-\emph{The Journal of Higher Education}, \emph{91}(6), 888--926.
-\url{https://doi.org/10.1080/00221546.2019.1706015}
-
-\bibitem[\citeproctext]{ref-bellei_estudio_2013}
-Bellei, C.
-(2013).
-El estudio de la segregaci{ó}n socioecon{ó}mica y acad{é}mica de la educaci{ó}n chilena.
-\emph{Estudios Pedag{ó}gicos (Valdivia)}, \emph{39}(1), 325--345.
-\url{https://doi.org/10.4067/S0718-07052013000100019}
-
-\bibitem[\citeproctext]{ref-breznau_welfare_2021}
-Breznau, N.
-(2021).
-The welfare state and risk perceptions: The {Novel Coronavirus Pandemic} and public concern in 70 countries.
-\emph{European Societies}, \emph{23}(sup1), S33--S46.
-\url{https://doi.org/10.1080/14616696.2020.1793215}
-
-\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
-Busemeyer, M.
-R., \& Iversen, T.
-(2020).
-The {Welfare State} with {Private Alternatives}: {The Transformation} of {Popular Support} for {Social Insurance}.
-\emph{The Journal of Politics}, \emph{82}(2), 671--686.
-\url{https://doi.org/10.1086/706980}
-
-\bibitem[\citeproctext]{ref-Castillo2011}
-Castillo, J.
-C.
-(2011).
-Legitimacy of {Inequality} in a {Highly Unequal Context}: {Evidence} from the {Chilean Case}.
-\emph{Social Justice Research}, \emph{24}(4), 314--340.
-\url{https://doi.org/10.1007/s11211-011-0144-5}
-
-\bibitem[\citeproctext]{ref-Castillo2012a_justice}
-Castillo, J.
-C.
-(2012).
-Is {Inequality Becoming Just}?
-{Changes} in {Public Opinion} about {Economic Distribution} in {Chile}.
-\emph{Bulletin of Latin American Research}, \emph{31}(1), 1--18.
-\url{https://doi.org/10.1111/j.1470-9856.2011.00605.x}
-
-\bibitem[\citeproctext]{ref-castillo_perception_2022}
-Castillo, J.
-C., García-Castro, J.-D., \& Venegas, M.
-(2022).
-Perception of economic inequality: Concepts, associated factors and prospects of a burgeoning research agenda.
-\emph{International Journal of Social Psychology}, \emph{37}(1), 180--207.
-\url{https://doi.org/10.1080/02134748.2021.2009275}
-
-\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
-Castillo, J.
-C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
-(2023).
-A {Multidimensional Approach} for {Measuring Meritocratic Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP Social Inequality Survey}.
-\emph{International Journal of Sociology}, 1--25.
-\url{https://doi.org/10.1080/00207659.2023.2274712}
-
-\bibitem[\citeproctext]{ref-confortiLegitimationInequalityAmerican1992}
-Conforti, J. M. (1992). The legitimation of inequality in {American} education. \emph{The Urban Review}, \emph{24}(4), 227--238. \url{https://doi.org/10.1007/BF01108357}
-
-\bibitem[\citeproctext]{ref-corvalanMercadoEscolarLibertad2016}
-Corvalán, J., Carrasco, A., \& García-Huidobro, J. E. (Eds.). (2016). \emph{Mercado escolar: {Libertad}, diversidad y desigualdad} (1st ed.). Ediciones UC. \url{https://doi.org/10.2307/j.ctv14rmrhn}
-
-\bibitem[\citeproctext]{ref-darmodyImpactsCOVID19Control2021}
-Darmody, M., Smyth, E., \& Russell, H.
-(2021).
-Impacts of the {COVID-19 Control Measures} on {Widening Educational Inequalities}.
-\emph{YOUNG}, \emph{29}(4), 366--380.
-\url{https://doi.org/10.1177/11033088211027412}
-
-\bibitem[\citeproctext]{ref-dayPerceivedIdealInequality2023}
-Day, M.
-V., \& Norton, M.
-I.
-(2023).
-Perceived and {Ideal Inequality} in {University Endowments} in the {United States}.
-\emph{Personality and Social Psychology Bulletin}, \emph{49}(8), 1151--1165.
-\url{https://doi.org/10.1177/01461672221083766}
-
-\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
-ELSOC, S. T. (2022). \emph{Estudio {Longitudinal Social} de {Chile}}. Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
-
-\bibitem[\citeproctext]{ref-Evans2017}
-Evans, M.
-D.
-R., \& Kelley, J.
-(2017).
-Communism, {Capitalism}, and {Images} of {Class}: {Effects} of {Reference Groups}, {Reality}, and {Regime} in 43 {Nations} and 110,000 {Individuals}, 1987-2009.
-\emph{Cross-Cultural Research}, \emph{51}(514), 315--359.
-\url{https://doi.org/10.1177/1069397116677963}
-
-\bibitem[\citeproctext]{ref-Evans1992}
-Evans, M.
-D.
-R., Kelley, J., \& Kolosi, T.
-(1992).
-Images of {Class}: {Public Perceptions} in {Hungary} and {Australia}.
-\emph{American Sociological Review}, \emph{57}(4), 461.
-\url{https://doi.org/10.2307/2096095}
-
-\bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
-García-Sánchez, E., \& De Carvalho Galvão, S.
-(2022).
-Las creencias que justifican la desigualdad moderan la relaci{ó}n entre el estatus socioecon{ó}mico y el apoyo a la redistribuci{ó}n.
-\emph{Revista Internacional de Sociolog{í}a}, \emph{80}(3), e210.
-\url{https://doi.org/10.3989/ris.2022.80.3.21.29}
-
-\bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
-Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality. \emph{Economics \& Politics}, \emph{30}(1), 27--54. \url{https://doi.org/10.1111/ecpo.12103}
-
-\bibitem[\citeproctext]{ref-goldin_meritocracy_2000}
-Goldin, C.
-(2000).
-Meritocracy and {Economic Inequality}.
-{Edited} by {Kenneth Arrow}, {Samuel Bowles}, and {Steven Durlauf} ( {Princeton}, {Princeton University Press}, 2000) 348 pp.
-\emph{Journal of Interdisciplinary History}, \emph{31}(3), 431a--433.
-\url{https://doi.org/10.1162/jinh.2000.31.3.431a}
-
-\bibitem[\citeproctext]{ref-hadjar_meritokratie_2008}
-Hadjar, A. (2008). \emph{{Meritokratie als Legitimationsprinzip}}. VS Verlag f{ü}r Sozialwissenschaften.
-
-\bibitem[\citeproctext]{ref-hadler_why_2005}
-Hadler, M.
-(2005).
-Why {Do People Accept Different Income Ratios}?: {A Multi-level Comparison} of {Thirty Countries}.
-\emph{Acta Sociologica}, \emph{48}(2), 131--154.
-\url{https://doi.org/10.1177/0001699305053768}
-
-\bibitem[\citeproctext]{ref-hankivsky_introduction_2022}
-Hankivsky, O., Morrow, M., \& Varcoe, C.
-(2022).
-{INTRODUCTION Women}'s {Health} in {Canada}: {Critical Intersectional Perspectives} on {Theory} and {Policy}.
-In M.
-Morrow, O.
-Hankivsky, \& C.
-Varcoe (Eds.), \emph{Women's {Health} in {Canada}} (pp.
-1--10).
-University of Toronto Press.
-\url{https://doi.org/10.3138/9781442623958-003}
-
-\bibitem[\citeproctext]{ref-hing_failure_2019}
-Hing, L.
-S.
-S., Wilson, A.
-E., Gourevitch, P., English, J., \& Sin, P.
-(2019).
-Failure to {Respond} to {Rising Income Inequality}: {Processes That Legitimize Growing Disparities}.
-\emph{Daedalus}, \emph{148}(3), 105--135.
-\url{https://doi.org/10.1162/daed_a_01752}
-
-\bibitem[\citeproctext]{ref-igliozziFairShareEffects2024}
-Igliozzi, D., Granot, Y., \& Ottati, V.
-(2024).
-A fair share: {Effects} of disparity, allocation strategy and system justification on perceptions of policy support in the education domain.
-\emph{European Journal of Social Psychology}, ejsp.3040.
-\url{https://doi.org/10.1002/ejsp.3040}
-
-\bibitem[\citeproctext]{ref-iturra_percepcion_2023}
-Iturra, J., Castillo, J.
-C., Rufs Orellana, C., \& Maldonado, L.
-(2023).
-La percepci{ó}n de desigualdad econ{ó}mica y su influencia sobre la justificaci{ó}n de las diferencias de ingreso leg{í}timas.
-\emph{Estudios Sociol{ó}gicos de El Colegio de M{é}xico}, \emph{41}(122), 309--352.
-\url{https://doi.org/10.24201/es.2023v41n122.2260}
-
-\bibitem[\citeproctext]{ref-janmaatSubjectiveInequalityReview2013}
-Janmaat, J.
-G.
-(2013).
-Subjective inequality: {A} review of international comparative studies on people's views about inequality.
-\emph{Archives Europeennes de Sociologie}, \emph{54}(3), 357--389.
-\url{https://doi.org/10.1017/S0003975613000209}
-
-\bibitem[\citeproctext]{ref-jassoJusticeEarningsNew1978}
-Jasso, G.
-(1978).
-On the {Justice} of {Earnings}: {A New Specification} of the {Justice Evaluation Function}.
-\emph{American Journal of Sociology}, \emph{83}(6), 1398--1419.
-\url{https://www.jstor.org/stable/2778110}
-
-\bibitem[\citeproctext]{ref-joikomujicaCuasimercadoEducativoChile2019}
-Joiko Mujica, S. (2019). {El cuasi-mercado educativo en Chile: desarrollo y consecuencias}. \emph{Revista Electr{ó}nica Di{á}logos Educativos; Vol. 12 N{ú}m. 23 (2012); 148-174}.
-
-\bibitem[\citeproctext]{ref-jost_theory_2020}
-Jost, J. (2020). \emph{A theory of system justification}. Harvard University Press.
-
-\bibitem[\citeproctext]{ref-jost_psychology_2003}
-Jost, J., \& Hunyady, O.
-(2003).
-The psychology of system justification and the palliative function of ideology.
-\emph{European Review of Social Psychology}, \emph{13}(1), 111--153.
-\url{https://doi.org/10.1080/10463280240000046}
-
-\bibitem[\citeproctext]{ref-kelleyLegitimationInequalityOccupational1993}
-Kelley, J., \& Evans, M.
-D.
-R.
-(1993).
-The legitimation of inequality: {Occupational} earnings in nine nations.
-\emph{American Journal of Sociology}, \emph{99}(1), 75--125.
-\url{https://www.jstor.org/stable/2781956}
-
-\bibitem[\citeproctext]{ref-kelley_legitimate_2021}
-Kelley, J., \& Evans, M.
-D.
-R.
-(2021).
-Legitimate earnings inequality and national welfare commitment: {Correspondence} between economic institutions and the pay 80,000+ people in 30 nations think legitimate for ordinary jobs and for elite jobs.
-\emph{Social Science Research}, \emph{94}, 102446.
-\url{https://doi.org/10.1016/j.ssresearch.2020.102446}
-
-\bibitem[\citeproctext]{ref-kluegel_social_1995}
-Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995). \emph{Social justice and political change: Public opinion in capitalist and post-communist states}. A. de Gruyter.
-
-\bibitem[\citeproctext]{ref-kluegelBeliefsStratification1981}
-Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}. \emph{Annual Review of Sociology}, 29--56.
-
-\bibitem[\citeproctext]{ref-kluegelBeliefsInequalityAmericans1986}
-Kluegel, J. R., \& Smith, E. R. (1986). \emph{Beliefs about {Inequality}: {Americans}' {Views} of {What Is} and {What Ought} to {Be}} (1st ed.). Routledge.
-
-\bibitem[\citeproctext]{ref-leeFairnessPerceptionsEducational2023}
-Lee, J.-S., \& Stacey, M.
-(2023).
-Fairness perceptions of educational inequality: The effects of self-interest and neoliberal orientations.
-\emph{The Australian Educational Researcher}.
-\url{https://doi.org/10.1007/s13384-023-00636-6}
-
-\bibitem[\citeproctext]{ref-liebigSociologyJustice2016}
-Liebig, S., \& Sauer, C. (2016). Sociology of {Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social Justice Theory} and {Research}} (pp. 37--59). Springer New York.
-
-\bibitem[\citeproctext]{ref-lindhPublicOpinionMarkets2015}
-Lindh, A.
-(2015).
-Public {Opinion} against {Markets}?
-{Attitudes} towards {Market Distribution} of {Social Services} -- {A Comparison} of 17 {Countries}.
-\emph{Social Policy \& Administration}, \emph{49}(7), 887--910.
-\url{https://doi.org/10.1111/spol.12105}
-
-\bibitem[\citeproctext]{ref-McCoy2007}
-McCoy, S.
-K., \& Major, B.
-(2007).
-Priming meritocracy and the psychological justification of inequality.
-\emph{Journal of Experimental Social Psychology}, \emph{43}(3), 341--351.
-\url{https://doi.org/10.1016/j.jesp.2006.04.009}
-
-\bibitem[\citeproctext]{ref-mcnamee_meritocracy_2004}
-McNamee, S., \& Miller, R. (2004). \emph{The meritocracy myth} (2004th ed.). Rowman \& Littlefield.
-
-\bibitem[\citeproctext]{ref-mijs_stratified_2016}
-Mijs, J.
-(2016).
-Stratified {Failure}: {Educational Stratification} and {Students}' {Attributions} of {Their Mathematics Performance} in 24 {Countries}.
-\emph{Sociology of Education}, \emph{89}(2), 137--153.
-\url{https://doi.org/10.1177/0038040716636434}
-
-\bibitem[\citeproctext]{ref-mijs_visualizing_2018}
-Mijs, J. (2018). Visualizing {Belief} in {Meritocracy}, 1930--2010. \emph{Socius}, \emph{4}, 2378023118811805. \url{https://doi.org/10.1177/2378023118811805}
-
-\bibitem[\citeproctext]{ref-mijs_paradox_2021}
-Mijs, J. (2021). The paradox of inequality: Income inequality and belief in meritocracy go hand in hand. \emph{Socio-Economic Review}, \emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
-
-\bibitem[\citeproctext]{ref-mijs_belief_2022}
-Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J.
-(2022).
-Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of {Subjective Social Mobility}.
-\emph{Social Psychology Quarterly}, \emph{85}(2), 131--141.
-\url{https://doi.org/10.1177/01902725211063818}
-
-\bibitem[\citeproctext]{ref-oecd_pisa_2023}
-OECD. (2023). \emph{{PISA} 2022 {Results} ({Volume I}): {The State} of {Learning} and {Equity} in {Education}}. OECD. \url{https://doi.org/10.1787/53f23881-en}
-
-\bibitem[\citeproctext]{ref-osberg_fair_2006}
-Osberg, L., \& Smeeding, T.
-(2006).
-{``{Fair}''} {Inequality}?
-{Attitudes} toward {Pay Differentials}: {The United States} in {Comparative Perspective}.
-\emph{American Sociological Review}, \emph{71}(3), 450--473.
-\url{https://doi.org/10.1177/000312240607100305}
-
-\bibitem[\citeproctext]{ref-palanza_chile_2023}
-Palanza, V., \& Sotomayor, P.
-(2023).
-Chile's failed constitutional intent: {Polarization}, fragmentation, haste and delegitimization.
-\emph{Global Constitutionalism}, 1--10.
-\url{https://doi.org/10.1017/S204538172300028X}
-
-\bibitem[\citeproctext]{ref-ramosarellanoEducacionPromesaIncumplida2022}
-Ramos Arellano, M.
-(Ed.).
-(2022).
-\emph{Educaci{ó}n: {La} promesa incumplida: Esfuerzo, miedos y esperanzas de familias chilenas en el mercado escolar} (Primera edici{ó}n).
-Catalonia : CIAE, Centro de Investigaci{ó}n Avanzada en Educaci{ó}n, Universidad de Chile.
-
-\bibitem[\citeproctext]{ref-Resh2016}
-Resh, N., \& Sabbagh, C. (2016). Justice and education. In \emph{Handbook of Social Justice Theory and Research}. \url{https://doi.org/10.1007/978-1-4939-3216-0_19}
-
-\bibitem[\citeproctext]{ref-rosvall_mapping_2010}
-Rosvall, M., \& Bergstrom, C. T. (2010). Mapping {Change} in {Large Networks}. \emph{PLoS ONE}, \emph{5}(1), e8694. \url{https://doi.org/10.1371/journal.pone.0008694}
-
-\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
-Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the common good?} (First edition). {Farrar, Straus and Giroux}.
-
-\bibitem[\citeproctext]{ref-sazo_chile_2023}
-Sazo, D. (2023). Chile 2022: {From} great expectations to rising pessimism. \emph{Revista de Ciencia Pol{í}tica (Santiago)}, \emph{ahead}. \url{https://doi.org/10.4067/s0718-090x2023005000118}
-
-\bibitem[\citeproctext]{ref-schroder_income_2017}
-Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for {Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47. \url{https://doi.org/10.1007/s11211-016-0276-8}
-
-\bibitem[\citeproctext]{ref-singer_applied_2009}
-Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data analysis: Modeling change and event occurence}. Oxford University Press, Incorporated.
-
-\bibitem[\citeproctext]{ref-sonhing_merit_2011}
-Son Hing, L.
-S., Ramona, D., Zanna, M.
-P., Garcia, D.
-M., Gee, S.
-S., \& Orazietti, K.
-(2011).
-The merit of meritocracy.
-\emph{Journal of Personality and Social Psychology}, \emph{101}(3), 433--450.
-\url{https://doi.org/10.1037/a0024618}
-
-\bibitem[\citeproctext]{ref-stoilovaFairnessEducationalOpportunities2023}
-Stoilova, R., \& Ilieva-Trichkova, P.
-(2023).
-Fairness of educational opportunities and income distribution: Gender-sensitive analysis in a {European} comparative perspective.
-\emph{International Journal of Sociology and Social Policy}, \emph{43}(1/2), 272--291.
-\url{https://doi.org/10.1108/IJSSP-02-2022-0065}
-
-\bibitem[\citeproctext]{ref-trump_income_2018}
-Trump, K.-S.
-(2018).
-Income {Inequality Influences Perceptions} of {Legitimate Income Differences}.
-\emph{British Journal of Political Science}, \emph{48}(4), 929--952.
-\url{https://doi.org/10.1017/S0007123416000326}
-
-\bibitem[\citeproctext]{ref-trump_when_2020}
-Trump, K.-S. (2020). When and why is economic inequality seen as fair. \emph{Current Opinion in Behavioral Sciences}, \emph{34}, 46--51. \url{https://doi.org/10.1016/j.cobeha.2019.12.001}
-
-\bibitem[\citeproctext]{ref-valantPoliticsAchievementGaps2016}
-Valant, J., \& Newark, D.
-A.
-(2016).
-The {Politics} of {Achievement Gaps}: {U}.{S}.
-{Public Opinion} on {Race-Based} and {Wealth-Based Differences} in {Test Scores}.
-\emph{Educational Researcher}, \emph{45}(6), 331--346.
-\url{https://doi.org/10.3102/0013189X16658447}
-
-\bibitem[\citeproctext]{ref-vandertoorn_twenty_2014}
-van der Toorn, J., \& Jost, J.
-(2014).
-Twenty years of system justification theory: {Introduction} to the special issue on {``{Ideology} and system justification processes.''} \emph{Group Processes \& Intergroup Relations}, \emph{17}(4), 413--419.
-\url{https://doi.org/10.1177/1368430214531509}
-
-\bibitem[\citeproctext]{ref-wegenerIllusionDistributiveJustice1987}
-Wegener, B. (1987). The {Illusion} of {Distributive Justice}. \emph{European Sociological Review}, \emph{3}(1), 1--13. \url{https://www.jstor.org/stable/522527}
-
-\bibitem[\citeproctext]{ref-Wegener1990}
-Wegener, B. (1990). Equity, relative deprivation, and the value consensus paradox. \emph{Social Justice Research}, \emph{4}(1), 65--86. \url{https://doi.org/10.1007/BF01048536}
-
-\bibitem[\citeproctext]{ref-young_rise_1962}
-Young, M. (1962). \emph{The rise of the meritocracy}. Penguin Books.
-
-\end{CSLReferences}
-
-\newpage
-
-\appendix
-
-\appendix \section{Appendix}
-
-\begin{table}[!ht]
-\caption{\label{tab:multilevel-full} Multilevel model for the justification of economic inequality}
-\begin{center}
-\scalebox{0.6}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\
-\hline
-Intercept                           & $1.86^{***}$  & $1.98^{***}$  & $1.72^{***}$  & $2.22^{***}$  & $2.11^{***}$  & $2.04^{***}$  \\
-                                    & $(0.02)$      & $(0.04)$      & $(0.09)$      & $(0.10)$      & $(0.10)$      & $(0.10)$      \\
-Wave (Ref.= Wave 2016)              &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Wave 2017                     & $-0.15^{***}$ &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-\quad Wave 2018                     & $-0.06^{*}$   &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2019                     & $-0.07^{**}$  &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2022                     & $0.21^{***}$  &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-Wave                                &               & $-0.17^{***}$ & $-0.17^{***}$ & $-0.13^{***}$ & $-0.14^{***}$ & $-0.14^{***}$ \\
-                                    &               & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      \\
-Wave$^2$                            &               & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  \\
-                                    &               & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
-Education (Ref.=Primary)            &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad High School                   &               &               & $0.02$        & $0.03$        & $0.04$        & $0.05$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Technical                     &               &               & $-0.01$       & $-0.00$       & $0.01$        & $0.01$        \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Universitary                  &               &               & $-0.09$       & $-0.05$       & $-0.04$       & $-0.04$       \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-Household Income (Ref.= Quintile 1) &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Quintile 2                    &               &               & $0.07$        & $0.07$        & $0.07$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 3                    &               &               & $0.08$        & $0.08$        & $0.08$        & $0.08$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 4                    &               &               & $0.12^{*}$    & $0.13^{*}$    & $0.13^{*}$    & $0.13^{*}$    \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile 5                    &               &               & $0.19^{**}$   & $0.19^{**}$   & $0.19^{***}$  & $0.20^{***}$  \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile (No information)     &               &               & $0.17^{*}$    & $0.17^{*}$    & $0.18^{*}$    & $0.18^{*}$    \\
-                                    &               &               & $(0.08)$      & $(0.08)$      & $(0.08)$      & $(0.08)$      \\
-Subjective Social Status            &               &               & $0.01$        & $0.01$        & $0.00$        & $0.00$        \\
-                                    &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-Age (Ref.= 18-29)                   &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Age 30-49                     &               &               & $0.02$        & $0.02$        & $0.02$        & $0.02$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 50-64                     &               &               & $0.07$        & $0.08$        & $0.08$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 65 or more                &               &               & $0.19^{**}$   & $0.20^{**}$   & $0.19^{**}$   & $0.18^{**}$   \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-Gender (Ref. Male)                  &               &               & $-0.08^{*}$   & $-0.08^{*}$   & $-0.08^{*}$   & $-0.07^{*}$   \\
-                                    &               &               & $(0.03)$      & $(0.03)$      & $(0.03)$      & $(0.03)$      \\
-Pol. pos (Ref.= Left)               &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Center                        &               &               & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Right                         &               &               & $0.29^{***}$  & $0.28^{***}$  & $0.27^{***}$  & $0.27^{***}$  \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Does not identify             &               &               & $0.15^{***}$  & $0.15^{***}$  & $0.14^{***}$  & $0.14^{***}$  \\
-                                    &               &               & $(0.04)$      & $(0.04)$      & $(0.04)$      & $(0.04)$      \\
-Inequality perception               &               &               &               & $-0.12^{***}$ & $-0.12^{***}$ & $-0.12^{***}$ \\
-                                    &               &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-Merit: Talent                       &               &               &               &               & $0.04^{***}$  & $0.01$        \\
-                                    &               &               &               &               & $(0.01)$      & $(0.01)$      \\
-Merit: Effort                       &               &               &               &               &               & $0.05^{***}$  \\
-                                    &               &               &               &               &               & $(0.01)$      \\
-\hline
-BIC                                 & $30911.29$    & $30907.80$    & $31053.70$    & $30942.30$    & $30940.78$    & $30936.16$    \\
-Num. obs.                           & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        \\
-Num. groups: Individuals            & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        \\
-Var: Individuals (Intercept)        & $0.19$        & $0.19$        & $0.18$        & $0.17$        & $0.17$        & $0.16$        \\
-Var: Residual                       & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.46$        \\
-\hline
-\multicolumn{7}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.}}
-\end{tabular}
-}
-\label{table:coefficients}
-\end{center}
-\end{table}
+Word count: 7663
 

### 2024-03-26 paper.Rmd
up tex files

@@
  # clean #in the yml
+ rmarkdown::render("paper.Rmd", output_format = "latex_document")
  rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")

### 2024-03-26 paper.tex
up tex files

@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{setspace}
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic},
-%   colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 \usepackage{times}
@@
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic}
-    \usepackage[]{authblk}
-    \setlength{\affilsep}{2em}
-    \renewcommand*{\Affilfont}{\normalsize\normalfont}
-    \renewcommand\Authsep{, }
-    \renewcommand\Authand{ \& }
-    \renewcommand\Authands{, \& }
-                                        \author[1, 3]{Juan Carlos Castillo}
-                                                            \affil[]{Department of Sociology, Universidad de Chile \thanks{\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}}}
-                                                                                \author[2]{Julio Iturra}
-                                                            \affil[]{Bremen International Graduate School of Social Sciences, Universität Bremen}
-                                                                                \author[3]{Kevin Carrasco}
-                                                            \affil[]{Center for Social Conflict and Cohesion Studies}
-                                            \date{}
-
-\providecommand{\keywords}[1]
-{
-  \small
-  \textbf{Keywords:} #1
-}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\title{Changes in the justification of educational inequalities: The
+role of perceptions of inequality and meritocracy during the COVID
+pandemic}
+\author{}
+\date{\vspace{-2.5em}}
 
@@
 \begin{abstract}
-Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-.
-\newline
-\newline
-
-\textbf{Keywords}: meritocracy, social inequality, inequality justification, COVID-19
+Education is considered a key tool for social mobility and equality of
+opportunities. However, despite the widespread value of education,
+disparities in educational outcomes persist over time. To what extent
+are such educational disparities justified in society? And, what are the
+main factors driving this kind of justification? This research focuses
+on the changes in the justification of inequality in access to education
+in Chile and their links with the perception of meritocracy. The Chilean
+case offers an interesting context for this study given its high
+economic inequality, deep neoliberal policies, and commodification of
+social services such as education. The central argument of this article
+is that the justification of inequalities weakens during periods of
+vulnerability and crisis (such as the health and economic crisis
+resulting from the COVID-19 pandemic), which could be linked to a
+challenge of meritocratic ideals. For testing the research hypotheses we
+estimate a series of longitudinal multilevel models with data from a
+Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
+Whereas we find support for the association between the perception of
+meritocracy and justification of inequality, the analyses that involved
+changes over time showed an increase in the justification of inequality
+in education. The discussion of these results delves into the social
+consequences of justifying inequality in a sensitive area as education,
+as well as the persistence of meritocratic ideals despite challenging
+events.
+
+. \newline \newline
+
+\textbf{Keywords}: meritocracy, social inequality, inequality
+justification, COVID-19
 \end{abstract}
@@
 
-\section{Introduction}\label{introduction}
-
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, ``Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries'' (\citeproc{ref-oecd_pisa_2023}{OECD, 2023}).
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 (\citeproc{ref-darmodyImpactsCOVID19Control2021}{Darmody et al., 2021}).
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education (\citeproc{ref-leeFairnessPerceptionsEducational2023}{Lee \& Stacey, 2023}).
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed (\citeproc{ref-hadjar_meritokratie_2008}{Hadjar, 2008}; \citeproc{ref-mcnamee_meritocracy_2004}{McNamee \& Miller, 2004}; \citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}).
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
-
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system (\citeproc{ref-bellei_estudio_2013}{Bellei, 2013}; \citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}).
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation (\citeproc{ref-joikomujicaCuasimercadoEducativoChile2019}{Joiko Mujica, 2019}).
-Previous research in Chile has emphasized the concept of an ``educational market'' (\citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}), characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences (\citeproc{ref-ramosarellanoEducacionPromesaIncumplida2022}{Ramos Arellano, 2022}).
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
-
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}; \citeproc{ref-hankivsky_introduction_2022}{Hankivsky et al., 2022}).
-A pertinent aspect of this discourse is articulated in the works of Mijs (\citeproc{ref-mijs_stratified_2016}{2016}, \citeproc{ref-mijs_visualizing_2018}{2018}), showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
-\section{Inequality, beliefs and justification}\label{inequality-beliefs-and-justification}
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as ``the defensibility of unequal relations between people'' (\citeproc{ref-barry_theories_1989}{1989, p.
-3}), whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification (\citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}; \citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-trump_income_2018}{Trump, 2018}).
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research (\citeproc{ref-Resh2016}{Resh \& Sabbagh, 2016}).
-From a sociological point of view, the works of Kluegel \& Smith (\citeproc{ref-kluegelBeliefsStratification1981}{1981}) can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of ``beliefs'' in a ``broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action.'' (\citeproc{ref-kluegelBeliefsStratification1981}{Kluegel \& Smith, 1981, p.
-30}).
-However, the literature has strained the concept of ``beliefs'' by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, Wegener (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{1987}) has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987, p.
-2}).
-This approach established that status positions influence how people perceive things \emph{are} and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust (\citeproc{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}).
-Additionally, Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}) has suggested more detailed conceptual distinctions about how people perceive economic inequality \emph{is}, how they believe or prefer it \emph{should be}, and the process of evaluating such differences as fair.
-
-Theoretically, Liebig \& Sauer (\citeproc{ref-liebigSociologyJustice2016}{2016}) establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, \emph{order-related attitudes} refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, \emph{procedural justice attitudes} account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be \emph{outcome-related attitudes} which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles (\citeproc{ref-kluegel_social_1995}{Kluegel et al., 1995}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011})
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{Janmaat, 2013}; \citeproc{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \& Smith, 1986}).
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries (\citeproc{ref-Evans1992}{Evans et al., 1992}; \citeproc{ref-Evans2017}{Evans \& Kelley, 2017}) and greater justification of salary gaps (\citeproc{ref-hadler_why_2005}{Hadler, 2005}; \citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}).
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}; \citeproc{ref-vandertoorn_twenty_2014}{van der Toorn \& Jost, 2014}).
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies (\citeproc{ref-trump_when_2020}{Trump, 2020}).
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position (\citeproc{ref-jost_theory_2020}{Jost, 2020}).
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-kelleyLegitimationInequalityOccupational1993}{Kelley \& Evans, 1993}; \citeproc{ref-osberg_fair_2006}{Osberg \& Smeeding, 2006}).
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987}, \citeproc{ref-Wegener1990}{1990}).
-Thus, both the role of social status and economic inequality perceptions have found support in correlational (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012}; \citeproc{ref-schroder_income_2017}{Schröder, 2017}), as well as in experimental studies that have manipulated inequality perception (\citeproc{ref-iturra_percepcion_2023}{Iturra et al., 2023}; \citeproc{ref-trump_income_2018}{Trump, 2018}).
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification (\citeproc{ref-Castillo2011}{Castillo, 2011}).
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences (\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De Carvalho Galvão, 2022}).
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-\section{The justification of educational inequality}\label{the-justification-of-educational-inequality}
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education, attending to the degree of justification of market-based access to social services.
-For instance, Lindh (\citeproc{ref-lindhPublicOpinionMarkets2015}{2015}) found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, Busemeyer \& Iversen (\citeproc{ref-busemeyer_welfare_2020}{2020}) shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by Conforti (\citeproc{ref-confortiLegitimationInequalityAmerican1992}{1992}) as ``while equality is embraced as a value, inequality is accepted as a reality'' (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that people with higher incomes can buy better education for their children than people with lower incomes?}), and the European Social Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I have had a fair chance of achieving the level of education I was seeking}).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about---and more supportive of proposals to close---wealth-based achievement gaps than Black-White or Hispanic-White gaps (\citeproc{ref-valantPoliticsAchievementGaps2016}{Valant \& Newark, 2016}).
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification (\citeproc{ref-stoilovaFairnessEducationalOpportunities2023}{Stoilova \& Ilieva-Trichkova, 2023}).
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}), justifications are usually confused under the labels of ``beliefs'' and ``attitudes,'' and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011}).
-For instance, a recent study by Day \& Norton (\citeproc{ref-dayPerceivedIdealInequality2023}{2023}) on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman, 2018})
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by Igliozzi et al.
-(\citeproc{ref-igliozziFairShareEffects2024}{2024}), they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}) is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance (`Students who place within the highest 50\% of testing scores from the previous year can attend the magnet school').
-In the same line, Lee \& Stacey (\citeproc{ref-leeFairnessPerceptionsEducational2023}{2023}) show that performance assessments (in what they call ``neoliberal orientations'') predicted people's fair perceptions of educational inequality.
-
-\emph{Justificacion, inequality and meritocracy in the educational domain}
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work ``The Rise of the Meritocracy'' (\citeproc{ref-young_rise_1962}{1962}), has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities (\citeproc{ref-bell_politics_2020}{Bell, 2020}).
-This would contribute to the justification of inequality in the educational domain (\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}), as individual ability is often inherited, and educational attainment is influenced by family background (\citeproc{ref-goldin_meritocracy_2000}{Goldin, 2000}).
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences (\citeproc{ref-McCoy2007}{McCoy \& Major, 2007}).
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}).
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities (\citeproc{ref-dayPerceivedIdealInequality2023}{Day \& Norton, 2023}).
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+Despite the widespread recognition of education as a key factor for
+social mobility and equality of opportunities, disparities in this
+domain are far from being eliminated. As the last PISA Report points
+out, ``Socio-economically disadvantaged students are seven times more
+likely than advantaged students to score below Level 2 in mathematics on
+average across OECD countries''
+(\protect\hyperlink{ref-oecd_pisa_2023}{OECD, 2023}). Furthermore,
+empirical evidence from social science research has revealed an
+exacerbation of educational inequalities across various domains, such as
+family socioeconomic status, gender, ethnicity, and migratory
+background, particularly after COVID-19
+(\protect\hyperlink{ref-darmodyImpactsCOVID19Control2021}{Darmody et
+al., 2021}). Nevertheless, and despite the consistent empirical evidence
+on educational inequalities and their consequences, research on
+distributive justice indicates remarkable individual differences in the
+justification of unequal access to education
+(\protect\hyperlink{ref-leeFairnessPerceptionsEducational2023}{Lee \&
+Stacey, 2023}). In the present paper, we argue that the perception of
+meritocracy in society might play a role in the justification of
+economic inequalities in this particular domain.
+
+Meritocracy is defined as a distributive system where individual factors
+such as effort and talent are conceived as the main determinants of
+individual performance, usually sidelining the role of opportunities and
+social background. In recent years, there has been a growing interest in
+studying meritocratic beliefs as well as their variations according to
+individual and contextual variables
+(\protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021};
+\protect\hyperlink{ref-mijs_belief_2022}{Mijs et al., 2022}). This line
+of research has highlighted how meritocratic beliefs are associated with
+justifying unequal access to resources and rewards in society, by
+legitimizing social hierarchies in which those who already have
+resources and advantages are more likely to succeed
+(\protect\hyperlink{ref-hadjar_meritokratie_2008}{Hadjar, 2008};
+\protect\hyperlink{ref-mcnamee_meritocracy_2004}{McNamee \& Miller,
+2004}; \protect\hyperlink{ref-sandel_tyranny_2020}{Sandel, 2020}).
+Within this framework, this study focuses on the changes in the
+justification of inequalities in access to education in Chile, and to
+what extent this phenomenon is associated with the perception of
+meritocracy and its variations during the COVID pandemic.
+
+The Chilean case is particularly relevant when it comes to the study of
+inequalities in education. The country is characterized by a high level
+of economic inequality associated with the privatization and
+commodification of several social policy domains, among them the
+educational system (\protect\hyperlink{ref-bellei_estudio_2013}{Bellei,
+2013};
+\protect\hyperlink{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et
+al., 2016}). The neoliberal reforms implemented during and since the
+military dictatorship (1973-1989) have deepened educational policies
+based on private subsidies, achievement incentives, selection, and
+segregation (\protect\hyperlink{ref-joiko_cuasimercado_2019}{Joiko,
+2019}). Previous research in Chile has emphasized the concept of an
+``educational market''
+(\protect\hyperlink{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et
+al., 2016}), characterized by selection, competition, and the provision
+of goods and services associated with payment capacities. Many studies
+have addressed this problem from economic, political, and sociological
+perspectives, generating relevant evidence on their academic and social
+consequences (\protect\hyperlink{ref-ramos_educacion_2022}{Ramos,
+2022}). Likewise, massive social mobilizations in the country
+(particularly in 2006, 2011, and 2018) have had educational inequalities
+as one of the main targets, putting pressure on governments towards a
+series of reforms that have advanced free higher education reforms and
+have diminished the role of private actors in the educational system.
+
+This research endeavors to dissect the nuanced alterations in the
+justifications for educational disparities within the Chilean framework.
+Central to our investigation is the premise that the rationale
+underpinning these inequalities experiences a significant recalibration
+during periods of vulnerability and crises, specifically under the
+unprecedented strains of the global health and economic cataclysm
+triggered by the COVID-19 pandemic
+(\protect\hyperlink{ref-breznau_welfare_2021}{Breznau, 2021};
+\protect\hyperlink{ref-hankivsky_introduction_2022}{Hankivsky et al.,
+2022}). A pertinent aspect of this discourse is articulated in the works
+of Mijs (\protect\hyperlink{ref-mijs_stratified_2016}{2016},
+\protect\hyperlink{ref-mijs_visualizing_2018}{2018}), showing the
+interplay between meritocratic beliefs and the legitimization of
+disparities, as well as the potential erosion of meritocratic principles
+amidst crises. Within this framework, we point out that periods of
+vulnerability (such as the health and economic crisis resulting from the
+COVID-19 pandemic) could weaken the justification of social
+inequalities, a change that would be at least in part linked to a
+challenge of the meritocratic ideals. Furthermore, we suggest that the
+association between meritocratic perception and inequality justification
+would be weaker among lower-status groups, as they are the most affected
+in critical times.
+
+\hypertarget{inequality-beliefs-and-justification}{%
+\section{Inequality, beliefs and
+justification}\label{inequality-beliefs-and-justification}}
+
+The justification of economic inequalities is a controversial field of
+study. As equality is usually considered a synonym of justice in common
+sense, the idea that someone could be willing to justify inequalities
+seems (in principle) far from what is considerably reasonable.
+Nevertheless, and as pointed out by Barry, the whole normative
+discussion in social justice can be conceived as ``the defensibility of
+unequal relations between people''
+(\protect\hyperlink{ref-barry_theories_1989}{1989, p. 3}), whereby
+empirical research has consistently shown strong variations between
+individuals and societies regarding inequality justification
+(\protect\hyperlink{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021};
+\protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021};
+\protect\hyperlink{ref-trump_income_2018}{Trump, 2018}). Both
+theoretical and empirical approaches have been able to establish a clear
+difference between inequality and justice, which opens the question of
+to what extent inequality can be tolerated or even justified.
+
+The field of studies of subjective economic inequality has been
+extensively developed in the social sciences, particularly in the area
+of social justice research (\protect\hyperlink{ref-Resh2016}{Resh \&
+Sabbagh, 2016}). From a sociological point of view, the works of Kluegel
+\& Smith
+(\protect\hyperlink{ref-kluegelBeliefsStratification1981}{1981}) can be
+considered pioneers in framing a research area comprising perceptions,
+beliefs, and justifications in topics such as inequality of
+opportunities, poverty, and distributive justice. Theoretically, the
+authors understand the concept of ``beliefs'' in a ``broad sense to
+refer to the information (veridical or non-veridical) about a phenomenon
+that an individual uses as a basis both for inferring other information
+and for action.''
+(\protect\hyperlink{ref-kluegelBeliefsStratification1981}{Kluegel \&
+Smith, 1981, p. 30}). However, the literature has strained the concept
+of ``beliefs'' by more specific distinctions around the individual and
+structural dimensions of economic inequality. In this regard, Wegener
+(\protect\hyperlink{ref-wegenerIllusionDistributiveJustice1987}{1987})
+has suggested that perceptions of social status and reward hierarchy are
+relative to the observer's status in such hierarchies
+(\protect\hyperlink{ref-wegenerIllusionDistributiveJustice1987}{Wegener,
+1987, p. 2}). This approach established that status positions influence
+how people perceive things \emph{are} and, in turn, these perceptions
+anchor how people evaluate this inequality as just or unjust
+(\protect\hyperlink{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}).
+Additionally, Janmaat
+(\protect\hyperlink{ref-janmaatSubjectiveInequalityReview2013}{2013})
+has suggested more detailed conceptual distinctions about how people
+perceive economic inequality \emph{is}, how they believe or prefer it
+\emph{should be}, and the process of evaluating such differences as
+fair.
+
+Theoretically, Liebig \& Sauer
+(\protect\hyperlink{ref-liebigSociologyJustice2016}{2016}) establish
+three fundamental distinctions in justice attitudes in the field of
+sociological justice research. First, \emph{order-related attitudes}
+refer to individual motives that explain preferences for resource
+allocation based on the principles of equality, need, or entitlement.
+Second, \emph{procedural justice attitudes} account for the degree of
+acceptance of unfavorable outcomes under conditions in which the
+procedure under which those decisions were made is considered fair
+(e.g., opportunities). Finally, they can be \emph{outcome-related
+attitudes} which refer primarily to the economic resources that a person
+currently obtains and the set of collective commitments associated with
+possessing those resources (e.g., transfers or taxes). Within this
+framework, we conceptualize the justification of inequality in the first
+area of order-related attitudes, as it expresses preferences for
+resource allocation that can be associated with concepts that reflect
+different distributive principles
+(\protect\hyperlink{ref-kluegel_social_1995}{Kluegel et al., 1995};
+\protect\hyperlink{ref-sonhing_merit_2011}{Son Hing et al., 2011})
+
+An important finding in the area of order-related attitudes is that
+beliefs about inequality are stratified by social status
+(\protect\hyperlink{ref-janmaatSubjectiveInequalityReview2013}{Janmaat,
+2013};
+\protect\hyperlink{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \&
+Smith, 1986}). The evidence suggests that advantaged positions in terms
+of socioeconomic status, such as income, education, or occupational
+prestige, are associated with preferences for less egalitarian class
+imageries (\protect\hyperlink{ref-Evans1992}{Evans et al., 1992};
+\protect\hyperlink{ref-Evans2017}{Evans \& Kelley, 2017}) and greater
+justification of salary gaps
+(\protect\hyperlink{ref-hadler_why_2005}{Hadler, 2005};
+\protect\hyperlink{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}).
+Nevertheless, and in contrast to this perspective, it has been argued
+that human beings are intrinsically motivated to justify the system they
+inhabit, even when the information available to them points in the
+opposite direction (\protect\hyperlink{ref-jost_psychology_2003}{Jost \&
+Hunyady, 2003}; \protect\hyperlink{ref-vandertoorn_twenty_2014}{van der
+Toorn \& Jost, 2014}). Individuals would be generally disposed to
+justify the economic status quo regardless of self- and group economic
+interest because their support partially depends on their beliefs in
+system-justifying ideologies
+(\protect\hyperlink{ref-trump_when_2020}{Trump, 2020}). As a result,
+what has been suggested is that low-status might perceive the status quo
+as just as they are more strongly attached to dominant ideologies, while
+high-status individuals are more prone to justify the system to a
+greater extent because it favors their interests and dominant position
+(\protect\hyperlink{ref-jost_theory_2020}{Jost, 2020}).
+
+Another factor that has been linked to inequality beliefs is the
+perception of economic inequality
+(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
+2022};
+\protect\hyperlink{ref-kelleyLegitimationInequalityOccupational1993}{Kelley
+\& Evans, 1993}; \protect\hyperlink{ref-osberg_fair_2006}{Osberg \&
+Smeeding, 2006}). Studies in this matter have argued that perceived
+economic inequality functions as an anchor for the justifications of
+inequality because if these inequalities are indeed seen as fair in
+their generative processes, then they will be considered as just
+(\protect\hyperlink{ref-wegenerIllusionDistributiveJustice1987}{Wegener,
+1987}, \protect\hyperlink{ref-Wegener1990}{1990}). Thus, both the role
+of social status and economic inequality perceptions have found support
+in correlational
+(\protect\hyperlink{ref-Castillo2012a_justice}{Castillo, 2012};
+\protect\hyperlink{ref-schroder_income_2017}{Schröder, 2017}), as well
+as in experimental studies that have manipulated inequality perception
+(\protect\hyperlink{ref-iturra_percepcion_2023}{Iturra et al., 2023};
+\protect\hyperlink{ref-trump_income_2018}{Trump, 2018}). In addition,
+normative or value-driven mechanisms can also explain to what extent
+people justify inequality. For instance, studies suggest that justice
+ideologies such as individualism or egalitarianism can mediate the
+influence of individual status on income inequality justification
+(\protect\hyperlink{ref-Castillo2011}{Castillo, 2011}). Likewise, system
+justification ideologies can strengthen the influence of socioeconomic
+status on economic inequality preferences
+(\protect\hyperlink{ref-garcia-sanchez_creencias_2022}{García-Sánchez \&
+De Carvalho, 2022}). Altogether, these studies have shown that not only
+socio-structural factors influence what people believe regarding the
+distribution of economic resources, but their normative stances and
+perceptions can also determine preferences for economic inequality in
+diverse domains.
+
+\hypertarget{the-justification-of-educational-inequality}{%
+\section{The justification of educational
+inequality}\label{the-justification-of-educational-inequality}}
+
+Studies addressing inequality justification have assessed the support
+for market ideologies, salary gaps, and distributive preferences. A
+strand of this research has focused on specific policy areas such as
+health, pensions, and education, attending to the degree of
+justification of market-based access to social services. For instance,
+Lindh (\protect\hyperlink{ref-lindhPublicOpinionMarkets2015}{2015})
+found that public support for the market distribution of social services
+is generally low in most countries, implying that most people believe it
+is unfair for market forces to determine access to basic social
+services. Similarly, Busemeyer \& Iversen
+(\protect\hyperlink{ref-busemeyer_welfare_2020}{2020}) shows that
+high-income citizens are less supportive of expanding public spending on
+universal benefit schemes. Instead, they tend to prefer a public basic
+insurance scheme, even if it provides relatively more benefits to
+low-income individuals.
+
+The study of inequality justification in the sphere of education is a
+rather specific and unattended field. Most studies on educational
+inequality relate performance indicators (such as grades and
+standardized test results) to stratification-related variables, whereby
+few refer specifically to if and how individuals tolerate and even
+justify educational inequalities. This is probably because in principle
+it seems counter-intuitive to think that someone would be willing to
+justify differences in such a sensitive area. This paradox is phrased by
+Conforti
+(\protect\hyperlink{ref-confortiLegitimationInequalityAmerican1992}{1992})
+as ``while equality is embraced as a value, inequality is accepted as a
+reality'' (p.230).
+
+In the area of empirical studies, the operationalization of educational
+inequality justification in terms of questionnaire items is diverse, and
+there is not a single instrument or scale dedicated specifically to this
+concept. Rather, some attitudinal surveys usually include
+education-related items that are posteriorly used as dependent or
+independent variables. Some examples are the International Social Survey
+Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that
+people with higher incomes can buy better education for their children
+than people with lower incomes?}), and the European Social Survey (ESS)
+2018 (\emph{Compared to other people in {[}country{]}, I have had a fair
+chance of achieving the level of education I was seeking}). Although
+both questions have been used to measure the justification of
+educational inequality, they refer to different aspects. The ESS can be
+classified as reflexive, as the respondent is involved as a referent for
+the question, whereas the ISSP question is non-reflexive. Furthermore,
+ISSP refers to educational access, while the ESS refers to educational
+outputs.
+
+Studies have addressed the justification of educational inequality from
+different conceptual perspectives and methodological strategies. Some
+perspectives emphasize the justification of differences between specific
+groups, finding for instance that Americans are more concerned
+about---and more supportive of proposals to close---wealth-based
+achievement gaps than Black-White or Hispanic-White gaps
+(\protect\hyperlink{ref-valantPoliticsAchievementGaps2016}{Valant \&
+Newark, 2016}). Other studies focus on the association between status
+differences and inequality justification in education, based on the
+rational premise that those who are worst-off would tend to show lower
+levels of justification
+(\protect\hyperlink{ref-stoilovaFairnessEducationalOpportunities2023}{Stoilova
+\& Ilieva-Trichkova, 2023}). However, most of the research in this area
+goes beyond status differences, focusing on how subjective factors such
+as perceptions and beliefs are related to different levels of
+justification.
+
+The perception of economic inequality is one of the factors associated
+with the justification of educational inequality. As pointed out by
+Janmaat
+(\protect\hyperlink{ref-janmaatSubjectiveInequalityReview2013}{2013}),
+justifications are usually confused under the labels of ``beliefs'' and
+``attitudes,'' and their distinction from perceptions is relevant as it
+helps state that justifying a certain level of inequality requires
+taking into account which level of (perceived) inequality is first and
+foremost being recognized
+(\protect\hyperlink{ref-castillo_multidimensional_2023}{Castillo et al.,
+2023}; \protect\hyperlink{ref-sonhing_merit_2011}{Son Hing et al.,
+2011}). For instance, a recent study by Day \& Norton
+(\protect\hyperlink{ref-dayPerceivedIdealInequality2023}{2023}) on the
+perception and justification of inequality in university endowments in
+the U.S. points out that individuals tend to underestimate the magnitude
+of existing inequality while simultaneously desiring greater equality.
+Furthermore, through an experiment, they demonstrate that information
+about inequality in university endowments (i.e., manipulating inequality
+perception) increases the perception of injustice in education. Results
+of this study are in line with the consistent evidence on biases related
+to the underestimation of economic inequality
+(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
+2022}; \protect\hyperlink{ref-gimpelson_misperceiving_2018}{Gimpelson \&
+Treisman, 2018})
+
+Besides focusing on the relationship between perception and
+justification of educational inequality, another group of studies has
+addressed how the justification of inequality is associated with the
+support of different distributive principles. In an experiment by
+Igliozzi et al.
+(\protect\hyperlink{ref-igliozziFairShareEffects2024}{2024}), they
+establish how the support for different justice principles (equity,
+equality, and need) is related to support for redistributive policies in
+domains such as education. They find that a larger system justification
+(\protect\hyperlink{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}) is
+related to higher support for the principles of equity and equality
+(over need) in distributing educational outcomes, where equity was
+measured in terms of access to education related to performance
+(`Students who place within the highest 50\% of testing scores from the
+previous year can attend the magnet school'). In the same line, Lee \&
+Stacey
+(\protect\hyperlink{ref-leeFairnessPerceptionsEducational2023}{2023})
+show that performance assessments (in what they call ``neoliberal
+orientations'') predicted people's fair perceptions of educational
+inequality.
+
+\emph{Justificacion, inequality and meritocracy in the educational
+domain}
+
+The relationship between performance and outputs brings us to the
+concepts of merit and meritocracy, key in educational justice research.
+The concept of meritocracy, coined by the sociologist Michael Young in
+his satirical work ``The Rise of the Meritocracy''
+(\protect\hyperlink{ref-young_rise_1962}{1962}), has a rich history and
+multifaceted meaning. Young's term was initially used to critique a
+hypothetical future society where social and economic positions were
+entirely determined by individuals' merits, talents, and efforts,
+ostensibly eliminating the influence of inherited privilege. However,
+over time, the term has evolved to describe a social system where
+advancement and success are perceived as directly linked to one's
+abilities and achievements, sidelining inequalities of opportunities
+(\protect\hyperlink{ref-bell_politics_2020}{Bell, 2020}). This would
+contribute to the justification of inequality in the educational domain
+(\protect\hyperlink{ref-sandel_tyranny_2020}{Sandel, 2020}), as
+individual ability is often inherited, and educational attainment is
+influenced by family background
+(\protect\hyperlink{ref-goldin_meritocracy_2000}{Goldin, 2000}).
+
+Empirical research examining the relationship between meritocratic
+perceptions and the justification of educational inequality provides
+valuable insights into the dynamics shaping individuals' attitudes and
+beliefs. Some studies have explored the extent to which people believe
+that educational systems are meritocratic and how these perceptions
+influence their acceptance or rejection of educational disparities, as
+well as status differences (\protect\hyperlink{ref-McCoy2007}{McCoy \&
+Major, 2007}). This belief is particularly influential in educational
+institutions, where it not only encourages academic success but also
+legitimizes social and income inequality
+(\protect\hyperlink{ref-batruch_belief_2023}{Batruch et al., 2023}).
+Most of the evidence so far suggests that those who believe in a direct
+link between effort, talent, and educational success are more inclined
+to rationalize existing disparities in educational outcomes. However,
+empirical findings also reveal nuances in this relationship. For
+instance, studies show that when individuals perceive structural
+barriers, such as economic disparities or discrimination, as hindering
+the meritocratic ideal, they may be less likely to justify educational
+inequalities
+(\protect\hyperlink{ref-dayPerceivedIdealInequality2023}{Day \& Norton,
+2023}). This suggests that while meritocratic beliefs play a significant
+role, individuals' awareness of broader societal issues can mediate
+their willingness to rationalize educational disparities.
 
@@
 
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}).
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-\(H_1\): The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences (\citeproc{ref-hing_failure_2019}{Hing et al., 2019}), expecting that:
-
-\(H_2\): Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-\(H_3\): A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-\(H_4\): Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-\(H_5\): The positive association between meritocracy and inequality justification mitigates along time
-
-\section{Data, variables \& methods}\label{data-variables-methods}
-
-\subsection{Data}\label{data}
-
-The main data source is the Chilean Longitudinal Social Survey 2016--2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77\% of the total national population and 93\% of the urban areas (\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}).
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4\% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41\%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit \url{https://coes.cl/encuesta-panel/}.
-
-\subsection{Variables}\label{variables}
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: ``It is just that high-income people have a better education for their children than people with lower incomes'' (``Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos'' in Spanish).
-Here, the respondents declared their preferences on a Likert scale from ``strongly disagree'' (1) to ``strongly agree'' (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort (``In Chile, people are rewarded for their efforts''), and the other related to talent (``In Chile, people are rewarded for their intelligence and skills'').
-In addition, we included an indicator for economic inequality perception (``In Chile, the income differences are too large'').
-Figure \ref{fig:frecuencies} shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2\%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7\%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6\%) and for their efforts (57\%).
+The present research uses longitudinal survey data from 2016 to 2022 to
+asses changes in inequality justification. As there are measures before
+and after the COVID pandemic, firstly we explore to what extent it is
+possible to identify variations in inequality justification in this time
+frame. Although we are aware that it is not possible to make a causal
+claim given that this period is characterized by many other
+circumstances besides COVID, we argue that the increase of risk
+perceptions in critical situations (such as the pandemic) impacts public
+concerns about equality and redistribution
+(\protect\hyperlink{ref-breznau_welfare_2021}{Breznau, 2021}). Although
+the pandemic might have triggered such concerns mostly regarding health,
+we believe that other policy areas, such as education, could have been
+impacted in this period. Consistently, our first hypothesis is:
+
+\(H_1\): The justification of educational inequalities decreases after
+the pandemic
+
+Our second argument is based on the evidence that relates rational
+interests with justifications and preferences
+(\protect\hyperlink{ref-sonhing_failure_2019}{Son Hing et al., 2019}),
+expecting that:
+
+\(H_2\): Those with lower status characteristics show lower
+justification of educational inequalities
+
+Thirdly, we address inequality perception. As referred to above, the
+justification of inequality requires, first and foremost, that such
+inequality is recognized. As a larger inequality in this domain
+threatens a basic equality ideal, the corresponding hypothesis in this
+case is:
+
+\(H_3\): A larger inequality perception is negatively associated with
+the justification of educational inequalities.
+
+Our fourth hypothesis refers to the central concept of meritocracy. As
+meritocracy justifies inequalities based on performance, we expect that:
+
+\(H_4\): Higher meritocratic perceptions are positively related to the
+justification of educational inequalities.
+
+Finally, and in line with hypotheses 1 and 4, we explore interactions
+between meritocracy and inequality justification over time. As COVID-19
+pandemic has affected the structural opportunities to attend schools,
+universities, or workplaces, the ideal of meritocracy might have been
+challenged. As a consequence, the justification of inequality could show
+a decrease over time:
+
+\(H_5\): The positive association between meritocracy and inequality
+justification mitigates along time
+
+\hypertarget{data-variables-methods}{%
+\section{Data, variables \& methods}\label{data-variables-methods}}
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+The main data source is the Chilean Longitudinal Social Survey
+2016--2022. The ELSOC has been designed as a yearly panel study to
+evaluate how individuals think, feel, and behave regarding a set of
+social issues concerning conflict and social cohesion in Chile. The
+sampling design is probabilistic, stratified, clustered, and multistage.
+It provides adequate coverage of the country's largest cities
+(Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller
+cities. Consequently, the ELSOC comprises a total of 2,927 participants
+aged between 18 and 75 years in wave 1. In addition, the sample
+represents 77\% of the total national population and 93\% of the urban
+areas (\protect\hyperlink{ref-elsoc_estudio_2022}{ELSOC, 2022}).
+
+The survey has been conducted yearly since 2016, except for the year
+2020, when it was suspended due to the pandemic. The waves 2016, 2017,
+2018, 2019 and 2022 were administered using computer-assisted personal
+interviewing (CAPI). However, a reduced version was conducted using
+computer-assisted telephone interviews (CATI) in 2021. In addition, wave
+3 included a refreshment sample to counter survey attrition. The same
+sampling strategy as that of wave 1 was implemented for selecting new
+cases. As a result, the total sample of wave 3 included 3,748 cases, of
+which 2,229 were part of the original sample and 1,519 were from the
+refreshment sample. The data from the refreshment sample is not included
+in the analytical sample because we wanted to analyze longer response
+trends. Regarding the original sample, the response rate was 62.4\% in
+wave 1, achieving N = 2,927 participants. In broader terms, the
+accumulated attrition between wave 1 and wave 6 is 41\%, achieving a
+final sample of N = 1728. The analysis used longitudinal weights to
+avoid bias due to the sampling design, allowing us to control for biases
+arising from systematic patterns of non-participation in the survey
+after the first wave. For a more detailed analysis of responses,
+attrition, and the construction of longitudinal weights, visit
+\url{https://coes.cl/encuesta-panel/}. The dataset is publicly available
+at: \url{https://dataverse.harvard.edu/dataverse/elsoc}.
+
+\hypertarget{variables}{%
+\subsection{Variables}\label{variables}}
+
+The dependent variable of this study is the justification of educational
+inequality. This construct is measured using the following statement:
+``It is just that high-income people have a better education for their
+children than people with lower incomes'' (``Es justo que las personas
+de altos ingresos tengan una mejor educación para sus hijos que las
+personas con ingresos más bajos'' in Spanish). Here, the respondents
+declared their preferences on a Likert scale from ``strongly disagree''
+(1) to ``strongly agree'' (5). The main independent variables refer to
+meritocratic perception, which is operationalized by two items, one
+related to effort (``In Chile, people are rewarded for their efforts''),
+and the other related to talent (``In Chile, people are rewarded for
+their intelligence and skills''). In addition, we included an indicator
+for economic inequality perception (``In Chile, the income differences
+are too large''). Figure @ref(fig:frecuencies) shows the average
+distribution throughout the years for these variables. We observe that
+there is a large group (88.2\%) of respondents who disagree or strongly
+disagree that people with higher incomes can access better education.
+Likewise, a large group of respondents (91.7\%) agree or strongly agree
+that income differences in Chile are too large. On the other hand, there
+is an important level of disagreement (disagree + strongly disagree)
+about the perception of meritocracy, this is, whether people are
+rewarded for their talent (47.6\%) and for their efforts (57\%).
 
@@
 
-\caption{Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:frecuencies}
+\caption{Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}(\#fig:frecuencies)
 \end{figure}
 
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \ref{tab:descriptives} shows the whole set of independent variables, the corresponding items, response categories, and frequencies.
+For testing the hypothesis regarding status we included educational
+level and household income quintiles, as well as subjective social
+status. Finally, political identification on a left-right scale, age,
+and gender were included as control variables. Table
+@ref(tab:descriptives) shows the whole set of independent variables, the
+corresponding items, response categories, and frequencies.
 
 \begin{longtable}[]{@{}l@{}}
-\caption{\label{tab:descriptives}Independent variables}\tabularnewline
+\caption{(\#tab:descriptives)Independent variables}\tabularnewline
 \toprule\noalign{}
@@
 
-\subsection{Methods}\label{methods}
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009}).
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library ``lme4'' (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009, p.
-4}).
-
-\section{Results}\label{results}
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
+
+Given the hierarchical structure of the data (observations nested in
+survey waves), we estimated a series of longitudinal multilevel models
+(\protect\hyperlink{ref-singer_applied_2009}{Singer \& Willett, 2009}).
+Such an approach is suited to account for the shared variance among
+units (in this case individuals), adjusting the estimation of the
+standard errors. The linear multilevel models are estimated using the R
+library ``lme4'' (\protect\hyperlink{ref-singer_applied_2009}{Singer \&
+Willett, 2009, p. 4}).
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 
-\caption{Change in the justification of educational inequality over time (2016-2022)}\label{fig:alluvial}
+\caption{Change in the justification of educational inequality over time (2016-2022)}(\#fig:alluvial)
 \end{figure}
 
-Figure \ref{fig:alluvial} illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data (\citeproc{ref-rosvall_mapping_2010}{Rosvall \& Bergstrom, 2010}).
-For instance, of the 32.2\% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90\% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
+Figure @ref(fig:alluvial) illustrates yearly frequencies in the
+justification of educational inequality between 2016 and 2022. Each year
+represents stacked percentual frequencies, and the flows in between
+reflect the within-subject change of opinions from one year to the next,
+as we are using longitudinal panel data
+(\protect\hyperlink{ref-rosvall_mapping_2010}{Rosvall \& Bergstrom,
+2010}). For instance, of the 32.2\% who strongly disagreed with
+inequality justification in 2016, about half of them kept responding the
+same in 2017, whereas the other half shifted their opinion to other
+response categories. In general, the large majority - between 80 and
+90\% - disagrees with the justification of educational inequality
+throughout the years. Despite this overall tendency, we also observe
+that the disagreement with inequality justification (disagree + strongly
+disagree) tends to go down in the last wave. This change is mostly a
+result of the radical decrease in the lowest response category (strongly
+disagree), which diminishes to half when compared with previous years.
 
@@
 
-\caption{Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years}\label{fig:means}
+\caption{Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years}(\#fig:means)
 \end{figure}
 
-Figure \ref{fig:means} shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
+Figure @ref(fig:means) shows the average changes in the main variables
+considered for this study. Here we observe that the justification of
+educational inequality has the lowest average throughout the years when
+compared with the other (independent) variables, whereas the highest
+average is consistently represented by inequality perception.
+Interestingly, in the last wave of the study (2022) the justification of
+inequality increases whereas the perception of inequality decreases. As
+the merit variables are concerned, they show a very similar pattern in
+terms of averages and changes over the years, being the perception of
+meritocracy related to effort always lower than the one associated with
+talent.
 
@@
 
-\caption{Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:correlation}
+\caption{Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}(\#fig:correlation)
 \end{figure}
 
-Figure \ref{fig:correlation} presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (\emph{r}=-0.29, p\textless.01) and a moderate and positive association with both meritocracy perception variables (\emph{r}=0.18, p\textless.01; \emph{r}=0.15, p\textless.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (\emph{r}=-0.15, p\textless.01; \emph{r}=-0.10, p\textless.01).
-Finally, the two meritocracy variables present a high and positive association with each other (\emph{r}=0.74, p\textless.01).
-
-\subsection{Multivariate analysis}\label{multivariate-analysis}
-
-Table \ref{tab:multilevel} shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 (\(\beta\)=-0.15, p\textless.001), 2018 (\(\beta\)=-0.06, p\textless.05), and 2019 (\(\beta\)=-0.07, p\textless.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality (\(\beta\)=0.21, p\textless.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 (\(\beta\)=0.04, p\textless.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality (\(\beta\)=0.05, p\textless.01), but it has a larger weight than the perception of talent.
+Figure @ref(fig:correlation) presents a correlation matrix of the main
+variables analyzed, using data from all survey waves. In this matrix,
+the correlations vary between low and moderate values. Justification of
+economic inequality depicts a moderate and negative association with the
+perception of inequality (\emph{r}=-0.29, p\textless.01) and a moderate
+and positive association with both meritocracy perception variables
+(\emph{r}=0.18, p\textless.01; \emph{r}=0.15, p\textless.01). Regarding
+the perception of inequality, it presents a moderate and negative
+association with both meritocracy perception variables (\emph{r}=-0.15,
+p\textless.01; \emph{r}=-0.10, p\textless.01). Finally, the two
+meritocracy variables present a high and positive association with each
+other (\emph{r}=0.74, p\textless.01).
+
+\hypertarget{multivariate-analysis}{%
+\subsection{Multivariate analysis}\label{multivariate-analysis}}
+
+Table @ref(tab:multilevel) shows the multilevel estimation results for
+the justification of educational inequality. Model 1 includes the survey
+waves to estimate intertemporal changes in the dependent variable.
+Taking 2016 as a reference point, we can observe a staggered decrease in
+2017 (\(\beta\)=-0.15, p\textless.001), 2018 (\(\beta\)=-0.06,
+p\textless.05), and 2019 (\(\beta\)=-0.07, p\textless.01). Nevertheless,
+in the last wave of 2022 there is a radical increase in level of
+justification of economic inequality (\(\beta\)=0.21, p\textless.001),
+suggesting a non-linear change in this variable. Attempting to model
+this path of change over time, Model 2 incorporates time (survey waves)
+as a continuous variable as well as its quadratic term, representing the
+nonlinear association initially observed in Model 1. On the one hand,
+the survey wave depicts a negative association, expressing an average
+decrease in inequality justification over time, but on the other hand,
+the quadratic wave term is positive, indicating the reversion of this
+path in the last measurement point.
+
+Model 3 adds the sociodemographic variables, where having a higher
+income and being older (65 or older compared to 18-29) have a
+significant and positive influence on the justification of educational
+inequality. On the contrary, the gender variable has a negative effect,
+expressing that women on average justify less inequality in education
+when compared to males. Educational level and subjective social status
+are not shown in this table for the sake of space (as they do not
+exhibit significant effects), but the complete estimation of the models
+is presented in the appendix. Model 3 also includes the control
+variables, where identifying oneself with the center, right-wing, or
+when reporting no political identification, has a significant and
+positive association on the justification of educational inequality -
+when compared to being left-wing oriented. Inequality perception is
+added in Model 4, showing a negative association with the justification
+of educational inequality as hypothesized, remaining stable when
+controlling for the rest of the variables.
+
+Models 5 and 6 introduce the meritocratic variables: talent (if
+intelligence and abilities are rewarded in society) and effort (if
+efforts are rewarded in society). In line with our hypotheses, the
+perception that talent is rewarded has a positive influence on the
+justification of educational inequality in Model 5 (\(\beta\)=0.04,
+p\textless.01). However, when controlling for the perception that effort
+is rewarded, this effect is no longer significant. In this sense, Model
+6 shows that the perception that effort is rewarded in society is not
+only positively associated with the justification of educational
+inequality (\(\beta\)=0.05, p\textless.01), but it has a larger weight
+than the perception of talent.
 
 \begin{table}[!ht]
-\caption{\label{tab:multilevel} Multilevel longitudinal models for the justification of inequality in education}
+\caption{(\#tab:multilevel) Multilevel longitudinal models for the justification of inequality in education}
 \begin{center}
@@
 \begin{table}[!ht]
-\caption{\label{tab:interact} Interaction effects for the justification of economic inequality}
+\caption{(\#tab:interact) Interaction effects for the justification of economic inequality}
 \begin{center}
@@
 
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \ref{tab:interact}.
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable (\(\beta\)=0.01, p\textless.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-\section{Discussion and conclusions}\label{discussion-and-conclusions}
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process (\citeproc{ref-palanza_chile_2023}{Palanza \& Sotomayor, 2023}; \citeproc{ref-sazo_chile_2023}{Sazo, 2023}).
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality (\(H_2\)).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude (\(H_3\)).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education (\(H_4\)).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis (\(H_5\)).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
+In this final part of the analysis, we contrast longitudinal hypotheses
+regarding changes in the relationship between meritocracy and
+justification of economic inequality over time. In hypothesis 5 we
+proposed that the association between the perception of meritocracy and
+inequality justification mitigates over time, as meritocratic ideals
+could have weakened due to critical situations associated with the COVID
+pandemic. We test this hypothesis through interaction effects, displayed
+in Table @ref(tab:interact). Model 1 is shown as a baseline model, it is
+the same as Model 6 in Table @ref(tab:multilevel) but it only displays
+the variables involved in the interaction for the sake of space (all
+other variables are controlled for). Model 2 adds the interaction
+between the perception of meritocracy based on talent and time (panel
+wave), whereas Model 3 does the same but now for the perception of
+meritocracy related to effort. As observed, only the effort variable
+shows a significant interaction with the time variable (\(\beta\)=0.01,
+p\textless.05), meaning that, on average, the association between the
+perception of meritocracy based on effort and the justification of
+inequality increases by 0.01 points in every measurement point.
+
+\hypertarget{discussion-and-conclusions}{%
+\section{Discussion and conclusions}\label{discussion-and-conclusions}}
+
+Our research sought to examine the relationship between the perception
+of meritocracy and the justification of inequality in education from a
+longitudinal perspective amid the health crisis generated by the
+COVID-19 pandemic. In general, our results showed mixed evidence about
+our hypotheses. Whereas we find support for the association between the
+perception of meritocracy and justification of inequality across time,
+the analyses that involved changes in justification of inequality are
+against our initial proposal. We argued that in times of crisis, greater
+exposure to risk due to the health context would result in a lower
+justification of inequality by citizens. In this regard, longitudinal
+models showed that the justification of inequality in education is far
+from being linear, and whereas a decreasing pattern was found between
+2017 and 2019, there is a striking increase in the last measurement
+point (2022), after the peak of the COVID pandemic. As we mentioned at
+the beginning, given our data limitations it is not possible to
+attribute such changes only to the pandemic and the related economic and
+social crises. As in the Chilean case, the pandemic period occurred
+along with a constitutional process, disentangling different forces
+driving public opinion it becomes difficult and deserves further
+qualitative and quantitative studies. A closer look to the political
+environment during the last year in the country could give additional
+hints. Some recent studies have argued that the election of a far
+left-wing constitutional assembly during this period generated a
+backslash effect, mainly due to several scandals that led to the
+delegitimation of this assembly. This could have driven preferences in a
+conservative direction, which was reflected in the election of a
+right-wing second assembly after the failed first constitutional process
+(\protect\hyperlink{ref-palanza_chile_2023}{Palanza \& Sotomayor, 2023};
+\protect\hyperlink{ref-sazo_chile_2023}{Sazo, 2023}).
+
+Regarding the status position, our central hypothesis was that
+individuals in more advantaged situations and with greater resources
+would tend to defend their interests and justify greater education
+inequality (\(H_2\)). In this regard, our results show a positive and
+robust relationship between income and the justification of inequality,
+which would align with our original hypothesis based on the rational
+interests of the better-off that would lead to a larger inequality
+justification. Besides status, we also considered inequality perception,
+as the literature on attitudes toward inequality has argued that a
+greater justification of inequality is associated with its perceived
+magnitude (\(H_3\)). Consistent with this hypothesis, our results showed
+that, throughout the years, a larger perceived economic inequality
+motivates a lower justification of inequality in education.
+
+The main focus of this study was on the relationship between meritocracy
+and justification of inequality, arguing that those who perceive that
+the society in which they live complies with meritocratic principles,
+would be more inclined to justify inequality in education (\(H_4\)). At
+first look, the results are consistent with the hypothesis; however,
+some nuances are worth attending concerning different aspects of
+meritocracy, namely effort and talent. Our results are favorable for the
+effort dimension but not confirmed for the talent dimension. In other
+words, the perception of meritocracy in terms of rewarded effort is more
+relevant in justifying access to education than the perception of
+talent. This could be explained as that talent could be more associated
+with luck in terms of random assignment, and in this terms would not be
+enough reason (as it is effort) to justify educational inequality.
+
+Finally, as our first hypothesis supported, we expected that the
+relationship between meritocratic perceptions and justification would
+tend to diminish over time. However, longitudinal evidence shows that
+the perception of meritocracy linked to effort strengthens its relation
+with the justification of inequality in education, against the temporal
+change hypothesis (\(H_5\)). In other words, meritocracy seems to gain
+force concerning justifying inequality and not the other way around as
+we expected given the social context in the post-covid era. An
+alternative interpretation is that in times of high exposure to risk,
+people tend to stick more to narratives that provide greater security
+regardless of the rest of the society - such as individual effort -,
+which might also be accompanied by lower degrees of solidarity to
+safeguard the well-being of themselves and their surrounding close
+peers. Again, this certainly needs further studies that include
+additional specific conceptual and methodological approaches to this
+phenomenon.
 
 Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}), the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
+First, we know that inequalities in education can manifest themselves in
+terms of outcomes as well as in areas such as access to social services
+in education as part of welfare policies. In this regard, the construct
+that we have captured with our indicator can be framed more in the
+latter, while the evaluation of justice involved in this process does
+not consider aspects such as equality of educational opportunities or
+observable outcomes such as the performance of individual or educational
+institutions. Second, as it has been established by recent research
+(\protect\hyperlink{ref-castillo_multidimensional_2023}{Castillo et al.,
+2023}), the perception of meritocracy can be understood through
+individual attributions regarding effort and talent, but also as a
+function of structural aspects such as family of origin status or social
+capital. In this sense, the available measurements restricted us to the
+use of single indicators for effort and talent, and further research
+should consider improving the measurement quality in this regard.
+Finally, there are limitations involving the longitudinal dimension of
+the data used. Given constraints for data collection during the
+pandemic, the ELSOC study had to shorten the questionnaire in 2021 and
+unfortunately, the educational justice indicator was excluded from that
+survey. Consequently, the results of temporal change have to be
+considered carefully, and further data waves could give us more
+information on this matter.
+
+The capabilities of the ELSOC longitudinal database are not limited to
+micro-level estimates. In this regard, thanks to the sampling strategy
+of the survey, it would be possible to make contextual estimates at the
+municipality level in future studies. For instance, statistical models
+could include contextual information from administrative data sources,
+allowing testing for hypotheses that include contextual socioeconomic
+inequality as well as its dynamics over time. Furthermore, this dataset
+could be used for comparing inequality justification in policy areas
+besides education, such as health and pensions. Future studies may shed
+light on how meritocratic perceptions and beliefs affect differently
+such areas and their changes over time, which would give relevant hints
+from social sciences for the discussion about societal and cultural
+changes as well as their impacts on solidarity and social cohesion.
 
@@
 
-\emph{Acknowledgments}
-
-The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022\_046
-
-\section{References}\label{references}
-
-\phantomsection\label{refs}
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-barry_theories_1989}
+\leavevmode\vadjust pre{\hypertarget{ref-barry_theories_1989}{}}%
 Barry, B. (1989). \emph{Theories of justice}. Univ. of California Pr.
 
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
+\leavevmode\vadjust pre{\hypertarget{ref-batruch_belief_2023}{}}%
 Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
-(2023).
-Belief in {School Meritocracy} and the {Legitimization} of {Social} and {Income Inequality}.
-\emph{Social Psychological and Personality Science}, \emph{14}(5), 621--635.
+(2023). Belief in {School Meritocracy} and the {Legitimization} of
+{Social} and {Income Inequality}. \emph{Social Psychological and
+Personality Science}, \emph{14}(5), 621--635.
 \url{https://doi.org/10.1177/19485506221111017}
 
-\bibitem[\citeproctext]{ref-bell_politics_2020}
-Bell, E.
-(2020).
-The {Politics} of {Designing Tuition-Free College}: {How Socially Constructed Target Populations Influence Policy Support}.
+\leavevmode\vadjust pre{\hypertarget{ref-bell_politics_2020}{}}%
+Bell, E. (2020). The {Politics} of {Designing Tuition-Free College}:
+{How Socially Constructed Target Populations Influence Policy Support}.
 \emph{The Journal of Higher Education}, \emph{91}(6), 888--926.
@@
 
-\bibitem[\citeproctext]{ref-bellei_estudio_2013}
-Bellei, C.
-(2013).
-El estudio de la segregaci{ó}n socioecon{ó}mica y acad{é}mica de la educaci{ó}n chilena.
-\emph{Estudios Pedag{ó}gicos (Valdivia)}, \emph{39}(1), 325--345.
+\leavevmode\vadjust pre{\hypertarget{ref-bellei_estudio_2013}{}}%
+Bellei, C. (2013). El estudio de la segregaci{ó}n socioecon{ó}mica y
+acad{é}mica de la educaci{ó}n chilena. \emph{Estudios Pedag{ó}gicos
+(Valdivia)}, \emph{39}(1), 325--345.
 \url{https://doi.org/10.4067/S0718-07052013000100019}
 
-\bibitem[\citeproctext]{ref-breznau_welfare_2021}
-Breznau, N.
-(2021).
-The welfare state and risk perceptions: The {Novel Coronavirus Pandemic} and public concern in 70 countries.
-\emph{European Societies}, \emph{23}(sup1), S33--S46.
+\leavevmode\vadjust pre{\hypertarget{ref-breznau_welfare_2021}{}}%
+Breznau, N. (2021). The welfare state and risk perceptions: The {Novel
+Coronavirus Pandemic} and public concern in 70 countries. \emph{European
+Societies}, \emph{23}(sup1), S33--S46.
 \url{https://doi.org/10.1080/14616696.2020.1793215}
 
-\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
-Busemeyer, M.
-R., \& Iversen, T.
-(2020).
-The {Welfare State} with {Private Alternatives}: {The Transformation} of {Popular Support} for {Social Insurance}.
-\emph{The Journal of Politics}, \emph{82}(2), 671--686.
-\url{https://doi.org/10.1086/706980}
-
-\bibitem[\citeproctext]{ref-Castillo2011}
-Castillo, J.
-C.
-(2011).
-Legitimacy of {Inequality} in a {Highly Unequal Context}: {Evidence} from the {Chilean Case}.
-\emph{Social Justice Research}, \emph{24}(4), 314--340.
+\leavevmode\vadjust pre{\hypertarget{ref-busemeyer_welfare_2020}{}}%
+Busemeyer, M. R., \& Iversen, T. (2020). The {Welfare State} with
+{Private Alternatives}: {The Transformation} of {Popular Support} for
+{Social Insurance}. \emph{The Journal of Politics}, \emph{82}(2),
+671--686. \url{https://doi.org/10.1086/706980}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Castillo2011}{}}%
+Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
+Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
+Research}, \emph{24}(4), 314--340.
 \url{https://doi.org/10.1007/s11211-011-0144-5}
 
-\bibitem[\citeproctext]{ref-Castillo2012a_justice}
-Castillo, J.
-C.
-(2012).
-Is {Inequality Becoming Just}?
-{Changes} in {Public Opinion} about {Economic Distribution} in {Chile}.
+\leavevmode\vadjust pre{\hypertarget{ref-Castillo2012a_justice}{}}%
+Castillo, J. C. (2012). Is {Inequality Becoming Just}? {Changes} in
+{Public Opinion} about {Economic Distribution} in {Chile}.
 \emph{Bulletin of Latin American Research}, \emph{31}(1), 1--18.
@@
 
-\bibitem[\citeproctext]{ref-castillo_perception_2022}
-Castillo, J.
-C., García-Castro, J.-D., \& Venegas, M.
-(2022).
-Perception of economic inequality: Concepts, associated factors and prospects of a burgeoning research agenda.
-\emph{International Journal of Social Psychology}, \emph{37}(1), 180--207.
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_perception_2022}{}}%
+Castillo, J. C., García-Castro, J.-D., \& Venegas, M. (2022). Perception
+of economic inequality: Concepts, associated factors and prospects of a
+burgeoning research agenda. \emph{International Journal of Social
+Psychology}, \emph{37}(1), 180--207.
 \url{https://doi.org/10.1080/02134748.2021.2009275}
 
-\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
-Castillo, J.
-C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
-(2023).
-A {Multidimensional Approach} for {Measuring Meritocratic Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP Social Inequality Survey}.
-\emph{International Journal of Sociology}, 1--25.
-\url{https://doi.org/10.1080/00207659.2023.2274712}
-
-\bibitem[\citeproctext]{ref-confortiLegitimationInequalityAmerican1992}
-Conforti, J. M. (1992). The legitimation of inequality in {American} education. \emph{The Urban Review}, \emph{24}(4), 227--238. \url{https://doi.org/10.1007/BF01108357}
-
-\bibitem[\citeproctext]{ref-corvalanMercadoEscolarLibertad2016}
-Corvalán, J., Carrasco, A., \& García-Huidobro, J. E. (Eds.). (2016). \emph{Mercado escolar: {Libertad}, diversidad y desigualdad} (1st ed.). Ediciones UC. \url{https://doi.org/10.2307/j.ctv14rmrhn}
-
-\bibitem[\citeproctext]{ref-darmodyImpactsCOVID19Control2021}
-Darmody, M., Smyth, E., \& Russell, H.
-(2021).
-Impacts of the {COVID-19 Control Measures} on {Widening Educational Inequalities}.
-\emph{YOUNG}, \emph{29}(4), 366--380.
-\url{https://doi.org/10.1177/11033088211027412}
-
-\bibitem[\citeproctext]{ref-dayPerceivedIdealInequality2023}
-Day, M.
-V., \& Norton, M.
-I.
-(2023).
-Perceived and {Ideal Inequality} in {University Endowments} in the {United States}.
-\emph{Personality and Social Psychology Bulletin}, \emph{49}(8), 1151--1165.
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_multidimensional_2023}{}}%
+Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
+(2023). A {Multidimensional Approach} for {Measuring Meritocratic
+Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
+Social Inequality Survey}. \emph{International Journal of Sociology},
+1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
+
+\leavevmode\vadjust pre{\hypertarget{ref-confortiLegitimationInequalityAmerican1992}{}}%
+Conforti, J. M. (1992). The legitimation of inequality in {American}
+education. \emph{The Urban Review}, \emph{24}(4), 227--238.
+\url{https://doi.org/10.1007/BF01108357}
+
+\leavevmode\vadjust pre{\hypertarget{ref-corvalanMercadoEscolarLibertad2016}{}}%
+Corvalán, J., Carrasco, A., \& García-Huidobro, J. E. (Eds.). (2016).
+\emph{Mercado escolar: {Libertad}, diversidad y desigualdad} (1st ed.).
+Ediciones UC. \url{https://doi.org/10.2307/j.ctv14rmrhn}
+
+\leavevmode\vadjust pre{\hypertarget{ref-darmodyImpactsCOVID19Control2021}{}}%
+Darmody, M., Smyth, E., \& Russell, H. (2021). Impacts of the {COVID-19
+Control Measures} on {Widening Educational Inequalities}. \emph{YOUNG},
+\emph{29}(4), 366--380. \url{https://doi.org/10.1177/11033088211027412}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dayPerceivedIdealInequality2023}{}}%
+Day, M. V., \& Norton, M. I. (2023). Perceived and {Ideal Inequality} in
+{University Endowments} in the {United States}. \emph{Personality and
+Social Psychology Bulletin}, \emph{49}(8), 1151--1165.
 \url{https://doi.org/10.1177/01461672221083766}
 
-\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
-ELSOC, S. T. (2022). \emph{Estudio {Longitudinal Social} de {Chile}}. Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
-
-\bibitem[\citeproctext]{ref-Evans2017}
-Evans, M.
-D.
-R., \& Kelley, J.
-(2017).
-Communism, {Capitalism}, and {Images} of {Class}: {Effects} of {Reference Groups}, {Reality}, and {Regime} in 43 {Nations} and 110,000 {Individuals}, 1987-2009.
+\leavevmode\vadjust pre{\hypertarget{ref-elsoc_estudio_2022}{}}%
+ELSOC, S. T. (2022). \emph{Estudio {Longitudinal Social} de {Chile}}.
+Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Evans2017}{}}%
+Evans, M. D. R., \& Kelley, J. (2017). Communism, {Capitalism}, and
+{Images} of {Class}: {Effects} of {Reference Groups}, {Reality}, and
+{Regime} in 43 {Nations} and 110,000 {Individuals}, 1987-2009.
 \emph{Cross-Cultural Research}, \emph{51}(514), 315--359.
@@
 
-\bibitem[\citeproctext]{ref-Evans1992}
-Evans, M.
-D.
-R., Kelley, J., \& Kolosi, T.
-(1992).
-Images of {Class}: {Public Perceptions} in {Hungary} and {Australia}.
-\emph{American Sociological Review}, \emph{57}(4), 461.
+\leavevmode\vadjust pre{\hypertarget{ref-Evans1992}{}}%
+Evans, M. D. R., Kelley, J., \& Kolosi, T. (1992). Images of {Class}:
+{Public Perceptions} in {Hungary} and {Australia}. \emph{American
+Sociological Review}, \emph{57}(4), 461.
 \url{https://doi.org/10.2307/2096095}
 
-\bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
-García-Sánchez, E., \& De Carvalho Galvão, S.
-(2022).
-Las creencias que justifican la desigualdad moderan la relaci{ó}n entre el estatus socioecon{ó}mico y el apoyo a la redistribuci{ó}n.
-\emph{Revista Internacional de Sociolog{í}a}, \emph{80}(3), e210.
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_creencias_2022}{}}%
+García-Sánchez, E., \& De Carvalho, S. (2022). Las creencias que
+justifican la desigualdad moderan la relaci{ó}n entre el estatus
+socioecon{ó}mico y el apoyo a la redistribuci{ó}n. \emph{Revista
+Internacional de Sociolog{í}a}, \emph{80}(3), e210.
 \url{https://doi.org/10.3989/ris.2022.80.3.21.29}
 
-\bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
-Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality. \emph{Economics \& Politics}, \emph{30}(1), 27--54. \url{https://doi.org/10.1111/ecpo.12103}
-
-\bibitem[\citeproctext]{ref-goldin_meritocracy_2000}
-Goldin, C.
-(2000).
-Meritocracy and {Economic Inequality}.
-{Edited} by {Kenneth Arrow}, {Samuel Bowles}, and {Steven Durlauf} ( {Princeton}, {Princeton University Press}, 2000) 348 pp.
-\emph{Journal of Interdisciplinary History}, \emph{31}(3), 431a--433.
+\leavevmode\vadjust pre{\hypertarget{ref-gimpelson_misperceiving_2018}{}}%
+Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality.
+\emph{Economics \& Politics}, \emph{30}(1), 27--54.
+\url{https://doi.org/10.1111/ecpo.12103}
+
+\leavevmode\vadjust pre{\hypertarget{ref-goldin_meritocracy_2000}{}}%
+Goldin, C. (2000). Meritocracy and {Economic Inequality}. {Edited} by
+{Kenneth Arrow}, {Samuel Bowles}, and {Steven Durlauf} ( {Princeton},
+{Princeton University Press}, 2000) 348 pp. \emph{Journal of
+Interdisciplinary History}, \emph{31}(3), 431a--433.
 \url{https://doi.org/10.1162/jinh.2000.31.3.431a}
 
-\bibitem[\citeproctext]{ref-hadjar_meritokratie_2008}
-Hadjar, A. (2008). \emph{{Meritokratie als Legitimationsprinzip}}. VS Verlag f{ü}r Sozialwissenschaften.
-
-\bibitem[\citeproctext]{ref-hadler_why_2005}
-Hadler, M.
-(2005).
-Why {Do People Accept Different Income Ratios}?: {A Multi-level Comparison} of {Thirty Countries}.
-\emph{Acta Sociologica}, \emph{48}(2), 131--154.
-\url{https://doi.org/10.1177/0001699305053768}
-
-\bibitem[\citeproctext]{ref-hankivsky_introduction_2022}
-Hankivsky, O., Morrow, M., \& Varcoe, C.
-(2022).
-{INTRODUCTION Women}'s {Health} in {Canada}: {Critical Intersectional Perspectives} on {Theory} and {Policy}.
-In M.
-Morrow, O.
-Hankivsky, \& C.
-Varcoe (Eds.), \emph{Women's {Health} in {Canada}} (pp.
-1--10).
-University of Toronto Press.
-\url{https://doi.org/10.3138/9781442623958-003}
-
-\bibitem[\citeproctext]{ref-hing_failure_2019}
-Hing, L.
-S.
-S., Wilson, A.
-E., Gourevitch, P., English, J., \& Sin, P.
-(2019).
-Failure to {Respond} to {Rising Income Inequality}: {Processes That Legitimize Growing Disparities}.
-\emph{Daedalus}, \emph{148}(3), 105--135.
-\url{https://doi.org/10.1162/daed_a_01752}
-
-\bibitem[\citeproctext]{ref-igliozziFairShareEffects2024}
-Igliozzi, D., Granot, Y., \& Ottati, V.
-(2024).
-A fair share: {Effects} of disparity, allocation strategy and system justification on perceptions of policy support in the education domain.
-\emph{European Journal of Social Psychology}, ejsp.3040.
+\leavevmode\vadjust pre{\hypertarget{ref-hadjar_meritokratie_2008}{}}%
+Hadjar, A. (2008). \emph{{Meritokratie als Legitimationsprinzip}}. VS
+Verlag f{ü}r Sozialwissenschaften.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hadler_why_2005}{}}%
+Hadler, M. (2005). Why {Do People Accept Different Income Ratios}?: {A
+Multi-level Comparison} of {Thirty Countries}. \emph{Acta Sociologica},
+\emph{48}(2), 131--154. \url{https://doi.org/10.1177/0001699305053768}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hankivsky_introduction_2022}{}}%
+Hankivsky, O., Morrow, M., \& Varcoe, C. (2022). {INTRODUCTION Women}'s
+{Health} in {Canada}: {Critical Intersectional Perspectives} on {Theory}
+and {Policy}. In M. Morrow, O. Hankivsky, \& C. Varcoe (Eds.),
+\emph{Women's {Health} in {Canada}} (pp. 1--10). University of Toronto
+Press. \url{https://doi.org/10.3138/9781442623958-003}
+
+\leavevmode\vadjust pre{\hypertarget{ref-igliozziFairShareEffects2024}{}}%
+Igliozzi, D., Granot, Y., \& Ottati, V. (2024). A fair share: {Effects}
+of disparity, allocation strategy and system justification on
+perceptions of policy support in the education domain. \emph{European
+Journal of Social Psychology}, ejsp.3040.
 \url{https://doi.org/10.1002/ejsp.3040}
 
-\bibitem[\citeproctext]{ref-iturra_percepcion_2023}
-Iturra, J., Castillo, J.
-C., Rufs Orellana, C., \& Maldonado, L.
-(2023).
-La percepci{ó}n de desigualdad econ{ó}mica y su influencia sobre la justificaci{ó}n de las diferencias de ingreso leg{í}timas.
-\emph{Estudios Sociol{ó}gicos de El Colegio de M{é}xico}, \emph{41}(122), 309--352.
+\leavevmode\vadjust pre{\hypertarget{ref-iturra_percepcion_2023}{}}%
+Iturra, J., Castillo, J. C., Rufs Orellana, C., \& Maldonado, L. (2023).
+La percepci{ó}n de desigualdad econ{ó}mica y su influencia sobre la
+justificaci{ó}n de las diferencias de ingreso leg{í}timas.
+\emph{Estudios Sociol{ó}gicos de El Colegio de M{é}xico},
+\emph{41}(122), 309--352.
 \url{https://doi.org/10.24201/es.2023v41n122.2260}
 
-\bibitem[\citeproctext]{ref-janmaatSubjectiveInequalityReview2013}
-Janmaat, J.
-G.
-(2013).
-Subjective inequality: {A} review of international comparative studies on people's views about inequality.
+\leavevmode\vadjust pre{\hypertarget{ref-janmaatSubjectiveInequalityReview2013}{}}%
+Janmaat, J. G. (2013). Subjective inequality: {A} review of
+international comparative studies on people's views about inequality.
 \emph{Archives Europeennes de Sociologie}, \emph{54}(3), 357--389.
@@
 
-\bibitem[\citeproctext]{ref-jassoJusticeEarningsNew1978}
-Jasso, G.
-(1978).
-On the {Justice} of {Earnings}: {A New Specification} of the {Justice Evaluation Function}.
-\emph{American Journal of Sociology}, \emph{83}(6), 1398--1419.
+\leavevmode\vadjust pre{\hypertarget{ref-jassoJusticeEarningsNew1978}{}}%
+Jasso, G. (1978). On the {Justice} of {Earnings}: {A New Specification}
+of the {Justice Evaluation Function}. \emph{American Journal of
+Sociology}, \emph{83}(6), 1398--1419.
 \url{https://www.jstor.org/stable/2778110}
 
-\bibitem[\citeproctext]{ref-joikomujicaCuasimercadoEducativoChile2019}
-Joiko Mujica, S. (2019). {El cuasi-mercado educativo en Chile: desarrollo y consecuencias}. \emph{Revista Electr{ó}nica Di{á}logos Educativos; Vol. 12 N{ú}m. 23 (2012); 148-174}.
-
-\bibitem[\citeproctext]{ref-jost_theory_2020}
-Jost, J. (2020). \emph{A theory of system justification}. Harvard University Press.
-
-\bibitem[\citeproctext]{ref-jost_psychology_2003}
-Jost, J., \& Hunyady, O.
-(2003).
-The psychology of system justification and the palliative function of ideology.
-\emph{European Review of Social Psychology}, \emph{13}(1), 111--153.
+\leavevmode\vadjust pre{\hypertarget{ref-joiko_cuasimercado_2019}{}}%
+Joiko, S. (2019). {El cuasi-mercado educativo en Chile: desarrollo y
+consecuencias}. \emph{Revista Electr{ó}nica Di{á}logos Educativos; Vol.
+12 N{ú}m. 23 (2012); 148-174}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jost_theory_2020}{}}%
+Jost, J. (2020). \emph{A theory of system justification}. Harvard
+University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jost_psychology_2003}{}}%
+Jost, J., \& Hunyady, O. (2003). The psychology of system justification
+and the palliative function of ideology. \emph{European Review of Social
+Psychology}, \emph{13}(1), 111--153.
 \url{https://doi.org/10.1080/10463280240000046}
 
-\bibitem[\citeproctext]{ref-kelleyLegitimationInequalityOccupational1993}
-Kelley, J., \& Evans, M.
-D.
-R.
-(1993).
-The legitimation of inequality: {Occupational} earnings in nine nations.
-\emph{American Journal of Sociology}, \emph{99}(1), 75--125.
+\leavevmode\vadjust pre{\hypertarget{ref-kelleyLegitimationInequalityOccupational1993}{}}%
+Kelley, J., \& Evans, M. D. R. (1993). The legitimation of inequality:
+{Occupational} earnings in nine nations. \emph{American Journal of
+Sociology}, \emph{99}(1), 75--125.
 \url{https://www.jstor.org/stable/2781956}
 
-\bibitem[\citeproctext]{ref-kelley_legitimate_2021}
-Kelley, J., \& Evans, M.
-D.
-R.
-(2021).
-Legitimate earnings inequality and national welfare commitment: {Correspondence} between economic institutions and the pay 80,000+ people in 30 nations think legitimate for ordinary jobs and for elite jobs.
-\emph{Social Science Research}, \emph{94}, 102446.
+\leavevmode\vadjust pre{\hypertarget{ref-kelley_legitimate_2021}{}}%
+Kelley, J., \& Evans, M. D. R. (2021). Legitimate earnings inequality
+and national welfare commitment: {Correspondence} between economic
+institutions and the pay 80,000+ people in 30 nations think legitimate
+for ordinary jobs and for elite jobs. \emph{Social Science Research},
+\emph{94}, 102446.
 \url{https://doi.org/10.1016/j.ssresearch.2020.102446}
 
-\bibitem[\citeproctext]{ref-kluegel_social_1995}
-Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995). \emph{Social justice and political change: Public opinion in capitalist and post-communist states}. A. de Gruyter.
-
-\bibitem[\citeproctext]{ref-kluegelBeliefsStratification1981}
-Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}. \emph{Annual Review of Sociology}, 29--56.
-
-\bibitem[\citeproctext]{ref-kluegelBeliefsInequalityAmericans1986}
-Kluegel, J. R., \& Smith, E. R. (1986). \emph{Beliefs about {Inequality}: {Americans}' {Views} of {What Is} and {What Ought} to {Be}} (1st ed.). Routledge.
-
-\bibitem[\citeproctext]{ref-leeFairnessPerceptionsEducational2023}
-Lee, J.-S., \& Stacey, M.
-(2023).
-Fairness perceptions of educational inequality: The effects of self-interest and neoliberal orientations.
+\leavevmode\vadjust pre{\hypertarget{ref-kluegel_social_1995}{}}%
+Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995).
+\emph{Social justice and political change: Public opinion in capitalist
+and post-communist states}. A. de Gruyter.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kluegelBeliefsStratification1981}{}}%
+Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
+\emph{Annual Review of Sociology}, 29--56.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kluegelBeliefsInequalityAmericans1986}{}}%
+Kluegel, J. R., \& Smith, E. R. (1986). \emph{Beliefs about
+{Inequality}: {Americans}' {Views} of {What Is} and {What Ought} to
+{Be}} (1st ed.). Routledge.
+
+\leavevmode\vadjust pre{\hypertarget{ref-leeFairnessPerceptionsEducational2023}{}}%
+Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
+inequality: The effects of self-interest and neoliberal orientations.
 \emph{The Australian Educational Researcher}.
@@
 
-\bibitem[\citeproctext]{ref-liebigSociologyJustice2016}
-Liebig, S., \& Sauer, C. (2016). Sociology of {Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social Justice Theory} and {Research}} (pp. 37--59). Springer New York.
-
-\bibitem[\citeproctext]{ref-lindhPublicOpinionMarkets2015}
-Lindh, A.
-(2015).
-Public {Opinion} against {Markets}?
-{Attitudes} towards {Market Distribution} of {Social Services} -- {A Comparison} of 17 {Countries}.
-\emph{Social Policy \& Administration}, \emph{49}(7), 887--910.
-\url{https://doi.org/10.1111/spol.12105}
-
-\bibitem[\citeproctext]{ref-McCoy2007}
-McCoy, S.
-K., \& Major, B.
-(2007).
-Priming meritocracy and the psychological justification of inequality.
-\emph{Journal of Experimental Social Psychology}, \emph{43}(3), 341--351.
+\leavevmode\vadjust pre{\hypertarget{ref-liebigSociologyJustice2016}{}}%
+Liebig, S., \& Sauer, C. (2016). Sociology of {Justice}. In C. Sabbagh
+\& M. Schmitt (Eds.), \emph{Handbook of {Social Justice Theory} and
+{Research}} (pp. 37--59). Springer New York.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lindhPublicOpinionMarkets2015}{}}%
+Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
+towards {Market Distribution} of {Social Services} -- {A Comparison} of
+17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
+887--910. \url{https://doi.org/10.1111/spol.12105}
+
+\leavevmode\vadjust pre{\hypertarget{ref-McCoy2007}{}}%
+McCoy, S. K., \& Major, B. (2007). Priming meritocracy and the
+psychological justification of inequality. \emph{Journal of Experimental
+Social Psychology}, \emph{43}(3), 341--351.
 \url{https://doi.org/10.1016/j.jesp.2006.04.009}
 
-\bibitem[\citeproctext]{ref-mcnamee_meritocracy_2004}
-McNamee, S., \& Miller, R. (2004). \emph{The meritocracy myth} (2004th ed.). Rowman \& Littlefield.
-
-\bibitem[\citeproctext]{ref-mijs_stratified_2016}
-Mijs, J.
-(2016).
-Stratified {Failure}: {Educational Stratification} and {Students}' {Attributions} of {Their Mathematics Performance} in 24 {Countries}.
-\emph{Sociology of Education}, \emph{89}(2), 137--153.
+\leavevmode\vadjust pre{\hypertarget{ref-mcnamee_meritocracy_2004}{}}%
+McNamee, S., \& Miller, R. (2004). \emph{The meritocracy myth} (2004th
+ed.). Rowman \& Littlefield.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_stratified_2016}{}}%
+Mijs, J. (2016). Stratified {Failure}: {Educational Stratification} and
+{Students}' {Attributions} of {Their Mathematics Performance} in 24
+{Countries}. \emph{Sociology of Education}, \emph{89}(2), 137--153.
 \url{https://doi.org/10.1177/0038040716636434}
 
-\bibitem[\citeproctext]{ref-mijs_visualizing_2018}
-Mijs, J. (2018). Visualizing {Belief} in {Meritocracy}, 1930--2010. \emph{Socius}, \emph{4}, 2378023118811805. \url{https://doi.org/10.1177/2378023118811805}
-
-\bibitem[\citeproctext]{ref-mijs_paradox_2021}
-Mijs, J. (2021). The paradox of inequality: Income inequality and belief in meritocracy go hand in hand. \emph{Socio-Economic Review}, \emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
-
-\bibitem[\citeproctext]{ref-mijs_belief_2022}
-Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J.
-(2022).
-Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of {Subjective Social Mobility}.
-\emph{Social Psychology Quarterly}, \emph{85}(2), 131--141.
-\url{https://doi.org/10.1177/01902725211063818}
-
-\bibitem[\citeproctext]{ref-oecd_pisa_2023}
-OECD. (2023). \emph{{PISA} 2022 {Results} ({Volume I}): {The State} of {Learning} and {Equity} in {Education}}. OECD. \url{https://doi.org/10.1787/53f23881-en}
-
-\bibitem[\citeproctext]{ref-osberg_fair_2006}
-Osberg, L., \& Smeeding, T.
-(2006).
-{``{Fair}''} {Inequality}?
-{Attitudes} toward {Pay Differentials}: {The United States} in {Comparative Perspective}.
-\emph{American Sociological Review}, \emph{71}(3), 450--473.
-\url{https://doi.org/10.1177/000312240607100305}
-
-\bibitem[\citeproctext]{ref-palanza_chile_2023}
-Palanza, V., \& Sotomayor, P.
-(2023).
-Chile's failed constitutional intent: {Polarization}, fragmentation, haste and delegitimization.
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_visualizing_2018}{}}%
+Mijs, J. (2018). Visualizing {Belief} in {Meritocracy}, 1930--2010.
+\emph{Socius}, \emph{4}, 2378023118811805.
+\url{https://doi.org/10.1177/2378023118811805}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_paradox_2021}{}}%
+Mijs, J. (2021). The paradox of inequality: Income inequality and belief
+in meritocracy go hand in hand. \emph{Socio-Economic Review},
+\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_belief_2022}{}}%
+Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022).
+Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of
+{Subjective Social Mobility}. \emph{Social Psychology Quarterly},
+\emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}
+
+\leavevmode\vadjust pre{\hypertarget{ref-oecd_pisa_2023}{}}%
+OECD. (2023). \emph{{PISA} 2022 {Results} ({Volume I}): {The State} of
+{Learning} and {Equity} in {Education}}. OECD.
+\url{https://doi.org/10.1787/53f23881-en}
+
+\leavevmode\vadjust pre{\hypertarget{ref-osberg_fair_2006}{}}%
+Osberg, L., \& Smeeding, T. (2006). {``{Fair}''} {Inequality}?
+{Attitudes} toward {Pay Differentials}: {The United States} in
+{Comparative Perspective}. \emph{American Sociological Review},
+\emph{71}(3), 450--473. \url{https://doi.org/10.1177/000312240607100305}
+
+\leavevmode\vadjust pre{\hypertarget{ref-palanza_chile_2023}{}}%
+Palanza, V., \& Sotomayor, P. (2023). Chile's failed constitutional
+intent: {Polarization}, fragmentation, haste and delegitimization.
 \emph{Global Constitutionalism}, 1--10.
@@
 
-\bibitem[\citeproctext]{ref-ramosarellanoEducacionPromesaIncumplida2022}
-Ramos Arellano, M.
-(Ed.).
-(2022).
-\emph{Educaci{ó}n: {La} promesa incumplida: Esfuerzo, miedos y esperanzas de familias chilenas en el mercado escolar} (Primera edici{ó}n).
-Catalonia : CIAE, Centro de Investigaci{ó}n Avanzada en Educaci{ó}n, Universidad de Chile.
-
-\bibitem[\citeproctext]{ref-Resh2016}
-Resh, N., \& Sabbagh, C. (2016). Justice and education. In \emph{Handbook of Social Justice Theory and Research}. \url{https://doi.org/10.1007/978-1-4939-3216-0_19}
-
-\bibitem[\citeproctext]{ref-rosvall_mapping_2010}
-Rosvall, M., \& Bergstrom, C. T. (2010). Mapping {Change} in {Large Networks}. \emph{PLoS ONE}, \emph{5}(1), e8694. \url{https://doi.org/10.1371/journal.pone.0008694}
-
-\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
-Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the common good?} (First edition). {Farrar, Straus and Giroux}.
-
-\bibitem[\citeproctext]{ref-sazo_chile_2023}
-Sazo, D. (2023). Chile 2022: {From} great expectations to rising pessimism. \emph{Revista de Ciencia Pol{í}tica (Santiago)}, \emph{ahead}. \url{https://doi.org/10.4067/s0718-090x2023005000118}
-
-\bibitem[\citeproctext]{ref-schroder_income_2017}
-Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for {Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47. \url{https://doi.org/10.1007/s11211-016-0276-8}
-
-\bibitem[\citeproctext]{ref-singer_applied_2009}
-Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data analysis: Modeling change and event occurence}. Oxford University Press, Incorporated.
-
-\bibitem[\citeproctext]{ref-sonhing_merit_2011}
-Son Hing, L.
-S., Ramona, D., Zanna, M.
-P., Garcia, D.
-M., Gee, S.
-S., \& Orazietti, K.
-(2011).
-The merit of meritocracy.
-\emph{Journal of Personality and Social Psychology}, \emph{101}(3), 433--450.
+\leavevmode\vadjust pre{\hypertarget{ref-ramos_educacion_2022}{}}%
+Ramos, M. (Ed.). (2022). \emph{Educaci{ó}n: {La} promesa incumplida:
+Esfuerzo, miedos y esperanzas de familias chilenas en el mercado
+escolar} (Primera edici{ó}n). Catalonia : CIAE, Centro de
+Investigaci{ó}n Avanzada en Educaci{ó}n, Universidad de Chile.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Resh2016}{}}%
+Resh, N., \& Sabbagh, C. (2016). Justice and education. In
+\emph{Handbook of Social Justice Theory and Research}.
+\url{https://doi.org/10.1007/978-1-4939-3216-0_19}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rosvall_mapping_2010}{}}%
+Rosvall, M., \& Bergstrom, C. T. (2010). Mapping {Change} in {Large
+Networks}. \emph{PLoS ONE}, \emph{5}(1), e8694.
+\url{https://doi.org/10.1371/journal.pone.0008694}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sandel_tyranny_2020}{}}%
+Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
+common good?} (First edition). {Farrar, Straus and Giroux}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sazo_chile_2023}{}}%
+Sazo, D. (2023). Chile 2022: {From} great expectations to rising
+pessimism. \emph{Revista de Ciencia Pol{í}tica (Santiago)},
+\emph{ahead}. \url{https://doi.org/10.4067/s0718-090x2023005000118}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schroder_income_2017}{}}%
+Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for
+{Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47.
+\url{https://doi.org/10.1007/s11211-016-0276-8}
+
+\leavevmode\vadjust pre{\hypertarget{ref-singer_applied_2009}{}}%
+Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data
+analysis: Modeling change and event occurence}. Oxford University Press,
+Incorporated.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sonhing_merit_2011}{}}%
+Son Hing, L. S., Ramona, D., Zanna, M. P., Garcia, D. M., Gee, S. S., \&
+Orazietti, K. (2011). The merit of meritocracy. \emph{Journal of
+Personality and Social Psychology}, \emph{101}(3), 433--450.
 \url{https://doi.org/10.1037/a0024618}
 
-\bibitem[\citeproctext]{ref-stoilovaFairnessEducationalOpportunities2023}
-Stoilova, R., \& Ilieva-Trichkova, P.
-(2023).
-Fairness of educational opportunities and income distribution: Gender-sensitive analysis in a {European} comparative perspective.
-\emph{International Journal of Sociology and Social Policy}, \emph{43}(1/2), 272--291.
+\leavevmode\vadjust pre{\hypertarget{ref-sonhing_failure_2019}{}}%
+Son Hing, L. S., Wilson, A. E., Gourevitch, P., English, J., \& Sin, P.
+(2019). Failure to {Respond} to {Rising Income Inequality}: {Processes
+That Legitimize Growing Disparities}. \emph{Daedalus}, \emph{148}(3),
+105--135. \url{https://doi.org/10.1162/daed_a_01752}
+
+\leavevmode\vadjust pre{\hypertarget{ref-stoilovaFairnessEducationalOpportunities2023}{}}%
+Stoilova, R., \& Ilieva-Trichkova, P. (2023). Fairness of educational
+opportunities and income distribution: Gender-sensitive analysis in a
+{European} comparative perspective. \emph{International Journal of
+Sociology and Social Policy}, \emph{43}(1/2), 272--291.
 \url{https://doi.org/10.1108/IJSSP-02-2022-0065}
 
-\bibitem[\citeproctext]{ref-trump_income_2018}
-Trump, K.-S.
-(2018).
-Income {Inequality Influences Perceptions} of {Legitimate Income Differences}.
-\emph{British Journal of Political Science}, \emph{48}(4), 929--952.
+\leavevmode\vadjust pre{\hypertarget{ref-trump_income_2018}{}}%
+Trump, K.-S. (2018). Income {Inequality Influences Perceptions} of
+{Legitimate Income Differences}. \emph{British Journal of Political
+Science}, \emph{48}(4), 929--952.
 \url{https://doi.org/10.1017/S0007123416000326}
 
-\bibitem[\citeproctext]{ref-trump_when_2020}
-Trump, K.-S. (2020). When and why is economic inequality seen as fair. \emph{Current Opinion in Behavioral Sciences}, \emph{34}, 46--51. \url{https://doi.org/10.1016/j.cobeha.2019.12.001}
-
-\bibitem[\citeproctext]{ref-valantPoliticsAchievementGaps2016}
-Valant, J., \& Newark, D.
-A.
-(2016).
-The {Politics} of {Achievement Gaps}: {U}.{S}.
-{Public Opinion} on {Race-Based} and {Wealth-Based Differences} in {Test Scores}.
-\emph{Educational Researcher}, \emph{45}(6), 331--346.
-\url{https://doi.org/10.3102/0013189X16658447}
-
-\bibitem[\citeproctext]{ref-vandertoorn_twenty_2014}
-van der Toorn, J., \& Jost, J.
-(2014).
-Twenty years of system justification theory: {Introduction} to the special issue on {``{Ideology} and system justification processes.''} \emph{Group Processes \& Intergroup Relations}, \emph{17}(4), 413--419.
+\leavevmode\vadjust pre{\hypertarget{ref-trump_when_2020}{}}%
+Trump, K.-S. (2020). When and why is economic inequality seen as fair.
+\emph{Current Opinion in Behavioral Sciences}, \emph{34}, 46--51.
+\url{https://doi.org/10.1016/j.cobeha.2019.12.001}
+
+\leavevmode\vadjust pre{\hypertarget{ref-valantPoliticsAchievementGaps2016}{}}%
+Valant, J., \& Newark, D. A. (2016). The {Politics} of {Achievement
+Gaps}: {U}.{S}. {Public Opinion} on {Race-Based} and {Wealth-Based
+Differences} in {Test Scores}. \emph{Educational Researcher},
+\emph{45}(6), 331--346. \url{https://doi.org/10.3102/0013189X16658447}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vandertoorn_twenty_2014}{}}%
+van der Toorn, J., \& Jost, J. (2014). Twenty years of system
+justification theory: {Introduction} to the special issue on
+{``{Ideology} and system justification processes.''} \emph{Group
+Processes \& Intergroup Relations}, \emph{17}(4), 413--419.
 \url{https://doi.org/10.1177/1368430214531509}
 
-\bibitem[\citeproctext]{ref-wegenerIllusionDistributiveJustice1987}
-Wegener, B. (1987). The {Illusion} of {Distributive Justice}. \emph{European Sociological Review}, \emph{3}(1), 1--13. \url{https://www.jstor.org/stable/522527}
-
-\bibitem[\citeproctext]{ref-Wegener1990}
-Wegener, B. (1990). Equity, relative deprivation, and the value consensus paradox. \emph{Social Justice Research}, \emph{4}(1), 65--86. \url{https://doi.org/10.1007/BF01048536}
-
-\bibitem[\citeproctext]{ref-young_rise_1962}
+\leavevmode\vadjust pre{\hypertarget{ref-wegenerIllusionDistributiveJustice1987}{}}%
+Wegener, B. (1987). The {Illusion} of {Distributive Justice}.
+\emph{European Sociological Review}, \emph{3}(1), 1--13.
+\url{https://www.jstor.org/stable/522527}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Wegener1990}{}}%
+Wegener, B. (1990). Equity, relative deprivation, and the value
+consensus paradox. \emph{Social Justice Research}, \emph{4}(1), 65--86.
+\url{https://doi.org/10.1007/BF01048536}
+
+\leavevmode\vadjust pre{\hypertarget{ref-young_rise_1962}{}}%
 Young, M. (1962). \emph{The rise of the meritocracy}. Penguin Books.
@@
 
-\appendix
+\hypertarget{appendix-appendix}{%
+\section*{(APPENDIX) Appendix}\label{appendix-appendix}}
+\addcontentsline{toc}{section}{(APPENDIX) Appendix}
 
@@
 \begin{table}[!ht]
-\caption{\label{tab:multilevel-full} Multilevel model for the justification of economic inequality}
+\caption{(\#tab:multilevel-full) Multilevel model for the justification of economic inequality}
 \begin{center}

### 2024-06-25 presentations/ISPP-santiago2024/ispp2024.qmd
up ispp presentation

--- 2024-03-25 production/prod_analysis-fix-t5.Rmd
@@
 ---
-title: "Analysis"
-date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
-output:
-  html_document:
-    toc: yes
-    toc_float:
-      collapsed: yes
-      smooth_scroll: no
-      number_sections: yes
-    code_folding: hide
-    number_sections: yes
-editor_options:
-  chunk_output_type: console
+pagetitle: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
+author: Juan Carlos Castillo, Julio Iturra & Kevin Carrasco
+bibliography: ../../input/bib/merit-edjust.bib 
+csl: ../../input/bib/apa7-singlespace.csl
+
+knitr:
+  opts_chunk: 
+    collapse: true
+    comment: "#>" 
+    R.options:
+      knitr.graphics.auto_pdf: true
+
+format: 
+  revealjs:
+    logo: images/edumer.png
+    slide-number: true
+    theme: [pp.scss]
+    #theme: [pp.scss]
+    title-slide-attributes:
+      visibility: false
+#      data-background-image: images/portada.jpg
+#      data-background-size: cover
+    auto-play-media: true
+
+editor: visual
 ---
-# Setup
 
-```{r setup}
-knitr::opts_chunk$set(
-	echo = FALSE,
-	results = "asis",
-	message = FALSE,
-	warning = FALSE, 
-	fig.height = 18, 
-	fig.width = 11, 
-	cache = TRUE
-)
-options(scipen=9999) # desactivar notacion cientifica
-remove(list = ls()) #limpieza del entorno de trabajo
-```
-
-```{css, echo=FALSE}
-# /* expandir el espacio del documento*/
-body .main-container {
-      max-width: 1600px !important;
-      width: 1600px !important;
-    }
-#/* expandir width de la TOC*/
-div.tocify {
-    width: 22% !important;
-    max-width: 331px !important;
-    max-height: 102% !important;
-}
-```
-
-# Descriptive analysis
-
-## Univariate
-
-```{r tab-desc, results="asis", warning=FALSE}
+```{r data, results=FALSE, warning=FALSE, echo=FALSE}
 if (!require("pacman")) install.packages("pacman") # instalar pacman
@@
 
-# Css
-st_css()
-st_options(lang = "en",
-           footnote = NA,
-           bootstrap.css = F,
-           custom.css = here::here("input/css/dfsummary.css"),
-           dfSummary.custom.1 = NA
-           )
+```
 
-# Df variables nivel 1
-# df<- dfSummary(df_study1_long %>% select(in_simpat:edad),
-#                plain.ascii = FALSE,
-#                style = "grid",
-#                tmp.img.dir = "/tmp",
-#                graph.magnif = TRUE,
-#                headings = F,  # encabezado
-#                varnumbers = F, # num variable
-#                labels.col = T, # etiquetas
-#                na.col = F,    # missing
-#                graph.col = T, # plot
-#                valid.col = T, # n valido
-#                col.widths = c(700,10,10,10,10)
-#                )
-# df$Variable <- NULL # delete variable column
-# view(df,
-#      file = here::here("output/tables/desc01.html")
-#      ) # Ver tabla en un archivo HTML
-# webshot::webshot(url =here::here("output/tables/desc01.html"),
-#         file =here::here("output/tables/desc01.png")) # Remover archivo HTML
-# # print(df, method = "render")
+::: {.columns .v-center-container}
+::: {.column width="30%"}
+![](qr.png){width="80%" fig-align="right"}
+<br>
+![](images/logocoes.png){width="90%" fig-align="right"}
+<br>
+![](images/edumer.png){width="80%" fig-align="left"}
+:::
 
-# Df variables nivel 1 _________________________________________________________
+::: {.column width="70%"}
+
+## **Changes in the justification of educational inequalities:**
+
+### The role of perceptions of inequality and meritocracy during the COVID pandemic
+
+----------------------------------------------------------------
+
+Juan Carlos Castillo, Julio Iturra & **Kevin Carrasco* **
+
+#### Fondecyt N° 1210847: Meritocracia en la escuela
+#### Center for social conflict and cohesion studies
+
+**ISPP Annual meeting** Chile, July 2024
+:::
+:::
+
+# Introduction {data-background-color="black"}
+
+## Introduction
+
+::: {.incremental}
+
+::: {.columns .v-center-container}
+::: {.column width="50%" .incremental}
+
+### Justification of inequality
+
+- Perceived economic inequality and justification of inequality [@wegenerIllusionDistributiveJustice1987; @jassoJusticeEarningsNew1978; @janmaatSubjectiveInequalityReview2013]
+
+- Cultural factors and policy feedback effects [@kelley_legitimate_2021; @mijs_paradox_2021; @trump_income_2018]
+
+- Chile, marketization & neoliberal reforms [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016]
+
+:::
+
+::: {.column width="50%" .incremental}
+### Meritocracy
+
+- Effort and talent [@young_rise_1962]
+
+- Normative and perceptual elements [@duru-bellat_who_2012; @castillo_meritocracia_2019]
+
+:::
+:::
+
+::: {.r-stack style="font-size: 130%;"}
+- Covid pandemic [@breznau_welfare_2021; @hankivsky_introduction_2022]
+:::
+
+:::
+
+## This study
+
+<br>
+
+![](general-scheme.png){width="100%" fig-align="center"}
+
+# Methods {data-background-color="black"}
+
+## Data
+
+::: {.incremental}
+
+- Longitudinal social study of Chile (ELSOC-COES) 2016-2022
+
+- 2021 (telephone version) the dependent variable was not asked
+
+- N total=8042; 1689 individuals for 5 waves
+
+:::
+
+## Variables
+
+```{r bivariate-plots, echo=FALSE}
+df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               educ, quintil1:sexo) %>% na.omit()
+merit_plot <- df_study1 %>% 
+  select(merit_effort, merit_talent, des_perc, just_educ) %>% 
+  sjPlot::plot_stackfrq(weight.by = df_study1$ponderador_long_total,
+                        expand.grid =TRUE,
+                      geom.colors = "PuBu",
+                      show.total = FALSE, 
+                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),4),
+                      legend.labels = c("Strongly disagree", "Disagree", "Neither agree
+nor disagree", "Agree", "Strongly agree"),
+                      axis.labels = c("Merit effort", "Merit talent", "Inequality 
+perception", "Justification 
+of 
+educational inequality")
+                      ) +
+  theme_light(base_size = 16) +
+  theme(legend.position="bottom", text = element_text(size = 14),
+        plot.caption = element_text(hjust = 1)) +
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
+merit_plot
+
+```
+
+## Variables {.smaller}
+
+```{r error=FALSE}
 df_study1 <- df_study_long_t6_comunas %>% dplyr::select(ola,
@@
 df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
+```
 
-df<- dfSummary(x = dplyr::select(df_study1, merit_talent, merit_effort, des_perc, educ, quintil1, ess, edad, sexo, pos_id),
+::: {.columns .v-center-container}
+
+::: {.column width="50%"}
+
+```{r error=FALSE}
+df<- dfSummary(x = select(df_study1, ola, quintil1, sexo),
                plain.ascii = FALSE,
@@
                na.col = F,    # missing
-               graph.col = T, # plot
+               graph.col = F, # plot
                valid.col = F, # n valido
-               col.widths = c(700,10,10,10)
+               col.widths = c(10,10,10)
                )
 df$Variable <- NULL # delete variable column
-summarytools::view(df,
-     file = here::here("output/tables/desc02.html")
-     ) # Ver tabla en un archivo HTML
-webshot::webshot(url =here::here("output/tables/desc02.html"),
-        file =here::here("output/tables/desc02.png")) # Remover archivo HTML
-# print(df, method = "render")
+
+print(df, method='render',
+      footnote = NA, 
+      headings = FALSE, 
+      style = list(
+        table = "font-size: 8px;",
+        th = "font-size: 8px;",
+        td = "font-size: 8px;"
+      ))
 ```
+
+:::
+
+::: {.column width="50%"}
+```{r error=FALSE}
+df<- dfSummary(x = select(df_study1, edad, pos_id),
+               plain.ascii = FALSE,
+               style = "grid",
+               tmp.img.dir = "/tmp",
+               graph.magnif = TRUE,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = F, # plot
+               valid.col = F, # n valido
+               col.widths = c(10,10,10)
+               )
+df$Variable <- NULL # delete variable column
+
+print(df, method='render',
+      footnote = NA, 
+      headings = FALSE, 
+      style = list(
+        table = "font-size: 8px;",
+        th = "font-size: 8px;",
+        td = "font-size: 8px;"
+      ))
+```
+:::
+
+:::
+
+# Results {data-background-color="black"}
+
+##
+
+### Changes in the justification of educational inequality (2016-2022)
 
@@
 alluvial_justeduc
-
-ggsave(alluvial_justeduc, file="output/graphs/alluvial_dep.png", height = 8, width = 7)
 ```
 
-```{r bivariate-plots}
-df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent,
-                                               educ, quintil1:sexo) %>% na.omit()
-merit_plot <- df_study1 %>% 
-  select(merit_effort, merit_talent, des_perc, just_educ) %>% 
-  sjPlot::plot_stackfrq(weight.by = df_study1$ponderador_long_total,
-                        expand.grid =TRUE,
-                      geom.colors = "PuBu",
-                      show.total = FALSE, 
-                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),4),
-                      legend.labels = c("Strongly disagree", "Disagree", "Neither agree
-nor disagree", "Agree", "Strongly agree"),
-                      axis.labels = c("Merit effort", "Merit talent", "Inequality 
-perception", "Justification 
-of 
-educational inequality")
-                      ) +
-  theme_light(base_size = 16) +
-  theme(legend.position="bottom", text = element_text(size = 14),
-        plot.caption = element_text(hjust = 1)) +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
-merit_plot
+##
 
-ggsave(merit_plot, file="output/graphs/merit_plot.png", height = 6, width = 10)
-```
-
-```{r}
-
-range <- c(1, 1)
-
-p_just_educ<-
-  dplyr::select(df_study1,
-                just_educ) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_des_perc<-
-  dplyr::select(df_study1,
-                des_perc) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_effort<-
-  dplyr::select(df_study1,
-                merit_effort) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = T,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_talent<-
-  dplyr::select(df_study1,
-                merit_talent) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE, 
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5), 
-    labels = c(
-      "Strongly disagree",
-      "Disagree",
-      "Neither disagree nor agree",
-      "Agree",
-      "Strongly agree"
-    )
-  )
-```
-
-##### Create plot
-```{r}
-pacman::p_load(cowplot)
-t <- 0.1
-s <- 0.45
-m <- 0.4
-xs <- 0.25
-
-caption <- 
-  stringr::str_wrap(
-    paste0("Note: Own elaboration based on ELSOC Survey ",
-           "(n=",dim(na.omit(df_study1))[1],"). "),
-    width = 100)
-
-title <- cowplot::ggdraw() + 
-  cowplot::draw_label(
-    "Frequencies",
-    fontface = 'bold',
-    size = 15,
-    x = 0,
-    hjust = 0
-  )
-
-likert1<-   
-  cowplot::add_sub(
-    cowplot::plot_grid(p_just_educ,
-                       p_des_perc,
-                       p_effort,
-                       p_talent,
-                       align = "hv",
-                       axis = "a",
-                       nrow = 7, #6 plots
-                       ncol = 1, #1 column   
-                       rel_heights = c(
-                                       s,s,s,s,
-                                       s,
-                                       s) #fig size within plot
-    ),
-    label = "", # Caption: source and sample size
-    size = 11,  # Caption font size
-    hjust = 0# Caption align 
-  ) %>% 
-  cowplot::ggdraw()
-jpeg(here::here("output/graphs/figure1.jpg"),
-    # units = "cm",
-    # res = 100,
-    width = 800,
-    height = 800)
-likert1
-dev.off()
-```
-
-justification education vs year
-
-```{r}
-just_educ_year = df_study1 %>%
-  group_by(ola) %>%
-  summarise(education = mean(just_educ, na.rm = T))
-
-categorias_x <- c(2016, 2017, 2018, 2019, 2022)
-
-educ_year_plot <- just_educ_year_survey %>% ggplot(aes(ola, education)) +
-  theme_bw(base_size = 14) +
-  geom_point(size = 4)+
-  ylab("Justification of inequality") +
-  xlab("Wave") +
-  scale_x_discrete(labels = categorias_x) +
-  ylim(1,3)
-educ_year_plot
-
-ggsave(educ_year_plot, file="output/graphs/educ_year_plot.png", width = 7, height = 7)
-```
-
-Bivariate descriptive by year
-
-probar ponderadores longitudinales
+### Changes in the main independent variables over years (2016-2022)
 
@@
   geom_point(data = label_data, aes(label_x - 0.05, label_y, color = variable, shape = variable), size = 4) +  # Capa de puntos para los símbolos de forma
-  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 0.05, nudge_y = 0, size = 4)  +
+  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 0.05, nudge_y = 0, size = 4, lineheight = 0.8) +  # Ajustar el interlineado del texto de las etiquetas
   labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
-
 years_plot
-
-ggsave("output/graphs/years_plot.png", plot = years_plot, height = 6, width = 7, dpi = 300)
-
 ```
 
-```{r}
-# Calcular cuantiles para las variables
+##
 
-boxplot <- df_study_long_t6_comunas %>% select(just_educ, des_perc, merit_talent, merit_effort, ola) %>% as.data.frame()
-
-boxplot <- boxplot %>% 
-  pivot_longer(cols = c("just_educ", 
-                        "des_perc", 
-                        "merit_talent", 
-                        "merit_effort"), 
-               names_to = "variable",
-               values_to = "value") %>% na.omit()
-
-boxplot$variable <- factor(boxplot$variable,
-                   levels = c("just_educ", 
-                        "des_perc", 
-                        "merit_talent", 
-                        "merit_effort"),
-                   labels = c("Justification
-educational
-inequality", "Inequality
-perception", "Merit: Talent", "Merit: Effort"))
-
-boxplot_plot <- boxplot %>%
-  ggplot(aes(x = as.factor(ola), y = value, fill=variable, color=variable)) + 
-  geom_boxplot() +
-  scale_y_continuous(name = "Value", limits = c(1, 5)) +
-  scale_x_discrete(name = "Wave", labels=categorias_x) +
-  labs(fill = "Variable", color = "Variable") +
-  theme_bw() +
-  theme(axis.text.x = element_text(margin = margin(b = 15)))
-boxplot_plot
-
-ggsave(boxplot_plot, file="output/graphs/boxplot_plot.png", height = 7, width = 7)
-```
-
-## Correlation
-
-```{r cor-plot, fig.dim=c(10,6)}
-rm(list=ls())
-if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               sjPlot,
-               sjmisc,
-               summarytools,
-               lavaan
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent,
-                                               educ, quintil1:sexo) %>% na.omit()
-# Full sample_______________________________________________________________________
-mat_full<- 
-df_study1 %>%
-  # filter(ola==1) %>%
-  select(just_educ, # distributive justice
-         des_perc,
-         merit_effort, merit_talent # meritocracy
-         ) %>% 
-  psych::mixedCor()
-
-cor_full <- as.matrix(mat_full$rho)
-diag(cor_full) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_full) <- c("A. Justification of educational inequality",
-                        "B. Inequality perception",
-                     "C. People are rewarded for their efforts",
-                     "D. People are rewarded for their intelligence")
-#set Column names of the matrix
-colnames(cor_full) <-c("(A)", "(B)","(C)","(D)")
-#Plot the matrix using corrplot
-png("output/graphs/corr.png",width=600,height=300, pointsize = 10)
-corrplot::corrplot(cor_full,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-")
-# Agregar caption
-grid::grid.text("Source: own elaboration with pooled data 
-from ELSOC 2016-2022 (n = 8042)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
-dev.off()
-# Wave 1_______________________________________________________________________
-mat_1<- 
-df_study1_long_comunas %>%
-  filter(ola==1) %>%
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_1 <- as.matrix(mat_1$rho)
-diag(cor_1) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_1) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_1,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 1")
-
-# Wave 2_______________________________________________________________________
-mat_2<-   
-df_study1_long_comunas %>%
-  filter(ola==2) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_2 <- as.matrix(mat_2$rho)
-diag(cor_2) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_2) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_2,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 2")
-
-# Wave 3_______________________________________________________________________
-mat_3 <- 
-df_study1_long_comunas %>%
-  filter(ola==3) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_3 <- as.matrix(mat_3$rho)
-diag(cor_3) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_3) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_3,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 3")
-
-# Wave 4_______________________________________________________________________
-mat_4 <- 
-df_study1_long_comunas %>%
-  filter(ola==4) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_4 <- as.matrix(mat_4$rho)
-diag(cor_4) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_4) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_4,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 4")
-
-# Wave 6_______________________________________________________________________
-mat_6 <- 
-df_study1_long_comunas %>%
-  filter(ola==6) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_6 <- as.matrix(mat_6$rho)
-diag(cor_6) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_6) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_6) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_6,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 6")
-```
-
-# Models
-
-## A. Justification of educational inequality
+### Multivariate analysis
 
 ```{r distribution-educ}
-rm(list=ls())
-#if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total, just_educ, des_perc, merit_effort, merit_talent, educ, quintil1:sexo7) %>% 
-  na.omit() %>% 
-  mutate(ola=as.factor(ola),
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h1 <- "ola"
-h2 <- "ola+merit_effort+merit_talent"
-h3 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo"
-h4 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17"
-h5 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ"
-h6 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+univlag22_17"
-h7 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ+univlag22_17"
-h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna"
-h9 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+var_simce_comuna"
-h10 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna+var_simce_comuna"
-h11 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado"
-h12 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+pagadolag22_16"
-h13 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16"
-
-h13.1 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadolag22_16+prom_simce_comuna+var_simce_comuna+prop_univ+univlag22_17+gini_ypch+ginilag22_17"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ9 <- lmer(formula(paste0("just_educ~",h9,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ10 <- lmer(formula(paste0("just_educ~",h10,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ11 <- lmer(formula(paste0("just_educ~",h11,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ12 <- lmer(formula(paste0("just_educ~",h12,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ13 <- lmer(formula(paste0("just_educ~",h13,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-educ13.1 <- lmer(formula(paste0("just_educ~",h13.1,"+(1|idencuesta)+(1|comuna)")),data = df_study1, weights = ponderador_long_total)
-
-#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::knitreg(list(educ1, educ2, educ3, educ4, educ5, educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13,educ13.1),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
-#                custom.coef.names = c("Intercepto",
-#                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
-#                                      "ola 2018",
-#                                      "ola 2019",
-##                                      "ola 2022",
-#                                      "Merito: esfuerzo",
-#                                      "Merito: Inteligencia",
-##                                      "Goahead: Familia",
-#                                      "Goahead: educacion",
-#                                      "Goahead: ambicion",
-#                                      "Goahead: trabajo duro",
-#                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
-#                                      "Ed. técnica",
-#                                      "Ed. universitaria",
-#                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
-#                                      "Quintil 3",
-#                                      "Quintil 4",
-#                                      "Quintil 5",
-#                                      "Quintil NA",
-#                                      "Estatus social subjetivo",
-#                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
-#                                      "Pos. política: Derecha",
-#                                      "Pos. política: No se identifica",
-#                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
-#                                      "Edad 50-64",
-#                                      "Edad 65 o más",
-#                                      "Mujer <br> <i>(Ref. Hombre)</i>",
-#                                      "Gini comunal",
-#                                      "Cambio gini 2022-2017",
-#                                      "Proporción educ universitaria",
-#                                      "Proporción matrícula privada",
-#                                      "Total matrícula privada",
-#                                      "Varianza Simce comuna",
-#                                      "Promedio Simce comuna"
-
-                              )
-          #omit.coef = omit
-
-```
-
-## A2. Education distribution justice (export tables)
-
-```{r}
-#rm(list=ls())
-#if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_indep_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna,just_educ, merit_effort, merit_talent, ahead_family:educ,quintil1:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17, loginc_mean, logincomelag22_17) %>%
-  na.omit() %>% 
-  mutate(ola=as.factor(ola),
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h1 <- "ola"
-h2 <- "ola+merit_effort+merit_talent"
-h3 <- "ola+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
-h4 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family"
-h5 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo"
-
-h6 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17"
-h7 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_univ+univlag22_17"
-h8 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17"
-h9 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadlag22_16"
-h10 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
-h11 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
-h12 <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+loginc_mean+logincomelag22_17"
-
-hfull <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+loginc_mean+logincomelag22_17"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ9 <- lmer(formula(paste0("just_educ~",h9,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ10 <- lmer(formula(paste0("just_educ~",h10,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ11 <- lmer(formula(paste0("just_educ~",h11,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ12 <- lmer(formula(paste0("just_educ~",h12,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ13 <- lmer(formula(paste0("just_educ~",hfull,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-# Table 1: individual
-
-#omit <- "(pos_id)|(edad)|(cuestion_mig)|(nation_t1)|(quintil1QNA)|(Intercept)"
-texreg::htmlreg(list(educ1, educ2, educ3, educ4, educ5),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
-                custom.coef.names = c("Intercepto",
-                                      "ola 2017 <br> <i>(Ref. ola 2016)</i>",
-                                      "ola 2018",
-                                      "ola 2019",
-                                      "ola 2022",
-                                      "Merito: esfuerzo",
-                                      "Merito: Inteligencia",
-                                      "Goahead: trabajo duro",
-                                      "Goahead: educacion",
-                                      "Goahead: ambicion",
-                                      "Goahead: Familia",
-                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
-                                      "Ed. técnica",
-                                      "Ed. universitaria",
-                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
-                                      "Quintil 3",
-                                      "Quintil 4",
-                                      "Quintil 5",
-                                      "Quintil NA",
-                                      "Estatus social subjetivo",
-                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
-                                      "Pos. política: Derecha",
-                                      "Pos. política: No se identifica",
-                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Edad 50-64",
-                                      "Edad 65 o más",
-                                      "Mujer <br> <i>(Ref. Hombre)</i>"),
-                file = "output/tables/table1-individuales.html")
-          #omit.coef = omit
-
-webshot2::webshot(url ="output/tables/table1-individuales.html" ,file ="output/tables/table1-individuales.png")
-
-```
-
-```{r}
-# table 2: contextuales
-omit <- "(ola)|(merit_effort)|(merit_talent)|(ahead_hardwork)|(ahead_educ)|(ahead_ambition)|(ahead_family)|(educ)|(quintil1)|(ess)|(pos_id)|(edad)|(sexo)|(Intercept)"
-texreg::htmlreg(list(educ6, educ7,educ8,educ9,educ10,educ11,educ12,educ13),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05", 
-                omit.coef = omit,
-                custom.coef.names = c(
-                                      "Gini comunal",
-                                      "Cambio gini 2022-2017",
-                                      "Proporción educ universitaria",
-                                      "Cambio prop. universitaria 2022-2017",
-                                      "Proporción matrícula privada",
-                                      "Cambio matrícula privada 2022-2016",
-                                      "Promedio Simce comuna",
-                                      "Varianza Simce comuna",
-                                      "Ingreso comunal",
-                                      "Cambio ingreso comunal 2022-2017"),
-                file = "output/tables/table2-contextual.html")
-          #omit.coef = omit
-
-webshot2::webshot(url ="output/tables/table2-contextual.html" ,file ="output/tables/table2-contextual.png")
-
-```
-
-## Interacciones
-
-```{r distribution-educ-context}
-#rm(list=ls())
-#if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg,
-               ggplot2
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna,just_educ, merit_effort, merit_talent, ahead_family:educ,quintil1:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17, loginc_mean, logincomelag22_17) %>%
-  na.omit() %>% 
-  mutate(ola=as.factor(ola),
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h1b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled"
-h2b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*merit_effort"
-h3b <- "merit_effort+merit_talent+ola+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+merit_talent*ola"
-h4b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_hardwork"
-h5b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_educ"
-h6b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_ambition"
-h7b <- "ola+merit_effort+merit_talent+ahead_hardwork+ahead_educ+ahead_ambition+ahead_family+educ+quintil1+ess+pos_id+edad+sexo+gini_ypch+ginilag22_17+prop_univ+univlag22_17+prop_pagado+pagadlag22_16+prom_simce_comuna_rescaled+var_simce_comuna_rescaled+ola*ahead_family"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)+(1|comuna)")),data = df_study1)
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)+(1|comuna)")),data = df_study1)
-
-# table 3: interacciones
-omit <- "(educ)|(quintil1)|(ess)|(pos_id)|(edad)|(sexo)|(Intercept)|(gini_ypch)|(ginilag22_17)|(prop_univ)|(univlag22_17)|(prop_pagado)|(pagadlag22_16)|(prom_simce_comuna_rescaled)|(var_simce_comuna_rescaled)"
-texreg::htmlreg(list(educ1, educ2,educ3,educ4,educ5,educ6,educ7),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05",
-                custom.coef.names = c("Intercepto",
-                                      "2017",
-                                      "2018",
-                                      "2019",
-                                      "2022",
-                                      "Merito: esfuerzo",
-                                      "Merito: Inteligencia",
-                                      "Goahead: trabajo duro",
-                                      "Goahead: educacion",
-                                      "Goahead: ambicion",
-                                      "Goahead: Familia",
-                                      "Ed. media <br> <i>(Ref. Ed. básica)</i>",
-                                      "Ed. técnica",
-                                      "Ed. universitaria",
-                                      "Quintil 2 <br> <i>(Ref. Quintil 1)</i>",
-                                      "Quintil 3",
-                                      "Quintil 4",
-                                      "Quintil 5",
-                                      "Quintil NA",
-                                      "Estatus social subjetivo",
-                                      "Pos. política: Centro <br> <i>(Ref. izquierda)</i>",
-                                      "Pos. política: Derecha",
-                                      "Pos. política: No se identifica",
-                                      "Edad 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Edad 50-64",
-                                      "Edad 65 o más",
-                                      "Mujer <br> <i>(Ref. Hombre)</i>",
-                                      "Gini comunal",
-                                      "Cambio gini 2022-2017",
-                                      "Proporción educ universitaria",
-                                      "Cambio prop. universitaria 2022-2017",
-                                      "Proporción matrícula privada",
-                                      "Cambio matrícula privada 2022-2016",
-                                      "Promedio Simce comuna",
-                                      "Varianza Simce comuna",
-                                      "2017*merit:esfuerzo",
-                                      "2018*merit:esfuerzo",
-                                      "2019*merit:esfuerzo",
-                                      "2022*merit:esfuerzo",
-                                      "2017*merit:talento",
-                                      "2018*merit:talento",
-                                      "2019*merit:talento",
-                                      "2022*merit:talento",
-                                      "2017*goahead:hardwork",
-                                      "2018*goahead:hardwork",
-                                      "2019*goahead:hardwork",
-                                      "2022*goahead:hardwork",
-                                      "2017*goahead:educ",
-                                      "2018*goahead:educ",
-                                      "2019*goahead:educ",
-                                      "2022*goahead:educ",
-                                      "2017*goahead:ambition",
-                                      "2018*goahead:ambition",
-                                      "2019*goahead:ambition",
-                                      "2022*goahead:ambition",
-                                      "2017*goahead:family",
-                                      "2018*goahead:family",
-                                      "2019*goahead:family",
-                                      "2022*goahead:family"),
-                file = "output/tables/table3-interac.html")
-          #omit.coef = omit
-
-webshot2::webshot(url ="output/tables/table3-interac.html" ,file ="output/tables/table3-interac.png")
-
-## graph interact
-#sjPlot::plot_model(educ2, type = "int", mdrt.values = "meansd")
-predict<-interplot::interplot(m = educ2,var1 = "merit_effort",var2 = "ola",point = T)
-predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
-predict_data <- predict_data %>% as.data.frame()
-
-predict_data$value <- c("ola1", 
-                             "ola2",
-                             "ola3", 
-                             "ola4",
-                             "ola6")
-
-predict_data$value <- factor(predict_data$value,
-                   levels = c("ola1", 
-                             "ola2",
-                             "ola3", 
-                             "ola4",
-                             "ola6"),
-                   labels = c("wave 1", 
-                              "wave 2",
-                              "wave 3",
-                              "wave 4",
-                              "wave 6"))
-
-interact_effect <- predict_data %>% ggplot(aes(value, coef1)) +
-  theme_bw(base_size = 14) +
-  geom_point(size = 4)+
-  geom_errorbar(aes(ymin=lb, ymax=ub),
-                  width=0.2) +
-  ylab("Coef") +
-  xlab("")
-interact_effect
-
-ggsave(interact_effect, file="output/graphs/interact_effect.png", width = 7, height = 7)
-```
-
-## Ponderadores
-
-```{r}
-#rm(list=ls())
-#if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
 pacman::p_load(dplyr,       # Manipulacion de datos
@@
          )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
 
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+h6 <- "ola+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
 educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::htmlreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. <br>
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017 <br> <i>(Ref. Wave 2016)</i>",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
-                                      "Quintile 2 <br> <i>(Ref. Quintile 1)</i>",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile NA",
-#                                      "Subjective Social Status",
-                                      "Age 30-49 <br> <i>(Ref. 18-29)</i>",
-                                      "Age 50-64",
-                                      "Age 65 o más",
-                                      "Gender <br> <i>(Ref. Male)</i>",
-                                      "Pol. pos: Center <br> <i>(Ref. Left)</i>",
-                                      "Pol. pos: Right",
-                                      "Pol. pos: Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                custom.gof.names = c("AIC",
-                                     "BIC",
-                                     "Log Likelihood",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-                file = "output/tables/table1-individuales-pond-v2.html")
-          #omit.coef = omit
-
-webshot2::webshot(url ="output/tables/table1-individuales-pond-v2.html" ,file ="output/tables/table1-individuales-pond-v2.png")
-
+plot_model(educ6, 
+            rm.terms = c("educationHigh school", "educationTechnical", "educationUniversitary", "ess", "quintil1Q5"), #quitar terminos
+            title = "", #quitar titulo
+            show.values = TRUE, #mostrar valor de efectos
+            dot.size = 1.5, #tamaño circulos
+            line.size = 1, #tamaño CI
+            value.size = 3.5, #tamaño valor efectoss
+            spacing = 1, #espacio entre efectos
+            vline.color = "red", # linea roja en punto neutro (1)
+            axis.labels = rev(c("Year 2017", "Year 2018", "Year 2019", "Year 2022", 
+                                "Quintile 2", "Quintile 3", "Quintile 4", "Quintile 5",
+                                "Age 30-49", "Age 50-64", "Age 65 or more",
+                                "Female", "Center", "Right", "Does not identify",
+                                "Inequality perception", "Merit: Talent", "Merit: Effort")), # con rev porque automatico los tira en otro orden
+            group.terms = c(1,1,1,2,3,3,3,3,4,4,4,1,5,5,5,1,2,2),
+            value.offset = 0.4  # Ajusta este valor para mover el texto de los coeficientes hacia arriba
+) + # variables dependientes
+  theme_bw(base_size = 16)
 ```
 
-## Interacciones
+# Conclusions {data-background-color="black"}
 
-```{r distribution-educ-context}
-#rm(list=ls())
-#if (!require("pacman")) install.packages("pacman") # instalar pacman
-                            # cargar librerias
+## Conclusions
+
+::: {.incremental}
+
+- Decreasing pattern of justification of educational inequality between 2017-2019, but a increase in the last measurement point (2022).
+
+- Mixed evidence about the association between the perception of meritocracy and justification of inequality across time.
+
+- Throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education
+
+:::
+
+## Future research
+
+::: {.incremental}
+
+- Access vs outcomes in education
+
+- 7 wave of ELSOC (2024)
+
+- Justification of inequality in pensions and health (market justice preferences)
+:::
+
+# Thank you!
+
+- Kevin Carrasco ([kevin.carrasco@ug.uchile.cl](kevin.carrasco@ug.uchile.cl))
+
+- Github Repo: [https://github.com/justicia-distributiva/merit-edjust](https://github.com/justicia-distributiva/merit-edjust)
+
+# Anexos {data-background-color="black"}
+
+## Interaction effects
+
+```{r}
 pacman::p_load(dplyr,       # Manipulacion de datos
@@
          )
+h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
-# Hipotesis
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-#h2b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola+merit_effort*ola_2"
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
-#h4b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+des_perc*ola+des_perc*ola_2"
-#h5b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*pos_id"
-#h6b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*sexo"
-#h7b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*quintil1"
-#h8b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+ola_2*edad"
-
-# A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-#educ2 <- lmer(formula(paste0("just_educ~",h2b,"+(ola+ola_2+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-#educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+ola_2+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-#educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-#educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-#educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-#educ8 <- lmer(formula(paste0("just_educ~",h8b,"+(ola+ola_2|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::htmlreg(list(educ6,educ3,educ1),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. <br>
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, <br> political position and inequality perception",
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Merit: Talent",
-                                      "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
-                                      ),
-                custom.gof.names = c("AIC", "BIC", "Log Likelihood",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),
-                file = "output/tables/table3-interac_pond.html")
-          #omit.coef = omit
-
-webshot2::webshot(url ="output/tables/table3-interac_pond.html" ,file ="output/tables/table3-interac-pond.png")
+plot_model(educ1, 
+            type="int"
+) + # variables dependientes
+  theme_bw(base_size = 16)+
+  labs(title="",
+       x="Wave",
+       y="Justification of educational inequality",
+       color="Merit: Effort")
 ```
 
-```{r }
-fit <- lmer(just_educ ~ ola + ola_2 + (ola + ola_2 | idencuesta), data = df_study1)
-df_study1$pred_inter_pend_aleatorio <- predict(fit)
-
-# Predictions that ignore random effects:
-df_study1$fitted <- predict(fit, re.form = ~0)
-
-# Plot just fixed effects:
-plot<-df_study1 %>% 
-  ggplot(aes(y = fitted, x = ola_2)) +
-  geom_line() +
-  theme_light()
-
-ggsave(plot, file="output/graphs/fixed_effects.png")
-
-predict<-interplot::interplot(m = educ1,var1 = "ola",point = T)
-predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
-predict_data <- predict_data %>% as.data.frame()
-
-predict_data$value <- c("ola1", 
-                             "ola2",
-                             "ola3", 
-                             "ola4",
-                             "ola6")
-
-predict_data$value <- factor(predict_data$value,
-                   levels = c("ola1", 
-                             "ola2",
-                             "ola3", 
-                             "ola4",
-                             "ola6"),
-                   labels = c("wave 1", 
-                              "wave 2",
-                              "wave 3",
-                              "wave 4",
-                              "wave 6"))
-
-interact_effect <- predict_data %>% ggplot(aes(value, coef1)) +
-  theme_bw(base_size = 14) +
-  geom_point(size = 4)+
-  geom_errorbar(aes(ymin=lb, ymax=ub),
-                  width=0.2) +
-  ylab("Coef") +
-  xlab("")
-interact_effect
-
-ggsave(interact_effect, file="output/graphs/interact_effect-pond.png", width = 7, height = 7)
-```
+## References

### 2024-07-05 presentations/ISPP-santiago2024/ispp2024.qmd
update presentation

@@
 
-editor: visual
+editor: source
 ---
@@
 ::: {.columns .v-center-container}
-::: {.column width="30%"}
-![](qr.png){width="80%" fig-align="right"}
-<br>
-![](images/logocoes.png){width="90%" fig-align="right"}
-<br>
-![](images/edumer.png){width="80%" fig-align="left"}
+::: {.column width="20%"}
+![](qr.png){width="80%" fig-align="right"} <br> ![](images/logocoes.png){width="90%" fig-align="right"} <br> ![](images/edumer.png){width="80%" fig-align="left"}
 :::
 
-::: {.column width="70%"}
-
-## **Changes in the justification of educational inequalities:**
+::: {.column width="80%"}
+## Changes in the justification of educational inequalities:
 
@@
 
-----------------------------------------------------------------
+------------------------------------------------------------------------
 
-Juan Carlos Castillo, Julio Iturra & **Kevin Carrasco* **
+Juan Carlos Castillo, Julio Iturra & **Kevin Carrasco\***
 
-#### Fondecyt N° 1210847: Meritocracia en la escuela
-#### Center for social conflict and cohesion studies
+Fondecyt N° 1210847: Meritocracia en la escuela
+
+Center for social conflict and cohesion studies
 
@@
 
-# Introduction {data-background-color="black"}
+## Research question {data-background-color="black"}
 
-## Introduction
+### To what extent has the justification of educational inequality changed over time?
 
-::: {.incremental}
+### to what extent does the perception of meritocracy influence the justification of educational inequality?
 
-::: {.columns .v-center-container}
-::: {.column width="50%" .incremental}
+## Team
 
-### Justification of inequality
+![](images/team.png){width="100%" fig-align="center"}
+
+## Justification of educational inequality
+
+::: incremental
 
@@
 
-::: {.column width="50%" .incremental}
-### Meritocracy
+::: {.notes}
+
+This research focuses on analyzing the differences that exist in the justifications for educational inequality in the Chilean context.
+Research in the field of social justice has distinguished, on the one hand, the perception of economic inequality and, on the other, the extent to which people are willing to tolerate these levels of inequality.
+In this sense, this research focuses on the second aspects, trying to address the degree to which people justify unequal access to better education.
+
+Furthermore, differences between individual and contextual variables have been evident in the literature, where the Chilean context becomes relevant as it is marked by a strong marketization of social services based on neoliberal reforms that have privatized different areas of the social sphere, such as education is.
+
+Added to this general framework, we also propose that this justification of inequality has varied over time and that, above all, it should decrease in times of crisis such as the pandemic, since it is in these moments when inequalities increase and therefore People should be less willing to tolerate greater inequality, especially among lower-status groups.
+
+:::
+
+## Meritocracy
+
+::: incremental
 
@@
 
-- Normative and perceptual elements [@duru-bellat_who_2012; @castillo_meritocracia_2019]
+-   Normative and perceptual elements [@duru-bellat_who_2012; @castillo_multidimensional_2023]
 
 :::
-:::
 
-::: {.r-stack style="font-size: 130%;"}
-- Covid pandemic [@breznau_welfare_2021; @hankivsky_introduction_2022]
-:::
+::: {.notes}
+
+The second central aspect of this research is that of meritocracy, which has its origins in the book Rise of the meritocracy by Michel Young, and which points to the belief or perception that social rewards are assigned based on effort and talent.
+This line of research has also been divided into two aspects, normative and perceptual, differentiated between whether people believe that rewards should be assigned based on effort and talent or whether this is the case in society.
+
+In this research, the focus is on the second aspect, addressing whether people perceive whether rewards are assigned based on effort and talent.
 
@@
 
+::: {.notes}
+
+In summary, this is the general scheme that we propose in this research, where the justification of educational inequality is our dependent variable and which should be affected by the perception of meritocracy, in addition to being conditioned by other factors such as the time and the pandemic of covid and individual factors of each respondent like perception of inequality, gender, age, political positions.
+
+:::
+
 # Methods {data-background-color="black"}
@@
 ## Data
-
-::: {.incremental}
 
@@
 
+-   More information in: [https://coes.cl/encuesta-panel/](https://coes.cl/encuesta-panel/)
+
+::: {.smaller}
+
+![](images/ELSOC.png){width="50%" fig-align="center"}
 :::
 
-## Variables
+::: {.notes}
 
-```{r bivariate-plots, echo=FALSE}
+To carry out this research, we used data from the Chilean social longitudinal study carried out by the COES between 2016 and 2022.
+However, this survey was not carried out in 2020 and in 2021 it was carried out a short version by telephone and in summary, not including the main dependent variable of this research.
+Thus, we have the responses of one thousand sixty-nine people for the 5 waves in which all these intems are available.
+
+:::
+
+##
+
+### **It's just that people with higher income have a better education for their children than people with lower income?**
+
+```{r dependent, echo=FALSE}
 df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
@@
 merit_plot <- df_study1 %>% 
-  select(merit_effort, merit_talent, des_perc, just_educ) %>% 
+  select(just_educ) %>% 
   sjPlot::plot_stackfrq(weight.by = df_study1$ponderador_long_total,
@@
                       show.total = FALSE, 
-                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),4),
-                      legend.labels = c("Strongly disagree", "Disagree", "Neither agree
-nor disagree", "Agree", "Strongly agree"),
-                      axis.labels = c("Merit effort", "Merit talent", "Inequality 
-perception", "Justification 
-of 
-educational inequality")
+                      vjust=c("bottom", "top", "bottom", "top", "bottom"),
+                      legend.labels = c("Strongly disagree", "Disagree", "Neither agree\n nor disagree", "Agree", "Strongly agree"),
+                      axis.labels = c("")
                       ) +
-  theme_light(base_size = 16) +
-  theme(legend.position="bottom", text = element_text(size = 14),
+  theme_light(base_size = 20) +
+  theme(legend.position="bottom", text = element_text(size = 16),
         plot.caption = element_text(hjust = 1)) +
@@
 
-## Variables {.smaller}
+::: {.notes}
 
-```{r error=FALSE}
-df_study1 <- df_study_long_t6_comunas %>% dplyr::select(ola,
-                                               just_educ, des_perc, merit_talent, merit_effort,
-                                               educ, quintil1:sexo)
-
-df_study1$ola <- factor(df_study1$ola,
-         labels = c("2016","2017","2018","2019", "2022"))
-df_study1$ola <- 
-sjlabelled::set_label(x = df_study1$ola,
-                      label = "Wave")
-
-df_study1$sexo <- factor(df_study1$sexo,
-         labels = c("Male", "Female"))
-df_study1$sexo <- 
-sjlabelled::set_label(x = df_study1$sexo,
-                      label = "Gender")
-
-for (i in c("just_educ", "des_perc",
-            "merit_effort","merit_talent")) {
-  df_study1[[i]] <- factor(df_study1[[i]],
-                            levels = 1:5,
-                            labels = c("Strongly disagree","Disagree",
-                                       "Neither disagree nor agree",
-                                       "Agree", "Strongly agree"))
-}
-
-df_study1$just_educ <- 
-sjlabelled::set_label(x = df_study1$just_educ, 
-                      label = "Justification of educational inequality")
-
-#
-df_study1$des_perc <- 
-sjlabelled::set_label(x = df_study1$des_perc, 
-                      label = "Income differences are too large")
-
-#
-df_study1$merit_effort <- 
-sjlabelled::set_label(x = df_study1$merit_effort, 
-                      label = "People are rewarded for their efforts")
-
-#
-df_study1$merit_talent <- 
-sjlabelled::set_label(x = df_study1$merit_talent, 
-                      label = "People are rewarded for their intelligence and skills")
-
-lab_study1 <- sjlabelled::get_label(df_study1) 
-df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
-```
-
-::: {.columns .v-center-container}
-
-::: {.column width="50%"}
-
-```{r error=FALSE}
-df<- dfSummary(x = select(df_study1, ola, quintil1, sexo),
-               plain.ascii = FALSE,
-               style = "grid",
-               tmp.img.dir = "/tmp",
-               graph.magnif = TRUE,
-               headings = F,  # encabezado
-               varnumbers = F, # num variable
-               labels.col = T, # etiquetas
-               na.col = F,    # missing
-               graph.col = F, # plot
-               valid.col = F, # n valido
-               col.widths = c(10,10,10)
-               )
-df$Variable <- NULL # delete variable column
-
-print(df, method='render',
-      footnote = NA, 
-      headings = FALSE, 
-      style = list(
-        table = "font-size: 8px;",
-        th = "font-size: 8px;",
-        td = "font-size: 8px;"
-      ))
-```
+Therefore, the main variable of this research is the justification of educational inequality, which in general has a wide degree of disagreement, that is, people who do not justify access to a better education for those who can afford it.
 
 :::
-
-::: {.column width="50%"}
-```{r error=FALSE}
-df<- dfSummary(x = select(df_study1, edad, pos_id),
-               plain.ascii = FALSE,
-               style = "grid",
-               tmp.img.dir = "/tmp",
-               graph.magnif = TRUE,
-               headings = F,  # encabezado
-               varnumbers = F, # num variable
-               labels.col = T, # etiquetas
-               na.col = F,    # missing
-               graph.col = F, # plot
-               valid.col = F, # n valido
-               col.widths = c(10,10,10)
-               )
-df$Variable <- NULL # delete variable column
-
-print(df, method='render',
-      footnote = NA, 
-      headings = FALSE, 
-      style = list(
-        table = "font-size: 8px;",
-        th = "font-size: 8px;",
-        td = "font-size: 8px;"
-      ))
-```
-:::
-
-:::
-
-# Results {data-background-color="black"}
 
@@
   mutate(idencuesta = 1)
-datos.justeduc$just_educ <- factor(datos.justeduc$just_educ, levels = rev(levels(datos.justeduc$just_educ)))
-etiquetas.justeduc$just_educ <- factor(etiquetas.justeduc$just_educ, levels = rev(levels(etiquetas.justeduc$just_educ)))
+datos.justeduc$just_educ <- factor(datos.justeduc$just_educ, levels = levels(datos.justeduc$just_educ))
+etiquetas.justeduc$just_educ <- factor(etiquetas.justeduc$just_educ, levels = levels(etiquetas.justeduc$just_educ))
 ```
@@
   scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022)) +
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)",
+          title="It's just that people with higher income have a better education for their children than people with lower income?")
+
+alluvial_justeduc +theme_light()
+```
+
+::: {.notes}
+
+In the same way, it is interesting to note that although in 2017 there was a decrease in the justification of inequality, it began to grow in the following years and reached its highest level after the pandemic.
+Therefore, contrary to our first hypothesis, this first result goes in the opposite direction, with a great variation among those who respond to the highest categories of disagreement.
+
+:::
+
+## Independent variables
+
+```{r independent, echo=FALSE}
+df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               educ, quintil1:sexo) %>% na.omit()
+merit_plot <- df_study1 %>% 
+  select(merit_effort, merit_talent) %>% 
+  sjPlot::plot_stackfrq(weight.by = df_study1$ponderador_long_total,
+                        expand.grid =TRUE,
+                      geom.colors = "PuBu",
+                      show.total = FALSE, 
+                      vjust=rep(c("bottom", "top", "bottom", "top", "bottom"),2),
+                      legend.labels = c("Strongly disagree", "Disagree", "Neither agree\n nor disagree", "Agree", "Strongly agree"),
+                      axis.labels = c("Merit effort", "Merit talent")
+                      ) +
+  theme_light(base_size = 20) +
+  theme(legend.position="bottom", text = element_text(size = 16),
+        plot.caption = element_text(hjust = 1)) +
   labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
+merit_plot
 
-alluvial_justeduc
 ```
+
+::: {.notes}
+
+On the other hand, the main independent variables are two variables of perception of meritocracy, that is, whether effort and talent are rewarded in Chilean society.
+
+:::
+
+## Independent variables
+
+::: {.incremental}
+
+- Inequality perception
+
+- Wave
+
+- Household income quintile per capita
+
+- Gender
+
+- Age groups
+
+- Political identification
+
+:::
+
+# Results {data-background-color="black"}
 
@@
 
-ineq_years = elsoc_diseno %>% #filter(!is.na(ola)) %>% 
-  group_by(ola) %>%
-  summarise(inequality = survey_mean(des_perc, vartype = "ci", na.rm = T),
-            ) %>% 
-  ungroup()
-
-ineq_years <- ineq_years %>% 
-  pivot_longer(cols = c(
-                        "inequality"), 
-               names_to = "variable",
-               values_to = "value") %>% 
-  cbind(type=c("Independent")) %>% 
-  select(ola, variable, value, type, value_upp=inequality_upp, value_low=inequality_low)
-
 talent_years = elsoc_diseno %>% #filter(!is.na(ola)) %>% 
@@
 
-vars_years <- rbind(just_years, ineq_years, talent_years, effort_years)
+vars_years <- rbind(just_years, talent_years, effort_years)
 
 categorias_x <- c(2016, 2017, 2018, 2019, 2022)
-vars_years$variable <- factor(vars_years$variable, levels = c("education", "inequality", "talent", "effort"), labels=  c("Justification
+vars_years$variable <- factor(vars_years$variable, levels = c("education", "talent", "effort"), labels=  c("Justification
 educational
 inequality
-(dependent)", "Inequality
-perception", "Merit: Talent", "Merit: Effort"),
+(dependent)", "Merit: Talent", "Merit: Effort"),
 ordered = TRUE)
@@
 inequality
-(dependent)", "Inequality
-perception
-(independent)", "Merit: Talent
+(dependent)", "Merit: Talent
 (independent)", "Merit: Effort
@@
 ## Crear el gráfico principal
-years_plot <- ggplot(vars_years, aes(ola, value, color = variable, shape = variable, linetype = variable)) +
+years_plot <- vars_years %>% filter(type=="Dependent") %>% 
+ggplot(aes(ola, value, color = variable, shape = variable, linetype = variable)) +
   theme_bw(base_size = 14) +
@@
   xlab("Wave") +
-  ylim(1, 5) +
+#  ylim(1, 5) +
+  scale_y_continuous(breaks = c(1, 2, 3, 4, 5),
+                     labels = c("Strongly\n disagree", "Disagree", "Neither agree\n nor disagree", "Agree", "Strongly\n agree"),
+                     limits = c(1, 5)) +
   scale_x_discrete(labels = categorias_x) +
-  scale_color_manual(labels = categorias_z, name = "Variable", values =  c("#bdc9e1ff","#0570b0ff","#74a9cfff","#b3b3b3ff")) +
-  scale_shape_manual(labels = categorias_z, name = "Variable", values = c("square", "circle", "triangle", "diamond")) +
+  scale_color_manual(labels = categorias_z, name = "Variable", values =  c("#bdc9e1ff")) +
+  scale_shape_manual(labels = categorias_z, name = "Variable", values = c("square")) +
   scale_linetype_manual(labels = categorias_z, name = "Variable", 
-                        values = c("solid", "dashed", "dotted", "dotdash")) +  # Definir diferentes tipos de línea
+                        values = c("solid")) +  # Definir diferentes tipos de línea
+   theme(legend.position = "none") +
+  coord_cartesian(xlim = c(1, 6)) +  # Ampliar el eje x
+  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022, ""))  # Personalizar etiquetas del eje x
+
+# Obtener los datos para etiquetar las líneas
+label_data <- vars_years %>% filter(type=="Dependent") %>% 
+  group_by(variable) %>%
+  filter(row_number() == n()) %>%
+  ungroup() %>%
+  mutate(label_x = 5.5, label_y = value)
+
+# Añadir etiquetas personalizadas
+years_plot <- years_plot +
+  geom_point(data = label_data, aes(label_x - 0.05, label_y, color = variable, shape = variable), size = 4) +  # Capa de puntos para los símbolos de forma
+  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 0.05, nudge_y = 0, size = 4, lineheight = 0.8) +  # Ajustar el interlineado del texto de las etiquetas
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
+years_plot
+```
+
+::: {.notes}
+
+As a summary, we have here the variation in the average justification of educational inequality during these years, with the maximum reached in 2022
+
+:::
+
+##
+
+### Changes in the main independent variables over years (2016-2022)
+
+```{r}
+## Crear el gráfico principal
+years_plot <- vars_years %>% 
+ggplot(aes(ola, value, color = variable, shape = variable, linetype = variable)) +
+  theme_bw(base_size = 14) +
+  geom_point(size = 4) +
+  geom_line(aes(group = variable)) +
+  geom_errorbar(aes(ymin = value_low, ymax = value_upp), width = 0.2) +
+  ylab("Mean") +
+  xlab("Wave") +
+#  ylim(1, 5) +
+  scale_y_continuous(breaks = c(1, 2, 3, 4, 5),
+                     labels = c("Strongly\n disagree", "Disagree", "Neither agree\n nor disagree", "Agree", "Strongly\n agree"),
+                     limits = c(1, 5)) +
+  scale_x_discrete(labels = categorias_x) +
+  scale_color_manual(labels = categorias_z, name = "Variable", values =  c("#bdc9e1ff","#0570b0ff","#74a9cfff")) +
+  scale_shape_manual(labels = categorias_z, name = "Variable", values = c("square", "circle", "triangle")) +
+  scale_linetype_manual(labels = categorias_z, name = "Variable", 
+                        values = c("solid", "dashed", "dotted")) +  # Definir diferentes tipos de línea
    theme(legend.position = "none") +
@@
 ```
+
+::: {.notes}
+
+But what I want to show is the comparison with the variables of perception of meritocracy based on talent and effort, which both follow a similar pattern, increasing in 2018 and then falling again until 2022.
+
+:::
 
@@
 
-h6 <- "ola+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+h6 <- "merit_talent+merit_effort+ola+des_perc+education+quintil1+ess+edad+sexo+pos_id"
 educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
@@
             vline.color = "red", # linea roja en punto neutro (1)
-            axis.labels = rev(c("Year 2017", "Year 2018", "Year 2019", "Year 2022", 
+            axis.labels = rev(c("Merit: Talent", "Merit: Effort",
+                                "Year 2017", "Year 2018", "Year 2019", "Year 2022",
+                                "Inequality perception", 
                                 "Quintile 2", "Quintile 3", "Quintile 4", "Quintile 5",
                                 "Age 30-49", "Age 50-64", "Age 65 or more",
-                                "Female", "Center", "Right", "Does not identify",
-                                "Inequality perception", "Merit: Talent", "Merit: Effort")), # con rev porque automatico los tira en otro orden
-            group.terms = c(1,1,1,2,3,3,3,3,4,4,4,1,5,5,5,1,2,2),
+                                "Female", "Center", "Right", "Does not identify")), # con rev porque automatico los tira en otro orden
+            group.terms = c(2,2,1,1,1,2,1,3,3,3,3,4,4,4,1,5,5,5),
             value.offset = 0.4  # Ajusta este valor para mover el texto de los coeficientes hacia arriba
@@
 
-# Conclusions {data-background-color="black"}
+::: {.notes}
 
-## Conclusions
+Finally, we carried out a linear regression analysis to evaluate each of the variables, and following our main hypotheses, the perception of meritocracy based on effort, but not talent, influences positively the justification of educational inequality.
 
-::: {.incremental}
+Also, in this results we can see that although in the first years the justification of inequality in education decreases compared to 2016, it is greater for 2022.
 
+In relation with the rest of control variables, the perception of inequality negatively influences the justification of educational inequality, the income quantiles, only the two greatest have a significant effect that is positive, while in age only the category of those over sixty-five years of age is significant.
+On the other hand, being a woman is negatively associated with the justification of educational inequality, while, compared to those who are on the left, being from the center, right, or those who do not identify with any of these options justify greater educational inequality.
+
+:::
+
+## Conclusions {data-background-color="black"}
+
+::: incremental
 -   Decreasing pattern of justification of educational inequality between 2017-2019, but a increase in the last measurement point (2022).
@@
 -   Throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education
+:::
+
+::: {.notes}
+
+In this way, we can conclude that there is a pattern of decreasing justification of educational inequality between 2017 and 2019, but with an increase in the last measurement in 2022.
+
+We found mixed evidence on the association of the perception of meritocracy and the justification of educational inequality, since effort is significant but not talent, which could be attributed to the fact that talent is rather randomly assigned and does not depend so much on the individual capabilities.
+
+In any case, over the years a greater perception of economic inequality influences a lower justification of educational inequality, coinciding with previous literature.
 
@@
 
-::: {.incremental}
-
+::: incremental
 -   Access vs outcomes in education
 
-- 7 wave of ELSOC (2024)
+-   7 wave of ELSOC (2023)
 
@@
 
+::: {.notes}
+
+For future research, it would be interesting to delve deeper into what is perceived as justification for educational inequality, since there may be differences in inequality in access to education with inequality in educational results, as well as by level, such as at the school level.
+or university level.
+
+We will now also soon have the seventh wave of elsoc available, which will be interesting to evaluate how it has varied in the last year
+
+and finally we also aim to be able to evaluate similar factors that could influence other aspects of society such as pensions and health, which in other research we have called market justice preferences.
+
+:::
+
 # Thank you!
 
-- Kevin Carrasco ([kevin.carrasco@ug.uchile.cl](kevin.carrasco@ug.uchile.cl))
+-   Kevin Carrasco ([kevin.carrasco\@ug.uchile.cl](kevin.carrasco@ug.uchile.cl))
 
@@
 
-# Anexos {data-background-color="black"}
+## References
 
-## Interaction effects
+::: {#refs}
+:::
 
-```{r}
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg,
-               ggplot2
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
+## {.appendix}
 
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+::: {.columns}
+::: {.column width="20%"}
+![](qr.png){width="80%" fig-align="right"} <br> ![](images/logocoes.png){width="90%" fig-align="right"} <br> ![](images/edumer.png){width="80%" fig-align="left"}
+:::
 
-plot_model(educ1, 
-            type="int"
-) + # variables dependientes
-  theme_bw(base_size = 16)+
-  labs(title="",
-       x="Wave",
-       y="Justification of educational inequality",
-       color="Merit: Effort")
-```
+::: {.column width="80%"}
+## Changes in the justification of educational inequalities:
 
-## References
+### The role of perceptions of inequality and meritocracy during the COVID pandemic
+
+------------------------------------------------------------------------
+
+Juan Carlos Castillo, Julio Iturra & **Kevin Carrasco\***
+
+Fondecyt N° 1210847: Meritocracia en la escuela
+
+Center for social conflict and cohesion studies
+
+**ISPP Annual meeting** Chile, July 2024
+:::
+:::

### 2024-12-16 01-introduction.qmd
change to quarto + add ola 7

--- 2024-03-26 paper.Rmd
@@
----
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
-css: "custom.css"
-linestretch: '1.5'
-link-citations: yes
-# params:
-# lang: en-GB
-# author: 
-# - name: Juan Carlos Castillo
-#   affiliation: Department of Sociology, Universidad de Chile
-#   email: juancastillov@uchile.cl
-#   number: 1, 3 
-# - name: Julio Iturra
-#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-#   number: 2
-# - name: Kevin Carrasco
-#   affiliation: Center for Social Conflict and Cohesion Studies
-#   number: 3
-abstract: | 
-  Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-  .
-  \newline
-  \newline
-
-   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-output:
-  bookdown::pdf_document2:
-    template: null
-    number_sections: true    
-    toc: false
-    keep_tex: false
-    pandoc_args:
-      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
-  bookdown::word_document2:
-    number_sections: false
-    # reference_docx: input/template.docx
-    toc: false
-  bookdown::html_document2:
-    number_sections: false
-linkcolor: blue
-bibliography:
-  - input/bib/merit-edjust.bib
-csl: input/bib/apa7-singlespace.csl
-editor_options:
-  chunk_output_type: console
-geometry: margin=1in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-always_allow_html: true # for word output
-fontsize: 12pt
-papersize: a4
----
-
-```{r eval=FALSE, include=FALSE}
- # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
- # clean #in the yml
- rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
-```
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-\pagebreak
-
 # Introduction
@@
 Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
-# Inequality, beliefs and justification
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p.
-3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021].
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
-30].
-However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p.
-2].
-This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978].
-Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.
-
- Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986].
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
-Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-# The justification of educational inequality
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services.
-For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-_Justificacion, inequality and meritocracy in the educational domain_
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-_This study_
-
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-$H_1$: The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
-
-$H_2$: Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-$H_5$: The positive association between meritocracy and inequality justification mitigates along time
-
-#  Data, variables & methods
-
-## Data
-
-The main data source is the Chilean Longitudinal Social Survey 2016–2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
-The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.
-
-## Variables
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
-In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
-
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/merit_plot.png")
-```
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
-
-```{r descriptives, echo=FALSE, results='asis'}
-pacman::p_load(kableExtra)
-kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
-      format="pipe",
-             caption="Independent variables",
-             col.names = "",
-             row.names = F)
-```
-
-## Methods
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models [@singer_applied_2009].
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
-
-```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-  #RUN THIS BEFORE THE ANALYSIS
-  pacman::p_load(knitr)
-  table_format = if(is_html_output()) {
-    "html"
-  } else if(is_latex_output()) {
-    "latex"
-  }
-
-  table_format2 = if(is_html_output()) {
-    T
-  } else if(is_latex_output()) {
-    F
-  }
-```
-
-# Results
-
-```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
- knitr::include_graphics("output/graphs/alluvial_dep.png")
-```
-
-Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
-
-```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
- knitr::include_graphics("output/graphs/years_plot.png")
-```
-
-Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
-
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/corr.png")
-```
-
-Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
-
-## Multivariate analysis
-
-Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
-
-```{r multilevel, echo=FALSE, results='asis'}
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-#                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Household Income (Ref.= Quintile 1)" = 8:12,
-                  "Age (Ref.= 18-29)" = 13:15,
-                  "Pol. pos (Ref.= Left)" = 17:19
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```
-
-```{r interact, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
-
-# Hipotesis
-
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
-
-# A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::texreg(list(educ6,educ3,educ1),
-                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
-                omit.coef = omit,
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Merit: Talent",
-                                      "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
-                                      ),
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),  
-               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = TRUE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE,
-               threeparttable = TRUE)
-```
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-# Discussion and conclusions
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
-
-Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-<!-- _Acknowledgments_ -->
-
-<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
-
-# References
-
-::: {#refs}
-:::
-
-\newpage 
-# (APPENDIX) Appendix {.unnumbered}
-
-`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
-
-```{r multilevel-full, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
-#                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                              "Education (Ref.=Primary)"= 8:10,
-                  "Household Income (Ref.= Quintile 1)" = 11:15,
-                  "Age (Ref.= 18-29)" = 17:19,
-                  "Pol. pos (Ref.= Left)" = 21:23
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.60,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```

### 2024-12-16 02-antecedentes.qmd
change to quarto + add ola 7

--- 2024-03-26 paper.Rmd
@@
----
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
-css: "custom.css"
-linestretch: '1.5'
-link-citations: yes
-# params:
-# lang: en-GB
-# author: 
-# - name: Juan Carlos Castillo
-#   affiliation: Department of Sociology, Universidad de Chile
-#   email: juancastillov@uchile.cl
-#   number: 1, 3 
-# - name: Julio Iturra
-#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-#   number: 2
-# - name: Kevin Carrasco
-#   affiliation: Center for Social Conflict and Cohesion Studies
-#   number: 3
-abstract: | 
-  Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-  .
-  \newline
-  \newline
-
-   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-output:
-  bookdown::pdf_document2:
-    template: null
-    number_sections: true    
-    toc: false
-    keep_tex: false
-    pandoc_args:
-      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
-  bookdown::word_document2:
-    number_sections: false
-    # reference_docx: input/template.docx
-    toc: false
-  bookdown::html_document2:
-    number_sections: false
-linkcolor: blue
-bibliography:
-  - input/bib/merit-edjust.bib
-csl: input/bib/apa7-singlespace.csl
-editor_options:
-  chunk_output_type: console
-geometry: margin=1in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-always_allow_html: true # for word output
-fontsize: 12pt
-papersize: a4
----
-
-```{r eval=FALSE, include=FALSE}
- # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
- # clean #in the yml
- rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
-```
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-\pagebreak
-
-# Introduction
-
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
-
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
-Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
-
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021].
-A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
 # Inequality, beliefs and justification
@@
 $H_5$: The positive association between meritocracy and inequality justification mitigates along time
-
-#  Data, variables & methods
-
-## Data
-
-The main data source is the Chilean Longitudinal Social Survey 2016–2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
-The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.
-
-## Variables
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
-In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
-
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/merit_plot.png")
-```
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
-
-```{r descriptives, echo=FALSE, results='asis'}
-pacman::p_load(kableExtra)
-kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
-      format="pipe",
-             caption="Independent variables",
-             col.names = "",
-             row.names = F)
-```
-
-## Methods
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models [@singer_applied_2009].
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
-
-```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-  #RUN THIS BEFORE THE ANALYSIS
-  pacman::p_load(knitr)
-  table_format = if(is_html_output()) {
-    "html"
-  } else if(is_latex_output()) {
-    "latex"
-  }
-
-  table_format2 = if(is_html_output()) {
-    T
-  } else if(is_latex_output()) {
-    F
-  }
-```
-
-# Results
-
-```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
- knitr::include_graphics("output/graphs/alluvial_dep.png")
-```
-
-Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
-
-```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
- knitr::include_graphics("output/graphs/years_plot.png")
-```
-
-Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
-
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/corr.png")
-```
-
-Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
-
-## Multivariate analysis
-
-Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
-
-```{r multilevel, echo=FALSE, results='asis'}
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-#                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Household Income (Ref.= Quintile 1)" = 8:12,
-                  "Age (Ref.= 18-29)" = 13:15,
-                  "Pol. pos (Ref.= Left)" = 17:19
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```
-
-```{r interact, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
-
-# Hipotesis
-
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
-
-# A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::texreg(list(educ6,educ3,educ1),
-                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
-                omit.coef = omit,
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Merit: Talent",
-                                      "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
-                                      ),
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),  
-               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = TRUE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE,
-               threeparttable = TRUE)
-```
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-# Discussion and conclusions
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
-
-Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-<!-- _Acknowledgments_ -->
-
-<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
-
-# References
-
-::: {#refs}
-:::
-
-\newpage 
-# (APPENDIX) Appendix {.unnumbered}
-
-`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
-
-```{r multilevel-full, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
-#                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                              "Education (Ref.=Primary)"= 8:10,
-                  "Household Income (Ref.= Quintile 1)" = 11:15,
-                  "Age (Ref.= 18-29)" = 17:19,
-                  "Pol. pos (Ref.= Left)" = 21:23
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.60,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```

### 2024-12-16 03-methods.qmd
change to quarto + add ola 7

--- 2024-03-26 paper.Rmd
@@
----
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
-css: "custom.css"
-linestretch: '1.5'
-link-citations: yes
-# params:
-# lang: en-GB
-# author: 
-# - name: Juan Carlos Castillo
-#   affiliation: Department of Sociology, Universidad de Chile
-#   email: juancastillov@uchile.cl
-#   number: 1, 3 
-# - name: Julio Iturra
-#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-#   number: 2
-# - name: Kevin Carrasco
-#   affiliation: Center for Social Conflict and Cohesion Studies
-#   number: 3
-abstract: | 
-  Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-  .
-  \newline
-  \newline
-
-   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-output:
-  bookdown::pdf_document2:
-    template: null
-    number_sections: true    
-    toc: false
-    keep_tex: false
-    pandoc_args:
-      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
-  bookdown::word_document2:
-    number_sections: false
-    # reference_docx: input/template.docx
-    toc: false
-  bookdown::html_document2:
-    number_sections: false
-linkcolor: blue
-bibliography:
-  - input/bib/merit-edjust.bib
-csl: input/bib/apa7-singlespace.csl
-editor_options:
-  chunk_output_type: console
-geometry: margin=1in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-always_allow_html: true # for word output
-fontsize: 12pt
-papersize: a4
----
-
-```{r eval=FALSE, include=FALSE}
- # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
- # clean #in the yml
- rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
-```
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-\pagebreak
-
-# Introduction
-
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
-
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
-Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
-
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021].
-A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
-# Inequality, beliefs and justification
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p.
-3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021].
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
-30].
-However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p.
-2].
-This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978].
-Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.
-
- Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986].
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
-Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-# The justification of educational inequality
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services.
-For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-_Justificacion, inequality and meritocracy in the educational domain_
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-_This study_
-
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-$H_1$: The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
-
-$H_2$: Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-$H_5$: The positive association between meritocracy and inequality justification mitigates along time
-
 #  Data, variables & methods
@@
 In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
+@fig-frecuencies shows the average distribution throughout the years for these variables.
 We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
@@
 
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
+```{r echo=FALSE, fig.align='center'}
+#| label: fig-frecuencies
+#| fig-cap: "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed"
+
  knitr::include_graphics("output/graphs/merit_plot.png")
@@
 Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
+@tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
 
-```{r descriptives, echo=FALSE, results='asis'}
+```{r echo=FALSE, results='asis'}
+#| label: tbl-descriptives
+#| tbl-cap: "Independent variables"
+
 pacman::p_load(kableExtra)
@@
       format="pipe",
-             caption="Independent variables",
              col.names = "",
@@
 The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
-
-```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-  #RUN THIS BEFORE THE ANALYSIS
-  pacman::p_load(knitr)
-  table_format = if(is_html_output()) {
-    "html"
-  } else if(is_latex_output()) {
-    "latex"
-  }
-
-  table_format2 = if(is_html_output()) {
-    T
-  } else if(is_latex_output()) {
-    F
-  }
-```
-
-# Results
-
-```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
- knitr::include_graphics("output/graphs/alluvial_dep.png")
-```
-
-Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
-
-```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
- knitr::include_graphics("output/graphs/years_plot.png")
-```
-
-Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
-
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/corr.png")
-```
-
-Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
-
-## Multivariate analysis
-
-Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
-
-```{r multilevel, echo=FALSE, results='asis'}
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-#                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Household Income (Ref.= Quintile 1)" = 8:12,
-                  "Age (Ref.= 18-29)" = 13:15,
-                  "Pol. pos (Ref.= Left)" = 17:19
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```
-
-```{r interact, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
-
-# Hipotesis
-
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
-
-# A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::texreg(list(educ6,educ3,educ1),
-                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
-                omit.coef = omit,
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Merit: Talent",
-                                      "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
-                                      ),
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),  
-               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = TRUE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE,
-               threeparttable = TRUE)
-```
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-# Discussion and conclusions
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
-
-Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-<!-- _Acknowledgments_ -->
-
-<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
-
-# References
-
-::: {#refs}
-:::
-
-\newpage 
-# (APPENDIX) Appendix {.unnumbered}
-
-`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
-
-```{r multilevel-full, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
-#                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                              "Education (Ref.=Primary)"= 8:10,
-                  "Household Income (Ref.= Quintile 1)" = 11:15,
-                  "Age (Ref.= 18-29)" = 17:19,
-                  "Pol. pos (Ref.= Left)" = 21:23
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.60,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```

### 2024-12-16 04-analisis.qmd
change to quarto + add ola 7

--- 2024-03-26 paper.Rmd
@@
----
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
-css: "custom.css"
-linestretch: '1.5'
-link-citations: yes
-# params:
-# lang: en-GB
-# author: 
-# - name: Juan Carlos Castillo
-#   affiliation: Department of Sociology, Universidad de Chile
-#   email: juancastillov@uchile.cl
-#   number: 1, 3 
-# - name: Julio Iturra
-#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-#   number: 2
-# - name: Kevin Carrasco
-#   affiliation: Center for Social Conflict and Cohesion Studies
-#   number: 3
-abstract: | 
-  Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-  .
-  \newline
-  \newline
-
-   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-output:
-  bookdown::pdf_document2:
-    template: null
-    number_sections: true    
-    toc: false
-    keep_tex: false
-    pandoc_args:
-      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
-  bookdown::word_document2:
-    number_sections: false
-    # reference_docx: input/template.docx
-    toc: false
-  bookdown::html_document2:
-    number_sections: false
-linkcolor: blue
-bibliography:
-  - input/bib/merit-edjust.bib
-csl: input/bib/apa7-singlespace.csl
-editor_options:
-  chunk_output_type: console
-geometry: margin=1in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-always_allow_html: true # for word output
-fontsize: 12pt
-papersize: a4
----
-
-```{r eval=FALSE, include=FALSE}
- # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
- # clean #in the yml
- rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
-```
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-\pagebreak
-
-# Introduction
-
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
-
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
-Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
-
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021].
-A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
-# Inequality, beliefs and justification
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p.
-3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021].
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
-30].
-However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p.
-2].
-This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978].
-Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.
-
- Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986].
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
-Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-# The justification of educational inequality
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services.
-For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-_Justificacion, inequality and meritocracy in the educational domain_
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-_This study_
-
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-$H_1$: The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
-
-$H_2$: Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-$H_5$: The positive association between meritocracy and inequality justification mitigates along time
-
-#  Data, variables & methods
-
-## Data
-
-The main data source is the Chilean Longitudinal Social Survey 2016–2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
-The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.
-
-## Variables
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
-In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
-
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/merit_plot.png")
-```
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
-
-```{r descriptives, echo=FALSE, results='asis'}
-pacman::p_load(kableExtra)
-kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
-      format="pipe",
-             caption="Independent variables",
-             col.names = "",
-             row.names = F)
-```
-
-## Methods
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models [@singer_applied_2009].
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
-
-```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-  #RUN THIS BEFORE THE ANALYSIS
-  pacman::p_load(knitr)
-  table_format = if(is_html_output()) {
-    "html"
-  } else if(is_latex_output()) {
-    "latex"
-  }
-
-  table_format2 = if(is_html_output()) {
-    T
-  } else if(is_latex_output()) {
-    F
-  }
-```
-
 # Results
 
-```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
+```{r echo=FALSE, fig.align='center'}
+#| label: fig-alluvial
+#| fig-cap: "Change in the justification of educational inequality over time (2016-2022)"
+#| 
  knitr::include_graphics("output/graphs/alluvial_dep.png")
@@
 
-Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
+@fig-alluvial illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
 Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
@@
 
-```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
+```{r echo=FALSE, fig.align='center'}
+#| label: fig-means
+#| fig-cap: "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years"
+
  knitr::include_graphics("output/graphs/years_plot.png")
@@
 
-Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
+@fig-means shows the average changes in the main variables considered for this study.
 Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
@@
 
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
+```{r echo=FALSE, fig.align='center'}
+#| label: fig-correlation
+#| fig-cap: "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed"
+
  knitr::include_graphics("output/graphs/corr.png")
@@
 
-Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
+@fig-correlation presents a correlation matrix of the main variables analyzed, using data from all survey waves.
 In this matrix, the correlations vary between low and moderate values.
@@
 
-Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
+@tbl-multilevel shows the multilevel estimation results for the justification of educational inequality.
 Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
@@
 
-```{r multilevel, echo=FALSE, results='asis'}
+```{r echo=FALSE, results='asis'}
+#| label: tbl-multilevel
+#| tbl-cap: "Multilevel longitudinal models for the justification of inequality in education"
+
 pacman::p_load(dplyr,       # Manipulacion de datos
@@
 df_study1 <- 
-  df_study_long_t6_comunas %>%
+  df_study_long_t7_comunas %>%
   select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
@@
 omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
+texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                                       "Wave 2022",
+                                      "Wave 2023",
                                       "Wave",
@@
                                       "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Household Income (Ref.= Quintile 1)" = 8:12,
-                  "Age (Ref.= 18-29)" = 13:15,
-                  "Pol. pos (Ref.= Left)" = 17:19
+                groups = list("Wave (Ref.= Wave 2016)" = 2:6,
+                  "Household Income (Ref.= Quintile 1)" = 9:13,
+                  "Age (Ref.= 18-29)" = 14:16,
+                  "Pol. pos (Ref.= Left)" = 18:20
                   ),
@@
                                      "Var: Residual"),
-               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
                center = TRUE,
@@
 ```{r interact, echo=FALSE, results='asis'}
+#| label: tbl-interact
+#| fig-cap: "Interaction effects for the justification of economic inequality"
+
 df_study1 <- 
-  df_study_long_t6_comunas %>%
+  df_study_long_t7_comunas %>%
   select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
@@
 omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::texreg(list(educ6,educ3,educ1),
+texreg::knitreg(list(educ6,educ3,educ1),
                 custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                                      "Cov: Individuals Wave Merit: Effort"),  
-               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
                center = TRUE,
@@
 In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
+We test this hypothesis through interaction effects, displayed in @tbl-interact.
+Model 1 is shown as a baseline model, it is the same as Model 6 in @tbl-interact but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
 Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
 As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-# Discussion and conclusions
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
-
-Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-<!-- _Acknowledgments_ -->
-
-<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
-
-# References
-
-::: {#refs}
-:::
-
-\newpage 
-# (APPENDIX) Appendix {.unnumbered}
-
-`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
-
-```{r multilevel-full, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
-#                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                              "Education (Ref.=Primary)"= 8:10,
-                  "Household Income (Ref.= Quintile 1)" = 11:15,
-                  "Age (Ref.= 18-29)" = 17:19,
-                  "Pol. pos (Ref.= Left)" = 21:23
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.60,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```

### 2024-12-16 05-discussion.qmd
change to quarto + add ola 7

--- 2024-03-26 paper.Rmd
@@
----
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
-css: "custom.css"
-linestretch: '1.5'
-link-citations: yes
-# params:
-# lang: en-GB
-# author: 
-# - name: Juan Carlos Castillo
-#   affiliation: Department of Sociology, Universidad de Chile
-#   email: juancastillov@uchile.cl
-#   number: 1, 3 
-# - name: Julio Iturra
-#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-#   number: 2
-# - name: Kevin Carrasco
-#   affiliation: Center for Social Conflict and Cohesion Studies
-#   number: 3
-abstract: | 
-  Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-  .
-  \newline
-  \newline
-
-   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-output:
-  bookdown::pdf_document2:
-    template: null
-    number_sections: true    
-    toc: false
-    keep_tex: false
-    pandoc_args:
-      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
-  bookdown::word_document2:
-    number_sections: false
-    # reference_docx: input/template.docx
-    toc: false
-  bookdown::html_document2:
-    number_sections: false
-linkcolor: blue
-bibliography:
-  - input/bib/merit-edjust.bib
-csl: input/bib/apa7-singlespace.csl
-editor_options:
-  chunk_output_type: console
-geometry: margin=1in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-always_allow_html: true # for word output
-fontsize: 12pt
-papersize: a4
----
-
-```{r eval=FALSE, include=FALSE}
- # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
- # clean #in the yml
- rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
-```
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-\pagebreak
-
-# Introduction
-
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
-
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
-Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
-
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021].
-A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
-# Inequality, beliefs and justification
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p.
-3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021].
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
-30].
-However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p.
-2].
-This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978].
-Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.
-
- Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986].
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
-Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-# The justification of educational inequality
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services.
-For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-_Justificacion, inequality and meritocracy in the educational domain_
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-_This study_
-
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-$H_1$: The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
-
-$H_2$: Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-$H_5$: The positive association between meritocracy and inequality justification mitigates along time
-
-#  Data, variables & methods
-
-## Data
-
-The main data source is the Chilean Longitudinal Social Survey 2016–2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
-The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.
-
-## Variables
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
-In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
-
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/merit_plot.png")
-```
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
-
-```{r descriptives, echo=FALSE, results='asis'}
-pacman::p_load(kableExtra)
-kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
-      format="pipe",
-             caption="Independent variables",
-             col.names = "",
-             row.names = F)
-```
-
-## Methods
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models [@singer_applied_2009].
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
-
-```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-  #RUN THIS BEFORE THE ANALYSIS
-  pacman::p_load(knitr)
-  table_format = if(is_html_output()) {
-    "html"
-  } else if(is_latex_output()) {
-    "latex"
-  }
-
-  table_format2 = if(is_html_output()) {
-    T
-  } else if(is_latex_output()) {
-    F
-  }
-```
-
-# Results
-
-```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
- knitr::include_graphics("output/graphs/alluvial_dep.png")
-```
-
-Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
-
-```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
- knitr::include_graphics("output/graphs/years_plot.png")
-```
-
-Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
-
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/corr.png")
-```
-
-Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
-
-## Multivariate analysis
-
-Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
-
-```{r multilevel, echo=FALSE, results='asis'}
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-#                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Household Income (Ref.= Quintile 1)" = 8:12,
-                  "Age (Ref.= 18-29)" = 13:15,
-                  "Pol. pos (Ref.= Left)" = 17:19
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```
-
-```{r interact, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
-
-# Hipotesis
-
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
-
-# A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::texreg(list(educ6,educ3,educ1),
-                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
-                omit.coef = omit,
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Merit: Talent",
-                                      "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
-                                      ),
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),  
-               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = TRUE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE,
-               threeparttable = TRUE)
-```
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
 # Discussion and conclusions
@@
 Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-<!-- _Acknowledgments_ -->
-
-<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
-
-# References
-
-::: {#refs}
-:::
-
-\newpage 
-# (APPENDIX) Appendix {.unnumbered}
-
-`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
-
-```{r multilevel-full, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
-#                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                              "Education (Ref.=Primary)"= 8:10,
-                  "Household Income (Ref.= Quintile 1)" = 11:15,
-                  "Age (Ref.= 18-29)" = 17:19,
-                  "Pol. pos (Ref.= Left)" = 21:23
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.60,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```

### 2024-12-16 07-apendice.qmd
change to quarto + add ola 7

--- 2024-03-26 paper.Rmd
@@
----
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
-css: "custom.css"
-linestretch: '1.5'
-link-citations: yes
-# params:
-# lang: en-GB
-# author: 
-# - name: Juan Carlos Castillo
-#   affiliation: Department of Sociology, Universidad de Chile
-#   email: juancastillov@uchile.cl
-#   number: 1, 3 
-# - name: Julio Iturra
-#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-#   number: 2
-# - name: Kevin Carrasco
-#   affiliation: Center for Social Conflict and Cohesion Studies
-#   number: 3
-abstract: | 
-  Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
+# (APPENDIX) Appendix {.unnumbered}
 
-  .
-  \newline
-  \newline
+`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
 
-   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-output:
-  bookdown::pdf_document2:
-    template: null
-    number_sections: true    
-    toc: false
-    keep_tex: false
-    pandoc_args:
-      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
-  bookdown::word_document2:
-    number_sections: false
-    # reference_docx: input/template.docx
-    toc: false
-  bookdown::html_document2:
-    number_sections: false
-linkcolor: blue
-bibliography:
-  - input/bib/merit-edjust.bib
-csl: input/bib/apa7-singlespace.csl
-editor_options:
-  chunk_output_type: console
-geometry: margin=1in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-always_allow_html: true # for word output
-fontsize: 12pt
-papersize: a4
----
-
-```{r eval=FALSE, include=FALSE}
- # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
- # clean #in the yml
- rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
-```
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-\pagebreak
-
-# Introduction
-
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
-
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
-Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
-
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021].
-A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
-# Inequality, beliefs and justification
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p.
-3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021].
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
-30].
-However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p.
-2].
-This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978].
-Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.
-
- Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986].
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
-Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-# The justification of educational inequality
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services.
-For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-_Justificacion, inequality and meritocracy in the educational domain_
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-_This study_
-
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-$H_1$: The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
-
-$H_2$: Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-$H_5$: The positive association between meritocracy and inequality justification mitigates along time
-
-#  Data, variables & methods
-
-## Data
-
-The main data source is the Chilean Longitudinal Social Survey 2016–2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
-The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.
-
-## Variables
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
-In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
-
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/merit_plot.png")
-```
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
-
-```{r descriptives, echo=FALSE, results='asis'}
-pacman::p_load(kableExtra)
-kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
-      format="pipe",
-             caption="Independent variables",
-             col.names = "",
-             row.names = F)
-```
-
-## Methods
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models [@singer_applied_2009].
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
-
-```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-  #RUN THIS BEFORE THE ANALYSIS
-  pacman::p_load(knitr)
-  table_format = if(is_html_output()) {
-    "html"
-  } else if(is_latex_output()) {
-    "latex"
-  }
-
-  table_format2 = if(is_html_output()) {
-    T
-  } else if(is_latex_output()) {
-    F
-  }
-```
-
-# Results
-
-```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
- knitr::include_graphics("output/graphs/alluvial_dep.png")
-```
-
-Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
-
-```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
- knitr::include_graphics("output/graphs/years_plot.png")
-```
-
-Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
-
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/corr.png")
-```
-
-Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
-
-## Multivariate analysis
-
-Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
-
-```{r multilevel, echo=FALSE, results='asis'}
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
+```{r multilevel-full, echo=FALSE, results='asis'}
 df_study1 <- 
-  df_study_long_t6_comunas %>%
+  df_study_long_t7_comunas %>%
   select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
@@
 omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-#                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Household Income (Ref.= Quintile 1)" = 8:12,
-                  "Age (Ref.= 18-29)" = 13:15,
-                  "Pol. pos (Ref.= Left)" = 17:19
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```
-
-```{r interact, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
-
-# Hipotesis
-
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
-
-# A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::texreg(list(educ6,educ3,educ1),
-                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
-                omit.coef = omit,
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Merit: Talent",
-                                      "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
-                                      ),
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),  
-               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = TRUE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE,
-               threeparttable = TRUE)
-```
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-# Discussion and conclusions
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
-
-Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-<!-- _Acknowledgments_ -->
-
-<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
-
-# References
-
-::: {#refs}
-:::
-
-\newpage 
-# (APPENDIX) Appendix {.unnumbered}
-
-`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
-
-```{r multilevel-full, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
+texreg::htmlreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
@@
                                       "Wave 2022",
+                                      "wave 2023",
                                       "Wave",
@@
                                       "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                              "Education (Ref.=Primary)"= 8:10,
-                  "Household Income (Ref.= Quintile 1)" = 11:15,
-                  "Age (Ref.= 18-29)" = 17:19,
-                  "Pol. pos (Ref.= Left)" = 21:23
+                groups = list("Wave (Ref.= Wave 2016)" = 2:6,
+                              "Education (Ref.=Primary)"= 9:11,
+                  "Household Income (Ref.= Quintile 1)" = 12:16,
+                  "Age (Ref.= 18-29)" = 18:20,
+                  "Pol. pos (Ref.= Left)" = 22:24
                   ),

### 2024-12-16 paper.qmd
change to quarto + add ola 7

--- 2024-03-26 paper.Rmd
@@
 ---
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
-css: "custom.css"
-linestretch: '1.5'
-link-citations: yes
-# params:
-# lang: en-GB
-# author: 
-# - name: Juan Carlos Castillo
-#   affiliation: Department of Sociology, Universidad de Chile
-#   email: juancastillov@uchile.cl
-#   number: 1, 3 
-# - name: Julio Iturra
-#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
-#   number: 2
-# - name: Kevin Carrasco
-#   affiliation: Center for Social Conflict and Cohesion Studies
-#   number: 3
+title: Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.
+format:
+  docx:
+    number-sections: true
+  html: 
+    code-fold: true
+    html-math-method: katex
+    number-sections: true
+  pdf: 
+    geometry: 
+      - top=30mm
+      - left=20mm
+    template-partials: 
+      - title.tex
+    include-in-header:
+      text: |
+        \usepackage[noblocks]{authblk}
+        \renewcommand*{\Authsep}{, }
+        \renewcommand*{\Authand}{, }
+        \renewcommand*{\Authands}{, }
+        \renewcommand\Affilfont{\small}
+    number-sections: true
+editor: source
+bibliography: "input/bib/merit-edjust.bib"
+author:
+  - name: Juan Carlos Castillo
+    corresponding: true
+    orcid: 0000-0003-1265-7854
+    email: juancastillov@uchile.cl
+    affiliations:
+      - ref: 1
+      - ref: 2
+      - ref: 3
+  - name: Julio Iturra
+    orcid: 0000-0003-1993-3479
+    email: julioiturrasanhueza@gmail.com 
+    affiliations:
+      - ref: 4
+  - name: Kevin Carrasco
+    orcid: 0000-0002-9348-8361
+    email: kevin.carrasco@ug.uchile.cl
+    affiliations:
+      - ref: 2
+affiliations: 
+  - id: 1
+    name: Universidad de Chile
+  - id: 2
+    name: Centro de estudios del conflicto y cohesión social (COES)
+  - id: 3
+    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
+  - id: 4
+    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
+citeproc: true
 abstract: | 
@@
 
-  .
   \newline
@@
    **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-output:
-  bookdown::pdf_document2:
-    template: null
-    number_sections: true    
-    toc: false
-    keep_tex: false
-    pandoc_args:
-      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
-  bookdown::word_document2:
-    number_sections: false
-    # reference_docx: input/template.docx
-    toc: false
-  bookdown::html_document2:
-    number_sections: false
-linkcolor: blue
-bibliography:
-  - input/bib/merit-edjust.bib
-csl: input/bib/apa7-singlespace.csl
-editor_options:
-  chunk_output_type: console
-geometry: margin=1in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
-always_allow_html: true # for word output
-fontsize: 12pt
-papersize: a4
 ---
-
-```{r eval=FALSE, include=FALSE}
- # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
- # clean #in the yml
- rmarkdown::render("paper.Rmd", output_format = "bookdown::pdf_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::html_document2")
- rmarkdown::render("paper.Rmd", output_format = "bookdown::word_document2")
-```
 
@@
 
-\pagebreak
+This document was last modified at {{< modified >}}
 
-# Introduction
+and it was last rendered at {{< now >}}
 
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
+{{< include 01-introduction.qmd >}}
 
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
+{{< include 02-antecedentes.qmd >}}
 
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
-Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
+{{< include 03-methods.qmd >}}
 
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021].
-A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
+{{< include 04-analisis.qmd >}}
 
-# Inequality, beliefs and justification
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as “the defensibility of unequal relations between people” [-@barry_theories_1989, p.
-3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021].
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
-30].
-However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p.
-2].
-This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978].
-Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.
-
- Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986].
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
-Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-# The justification of educational inequality
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services.
-For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-_Justificacion, inequality and meritocracy in the educational domain_
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-_This study_
-
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-$H_1$: The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
-
-$H_2$: Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-$H_5$: The positive association between meritocracy and inequality justification mitigates along time
-
-#  Data, variables & methods
-
-## Data
-
-The main data source is the Chilean Longitudinal Social Survey 2016–2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/.
-The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.
-
-## Variables
-
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
-In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large").
-Figure \@ref(fig:frecuencies) shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).
-
-```{r frecuencies, echo=FALSE, fig.cap = "Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/merit_plot.png")
-```
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \@ref(tab:descriptives) shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
-
-```{r descriptives, echo=FALSE, results='asis'}
-pacman::p_load(kableExtra)
-kable(data.frame(image='![](output/tables/desc02.png)'),    # generar tabla
-      format="pipe",
-             caption="Independent variables",
-             col.names = "",
-             row.names = F)
-```
-
-## Methods
-
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models [@singer_applied_2009].
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library "lme4" [@singer_applied_2009, p.4].
-
-```{r message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-  #RUN THIS BEFORE THE ANALYSIS
-  pacman::p_load(knitr)
-  table_format = if(is_html_output()) {
-    "html"
-  } else if(is_latex_output()) {
-    "latex"
-  }
-
-  table_format2 = if(is_html_output()) {
-    T
-  } else if(is_latex_output()) {
-    F
-  }
-```
-
-# Results
-
-```{r alluvial, echo=FALSE, fig.cap = "Change in the justification of educational inequality over time (2016-2022)", fig.align='center'}
- knitr::include_graphics("output/graphs/alluvial_dep.png")
-```
-
-Figure \@ref(fig:alluvial) illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
-
-```{r means, echo=FALSE, fig.cap = "Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years", fig.align='center'}
- knitr::include_graphics("output/graphs/years_plot.png")
-```
-
-Figure \@ref(fig:means) shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
-
-```{r correlation, echo=FALSE, fig.cap = "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed", fig.align='center'}
- knitr::include_graphics("output/graphs/corr.png")
-```
-
-Figure \@ref(fig:correlation) presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
-
-## Multivariate analysis
-
-Table \@ref(tab:multilevel) shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
-
-```{r multilevel, echo=FALSE, results='asis'}
-pacman::p_load(dplyr,       # Manipulacion de datos
-               knitr,
-               kableExtra,
-               summarytools,
-               lme4,
-               texreg
-               )
-load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-
-# generate analytical sample
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
-                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-#                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                  "Household Income (Ref.= Quintile 1)" = 8:12,
-                  "Age (Ref.= 18-29)" = 13:15,
-                  "Pol. pos (Ref.= Left)" = 17:19
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel)","Multilevel longitudinal models for the justification of inequality in education"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```
-
-```{r interact, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
-
-# Hipotesis
-
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
-
-# A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
-texreg::texreg(list(educ6,educ3,educ1),
-                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
-                omit.coef = omit,
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.coef.names = c("Intercept",
-                                      "Wave",
-                                      "Wave^2",
-                                      "Merit: Talent",
-                                      "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
-                                      ),
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),  
-               caption = paste("(\\#tab:interact)","Interaction effects for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.765,
-               float.pos = "!ht", 
-               booktabs = TRUE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE,
-               threeparttable = TRUE)
-```
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \@ref(tab:interact).
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
-
-# Discussion and conclusions
-
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
-
-Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-<!-- _Acknowledgments_ -->
-
-<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
+{{< include 05-discussion.qmd >}}
 
@@
 
-\newpage 
-# (APPENDIX) Appendix {.unnumbered}
-
-`r if (knitr::is_latex_output()){ '\\appendix'}` `r if (knitr::is_latex_output()){ '\\section{Appendix}'}`
-
-```{r multilevel-full, echo=FALSE, results='asis'}
-df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
-         )
-# Hipotesis
-h0 <- "ola"
-h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-
-# A. Education distributive justice
-educ.null <- lmer(formula(paste0("just_educ~","1 + (1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ0 <- lmer(formula(paste0("just_educ~",h0,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ1 <- lmer(formula(paste0("just_educ~",h1,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ2 <- lmer(formula(paste0("just_educ~",h2,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-
-# Table 1: individual
-
-omit <- "(education)|(ess)"
-texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
-                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
-#                omit.coef = omit,
-                custom.coef.names = c("Intercept",
-                                      "Wave 2017",
-                                      "Wave 2018",
-                                      "Wave 2019",
-                                      "Wave 2022",
-                                      "Wave",
-                                      "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
-                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort" ),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:5,
-                              "Education (Ref.=Primary)"= 8:10,
-                  "Household Income (Ref.= Quintile 1)" = 11:15,
-                  "Age (Ref.= 18-29)" = 17:19,
-                  "Pol. pos (Ref.= Left)" = 21:23
-                  ),
-                include.loglik = FALSE,
-                include.aic = FALSE,
-                custom.gof.names = c("BIC",
-                                     "Num. obs.",
-                                     "Num. groups: Individuals",
-                                     "Var: Individuals (Intercept)",
-                                     "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
-               center = TRUE,
-               caption.above = TRUE,
-               scalebox = 0.60,
-               float.pos = "!ht", 
-               booktabs = FALSE, 
-               use.packages = FALSE, 
-               include.rs = FALSE, 
-               include.rmse = FALSE,
-               include.ci = FALSE)
-
-```
+{{< include 07-apendice.qmd >}}

### 2024-12-16 paper.tex
change to quarto + add ola 7

--- 2024-03-26 paper.tex
@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
 \documentclass[
-  12pt,
-  a4paper,
 ]{article}
+
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{setspace}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
 \fi
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic},
-%   colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{longtable,booktabs,array}
+\usepackage[top=30mm,left=20mm]{geometry}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
+% Make \paragraph and \subparagraph free-standing
+\makeatletter
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\makeatother
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage{times}
-\hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}
-\usepackage{caption}
-\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-\usepackage{graphicx}
-\usepackage{float}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
   % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\usepackage[noblocks]{authblk}
+\renewcommand*{\Authsep}{, }
+\renewcommand*{\Authand}{, }
+\renewcommand*{\Authands}{, }
+\renewcommand\Affilfont{\small}
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
 \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic}
-    \usepackage[]{authblk}
-    \setlength{\affilsep}{2em}
-    \renewcommand*{\Affilfont}{\normalsize\normalfont}
-    \renewcommand\Authsep{, }
-    \renewcommand\Authand{ \& }
-    \renewcommand\Authands{, \& }
-                                        \author[1, 3]{Juan Carlos Castillo}
-                                                            \affil[]{Department of Sociology, Universidad de Chile \thanks{\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}}}
-                                                                                \author[2]{Julio Iturra}
-                                                            \affil[]{Bremen International Graduate School of Social Sciences, Universität Bremen}
-                                                                                \author[3]{Kevin Carrasco}
-                                                            \affil[]{Center for Social Conflict and Cohesion Studies}
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage{bookmark}
+
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.},
+  pdfauthor={Juan Carlos Castillo; Julio Iturra; Kevin Carrasco},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\title{Changes in the justification of educational inequalities. The
+role of perceptions of inequality and meritocracy during the COVID
+pandemic.}
+
+  \author{Juan Carlos Castillo}
+            \affil{%
+                  Universidad de Chile
+              }
+          \affil{%
+                  Centro de estudios del conflicto y cohesión social
+                  (COES)
+              }
+          \affil{%
+                  Núcleo milenio de desigualdades y oportunidades
+                  digitales (NUDOS)
+              }
+        \author{Julio Iturra}
+            \affil{%
+                  International Graduate School of Social Sciencies
+                  (BIGSSS), University of Bremen, Germany
+              }
+        \author{Kevin Carrasco}
+            \affil{%
+                  Centro de estudios del conflicto y cohesión social
+                  (COES)
+              }
+
 \date{}
-
-\providecommand{\keywords}[1]
-{
-  \small
-  \textbf{Keywords:} #1
-}
-
 \begin{document}
@@
 \begin{abstract}
-Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
-And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-.
+Education is considered a key tool for social mobility and equality of
+opportunities. However, despite the widespread value of education,
+disparities in educational outcomes persist over time. To what extent
+are such educational disparities justified in society? And, what are the
+main factors driving this kind of justification? This research focuses
+on the changes in the justification of inequality in access to education
+in Chile and their links with the perception of meritocracy. The Chilean
+case offers an interesting context for this study given its high
+economic inequality, deep neoliberal policies, and commodification of
+social services such as education. The central argument of this article
+is that the justification of inequalities weakens during periods of
+vulnerability and crisis (such as the health and economic crisis
+resulting from the COVID-19 pandemic), which could be linked to a
+challenge of meritocratic ideals. For testing the research hypotheses we
+estimate a series of longitudinal multilevel models with data from a
+Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
+Whereas we find support for the association between the perception of
+meritocracy and justification of inequality, the analyses that involved
+changes over time showed an increase in the justification of inequality
+in education. The discussion of these results delves into the social
+consequences of justifying inequality in a sensitive area as education,
+as well as the persistence of meritocratic ideals despite challenging
+events.
+
 \newline
@@
 
-\textbf{Keywords}: meritocracy, social inequality, inequality justification, COVID-19
+\textbf{Keywords}: meritocracy, social inequality, inequality
+justification, COVID-19
 \end{abstract}
 
-\setstretch{1.5}
-\pagebreak
+This document was last modified at 2024-12-15 23:10:25
+
+and it was last rendered at 2024-12-15 23:10:25
 
@@
 
-Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, ``Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries'' (\citeproc{ref-oecd_pisa_2023}{OECD, 2023}).
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 (\citeproc{ref-darmodyImpactsCOVID19Control2021}{Darmody et al., 2021}).
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education (\citeproc{ref-leeFairnessPerceptionsEducational2023}{Lee \& Stacey, 2023}).
-In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed (\citeproc{ref-hadjar_meritokratie_2008}{Hadjar, 2008}; \citeproc{ref-mcnamee_meritocracy_2004}{McNamee \& Miller, 2004}; \citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}).
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
-
-The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system (\citeproc{ref-bellei_estudio_2013}{Bellei, 2013}; \citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}).
-The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation (\citeproc{ref-joikomujicaCuasimercadoEducativoChile2019}{Joiko Mujica, 2019}).
-Previous research in Chile has emphasized the concept of an ``educational market'' (\citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}), characterized by selection, competition, and the provision of goods and services associated with payment capacities.
-Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences (\citeproc{ref-ramosarellanoEducacionPromesaIncumplida2022}{Ramos Arellano, 2022}).
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
-
-This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}; \citeproc{ref-hankivsky_introduction_2022}{Hankivsky et al., 2022}).
-A pertinent aspect of this discourse is articulated in the works of Mijs (\citeproc{ref-mijs_stratified_2016}{2016}, \citeproc{ref-mijs_visualizing_2018}{2018}), showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
-Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
-
-\section{Inequality, beliefs and justification}\label{inequality-beliefs-and-justification}
-
-The justification of economic inequalities is a controversial field of study.
-As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as ``the defensibility of unequal relations between people'' (\citeproc{ref-barry_theories_1989}{1989, p.
-3}), whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification (\citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}; \citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-trump_income_2018}{Trump, 2018}).
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
-
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research (\citeproc{ref-Resh2016}{Resh \& Sabbagh, 2016}).
-From a sociological point of view, the works of Kluegel \& Smith (\citeproc{ref-kluegelBeliefsStratification1981}{1981}) can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of ``beliefs'' in a ``broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action.'' (\citeproc{ref-kluegelBeliefsStratification1981}{Kluegel \& Smith, 1981, p.
-30}).
-However, the literature has strained the concept of ``beliefs'' by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, Wegener (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{1987}) has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987, p.
-2}).
-This approach established that status positions influence how people perceive things \emph{are} and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust (\citeproc{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}).
-Additionally, Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}) has suggested more detailed conceptual distinctions about how people perceive economic inequality \emph{is}, how they believe or prefer it \emph{should be}, and the process of evaluating such differences as fair.
-
-Theoretically, Liebig \& Sauer (\citeproc{ref-liebigSociologyJustice2016}{2016}) establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, \emph{order-related attitudes} refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, \emph{procedural justice attitudes} account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be \emph{outcome-related attitudes} which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles (\citeproc{ref-kluegel_social_1995}{Kluegel et al., 1995}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011})
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{Janmaat, 2013}; \citeproc{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \& Smith, 1986}).
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries (\citeproc{ref-Evans1992}{Evans et al., 1992}; \citeproc{ref-Evans2017}{Evans \& Kelley, 2017}) and greater justification of salary gaps (\citeproc{ref-hadler_why_2005}{Hadler, 2005}; \citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}).
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}; \citeproc{ref-vandertoorn_twenty_2014}{van der Toorn \& Jost, 2014}).
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies (\citeproc{ref-trump_when_2020}{Trump, 2020}).
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position (\citeproc{ref-jost_theory_2020}{Jost, 2020}).
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-kelleyLegitimationInequalityOccupational1993}{Kelley \& Evans, 1993}; \citeproc{ref-osberg_fair_2006}{Osberg \& Smeeding, 2006}).
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987}, \citeproc{ref-Wegener1990}{1990}).
-Thus, both the role of social status and economic inequality perceptions have found support in correlational (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012}; \citeproc{ref-schroder_income_2017}{Schröder, 2017}), as well as in experimental studies that have manipulated inequality perception (\citeproc{ref-iturra_percepcion_2023}{Iturra et al., 2023}; \citeproc{ref-trump_income_2018}{Trump, 2018}).
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification (\citeproc{ref-Castillo2011}{Castillo, 2011}).
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences (\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De Carvalho Galvão, 2022}).
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-\section{The justification of educational inequality}\label{the-justification-of-educational-inequality}
-
-Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
-A strand of this research has focused on specific policy areas such as health, pensions, and education, attending to the degree of justification of market-based access to social services.
-For instance, Lindh (\citeproc{ref-lindhPublicOpinionMarkets2015}{2015}) found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
-Similarly, Busemeyer \& Iversen (\citeproc{ref-busemeyer_welfare_2020}{2020}) shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
-Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by Conforti (\citeproc{ref-confortiLegitimationInequalityAmerican1992}{1992}) as ``while equality is embraced as a value, inequality is accepted as a reality'' (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that people with higher incomes can buy better education for their children than people with lower incomes?}), and the European Social Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I have had a fair chance of achieving the level of education I was seeking}).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about---and more supportive of proposals to close---wealth-based achievement gaps than Black-White or Hispanic-White gaps (\citeproc{ref-valantPoliticsAchievementGaps2016}{Valant \& Newark, 2016}).
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification (\citeproc{ref-stoilovaFairnessEducationalOpportunities2023}{Stoilova \& Ilieva-Trichkova, 2023}).
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}), justifications are usually confused under the labels of ``beliefs'' and ``attitudes,'' and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011}).
-For instance, a recent study by Day \& Norton (\citeproc{ref-dayPerceivedIdealInequality2023}{2023}) on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman, 2018})
-
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by Igliozzi et al.
-(\citeproc{ref-igliozziFairShareEffects2024}{2024}), they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}) is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance (`Students who place within the highest 50\% of testing scores from the previous year can attend the magnet school').
-In the same line, Lee \& Stacey (\citeproc{ref-leeFairnessPerceptionsEducational2023}{2023}) show that performance assessments (in what they call ``neoliberal orientations'') predicted people's fair perceptions of educational inequality.
-
-\emph{Justificacion, inequality and meritocracy in the educational domain}
-
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work ``The Rise of the Meritocracy'' (\citeproc{ref-young_rise_1962}{1962}), has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities (\citeproc{ref-bell_politics_2020}{Bell, 2020}).
-This would contribute to the justification of inequality in the educational domain (\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}), as individual ability is often inherited, and educational attainment is influenced by family background (\citeproc{ref-goldin_meritocracy_2000}{Goldin, 2000}).
-
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences (\citeproc{ref-McCoy2007}{McCoy \& Major, 2007}).
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}).
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities (\citeproc{ref-dayPerceivedIdealInequality2023}{Day \& Norton, 2023}).
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
+Despite the widespread recognition of education as a key factor for
+social mobility and equality of opportunities, disparities in this
+domain are far from being eliminated. As the last PISA Report points
+out, ``Socio-economically disadvantaged students are seven times more
+likely than advantaged students to score below Level 2 in mathematics on
+average across OECD countries'' (OECD 2023). Furthermore, empirical
+evidence from social science research has revealed an exacerbation of
+educational inequalities across various domains, such as family
+socioeconomic status, gender, ethnicity, and migratory background,
+particularly after COVID-19 (Darmody, Smyth, and Russell 2021).
+Nevertheless, and despite the consistent empirical evidence on
+educational inequalities and their consequences, research on
+distributive justice indicates remarkable individual differences in the
+justification of unequal access to education (Lee and Stacey 2023). In
+the present paper, we argue that the perception of meritocracy in
+society might play a role in the justification of economic inequalities
+in this particular domain.
+
+Meritocracy is defined as a distributive system where individual factors
+such as effort and talent are conceived as the main determinants of
+individual performance, usually sidelining the role of opportunities and
+social background. In recent years, there has been a growing interest in
+studying meritocratic beliefs as well as their variations according to
+individual and contextual variables (Mijs 2021; Mijs et al. 2022). This
+line of research has highlighted how meritocratic beliefs are associated
+with justifying unequal access to resources and rewards in society, by
+legitimizing social hierarchies in which those who already have
+resources and advantages are more likely to succeed (Sandel 2020;
+McNamee and Miller 2004; Hadjar 2008). Within this framework, this study
+focuses on the changes in the justification of inequalities in access to
+education in Chile, and to what extent this phenomenon is associated
+with the perception of meritocracy and its variations during the COVID
+pandemic.
+
+The Chilean case is particularly relevant when it comes to the study of
+inequalities in education. The country is characterized by a high level
+of economic inequality associated with the privatization and
+commodification of several social policy domains, among them the
+educational system (Bellei 2013; Corvalán, Carrasco, and García-Huidobro
+2016). The neoliberal reforms implemented during and since the military
+dictatorship (1973-1989) have deepened educational policies based on
+private subsidies, achievement incentives, selection, and segregation
+(Joiko 2019). Previous research in Chile has emphasized the concept of
+an ``educational market'' (Corvalán, Carrasco, and García-Huidobro
+2016), characterized by selection, competition, and the provision of
+goods and services associated with payment capacities. Many studies have
+addressed this problem from economic, political, and sociological
+perspectives, generating relevant evidence on their academic and social
+consequences (Ramos 2022). Likewise, massive social mobilizations in the
+country (particularly in 2006, 2011, and 2018) have had educational
+inequalities as one of the main targets, putting pressure on governments
+towards a series of reforms that have advanced free higher education
+reforms and have diminished the role of private actors in the
+educational system.
+
+This research endeavors to dissect the nuanced alterations in the
+justifications for educational disparities within the Chilean framework.
+Central to our investigation is the premise that the rationale
+underpinning these inequalities experiences a significant recalibration
+during periods of vulnerability and crises, specifically under the
+unprecedented strains of the global health and economic cataclysm
+triggered by the COVID-19 pandemic (Hankivsky, Morrow, and Varcoe 2022;
+Breznau 2021). A pertinent aspect of this discourse is articulated in
+the works of Mijs (2016, 2018), showing the interplay between
+meritocratic beliefs and the legitimization of disparities, as well as
+the potential erosion of meritocratic principles amidst crises. Within
+this framework, we point out that periods of vulnerability (such as the
+health and economic crisis resulting from the COVID-19 pandemic) could
+weaken the justification of social inequalities, a change that would be
+at least in part linked to a challenge of the meritocratic ideals.
+Furthermore, we suggest that the association between meritocratic
+perception and inequality justification would be weaker among
+lower-status groups, as they are the most affected in critical times.
+
+\section{Inequality, beliefs and
+justification}\label{inequality-beliefs-and-justification}
+
+The justification of economic inequalities is a controversial field of
+study. As equality is usually considered a synonym of justice in common
+sense, the idea that someone could be willing to justify inequalities
+seems (in principle) far from what is considerably reasonable.
+Nevertheless, and as pointed out by Barry, the whole normative
+discussion in social justice can be conceived as ``the defensibility of
+unequal relations between people'' (1989, 3), whereby empirical research
+has consistently shown strong variations between individuals and
+societies regarding inequality justification (Trump 2018; Kelley and
+Evans 2021; Mijs 2021). Both theoretical and empirical approaches have
+been able to establish a clear difference between inequality and
+justice, which opens the question of to what extent inequality can be
+tolerated or even justified.
+
+The field of studies of subjective economic inequality has been
+extensively developed in the social sciences, particularly in the area
+of social justice research (Resh and Sabbagh 2016). From a sociological
+point of view, the works of Kluegel and Smith (1981) can be considered
+pioneers in framing a research area comprising perceptions, beliefs, and
+justifications in topics such as inequality of opportunities, poverty,
+and distributive justice. Theoretically, the authors understand the
+concept of ``beliefs'' in a ``broad sense to refer to the information
+(veridical or non-veridical) about a phenomenon that an individual uses
+as a basis both for inferring other information and for action.''
+(Kluegel and Smith 1981, 30). However, the literature has strained the
+concept of ``beliefs'' by more specific distinctions around the
+individual and structural dimensions of economic inequality. In this
+regard, Wegener (1987) has suggested that perceptions of social status
+and reward hierarchy are relative to the observer's status in such
+hierarchies (Wegener 1987, 2). This approach established that status
+positions influence how people perceive things \emph{are} and, in turn,
+these perceptions anchor how people evaluate this inequality as just or
+unjust (Jasso 1978). Additionally, Janmaat (2013) has suggested more
+detailed conceptual distinctions about how people perceive economic
+inequality \emph{is}, how they believe or prefer it \emph{should be},
+and the process of evaluating such differences as fair.
+
+Theoretically, Liebig and Sauer (2016) establish three fundamental
+distinctions in justice attitudes in the field of sociological justice
+research. First, \emph{order-related attitudes} refer to individual
+motives that explain preferences for resource allocation based on the
+principles of equality, need, or entitlement. Second, \emph{procedural
+justice attitudes} account for the degree of acceptance of unfavorable
+outcomes under conditions in which the procedure under which those
+decisions were made is considered fair (e.g., opportunities). Finally,
+they can be \emph{outcome-related attitudes} which refer primarily to
+the economic resources that a person currently obtains and the set of
+collective commitments associated with possessing those resources (e.g.,
+transfers or taxes). Within this framework, we conceptualize the
+justification of inequality in the first area of order-related
+attitudes, as it expresses preferences for resource allocation that can
+be associated with concepts that reflect different distributive
+principles (Kluegel, Mason, and Wegener 1995; Son Hing et al. 2011)
+
+An important finding in the area of order-related attitudes is that
+beliefs about inequality are stratified by social status (Janmaat 2013;
+Kluegel and Smith 1986). The evidence suggests that advantaged positions
+in terms of socioeconomic status, such as income, education, or
+occupational prestige, are associated with preferences for less
+egalitarian class imageries (Evans, Kelley, and Kolosi 1992; Evans and
+Kelley 2017) and greater justification of salary gaps (Hadler 2005;
+Kelley and Evans 2021). Nevertheless, and in contrast to this
+perspective, it has been argued that human beings are intrinsically
+motivated to justify the system they inhabit, even when the information
+available to them points in the opposite direction (Jost and Hunyady
+2003; van der Toorn and Jost 2014). Individuals would be generally
+disposed to justify the economic status quo regardless of self- and
+group economic interest because their support partially depends on their
+beliefs in system-justifying ideologies (Trump 2020). As a result, what
+has been suggested is that low-status might perceive the status quo as
+just as they are more strongly attached to dominant ideologies, while
+high-status individuals are more prone to justify the system to a
+greater extent because it favors their interests and dominant position
+(Jost 2020).
+
+Another factor that has been linked to inequality beliefs is the
+perception of economic inequality (Kelley and Evans 1993; Osberg and
+Smeeding 2006; Castillo, García-Castro, and Venegas 2022). Studies in
+this matter have argued that perceived economic inequality functions as
+an anchor for the justifications of inequality because if these
+inequalities are indeed seen as fair in their generative processes, then
+they will be considered as just (Wegener 1987, 1990). Thus, both the
+role of social status and economic inequality perceptions have found
+support in correlational (Castillo 2012; Schröder 2017), as well as in
+experimental studies that have manipulated inequality perception (Iturra
+et al. 2023; Trump 2018). In addition, normative or value-driven
+mechanisms can also explain to what extent people justify inequality.
+For instance, studies suggest that justice ideologies such as
+individualism or egalitarianism can mediate the influence of individual
+status on income inequality justification (Castillo 2011). Likewise,
+system justification ideologies can strengthen the influence of
+socioeconomic status on economic inequality preferences (García-Sánchez
+and De Carvalho 2022). Altogether, these studies have shown that not
+only socio-structural factors influence what people believe regarding
+the distribution of economic resources, but their normative stances and
+perceptions can also determine preferences for economic inequality in
+diverse domains.
+
+\section{The justification of educational
+inequality}\label{the-justification-of-educational-inequality}
+
+Studies addressing inequality justification have assessed the support
+for market ideologies, salary gaps, and distributive preferences. A
+strand of this research has focused on specific policy areas such as
+health, pensions, and education, attending to the degree of
+justification of market-based access to social services. For instance,
+Lindh (2015) found that public support for the market distribution of
+social services is generally low in most countries, implying that most
+people believe it is unfair for market forces to determine access to
+basic social services. Similarly, Busemeyer and Iversen (2020) shows
+that high-income citizens are less supportive of expanding public
+spending on universal benefit schemes. Instead, they tend to prefer a
+public basic insurance scheme, even if it provides relatively more
+benefits to low-income individuals.
+
+The study of inequality justification in the sphere of education is a
+rather specific and unattended field. Most studies on educational
+inequality relate performance indicators (such as grades and
+standardized test results) to stratification-related variables, whereby
+few refer specifically to if and how individuals tolerate and even
+justify educational inequalities. This is probably because in principle
+it seems counter-intuitive to think that someone would be willing to
+justify differences in such a sensitive area. This paradox is phrased by
+Conforti (1992) as ``while equality is embraced as a value, inequality
+is accepted as a reality'' (p.230).
+
+In the area of empirical studies, the operationalization of educational
+inequality justification in terms of questionnaire items is diverse, and
+there is not a single instrument or scale dedicated specifically to this
+concept. Rather, some attitudinal surveys usually include
+education-related items that are posteriorly used as dependent or
+independent variables. Some examples are the International Social Survey
+Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that
+people with higher incomes can buy better education for their children
+than people with lower incomes?}), and the European Social Survey (ESS)
+2018 (\emph{Compared to other people in {[}country{]}, I have had a fair
+chance of achieving the level of education I was seeking}). Although
+both questions have been used to measure the justification of
+educational inequality, they refer to different aspects. The ESS can be
+classified as reflexive, as the respondent is involved as a referent for
+the question, whereas the ISSP question is non-reflexive. Furthermore,
+ISSP refers to educational access, while the ESS refers to educational
+outputs.
+
+Studies have addressed the justification of educational inequality from
+different conceptual perspectives and methodological strategies. Some
+perspectives emphasize the justification of differences between specific
+groups, finding for instance that Americans are more concerned
+about---and more supportive of proposals to close---wealth-based
+achievement gaps than Black-White or Hispanic-White gaps (Valant and
+Newark 2016). Other studies focus on the association between status
+differences and inequality justification in education, based on the
+rational premise that those who are worst-off would tend to show lower
+levels of justification (Stoilova and Ilieva-Trichkova 2023). However,
+most of the research in this area goes beyond status differences,
+focusing on how subjective factors such as perceptions and beliefs are
+related to different levels of justification.
+
+The perception of economic inequality is one of the factors associated
+with the justification of educational inequality. As pointed out by
+Janmaat (2013), justifications are usually confused under the labels of
+``beliefs'' and ``attitudes,'' and their distinction from perceptions is
+relevant as it helps state that justifying a certain level of inequality
+requires taking into account which level of (perceived) inequality is
+first and foremost being recognized (Son Hing et al. 2011; Castillo et
+al. 2023). For instance, a recent study by Day and Norton (2023) on the
+perception and justification of inequality in university endowments in
+the U.S. points out that individuals tend to underestimate the magnitude
+of existing inequality while simultaneously desiring greater equality.
+Furthermore, through an experiment, they demonstrate that information
+about inequality in university endowments (i.e., manipulating inequality
+perception) increases the perception of injustice in education. Results
+of this study are in line with the consistent evidence on biases related
+to the underestimation of economic inequality (Castillo, García-Castro,
+and Venegas 2022; Gimpelson and Treisman 2018)
+
+Besides focusing on the relationship between perception and
+justification of educational inequality, another group of studies has
+addressed how the justification of inequality is associated with the
+support of different distributive principles. In an experiment by
+Igliozzi, Granot, and Ottati (2024), they establish how the support for
+different justice principles (equity, equality, and need) is related to
+support for redistributive policies in domains such as education. They
+find that a larger system justification (Jost and Hunyady 2003) is
+related to higher support for the principles of equity and equality
+(over need) in distributing educational outcomes, where equity was
+measured in terms of access to education related to performance
+(`Students who place within the highest 50\% of testing scores from the
+previous year can attend the magnet school'). In the same line, Lee and
+Stacey (2023) show that performance assessments (in what they call
+``neoliberal orientations'') predicted people's fair perceptions of
+educational inequality.
+
+\emph{Justificacion, inequality and meritocracy in the educational
+domain}
+
+The relationship between performance and outputs brings us to the
+concepts of merit and meritocracy, key in educational justice research.
+The concept of meritocracy, coined by the sociologist Michael Young in
+his satirical work ``The Rise of the Meritocracy'' (1962), has a rich
+history and multifaceted meaning. Young's term was initially used to
+critique a hypothetical future society where social and economic
+positions were entirely determined by individuals' merits, talents, and
+efforts, ostensibly eliminating the influence of inherited privilege.
+However, over time, the term has evolved to describe a social system
+where advancement and success are perceived as directly linked to one's
+abilities and achievements, sidelining inequalities of opportunities
+(Bell 2020). This would contribute to the justification of inequality in
+the educational domain (Sandel 2020), as individual ability is often
+inherited, and educational attainment is influenced by family background
+(Goldin 2000).
+
+Empirical research examining the relationship between meritocratic
+perceptions and the justification of educational inequality provides
+valuable insights into the dynamics shaping individuals' attitudes and
+beliefs. Some studies have explored the extent to which people believe
+that educational systems are meritocratic and how these perceptions
+influence their acceptance or rejection of educational disparities, as
+well as status differences (McCoy and Major 2007). This belief is
+particularly influential in educational institutions, where it not only
+encourages academic success but also legitimizes social and income
+inequality (Batruch et al. 2023). Most of the evidence so far suggests
+that those who believe in a direct link between effort, talent, and
+educational success are more inclined to rationalize existing
+disparities in educational outcomes. However, empirical findings also
+reveal nuances in this relationship. For instance, studies show that
+when individuals perceive structural barriers, such as economic
+disparities or discrimination, as hindering the meritocratic ideal, they
+may be less likely to justify educational inequalities (Day and Norton
+2023). This suggests that while meritocratic beliefs play a significant
+role, individuals' awareness of broader societal issues can mediate
+their willingness to rationalize educational disparities.
 
@@
 
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}).
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
-Consistently, our first hypothesis is:
-
-\(H_1\): The justification of educational inequalities decreases after the pandemic
-
-Our second argument is based on the evidence that relates rational interests with justifications and preferences (\citeproc{ref-hing_failure_2019}{Hing et al., 2019}), expecting that:
-
-\(H_2\): Those with lower status characteristics show lower justification of educational inequalities
-
-Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
-As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
-
-\(H_3\): A larger inequality perception is negatively associated with the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
-
-\(H_4\): Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
-As a consequence, the justification of inequality could show a decrease over time:
-
-\(H_5\): The positive association between meritocracy and inequality justification mitigates along time
+The present research uses longitudinal survey data from 2016 to 2022 to
+asses changes in inequality justification. As there are measures before
+and after the COVID pandemic, firstly we explore to what extent it is
+possible to identify variations in inequality justification in this time
+frame. Although we are aware that it is not possible to make a causal
+claim given that this period is characterized by many other
+circumstances besides COVID, we argue that the increase of risk
+perceptions in critical situations (such as the pandemic) impacts public
+concerns about equality and redistribution (Breznau 2021). Although the
+pandemic might have triggered such concerns mostly regarding health, we
+believe that other policy areas, such as education, could have been
+impacted in this period. Consistently, our first hypothesis is:
+
+\(H_1\): The justification of educational inequalities decreases after
+the pandemic
+
+Our second argument is based on the evidence that relates rational
+interests with justifications and preferences (Son Hing et al. 2019),
+expecting that:
+
+\(H_2\): Those with lower status characteristics show lower
+justification of educational inequalities
+
+Thirdly, we address inequality perception. As referred to above, the
+justification of inequality requires, first and foremost, that such
+inequality is recognized. As a larger inequality in this domain
+threatens a basic equality ideal, the corresponding hypothesis in this
+case is:
+
+\(H_3\): A larger inequality perception is negatively associated with
+the justification of educational inequalities.
+
+Our fourth hypothesis refers to the central concept of meritocracy. As
+meritocracy justifies inequalities based on performance, we expect that:
+
+\(H_4\): Higher meritocratic perceptions are positively related to the
+justification of educational inequalities.
+
+Finally, and in line with hypotheses 1 and 4, we explore interactions
+between meritocracy and inequality justification over time. As COVID-19
+pandemic has affected the structural opportunities to attend schools,
+universities, or workplaces, the ideal of meritocracy might have been
+challenged. As a consequence, the justification of inequality could show
+a decrease over time:
+
+\(H_5\): The positive association between meritocracy and inequality
+justification mitigates along time
 
@@
 
-The main data source is the Chilean Longitudinal Social Survey 2016--2022.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
-The sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77\% of the total national population and 93\% of the urban areas (\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}).
-
-The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
-However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
-In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
-As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
-Regarding the original sample, the response rate was 62.4\% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41\%, achieving a final sample of N = 1728.
-The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
-For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit \url{https://coes.cl/encuesta-panel/}.
+The main data source is the Chilean Longitudinal Social Survey
+2016--2022. The ELSOC has been designed as a yearly panel study to
+evaluate how individuals think, feel, and behave regarding a set of
+social issues concerning conflict and social cohesion in Chile. The
+sampling design is probabilistic, stratified, clustered, and multistage.
+It provides adequate coverage of the country's largest cities
+(Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller
+cities. Consequently, the ELSOC comprises a total of 2,927 participants
+aged between 18 and 75 years in wave 1. In addition, the sample
+represents 77\% of the total national population and 93\% of the urban
+areas (ELSOC 2022).
+
+The survey has been conducted yearly since 2016, except for the year
+2020, when it was suspended due to the pandemic. The waves 2016, 2017,
+2018, 2019 and 2022 were administered using computer-assisted personal
+interviewing (CAPI). However, a reduced version was conducted using
+computer-assisted telephone interviews (CATI) in 2021. In addition, wave
+3 included a refreshment sample to counter survey attrition. The same
+sampling strategy as that of wave 1 was implemented for selecting new
+cases. As a result, the total sample of wave 3 included 3,748 cases, of
+which 2,229 were part of the original sample and 1,519 were from the
+refreshment sample. The data from the refreshment sample is not included
+in the analytical sample because we wanted to analyze longer response
+trends. Regarding the original sample, the response rate was 62.4\% in
+wave 1, achieving N = 2,927 participants. In broader terms, the
+accumulated attrition between wave 1 and wave 6 is 41\%, achieving a
+final sample of N = 1728. The analysis used longitudinal weights to
+avoid bias due to the sampling design, allowing us to control for biases
+arising from systematic patterns of non-participation in the survey
+after the first wave. For a more detailed analysis of responses,
+attrition, and the construction of longitudinal weights, visit
+https://coes.cl/encuesta-panel/. The dataset is publicly available at:
+https://dataverse.harvard.edu/dataverse/elsoc.
 
@@
 
-The dependent variable of this study is the justification of educational inequality.
-This construct is measured using the following statement: ``It is just that high-income people have a better education for their children than people with lower incomes'' (``Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos'' in Spanish).
-Here, the respondents declared their preferences on a Likert scale from ``strongly disagree'' (1) to ``strongly agree'' (5).
-The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort (``In Chile, people are rewarded for their efforts''), and the other related to talent (``In Chile, people are rewarded for their intelligence and skills'').
-In addition, we included an indicator for economic inequality perception (``In Chile, the income differences are too large'').
-Figure \ref{fig:frecuencies} shows the average distribution throughout the years for these variables.
-We observe that there is a large group (88.2\%) of respondents who disagree or strongly disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7\%) agree or strongly agree that income differences in Chile are too large.
-On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6\%) and for their efforts (57\%).
+The dependent variable of this study is the justification of educational
+inequality. This construct is measured using the following statement:
+``It is just that high-income people have a better education for their
+children than people with lower incomes'' (``Es justo que las personas
+de altos ingresos tengan una mejor educación para sus hijos que las
+personas con ingresos más bajos'' in Spanish). Here, the respondents
+declared their preferences on a Likert scale from ``strongly disagree''
+(1) to ``strongly agree'' (5). The main independent variables refer to
+meritocratic perception, which is operationalized by two items, one
+related to effort (``In Chile, people are rewarded for their efforts''),
+and the other related to talent (``In Chile, people are rewarded for
+their intelligence and skills''). In addition, we included an indicator
+for economic inequality perception (``In Chile, the income differences
+are too large''). Figure~\ref{fig-frecuencies} shows the average
+distribution throughout the years for these variables. We observe that
+there is a large group (88.2\%) of respondents who disagree or strongly
+disagree that people with higher incomes can access better education.
+Likewise, a large group of respondents (91.7\%) agree or strongly agree
+that income differences in Chile are too large. On the other hand, there
+is an important level of disagreement (disagree + strongly disagree)
+about the perception of meritocracy, this is, whether people are
+rewarded for their talent (47.6\%) and for their efforts (57\%).
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/merit_plot}
+\centering{
+
+\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/merit_plot.png}
 
@@
 
-\caption{Frequency of responses for justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:frecuencies}
-\end{figure}
-
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
-Table \ref{tab:descriptives} shows the whole set of independent variables, the corresponding items, response categories, and frequencies.
+\caption{\label{fig-frecuencies}Frequency of responses for justification
+of educational inequality, perception of inequality, and perception of
+meritocracy for all the years analyzed}
+
+\end{figure}%
+
+For testing the hypothesis regarding status we included educational
+level and household income quintiles, as well as subjective social
+status. Finally, political identification on a left-right scale, age,
+and gender were included as control variables.
+Table~\ref{tbl-descriptives} shows the whole set of independent
+variables, the corresponding items, response categories, and
+frequencies.
 
 \begin{longtable}[]{@{}l@{}}
-\caption{\label{tab:descriptives}Independent variables}\tabularnewline
+\caption{Independent variables}\label{tbl-descriptives}\tabularnewline
 \toprule\noalign{}
@@
 
-Given the hierarchical structure of the data (observations nested in survey waves), we estimated a series of longitudinal multilevel models (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009}).
-Such an approach is suited to account for the shared variance among units (in this case individuals), adjusting the estimation of the standard errors.
-The linear multilevel models are estimated using the R library ``lme4'' (\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009, p.
-4}).
+Given the hierarchical structure of the data (observations nested in
+survey waves), we estimated a series of longitudinal multilevel models
+(Singer and Willett 2009). Such an approach is suited to account for the
+shared variance among units (in this case individuals), adjusting the
+estimation of the standard errors. The linear multilevel models are
+estimated using the R library ``lme4'' (Singer and Willett 2009, 4).
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/alluvial_dep}
+\centering{
+
+\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/alluvial_dep.png}
 
@@
 
-\caption{Change in the justification of educational inequality over time (2016-2022)}\label{fig:alluvial}
-\end{figure}
-
-Figure \ref{fig:alluvial} illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data (\citeproc{ref-rosvall_mapping_2010}{Rosvall \& Bergstrom, 2010}).
-For instance, of the 32.2\% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90\% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
+\caption{\label{fig-alluvial}Change in the justification of educational
+inequality over time (2016-2022)}
+
+\end{figure}%
+
+Figure~\ref{fig-alluvial} illustrates yearly frequencies in the
+justification of educational inequality between 2016 and 2022. Each year
+represents stacked percentual frequencies, and the flows in between
+reflect the within-subject change of opinions from one year to the next,
+as we are using longitudinal panel data (Rosvall and Bergstrom 2010).
+For instance, of the 32.2\% who strongly disagreed with inequality
+justification in 2016, about half of them kept responding the same in
+2017, whereas the other half shifted their opinion to other response
+categories. In general, the large majority - between 80 and 90\% -
+disagrees with the justification of educational inequality throughout
+the years. Despite this overall tendency, we also observe that the
+disagreement with inequality justification (disagree + strongly
+disagree) tends to go down in the last wave. This change is mostly a
+result of the radical decrease in the lowest response category (strongly
+disagree), which diminishes to half when compared with previous years.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/years_plot}
+\centering{
+
+\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/years_plot.png}
 
@@
 
-\caption{Change in the mean of justification of educational inequality, perception of inequality, and perception of meritocracy over the years}\label{fig:means}
-\end{figure}
-
-Figure \ref{fig:means} shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
+\caption{\label{fig-means}Change in the mean of justification of
+educational inequality, perception of inequality, and perception of
+meritocracy over the years}
+
+\end{figure}%
+
+Figure~\ref{fig-means} shows the average changes in the main variables
+considered for this study. Here we observe that the justification of
+educational inequality has the lowest average throughout the years when
+compared with the other (independent) variables, whereas the highest
+average is consistently represented by inequality perception.
+Interestingly, in the last wave of the study (2022) the justification of
+inequality increases whereas the perception of inequality decreases. As
+the merit variables are concerned, they show a very similar pattern in
+terms of averages and changes over the years, being the perception of
+meritocracy related to effort always lower than the one associated with
+talent.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{output/graphs/corr}
+\centering{
+
+\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/corr.png}
 
@@
 
-\caption{Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed}\label{fig:correlation}
-\end{figure}
-
-Figure \ref{fig:correlation} presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (\emph{r}=-0.29, p\textless.01) and a moderate and positive association with both meritocracy perception variables (\emph{r}=0.18, p\textless.01; \emph{r}=0.15, p\textless.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (\emph{r}=-0.15, p\textless.01; \emph{r}=-0.10, p\textless.01).
-Finally, the two meritocracy variables present a high and positive association with each other (\emph{r}=0.74, p\textless.01).
+\caption{\label{fig-correlation}Correlation matrix of justification of
+educational inequality, perception of inequality, and perception of
+meritocracy for all the years analyzed}
+
+\end{figure}%
+
+Figure~\ref{fig-correlation} presents a correlation matrix of the main
+variables analyzed, using data from all survey waves. In this matrix,
+the correlations vary between low and moderate values. Justification of
+economic inequality depicts a moderate and negative association with the
+perception of inequality (\emph{r}=-0.29, p\textless.01) and a moderate
+and positive association with both meritocracy perception variables
+(\emph{r}=0.18, p\textless.01; \emph{r}=0.15, p\textless.01). Regarding
+the perception of inequality, it presents a moderate and negative
+association with both meritocracy perception variables (\emph{r}=-0.15,
+p\textless.01; \emph{r}=-0.10, p\textless.01). Finally, the two
+meritocracy variables present a high and positive association with each
+other (\emph{r}=0.74, p\textless.01).
 
@@
 
-Table \ref{tab:multilevel} shows the multilevel estimation results for the justification of educational inequality.
-Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 (\(\beta\)=-0.15, p\textless.001), 2018 (\(\beta\)=-0.06, p\textless.05), and 2019 (\(\beta\)=-0.07, p\textless.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality (\(\beta\)=0.21, p\textless.001), suggesting a non-linear change in this variable.
-Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 (\(\beta\)=0.04, p\textless.01).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality (\(\beta\)=0.05, p\textless.01), but it has a larger weight than the perception of talent.
-
-\begin{table}[!ht]
-\caption{\label{tab:multilevel} Multilevel longitudinal models for the justification of inequality in education}
-\begin{center}
-\scalebox{0.765}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\
-\hline
-Intercept                           & $1.86^{***}$  & $1.98^{***}$  & $1.72^{***}$  & $2.22^{***}$  & $2.11^{***}$  & $2.04^{***}$  \\
-                                    & $(0.02)$      & $(0.04)$      & $(0.09)$      & $(0.10)$      & $(0.10)$      & $(0.10)$      \\
-Wave (Ref.= Wave 2016)              &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Wave 2017                     & $-0.15^{***}$ &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-\quad Wave 2018                     & $-0.06^{*}$   &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2019                     & $-0.07^{**}$  &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2022                     & $0.21^{***}$  &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-Wave                                &               & $-0.17^{***}$ & $-0.17^{***}$ & $-0.13^{***}$ & $-0.14^{***}$ & $-0.14^{***}$ \\
-                                    &               & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      \\
-Wave$^2$                            &               & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  \\
-                                    &               & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
-Household Income (Ref.= Quintile 1) &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Quintile 2                    &               &               & $0.07$        & $0.07$        & $0.07$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 3                    &               &               & $0.08$        & $0.08$        & $0.08$        & $0.08$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 4                    &               &               & $0.12^{*}$    & $0.13^{*}$    & $0.13^{*}$    & $0.13^{*}$    \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile 5                    &               &               & $0.19^{**}$   & $0.19^{**}$   & $0.19^{***}$  & $0.20^{***}$  \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile (No information)     &               &               & $0.17^{*}$    & $0.17^{*}$    & $0.18^{*}$    & $0.18^{*}$    \\
-                                    &               &               & $(0.08)$      & $(0.08)$      & $(0.08)$      & $(0.08)$      \\
-Age (Ref.= 18-29)                   &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Age 30-49                     &               &               & $0.02$        & $0.02$        & $0.02$        & $0.02$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 50-64                     &               &               & $0.07$        & $0.08$        & $0.08$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 65 or more                &               &               & $0.19^{**}$   & $0.20^{**}$   & $0.19^{**}$   & $0.18^{**}$   \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-Gender (Ref. Male)                  &               &               & $-0.08^{*}$   & $-0.08^{*}$   & $-0.08^{*}$   & $-0.07^{*}$   \\
-                                    &               &               & $(0.03)$      & $(0.03)$      & $(0.03)$      & $(0.03)$      \\
-Pol. pos (Ref.= Left)               &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Center                        &               &               & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Right                         &               &               & $0.29^{***}$  & $0.28^{***}$  & $0.27^{***}$  & $0.27^{***}$  \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Does not identify             &               &               & $0.15^{***}$  & $0.15^{***}$  & $0.14^{***}$  & $0.14^{***}$  \\
-                                    &               &               & $(0.04)$      & $(0.04)$      & $(0.04)$      & $(0.04)$      \\
-Inequality perception               &               &               &               & $-0.12^{***}$ & $-0.12^{***}$ & $-0.12^{***}$ \\
-                                    &               &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-Merit: Talent                       &               &               &               &               & $0.04^{***}$  & $0.01$        \\
-                                    &               &               &               &               & $(0.01)$      & $(0.01)$      \\
-Merit: Effort                       &               &               &               &               &               & $0.05^{***}$  \\
-                                    &               &               &               &               &               & $(0.01)$      \\
-\hline
-BIC                                 & $30911.29$    & $30907.80$    & $31053.70$    & $30942.30$    & $30940.78$    & $30936.16$    \\
-Num. obs.                           & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        \\
-Num. groups: Individuals            & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        \\
-Var: Individuals (Intercept)        & $0.19$        & $0.19$        & $0.18$        & $0.17$        & $0.17$        & $0.16$        \\
-Var: Residual                       & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.46$        \\
-\hline
-\multicolumn{7}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.}}
-\end{tabular}
+Table~\ref{tbl-multilevel} shows the multilevel estimation results for
+the justification of educational inequality. Model 1 includes the survey
+waves to estimate intertemporal changes in the dependent variable.
+Taking 2016 as a reference point, we can observe a staggered decrease in
+2017 (\(\beta\)=-0.15, p\textless.001), 2018 (\(\beta\)=-0.06,
+p\textless.05), and 2019 (\(\beta\)=-0.07, p\textless.01). Nevertheless,
+in the last wave of 2022 there is a radical increase in level of
+justification of economic inequality (\(\beta\)=0.21, p\textless.001),
+suggesting a non-linear change in this variable. Attempting to model
+this path of change over time, Model 2 incorporates time (survey waves)
+as a continuous variable as well as its quadratic term, representing the
+nonlinear association initially observed in Model 1. On the one hand,
+the survey wave depicts a negative association, expressing an average
+decrease in inequality justification over time, but on the other hand,
+the quadratic wave term is positive, indicating the reversion of this
+path in the last measurement point.
+
+Model 3 adds the sociodemographic variables, where having a higher
+income and being older (65 or older compared to 18-29) have a
+significant and positive influence on the justification of educational
+inequality. On the contrary, the gender variable has a negative effect,
+expressing that women on average justify less inequality in education
+when compared to males. Educational level and subjective social status
+are not shown in this table for the sake of space (as they do not
+exhibit significant effects), but the complete estimation of the models
+is presented in the appendix. Model 3 also includes the control
+variables, where identifying oneself with the center, right-wing, or
+when reporting no political identification, has a significant and
+positive association on the justification of educational inequality -
+when compared to being left-wing oriented. Inequality perception is
+added in Model 4, showing a negative association with the justification
+of educational inequality as hypothesized, remaining stable when
+controlling for the rest of the variables.
+
+Models 5 and 6 introduce the meritocratic variables: talent (if
+intelligence and abilities are rewarded in society) and effort (if
+efforts are rewarded in society). In line with our hypotheses, the
+perception that talent is rewarded has a positive influence on the
+justification of educational inequality in Model 5 (\(\beta\)=0.04,
+p\textless.01). However, when controlling for the perception that effort
+is rewarded, this effect is no longer significant. In this sense, Model
+6 shows that the perception that effort is rewarded in society is not
+only positively associated with the justification of educational
+inequality (\(\beta\)=0.05, p\textless.01), but it has a larger weight
+than the perception of talent.
+
+\begin{table}
+
+\caption{\label{tbl-multilevel}Multilevel longitudinal models for the
+justification of inequality in education}
+
+\centering{
+
+Statistical models
+
+~
+
+Model 1
+
+Model 2
+
+Model 3
+
+Model 4
+
+Model 5
+
+Model 6
+
+Intercept
+
+1.84***
+
+1.87***
+
+1.80***
+
+2.39***
+
+2.22***
+
+2.13***
+
+~
+
+(0.02)
+
+(0.03)
+
+(0.09)
+
+(0.10)
+
+(0.10)
+
+(0.10)
+
+Wave (Ref.= Wave 2016)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2017
+
+-0.14***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2018
+
+-0.04
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2019
+
+-0.04
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2022
+
+0.23***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2023
+
+0.28***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Wave
+
+~
+
+-0.08***
+
+-0.08***
+
+-0.05**
+
+-0.06**
+
+-0.06**
+
+~
+
+~
+
+(0.02)
+
+(0.02)
+
+(0.02)
+
+(0.02)
+
+(0.02)
+
+Wave\^{}2
+
+~
+
+0.02***
+
+0.02***
+
+0.01***
+
+0.01***
+
+0.01***
+
+~
+
+~
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+Household Income (Ref.= Quintile 1)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Quintile 2
+
+~
+
+~
+
+0.05
+
+0.04
+
+0.04
+
+0.04
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Quintile 3
+
+~
+
+~
+
+0.10
+
+0.09
+
+0.09
+
+0.10
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Quintile 4
+
+~
+
+~
+
+0.11
+
+0.11*
+
+0.11*
+
+0.11*
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Quintile 5
+
+~
+
+~
+
+0.15**
+
+0.15**
+
+0.16**
+
+0.16**
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+~~~~~Quintile (No information)
+
+~
+
+~
+
+0.16*
+
+0.17*
+
+0.18*
+
+0.17*
+
+~
+
+~
+
+~
+
+(0.08)
+
+(0.08)
+
+(0.08)
+
+(0.08)
+
+Age (Ref.= 18-29)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Age 30-49
+
+~
+
+~
+
+-0.02
+
+-0.01
+
+-0.01
+
+-0.02
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Age 50-64
+
+~
+
+~
+
+0.03
+
+0.04
+
+0.03
+
+0.03
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Age 65 or more
+
+~
+
+~
+
+0.14*
+
+0.14*
+
+0.13*
+
+0.12*
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+Gender (Ref. Male)
+
+~
+
+~
+
+-0.10**
+
+-0.10**
+
+-0.09**
+
+-0.08*
+
+~
+
+~
+
+~
+
+(0.03)
+
+(0.03)
+
+(0.03)
+
+(0.03)
+
+Pol. pos (Ref.= Left)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Center
+
+~
+
+~
+
+0.08
+
+0.07
+
+0.07
+
+0.07
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Right
+
+~
+
+~
+
+0.28***
+
+0.26***
+
+0.25***
+
+0.25***
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Does not identify
+
+~
+
+~
+
+0.09*
+
+0.08
+
+0.08
+
+0.08
+
+~
+
+~
+
+~
+
+(0.04)
+
+(0.04)
+
+(0.04)
+
+(0.04)
+
+Inequality perception
+
+~
+
+~
+
+~
+
+-0.15***
+
+-0.14***
+
+-0.14***
+
+~
+
+~
+
+~
+
+~
+
+(0.01)
+
+(0.01)
+
+(0.01)
+
+Merit: Talent
+
+~
+
+~
+
+~
+
+~
+
+0.06***
+
+0.01
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.01)
+
+(0.01)
+
+Merit: Effort
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.08***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.01)
+
+BIC
+
+37741.49
+
+37740.80
+
+37897.19
+
+37723.09
+
+37697.94
+
+37668.38
+
+Num. obs.
+
+9519
+
+9519
+
+9519
+
+9519
+
+9519
+
+9519
+
+Num. groups: Individuals
+
+1701
+
+1701
+
+1701
+
+1701
+
+1701
+
+1701
+
+Var: Individuals (Intercept)
+
+0.18
+
+0.18
+
+0.17
+
+0.17
+
+0.16
+
+0.16
+
+Var: Residual
+
+0.53
+
+0.53
+
+0.53
+
+0.52
+
+0.52
+
+0.52
+
+*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
+Note: Model 3 to 6 are controlled by educational level, and subjective
+social status with no significant effects.
+
 }
-\label{table:coefficients}
-\end{center}
-\end{table}
-
-\begin{table}[!ht]
-\caption{\label{tab:interact} Interaction effects for the justification of economic inequality}
-\begin{center}
-\scalebox{0.765}{
-\begin{threeparttable}
-\begin{tabular}{l c c c}
-\toprule
- & Model 1 & Model 2 & Model 3 \\
-\midrule
-Intercept                                  & $2.04^{***}$  & $1.99^{***}$  & $2.17^{***}$  \\
-                                           & $(0.10)$      & $(0.11)$      & $(0.11)$      \\
-Wave                                       & $-0.14^{***}$ & $-0.11^{***}$ & $-0.17^{***}$ \\
-                                           & $(0.02)$      & $(0.03)$      & $(0.03)$      \\
-Wave$^2$                                   & $0.03^{***}$  & $0.02^{***}$  & $0.03^{***}$  \\
-                                           & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
-Merit: Talent                              & $0.01$        & $0.03$        & $0.01$        \\
-                                           & $(0.01)$      & $(0.02)$      & $(0.01)$      \\
-Merit: Effort                              & $0.05^{***}$  & $0.06^{***}$  & $0.02$        \\
-                                           & $(0.01)$      & $(0.01)$      & $(0.02)$      \\
-Wave * Merit: Talent                       &               & $-0.00$       &               \\
-                                           &               & $(0.01)$      &               \\
-Wave * Merit: Effort                       &               &               & $0.01^{*}$    \\
-                                           &               &               & $(0.01)$      \\
-\midrule
-BIC                                        & $30936.16$    & $29763.63$    & $29921.16$    \\
-Num. obs.                                  & $8042$        & $8042$        & $8042$        \\
-Num. groups: Individuals                   & $1689$        & $1689$        & $1689$        \\
-Var: Individuals (Intercept)               & $0.16$        & $1.13$        & $1.03$        \\
-Var: Residual                              & $0.46$        & $0.32$        & $0.33$        \\
-Var: Individuals Wave                      & $$            & $0.02$        & $0.02$        \\
-Var: Individuals Merit: Talent             & $$            & $0.11$        & $$            \\
-Cov: Individuals (Intercept) Wave          & $$            & $-0.07$       & $-0.07$       \\
-Cov: Individuals (Intercept) Merit: Talent & $$            & $-0.30$       & $$            \\
-Cov: Individuals Wave Merit: Talent        & $$            & $0.00$        & $$            \\
-Var: Individuals Merit: Effort             & $$            & $$            & $0.10$        \\
-Cov: Individuals (Intercept) Merit: Effort & $$            & $$            & $-0.27$       \\
-Cov: Individuals Wave Merit: Effort        & $$            & $$            & $0.01$        \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}[flushleft]
-\scriptsize{\item *** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception}
-\end{tablenotes}
-\end{threeparttable}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-interact}}
+
+\centering{
+
+Statistical models
+
+~
+
+Model 1
+
+Model 2
+
+Model 3
+
+Intercept
+
+2.13***
+
+1.96***
+
+2.10***
+
+~
+
+(0.10)
+
+(0.11)
+
+(0.11)
+
+Wave
+
+-0.06**
+
+-0.03
+
+-0.06**
+
+~
+
+(0.02)
+
+(0.02)
+
+(0.02)
+
+Wave\^{}2
+
+0.01***
+
+0.01***
+
+0.01***
+
+~
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+Merit: Talent
+
+0.01
+
+0.04
+
+0.01
+
+~
+
+(0.01)
+
+(0.02)
+
+(0.01)
+
+Merit: Effort
+
+0.08***
+
+0.08***
+
+0.07**
+
+~
+
+(0.01)
+
+(0.01)
+
+(0.02)
+
+Wave * Merit: Talent
+
+~
+
+-0.01
+
+~
+
+~
+
+~
+
+(0.00)
+
+~
+
+Wave * Merit: Effort
+
+~
+
+~
+
+0.00
+
+~
+
+~
+
+~
+
+(0.00)
+
+BIC
+
+37668.38
+
+36196.86
+
+36307.66
+
+Num. obs.
+
+9519
+
+9519
+
+9519
+
+Num. groups: Individuals
+
+1701
+
+1701
+
+1701
+
+Var: Individuals (Intercept)
+
+0.16
+
+0.96
+
+0.88
+
+Var: Residual
+
+0.52
+
+0.36
+
+0.37
+
+Var: Individuals Wave
+
+~
+
+0.02
+
+0.02
+
+Var: Individuals Merit: Talent
+
+~
+
+0.10
+
+~
+
+Cov: Individuals (Intercept) Wave
+
+~
+
+-0.05
+
+-0.05
+
+Cov: Individuals (Intercept) Merit: Talent
+
+~
+
+-0.26
+
+~
+
+Cov: Individuals Wave Merit: Talent
+
+~
+
+-0.00
+
+~
+
+Var: Individuals Merit: Effort
+
+~
+
+~
+
+0.10
+
+Cov: Individuals (Intercept) Merit: Effort
+
+~
+
+~
+
+-0.24
+
+Cov: Individuals Wave Merit: Effort
+
+~
+
+~
+
+0.00
+
+\item *
+
+** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
+Note: All the models are controlled by educational level, income
+quintile, subjective social status, age, gender, political position, and
+inequality perception
+
 }
-\label{table:coefficients}
-\end{center}
-\end{table}
-
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in Table \ref{tab:interact}.
-Model 1 is shown as a baseline model, it is the same as Model 6 in Table 1 but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable (\(\beta\)=0.01, p\textless.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
+
+\end{table}%
+
+In this final part of the analysis, we contrast longitudinal hypotheses
+regarding changes in the relationship between meritocracy and
+justification of economic inequality over time. In hypothesis 5 we
+proposed that the association between the perception of meritocracy and
+inequality justification mitigates over time, as meritocratic ideals
+could have weakened due to critical situations associated with the COVID
+pandemic. We test this hypothesis through interaction effects, displayed
+in Table~\ref{tbl-interact}. Model 1 is shown as a baseline model, it is
+the same as Model 6 in Table~\ref{tbl-interact} but it only displays the
+variables involved in the interaction for the sake of space (all other
+variables are controlled for). Model 2 adds the interaction between the
+perception of meritocracy based on talent and time (panel wave), whereas
+Model 3 does the same but now for the perception of meritocracy related
+to effort. As observed, only the effort variable shows a significant
+interaction with the time variable (\(\beta\)=0.01, p\textless.05),
+meaning that, on average, the association between the perception of
+meritocracy based on effort and the justification of inequality
+increases by 0.01 points in every measurement point.
 
@@
 
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
-In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
-We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process (\citeproc{ref-palanza_chile_2023}{Palanza \& Sotomayor, 2023}; \citeproc{ref-sazo_chile_2023}{Sazo, 2023}).
-
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality (\(H_2\)).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude (\(H_3\)).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education (\(H_4\)).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
-In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
-This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis (\(H_5\)).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
+Our research sought to examine the relationship between the perception
+of meritocracy and the justification of inequality in education from a
+longitudinal perspective amid the health crisis generated by the
+COVID-19 pandemic. In general, our results showed mixed evidence about
+our hypotheses. Whereas we find support for the association between the
+perception of meritocracy and justification of inequality across time,
+the analyses that involved changes in justification of inequality are
+against our initial proposal. We argued that in times of crisis, greater
+exposure to risk due to the health context would result in a lower
+justification of inequality by citizens. In this regard, longitudinal
+models showed that the justification of inequality in education is far
+from being linear, and whereas a decreasing pattern was found between
+2017 and 2019, there is a striking increase in the last measurement
+point (2022), after the peak of the COVID pandemic. As we mentioned at
+the beginning, given our data limitations it is not possible to
+attribute such changes only to the pandemic and the related economic and
+social crises. As in the Chilean case, the pandemic period occurred
+along with a constitutional process, disentangling different forces
+driving public opinion it becomes difficult and deserves further
+qualitative and quantitative studies. A closer look to the political
+environment during the last year in the country could give additional
+hints. Some recent studies have argued that the election of a far
+left-wing constitutional assembly during this period generated a
+backslash effect, mainly due to several scandals that led to the
+delegitimation of this assembly. This could have driven preferences in a
+conservative direction, which was reflected in the election of a
+right-wing second assembly after the failed first constitutional process
+(Palanza and Sotomayor 2023; Sazo 2023).
+
+Regarding the status position, our central hypothesis was that
+individuals in more advantaged situations and with greater resources
+would tend to defend their interests and justify greater education
+inequality (\(H_2\)). In this regard, our results show a positive and
+robust relationship between income and the justification of inequality,
+which would align with our original hypothesis based on the rational
+interests of the better-off that would lead to a larger inequality
+justification. Besides status, we also considered inequality perception,
+as the literature on attitudes toward inequality has argued that a
+greater justification of inequality is associated with its perceived
+magnitude (\(H_3\)). Consistent with this hypothesis, our results showed
+that, throughout the years, a larger perceived economic inequality
+motivates a lower justification of inequality in education.
+
+The main focus of this study was on the relationship between meritocracy
+and justification of inequality, arguing that those who perceive that
+the society in which they live complies with meritocratic principles,
+would be more inclined to justify inequality in education (\(H_4\)). At
+first look, the results are consistent with the hypothesis; however,
+some nuances are worth attending concerning different aspects of
+meritocracy, namely effort and talent. Our results are favorable for the
+effort dimension but not confirmed for the talent dimension. In other
+words, the perception of meritocracy in terms of rewarded effort is more
+relevant in justifying access to education than the perception of
+talent. This could be explained as that talent could be more associated
+with luck in terms of random assignment, and in this terms would not be
+enough reason (as it is effort) to justify educational inequality.
+
+Finally, as our first hypothesis supported, we expected that the
+relationship between meritocratic perceptions and justification would
+tend to diminish over time. However, longitudinal evidence shows that
+the perception of meritocracy linked to effort strengthens its relation
+with the justification of inequality in education, against the temporal
+change hypothesis (\(H_5\)). In other words, meritocracy seems to gain
+force concerning justifying inequality and not the other way around as
+we expected given the social context in the post-covid era. An
+alternative interpretation is that in times of high exposure to risk,
+people tend to stick more to narratives that provide greater security
+regardless of the rest of the society - such as individual effort -,
+which might also be accompanied by lower degrees of solidarity to
+safeguard the well-being of themselves and their surrounding close
+peers. Again, this certainly needs further studies that include
+additional specific conceptual and methodological approaches to this
+phenomenon.
 
 Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}), the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
-Finally, there are limitations involving the longitudinal dimension of the data used.
-Given constraints for data collection during the pandemic, the ELSOC study had to shorten the questionnaire in 2021 and unfortunately, the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered carefully, and further data waves could give us more information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
-For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.
-Furthermore, this dataset could be used for comparing inequality justification in policy areas besides education, such as health and pensions.
-Future studies may shed light on how meritocratic perceptions and beliefs affect differently such areas and their changes over time, which would give relevant hints from social sciences for the discussion about societal and cultural changes as well as their impacts on solidarity and social cohesion.
-
-\pagebreak
-
-\emph{Acknowledgments}
-
-The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022\_046
+First, we know that inequalities in education can manifest themselves in
+terms of outcomes as well as in areas such as access to social services
+in education as part of welfare policies. In this regard, the construct
+that we have captured with our indicator can be framed more in the
+latter, while the evaluation of justice involved in this process does
+not consider aspects such as equality of educational opportunities or
+observable outcomes such as the performance of individual or educational
+institutions. Second, as it has been established by recent research
+(Castillo et al. 2023), the perception of meritocracy can be understood
+through individual attributions regarding effort and talent, but also as
+a function of structural aspects such as family of origin status or
+social capital. In this sense, the available measurements restricted us
+to the use of single indicators for effort and talent, and further
+research should consider improving the measurement quality in this
+regard. Finally, there are limitations involving the longitudinal
+dimension of the data used. Given constraints for data collection during
+the pandemic, the ELSOC study had to shorten the questionnaire in 2021
+and unfortunately, the educational justice indicator was excluded from
+that survey. Consequently, the results of temporal change have to be
+considered carefully, and further data waves could give us more
+information on this matter.
+
+The capabilities of the ELSOC longitudinal database are not limited to
+micro-level estimates. In this regard, thanks to the sampling strategy
+of the survey, it would be possible to make contextual estimates at the
+municipality level in future studies. For instance, statistical models
+could include contextual information from administrative data sources,
+allowing testing for hypotheses that include contextual socioeconomic
+inequality as well as its dynamics over time. Furthermore, this dataset
+could be used for comparing inequality justification in policy areas
+besides education, such as health and pensions. Future studies may shed
+light on how meritocratic perceptions and beliefs affect differently
+such areas and their changes over time, which would give relevant hints
+from social sciences for the discussion about societal and cultural
+changes as well as their impacts on solidarity and social cohesion.
 
@@
 \bibitem[\citeproctext]{ref-barry_theories_1989}
-Barry, B. (1989). \emph{Theories of justice}. Univ. of California Pr.
+Barry, Brian. 1989. \emph{Theories of Justice}. A Treatise on Social
+Justice 1. Berkeley: Univ. of California Pr.
 
 \bibitem[\citeproctext]{ref-batruch_belief_2023}
-Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
-(2023).
-Belief in {School Meritocracy} and the {Legitimization} of {Social} and {Income Inequality}.
-\emph{Social Psychological and Personality Science}, \emph{14}(5), 621--635.
-\url{https://doi.org/10.1177/19485506221111017}
+Batruch, Anatolia, Jolanda Jetten, Herman Van De Werfhorst, Céline
+Darnon, and Fabrizio Butera. 2023. {``Belief in {School Meritocracy} and
+the {Legitimization} of {Social} and {Income Inequality}.''}
+\emph{Social Psychological and Personality Science} 14 (5): 621--35.
+\url{https://doi.org/10.1177/19485506221111017}.
 
 \bibitem[\citeproctext]{ref-bell_politics_2020}
-Bell, E.
-(2020).
-The {Politics} of {Designing Tuition-Free College}: {How Socially Constructed Target Populations Influence Policy Support}.
-\emph{The Journal of Higher Education}, \emph{91}(6), 888--926.
-\url{https://doi.org/10.1080/00221546.2019.1706015}
+Bell, Elizabeth. 2020. {``The {Politics} of {Designing Tuition-Free
+College}: {How Socially Constructed Target Populations Influence Policy
+Support}.''} \emph{The Journal of Higher Education} 91 (6): 888--926.
+\url{https://doi.org/10.1080/00221546.2019.1706015}.
 
 \bibitem[\citeproctext]{ref-bellei_estudio_2013}
-Bellei, C.
-(2013).
-El estudio de la segregaci{ó}n socioecon{ó}mica y acad{é}mica de la educaci{ó}n chilena.
-\emph{Estudios Pedag{ó}gicos (Valdivia)}, \emph{39}(1), 325--345.
-\url{https://doi.org/10.4067/S0718-07052013000100019}
+Bellei, Cristián. 2013. {``El Estudio de La Segregaci{ó}n
+Socioecon{ó}mica y Acad{é}mica de La Educaci{ó}n Chilena.''}
+\emph{Estudios Pedag{ó}gicos (Valdivia)} 39 (1): 325--45.
+\url{https://doi.org/10.4067/S0718-07052013000100019}.
 
 \bibitem[\citeproctext]{ref-breznau_welfare_2021}
-Breznau, N.
-(2021).
-The welfare state and risk perceptions: The {Novel Coronavirus Pandemic} and public concern in 70 countries.
-\emph{European Societies}, \emph{23}(sup1), S33--S46.
-\url{https://doi.org/10.1080/14616696.2020.1793215}
+Breznau, Nate. 2021. {``The Welfare State and Risk Perceptions: The
+{Novel Coronavirus Pandemic} and Public Concern in 70 Countries.''}
+\emph{European Societies} 23 (sup1): S33--46.
+\url{https://doi.org/10.1080/14616696.2020.1793215}.
 
 \bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
-Busemeyer, M.
-R., \& Iversen, T.
-(2020).
-The {Welfare State} with {Private Alternatives}: {The Transformation} of {Popular Support} for {Social Insurance}.
-\emph{The Journal of Politics}, \emph{82}(2), 671--686.
-\url{https://doi.org/10.1086/706980}
+Busemeyer, Marius R., and Torben Iversen. 2020. {``The {Welfare State}
+with {Private Alternatives}: {The Transformation} of {Popular Support}
+for {Social Insurance}.''} \emph{The Journal of Politics} 82 (2):
+671--86. \url{https://doi.org/10.1086/706980}.
 
 \bibitem[\citeproctext]{ref-Castillo2011}
-Castillo, J.
-C.
-(2011).
-Legitimacy of {Inequality} in a {Highly Unequal Context}: {Evidence} from the {Chilean Case}.
-\emph{Social Justice Research}, \emph{24}(4), 314--340.
-\url{https://doi.org/10.1007/s11211-011-0144-5}
+Castillo, Juan Carlos. 2011. {``Legitimacy of {Inequality} in a {Highly
+Unequal Context}: {Evidence} from the {Chilean Case}.''} \emph{Social
+Justice Research} 24 (4): 314--40.
+\url{https://doi.org/10.1007/s11211-011-0144-5}.
 
 \bibitem[\citeproctext]{ref-Castillo2012a_justice}
-Castillo, J.
-C.
-(2012).
-Is {Inequality Becoming Just}?
-{Changes} in {Public Opinion} about {Economic Distribution} in {Chile}.
-\emph{Bulletin of Latin American Research}, \emph{31}(1), 1--18.
-\url{https://doi.org/10.1111/j.1470-9856.2011.00605.x}
+---------. 2012. {``Is {Inequality Becoming Just}? {Changes} in {Public
+Opinion} about {Economic Distribution} in {Chile}.''} \emph{Bulletin of
+Latin American Research} 31 (1): 1--18.
+\url{https://doi.org/10.1111/j.1470-9856.2011.00605.x}.
 
 \bibitem[\citeproctext]{ref-castillo_perception_2022}
-Castillo, J.
-C., García-Castro, J.-D., \& Venegas, M.
-(2022).
-Perception of economic inequality: Concepts, associated factors and prospects of a burgeoning research agenda.
-\emph{International Journal of Social Psychology}, \emph{37}(1), 180--207.
-\url{https://doi.org/10.1080/02134748.2021.2009275}
+Castillo, Juan Carlos, Juan-Diego García-Castro, and Martín Venegas.
+2022. {``Perception of Economic Inequality: Concepts, Associated Factors
+and Prospects of a Burgeoning Research Agenda.''} \emph{International
+Journal of Social Psychology} 37 (1): 180--207.
+\url{https://doi.org/10.1080/02134748.2021.2009275}.
 
 \bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
-Castillo, J.
-C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
-(2023).
-A {Multidimensional Approach} for {Measuring Meritocratic Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP Social Inequality Survey}.
-\emph{International Journal of Sociology}, 1--25.
-\url{https://doi.org/10.1080/00207659.2023.2274712}
+Castillo, Juan Carlos, Julio Iturra, Luis Maldonado, Jorge Atria, and
+Francisco Meneses. 2023. {``A {Multidimensional Approach} for {Measuring
+Meritocratic Beliefs}: {Advantages}, {Limitations} and {Alternatives} to
+the {ISSP Social Inequality Survey}.''} \emph{International Journal of
+Sociology}, October, 1--25.
+\url{https://doi.org/10.1080/00207659.2023.2274712}.
 
 \bibitem[\citeproctext]{ref-confortiLegitimationInequalityAmerican1992}
-Conforti, J. M. (1992). The legitimation of inequality in {American} education. \emph{The Urban Review}, \emph{24}(4), 227--238. \url{https://doi.org/10.1007/BF01108357}
+Conforti, Joseph M. 1992. {``The Legitimation of Inequality in
+{American} Education.''} \emph{The Urban Review} 24 (4): 227--38.
+\url{https://doi.org/10.1007/BF01108357}.
 
 \bibitem[\citeproctext]{ref-corvalanMercadoEscolarLibertad2016}
-Corvalán, J., Carrasco, A., \& García-Huidobro, J. E. (Eds.). (2016). \emph{Mercado escolar: {Libertad}, diversidad y desigualdad} (1st ed.). Ediciones UC. \url{https://doi.org/10.2307/j.ctv14rmrhn}
+Corvalán, Javier, Alejandro Carrasco, and J. E. García-Huidobro, eds.
+2016. \emph{Mercado Escolar: {Libertad}, Diversidad y Desigualdad}. 1st
+ed. Santiago: Ediciones UC. \url{https://doi.org/10.2307/j.ctv14rmrhn}.
 
 \bibitem[\citeproctext]{ref-darmodyImpactsCOVID19Control2021}
-Darmody, M., Smyth, E., \& Russell, H.
-(2021).
-Impacts of the {COVID-19 Control Measures} on {Widening Educational Inequalities}.
-\emph{YOUNG}, \emph{29}(4), 366--380.
-\url{https://doi.org/10.1177/11033088211027412}
+Darmody, Merike, Emer Smyth, and Helen Russell. 2021. {``Impacts of the
+{COVID-19 Control Measures} on {Widening Educational Inequalities}.''}
+\emph{YOUNG} 29 (4): 366--80.
+\url{https://doi.org/10.1177/11033088211027412}.
 
 \bibitem[\citeproctext]{ref-dayPerceivedIdealInequality2023}
-Day, M.
-V., \& Norton, M.
-I.
-(2023).
-Perceived and {Ideal Inequality} in {University Endowments} in the {United States}.
-\emph{Personality and Social Psychology Bulletin}, \emph{49}(8), 1151--1165.
-\url{https://doi.org/10.1177/01461672221083766}
+Day, Martin V., and Michael I. Norton. 2023. {``Perceived and {Ideal
+Inequality} in {University Endowments} in the {United States}.''}
+\emph{Personality and Social Psychology Bulletin} 49 (8): 1151--65.
+\url{https://doi.org/10.1177/01461672221083766}.
 
 \bibitem[\citeproctext]{ref-elsoc_estudio_2022}
-ELSOC, S. T. (2022). \emph{Estudio {Longitudinal Social} de {Chile}}. Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
+ELSOC, Survey Team. 2022. {``Estudio {Longitudinal Social} de
+{Chile}.''} Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}.
 
 \bibitem[\citeproctext]{ref-Evans2017}
-Evans, M.
-D.
-R., \& Kelley, J.
-(2017).
-Communism, {Capitalism}, and {Images} of {Class}: {Effects} of {Reference Groups}, {Reality}, and {Regime} in 43 {Nations} and 110,000 {Individuals}, 1987-2009.
-\emph{Cross-Cultural Research}, \emph{51}(514), 315--359.
-\url{https://doi.org/10.1177/1069397116677963}
+Evans, M. D. R., and Jonathan Kelley. 2017. {``Communism, {Capitalism},
+and {Images} of {Class}: {Effects} of {Reference Groups}, {Reality}, and
+{Regime} in 43 {Nations} and 110,000 {Individuals}, 1987-2009.''}
+\emph{Cross-Cultural Research} 51 (514): 315--59.
+\url{https://doi.org/10.1177/1069397116677963}.
 
 \bibitem[\citeproctext]{ref-Evans1992}
-Evans, M.
-D.
-R., Kelley, J., \& Kolosi, T.
-(1992).
-Images of {Class}: {Public Perceptions} in {Hungary} and {Australia}.
-\emph{American Sociological Review}, \emph{57}(4), 461.
-\url{https://doi.org/10.2307/2096095}
+Evans, M. D. R., Jonathan Kelley, and Tamas Kolosi. 1992. {``Images of
+{Class}: {Public Perceptions} in {Hungary} and {Australia}.''}
+\emph{American Sociological Review} 57 (4): 461.
+\url{https://doi.org/10.2307/2096095}.
 
 \bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
-García-Sánchez, E., \& De Carvalho Galvão, S.
-(2022).
-Las creencias que justifican la desigualdad moderan la relaci{ó}n entre el estatus socioecon{ó}mico y el apoyo a la redistribuci{ó}n.
-\emph{Revista Internacional de Sociolog{í}a}, \emph{80}(3), e210.
-\url{https://doi.org/10.3989/ris.2022.80.3.21.29}
+García-Sánchez, Efraín, and Sofia De Carvalho. 2022. {``Las Creencias
+Que Justifican La Desigualdad Moderan La Relaci{ó}n Entre El Estatus
+Socioecon{ó}mico y El Apoyo a La Redistribuci{ó}n.''} \emph{Revista
+Internacional de Sociolog{í}a} 80 (3): e210.
+\url{https://doi.org/10.3989/ris.2022.80.3.21.29}.
 
 \bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
-Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality. \emph{Economics \& Politics}, \emph{30}(1), 27--54. \url{https://doi.org/10.1111/ecpo.12103}
+Gimpelson, Vladimir, and Daniel Treisman. 2018. {``Misperceiving
+Inequality.''} \emph{Economics \& Politics} 30 (1): 27--54.
+\url{https://doi.org/10.1111/ecpo.12103}.
 
 \bibitem[\citeproctext]{ref-goldin_meritocracy_2000}
-Goldin, C.
-(2000).
-Meritocracy and {Economic Inequality}.
-{Edited} by {Kenneth Arrow}, {Samuel Bowles}, and {Steven Durlauf} ( {Princeton}, {Princeton University Press}, 2000) 348 pp.
-\emph{Journal of Interdisciplinary History}, \emph{31}(3), 431a--433.
-\url{https://doi.org/10.1162/jinh.2000.31.3.431a}
+Goldin, Claudia. 2000. {``Meritocracy and {Economic Inequality}.
+{Edited} by {Kenneth Arrow}, {Samuel Bowles}, and {Steven Durlauf} (
+{Princeton}, {Princeton University Press}, 2000) 348 Pp.''}
+\emph{Journal of Interdisciplinary History} 31 (3): 431a--433.
+\url{https://doi.org/10.1162/jinh.2000.31.3.431a}.
 
 \bibitem[\citeproctext]{ref-hadjar_meritokratie_2008}
-Hadjar, A. (2008). \emph{{Meritokratie als Legitimationsprinzip}}. VS Verlag f{ü}r Sozialwissenschaften.
+Hadjar, Andreas. 2008. \emph{{Meritokratie als Legitimationsprinzip}}.
+Wiesbaden: VS Verlag f{ü}r Sozialwissenschaften.
 
 \bibitem[\citeproctext]{ref-hadler_why_2005}
-Hadler, M.
-(2005).
-Why {Do People Accept Different Income Ratios}?: {A Multi-level Comparison} of {Thirty Countries}.
-\emph{Acta Sociologica}, \emph{48}(2), 131--154.
-\url{https://doi.org/10.1177/0001699305053768}
+Hadler, Markus. 2005. {``Why {Do People Accept Different Income
+Ratios}?: {A Multi-level Comparison} of {Thirty Countries}.''}
+\emph{Acta Sociologica} 48 (2): 131--54.
+\url{https://doi.org/10.1177/0001699305053768}.
 
 \bibitem[\citeproctext]{ref-hankivsky_introduction_2022}
-Hankivsky, O., Morrow, M., \& Varcoe, C.
-(2022).
-{INTRODUCTION Women}'s {Health} in {Canada}: {Critical Intersectional Perspectives} on {Theory} and {Policy}.
-In M.
-Morrow, O.
-Hankivsky, \& C.
-Varcoe (Eds.), \emph{Women's {Health} in {Canada}} (pp.
-1--10).
-University of Toronto Press.
-\url{https://doi.org/10.3138/9781442623958-003}
-
-\bibitem[\citeproctext]{ref-hing_failure_2019}
-Hing, L.
-S.
-S., Wilson, A.
-E., Gourevitch, P., English, J., \& Sin, P.
-(2019).
-Failure to {Respond} to {Rising Income Inequality}: {Processes That Legitimize Growing Disparities}.
-\emph{Daedalus}, \emph{148}(3), 105--135.
-\url{https://doi.org/10.1162/daed_a_01752}
+Hankivsky, Olena, Marina Morrow, and Colleen Varcoe. 2022.
+{``{INTRODUCTION Women}'s {Health} in {Canada}: {Critical Intersectional
+Perspectives} on {Theory} and {Policy}.''} In \emph{Women's {Health} in
+{Canada}}, edited by Marina Morrow, Olena Hankivsky, and Colleen Varcoe,
+1--10. University of Toronto Press.
+\url{https://doi.org/10.3138/9781442623958-003}.
 
 \bibitem[\citeproctext]{ref-igliozziFairShareEffects2024}
-Igliozzi, D., Granot, Y., \& Ottati, V.
-(2024).
-A fair share: {Effects} of disparity, allocation strategy and system justification on perceptions of policy support in the education domain.
-\emph{European Journal of Social Psychology}, ejsp.3040.
-\url{https://doi.org/10.1002/ejsp.3040}
+Igliozzi, David, Yael Granot, and Victor Ottati. 2024. {``A Fair Share:
+{Effects} of Disparity, Allocation Strategy and System Justification on
+Perceptions of Policy Support in the Education Domain.''} \emph{European
+Journal of Social Psychology}, February, ejsp.3040.
+\url{https://doi.org/10.1002/ejsp.3040}.
 
 \bibitem[\citeproctext]{ref-iturra_percepcion_2023}
-Iturra, J., Castillo, J.
-C., Rufs Orellana, C., \& Maldonado, L.
-(2023).
-La percepci{ó}n de desigualdad econ{ó}mica y su influencia sobre la justificaci{ó}n de las diferencias de ingreso leg{í}timas.
-\emph{Estudios Sociol{ó}gicos de El Colegio de M{é}xico}, \emph{41}(122), 309--352.
-\url{https://doi.org/10.24201/es.2023v41n122.2260}
+Iturra, Julio, Juan Carlos Castillo, Catalina Rufs Orellana, and Luis
+Maldonado. 2023. {``La Percepci{ó}n de Desigualdad Econ{ó}mica y Su
+Influencia Sobre La Justificaci{ó}n de Las Diferencias de Ingreso
+Leg{í}timas.''} \emph{Estudios Sociol{ó}gicos de El Colegio de M{é}xico}
+41 (122): 309--52. \url{https://doi.org/10.24201/es.2023v41n122.2260}.
 
 \bibitem[\citeproctext]{ref-janmaatSubjectiveInequalityReview2013}
-Janmaat, J.
-G.
-(2013).
-Subjective inequality: {A} review of international comparative studies on people's views about inequality.
-\emph{Archives Europeennes de Sociologie}, \emph{54}(3), 357--389.
-\url{https://doi.org/10.1017/S0003975613000209}
+Janmaat, Jan Germen. 2013. {``Subjective Inequality: {A} Review of
+International Comparative Studies on People's Views about Inequality.''}
+\emph{Archives Europeennes de Sociologie} 54 (3): 357--89.
+\url{https://doi.org/10.1017/S0003975613000209}.
 
 \bibitem[\citeproctext]{ref-jassoJusticeEarningsNew1978}
-Jasso, G.
-(1978).
-On the {Justice} of {Earnings}: {A New Specification} of the {Justice Evaluation Function}.
-\emph{American Journal of Sociology}, \emph{83}(6), 1398--1419.
-\url{https://www.jstor.org/stable/2778110}
-
-\bibitem[\citeproctext]{ref-joikomujicaCuasimercadoEducativoChile2019}
-Joiko Mujica, S. (2019). {El cuasi-mercado educativo en Chile: desarrollo y consecuencias}. \emph{Revista Electr{ó}nica Di{á}logos Educativos; Vol. 12 N{ú}m. 23 (2012); 148-174}.
+Jasso, Guillermina. 1978. {``On the {Justice} of {Earnings}: {A New
+Specification} of the {Justice Evaluation Function}.''} \emph{American
+Journal of Sociology} 83 (6): 1398--1419.
+\url{https://www.jstor.org/stable/2778110}.
+
+\bibitem[\citeproctext]{ref-joiko_cuasimercado_2019}
+Joiko, Sara. 2019. {``{El cuasi-mercado educativo en Chile: desarrollo y
+consecuencias}.''} \emph{Revista Electr{ó}nica Di{á}logos Educativos;
+Vol. 12 N{ú}m. 23 (2012); 148-174}, April.
 
 \bibitem[\citeproctext]{ref-jost_theory_2020}
-Jost, J. (2020). \emph{A theory of system justification}. Harvard University Press.
+Jost, John. 2020. \emph{A Theory of System Justification}. Cambridge,
+Massachusetts: Harvard University Press.
 
 \bibitem[\citeproctext]{ref-jost_psychology_2003}
-Jost, J., \& Hunyady, O.
-(2003).
-The psychology of system justification and the palliative function of ideology.
-\emph{European Review of Social Psychology}, \emph{13}(1), 111--153.
-\url{https://doi.org/10.1080/10463280240000046}
+Jost, John, and Orsolya Hunyady. 2003. {``The Psychology of System
+Justification and the Palliative Function of Ideology.''} \emph{European
+Review of Social Psychology} 13 (1): 111--53.
+\url{https://doi.org/10.1080/10463280240000046}.
 
 \bibitem[\citeproctext]{ref-kelleyLegitimationInequalityOccupational1993}
-Kelley, J., \& Evans, M.
-D.
-R.
-(1993).
-The legitimation of inequality: {Occupational} earnings in nine nations.
-\emph{American Journal of Sociology}, \emph{99}(1), 75--125.
-\url{https://www.jstor.org/stable/2781956}
+Kelley, Jonathan, and M. D. R. Evans. 1993. {``The Legitimation of
+Inequality: {Occupational} Earnings in Nine Nations.''} \emph{American
+Journal of Sociology} 99 (1): 75--125.
+\url{https://www.jstor.org/stable/2781956}.
 
 \bibitem[\citeproctext]{ref-kelley_legitimate_2021}
-Kelley, J., \& Evans, M.
-D.
-R.
-(2021).
-Legitimate earnings inequality and national welfare commitment: {Correspondence} between economic institutions and the pay 80,000+ people in 30 nations think legitimate for ordinary jobs and for elite jobs.
-\emph{Social Science Research}, \emph{94}, 102446.
-\url{https://doi.org/10.1016/j.ssresearch.2020.102446}
+---------. 2021. {``Legitimate Earnings Inequality and National Welfare
+Commitment: {Correspondence} Between Economic Institutions and the Pay
+80,000+ People in 30 Nations Think Legitimate for Ordinary Jobs and for
+Elite Jobs.''} \emph{Social Science Research} 94 (February): 102446.
+\url{https://doi.org/10.1016/j.ssresearch.2020.102446}.
 
 \bibitem[\citeproctext]{ref-kluegel_social_1995}
-Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995). \emph{Social justice and political change: Public opinion in capitalist and post-communist states}. A. de Gruyter.
+Kluegel, James R., David S. Mason, and Bernd Wegener, eds. 1995.
+\emph{Social Justice and Political Change: Public Opinion in Capitalist
+and Post-Communist States}. Social Institutions and Social Change. New
+York: A. de Gruyter.
 
 \bibitem[\citeproctext]{ref-kluegelBeliefsStratification1981}
-Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}. \emph{Annual Review of Sociology}, 29--56.
+Kluegel, James R., and Eliot R. Smith. 1981. {``Beliefs {About
+Stratification}.''} \emph{Annual Review of Sociology}, 29--56.
 
 \bibitem[\citeproctext]{ref-kluegelBeliefsInequalityAmericans1986}
-Kluegel, J. R., \& Smith, E. R. (1986). \emph{Beliefs about {Inequality}: {Americans}' {Views} of {What Is} and {What Ought} to {Be}} (1st ed.). Routledge.
+---------. 1986. \emph{Beliefs about {Inequality}: {Americans}' {Views}
+of {What Is} and {What Ought} to {Be}}. 1st ed. Routledge.
 
 \bibitem[\citeproctext]{ref-leeFairnessPerceptionsEducational2023}
-Lee, J.-S., \& Stacey, M.
-(2023).
-Fairness perceptions of educational inequality: The effects of self-interest and neoliberal orientations.
-\emph{The Australian Educational Researcher}.
-\url{https://doi.org/10.1007/s13384-023-00636-6}
+Lee, Jung-Sook, and Meghan Stacey. 2023. {``Fairness Perceptions of
+Educational Inequality: The Effects of Self-Interest and Neoliberal
+Orientations.''} \emph{The Australian Educational Researcher}, May.
+\url{https://doi.org/10.1007/s13384-023-00636-6}.
 
 \bibitem[\citeproctext]{ref-liebigSociologyJustice2016}
-Liebig, S., \& Sauer, C. (2016). Sociology of {Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social Justice Theory} and {Research}} (pp. 37--59). Springer New York.
+Liebig, Stefan, and Carsten Sauer. 2016. {``Sociology of {Justice}.''}
+In \emph{Handbook of {Social Justice Theory} and {Research}}, edited by
+Clara Sabbagh and Manfred Schmitt, 37--59. New York, NY: Springer New
+York.
 
 \bibitem[\citeproctext]{ref-lindhPublicOpinionMarkets2015}
-Lindh, A.
-(2015).
-Public {Opinion} against {Markets}?
-{Attitudes} towards {Market Distribution} of {Social Services} -- {A Comparison} of 17 {Countries}.
-\emph{Social Policy \& Administration}, \emph{49}(7), 887--910.
-\url{https://doi.org/10.1111/spol.12105}
+Lindh, Arvid. 2015. {``Public {Opinion} Against {Markets}? {Attitudes}
+Towards {Market Distribution} of {Social Services} -- {A Comparison} of
+17 {Countries}.''} \emph{Social Policy \& Administration} 49 (7):
+887--910. \url{https://doi.org/10.1111/spol.12105}.
 
 \bibitem[\citeproctext]{ref-McCoy2007}
-McCoy, S.
-K., \& Major, B.
-(2007).
-Priming meritocracy and the psychological justification of inequality.
-\emph{Journal of Experimental Social Psychology}, \emph{43}(3), 341--351.
-\url{https://doi.org/10.1016/j.jesp.2006.04.009}
+McCoy, Shannon K., and Brenda Major. 2007. {``Priming Meritocracy and
+the Psychological Justification of Inequality.''} \emph{Journal of
+Experimental Social Psychology} 43 (3): 341--51.
+\url{https://doi.org/10.1016/j.jesp.2006.04.009}.
 
 \bibitem[\citeproctext]{ref-mcnamee_meritocracy_2004}
-McNamee, S., \& Miller, R. (2004). \emph{The meritocracy myth} (2004th ed.). Rowman \& Littlefield.
+McNamee, Stephen, and Robert Miller. 2004. \emph{The Meritocracy Myth}.
+2004th ed. Lanham Md.: Rowman \& Littlefield.
 
 \bibitem[\citeproctext]{ref-mijs_stratified_2016}
-Mijs, J.
-(2016).
-Stratified {Failure}: {Educational Stratification} and {Students}' {Attributions} of {Their Mathematics Performance} in 24 {Countries}.
-\emph{Sociology of Education}, \emph{89}(2), 137--153.
-\url{https://doi.org/10.1177/0038040716636434}
+Mijs, Jonathan. 2016. {``Stratified {Failure}: {Educational
+Stratification} and {Students}' {Attributions} of {Their Mathematics
+Performance} in 24 {Countries}.''} \emph{Sociology of Education} 89 (2):
+137--53. \url{https://doi.org/10.1177/0038040716636434}.
 
 \bibitem[\citeproctext]{ref-mijs_visualizing_2018}
-Mijs, J. (2018). Visualizing {Belief} in {Meritocracy}, 1930--2010. \emph{Socius}, \emph{4}, 2378023118811805. \url{https://doi.org/10.1177/2378023118811805}
+---------. 2018. {``Visualizing {Belief} in {Meritocracy},
+1930--2010.''} \emph{Socius} 4 (January): 2378023118811805.
+\url{https://doi.org/10.1177/2378023118811805}.
 
 \bibitem[\citeproctext]{ref-mijs_paradox_2021}
-Mijs, J. (2021). The paradox of inequality: Income inequality and belief in meritocracy go hand in hand. \emph{Socio-Economic Review}, \emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
+---------. 2021. {``The Paradox of Inequality: Income Inequality and
+Belief in Meritocracy Go Hand in Hand.''} \emph{Socio-Economic Review}
+19 (1): 7--35. \url{https://doi.org/10.1093/ser/mwy051}.
 
 \bibitem[\citeproctext]{ref-mijs_belief_2022}
-Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J.
-(2022).
-Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of {Subjective Social Mobility}.
-\emph{Social Psychology Quarterly}, \emph{85}(2), 131--141.
-\url{https://doi.org/10.1177/01902725211063818}
+Mijs, Jonathan, Stijn Daenekindt, Willem de Koster, and Jeroen van der
+Waal. 2022. {``Belief in {Meritocracy Reexamined}: {Scrutinizing} the
+{Role} of {Subjective Social Mobility}.''} \emph{Social Psychology
+Quarterly} 85 (2): 131--41.
+\url{https://doi.org/10.1177/01902725211063818}.
 
 \bibitem[\citeproctext]{ref-oecd_pisa_2023}
-OECD. (2023). \emph{{PISA} 2022 {Results} ({Volume I}): {The State} of {Learning} and {Equity} in {Education}}. OECD. \url{https://doi.org/10.1787/53f23881-en}
+OECD. 2023. \emph{{PISA} 2022 {Results} ({Volume I}): {The State} of
+{Learning} and {Equity} in {Education}}. {PISA}. OECD.
+\url{https://doi.org/10.1787/53f23881-en}.
 
 \bibitem[\citeproctext]{ref-osberg_fair_2006}
-Osberg, L., \& Smeeding, T.
-(2006).
-{``{Fair}''} {Inequality}?
-{Attitudes} toward {Pay Differentials}: {The United States} in {Comparative Perspective}.
-\emph{American Sociological Review}, \emph{71}(3), 450--473.
-\url{https://doi.org/10.1177/000312240607100305}
+Osberg, Lars, and Timothy Smeeding. 2006. {``{`{Fair}'} {Inequality}?
+{Attitudes} Toward {Pay Differentials}: {The United States} in
+{Comparative Perspective}.''} \emph{American Sociological Review} 71
+(3): 450--73. \url{https://doi.org/10.1177/000312240607100305}.
 
 \bibitem[\citeproctext]{ref-palanza_chile_2023}
-Palanza, V., \& Sotomayor, P.
-(2023).
-Chile's failed constitutional intent: {Polarization}, fragmentation, haste and delegitimization.
-\emph{Global Constitutionalism}, 1--10.
-\url{https://doi.org/10.1017/S204538172300028X}
-
-\bibitem[\citeproctext]{ref-ramosarellanoEducacionPromesaIncumplida2022}
-Ramos Arellano, M.
-(Ed.).
-(2022).
-\emph{Educaci{ó}n: {La} promesa incumplida: Esfuerzo, miedos y esperanzas de familias chilenas en el mercado escolar} (Primera edici{ó}n).
-Catalonia : CIAE, Centro de Investigaci{ó}n Avanzada en Educaci{ó}n, Universidad de Chile.
+Palanza, Valeria, and Patricia Sotomayor. 2023. {``Chile's Failed
+Constitutional Intent: {Polarization}, Fragmentation, Haste and
+Delegitimization.''} \emph{Global Constitutionalism}, September, 1--10.
+\url{https://doi.org/10.1017/S204538172300028X}.
+
+\bibitem[\citeproctext]{ref-ramos_educacion_2022}
+Ramos, Marcela, ed. 2022. \emph{Educaci{ó}n: {La} Promesa Incumplida:
+Esfuerzo, Miedos y Esperanzas de Familias Chilenas En El Mercado
+Escolar}. Primera edici{ó}n. Santiago, Chile: Catalonia : CIAE, Centro
+de Investigaci{ó}n Avanzada en Educaci{ó}n, Universidad de Chile.
 
 \bibitem[\citeproctext]{ref-Resh2016}
-Resh, N., \& Sabbagh, C. (2016). Justice and education. In \emph{Handbook of Social Justice Theory and Research}. \url{https://doi.org/10.1007/978-1-4939-3216-0_19}
+Resh, Nura, and Clara Sabbagh. 2016. \emph{Justice and Education}.
+\emph{Handbook of Social Justice Theory and Research}.
+\url{https://doi.org/10.1007/978-1-4939-3216-0_19}.
 
 \bibitem[\citeproctext]{ref-rosvall_mapping_2010}
-Rosvall, M., \& Bergstrom, C. T. (2010). Mapping {Change} in {Large Networks}. \emph{PLoS ONE}, \emph{5}(1), e8694. \url{https://doi.org/10.1371/journal.pone.0008694}
+Rosvall, Martin, and Carl T. Bergstrom. 2010. {``Mapping {Change} in
+{Large Networks}.''} Edited by Fabio Rapallo. \emph{PLoS ONE} 5 (1):
+e8694. \url{https://doi.org/10.1371/journal.pone.0008694}.
 
 \bibitem[\citeproctext]{ref-sandel_tyranny_2020}
-Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the common good?} (First edition). {Farrar, Straus and Giroux}.
+Sandel, Michael J. 2020. \emph{The Tyranny of Merit: {What}'s Become of
+the Common Good?} First edition. New York: {Farrar, Straus and Giroux}.
 
 \bibitem[\citeproctext]{ref-sazo_chile_2023}
-Sazo, D. (2023). Chile 2022: {From} great expectations to rising pessimism. \emph{Revista de Ciencia Pol{í}tica (Santiago)}, \emph{ahead}. \url{https://doi.org/10.4067/s0718-090x2023005000118}
+Sazo, Diego. 2023. {``Chile 2022: {From} Great Expectations to Rising
+Pessimism.''} \emph{Revista de Ciencia Pol{í}tica (Santiago)}, no.
+ahead. \url{https://doi.org/10.4067/s0718-090x2023005000118}.
 
 \bibitem[\citeproctext]{ref-schroder_income_2017}
-Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for {Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47. \url{https://doi.org/10.1007/s11211-016-0276-8}
+Schröder, Martin. 2017. {``Is {Income Inequality Related} to {Tolerance}
+for {Inequality}?''} \emph{Social Justice Research} 30 (1): 23--47.
+\url{https://doi.org/10.1007/s11211-016-0276-8}.
 
 \bibitem[\citeproctext]{ref-singer_applied_2009}
-Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data analysis: Modeling change and event occurence}. Oxford University Press, Incorporated.
+Singer, Judith D., and John B. Willett. 2009. \emph{Applied Longitudinal
+Data Analysis: Modeling Change and Event Occurence}. New York: Oxford
+University Press, Incorporated.
 
 \bibitem[\citeproctext]{ref-sonhing_merit_2011}
-Son Hing, L.
-S., Ramona, D., Zanna, M.
-P., Garcia, D.
-M., Gee, S.
-S., \& Orazietti, K.
-(2011).
-The merit of meritocracy.
-\emph{Journal of Personality and Social Psychology}, \emph{101}(3), 433--450.
-\url{https://doi.org/10.1037/a0024618}
+Son Hing, Leanne S., D. Ramona, Mark P. Zanna, Donna M. Garcia,
+Stephanie S. Gee, and Katie Orazietti. 2011. {``The Merit of
+Meritocracy.''} \emph{Journal of Personality and Social Psychology} 101
+(3): 433--50. \url{https://doi.org/10.1037/a0024618}.
+
+\bibitem[\citeproctext]{ref-sonhing_failure_2019}
+Son Hing, Leanne S., Anne E. Wilson, Peter Gourevitch, Jaslyn English,
+and Parco Sin. 2019. {``Failure to {Respond} to {Rising Income
+Inequality}: {Processes That Legitimize Growing Disparities}.''}
+\emph{Daedalus} 148 (3): 105--35.
+\url{https://doi.org/10.1162/daed_a_01752}.
 
 \bibitem[\citeproctext]{ref-stoilovaFairnessEducationalOpportunities2023}
-Stoilova, R., \& Ilieva-Trichkova, P.
-(2023).
-Fairness of educational opportunities and income distribution: Gender-sensitive analysis in a {European} comparative perspective.
-\emph{International Journal of Sociology and Social Policy}, \emph{43}(1/2), 272--291.
-\url{https://doi.org/10.1108/IJSSP-02-2022-0065}
+Stoilova, Rumiana, and Petya Ilieva-Trichkova. 2023. {``Fairness of
+Educational Opportunities and Income Distribution: Gender-Sensitive
+Analysis in a {European} Comparative Perspective.''} \emph{International
+Journal of Sociology and Social Policy} 43 (1/2): 272--91.
+\url{https://doi.org/10.1108/IJSSP-02-2022-0065}.
 
 \bibitem[\citeproctext]{ref-trump_income_2018}
-Trump, K.-S.
-(2018).
-Income {Inequality Influences Perceptions} of {Legitimate Income Differences}.
-\emph{British Journal of Political Science}, \emph{48}(4), 929--952.
-\url{https://doi.org/10.1017/S0007123416000326}
+Trump, Kris-Stella. 2018. {``Income {Inequality Influences Perceptions}
+of {Legitimate Income Differences}.''} \emph{British Journal of
+Political Science} 48 (4): 929--52.
+\url{https://doi.org/10.1017/S0007123416000326}.
 
 \bibitem[\citeproctext]{ref-trump_when_2020}
-Trump, K.-S. (2020). When and why is economic inequality seen as fair. \emph{Current Opinion in Behavioral Sciences}, \emph{34}, 46--51. \url{https://doi.org/10.1016/j.cobeha.2019.12.001}
+---------. 2020. {``When and Why Is Economic Inequality Seen as Fair.''}
+\emph{Current Opinion in Behavioral Sciences} 34 (August): 46--51.
+\url{https://doi.org/10.1016/j.cobeha.2019.12.001}.
 
 \bibitem[\citeproctext]{ref-valantPoliticsAchievementGaps2016}
-Valant, J., \& Newark, D.
-A.
-(2016).
-The {Politics} of {Achievement Gaps}: {U}.{S}.
-{Public Opinion} on {Race-Based} and {Wealth-Based Differences} in {Test Scores}.
-\emph{Educational Researcher}, \emph{45}(6), 331--346.
-\url{https://doi.org/10.3102/0013189X16658447}
+Valant, Jon, and Daniel A. Newark. 2016. {``The {Politics} of
+{Achievement Gaps}: {U}.{S}. {Public Opinion} on {Race-Based} and
+{Wealth-Based Differences} in {Test Scores}.''} \emph{Educational
+Researcher} 45 (6): 331--46.
+\url{https://doi.org/10.3102/0013189X16658447}.
 
 \bibitem[\citeproctext]{ref-vandertoorn_twenty_2014}
-van der Toorn, J., \& Jost, J.
-(2014).
-Twenty years of system justification theory: {Introduction} to the special issue on {``{Ideology} and system justification processes.''} \emph{Group Processes \& Intergroup Relations}, \emph{17}(4), 413--419.
-\url{https://doi.org/10.1177/1368430214531509}
+van der Toorn, Jojanneke, and John Jost. 2014. {``Twenty Years of System
+Justification Theory: {Introduction} to the Special Issue on
+{`{Ideology} and System Justification Processes'}.''} \emph{Group
+Processes \& Intergroup Relations} 17 (4): 413--19.
+\url{https://doi.org/10.1177/1368430214531509}.
 
 \bibitem[\citeproctext]{ref-wegenerIllusionDistributiveJustice1987}
-Wegener, B. (1987). The {Illusion} of {Distributive Justice}. \emph{European Sociological Review}, \emph{3}(1), 1--13. \url{https://www.jstor.org/stable/522527}
+Wegener, Bernd. 1987. {``The {Illusion} of {Distributive Justice}.''}
+\emph{European Sociological Review} 3 (1): 1--13.
+\url{https://www.jstor.org/stable/522527}.
 
 \bibitem[\citeproctext]{ref-Wegener1990}
-Wegener, B. (1990). Equity, relative deprivation, and the value consensus paradox. \emph{Social Justice Research}, \emph{4}(1), 65--86. \url{https://doi.org/10.1007/BF01048536}
+---------. 1990. {``Equity, Relative Deprivation, and the Value
+Consensus Paradox.''} \emph{Social Justice Research} 4 (1): 65--86.
+\url{https://doi.org/10.1007/BF01048536}.
 
 \bibitem[\citeproctext]{ref-young_rise_1962}
-Young, M. (1962). \emph{The rise of the meritocracy}. Penguin Books.
+Young, M. 1962. \emph{The Rise of the Meritocracy}. Baltimore: Penguin
+Books.
 
@@
 
-\newpage
-
-\appendix
+\section*{(APPENDIX) Appendix}\label{appendix-appendix}
+\addcontentsline{toc}{section}{(APPENDIX) Appendix}
 
@@
 
-\begin{table}[!ht]
-\caption{\label{tab:multilevel-full} Multilevel model for the justification of economic inequality}
-\begin{center}
-\scalebox{0.6}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\
-\hline
-Intercept                           & $1.86^{***}$  & $1.98^{***}$  & $1.72^{***}$  & $2.22^{***}$  & $2.11^{***}$  & $2.04^{***}$  \\
-                                    & $(0.02)$      & $(0.04)$      & $(0.09)$      & $(0.10)$      & $(0.10)$      & $(0.10)$      \\
-Wave (Ref.= Wave 2016)              &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Wave 2017                     & $-0.15^{***}$ &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-\quad Wave 2018                     & $-0.06^{*}$   &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2019                     & $-0.07^{**}$  &               &               &               &               &               \\
-                                    & $(0.03)$      &               &               &               &               &               \\
-\quad Wave 2022                     & $0.21^{***}$  &               &               &               &               &               \\
-                                    & $(0.02)$      &               &               &               &               &               \\
-Wave                                &               & $-0.17^{***}$ & $-0.17^{***}$ & $-0.13^{***}$ & $-0.14^{***}$ & $-0.14^{***}$ \\
-                                    &               & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      & $(0.02)$      \\
-Wave$^2$                            &               & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  & $0.03^{***}$  \\
-                                    &               & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      & $(0.00)$      \\
-Education (Ref.=Primary)            &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad High School                   &               &               & $0.02$        & $0.03$        & $0.04$        & $0.05$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Technical                     &               &               & $-0.01$       & $-0.00$       & $0.01$        & $0.01$        \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Universitary                  &               &               & $-0.09$       & $-0.05$       & $-0.04$       & $-0.04$       \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-Household Income (Ref.= Quintile 1) &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Quintile 2                    &               &               & $0.07$        & $0.07$        & $0.07$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 3                    &               &               & $0.08$        & $0.08$        & $0.08$        & $0.08$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Quintile 4                    &               &               & $0.12^{*}$    & $0.13^{*}$    & $0.13^{*}$    & $0.13^{*}$    \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile 5                    &               &               & $0.19^{**}$   & $0.19^{**}$   & $0.19^{***}$  & $0.20^{***}$  \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-\quad Quintile (No information)     &               &               & $0.17^{*}$    & $0.17^{*}$    & $0.18^{*}$    & $0.18^{*}$    \\
-                                    &               &               & $(0.08)$      & $(0.08)$      & $(0.08)$      & $(0.08)$      \\
-Subjective Social Status            &               &               & $0.01$        & $0.01$        & $0.00$        & $0.00$        \\
-                                    &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-Age (Ref.= 18-29)                   &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Age 30-49                     &               &               & $0.02$        & $0.02$        & $0.02$        & $0.02$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 50-64                     &               &               & $0.07$        & $0.08$        & $0.08$        & $0.07$        \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Age 65 or more                &               &               & $0.19^{**}$   & $0.20^{**}$   & $0.19^{**}$   & $0.18^{**}$   \\
-                                    &               &               & $(0.06)$      & $(0.06)$      & $(0.06)$      & $(0.06)$      \\
-Gender (Ref. Male)                  &               &               & $-0.08^{*}$   & $-0.08^{*}$   & $-0.08^{*}$   & $-0.07^{*}$   \\
-                                    &               &               & $(0.03)$      & $(0.03)$      & $(0.03)$      & $(0.03)$      \\
-Pol. pos (Ref.= Left)               &               &               &               &               &               &               \\
-                                    &               &               &               &               &               &               \\
-\quad Center                        &               &               & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    & $0.12^{*}$    \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Right                         &               &               & $0.29^{***}$  & $0.28^{***}$  & $0.27^{***}$  & $0.27^{***}$  \\
-                                    &               &               & $(0.05)$      & $(0.05)$      & $(0.05)$      & $(0.05)$      \\
-\quad Does not identify             &               &               & $0.15^{***}$  & $0.15^{***}$  & $0.14^{***}$  & $0.14^{***}$  \\
-                                    &               &               & $(0.04)$      & $(0.04)$      & $(0.04)$      & $(0.04)$      \\
-Inequality perception               &               &               &               & $-0.12^{***}$ & $-0.12^{***}$ & $-0.12^{***}$ \\
-                                    &               &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
-Merit: Talent                       &               &               &               &               & $0.04^{***}$  & $0.01$        \\
-                                    &               &               &               &               & $(0.01)$      & $(0.01)$      \\
-Merit: Effort                       &               &               &               &               &               & $0.05^{***}$  \\
-                                    &               &               &               &               &               & $(0.01)$      \\
-\hline
-BIC                                 & $30911.29$    & $30907.80$    & $31053.70$    & $30942.30$    & $30940.78$    & $30936.16$    \\
-Num. obs.                           & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        & $8042$        \\
-Num. groups: Individuals            & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        & $1689$        \\
-Var: Individuals (Intercept)        & $0.19$        & $0.19$        & $0.18$        & $0.17$        & $0.17$        & $0.16$        \\
-Var: Residual                       & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.47$        & $0.46$        \\
-\hline
-\multicolumn{7}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.}}
-\end{tabular}
-}
-\label{table:coefficients}
-\end{center}
-\end{table}
+(\#tab:multilevel-full) Multilevel model for the justification of
+economic inequality
+
+~
+
+Model 1
+
+Model 2
+
+Model 3
+
+Model 4
+
+Model 5
+
+Model 6
+
+Intercept
+
+1.84***
+
+1.87***
+
+1.80***
+
+2.39***
+
+2.22***
+
+2.13***
+
+~
+
+(0.02)
+
+(0.03)
+
+(0.09)
+
+(0.10)
+
+(0.10)
+
+(0.10)
+
+Wave (Ref.= Wave 2016)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2017
+
+-0.14***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2018
+
+-0.04
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2019
+
+-0.04
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2022
+
+0.23***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~wave 2023
+
+0.28***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.03)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Wave
+
+~
+
+-0.08***
+
+-0.08***
+
+-0.05**
+
+-0.06**
+
+-0.06**
+
+~
+
+~
+
+(0.02)
+
+(0.02)
+
+(0.02)
+
+(0.02)
+
+(0.02)
+
+Wave\^{}2
+
+~
+
+0.02***
+
+0.02***
+
+0.01***
+
+0.01***
+
+0.01***
+
+~
+
+~
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+Education (Ref.=Primary)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~High School
+
+~
+
+~
+
+-0.01
+
+-0.01
+
+-0.00
+
+0.00
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Technical
+
+~
+
+~
+
+-0.09
+
+-0.07
+
+-0.06
+
+-0.05
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+~~~~~Universitary
+
+~
+
+~
+
+-0.15*
+
+-0.11
+
+-0.10
+
+-0.10
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+Household Income (Ref.= Quintile 1)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Quintile 2
+
+~
+
+~
+
+0.05
+
+0.04
+
+0.04
+
+0.04
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Quintile 3
+
+~
+
+~
+
+0.10
+
+0.09
+
+0.09
+
+0.10
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Quintile 4
+
+~
+
+~
+
+0.11
+
+0.11*
+
+0.11*
+
+0.11*
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Quintile 5
+
+~
+
+~
+
+0.15**
+
+0.15**
+
+0.16**
+
+0.16**
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+~~~~~Quintile (No information)
+
+~
+
+~
+
+0.16*
+
+0.17*
+
+0.18*
+
+0.17*
+
+~
+
+~
+
+~
+
+(0.08)
+
+(0.08)
+
+(0.08)
+
+(0.08)
+
+Subjective Social Status
+
+~
+
+~
+
+-0.00
+
+-0.00
+
+-0.01
+
+-0.01
+
+~
+
+~
+
+~
+
+(0.01)
+
+(0.01)
+
+(0.01)
+
+(0.01)
+
+Age (Ref.= 18-29)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Age 30-49
+
+~
+
+~
+
+-0.02
+
+-0.01
+
+-0.01
+
+-0.02
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Age 50-64
+
+~
+
+~
+
+0.03
+
+0.04
+
+0.03
+
+0.03
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Age 65 or more
+
+~
+
+~
+
+0.14*
+
+0.14*
+
+0.13*
+
+0.12*
+
+~
+
+~
+
+~
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+(0.06)
+
+Gender (Ref. Male)
+
+~
+
+~
+
+-0.10**
+
+-0.10**
+
+-0.09**
+
+-0.08*
+
+~
+
+~
+
+~
+
+(0.03)
+
+(0.03)
+
+(0.03)
+
+(0.03)
+
+Pol. pos (Ref.= Left)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Center
+
+~
+
+~
+
+0.08
+
+0.07
+
+0.07
+
+0.07
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Right
+
+~
+
+~
+
+0.28***
+
+0.26***
+
+0.25***
+
+0.25***
+
+~
+
+~
+
+~
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+(0.05)
+
+~~~~~Does not identify
+
+~
+
+~
+
+0.09*
+
+0.08
+
+0.08
+
+0.08
+
+~
+
+~
+
+~
+
+(0.04)
+
+(0.04)
+
+(0.04)
+
+(0.04)
+
+Inequality perception
+
+~
+
+~
+
+~
+
+-0.15***
+
+-0.14***
+
+-0.14***
+
+~
+
+~
+
+~
+
+~
+
+(0.01)
+
+(0.01)
+
+(0.01)
+
+Merit: Talent
+
+~
+
+~
+
+~
+
+~
+
+0.06***
+
+0.01
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.01)
+
+(0.01)
+
+Merit: Effort
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.08***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.01)
+
+BIC
+
+37741.49
+
+37740.80
+
+37897.19
+
+37723.09
+
+37697.94
+
+37668.38
+
+Num. obs.
+
+9519
+
+9519
+
+9519
+
+9519
+
+9519
+
+9519
+
+Num. groups: Individuals
+
+1701
+
+1701
+
+1701
+
+1701
+
+1701
+
+1701
+
+Var: Individuals (Intercept)
+
+0.18
+
+0.18
+
+0.17
+
+0.17
+
+0.16
+
+0.16
+
+Var: Residual
+
+0.53
+
+0.53
+
+0.53
+
+0.52
+
+0.52
+
+0.52
+
+*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
 

### 2024-12-16 production/prod_analysis-fix-t5.Rmd
change to quarto + add ola 7

@@
 # Df variables nivel 1 _________________________________________________________
-df_study1 <- df_study_long_t6_comunas %>% dplyr::select(ola,
+df_study1 <- df_study_long_t7_comunas %>% dplyr::select(ola,
                                                just_educ, des_perc, merit_talent, merit_effort,
@@
 df_study1$ola <- factor(df_study1$ola,
-         labels = c("2016","2017","2018","2019", "2022"))
+         labels = c("2016","2017","2018","2019", "2022", "2023"))
 df_study1$ola <- 
@@
 lab_study1 <- sjlabelled::get_label(df_study1) 
-df_study1 <- df_study1 %>% na.omit() %>% sjlabelled::set_label(lab_study1)
-
-df<- dfSummary(x = dplyr::select(df_study1, merit_talent, merit_effort, des_perc, educ, quintil1, ess, edad, sexo, pos_id),
+df_study1 <- df_study1 %>% sjlabelled::set_label(lab_study1)
+
+df<- dfSummary(x = dplyr::select(df_study1, ola, merit_talent, merit_effort, des_perc, educ, quintil1, ess, edad, sexo, pos_id),
                plain.ascii = FALSE,
@@
 load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
-df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
                                                just_educ, des_perc, merit_effort, merit_talent,
@@
               bg.colour='grey30')+
-  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022)) +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
+  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022, 2023)) +
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 9519)")
 
@@
 ```{r bivariate-plots}
-df_study1 <- df_study_long_t6_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
+df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
                                                just_educ, des_perc, merit_effort, merit_talent,
@@
         plot.caption = element_text(hjust = 1)) +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 9519)")
 merit_plot
@@
 ggsave(merit_plot, file="output/graphs/merit_plot.png", height = 6, width = 10)
-```
-
-```{r}
-
-range <- c(1, 1)
-
-p_just_educ<-
-  dplyr::select(df_study1,
-                just_educ) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_des_perc<-
-  dplyr::select(df_study1,
-                des_perc) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_effort<-
-  dplyr::select(df_study1,
-                merit_effort) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = T,
-                      reverse.scale = T,
-                      show.n = FALSE,
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5),
-  ) +
-  theme(legend.position = "none")
-
-p_talent<-
-  dplyr::select(df_study1,
-                merit_talent) %>%
-  sjPlot::plot_likert(geom.colors = "PuBu",
-                      geom.size = 0.8,
-                      catcount = 5,
-                      grid.range  =  range,
-                      show.prc.sign = T,
-                      reverse.colors = F,
-                      reverse.scale = T,
-                      show.n = FALSE, 
-                      intercept.line.color = "white")  +
-  ggplot2::theme(legend.position = "bottom",
-    text = ggplot2::element_text(size = 16))+
-  ggplot2::scale_fill_manual(
-    values = colors,
-    breaks = c(1,2,3,4,5), 
-    labels = c(
-      "Strongly disagree",
-      "Disagree",
-      "Neither disagree nor agree",
-      "Agree",
-      "Strongly agree"
-    )
-  )
-```
-
-##### Create plot
-```{r}
-pacman::p_load(cowplot)
-t <- 0.1
-s <- 0.45
-m <- 0.4
-xs <- 0.25
-
-caption <- 
-  stringr::str_wrap(
-    paste0("Note: Own elaboration based on ELSOC Survey ",
-           "(n=",dim(na.omit(df_study1))[1],"). "),
-    width = 100)
-
-title <- cowplot::ggdraw() + 
-  cowplot::draw_label(
-    "Frequencies",
-    fontface = 'bold',
-    size = 15,
-    x = 0,
-    hjust = 0
-  )
-
-likert1<-   
-  cowplot::add_sub(
-    cowplot::plot_grid(p_just_educ,
-                       p_des_perc,
-                       p_effort,
-                       p_talent,
-                       align = "hv",
-                       axis = "a",
-                       nrow = 7, #6 plots
-                       ncol = 1, #1 column   
-                       rel_heights = c(
-                                       s,s,s,s,
-                                       s,
-                                       s) #fig size within plot
-    ),
-    label = "", # Caption: source and sample size
-    size = 11,  # Caption font size
-    hjust = 0# Caption align 
-  ) %>% 
-  cowplot::ggdraw()
-jpeg(here::here("output/graphs/figure1.jpg"),
-    # units = "cm",
-    # res = 100,
-    width = 800,
-    height = 800)
-likert1
-dev.off()
-```
-
-justification education vs year
-
-```{r}
-just_educ_year = df_study1 %>%
-  group_by(ola) %>%
-  summarise(education = mean(just_educ, na.rm = T))
-
-categorias_x <- c(2016, 2017, 2018, 2019, 2022)
-
-educ_year_plot <- just_educ_year_survey %>% ggplot(aes(ola, education)) +
-  theme_bw(base_size = 14) +
-  geom_point(size = 4)+
-  ylab("Justification of inequality") +
-  xlab("Wave") +
-  scale_x_discrete(labels = categorias_x) +
-  ylim(1,3)
-educ_year_plot
-
-ggsave(educ_year_plot, file="output/graphs/educ_year_plot.png", width = 7, height = 7)
 ```
@@
 
-categorias_x <- c(2016, 2017, 2018, 2019, 2022)
+categorias_x <- c(2016, 2017, 2018, 2019, 2022, 2023)
 vars_years$variable <- factor(vars_years$variable, levels = c("education", "inequality", "talent", "effort"), labels=  c("Justification
@@
    theme(legend.position = "none") +
-  coord_cartesian(xlim = c(1, 6)) +  # Ampliar el eje x
-  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022, ""))  # Personalizar etiquetas del eje x
+  coord_cartesian(xlim = c(1, 7)) +  # Ampliar el eje x
+  scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022, 2023, ""))  # Personalizar etiquetas del eje x
 
@@
 years_plot <- years_plot +
-  geom_point(data = label_data, aes(label_x - 0.05, label_y, color = variable, shape = variable), size = 4) +  # Capa de puntos para los símbolos de forma
-  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 0.05, nudge_y = 0, size = 4)  +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 8042)")
+  geom_point(data = label_data, aes(label_x + 1, label_y, color = variable, shape = variable), size = 4) +  # Capa de puntos para los símbolos de forma
+  geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 1.1, nudge_y = 0, size = 4)  +
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 9519)")
 
@@
 ggsave("output/graphs/years_plot.png", plot = years_plot, height = 6, width = 7, dpi = 300)
-
-```
-
-```{r}
-# Calcular cuantiles para las variables
-
-boxplot <- df_study_long_t6_comunas %>% select(just_educ, des_perc, merit_talent, merit_effort, ola) %>% as.data.frame()
-
-boxplot <- boxplot %>% 
-  pivot_longer(cols = c("just_educ", 
-                        "des_perc", 
-                        "merit_talent", 
-                        "merit_effort"), 
-               names_to = "variable",
-               values_to = "value") %>% na.omit()
-
-boxplot$variable <- factor(boxplot$variable,
-                   levels = c("just_educ", 
-                        "des_perc", 
-                        "merit_talent", 
-                        "merit_effort"),
-                   labels = c("Justification
-educational
-inequality", "Inequality
-perception", "Merit: Talent", "Merit: Effort"))
-
-boxplot_plot <- boxplot %>%
-  ggplot(aes(x = as.factor(ola), y = value, fill=variable, color=variable)) + 
-  geom_boxplot() +
-  scale_y_continuous(name = "Value", limits = c(1, 5)) +
-  scale_x_discrete(name = "Wave", labels=categorias_x) +
-  labs(fill = "Variable", color = "Variable") +
-  theme_bw() +
-  theme(axis.text.x = element_text(margin = margin(b = 15)))
-boxplot_plot
-
-ggsave(boxplot_plot, file="output/graphs/boxplot_plot.png", height = 7, width = 7)
 ```
@@
 corrplot::corrplot(cor_full,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
+  method = "color", # Cambia los círculos por color completo de cada cuadrante
+  addCoef.col = "#000390", # Color de los coeficientes
+  type = "upper", # Deja solo las correlaciones de arriba
+  tl.col = "black", # COlor letras, rojo por defecto
   na.label = "-")
@@
 grid::grid.text("Source: own elaboration with pooled data 
-from ELSOC 2016-2022 (n = 8042)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
+from ELSOC 2016-2023 (n = 9519)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
 dev.off()
 # Wave 1_______________________________________________________________________
-mat_1<- 
-df_study1_long_comunas %>%
-  filter(ola==1) %>%
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_1 <- as.matrix(mat_1$rho)
-diag(cor_1) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_1) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_1,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 1")
-
-# Wave 2_______________________________________________________________________
-mat_2<-   
-df_study1_long_comunas %>%
-  filter(ola==2) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_2 <- as.matrix(mat_2$rho)
-diag(cor_2) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_2) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_2,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 2")
-
-# Wave 3_______________________________________________________________________
-mat_3 <- 
-df_study1_long_comunas %>%
-  filter(ola==3) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_3 <- as.matrix(mat_3$rho)
-diag(cor_3) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_3) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_3,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 3")
-
-# Wave 4_______________________________________________________________________
-mat_4 <- 
-df_study1_long_comunas %>%
-  filter(ola==4) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_4 <- as.matrix(mat_4$rho)
-diag(cor_4) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_4) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_4,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 4")
-
-# Wave 6_______________________________________________________________________
-mat_6 <- 
-df_study1_long_comunas %>%
-  filter(ola==6) %>% 
-  select(just_educ, # distributive justice
-         merit_effort, merit_talent, # meritocracy
-         ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
-         ) %>% 
-  psych::mixedCor()
-
-cor_6 <- as.matrix(mat_6$rho)
-diag(cor_6) = NA #set diagonal values to NA
-# Set Row names of the matrix
-rownames(cor_6) <- c("A. Education distributive justice",
-                     "B. People are rewarded for their efforts",
-                     "C. People are rewarded for their intelligence",
-                     "D. Go ahead: wealthy family",
-                     "E. Go ahead: Having ambition",
-                     "F. Go ahead: Good level of education",
-                     "G. Go ahead: Hard work")
-#set Column names of the matrix
-colnames(cor_6) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
-#Plot the matrix using corrplot
-corrplot::corrplot(cor_6,
-  method = "color",
-  addCoef.col = "#000390",
-  type = "upper",
-  tl.col = "black",
-  col=colorRampPalette(c("white","#0068DC"))(12),
-  bg = "white",
-  na.label = "-",
-  title = "\n Wave 6")
+# mat_1<- 
+# df_study1_long_comunas %>%
+#   filter(ola==1) %>%
+#   select(just_educ, # distributive justice
+#          merit_effort, merit_talent, # meritocracy
+#          ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# cor_1 <- as.matrix(mat_1$rho)
+# diag(cor_1) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_1) <- c("A. Education distributive justice",
+#                      "B. People are rewarded for their efforts",
+#                      "C. People are rewarded for their intelligence",
+#                      "D. Go ahead: wealthy family",
+#                      "E. Go ahead: Having ambition",
+#                      "F. Go ahead: Good level of education",
+#                      "G. Go ahead: Hard work")
+# #set Column names of the matrix
+# colnames(cor_1) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_1,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 1")
+# 
+# # Wave 2_______________________________________________________________________
+# mat_2<-   
+# df_study1_long_comunas %>%
+#   filter(ola==2) %>% 
+#   select(just_educ, # distributive justice
+#          merit_effort, merit_talent, # meritocracy
+#          ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# cor_2 <- as.matrix(mat_2$rho)
+# diag(cor_2) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_2) <- c("A. Education distributive justice",
+#                      "B. People are rewarded for their efforts",
+#                      "C. People are rewarded for their intelligence",
+#                      "D. Go ahead: wealthy family",
+#                      "E. Go ahead: Having ambition",
+#                      "F. Go ahead: Good level of education",
+#                      "G. Go ahead: Hard work")
+# #set Column names of the matrix
+# colnames(cor_2) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_2,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 2")
+# 
+# 
+# # Wave 3_______________________________________________________________________
+# mat_3 <- 
+# df_study1_long_comunas %>%
+#   filter(ola==3) %>% 
+#   select(just_educ, # distributive justice
+#          merit_effort, merit_talent, # meritocracy
+#          ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+#          ) %>% 
+#   psych::mixedCor()  
+# 
+# cor_3 <- as.matrix(mat_3$rho)
+# diag(cor_3) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_3) <- c("A. Education distributive justice",
+#                      "B. People are rewarded for their efforts",
+#                      "C. People are rewarded for their intelligence",
+#                      "D. Go ahead: wealthy family",
+#                      "E. Go ahead: Having ambition",
+#                      "F. Go ahead: Good level of education",
+#                      "G. Go ahead: Hard work")
+# #set Column names of the matrix
+# colnames(cor_3) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_3,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 3")
+# 
+# # Wave 4_______________________________________________________________________
+# mat_4 <- 
+# df_study1_long_comunas %>%
+#   filter(ola==4) %>% 
+#   select(just_educ, # distributive justice
+#          merit_effort, merit_talent, # meritocracy
+#          ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# 
+# cor_4 <- as.matrix(mat_4$rho)
+# diag(cor_4) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_4) <- c("A. Education distributive justice",
+#                      "B. People are rewarded for their efforts",
+#                      "C. People are rewarded for their intelligence",
+#                      "D. Go ahead: wealthy family",
+#                      "E. Go ahead: Having ambition",
+#                      "F. Go ahead: Good level of education",
+#                      "G. Go ahead: Hard work")
+# #set Column names of the matrix
+# colnames(cor_4) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_4,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 4")
+# 
+# 
+# # Wave 6_______________________________________________________________________
+# mat_6 <- 
+# df_study1_long_comunas %>%
+#   filter(ola==6) %>% 
+#   select(just_educ, # distributive justice
+#          merit_effort, merit_talent, # meritocracy
+#          ahead_family, ahead_ambition, ahead_educ, ahead_hardwork # go ahead
+#          ) %>% 
+#   psych::mixedCor() 
+# 
+# cor_6 <- as.matrix(mat_6$rho)
+# diag(cor_6) = NA #set diagonal values to NA
+# # Set Row names of the matrix
+# rownames(cor_6) <- c("A. Education distributive justice",
+#                      "B. People are rewarded for their efforts",
+#                      "C. People are rewarded for their intelligence",
+#                      "D. Go ahead: wealthy family",
+#                      "E. Go ahead: Having ambition",
+#                      "F. Go ahead: Good level of education",
+#                      "G. Go ahead: Hard work")
+# #set Column names of the matrix
+# colnames(cor_6) <-c("(A)", "(B)","(C)","(D)","(E)","(F)","(G)")
+# #Plot the matrix using corrplot
+# corrplot::corrplot(cor_6,
+#   method = "color",
+#   addCoef.col = "#000390",
+#   type = "upper",
+#   tl.col = "black",
+#   col=colorRampPalette(c("white","#0068DC"))(12),
+#   bg = "white",
+#   na.label = "-",
+#   title = "\n Wave 6")
 ```
@@
 
-```{r distribution-educ}
+```{r distribution-educ, eval=FALSE}
 rm(list=ls())
@@
 df_study1 <- 
-  df_study_long_t6_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total, just_educ, des_perc, merit_effort, merit_talent, educ, quintil1:sexo7) %>% 
+  df_study_long_t7_comunas %>%
+  select(idencuesta,ola,comuna, ponderador_long_total, just_educ, des_perc, merit_effort, merit_talent, educ, quintil1:sexo) %>% 
   na.omit() %>% 
@@
 
-```{r}
+```{r eval=FALSE}
 #rm(list=ls())
@@
                )
-load(file = here::here("input/data-proc/df_study_t6_indep_comunas.RData"))
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
 
@@
 df_study1 <- 
-  df_study_long_t6_comunas %>%
+  df_study_long_t7_comunas %>%
   select(idencuesta,ola,comuna,just_educ, merit_effort, merit_talent, ahead_family:educ,quintil1:sexo, ginilag22_17, gini_ypch_2022, gini_ypch, prop_univ, gini_gc, prop_pagado, pagadlag22_16=pagadolag22_16, total_mat_pagado, var_simce_comuna_rescaled, prom_simce_comuna_rescaled, univlag22_17, loginc_mean, logincomelag22_17) %>%
@@
 
-```{r}
+```{r eval=FALSE}
 # table 2: contextuales
@@
 df_study1 <- 
-  df_study_long_t6_comunas %>%
+  df_study_long_t7_comunas %>%
   select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
@@
 fit <- lmer(just_educ ~ ola + ola_2 + (ola + ola_2 | idencuesta), data = df_study1)
+
 df_study1$pred_inter_pend_aleatorio <- predict(fit)
@@
   theme_light()
-
+plot
 ggsave(plot, file="output/graphs/fixed_effects.png")
@@
 predict<-interplot::interplot(m = educ1,var1 = "ola",point = T)
-predict_data <- predict$data[c(1, 2, 4, 6, 8), ]
+predict_data <- predict$data[c(1, 2, 4, 6, 7, 8), ]
 predict_data <- predict_data %>% as.data.frame()
@@
                              "ola4",
-                             "ola6")
+                             "ola6",
+                        "ola7")
 

### 2024-12-16 production/prod_prep-mergedata.Rmd
change to quarto + add ola 7

--- 2024-01-26 production/prod_prep-mergedata.Rmd
@@
 load(file = here::here("input/data-proc/df_study2_long.RData"))
-load(file = here::here("input/data-proc/df_study1_long_t5.RData"))
+load(file = here::here("input/data-proc/df_study1_long_t7.RData"))
 load(file = here::here("input/data-proc/study_1_comunas.RData"))
@@
 load(file = here::here("input/data-proc/df_simce_comunas_long.RData")) # contiene las variables de Simce a nivel comunal promedio y la varianza dentro de la comuna
+```
 
+```{r}
 ## Juntar Gini y prop universitarios por comuna
@@
 df_gini_comunas_wide$comunacod <- as.numeric(df_gini_comunas_wide$comunacod)
+
+df_study1_comunas <- df_study1_comunas[!duplicated(df_study1_comunas$comuna), ]
 
@@
                                                                             ola==4~2022,
-                                                                            ola==6~2022))
+                                                                            ola==6~2022,
+                                                                            ola==7~2023))
 
@@
 
-df_study_long_t5_comunas <- 
- dplyr::left_join(x = df_study1_long_t5,
+df_study_long_t7_comunas <- 
+ dplyr::left_join(x = df_study1_long_t7,
                   y = df_study1_comunas, 
@@
  dplyr::rename(comuna=comuna.x)
+
+df_study_long_t7_comunas <- df_study_long_t7_comunas %>% mutate(year= case_when(ola==1~2017,
+                                                                            ola==2~2017,
+                                                                            ola==3~2017,
+                                                                            ola==4~2022,
+                                                                            ola==6~2022,
+                                                                            ola==7~2023))
+
+df_study_long_t7_comunas <- 
+ dplyr::left_join(x = df_study_long_t7_comunas,
+                  y = select(df_gini_comunas_long, comunacod=COMUNA_COD,year, starts_with("gini"), prop_univ, starts_with("inc"), starts_with("log")), 
+                  by = c("comunacod", "year"))
+
+df_study_long_t7_comunas <- 
+ dplyr::left_join(x = df_study_long_t7_comunas,
+                  y = select(df_gini_comunas_wide, comunacod,starts_with("gini"), starts_with("prop_univ"), starts_with("univlag"), starts_with("inc"), starts_with("log")), 
+                  by = c("comunacod"))
 
@@
                                                                             ola==4~2019,
-                                                                            ola==6~2022))
+                                                                            ola==6~2022,
+                                                                            ola==7~2023))
 
@@
                                                                             ola==5~2021,
-                                                                            ola==6~2022))
+                                                                            ola==6~2022,
+                                                                            ola==7~2023))
+
+df_study_long_t7_comunas <- df_study_long_t7_comunas %>% mutate(year= case_when(ola==1~2016,
+                                                                            ola==2~2017,
+                                                                            ola==3~2018,
+                                                                            ola==4~2019,
+                                                                            ola==5~2021,
+                                                                            ola==6~2022,
+                                                                            ola==7~2023))
 
@@
  dplyr::left_join(x = df_study2_long_comunas,
+                  y = select(df_matricula_comunas_wide, comunacod=cod_com_rbd_16,starts_with("total_mat_pagado"), starts_with("prop_pagado"), pagadolag22_16),
+                  by = c("comunacod"))
+
+df_study_long_t7_comunas <- 
+ dplyr::left_join(x = df_study_long_t7_comunas,
+                  y = select(df_matricula_comunas_long, comunacod=cod_com_rbd_15,year, total_mat_pagado,prop_pagado), 
+                  by = c("comunacod", "year"))
+
+df_study_long_t7_comunas <- 
+ dplyr::left_join(x = df_study_long_t7_comunas,
                   y = select(df_matricula_comunas_wide, comunacod=cod_com_rbd_16,starts_with("total_mat_pagado"), starts_with("prop_pagado"), pagadolag22_16), 
@@
 
+df_study_long_t7_comunas <- 
+ dplyr::left_join(x = df_study_long_t7_comunas,
+                  y = select(df_simce_comunas_long, comunacod=cod_com_15,year, prom_simce_comuna,var_simce_comuna), 
+                  by = c("comunacod", "year"))
+
+df_study_long_t7_comunas <- 
+ dplyr::left_join(x = df_study_long_t7_comunas,
+                  y = select(df_comuna_rbd_wide, comunacod=COMUNA,starts_with("prom_simce"), starts_with("var_simce")), 
+                  by = c("comunacod"))
+
 ## Rescale simce
@@
 df_study1_long_comunas$var_simce_comuna_rescaled <- df_study1_long_comunas$var_simce_comuna / 100
+df_study_long_t7_comunas$prom_simce_comuna_rescaled <- df_study_long_t7_comunas$prom_simce_comuna / 10
+df_study_long_t7_comunas$var_simce_comuna_rescaled <- df_study_long_t7_comunas$var_simce_comuna / 100
 # Verificar el rescalado
@@
      file = here::here("input/data-proc/df_study2_long_comunas.RData"))
-save(df_study_long_t5_comunas,
-     file = here::here("input/data-proc/df_study_t5_ind_comunas.RData"))
+save(df_study_long_t7_comunas,
+     file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
 
@@
 
-sjPlot::view_df(df_study_long_t5_comunas,
+sjPlot::view_df(df_study_long_t7_comunas,
         show.frq = T, 

### 2024-12-16 production/prod_prep.Rmd
change to quarto + add ola 7

@@
 
-file.exists(here::here("input/data-original/elsoc_long.Rdata"))
+#file.exists(here::here("input/data-original/elsoc_long.Rdata"))
 
-load(url("https://dataverse.harvard.edu/api/access/datafile/7245118")) #ELSOC
-save(elsoc_long_2016_2022.2,file = here::here("input/data-original/elsoc_long.Rdata"))
+#load(url("https://dataverse.harvard.edu/api/access/datafile/7245118")) #ELSOC
+#save(elsoc_long_2016_2023,file = here::here("input/data-original/elsoc_long.Rdata"))
 
@@
 
-elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-999] <- NA
-elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-888] <- NA
-elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-777] <- NA
-elsoc_long_2016_2022.2[elsoc_long_2016_2022.2 ==-666] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
 
 # Attitudes toward inmigrants__________________________________________________
-elsoc_long_2016_2022.2 <- 
-elsoc_long_2016_2022.2 %>% 
+elsoc_long_2016_2023 <- 
+elsoc_long_2016_2023 %>% 
   #create new variables
@@
 #
-sjmisc::frq(elsoc_long_2016_2022.2$just_educ)
-elsoc_long_2016_2022.2$just_educ <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$just_educ, 
+sjmisc::frq(elsoc_long_2016_2023$just_educ)
+elsoc_long_2016_2023$just_educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$just_educ, 
                       label = "Education distributive justice")
@@
 #
-elsoc_long_2016_2022.2$just_pension <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$just_pension, 
+elsoc_long_2016_2023$just_pension <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$just_pension, 
                       label = "Pension distributive justice")
@@
 #
-elsoc_long_2016_2022.2$just_salud <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$just_salud, 
+elsoc_long_2016_2023$just_salud <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$just_salud, 
                       label = "Health distributive justice")
@@
 #
-elsoc_long_2016_2022.2$des_perc <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$des_perc, 
+elsoc_long_2016_2023$des_perc <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$des_perc, 
                       label = "Income differences are too large")
@@
 #
-sjmisc::frq(elsoc_long_2016_2022.2$merit_effort)
-elsoc_long_2016_2022.2$merit_effort <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$merit_effort, 
+sjmisc::frq(elsoc_long_2016_2023$merit_effort)
+elsoc_long_2016_2023$merit_effort <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$merit_effort, 
                       label = "People are rewarded for their efforts")
@@
 #
-sjmisc::frq(elsoc_long_2016_2022.2$merit_talent)
-elsoc_long_2016_2022.2$merit_talent <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$merit_talent, 
+sjmisc::frq(elsoc_long_2016_2023$merit_talent)
+elsoc_long_2016_2023$merit_talent <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$merit_talent, 
                       label = "People are rewarded for their intelligence")
@@
 #
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_family)
-elsoc_long_2016_2022.2$ahead_family <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_family, 
+sjmisc::frq(elsoc_long_2016_2023$ahead_family)
+elsoc_long_2016_2023$ahead_family <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$ahead_family, 
                       label = "Coming from a wealthy family")
@@
 #
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_educ)
-elsoc_long_2016_2022.2$ahead_educ <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_educ, 
+sjmisc::frq(elsoc_long_2016_2023$ahead_educ)
+elsoc_long_2016_2023$ahead_educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$ahead_educ, 
                       label = "Having a good level of education")
@@
 #
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_ambition)
-elsoc_long_2016_2022.2$ahead_ambition <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_ambition, 
+sjmisc::frq(elsoc_long_2016_2023$ahead_ambition)
+elsoc_long_2016_2023$ahead_ambition <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$ahead_ambition, 
                       label = "Having ambition")
@@
 #
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_hardwork)
-elsoc_long_2016_2022.2$ahead_hardwork <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$ahead_hardwork, 
+sjmisc::frq(elsoc_long_2016_2023$ahead_hardwork)
+elsoc_long_2016_2023$ahead_hardwork <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$ahead_hardwork, 
                       label = "Hard work")
@@
             "ahead_family","ahead_educ","ahead_ambition", "ahead_hardwork")) {
-elsoc_long_2016_2022.2[[i]] <-
-sjlabelled::set_labels(x = elsoc_long_2016_2022.2[[i]],
+elsoc_long_2016_2023[[i]] <-
+sjlabelled::set_labels(x = elsoc_long_2016_2023[[i]],
                        labels = c("Strongly disagree","Disagree",
@@
 
-sjmisc::frq(elsoc_long_2016_2022.2$just_educ)
-sjmisc::frq(elsoc_long_2016_2022.2$merit_effort)
-sjmisc::frq(elsoc_long_2016_2022.2$merit_talent)
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_family)
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_educ)
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_ambition)
-sjmisc::frq(elsoc_long_2016_2022.2$ahead_hardwork)
+sjmisc::frq(elsoc_long_2016_2023$just_educ)
+sjmisc::frq(elsoc_long_2016_2023$merit_effort)
+sjmisc::frq(elsoc_long_2016_2023$merit_talent)
+sjmisc::frq(elsoc_long_2016_2023$ahead_family)
+sjmisc::frq(elsoc_long_2016_2023$ahead_educ)
+sjmisc::frq(elsoc_long_2016_2023$ahead_ambition)
+sjmisc::frq(elsoc_long_2016_2023$ahead_hardwork)
 # Socioeconomic_________________________________________________________________
@@
 # Education_______________________________________
-elsoc_long_2016_2022.2$educ <- 
-  car::recode(elsoc_long_2016_2022.2$m01,
+elsoc_long_2016_2023$educ <- 
+  car::recode(elsoc_long_2016_2023$m01,
               "c(1,2,3)=1;c(4,5)=2;c(6,7)=3;c(8,9,10)=4; c(-888,-999)=NA")
-elsoc_long_2016_2022.2$educ <-
-  factor(elsoc_long_2016_2022.2$educ,
+elsoc_long_2016_2023$educ <-
+  factor(elsoc_long_2016_2023$educ,
          labels = c("Primary","High school","Technical","Universitary"))
@@
 #reverse education, reference level is the highest level
-#elsoc_long_2016_2022.2$educ <- forcats::fct_rev(elsoc_long_2016_2022.2$educ)
+#elsoc_long_2016_2023$educ <- forcats::fct_rev(elsoc_long_2016_2023$educ)
 
-elsoc_long_2016_2022.2$educ <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$educ,
+elsoc_long_2016_2023$educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$educ,
                       label = "Education")
-sjmisc::frq(elsoc_long_2016_2022.2$educ)
+sjmisc::frq(elsoc_long_2016_2023$educ)
 
 #Recoding of education to years based on casen 2017.
-elsoc_long_2016_2022.2$educyear<- as.numeric(
-  car::recode(elsoc_long_2016_2022.2$m01, 
+elsoc_long_2016_2023$educyear<- as.numeric(
+  car::recode(elsoc_long_2016_2023$m01, 
               "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
@@
 
-elsoc_long_2016_2022.2$educyear <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$educyear,
+elsoc_long_2016_2023$educyear <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$educyear,
                       label = "Education in years")
 
-class(elsoc_long_2016_2022.2$educyear)
-sjmisc::frq(elsoc_long_2016_2022.2$educyear)
+class(elsoc_long_2016_2023$educyear)
+sjmisc::frq(elsoc_long_2016_2023$educyear)
 
@@
 #Impute midpoint of income ranges
-elsoc_long_2016_2022.2$m30_rec <-
-as.numeric(car::recode(elsoc_long_2016_2022.2$m30,
+elsoc_long_2016_2023$m30_rec <-
+as.numeric(car::recode(elsoc_long_2016_2023$m30,
            "1=110000;2=251000;3=305000;4=355000;5=400000;
@@
 #Impute midpoint of income ranges (2021)
-elsoc_long_2016_2022.2$m30b_rec <-
-as.numeric(car::recode(elsoc_long_2016_2022.2$m30b,
+elsoc_long_2016_2023$m30b_rec <-
+as.numeric(car::recode(elsoc_long_2016_2023$m30b,
            "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))
 
-sjmisc::frq(elsoc_long_2016_2022.2$m30b_rec)
+sjmisc::frq(elsoc_long_2016_2023$m30b_rec)
 
 #Recode DK/DA of Income to NA
-elsoc_long_2016_2022.2$m29_rec <-
-  as.numeric(car::recode(elsoc_long_2016_2022.2$m29,"c(-888,-999)=NA"))
+elsoc_long_2016_2023$m29_rec <-
+  as.numeric(car::recode(elsoc_long_2016_2023$m29,"c(-888,-999)=NA"))
 
 #replace NA of income with new imputed variable
-elsoc_long_2016_2022.2$m29_imp <- 
-  ifelse(test = !is.na(elsoc_long_2016_2022.2$m29_rec),
-         yes =  elsoc_long_2016_2022.2$m29_rec,
-         no =  elsoc_long_2016_2022.2$m30_rec)
-summary(elsoc_long_2016_2022.2$m29_imp)
+elsoc_long_2016_2023$m29_imp <- 
+  ifelse(test = !is.na(elsoc_long_2016_2023$m29_rec),
+         yes =  elsoc_long_2016_2023$m29_rec,
+         no =  elsoc_long_2016_2023$m30_rec)
+summary(elsoc_long_2016_2023$m29_imp)
 
-elsoc_long_2016_2022.2$m29_imp <- 
-  ifelse(test = is.na(elsoc_long_2016_2022.2$m29_imp),
-         yes =  elsoc_long_2016_2022.2$m30b_rec,
-         no =  elsoc_long_2016_2022.2$m29_imp)
-summary(elsoc_long_2016_2022.2$m29_imp)
+elsoc_long_2016_2023$m29_imp <- 
+  ifelse(test = is.na(elsoc_long_2016_2023$m29_imp),
+         yes =  elsoc_long_2016_2023$m30b_rec,
+         no =  elsoc_long_2016_2023$m29_imp)
+summary(elsoc_long_2016_2023$m29_imp)
 
 # deflate at each year's prices
-elsoc_long_2016_2022.2$deflactor <-
-  with(elsoc_long_2016_2022.2, case_when(
+elsoc_long_2016_2023$deflactor <-
+  with(elsoc_long_2016_2023, case_when(
     ola == 2016 ~ 113.88 / 123.82,
@@
 # N Household:
-elsoc_long_2016_2022.2 <-
-  elsoc_long_2016_2022.2 %>%
+elsoc_long_2016_2023 <-
+  elsoc_long_2016_2023 %>%
   mutate(n_hogar =
@@
                             ola == 4 ~ m54,
-                            ola == 5 ~ m54))
-sjmisc::frq(elsoc_long_2016_2022.2$n_hogar)
+                            ola == 5 ~ m54,
+                            ola == 6 ~ m54,
+                            ola == 7 ~ m54))
+sjmisc::frq(elsoc_long_2016_2023$n_hogar)
 
 #Recode DK/DA to NA
-elsoc_long_2016_2022.2$n_hogar_r<-
-  car::recode(elsoc_long_2016_2022.2$n_hogar,"c(-888,-999)=NA")
+elsoc_long_2016_2023$n_hogar_r<-
+  car::recode(elsoc_long_2016_2023$n_hogar,"c(-888,-999)=NA")
 
 # Per capita household income:
-elsoc_long_2016_2022.2$ing_pc <- 
-  (elsoc_long_2016_2022.2$m29_imp/elsoc_long_2016_2022.2$n_hogar_r)
+elsoc_long_2016_2023$ing_pc <- 
+  (elsoc_long_2016_2023$m29_imp/elsoc_long_2016_2023$n_hogar_r)
 
-elsoc_long_2016_2022.2$ing_pc <-
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$ing_pc,
+elsoc_long_2016_2023$ing_pc <-
+sjlabelled::set_label(x = elsoc_long_2016_2023$ing_pc,
                       label = "Household income per capita")
 
-sjmisc::descr(elsoc_long_2016_2022.2$ing_pc)
+sjmisc::descr(elsoc_long_2016_2023$ing_pc)
 
 # Compute income quintiles
-elsoc_long_2016_2022.2 <- elsoc_long_2016_2022.2 %>% 
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
   group_by(ola) %>% 
@@
 
-elsoc_long_2016_2022.2$quintil <- 
-  factor(elsoc_long_2016_2022.2$quintil,
+elsoc_long_2016_2023$quintil <- 
+  factor(elsoc_long_2016_2023$quintil,
          levels = c(1, 2, 3, 4, 5),
@@
 #reverse quintile, reference level is the highest quintile
-#elsoc_long_2016_2022.2$quintil <- forcats::fct_rev(elsoc_long_2016_2022.2$quintil)
+#elsoc_long_2016_2023$quintil <- forcats::fct_rev(elsoc_long_2016_2023$quintil)
 
-elsoc_long_2016_2022.2$quintil <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$quintil,
+elsoc_long_2016_2023$quintil <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$quintil,
                       label = "Household income quintile per capita")
 
-sjmisc::frq(elsoc_long_2016_2022.2$quintil)
+sjmisc::frq(elsoc_long_2016_2023$quintil)
 
 #include new quintile category with missing cases
-elsoc_long_2016_2022.2$quintil1<-
-  car::recode(elsoc_long_2016_2022.2$quintil, 
+elsoc_long_2016_2023$quintil1<-
+  car::recode(elsoc_long_2016_2023$quintil, 
               "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")
 
-#elsoc_long_2016_2022.2$quintil1 <- factor(elsoc_long_2016_2022.2$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))
+#elsoc_long_2016_2023$quintil1 <- factor(elsoc_long_2016_2023$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))
 
-elsoc_long_2016_2022.2$quintil1 <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$quintil1,
+elsoc_long_2016_2023$quintil1 <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$quintil1,
                       label = "Household income quintile per capita (NA)") 
-sjmisc::frq(elsoc_long_2016_2022.2$quintil1)
+sjmisc::frq(elsoc_long_2016_2023$quintil1)
 
 #Age Ranges
-elsoc_long_2016_2022.2$edad <- 
-  factor(car::recode(elsoc_long_2016_2022.2$m0_edad, 
+elsoc_long_2016_2023$edad <- 
+  factor(car::recode(elsoc_long_2016_2023$m0_edad, 
                      "18:29=1;30:49=2;50:64=3;65:150=4"),
          labels = c('18-29', '30-49', '50-64', '65 or more'))
-elsoc_long_2016_2022.2$edad <-
-  sjlabelled::set_label(elsoc_long_2016_2022.2$edad, 
+elsoc_long_2016_2023$edad <-
+  sjlabelled::set_label(elsoc_long_2016_2023$edad, 
                         label = c("Age groups")) 
@@
 
-elsoc_long_2016_2022.2$ess <- 
-as.numeric(elsoc_long_2016_2022.2$d01_01)
-sjmisc::frq(elsoc_long_2016_2022.2$ess)
-#elsoc_long_2016_2022.2$ess <- sjmisc::rec(elsoc_long_2016_2022.2$ess,rec = "rev")
-#sjmisc::frq(elsoc_long_2016_2022.2$ess)
+elsoc_long_2016_2023$ess <- 
+as.numeric(elsoc_long_2016_2023$d01_01)
+sjmisc::frq(elsoc_long_2016_2023$ess)
+#elsoc_long_2016_2023$ess <- sjmisc::rec(elsoc_long_2016_2023$ess,rec = "rev")
+#sjmisc::frq(elsoc_long_2016_2023$ess)
 
-elsoc_long_2016_2022.2$ess <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$ess,
+elsoc_long_2016_2023$ess <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$ess,
                       label = "Subjective Social Status: individual")
 
-sjmisc::frq(elsoc_long_2016_2022.2$ess)
-# sjPlot::plot_frq(elsoc_long_2016_2022.2$ess)
+sjmisc::frq(elsoc_long_2016_2023$ess)
+# sjPlot::plot_frq(elsoc_long_2016_2023$ess)
 
 # Political position ____________________________
-elsoc_long_2016_2022.2$pos_id <-
+elsoc_long_2016_2023$pos_id <-
 factor(
   car::recode(
-    elsoc_long_2016_2022.2$c15,
+    elsoc_long_2016_2023$c15,
     "c(11,12,-888,-999)='Does not identify';c(0,1,2,3,4)='Left';
@@
 
-elsoc_long_2016_2022.2$pos_id <- factor(elsoc_long_2016_2022.2$pos_id,
-                                      levels = levels(elsoc_long_2016_2022.2$pos_id))
+elsoc_long_2016_2023$pos_id <- factor(elsoc_long_2016_2023$pos_id,
+                                      levels = levels(elsoc_long_2016_2023$pos_id))
 
-elsoc_long_2016_2022.2$pos_id <- 
-sjlabelled::set_label(x = elsoc_long_2016_2022.2$pos_id, 
+elsoc_long_2016_2023$pos_id <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$pos_id, 
                       label = "Political identification")
 
-sjmisc::frq(elsoc_long_2016_2022.2$pos_id)
+sjmisc::frq(elsoc_long_2016_2023$pos_id)
 
@@
 df_study1 <- 
-elsoc_long_2016_2022.2 %>% 
+elsoc_long_2016_2023 %>% 
   filter(muestra == 1) %>% #keep original sample (vs refresh)
@@
 df_study1_wide$edad_6 <-df_study1_wide$edad_1
+df_study1_wide$edad_7 <-df_study1_wide$edad_1
 
@@
 df_study1_wide$sexo_6 <-df_study1_wide$sexo_1
+df_study1_wide$sexo_7 <-df_study1_wide$sexo_1
 
@@
 df_study1_wide$educ_6 <-df_study1_wide$educ_1
+df_study1_wide$educ_7 <-df_study1_wide$educ_1
 
@@
 df_study1_wide$educyear_6 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_7 <-df_study1_wide$educyear_1
 
@@
 df_study1_wide$quintil1_6 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_7 <-df_study1_wide$quintil1_1
 
@@
 df_study1_wide$ess_6 <-df_study1_wide$ess_1
+df_study1_wide$ess_7 <-df_study1_wide$ess_1
 
@@
 df_study1_wide$pos_id_6 <-df_study1_wide$pos_id_1
+df_study1_wide$pos_id_7 <-df_study1_wide$pos_id_1
 
@@
 df_study1_wide$comuna_6 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_7 <-df_study1_wide$comuna_1
 
@@
 df_study1_wide$comunacod_6 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_7 <-df_study1_wide$comunacod_1
 
@@
 df_study1_long %>% 
-  filter(ola %in% c(1,2,3,4,6)) %>% 
-  mutate(ola=factor(ola,levels = 1:6,labels = 1:6))
+  filter(ola %in% c(1,2,3,4,6,7)) %>% 
+  mutate(ola=factor(ola,levels = 1:7,labels = 1:7))
 dim(df_study1_long) #check, now is OK
@@
 #______________________________________________________________________________
-# obtain the idencuesta for wave 6
+# obtain the idencuesta for wave 7
 ids <- 
@@
   select(idencuesta,ola) %>% 
-  filter(ola==6) %>% 
+  filter(ola==7) %>% 
   sjmisc::frq(idencuesta,show.na = F) %>% as.data.frame()
 
-# filter data by the idencuesta of t5
-df_study1_long_t6 <- 
+# filter data by the idencuesta of t7
+df_study1_long_t7 <- 
   df_study1_long %>%
@@
 
-names(df_study1_long_t6)
-dim(df_study1_long_t6)
-sjmisc::frq(df_study1_long_t6$ola)
+names(df_study1_long_t7)
+dim(df_study1_long_t7)
+sjmisc::frq(df_study1_long_t7$ola)
 
@@
 save(df_study1_long,file = here::here("input/data-proc/df_study1_long.RData"))
-save(df_study1_long_t6,file = here::here("input/data-proc/df_study1_long_t6.RData"))
+save(df_study1_long_t7,file = here::here("input/data-proc/df_study1_long_t7.RData"))
 save(df_study2_long,file = here::here("input/data-proc/df_study2_long.RData"))
@@
 
-sjPlot::view_df(df_study1_long_t6,
+sjPlot::view_df(df_study1_long_t7,
         show.frq = T,
@@
         show.na = T,
-        file = here::here("output/df_study1_long_t6.html"))
+        file = here::here("output/df_study1_long_t7.html"))
 ```
@@
 df_elsoc <- 
-elsoc_long_2016_2022.2 %>% 
+elsoc_long_2016_2023 %>% 
   select(comuna,comuna_cod) %>%
@@
 df_cit_a <- 
-elsoc_long_2016_2022.2 %>% 
+elsoc_long_2016_2023 %>% 
   select(comuna,comuna_cod,idencuesta) %>% 

### 2024-12-16 04-analisis.qmd
add covid vars
significant effect with:
- it is more important to give priority to economic activity than to the health of the population
- receive State benefits (Food Box, Emergency Family Income, Contribution for the Middle Class or other)

@@
   df_study_long_t7_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo, covid_economia, covid_beneficios) %>% 
   na.omit() %>% 
@@
          )
+
+df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
+
 # Hipotesis
@@
 h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+h7 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia"
+h8 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios"
 
@@
 educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ7 <- lmer(formula(paste0("just_educ~",h7,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ8 <- lmer(formula(paste0("just_educ~",h8,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
@@
 omit <- "(education)|(ess)"
-texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
+texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                                       "Merit: Talent",
-                                      "Merit: Effort" ),
+                                      "Merit: Effort" ,
+                                      "Covid: economy more important than health",
+                                      "Covid: receive state benefits"),
                 groups = list("Wave (Ref.= Wave 2016)" = 2:6,
@@
   df_study_long_t7_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo) %>% 
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo, covid_economia, covid_beneficios) %>% 
   na.omit() %>% 
@@
 
+df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
 # Hipotesis
 
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+merit_talent*ola"
+h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios"
+h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios+merit_effort*ola"
+h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios+merit_talent*ola"
 
@@
 # table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)"
+omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)|(covid_beneficios)|(covid_economia)"
 texreg::knitreg(list(educ6,educ3,educ1),

### 2024-12-16 production/prod_prep.Rmd
add covid vars
significant effect with:
- it is more important to give priority to economic activity than to the health of the population
- receive State benefits (Food Box, Emergency Family Income, Contribution for the Middle Class or other)

@@
          ahead_ambition=r12_04,
-         ahead_hardwork=r12_05) %>% 
+         ahead_hardwork=r12_05,
+         covid_conocidos=r03_09,
+         covid_economia=c37_09,
+         covid_cuarentena=c37_10,
+         covid_propio=s31,
+         covid_alimentos=m62,
+         covid_beneficios=m63_01,
+         covid_retirouno=m63_02,
+         covid_retirodos=m63_03
+         ) %>% 
   sjlabelled::drop_labels(., drop.na = FALSE)
@@
 sjmisc::frq(elsoc_long_2016_2023$pos_id)
+```
 
+```{r}
 # Select variables______________________________________________________________
@@
          ess,
+         covid_conocidos,
+         covid_economia,
+         covid_cuarentena,
+         covid_propio,
+         covid_alimentos,
+         covid_beneficios,
+         covid_retirouno,
+         covid_retirodos,
          # ess_inm,
@@
 
+df_study1_wide$covid_conocidos_1 <-df_study1_wide$covid_conocidos_5 #conocidos covid
+df_study1_wide$covid_conocidos_2 <-df_study1_wide$covid_conocidos_5
+df_study1_wide$covid_conocidos_3 <-df_study1_wide$covid_conocidos_5
+df_study1_wide$covid_conocidos_4 <-df_study1_wide$covid_conocidos_5
+df_study1_wide$covid_conocidos_6 <-df_study1_wide$covid_conocidos_5
+df_study1_wide$covid_conocidos_7 <-df_study1_wide$covid_conocidos_5
+
+df_study1_wide$covid_economia_1 <-df_study1_wide$covid_economia_5 #Importancia economia sobre salud
+df_study1_wide$covid_economia_2 <-df_study1_wide$covid_economia_5
+df_study1_wide$covid_economia_3 <-df_study1_wide$covid_economia_5
+df_study1_wide$covid_economia_4 <-df_study1_wide$covid_economia_5
+df_study1_wide$covid_economia_6 <-df_study1_wide$covid_economia_5
+df_study1_wide$covid_economia_7 <-df_study1_wide$covid_economia_5
+
+df_study1_wide$covid_cuarentena_1 <-df_study1_wide$covid_cuarentena_5 #importancia cuarentena
+df_study1_wide$covid_cuarentena_2 <-df_study1_wide$covid_cuarentena_5
+df_study1_wide$covid_cuarentena_3 <-df_study1_wide$covid_cuarentena_5
+df_study1_wide$covid_cuarentena_4 <-df_study1_wide$covid_cuarentena_5
+df_study1_wide$covid_cuarentena_6 <-df_study1_wide$covid_cuarentena_5
+df_study1_wide$covid_cuarentena_7 <-df_study1_wide$covid_cuarentena_5
+
+df_study1_wide$covid_propio_1 <-df_study1_wide$covid_propio_5 #Diagnostico covid propio
+df_study1_wide$covid_propio_2 <-df_study1_wide$covid_propio_5
+df_study1_wide$covid_propio_3 <-df_study1_wide$covid_propio_5
+df_study1_wide$covid_propio_4 <-df_study1_wide$covid_propio_5
+df_study1_wide$covid_propio_6 <-df_study1_wide$covid_propio_5
+df_study1_wide$covid_propio_7 <-df_study1_wide$covid_propio_5
+
+df_study1_wide$covid_alimentos_1 <-df_study1_wide$covid_alimentos_5 # Dificultades conseguir alimentos
+df_study1_wide$covid_alimentos_2 <-df_study1_wide$covid_alimentos_5
+df_study1_wide$covid_alimentos_3 <-df_study1_wide$covid_alimentos_5
+df_study1_wide$covid_alimentos_4 <-df_study1_wide$covid_alimentos_5
+df_study1_wide$covid_alimentos_6 <-df_study1_wide$covid_alimentos_5
+df_study1_wide$covid_alimentos_7 <-df_study1_wide$covid_alimentos_5
+
+df_study1_wide$covid_beneficios_1 <-df_study1_wide$covid_beneficios_5 #Recibe beneficios
+df_study1_wide$covid_beneficios_2 <-df_study1_wide$covid_beneficios_5
+df_study1_wide$covid_beneficios_3 <-df_study1_wide$covid_beneficios_5
+df_study1_wide$covid_beneficios_4 <-df_study1_wide$covid_beneficios_5
+df_study1_wide$covid_beneficios_6 <-df_study1_wide$covid_beneficios_5
+df_study1_wide$covid_beneficios_7 <-df_study1_wide$covid_beneficios_5
+
+df_study1_wide$covid_retirouno_1 <-df_study1_wide$covid_retirouno_5 #Primer retiro afp
+df_study1_wide$covid_retirouno_2 <-df_study1_wide$covid_retirouno_5
+df_study1_wide$covid_retirouno_3 <-df_study1_wide$covid_retirouno_5
+df_study1_wide$covid_retirouno_4 <-df_study1_wide$covid_retirouno_5
+df_study1_wide$covid_retirouno_6 <-df_study1_wide$covid_retirouno_5
+df_study1_wide$covid_retirouno_7 <-df_study1_wide$covid_retirouno_5
+
+df_study1_wide$covid_retirodos_1 <-df_study1_wide$covid_retirodos_5 #Segundo retiro afp
+df_study1_wide$covid_retirodos_2 <-df_study1_wide$covid_retirodos_5
+df_study1_wide$covid_retirodos_3 <-df_study1_wide$covid_retirodos_5
+df_study1_wide$covid_retirodos_4 <-df_study1_wide$covid_retirodos_5
+df_study1_wide$covid_retirodos_6 <-df_study1_wide$covid_retirodos_5
+df_study1_wide$covid_retirodos_7 <-df_study1_wide$covid_retirodos_5
+
 dim(df_study1_wide)
@@
                                  "^ahead_family_", "^ahead_educ_", "^ahead_ambition_", "^ahead_hardwork_",
+                                 "^covid_conocidos_", "^covid_economia_", "^covid_cuarentena_", "^covid_propio_",
+                                 "^covid_alimentos_", "^covid_beneficios_", "^covid_retirouno_", "^covid_retirodos_",
                                  "^educ_","^educyear_","^quintil1_","^ess_",
@@
                              "ahead_family","ahead_educ","ahead_ambition","ahead_hardwork",
+                             "covid_conocidos", "covid_economia", "covid_cuarentena", "covid_propio",
+                             "covid_alimentos", "covid_beneficios", "covid_retirouno", "covid_retirodos",
                              "educ","educyear","quintil1","ess",

### 2024-12-20 01-introduction.qmd
cambios intro+antecedentes

@@
 In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
+We examine this association in the context of COVID 19 in Chile, a period characterized by several challenges regarding the political and economic system that could have affected the way in which the role of individual performance and social outcomes are deemed.
 
@@
 This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.
+Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile during the COVID pandemic, and to what extent this phenomenon is associated with the perception of meritocracy and its variations.
 
@@
 Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.
+Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they would be the most affected in critical times.
+
+[[no sé si de pronto aclarar más la idea de que no es posible aislar el efecto COVID propiamente tal ...]]

### 2024-12-20 02-antecedentes.qmd
cambios intro+antecedentes

@@
-# Inequality, beliefs and justification
+# The justification of educational inequalities
 
@@
 
-The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016].
-From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice.
-Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p.
-30].
-However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality.
-In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p.
-2].
-This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978].
-Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.
-
- Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research.
-First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement.
-Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities).
-Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes).
-Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]
-
-An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986].
-The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021].
-Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014].
-Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020].
-As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].
-
-Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022].
-Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990].
-Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018].
-In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011].
-Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022].
-Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.
-
-# The justification of educational inequality
-
 Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences.
@@
 
-The study of inequality justification in the sphere of education is a rather specific and unattended field.
-Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities.
-This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area.
-This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).
-
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept.
-Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
-Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
-Although both questions have been used to measure the justification of educational inequality, they refer to different aspects.
-The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
-Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
-
-Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies.
-Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016].
-Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
+Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
+For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps.
 However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
 
-The perception of economic inequality is one of the factors associated with the justification of educational inequality.
+[[ver si esto queda acá o baja]] The perception of economic inequality is one of the factors associated with the justification of educational inequality.
 As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
@@
 
-Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles.
-In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
+In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items has been diverse, and there is not a single instrument or scale dedicated specifically to this concept.
+Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
+Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_).
+Although both questions have been used to measure the justification of educational inequality, they refer to different aspects: the ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive.
+Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
 
-_Justificacion, inequality and meritocracy in the educational domain_
+_Meritocracy and educational inequalities_
 
-The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research.
+The relationship between performance and outputs brings about to the concepts of merit and meritocracy, key in educational justice research.
 The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
@@
 
-Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
+Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find related research which incorporates the support of different distributive principles with regard to education.
+For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
+They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
+In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
+
+More specific empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
 Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
@@
 
-The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification.
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame.
+The present research aims at identifying changes in the justification of educational inequalities in Chile, with a particular focus in its relationship with meritocratic perceptions.
+The time frame for this analysis covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile).
+Although this period comprises the COVID pandemic, the changes in the variables analyzed are not possible to be attributed only to this event in terms of a health crisis, as several economic, social and cultural processes overlapped worldwide.
+Besides, in the specific case of Chile there were two constitutional processes happening in this very same period after a massive social upheaval.
+[[redondear-idea]]
+
+As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame.
 Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
-Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period.
+Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period.
 Consistently, our first hypothesis is:

### 2024-12-26 02-antecedentes.qmd
+ ajustes antecedentes

@@
 3], whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification [@trump_income_2018;@kelley_legitimate_2021;@mijs_paradox_2021].
-Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.
+Both theoretical and empirical approaches have been able to establish a clear distinction between inequality and justice, opening the question of to what extent inequality can be tolerated or even justified.
 
@@
 Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.
-
-Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps.
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
-
-[[ver si esto queda acá o baja]] The perception of economic inequality is one of the factors associated with the justification of educational inequality.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
-points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
-Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]
 
@@
 
-As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame.
+As there are measures before and after the pandemic period, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame.
 Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
@@
 
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:
+Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019].
+Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
+For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps.
+However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
+In line with this evidence, we expect that:
 
@@
 Thirdly, we address inequality perception.
-As referred to above, the justification of inequality requires, first and foremost, that such inequality is recognized.
+As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
+For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
+points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
+Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education.
+Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018].
 As a larger inequality in this domain threatens a basic equality ideal, the corresponding hypothesis in this case is:
@@
 Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged.
+As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged in the time frame analyzed.
 As a consequence, the justification of inequality could show a decrease over time:

### 2024-12-26 03-methods.qmd
changes methods + analisis

@@
 
-The main data source is the Chilean Longitudinal Social Survey 2016–2022.
+The main data source is the Chilean Longitudinal Social Survey 2016–2023.
 The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
@@
 Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728.
+In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1737.
 The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
@@
 This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish).
-Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5).
+Here, the respondents declared their preferences on a Likert scale from (1) "strongly disagree" to (5) “strongly agree” (5).
 The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills").
@@
 
+A third group of variables is included in the analysis to assess the effect of elements related to the pandemic.
+These variables were only included in the 2021 wave, so each individual's responses were replicated for the remaining years.
+The first variable addresses the belief regarding whether, in the context of the pandemic, it is more important to give priority to economic activity than to the health of the population, measured on a Likert scale from (1) "strongly disagree" to (2) "strongly agree".
+The second variable is a dummy (No/Yes) that captures whether individuals received state benefits during the pandemic (e.g., Food Box, Emergency Family Income, Contribution for the Middle Class, or others).
+
 For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
-Finally, political identification on a left-right scale, age, and gender were included as control variables.
+Finally, political identification on a left-right scale, age groups, and gender were included as control variables.
 @tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.

### 2024-12-26 04-analisis.qmd
changes methods + analisis

@@
 #| label: fig-alluvial
-#| fig-cap: "Change in the justification of educational inequality over time (2016-2022)"
+#| fig-cap: "Change in the justification of educational inequality over time (2016-2023)"
 #| 
@@
 
-@fig-alluvial illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022.
+@fig-alluvial illustrates yearly frequencies in the justification of educational inequality between 2016 and 2023.
 Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
+For instance, of the 32.5% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
 In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave.
-This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.
+Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last waves.
+This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11.1% in 2023).
 
@@
 Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases.
+Interestingly, in the last waves of the study (2022-2023) the justification of inequality increases whereas the perception of inequality decreases.
 As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
@@
 In this matrix, the correlations vary between low and moderate values.
-Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01).
+Justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
+Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.11, p<.01).
 Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
@@
 Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01).
-Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
+Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.17, p<.001), 2018 ($\beta$=-0.07, p<.05), and 2019 ($\beta$=-0.07, p>.05).
+Nevertheless, in the last waves of 2022 and 2023 there is a radical increase in level of justification of economic inequality ($\beta$=0.2, p<.001 and $\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
 Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.
+On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.
 
-Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality.
-On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males.
-Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix.
-Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented.
-Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
+Model 3 adds the socioeconomic variables to test hypothesis 2.
+Educational level and household income quintiles were added, however, none of them had a significant effect.
+In addition, from this model onwards the demographic control variables were added, where only being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001).
+Subjective social status, age and gender had no significant effect when controlling for all the variables analysed.
 
-Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01).
+Models 4 and 5 add the two variables about the Covid context, where we can see that those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.05, p<.01).
+On the other hand, those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.08, p<.01).
+
+Inequality perception is added in Model 6, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
+Model 7 and 8 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
+In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 7 ($\beta$=0.05, p<.001).
 However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.
+In this sense, Model 8 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.08, p<.001), but it has a larger weight than the perception of talent.
 
@@
 h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
-h7 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia"
-h8 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios"
+h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia"
+h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
+h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
+h7 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
+h8 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
 
@@
 
-omit <- "(education)|(ess)"
+omit <- "(edad)|(ess)|(sexo)|(pos_id)"
 texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
+Note: Model 3 to 8 are controlled by gender, political position, and subjective social status with no significant effects.", 
                 omit.coef = omit,
@@
                                       "Wave^2",
-#                                      "High School <br> <i>(Ref. Primary)</i>",
-#                                      "Technical",
-#                                      "Universitary",
+                                      "High School",
+                                      "Technical",
+                                      "Universitary",
                                       "Quintile 2",
@@
 #                                      "Subjective Social Status",
-                                      "Age 30-49",
-                                      "Age 50-64",
-                                      "Age 65 or more",
-                                      "Gender (Ref. Male)",
-                                      "Center",
-                                      "Right",
-                                      "Does not identify",
+                                      # "Age 30-49",
+                                      # "Age 50-64",
+                                      # "Age 65 or more",
+                                      # "Gender (Ref. Male)",
+                                      # "Center",
+                                      # "Right",
+                                      # "Does not identify",
+                                      "Covid: economy more important than health",
+                                      "Covid: receive state benefits",
                                       "Inequality perception",
                                       "Merit: Talent",
-                                      "Merit: Effort" ,
-                                      "Covid: economy more important than health",
-                                      "Covid: receive state benefits"),
+                                      "Merit: Effort"),
                 groups = list("Wave (Ref.= Wave 2016)" = 2:6,
-                  "Household Income (Ref.= Quintile 1)" = 9:13,
-                  "Age (Ref.= 18-29)" = 14:16,
-                  "Pol. pos (Ref.= Left)" = 18:20
+                              "Education (Ref.= Primary)" = 9:11,
+                  "Household Income (Ref.= Quintile 1)" = 12:16
+                  #"Age (Ref.= 18-29)" = 14:16,
+                  #"Pol. pos (Ref.= Left)" = 18:20
                   ),
@@
                 include.aic = FALSE,
+                custom.gof.rows = list("Controls"= c("No", "No", rep("Yes", 6))),
                 custom.gof.names = c("BIC",
@@
 
-```{r interact, echo=FALSE, results='asis'}
+```{r interact, echo=FALSE, results='asis', warning=FALSE}
 #| label: tbl-interact
-#| fig-cap: "Interaction effects for the justification of economic inequality"
+#| tbl-cap: "Interaction effects for the justification of economic inequality"
 
@@
 
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios+merit_talent*ola"
+h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
+h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*ola"
+h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*ola"
 
 # A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
@@
                 include.aic = FALSE,
+                custom.model.names = c("Model 8", "Model 9", "Model 10"),
                 custom.coef.names = c("Intercept",
@@
                                      "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort"),  
+                                     "Cov: Individuals (Intercept) Wave"),  
                center = TRUE,
@@
                include.rmse = FALSE,
-               include.ci = FALSE,
-               threeparttable = TRUE)
+               include.ci = FALSE)
 ```
 
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time.
+In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of educational inequality over time.
 In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
 We test this hypothesis through interaction effects, displayed in @tbl-interact.
-Model 1 is shown as a baseline model, it is the same as Model 6 in @tbl-interact but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort.
-As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.
+Model 1 is shown as a baseline model, it is the same as Model 8 in @tbl-interact but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
+Model 9 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 10 does the same but now for the perception of meritocracy related to effort.
+As observed, only the talent variable shows a significant interaction with the time variable ($\beta$=-0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on talent and the justification of inequality decreases by 0.01 points in every measurement point.

### 2024-12-26 production/prod_analysis-fix-t5.Rmd
changes methods + analisis

@@
 # Df variables nivel 1 _________________________________________________________
-df_study1 <- df_study_long_t7_comunas %>% dplyr::select(ola,
-                                               just_educ, des_perc, merit_talent, merit_effort,
+df_study1 <- df_study_long_t7_comunas %>% dplyr::select(ola, idencuesta,
+                                               just_educ, covid_economia, covid_beneficios, des_perc, merit_talent, merit_effort,
                                                educ, quintil1:sexo)
+
+ids <- 
+  df_study1 %>% 
+  select(idencuesta,ola) %>% 
+  filter(ola==1) %>% 
+  sjmisc::frq(idencuesta,show.na = F) %>% as.data.frame()
+
+# filter data by the idencuesta of t7
+df_study1 <- 
+  df_study1 %>%
+  filter(idencuesta %in% ids$val)
 
@@
 
-for (i in c("just_educ", "des_perc",
+for (i in c("just_educ", "des_perc", "covid_economia",
             "merit_effort","merit_talent")) {
@@
 
+df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
+df_study1$covid_beneficios <- factor(df_study1$covid_beneficios,
+                                     levels = 1:2,
+                                     labels = c("No", "Yes"))
+
 df_study1$just_educ <- 
@@
 
+df_study1$covid_economia <- 
+sjlabelled::set_label(x = df_study1$covid_economia, 
+                      label = "It is more important to give priority to economic activity than to the health of the population")
+
+df_study1$covid_beneficios <- 
+sjlabelled::set_label(x = df_study1$covid_beneficios, 
+                      label = "Obtained state benefits during the covid-19 pandemic")
 #
@@
 
-df<- dfSummary(x = dplyr::select(df_study1, ola, merit_talent, merit_effort, des_perc, educ, quintil1, ess, edad, sexo, pos_id),
+df<- dfSummary(x = dplyr::select(df_study1, ola, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, educ, quintil1, ess, edad, sexo, pos_id),
                plain.ascii = FALSE,
@@
 df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               just_educ, des_perc, merit_effort, merit_talent,covid_economia, covid_beneficios,
                                                educ, quintil1:sexo) %>% na.omit()
@@
   scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022, 2023)) +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 9519)")
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7894)")
 
@@
 df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent,
+                                               just_educ, des_perc, merit_effort, merit_talent, covid_economia, covid_beneficios,
                                                educ, quintil1:sexo) %>% na.omit()
@@
         plot.caption = element_text(hjust = 1)) +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2022 (n = 9519)")
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7894)")
 merit_plot
@@
   geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 1.1, nudge_y = 0, size = 4)  +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 9519)")
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7894)")
 
@@
 grid::grid.text("Source: own elaboration with pooled data 
-from ELSOC 2016-2023 (n = 9519)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
+from ELSOC 2016-2023 (n = 7894)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
 dev.off()

### 2024-12-30 01-introduction.qmd
uodates para respuestas reviewers

@@
 
-[[no sé si de pronto aclarar más la idea de que no es posible aislar el efecto COVID propiamente tal ...]]
+Against this backdrop, Chile serves as an example among Latin American countries due to its strong sanitary measures.
+These measures significantly impacted the labor market, leading to important consequences for unemployment and income distribution.
+It is importantto approach any claims about the influence of the COVID-19 pandemic with caution.
+Our findings are situated within a broader context of increasing socioeconomic risks and vulnerabilities, which were heightened by the sanitary restrictions implemented during the pandemic.

### 2024-12-30 02-antecedentes.qmd
uodates para respuestas reviewers

@@
 
+Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health.
+However, the relationship between meritocracy and the justification of inequality within the educational domain remains underexplored.
+For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit.
+Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
+
 More specific empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
@@
 Besides, in the specific case of Chile there were two constitutional processes happening in this very same period after a massive social upheaval.
-[[redondear-idea]]
 
@@
 Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
+
+Within this framework, we point out that periods of rising socioeconomic and health vulnerabilities could weaken the justification of social inequalities.
+This change would be at least in part linked to a challenge of the meritocratic ideals.
+Chile provides a significant case for studying inequality, as it exemplifies a prosperous yet highly unequal society within Latin America.
+Under such conditions, a shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions.
+Chile implemented sanitary controls early in the pandemic, initiating lockdowns on March 18, 2020, and lifting all restrictions by March 5, 2023.
+In this context, Chile stands out as one of the strictest countries globally and within South America in terms of pandemic-related policies.
+According to the Stringency Index [@hale_global_2021]—ranging from 0 to 100 (100 being the strictest) and encompassing measures like school closures, workplace shutdowns, and travel bans—the global median was 42.54, while Chile scored 58.1.
+This placed Chile 36th among 184 countries and 4th in South America, following Peru (60.1), Suriname (60.5), and Venezuela (80.5).
+
+Within the period analyzed (2016-2022), an important cycle of protests known as the "social outburst" took place between October 2019 and March 2020 [@somma_no_2021].
+However, the protest drastically decreased as a result of the lockdown [@joignant_informe_2020].
+In terms of public views on economic inequality, it has been documented that critical views toward inequality increased in 2019 and deepened under the pandemic [@coes_radiograficambio_2023], which we argue can be in part explained by the increased risk exposure and vulnerability in the context of the COVID-19 pandemic.
+Hereby, a relevant aspect is the transformation of the labor market and its impact on income distribution [@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,].
+In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6.
+In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade.
+By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.
+
 Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period.

### 2024-12-30 03-methods.qmd
uodates para respuestas reviewers

@@
 
-For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status.
+For testing the hypothesis regarding status we included educational level and household income quintiles.
+Also, we included subjective social status as it has been argued that perceived social status is a relevant predictor of attitudes toward economic inequality [@schneider_poverty_2015;@castillo_meritocracia_2019].
+In addition, subjective measures can complement objective measures in predicting life chances [@oesch_subjective_2023], which are arguably connected to how individuals experience economic inequality.
+In this regard, it has been argued that how people form their views on their standing in society results from experiences with direct socioeconomic circumstances and social comparison processes with reference groups [@Condon2020].
 Finally, political identification on a left-right scale, age groups, and gender were included as control variables.

### 2025-01-09 07-apendice.qmd
change income vars

@@
 
-```{r multilevel-full, echo=FALSE, results='asis'}
+```{r echo=FALSE, results='asis'}
+#| label: tbl-multilevel-full
+#| tbl-cap: "Complete multilevel longitudinal models for the justification of inequality in education"
+
+pacman::p_load(dplyr,       # Manipulacion de datos
+               knitr,
+               kableExtra,
+               summarytools,
+               lme4,
+               texreg
+               )
+load(file = here::here("input/data-proc/df_study_t6_ind_comunas.RData"))
+
+# generate analytical sample
 df_study1 <- 
   df_study_long_t7_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo) %>% 
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo, covid_economia, covid_beneficios, inc_pcmed) %>% 
   na.omit() %>% 
@@
          )
+
+df_study1 <- df_study1 %>%
+  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
+  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
+  ungroup()
+#sjmisc::frq(df_study1$n_participaciones)
+df_study1 <- df_study1 %>% filter(n_participaciones>1)
+
+#, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp
+df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
+
 # Hipotesis
@@
 h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
+h2 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo"
+h3 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id"
+h4 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia"
+h5 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
+h6 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
+h7 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
+h8 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
 
@@
 
-omit <- "(education)|(ess)"
-texreg::htmlreg(list(educ0, educ1, educ3, educ4, educ5, educ6),
+#omit <- "(edad)|(ess)|(sexo)|(pos_id)"
+texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
@@
                                       "Wave 2022",
-                                      "wave 2023",
+                                      "Wave 2023",
                                       "Wave",
                                       "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
+                                      "Universitary (Ref.= Less than universitary)",
+                                      "Income",
+                                      # "High School",
+                                      # "Technical",
+                                      # "Universitary",
+                                      # "Quintile 2",
+                                      # "Quintile 3",
+                                      # "Quintile 4",
+                                      # "Quintile 5",
+                                      # "Quintile (No information)",
                                       "Subjective Social Status",
@@
                                        "Does not identify",
+                                      "Covid: economy more important than health",
+                                      "Covid: receive state benefits",
                                       "Inequality perception",
                                       "Merit: Talent",
-                                      "Merit: Effort" ),
+                                      "Merit: Effort"),
                 groups = list("Wave (Ref.= Wave 2016)" = 2:6,
-                              "Education (Ref.=Primary)"= 9:11,
-                  "Household Income (Ref.= Quintile 1)" = 12:16,
-                  "Age (Ref.= 18-29)" = 18:20,
-                  "Pol. pos (Ref.= Left)" = 22:24
+                          #    "Education (Ref.= Primary)" = 9:11,
+                  #"Household Income (Ref.= Quintile 1)" = 12:16
+                  "Age (Ref.= 18-29)" = 12:14,
+                  "Pol. pos (Ref.= Left)" = 16:18
                   ),
@@
                 include.aic = FALSE,
+ #               custom.gof.rows = list("Controls"= c("No", "No", rep("Yes", 6))),
                 custom.gof.names = c("BIC",
@@
                                      "Var: Residual"),
-               caption = paste("(\\#tab:multilevel-full)","Multilevel model for the justification of economic inequality"),
                center = TRUE,
                caption.above = TRUE,
-               scalebox = 0.60,
+               scalebox = 0.765,
                float.pos = "!ht", 

### 2025-01-09 production/prod_prep-mergedata.Rmd
change income vars

@@
                   by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo,
+ dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp,
                esc_cit17) %>% 
@@
                   by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo,
+ dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp, 
                esc_cit17) %>% 
@@
 summary(df_study1_long_comunas$var_simce_comuna_rescaled)
+```
+
+```{r}
 # save data____________________________________________________________________

### 2025-01-09 production/prod_prep.Rmd
change income vars

@@
 # load libraries
-pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here)
+pacman::p_load(dplyr,readxl,sjmisc, sjlabelled, questionr, car,here, ggstats)
 load(here::here("input/data-original/elsoc_long.Rdata"))
@@
   sjlabelled::drop_labels(., drop.na = FALSE)
+```
 
+```{r}
 #
@@
 sjmisc::frq(elsoc_long_2016_2023$ahead_hardwork)
-# Socioeconomic_________________________________________________________________
+```
 
-# Education_______________________________________
-elsoc_long_2016_2023$educ <- 
-  car::recode(elsoc_long_2016_2023$m01,
-              "c(1,2,3)=1;c(4,5)=2;c(6,7)=3;c(8,9,10)=4; c(-888,-999)=NA")
-elsoc_long_2016_2023$educ <-
-  factor(elsoc_long_2016_2023$educ,
-         labels = c("Primary","High school","Technical","Universitary"))
-
-#reverse education, reference level is the highest level
-#elsoc_long_2016_2023$educ <- forcats::fct_rev(elsoc_long_2016_2023$educ)
-
-elsoc_long_2016_2023$educ <- 
-sjlabelled::set_label(x = elsoc_long_2016_2023$educ,
-                      label = "Education")
-sjmisc::frq(elsoc_long_2016_2023$educ)
-
-#Recoding of education to years based on casen 2017.
-elsoc_long_2016_2023$educyear<- as.numeric(
-  car::recode(elsoc_long_2016_2023$m01, 
-              "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
-               7=14.8;8=14.9;9=16.9;10=19.07;c(-888,-999)=NA", 
-              as.numeric = T))
-
-elsoc_long_2016_2023$educyear <- 
-sjlabelled::set_label(x = elsoc_long_2016_2023$educyear,
-                      label = "Education in years")
-
-class(elsoc_long_2016_2023$educyear)
-sjmisc::frq(elsoc_long_2016_2023$educyear)
-
-# Household income_________________________________________
-
-#Impute midpoint of income ranges
-elsoc_long_2016_2023$m30_rec <-
-as.numeric(car::recode(elsoc_long_2016_2023$m30,
-           "1=110000;2=251000;3=305000;4=355000;5=400000;
-            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
-            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
-            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))
-
-#Impute midpoint of income ranges (2021)
-elsoc_long_2016_2023$m30b_rec <-
-as.numeric(car::recode(elsoc_long_2016_2023$m30b,
-           "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))
-
-sjmisc::frq(elsoc_long_2016_2023$m30b_rec)
-
-#Recode DK/DA of Income to NA
-elsoc_long_2016_2023$m29_rec <-
-  as.numeric(car::recode(elsoc_long_2016_2023$m29,"c(-888,-999)=NA"))
-
-#replace NA of income with new imputed variable
-elsoc_long_2016_2023$m29_imp <- 
-  ifelse(test = !is.na(elsoc_long_2016_2023$m29_rec),
-         yes =  elsoc_long_2016_2023$m29_rec,
-         no =  elsoc_long_2016_2023$m30_rec)
-summary(elsoc_long_2016_2023$m29_imp)
-
-elsoc_long_2016_2023$m29_imp <- 
-  ifelse(test = is.na(elsoc_long_2016_2023$m29_imp),
-         yes =  elsoc_long_2016_2023$m30b_rec,
-         no =  elsoc_long_2016_2023$m29_imp)
-summary(elsoc_long_2016_2023$m29_imp)
-
-# deflate at each year's prices
-elsoc_long_2016_2023$deflactor <-
-  with(elsoc_long_2016_2023, case_when(
-    ola == 2016 ~ 113.88 / 123.82,
-    ola == 2017 ~ 116.46 / 123.82,
-    ola == 2018 ~ 119.45 / 123.82,
-    ola == 2019 ~ 123.82 / 123.82
-  ))
-
-# N Household:
-elsoc_long_2016_2023 <-
-  elsoc_long_2016_2023 %>%
-  mutate(n_hogar =
-           dplyr::case_when(ola == 1 ~ nhogar1,
-                            ola == 2 ~ m46_nhogar,
-                            ola == 3 ~ m54,
-                            ola == 4 ~ m54,
-                            ola == 5 ~ m54,
-                            ola == 6 ~ m54,
-                            ola == 7 ~ m54))
-sjmisc::frq(elsoc_long_2016_2023$n_hogar)
-
-#Recode DK/DA to NA
-elsoc_long_2016_2023$n_hogar_r<-
-  car::recode(elsoc_long_2016_2023$n_hogar,"c(-888,-999)=NA")
-
-# Per capita household income:
-elsoc_long_2016_2023$ing_pc <- 
-  (elsoc_long_2016_2023$m29_imp/elsoc_long_2016_2023$n_hogar_r)
-
-elsoc_long_2016_2023$ing_pc <-
-sjlabelled::set_label(x = elsoc_long_2016_2023$ing_pc,
-                      label = "Household income per capita")
-
-sjmisc::descr(elsoc_long_2016_2023$ing_pc)
-
-# Compute income quintiles
-elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
-  group_by(ola) %>% 
-  mutate(quintil = ntile(-desc(ing_pc), 5)) %>% 
-  ungroup()
-
-elsoc_long_2016_2023$quintil <- 
-  factor(elsoc_long_2016_2023$quintil,
-         levels = c(1, 2, 3, 4, 5),
-         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors
-
-#reverse quintile, reference level is the highest quintile
-#elsoc_long_2016_2023$quintil <- forcats::fct_rev(elsoc_long_2016_2023$quintil)
-
-elsoc_long_2016_2023$quintil <- 
-sjlabelled::set_label(x = elsoc_long_2016_2023$quintil,
-                      label = "Household income quintile per capita")
-
-sjmisc::frq(elsoc_long_2016_2023$quintil)
-
-#include new quintile category with missing cases
-elsoc_long_2016_2023$quintil1<-
-  car::recode(elsoc_long_2016_2023$quintil, 
-              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")
-
-#elsoc_long_2016_2023$quintil1 <- factor(elsoc_long_2016_2023$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))
-
-elsoc_long_2016_2023$quintil1 <- 
-sjlabelled::set_label(x = elsoc_long_2016_2023$quintil1,
-                      label = "Household income quintile per capita (NA)") 
-sjmisc::frq(elsoc_long_2016_2023$quintil1)
-
+```{r}
 #Age Ranges
@@
 ```{r}
+# Socioeconomic_________________________________________________________________
+
+# Education_______________________________________
+elsoc_long_2016_2023$educ <- 
+  car::recode(elsoc_long_2016_2023$m01,
+              "c(1,2,3,4,5,6,7)=1;c(8,9,10)=2; c(-888,-999)=NA")
+elsoc_long_2016_2023$educ <-
+  factor(elsoc_long_2016_2023$educ,
+         labels = c("Less than Universitary","Universitary"))
+
+#reverse education, reference level is the highest level
+#elsoc_long_2016_2023$educ <- forcats::fct_rev(elsoc_long_2016_2023$educ)
+
+elsoc_long_2016_2023$educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$educ,
+                      label = "Education")
+sjmisc::frq(elsoc_long_2016_2023$educ)
+
+#Recoding of education to years based on casen 2017.
+elsoc_long_2016_2023$educyear<- as.numeric(
+  car::recode(elsoc_long_2016_2023$m01, 
+              "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
+               7=14.8;8=14.9;9=16.9;10=19.07;c(-888,-999)=NA", 
+              as.numeric = T))
+
+elsoc_long_2016_2023$educyear <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$educyear,
+                      label = "Education in years")
+
+class(elsoc_long_2016_2023$educyear)
+sjmisc::frq(elsoc_long_2016_2023$educyear)
+```
+
+```{r}
+# N Household:
 # Select variables______________________________________________________________
@@
          sexo,
-         # cuestion_mig,
-         # starts_with("in_"),
          just_educ,
@@
          # quintil,
-         quintil1,
+#         quintil1,
          ess,
@@
          covid_retirodos,
-         # ess_inm,
-         # nation_t1,
-         # know_inm,
-         # know_inm_bi,
-         # frien_inm_bi_t4,
-         # frien_inm_bi_t1,
-         # know_inm_t1t4,
-         # frien_inm,
-         # frien_inm_bi,
-         # frien_inm_t1t4,
          pos_id,
-         edad)
+         edad,
+         nhogar1,
+         m46_nhogar,
+         m54, m30, m30b, m29)
+
+# Household income_________________________________________
+
+#Impute midpoint of income ranges
+df_study1$m30_rec <-
+as.numeric(car::recode(df_study1$m30,
+           "1=110000;2=251000;3=305000;4=355000;5=400000;
+            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
+            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
+            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))
+
+#Impute midpoint of income ranges (2021)
+df_study1$m30b_rec <-
+as.numeric(car::recode(df_study1$m30b,
+           "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))
+
+sjmisc::frq(df_study1$m30b_rec)
+
+#Recode DK/DA of Income to NA
+df_study1$m29_rec <-
+  as.numeric(car::recode(df_study1$m29,"c(-888,-999)=NA"))
+
+#replace NA of income with new imputed variable
+df_study1$m29_imp <- 
+  ifelse(test = !is.na(df_study1$m29_rec),
+         yes =  df_study1$m29_rec,
+         no =  df_study1$m30_rec)
+summary(df_study1$m29_imp)
+
+df_study1$m29_imp <- 
+  ifelse(test = is.na(df_study1$m29_imp),
+         yes =  df_study1$m30b_rec,
+         no =  df_study1$m29_imp)
+summary(df_study1$m29_imp)
+
+# deflate at each year's prices
+df_study1$deflactor <-
+  with(df_study1, case_when(
+    ola == 2016 ~ 113.88 / 123.82,
+    ola == 2017 ~ 116.46 / 123.82,
+    ola == 2018 ~ 119.45 / 123.82,
+    ola == 2019 ~ 123.82 / 123.82
+  ))
 
@@
                      # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
-                     values_from = names(select(df_study1,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:edad,sexo))
+                     values_from = names(select(df_study1,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:edad,sexo, nhogar1, m46_nhogar, m54, m30, m30b, m29_imp))
+                     )
+
+df_study1_wide$m54_6 <- df_study1_wide$m54_7
+
+# reshape from long to wide
+pacman::p_load(datasets,data.table)
+df_study1_long <- data.table::melt.data.table(data.table::setDT(df_study1_wide),
+              id.vars = c("idencuesta","muestra"),
+              variable.name = c("ola"),
+              measure = patterns("^comuna_","^comunacod_","^ponderador_long_total", "^segmento", "^estrato",
+                                 "^just_educ_", "^just_pension_", "^just_salud_", "^des_perc_",
+                                 "^merit_effort_","^merit_talent_",
+                                 "^ahead_family_", "^ahead_educ_", "^ahead_ambition_", "^ahead_hardwork_",
+                                 "^covid_conocidos_", "^covid_economia_", "^covid_cuarentena_", "^covid_propio_",
+                                 "^covid_alimentos_", "^covid_beneficios_", "^covid_retirouno_", "^covid_retirodos_",
+                                 "^educ_","^educyear_","^ess_",
+                                 "^pos_id_",
+                                 "^edad_","^sexo_", "^m54_", "^nhogar1_", "^m46_nhogar_", "^m30_", "^m30b_", "^m29_imp_"
+                                 ),
+              value.name = c("comuna","comunacod","ponderador_long_total","segmento","estrato",
+                             "just_educ", "just_pension","just_salud", "des_perc",
+                             "merit_effort","merit_talent",
+                             "ahead_family","ahead_educ","ahead_ambition","ahead_hardwork",
+                             "covid_conocidos", "covid_economia", "covid_cuarentena", "covid_propio",
+                             "covid_alimentos", "covid_beneficios", "covid_retirouno", "covid_retirodos",
+                             "educ","educyear","ess",
+                             "pos_id",
+                             "edad","sexo", "m54", "nhogar1", "m46_nhogar", "m30", "m30b", "m29_imp"),
+              na.rm = F,value.factor = TRUE
+              )
+
+df_study1_long <-
+  df_study1_long %>%
+  mutate(n_hogar =
+           dplyr::case_when(ola == 1 ~ nhogar1,
+                            ola == 2 ~ m46_nhogar,
+                            ola == 3 ~ m54,
+                            ola == 4 ~ m54,
+                            ola == 5 ~ m54,
+                            ola == 6 ~ m54,
+                            ola == 7 ~ m54))
+sjmisc::frq(df_study1_long$n_hogar)
+
+#Recode DK/DA to NA
+df_study1_long$n_hogar_r<-
+  car::recode(df_study1_long$n_hogar,"c(-888,-999)=NA")
+
+# Per capita household income:
+df_study1_long$ing_pc <- 
+  (df_study1_long$m29_imp/df_study1_long$n_hogar_r)
+
+df_study1_long$ing_pc <-
+sjlabelled::set_label(x = df_study1_long$ing_pc,
+                      label = "Household income per capita")
+
+sjmisc::descr(df_study1_long$ing_pc)
+
+# Compute income quintiles
+df_study1_long <- df_study1_long %>% 
+  group_by(ola) %>% 
+  mutate(quintil = ntile(-desc(ing_pc), 5)) %>% 
+  ungroup()
+
+df_study1_long$quintil <- 
+  factor(df_study1_long$quintil,
+         levels = c(1, 2, 3, 4, 5),
+         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors
+
+#reverse quintile, reference level is the highest quintile
+#elsoc_long_2016_2023$quintil <- forcats::fct_rev(elsoc_long_2016_2023$quintil)
+
+df_study1_long$quintil <- 
+sjlabelled::set_label(x = df_study1_long$quintil,
+                      label = "Household income quintile per capita")
+
+sjmisc::frq(df_study1_long$quintil)
+
+#include new quintile category with missing cases
+df_study1_long$quintil1<-
+  car::recode(df_study1_long$quintil, 
+              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")
+
+#elsoc_long_2016_2023$quintil1 <- factor(elsoc_long_2016_2023$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))
+
+df_study1_long$quintil1 <- 
+sjlabelled::set_label(x = df_study1_long$quintil1,
+                      label = "Household income quintile per capita (NA)") 
+sjmisc::frq(df_study1_long$quintil1)
+```
+
+# Alternative operationalization
+_Income_
+```{r}
+df_study1_long <-
+df_study1_long %>%
+  # group_by(ola) %>%
+  mutate(percentile_wave=ntile(ing_pc,100),
+         bot_mid_top_wave=car::recode(percentile_wave,"1:20=1;21:79=2;80:100=3"))
+
+elsoc_wide <- panelr::widen_panel(panelr::panel_data(df_study1_long %>% select(idencuesta,ola,percentile_wave,bot_mid_top_wave), id = idencuesta, wave = ola), separator = "_")
+elsoc_wide$income_prof <- paste0(elsoc_wide$bot_mid_top_wave_1,elsoc_wide$bot_mid_top_wave_5)
+frq(elsoc_wide$income_prof)
+# a. Stable low income: In the lowest tertile in all three periods
+# b. Stable middle income: In the middle tertile in all three periods
+# c. Stable high income: In the highest tertile in all three periods
+# d. Upward mobility: In the lowest tertile in the first period and either in tertile 2 or 3 in the last period
+# e. Downward mobility: In the highest tertile in the first period and either in tertile 1 or 2 in the last period
+# f. Other: All other trajectories
+elsoc_wide$ind_inc_prof <- car::recode(elsoc_wide$income_prof,"'11'='1stablelow';'33'='2stablemid';'22'='3stablehig';c('12','13')='4upwardmob';c('31','21')='5downwardmob';else = '99other'")
+frq(elsoc_wide$ind_inc_prof)
+# • Stable low income: Start and end at the 20th percentile in income (P20).
+# • Stable middle income: Start and end at the 50th percentile in income (P50).
+# • Stable high income: Start and end at the 80th percentile in income (P80).
+# • Upward mobility: Start at 20th percentile, end at 80th percentile (P20 → P80).
+# • Downward mobility: Start at 80th percentile, end at 20th percentile (P80 → P20).
+df_profiles<- select(elsoc_wide,"idencuesta","ind_inc_prof",percentile_wave_1,percentile_wave_5)
+df_profiles$idencuesta <- as.numeric(as.character(df_profiles$idencuesta))
+df_study1_long <-
+  df_study1_long %>%
+  left_join(select(df_profiles,"idencuesta","ind_inc_prof",percentile_wave_1,percentile_wave_5),by="idencuesta")
+```
+
+```{r}
+# Income group as percentage of median
+# poverty (<50 per cent of median income), below average (>¼50 per cent and <100 per cent), above average (>¼100 per cent and <200 per cent), and affluence (>¼200 per cent).
+df_study1_long<-
+  df_study1_long %>%
+  group_by(ola) %>%
+  #compute median household pc income
+  mutate(inc_pcmed=weighted.median(ing_pc,ponderador_long_total,na.rm = T)) %>%
+  mutate(ratioincome=(ing_pc/inc_pcmed)*100) %>%
+  ungroup()
+# elsoc_long_2016_2022.2[,c('idencuesta',"ola", 'm29_imp',"n_hogar","ing_pc","inc_pcmed")] %>% View()
+df_study1_long$incomegrp=
+  car::recode(df_study1_long$ratioincome,
+              "0:49.9='Low';50:99.9='Below average';100:200='Above average';200.1:999999999999='Affluence'",
+              levels =c("Low","Below average","Above average","Affluence"),as.factor = T)
+frq(df_study1_long$incomegrp)
+```
+
+```{r}
+# Reshape long to wide
+df_study1_wide <- df_study1_long %>% 
+  tidyr::pivot_wider(id_cols = c("idencuesta","muestra"),
+                     names_from = "ola",
+                     # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
+                     values_from = names(select(df_study1_long,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:edad,sexo, quintil1, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp))
                      )
@@
                 show.frq = T,show.values = T,show.na = T,show.prc = T, show.type = T)
+```
 
+```{r}
 # reshape from long to wide
@@
                                  "^pos_id_",
-                                 "^edad_","^sexo_"
+                                 "^edad_","^sexo_",
+                                 "^percentile_wave_", "^bot_mid_top_wave_", "^inc_pcmed_", "^ratioincome_", "^incomegrp_"
                                  ),
@@
                              "pos_id",
-                             "edad","sexo"),
+                             "edad","sexo",
+                             "percentile_wave", "bot_mid_top_wave", "inc_pcmed", "ratioincome", "incomegrp"),
               na.rm = F,value.factor = TRUE
               )
+```
 
+```{r}
 names(df_study1_long) #check names of long dataset
@@
 
-df_study1_long <- 
-set_label(x = df_study1_long,
-          label = get_label(select(df_study1,names(df_study1_long))))
+# df_study1_long <- 
+# set_label(x = df_study1_long,
+#           label = get_label(select(df_study1,names(df_study1_long))))
 
@@
 sjmisc::frq(df_study1_long_t7$ola)
+```
 
+```{r}
 # SAVE DATA____________________________________________________________________

### 2025-01-09 04-analisis.qmd
add inc_pcmed analisis

@@
   df_study_long_t7_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo, covid_economia, covid_beneficios) %>% 
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo, covid_economia, covid_beneficios, inc_pcmed) %>% 
   na.omit() %>% 
@@
 
+df_study1 <- df_study1 %>%
+  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
+  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
+  ungroup()
+#sjmisc::frq(df_study1$n_participaciones)
+df_study1 <- df_study1 %>% filter(n_participaciones>1)
+
+#, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp
 df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
@@
 h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia"
-h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
-h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
-h7 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
-h8 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
+h2 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo"
+h3 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id"
+h4 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia"
+h5 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
+h6 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
+h7 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
+h8 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
 
@@
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 8 are controlled by gender, political position, and subjective social status with no significant effects.", 
+Note: Model 3 to 8 are controlled by age, gender, political position, and subjective social status.", 
                 omit.coef = omit,
@@
                                       "Wave^2",
-                                      "High School",
-                                      "Technical",
-                                      "Universitary",
-                                      "Quintile 2",
-                                      "Quintile 3",
-                                      "Quintile 4",
-                                      "Quintile 5",
-                                      "Quintile (No information)",
+                                      "Universitary (Ref.= Less than universitary)",
+                                      "Income",
+                                      # "High School",
+                                      # "Technical",
+                                      # "Universitary",
+                                      # "Quintile 2",
+                                      # "Quintile 3",
+                                      # "Quintile 4",
+                                      # "Quintile 5",
+                                      # "Quintile (No information)",
 #                                      "Subjective Social Status",
@@
                                       "Merit: Effort"),
-                groups = list("Wave (Ref.= Wave 2016)" = 2:6,
-                              "Education (Ref.= Primary)" = 9:11,
-                  "Household Income (Ref.= Quintile 1)" = 12:16
+                groups = list("Wave (Ref.= Wave 2016)" = 2:6
+                          #    "Education (Ref.= Primary)" = 9:11,
+                  #"Household Income (Ref.= Quintile 1)" = 12:16
                   #"Age (Ref.= 18-29)" = 14:16,
@@
 
-df_study1 <- 
-  df_study_long_t7_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo, covid_economia, covid_beneficios) %>% 
-  na.omit() %>% 
-  mutate(ola = as.numeric(ola),
-         ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
- #        edad=as.numeric(edad)
-         #pos_id=as.numeric(pos_id)
-         )
+# df_study1 <- 
+#   df_study_long_t7_comunas %>%
+#   select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent,education=educ,quintil1:sexo, covid_economia, covid_beneficios) %>% 
+#   na.omit() %>% 
+#   mutate(ola = as.numeric(ola),
+#          ola_2=as.numeric(ola)^2,
+#          sexo=as.factor(sexo)
+#  #        edad=as.numeric(edad)
+#          #pos_id=as.numeric(pos_id)
+#          )
 
-df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
+df_study1$ola <- car::recode(df_study1$ola, c("1=1; 2=1;3=1;4=1; 6=2;7=2"))
+df_study1$ola <- as.factor(df_study1$ola)
+#df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
 # Hipotesis
 
-h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*ola"
-
+h6 <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
+h1b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*inc_pcmed"
+h3b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*inc_pcmed"
+h4b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+des_perc*inc_pcmed"
+h5b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+inc_pcmed*ola"
 # A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(inc_pcmed+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(inc_pcmed+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(inc_pcmed+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+inc_pcmed|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
@@
 # table 3: interacciones
-omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)|(covid_beneficios)|(covid_economia)"
-texreg::knitreg(list(educ6,educ3,educ1),
+omit <- "(education)|(ess)|(edad)|(sexo)|(pos_id)|(covid_beneficios)|(covid_economia)"
+texreg::knitreg(list(educ6,educ3,educ1, educ4, educ5),
                 custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                 include.aic = FALSE,
-                custom.model.names = c("Model 8", "Model 9", "Model 10"),
+                custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11", "Model 12"),
                 custom.coef.names = c("Intercept",
@@
                                       "Wave^2",
+                                      "Income",
+                                      "Inequality perception",
                                       "Merit: Talent",
                                       "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort"
+                                      "Income * Merit: Talent",
+                                      "Income * Merit: Effort",
+                                      "Income * Inequality perception",
+                                      "Income * Wave"
                                       ),
@@
                                      "Var: Residual",
+                                     "Var: Individuals Income",
+                                     "Var: Individuals Merit: Talent",
+                                     "Cov: Individuals (Intercept) Income",
+                                     "Cov: Individuals (Intercept) Merit: Talent",
+                                     "Cov: Individuals Income Merit: Talent",
+                                     "Var: Individuals Merit: Effort",
+                                     "Cov: Individuals (Intercept) Merit: Effort",
+                                     "Cov: Individuals Income Merit: Effort",
+                                     "Var: Individuals Inequality perception",
+                                     "Cov: Individuals (Intercept) Inequality perception",
+                                     "Cov: Individuals Income Inequality perception",
                                      "Var: Individuals Wave",
-                                     "Cov: Individuals (Intercept) Wave"),  
+                                     "Cov: Individuals (Intercept) Wave",
+                                     "Cov: Individuals Wave Income"),  
                center = TRUE,

### 2025-01-10 04-analisis.qmd
add income per capita

@@
   df_study_long_t7_comunas %>%
-  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo, covid_economia, covid_beneficios, inc_pcmed) %>% 
+  select(idencuesta,ola,comuna, ponderador_long_total,just_educ, des_perc, merit_effort, merit_talent, education=educ,quintil1:sexo, covid_economia, covid_beneficios, ing_pc) %>% 
   na.omit() %>% 
@@
          ola_2=as.numeric(ola)^2,
-         sexo=as.factor(sexo)
+         sexo=as.factor(sexo),
+         ing_pc_rec = (ing_pc/100000)
          )
@@
 h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia"
-h5 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
-h6 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
-h7 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
-h8 <- "ola_num+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
+h2 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo"
+h3 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id"
+h4 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia"
+h5 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
+h6 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
+h7 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
+h8 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
 
@@
                                       "Universitary (Ref.= Less than universitary)",
-                                      "Income",
+                                      "Income per capita",
                                       # "High School",
@@
                include.ci = FALSE)
-
 ```
@@
 
-h6 <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
-h1b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*inc_pcmed"
-h3b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*inc_pcmed"
-h4b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+des_perc*inc_pcmed"
-h5b <- "ola+ola_2+education+inc_pcmed+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+inc_pcmed*ola"
+h6 <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
+h1b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*ola"
+h3b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*ola"
+h4b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+des_perc*ola"
+h5b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+ing_pc_rec*ola"
 # A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(inc_pcmed+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(inc_pcmed+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(inc_pcmed+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+inc_pcmed|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+ing_pc_rec|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
@@
                                       "Wave",
-                                      "Wave^2",
                                       "Income",
@@
                                       "Merit: Effort",
-                                      "Income * Merit: Talent",
-                                      "Income * Merit: Effort",
-                                      "Income * Inequality perception",
-                                      "Income * Wave"
+                                      "Wave * Merit: Talent",
+                                      "Wave * Merit: Effort",
+                                      "Wave * Inequality perception",
+                                      "Wave * Income"
                                       ),
@@
                                      "Var: Residual",
-                                     "Var: Individuals Income",
+                                     "Var: Individuals Wave",
                                      "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Income",
+                                     "Cov: Individuals (Intercept) Wave",
                                      "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Income Merit: Talent",
+                                     "Cov: Individuals Wave Merit: Talent",
                                      "Var: Individuals Merit: Effort",
                                      "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Income Merit: Effort",
+                                     "Cov: Individuals Wave Merit: Effort",
                                      "Var: Individuals Inequality perception",
                                      "Cov: Individuals (Intercept) Inequality perception",
-                                     "Cov: Individuals Income Inequality perception",
-                                     "Var: Individuals Wave",
-                                     "Cov: Individuals (Intercept) Wave",
+                                     "Cov: Individuals Wave Inequality perception",
+                                     "Var: Individuals Income",
+                                     "Cov: Individuals (Intercept) Income",
                                      "Cov: Individuals Wave Income"),  

### 2025-01-10 production/prod_prep-mergedata.Rmd
add income per capita

@@
                   by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp,
+ dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp, cambio_ingreso_pandemia, ing_pc,
                esc_cit17) %>% 
@@
                   by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp, 
+ dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp, cambio_ingreso_pandemia, ing_pc,
                esc_cit17) %>% 
@@
                   by = "comunacod") %>% 
- dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp,
+ dplyr::select(idencuesta:edad,sexo, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp,cambio_ingreso_pandemia, ing_pc, 
                esc_cit17) %>% 

### 2025-01-10 production/prod_prep.Rmd
add income per capita

@@
 ```{r}
+## Cambio entre primera ola con información y última ola con información
+df_study1_long <- df_study1_long %>%
+  group_by(idencuesta) %>% 
+  mutate(
+    ingreso_primera_ola = ing_pc[which.min(ola)],  # Ingreso en la primera ola
+    ingreso_ultima_ola = ing_pc[which.max(ola)]   # Ingreso en la última ola
+  ) %>%
+  mutate(
+    cambio_ingreso = case_when(
+      is.na(ingreso_primera_ola) | is.na(ingreso_ultima_ola) ~ "NA",  # Si alguno es NA, categoría NA
+      ingreso_ultima_ola > ingreso_primera_ola ~ "Mayor",
+      ingreso_ultima_ola < ingreso_primera_ola ~ "Menor",
+      ingreso_ultima_ola == ingreso_primera_ola ~ "Igual"
+    )
+  ) %>%
+  ungroup()
+
+df_study1_long$cambio_ingreso <- factor(df_study1_long$cambio_ingreso, levels = c("Menor", "Igual", "Mayor", "NA"))
+df_study1_long$cambio_ingreso <- as.numeric(df_study1_long$cambio_ingreso)
+sjmisc::frq(df_study1_long$cambio_ingreso)
+```
+
+```{r}
+### Cambio pre y post pandemia
+df_study1_long <- df_study1_long %>%
+  group_by(idencuesta) %>%
+  mutate(
+    promedio_prepandemia = mean(ing_pc[ola %in% 1:4], na.rm = TRUE),  # Promedio de las olas 1 a 4
+    promedio_postpandemia = mean(ing_pc[ola %in% 6:7], na.rm = TRUE)       # Promedio de las olas 6 y 7
+  ) %>%
+  mutate(
+    cambio_ingreso_pandemia = case_when(
+      is.na(promedio_prepandemia) | is.na(promedio_postpandemia) ~ NA_character_,  # Si faltan datos, NA
+      promedio_postpandemia > promedio_prepandemia ~ "Mayor",
+      promedio_postpandemia < promedio_prepandemia ~ "Menor",
+      promedio_postpandemia == promedio_prepandemia ~ "Igual"
+    )
+  ) %>%
+  ungroup()
+
+df_study1_long$cambio_ingreso_pandemia <- factor(df_study1_long$cambio_ingreso_pandemia, levels = c("Menor", "Igual", "Mayor"))
+sjmisc::frq(df_study1_long$cambio_ingreso_pandemia)
+```
+
+```{r}
 # Reshape long to wide
@@
                      # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
-                     values_from = names(select(df_study1_long,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:edad,sexo, quintil1, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp))
+                     values_from = names(select(df_study1_long,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:edad,sexo, quintil1, ing_pc, percentile_wave, bot_mid_top_wave, inc_pcmed, ratioincome, incomegrp, cambio_ingreso, cambio_ingreso_pandemia))
                      )
@@
 df_study1_wide$quintil1_7 <-df_study1_wide$quintil1_1
+
+df_study1_wide$ing_pc_2 <-df_study1_wide$ing_pc_1 #income per capita
+df_study1_wide$ing_pc_3 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_4 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_5 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_6 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_7 <-df_study1_wide$ing_pc_1
+
+## Alternative incomes
+df_study1_wide$percentile_wave_2 <- df_study1_wide$percentile_wave_1
+df_study1_wide$percentile_wave_3 <- df_study1_wide$percentile_wave_1
+df_study1_wide$percentile_wave_4 <- df_study1_wide$percentile_wave_1
+df_study1_wide$percentile_wave_5 <- df_study1_wide$percentile_wave_1
+df_study1_wide$percentile_wave_6 <- df_study1_wide$percentile_wave_1
+df_study1_wide$percentile_wave_7 <- df_study1_wide$percentile_wave_1
+
+df_study1_wide$bot_mid_top_wave_2 <- df_study1_wide$bot_mid_top_wave_1
+df_study1_wide$bot_mid_top_wave_3 <- df_study1_wide$bot_mid_top_wave_1
+df_study1_wide$bot_mid_top_wave_4 <- df_study1_wide$bot_mid_top_wave_1
+df_study1_wide$bot_mid_top_wave_5 <- df_study1_wide$bot_mid_top_wave_1
+df_study1_wide$bot_mid_top_wave_6 <- df_study1_wide$bot_mid_top_wave_1
+df_study1_wide$bot_mid_top_wave_7 <- df_study1_wide$bot_mid_top_wave_1
+
+df_study1_wide$inc_pcmed_2 <- df_study1_wide$inc_pcmed_1
+df_study1_wide$inc_pcmed_3 <- df_study1_wide$inc_pcmed_1
+df_study1_wide$inc_pcmed_4 <- df_study1_wide$inc_pcmed_1
+df_study1_wide$inc_pcmed_5 <- df_study1_wide$inc_pcmed_1
+df_study1_wide$inc_pcmed_6 <- df_study1_wide$inc_pcmed_1
+df_study1_wide$inc_pcmed_7 <- df_study1_wide$inc_pcmed_1
+
+df_study1_wide$ratioincome_2 <- df_study1_wide$ratioincome_1
+df_study1_wide$ratioincome_3 <- df_study1_wide$ratioincome_1
+df_study1_wide$ratioincome_4 <- df_study1_wide$ratioincome_1
+df_study1_wide$ratioincome_5 <- df_study1_wide$ratioincome_1
+df_study1_wide$ratioincome_6 <- df_study1_wide$ratioincome_1
+df_study1_wide$ratioincome_7 <- df_study1_wide$ratioincome_1
+
+df_study1_wide$incomegrp_2 <- df_study1_wide$incomegrp_1
+df_study1_wide$incomegrp_3 <- df_study1_wide$incomegrp_1
+df_study1_wide$incomegrp_4 <- df_study1_wide$incomegrp_1
+df_study1_wide$incomegrp_5 <- df_study1_wide$incomegrp_1
+df_study1_wide$incomegrp_6 <- df_study1_wide$incomegrp_1
+df_study1_wide$incomegrp_7 <- df_study1_wide$incomegrp_1
 
@@
                                  "^educ_","^educyear_","^quintil1_","^ess_",
-                                 "^pos_id_",
-                                 "^edad_","^sexo_",
-                                 "^percentile_wave_", "^bot_mid_top_wave_", "^inc_pcmed_", "^ratioincome_", "^incomegrp_"
+                                 "^pos_id_", "^edad_","^sexo_",
+                                 "^percentile_wave_", "^bot_mid_top_wave_", "^inc_pcmed_", "^ratioincome_", "^incomegrp_",
+                                 "^ing_pc_", "^cambio_ingreso_pandemia_"
                                  ),
@@
                              "educ","educyear","quintil1","ess",
-                             "pos_id",
-                             "edad","sexo",
-                             "percentile_wave", "bot_mid_top_wave", "inc_pcmed", "ratioincome", "incomegrp"),
+                             "pos_id", "edad","sexo",
+                             "percentile_wave", "bot_mid_top_wave", "inc_pcmed", "ratioincome", "incomegrp",
+                             "ing_pc", "cambio_ingreso_pandemia"),
               na.rm = F,value.factor = TRUE

### 2025-01-13 04-analisis.qmd
add covid interactions

@@
 Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.17, p<.001), 2018 ($\beta$=-0.07, p<.05), and 2019 ($\beta$=-0.07, p>.05).
-Nevertheless, in the last waves of 2022 and 2023 there is a radical increase in level of justification of economic inequality ($\beta$=0.2, p<.001 and $\beta$=0.21, p<.001), suggesting a non-linear change in this variable.
+Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.167, p<.001), 2018 ($\beta$=-0.059, p<.05), and 2019 ($\beta$=-0.03, p>.05).
+Nevertheless, in the last waves of 2022 and 2023 there is a radical increase in level of justification of educational inequality ($\beta$=0.204, p<.001 and $\beta$=0.233, p<.001), suggesting a non-linear change in this variable.
 Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
@@
 Model 3 adds the socioeconomic variables to test hypothesis 2.
-Educational level and household income quintiles were added, however, none of them had a significant effect.
-In addition, from this model onwards the demographic control variables were added, where only being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001).
-Subjective social status, age and gender had no significant effect when controlling for all the variables analysed.
+Educational level and household per capita income were added, but none of them had a significant effect.
+However, in the next models, controlling for all variables, household per capita income had a positive significant effect ($\beta$=0.001, p<.05).
+In addition, from this model onwards the demographic control variables were added, where being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001), and being women (compared to men) had a negative and significant effect ($\beta$=-0.08, p<.001).
+Subjective social status and age had no significant effect when controlling for all the variables analysed.
+The full models can be seen in the appendix (see @tbl-multilevel-full).
 
-Models 4 and 5 add the two variables about the Covid context, where we can see that those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.05, p<.01).
-On the other hand, those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.08, p<.01).
+Models 4 and 5 add the two variables about the Covid context, where we can see that those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.038, p<.05).
+On the other hand, those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.093, p<.01).
 
@@
 Model 7 and 8 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 7 ($\beta$=0.05, p<.001).
+In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 7 ($\beta$=0.044, p<.001).
 However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 8 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.08, p<.001), but it has a larger weight than the perception of talent.
+In this sense, Model 8 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.083, p<.001), but it has a larger weight than the perception of talent.
 
@@
                lme4,
-               texreg
+               texreg,
+               ggeffects,
+               ggplot2
                )
@@
 df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
+df_study1$covid_beneficios<- factor(df_study1$covid_beneficios)
 
@@
 texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
+                digits = 3,
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                                       "Universitary (Ref.= Less than universitary)",
-                                      "Income per capita",
+                                      "Household income",
                                       # "High School",
@@
 
-df_study1$ola <- car::recode(df_study1$ola, c("1=1; 2=1;3=1;4=1; 6=2;7=2"))
-df_study1$ola <- as.factor(df_study1$ola)
+# df_study1$ola <- car::recode(df_study1$ola, c("1=1; 2=1;3=1;4=1; 6=2;7=2"))
+# df_study1$ola <- as.factor(df_study1$ola)
 #df_study1$covid_beneficios<- car::recode(df_study1$covid_beneficios, c("1=2; 2=1"))
@@
 
-h6 <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
-h1b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*ola"
-h3b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*ola"
-h4b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+des_perc*ola"
-h5b <- "ola+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+ing_pc_rec*ola"
+h6 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
+h1b <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*covid_economia"
+h3b <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*covid_economia"
+h4b <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+des_perc*covid_economia"
+h5b <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*covid_beneficios"
+h6b <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*covid_beneficios"
+h7b <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+des_perc*covid_beneficios"
 # A. justification educational inequality
-educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(ola+des_perc|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(ola+ing_pc_rec|idencuesta)")),data = df_study1, weights = ponderador_long_total)
-educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ0 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ4 <- lmer(formula(paste0("just_educ~",h4b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ5 <- lmer(formula(paste0("just_educ~",h5b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ6 <- lmer(formula(paste0("just_educ~",h6b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
+educ7 <- lmer(formula(paste0("just_educ~",h7b,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)
 
 # table 3: interacciones
-omit <- "(education)|(ess)|(edad)|(sexo)|(pos_id)|(covid_beneficios)|(covid_economia)"
-texreg::knitreg(list(educ6,educ3,educ1, educ4, educ5),
+omit <- "(education)|(ing_pc_rec)|(ess)|(edad)|(sexo)|(pos_id)|(covid_economia)|(des_perc)"
+texreg::knitreg(list(educ0,educ6, educ5),
                 custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
@@
                 include.aic = FALSE,
-                custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11", "Model 12"),
+                custom.model.names = c("Model 8", "Model 9", "Model 10"),
                 custom.coef.names = c("Intercept",
                                       "Wave",
-                                      "Income",
-                                      "Inequality perception",
+                                      "Wave^2",
+                                      "Covid: receive state benefits",
                                       "Merit: Talent",
                                       "Merit: Effort",
-                                      "Wave * Merit: Talent",
-                                      "Wave * Merit: Effort",
-                                      "Wave * Inequality perception",
-                                      "Wave * Income"
+                                      "Receive state benefits * Merit: Talent",
+                                      "Receive state benefits * Merit: Effort"
                                       ),
@@
                                      "Var: Individuals (Intercept)",
-                                     "Var: Residual",
-                                     "Var: Individuals Wave",
-                                     "Var: Individuals Merit: Talent",
-                                     "Cov: Individuals (Intercept) Wave",
-                                     "Cov: Individuals (Intercept) Merit: Talent",
-                                     "Cov: Individuals Wave Merit: Talent",
-                                     "Var: Individuals Merit: Effort",
-                                     "Cov: Individuals (Intercept) Merit: Effort",
-                                     "Cov: Individuals Wave Merit: Effort",
-                                     "Var: Individuals Inequality perception",
-                                     "Cov: Individuals (Intercept) Inequality perception",
-                                     "Cov: Individuals Wave Inequality perception",
-                                     "Var: Individuals Income",
-                                     "Cov: Individuals (Intercept) Income",
-                                     "Cov: Individuals Wave Income"),  
+                                     "Var: Residual"),  
                center = TRUE,
@@
 
-In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of educational inequality over time.
-In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic.
-We test this hypothesis through interaction effects, displayed in @tbl-interact.
-Model 1 is shown as a baseline model, it is the same as Model 8 in @tbl-interact but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for).
-Model 9 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 10 does the same but now for the perception of meritocracy related to effort.
-As observed, only the talent variable shows a significant interaction with the time variable ($\beta$=-0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on talent and the justification of inequality decreases by 0.01 points in every measurement point.
+In this last part of the analysis, we test hypotheses about the effect of COVID-19 on changes in the relationship between meritocracy and justification of educational inequality.
+In hypothesis 5, we proposed that the association between the perception of meritocracy and justification of inequality is mitigated in times of crisis, as meritocratic ideals might have been weakened due to critical situations associated with the COVID-19 pandemic.
+We test this hypothesis through interaction effects, shown in @tbl-interact.
+The first model is shown as the base model.
+It is the same as model 8 in @tbl-multilevel, but only shows the variables involved in the interaction for space reasons (all other variables are controlled).
+Model 9 adds the interaction between the perception of talent-based meritocracy and whether state benefits were received during the pandemic.
+Meanwhile, model 10 does the same, but now, it is for the perception of effort-related meritocracy.
+As can be seen, both variables show a negative significant interaction with receiving state benefits ($\beta_{talent}$=-0.07, p<.01; $\beta_{effort}$=-0.05, p<.05), which means that, on average, the association between the perception of talent-based meritocracy and the justification of inequality is lower by 0.07 points for those who received state benefits and the association between the perception of effort-based meritocracy and the justification of inequality is lower by 0.05 points for those who received state benefits.
+
+```{r echo=FALSE, fig.align='center'}
+#| label: fig-interact-talent
+#| fig-cap: "Interaction effect of merit based on talent and receive state benefits on justification of educational inequality"
+
+ggeffects::ggpredict(educ6, terms = c("merit_talent", "covid_beneficios")) %>%
+  ggplot(aes(x=x, y=predicted, shape = group, color = group)) +
+  geom_line(aes(group=group,linetype = group),position = position_dodge(.1)) + 
+  geom_point(size = 2.5,position = position_dodge(.1))+
+  scale_x_continuous(name = "Merit: talent", breaks=c(1,2,3,4,5), labels = c("Strongly
+disagree", "Disagree", "Nor agree
+nor disagree", "Agree", "Strongly
+agree")) + 
+  scale_shape_discrete(name = "Receive state
+benefits",
+                       limits = c("1", "2"),
+                       labels = c("No", "Yes")) +
+  scale_color_manual(name = "Receive state
+benefits",
+                     limits = c("1", "2"),
+                     labels = c("No", "Yes"),
+                     values = c("black", "black")) +
+  scale_linetype_manual(name = "Receive state
+benefits",
+                        limits = c("1", "2"),
+                        labels = c("No", "Yes"),
+                        values = c("solid", "dashed"))+
+  theme_bw() +
+  labs(title="", y = "Predics values of justification of educational inequality") + 
+  theme(plot.title = element_text(size = 11),
+        axis.text=element_text(size=11))
+```
+
+To further explore this interaction, the interaction between receiving state benefits and the perception of meritocracy based on talent can be visualized in @fig-interact-talent.
+This figure shows that the positive effect of perceiving greater meritocracy based on talent on the justification of educational inequality is smaller for those who received state benefits during the pandemic.

### 2025-01-13 07-apendice.qmd
add covid interactions

@@
                                       "Universitary (Ref.= Less than universitary)",
-                                      "Income",
+                                      "Household Income",
                                       # "High School",

### 2025-01-14 01-introduction.qmd
render paper and minor changes

@@
 Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].
-Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.
+Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system [@donoso_when_2016; @reyes-housholder_chile_2019].
 

### 2025-01-14 03-methods.qmd
render paper and minor changes

@@
 Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1737.
+In broader terms, the accumulated attrition between wave 1 and wave 7 is 40%, achieving a final sample of N = 1741.
 The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
@@
 
-A third group of variables is included in the analysis to assess the effect of elements related to the pandemic.
+A third group of variables is included in the analysis to assess the effect of elements related to the pandemic context.
 These variables were only included in the 2021 wave, so each individual's responses were replicated for the remaining years.
@@
 
-For testing the hypothesis regarding status we included educational level and household income quintiles.
+For testing the hypothesis regarding status we included educational level and household income.
 Also, we included subjective social status as it has been argued that perceived social status is a relevant predictor of attitudes toward economic inequality [@schneider_poverty_2015;@castillo_meritocracia_2019].

### 2025-01-14 04-analisis.qmd
render paper and minor changes

@@
 Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.5% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
+For instance, of the 32.9% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
 In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
 Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last waves.
-This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11.1% in 2023).
+This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11% in 2023).
 
@@
 In this matrix, the correlations vary between low and moderate values.
-Justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.11, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).
+Justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.3, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
+Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.16, p<.01; _r_=-0.11, p<.01).
+Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.75, p<.01).
 

### 2025-01-14 production/prod_analysis-fix-t5.Rmd
render paper and minor changes

@@
                                                just_educ, covid_economia, covid_beneficios, des_perc, merit_talent, merit_effort,
-                                               educ, quintil1:sexo)
+                                               educ, ing_pc, ess:sexo)
+colSums(is.na(df_study1))
+
+df_study1 <- df_study1[complete.cases(df_study1$covid_economia), ]
+df_study1 <- df_study1[complete.cases(df_study1$ing_pc), ]
+df_study1 <- df_study1[complete.cases(df_study1$pos_id), ]
+df_study1 <- df_study1[complete.cases(df_study1$covid_beneficios), ]
+df_study1 <- df_study1[complete.cases(df_study1$ess), ]
+
+sjmisc::frq(df_study1$ola)
 
@@
 sjlabelled::set_label(x = df_study1$covid_economia, 
-                      label = "It is more important to give priority to economic activity than to the health of the population")
+                      label = "In the context of the coronavirus pandemic, it is more important to give priority to economic activity than to the health of the population")
 
@@
 
+#
+df_study1$educ <- 
+sjlabelled::set_label(x = df_study1$educ, 
+                      label = "Educational level")
+
+#
+df_study1$edad <- 
+sjlabelled::set_label(x = df_study1$edad, 
+                      label = "Age groups")
+
+#
+df_study1$pos_id <- 
+sjlabelled::set_label(x = df_study1$pos_id, 
+                      label = "Political identification")
+
 lab_study1 <- sjlabelled::get_label(df_study1) 
@@
 
-df<- dfSummary(x = dplyr::select(df_study1, ola, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, educ, quintil1, ess, edad, sexo, pos_id),
+df<- dfSummary(x = dplyr::select(df_study1, ola, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, educ, ing_pc, ess, edad, sexo, pos_id),
                plain.ascii = FALSE,
@@
 df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent,covid_economia, covid_beneficios,
-                                               educ, quintil1:sexo) %>% na.omit()
+                                               just_educ, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, educ, ing_pc, ess, edad, sexo, pos_id) %>% na.omit()
 df_study1 <- df_study1[complete.cases(df_study1$segmento), ]
@@
   scale_x_discrete(labels = c(2016, 2017, 2018, 2019, 2022, 2023)) +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7894)")
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7576)")
 
@@
 df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent, covid_economia, covid_beneficios,
-                                               educ, quintil1:sexo) %>% na.omit()
+                                               just_educ, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, 
+                                               educ, ing_pc, ess, edad, sexo, pos_id) %>% na.omit()
 merit_plot <- df_study1 %>% 
@@
         plot.caption = element_text(hjust = 1)) +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7894)")
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7576)")
 merit_plot
@@
 df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent, covid_economia, covid_beneficios,
-                                               educ, quintil1:sexo) %>% na.omit()
+                                               just_educ, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, 
+                                               educ, ing_pc, ess, edad, sexo, pos_id) %>% na.omit()
 df_study1 <- df_study1[complete.cases(df_study1$segmento), ]
@@
   geom_text(data = label_data, aes(label_x, label_y, label = variable), color = "black", hjust = 0, nudge_x = 1.1, nudge_y = 0, size = 4)  +
-  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7894)")
+  labs(caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 7576)")
 
@@
 
-```{r cor-plot, fig.dim=c(10,6)}
+```{r cor-plot, fig.dim=c(12,7)}
 rm(list=ls())
@@
 df_study1 <- df_study_long_t7_comunas %>% dplyr::select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-                                               just_educ, des_perc, merit_effort, merit_talent, covid_economia, covid_beneficios,
-                                               educ, quintil1:sexo) %>% na.omit()
+                                               just_educ, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, 
+                                               educ, ing_pc, ess, edad, sexo, pos_id) %>% na.omit()
 # Full sample_______________________________________________________________________
@@
 #Plot the matrix using corrplot
-png("output/graphs/corr.png",width=600,height=300, pointsize = 10)
+png("output/graphs/corr.png",width=700,height=600, pointsize = 10)
 corrplot::corrplot(cor_full,
@@
 grid::grid.text("Source: own elaboration with pooled data 
-from ELSOC 2016-2023 (n = 7894)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
+from ELSOC 2016-2023 (n = 7576)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
 dev.off()
-# Wave 1_______________________________________________________________________
 # mat_1<- 

### 2025-01-14 05-discussion.qmd
revpaper -  theory, methods & conclusions

@@
 
-Finally, as our first hypothesis supported, we expected that the relationship between meritocratic perceptions and justification would tend to diminish over time.
-However, longitudinal evidence shows that the perception of meritocracy linked to effort strengthens its relation with the justification of inequality in education, against the temporal change hypothesis ($H_5$).
-In other words, meritocracy seems to gain force concerning justifying inequality and not the other way around as we expected given the social context in the post-covid era.
-An alternative interpretation is that in times of high exposure to risk, people tend to stick more to narratives that provide greater security regardless of the rest of the society - such as individual effort -, which might also be accompanied by lower degrees of solidarity to safeguard the well-being of themselves and their surrounding close peers.
-Again, this certainly needs further studies that include additional specific conceptual and methodological approaches to this phenomenon.
+Our first hypothesis was supported, as we anticipated that the relationship between perceptions of meritocracy and the justification of educational inequality would tend to diminish over time.
+We based this on questions that related directly to experiences during the pandemic, such as receiving economic benefits during the lockdown period.
+Our findings revealed that those who received benefits justified educational inequality less than those who did not.
+Moreover, we observed that the connection between perceived meritocracy and the justification of educational inequality weakened for those who received benefits.
+This evidence partially supports our temporal change hypothesis ($H5$).
+In other words, the influence of meritocracy on justifying inequality appears to lessen, which aligns with our expectations given the social context in the post-COVID era.
+One possible interpretation is that, during times of increased risk, people may become more skeptical of narratives emphasizing individual effort over collective efforts.
+This shift in perspective might also foster stronger solidarity to protect individual and societal wellbeing.
+However, further studies are necessary to explore this phenomenon using more specific conceptual and methodological approaches.
 

### 2025-01-15 01-introduction.qmd
abstract , intro, antecedentes JC

@@
 Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated.
-As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
+As the last PISA report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
 Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
@@
 We examine this association in the context of COVID 19 in Chile, a period characterized by several challenges regarding the political and economic system that could have affected the way in which the role of individual performance and social outcomes are deemed.
-
-Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background.
-In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022].
-This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
-Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile during the COVID pandemic, and to what extent this phenomenon is associated with the perception of meritocracy and its variations.
 
@@
 A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises.
+This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008].
 Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they would be the most affected in critical times.
-
-Against this backdrop, Chile serves as an example among Latin American countries due to its strong sanitary measures.
-These measures significantly impacted the labor market, leading to important consequences for unemployment and income distribution.
-It is importantto approach any claims about the influence of the COVID-19 pandemic with caution.
-Our findings are situated within a broader context of increasing socioeconomic risks and vulnerabilities, which were heightened by the sanitary restrictions implemented during the pandemic.

### 2025-01-15 02-antecedentes.qmd
abstract , intro, antecedentes JC

@@
 
-In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items has been diverse, and there is not a single instrument or scale dedicated specifically to this concept.
+In the area of survey studies, the operationalization of educational inequality justification in terms of questionnaire items has been diverse, and there is not a single instrument or scale dedicated specifically to this concept.
 Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables.
@@
 Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.
+Although such items covering inequality justification are present in emblematic comparative studies, still there are few studies focusing specifically in the justification of educational inequality.
 
@@
 
-The relationship between performance and outputs brings about to the concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
+Whether inequalities in education are considered just or not lead us almost naturally to the concepts of merit and meritocracy.
+Meritocracy, a concept coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
+Young's term was initially used to criticize a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
 However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
+This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often considered inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
 
-Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find related research which incorporates the support of different distributive principles with regard to education.
+Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find research which incorporates the support of different distributive principles with regard to education.
 For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
+They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
 In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health.
-However, the relationship between meritocracy and the justification of inequality within the educational domain remains underexplored.
-For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit.
-Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
 
@@
 
-_This study_
+Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health.
+For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit.
+Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
 
-The present research aims at identifying changes in the justification of educational inequalities in Chile, with a particular focus in its relationship with meritocratic perceptions.
-The time frame for this analysis covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile).
+_The COVID pandemic and public concerns about inequality_
+
+The time frame for this study covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile).
 Although this period comprises the COVID pandemic, the changes in the variables analyzed are not possible to be attributed only to this event in terms of a health crisis, as several economic, social and cultural processes overlapped worldwide.
-Besides, in the specific case of Chile there were two constitutional processes happening in this very same period after a massive social upheaval.
+However, and as several studies have shown, this period appeared to trigger attitudinal changes that are relevant to explore, both in themselves and as antecedent to future studies.
+In this sense, in this paper we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
 
-As there are measures before and after the pandemic period, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame.
-Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].
+A shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions[@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,].
+In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6.
+In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade.
+By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.
 
-Within this framework, we point out that periods of rising socioeconomic and health vulnerabilities could weaken the justification of social inequalities.
-This change would be at least in part linked to a challenge of the meritocratic ideals.
-Chile provides a significant case for studying inequality, as it exemplifies a prosperous yet highly unequal society within Latin America.
-Under such conditions, a shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions.
 Chile implemented sanitary controls early in the pandemic, initiating lockdowns on March 18, 2020, and lifting all restrictions by March 5, 2023.
@@
 This placed Chile 36th among 184 countries and 4th in South America, following Peru (60.1), Suriname (60.5), and Venezuela (80.5).
+Besides, an important cycle of protests known as the "social outburst" took place between October 2019 and March 2020 [@somma_no_2021].
+As a consequence there were two constitutional processes happening in this very same period after the massive social upheavals.
+However, the protest drastically decreased as a result of the lockdown [@joignant_informe_2020].
 
-Within the period analyzed (2016-2022), an important cycle of protests known as the "social outburst" took place between October 2019 and March 2020 [@somma_no_2021].
-However, the protest drastically decreased as a result of the lockdown [@joignant_informe_2020].
 In terms of public views on economic inequality, it has been documented that critical views toward inequality increased in 2019 and deepened under the pandemic [@coes_radiograficambio_2023], which we argue can be in part explained by the increased risk exposure and vulnerability in the context of the COVID-19 pandemic.
-Hereby, a relevant aspect is the transformation of the labor market and its impact on income distribution [@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,].
-In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6.
-In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade.
-By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.
-
 Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period.
@@
 
-Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019].
+Our second argument is based on the evidence that relates rational interests to justifications and preferences [@sonhing_failure_2019].
 Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps.
-However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.
+For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps.
 In line with this evidence, we expect that:

### 2025-01-15 paper.qmd
abstract , intro, antecedentes JC

@@
 However, despite the widespread value of education, disparities in educational outcomes persist over time.
-To what extent are such educational disparities justified in society?
+This paper is guided by the following research questions: To what extent are such educational disparities justified in society?
+Is the justification of educational inequality affected in periods of personal and social vulnerability, as in the COVID pandemic?
 And, what are the main factors driving this kind of justification?
-This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy.
 The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927).
+This economic and cultural environment has promoted meritocratic ideals, whereby individual talent and effort are considered key to get ahead in life, disregarding opportunities linked to social origin.
+The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic).
+Furthermore, we argue that such changes could be linked to a challenge of meritocratic ideals.
+For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2023, 6 waves, N = 2,927).
 Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.

### 2025-01-17 02-antecedentes.qmd
reestructuracion antecedentes

@@
 Although such items covering inequality justification are present in emblematic comparative studies, still there are few studies focusing specifically in the justification of educational inequality.
-
-_Meritocracy and educational inequalities_
-
-Whether inequalities in education are considered just or not lead us almost naturally to the concepts of merit and meritocracy.
-Meritocracy, a concept coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
-Young's term was initially used to criticize a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
-This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often considered inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
-
-Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find research which incorporates the support of different distributive principles with regard to education.
-For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
-They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
-
-More specific empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
-This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
-Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
-However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
-This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
-
-Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health.
-For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit.
-Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
 
@@
 
-Our second argument is based on the evidence that relates rational interests to justifications and preferences [@sonhing_failure_2019].
-Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps.
-In line with this evidence, we expect that:
+Besides proposing a general decrease over time in the justification of inequalities, a second hypothesis refers to specific aspects of the pandemic.
+As some political measures from the state included several social benefits (as delivering food boxes, free vaccination, bonuses, among others), we argue that being recipient of such benefits could affect attitudes towards the redistributive role of the state as equalizer opportunities to face risks.
+As a consequence, preferences for market justice would be reduced:
 
-$H_2$: Those with lower status characteristics show lower justification of educational inequalities
+$H_2$: Recipient of state benefits show lower market justice preferences
 
-Thirdly, we address inequality perception.
+In the same line, we expect that those who give priority to take care of the health of the population instead of the country economy would exhibit larger market justice preferences:
+
+$H_3$: Those prioritizing health instead of the economic situation show higher market justice preferences
+
+_Meritocracy and educational inequalities_
+
+Whether inequalities in education are considered just or not lead us to take into account the concepts of merit and meritocracy.
+Meritocracy, a concept coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning.
+Young's term was initially used to criticize a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege.
+However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020].
+This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often considered inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].
+
+Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find research which incorporates the support of different distributive principles with regard to education.
+For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
+They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
+In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
+
+More specific empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
+Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
+This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
+Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes.
+However, empirical findings also reveal nuances in this relationship.
+For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
+This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
+
+Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health.
+For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit.
+Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
+In line with this evidence, we propose that
+
+$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
+
+Combining the arguments regarding the effect of the pandemic and its relationship with meritocracy, we explore interactions between meritocracy and inequality justification over time.
+As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged in the time frame analyzed.
+As a consequence, the justification of inequality could show a decrease over time:
+
+$H_5$: The positive association between meritocracy and inequality justification mitigates along time
+
+Along with meritocratic perceptions, in this study we incorporate inequality perception.
 As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
@@
 
-$H_3$: A larger inequality perception is negatively associated with the justification of educational inequalities.
+$H_6$: A larger inequality perception is negatively associated with the justification of educational inequalities.
 
-Our fourth hypothesis refers to the central concept of meritocracy. As meritocracy justifies inequalities based on performance, we expect that:
+Next, we take into account the evidence that relates rational interests to justifications and preferences [@sonhing_failure_2019].
+Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
+For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps.
+In line with this evidence, we expect that:
 
-$H_4$: Higher meritocratic perceptions are positively related to the justification of educational inequalities.
-
-Finally, and in line with hypotheses 1 and 4, we explore interactions between meritocracy and inequality justification over time.
-As COVID-19 pandemic has affected the structural opportunities to attend schools, universities, or workplaces, the ideal of meritocracy might have been challenged in the time frame analyzed.
-As a consequence, the justification of inequality could show a decrease over time:
-
-$H_5$: The positive association between meritocracy and inequality justification mitigates along time
+$H_7$: Those with lower status characteristics show lower justification of educational inequalities

### 2025-01-17 03-methods.qmd
revpaper -  theory, methods & conclusions
methods & descriptives

@@
 
-The main data source is the Chilean Longitudinal Social Survey 2016–2023.
-The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
+The main data source is the Chilean Longitudinal Social Survey ELSOC (2016–2023).
+ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile.
 The sampling design is probabilistic, stratified, clustered, and multistage.
 It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities.
-Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1.
-In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
+The sample covers participants aged between 18 and 75 years (in wave 1), representing  77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].
 
 The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic.
-The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI).
+The waves 2016, 2017, 2018, 2019, 2022, and 2023 were administered using computer-assisted personal interviewing (CAPI).
 However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021.
 In addition, wave 3 included a refreshment sample to counter survey attrition.
-The same sampling strategy as that of wave 1 was implemented for selecting new cases.
 As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample.
-The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends.
+The data from the refreshment sample is not included in the analytical sample for this study as we analyze longer response trends.
 Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants.
-In broader terms, the accumulated attrition between wave 1 and wave 7 is 40%, achieving a final sample of N = 1741.
+In broader terms, the accumulated attrition between wave 1 and wave 7 is 40%, achieving a final sample of N = 1,741.
 The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave.
@@
 
-For testing the hypothesis regarding status we included educational level and household income.
+For testing the hypothesis regarding status we considered educational level and household income.
 Also, we included subjective social status as it has been argued that perceived social status is a relevant predictor of attitudes toward economic inequality [@schneider_poverty_2015;@castillo_meritocracia_2019].
-In addition, subjective measures can complement objective measures in predicting life chances [@oesch_subjective_2023], which are arguably connected to how individuals experience economic inequality.
+In addition, such subjective measures can complement objective measures in predicting life chances [@oesch_subjective_2023], which are arguably connected to how individuals experience economic inequality.
 In this regard, it has been argued that how people form their views on their standing in society results from experiences with direct socioeconomic circumstances and social comparison processes with reference groups [@Condon2020].

### 2025-01-17 04-analisis.qmd
methods & descriptives

@@
 # Results
+
+Beginning with a descriptive analysis of changes in the justification of educational inequality between 2016 and 2023, @fig-alluvial illustrates yearly frequencies and transitions between years.
+Each year represents stacked porcentual frequencies, whereas the flows in between reflect the within-subject change of opinions from one year to the next [@rosvall_mapping_2010].
+For instance, of the 32.9% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
+In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
+Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last waves.
+This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11% in 2023).
 
@@
 
-@fig-alluvial illustrates yearly frequencies in the justification of educational inequality between 2016 and 2023.
-Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010].
-For instance, of the 32.9% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories.
-In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years.
-Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last waves.
-This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11% in 2023).
+@fig-means shows the average changes in the main variables considered for this study.
+Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
+Interestingly, in the last waves of the study (2022-2023) the justification of inequality increases whereas the perception of inequality decreases.
+As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
 
@@
 
-@fig-means shows the average changes in the main variables considered for this study.
-Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception.
-Interestingly, in the last waves of the study (2022-2023) the justification of inequality increases whereas the perception of inequality decreases.
-As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.
+@fig-correlation presents a correlation matrix of the main variables analyzed, using data from all survey waves.
+In this matrix, the correlations vary between low and moderate values.
+Justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.3, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
+Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.16, p<.01; _r_=-0.11, p<.01).
+Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.75, p<.01).
 
@@
 
-@fig-correlation presents a correlation matrix of the main variables analyzed, using data from all survey waves.
-In this matrix, the correlations vary between low and moderate values.
-Justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.3, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.16, p<.01; _r_=-0.11, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.75, p<.01).
-
-## Multivariate analysis
+_Multilevel models_
 
@@
 Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.167, p<.001), 2018 ($\beta$=-0.059, p<.05), and 2019 ($\beta$=-0.03, p>.05).
-Nevertheless, in the last waves of 2022 and 2023 there is a radical increase in level of justification of educational inequality ($\beta$=0.204, p<.001 and $\beta$=0.233, p<.001), suggesting a non-linear change in this variable.
+Nevertheless, in the last waves of 2022 and 2023 there is a clear increase in level of justification of educational inequality ($\beta$=0.204, p<.001 and $\beta$=0.233, p<.001), suggesting a non-linear change in this variable.
 Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1.
-On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.
+On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time.
+On the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.
 

### 2025-01-17 02-antecedentes.qmd
ajustes antecedentes

@@
 
-A shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions[@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,].
-In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6.
-In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade.
-By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.
+In Chile, the COVID-19 pandemic posed a significant threat to economically vulnerable groups and those in precarious labor market positions [@ministerio_de_desarrollo_social_y_familia_resultados_2022; @instituto_nacional_de_estadisticas_resultados_2022].
+The poverty rate, which was 11.2% in 2015, decreased to 8.5% by 2017, with a 10/10 ratio of 31.6.
+However, during the pandemic in 2020, poverty rose to 10.7%, the 10/10 ratio spiked to 301.7, and unemployment reached a decade-high of 10.5%.
+By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, with a slight rise to 8.7% in 2023.
 
@@
 Besides proposing a general decrease over time in the justification of inequalities, a second hypothesis refers to specific aspects of the pandemic.
-As some political measures from the state included several social benefits (as delivering food boxes, free vaccination, bonuses, among others), we argue that being recipient of such benefits could affect attitudes towards the redistributive role of the state as equalizer opportunities to face risks.
-As a consequence, preferences for market justice would be reduced:
+As some political measures from the state included several social benefits (as delivering food boxes, free vaccination, direct transfers, among others), we argue that being recipient of such benefits could affect attitudes towards the redistributive role of the state as equalizer opportunities to face risks.
+As a consequence, preferences for educational inequalities would be reduced:
 
-$H_2$: Recipient of state benefits show lower market justice preferences
+$H_2$: Recipient of state benefits show lower justification of educational inequalities
 
-In the same line, we expect that those who give priority to take care of the health of the population instead of the country economy would exhibit larger market justice preferences:
+In the same line, we expect that those who give priority to take care of the health of the population instead of the country economy would exhibit larger preferences for educational inequalities:
 
-$H_3$: Those prioritizing health instead of the economic situation show higher market justice preferences
+$H_3$: Those prioritizing health instead of the economic situation show higher justification of educational inequalities
 
@@
 They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.
+In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted perceived fairness of educational inequality.
 
@@
 Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
-In line with this evidence, we propose that
+In line with this evidence, we propose that:
 

### 2025-01-18 02-antecedentes.qmd
reorder vars
ajusta citations keys
ajusta n descriptivos
cambia orden tabla multinivel
render paper

@@
 A strand of this research has focused on specific policy areas such as health, pensions, and education,  attending to the degree of justification of market-based access to social services.
-For instance, @lindhPublicOpinionMarkets2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
+For instance, @lindh_public_2015 found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services.
 Similarly, @busemeyer_welfare_2020 shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes.
@@
 
-In Chile, the COVID-19 pandemic posed a significant threat to economically vulnerable groups and those in precarious labor market positions [@ministerio_de_desarrollo_social_y_familia_resultados_2022; @instituto_nacional_de_estadisticas_resultados_2022].
+In Chile, the COVID-19 pandemic posed a significant threat to economically vulnerable groups and those in precarious labor market positions [@ministeriodedesarrollosocialyfamilia_resultados_2023; @institutonacionaldeestadisticas_resultados_2022].
 The poverty rate, which was 11.2% in 2015, decreased to 8.5% by 2017, with a 10/10 ratio of 31.6.
@@
 
-In terms of public views on economic inequality, it has been documented that critical views toward inequality increased in 2019 and deepened under the pandemic [@coes_radiograficambio_2023], which we argue can be in part explained by the increased risk exposure and vulnerability in the context of the COVID-19 pandemic.
+In terms of public views on economic inequality, it has been documented that critical views toward inequality increased in 2019 and deepened under the pandemic [@coes_radiografia_2023], which we argue can be in part explained by the increased risk exposure and vulnerability in the context of the COVID-19 pandemic.
 Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period.
@@
 Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find research which incorporates the support of different distributive principles with regard to education.
-For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
+For instance, in an experiment by @igliozzi_fair_2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education.
 They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school').
-In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted perceived fairness of educational inequality.
+In the same line, @lee_fairness_2023 show that performance assessments (in what they call “neoliberal orientations”) predicted perceived fairness of educational inequality.
 
 More specific empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs.
-Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007].
+Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@mcnamee_meritocracy_2004].
 This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023].
@@
 However, empirical findings also reveal nuances in this relationship.
-For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023].
+For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@day_perceived_2023].
 This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.
@@
 For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit.
-Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
+Similarly, @castillo_percepcion_2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.
 In line with this evidence, we propose that:
@@
 Along with meritocratic perceptions, in this study we incorporate inequality perception.
-As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
-For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S.
+As pointed out by @janmaat_subjective_2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023].
+For instance, a recent study by @day_perceived_2023 on the perception and justification of inequality in university endowments in the U.S.
 points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality.
@@
 Next, we take into account the evidence that relates rational interests to justifications and preferences [@sonhing_failure_2019].
-Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023].
-For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps.
+Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@busemeyer_welfare_2020].
+For instance, @valant_politics_2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps.
 In line with this evidence, we expect that:

### 2025-01-18 03-methods.qmd
reorder vars
ajusta citations keys
ajusta n descriptivos
cambia orden tabla multinivel
render paper

@@
 In addition, such subjective measures can complement objective measures in predicting life chances [@oesch_subjective_2023], which are arguably connected to how individuals experience economic inequality.
-In this regard, it has been argued that how people form their views on their standing in society results from experiences with direct socioeconomic circumstances and social comparison processes with reference groups [@Condon2020].
+In this regard, it has been argued that how people form their views on their standing in society results from experiences with direct socioeconomic circumstances and social comparison processes with reference groups [@condon_inequality_2020].
 Finally, political identification on a left-right scale, age groups, and gender were included as control variables.

### 2025-01-18 04-analisis.qmd
reorder vars
ajusta citations keys
ajusta n descriptivos
cambia orden tabla multinivel
render paper

@@
 
-@fig-correlation presents a correlation matrix of the main variables analyzed, using data from all survey waves.
+@fig-correlation presents a correlation matrix of the main variables analyzed, using data from the first wave.
 In this matrix, the correlations vary between low and moderate values.
-Justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.3, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
+The correlations between justification of educational inequality with the two covid variables are positive, but low.
+In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.3, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
 Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.16, p<.01; _r_=-0.11, p<.01).
@@
 
-Model 3 adds the socioeconomic variables to test hypothesis 2.
+Models 3 and 4 adds the two variables about the Covid context, where we can see that those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.093, p<.01).
+On the other hand, those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.038, p<.05).
+
+Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
+In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.055, p<.001).
+However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
+In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.083, p<.001), but it has a larger weight than the perception of talent.
+Inequality perception is added in Model 7, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
+
+Model 8 adds the socioeconomic variables to test hypothesis 7.
 Educational level and household per capita income were added, but none of them had a significant effect.
 However, in the next models, controlling for all variables, household per capita income had a positive significant effect ($\beta$=0.001, p<.05).
-In addition, from this model onwards the demographic control variables were added, where being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001), and being women (compared to men) had a negative and significant effect ($\beta$=-0.08, p<.001).
+In addition, in this model the demographic control variables were added, where being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001), and being women (compared to men) had a negative and significant effect ($\beta$=-0.08, p<.001).
 Subjective social status and age had no significant effect when controlling for all the variables analysed.
 The full models can be seen in the appendix (see @tbl-multilevel-full).
-
-Models 4 and 5 add the two variables about the Covid context, where we can see that those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.038, p<.05).
-On the other hand, those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.093, p<.01).
-
-Inequality perception is added in Model 6, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
-Model 7 and 8 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society).
-In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 7 ($\beta$=0.044, p<.001).
-However, when controlling for the perception that effort is rewarded, this effect is no longer significant.
-In this sense, Model 8 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.083, p<.001), but it has a larger weight than the perception of talent.
 
@@
 h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia"
-h5 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
-h6 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
-h7 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
+h2 <- "ola_num+ola_2+covid_beneficios"
+h3 <- "ola_num+ola_2+covid_beneficios+covid_economia"
+h4 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent"
+h5 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent+merit_effort"
+h6 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent+merit_effort+des_perc"
+h7 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent+merit_effort+des_perc+education+ing_pc_rec+ess+edad+sexo+pos_id"
+
 h8 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
@@
 omit <- "(edad)|(ess)|(sexo)|(pos_id)"
-texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
+texreg::knitreg(list(educ0, educ1, educ2, educ3, educ4, educ5, educ6, educ7)
+                     ,
                 digits = 3,
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 3 to 8 are controlled by age, gender, political position, and subjective social status.", 
+Note: Model 6 is controlled by age, gender, political position, and subjective social status.", 
                 omit.coef = omit,
@@
                                       "Wave^2",
+                                      "Covid: receive state benefits",
+                                      "Covid: economy more important than health",
+                                      "Merit: Talent",
+                                      "Merit: Effort",
+                                      "Inequality perception",
                                       "Universitary (Ref.= Less than universitary)",
-                                      "Household income",
+                                      "Household income"
                                       # "High School",
@@
                                       # "Does not identify",
-                                      "Covid: economy more important than health",
-                                      "Covid: receive state benefits",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort"),
+                                      ),
                 groups = list("Wave (Ref.= Wave 2016)" = 2:6
@@
                 include.aic = FALSE,
-                custom.gof.rows = list("Controls"= c("No", "No", rep("Yes", 6))),
+                custom.gof.rows = list("Controls"= c(rep("No", 7), "Yes")),
                 custom.gof.names = c("BIC",
@@
   geom_point(size = 2.5,position = position_dodge(.1))+
+  scale_y_continuous(limits=c(1,5),  breaks=c(1,2,3,4,5))+
   scale_x_continuous(name = "Merit: talent", breaks=c(1,2,3,4,5), labels = c("Strongly
-disagree", "Disagree", "Nor agree
+disagree", "Disagree", "Neither agree
 nor disagree", "Agree", "Strongly

### 2025-01-18 05-discussion.qmd
reorder vars
ajusta citations keys
ajusta n descriptivos
cambia orden tabla multinivel
render paper

@@
 Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chile_2023;@sazo_chile_2023].
+This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chiles_2023;@sazo_chile_2023].
 

### 2025-01-18 07-apendice.qmd
reorder vars
ajusta citations keys
ajusta n descriptivos
cambia orden tabla multinivel
render paper

@@
 h1 <- "ola_num+ola_2"
-h2 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo"
-h3 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id"
-h4 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia"
-h5 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
-h6 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
-h7 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
-h8 <- "ola_num+ola_2+education+ing_pc_rec+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
+h2 <- "ola_num+ola_2+covid_beneficios"
+h3 <- "ola_num+ola_2+covid_beneficios+covid_economia"
+h4 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent"
+h5 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent+merit_effort"
+h6 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent+merit_effort+des_perc"
+h7 <- "ola_num+ola_2+covid_beneficios+covid_economia+merit_talent+merit_effort+des_perc+education+ing_pc_rec+ess+edad+sexo+pos_id"
 
@@
 #omit <- "(edad)|(ess)|(sexo)|(pos_id)"
-texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
+texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 
@@
                                       "Wave^2",
+                                      "Covid: receive state benefits",
+                                      "Covid: economy more important than health",
+                                      "Merit: Talent",
+                                      "Merit: Effort",
+                                      "Inequality perception",
                                       "Universitary (Ref.= Less than universitary)",
@@
                                        "Right",
-                                       "Does not identify",
-                                      "Covid: economy more important than health",
-                                      "Covid: receive state benefits",
-                                      "Inequality perception",
-                                      "Merit: Talent",
-                                      "Merit: Effort"),
+                                       "Does not identify"
+                                      ),
                 groups = list("Wave (Ref.= Wave 2016)" = 2:6,

### 2025-01-18 production/prod_analysis-fix-t5.Rmd
reorder vars
ajusta citations keys
ajusta n descriptivos
cambia orden tabla multinivel
render paper

@@
 st_options(lang = "en",
-           footnote = NA,
+           footnote = "Note: Descriptives presented only for wave 1",
            bootstrap.css = F,
@@
 # Df variables nivel 1 _________________________________________________________
-df_study1 <- df_study_long_t7_comunas %>% dplyr::select(ola, idencuesta,
+df_study1 <- df_study_long_t7_comunas %>% 
+  dplyr::select(ola, idencuesta,
                 just_educ, covid_economia, covid_beneficios, des_perc, merit_talent, merit_effort,
-                                               educ, ing_pc, ess:sexo)
+                educ, ing_pc, ess:sexo) %>% 
+  filter(ola==1)
 colSums(is.na(df_study1))
@@
 
-df_study1$ola <- factor(df_study1$ola,
-         labels = c("2016","2017","2018","2019", "2022", "2023"))
-df_study1$ola <- 
-sjlabelled::set_label(x = df_study1$ola,
-                      label = "Wave")
+# df_study1$ola <- factor(df_study1$ola,
+#          labels = c("2016","2017","2018","2019", "2022", "2023"))
+# df_study1$ola <- 
+# sjlabelled::set_label(x = df_study1$ola,
+#                       label = "Wave")
 
@@
 
+df_study1$ess <- 
+sjlabelled::set_label(x = df_study1$ess, 
+                      label = "Subjective social status")
+
+df_study1$ing_pc <- 
+sjlabelled::set_label(x = df_study1$ing_pc, 
+                      label = "Household income per capita")
+
 lab_study1 <- sjlabelled::get_label(df_study1) 
@@
 
-df<- dfSummary(x = dplyr::select(df_study1, ola, covid_economia, covid_beneficios, merit_talent, merit_effort, des_perc, educ, ing_pc, ess, edad, sexo, pos_id),
+df<- dfSummary(x = dplyr::select(df_study1, covid_economia, covid_beneficios, 
+                                 merit_talent, merit_effort, des_perc, 
+                                 educ, ing_pc, ess, edad, sexo, pos_id),
                plain.ascii = FALSE,
@@
 
-ggsave(alluvial_justeduc, file="output/graphs/alluvial_dep.png", height = 8, width = 7)
+ggsave(alluvial_justeduc, file="output/graphs/alluvial_dep.png", height = 7, width = 9)
 ```
@@
                                                educ, ing_pc, ess, edad, sexo, pos_id) %>% na.omit()
-# Full sample_______________________________________________________________________
+# Full sample______________________________________________________________________
 mat_full<- 
 df_study1 %>%
-  # filter(ola==1) %>%
+   filter(ola==1) %>%
   select(just_educ, # distributive justice
+         covid_economia,
+         covid_beneficios,
          des_perc,
@@
 rownames(cor_full) <- c("A. Justification of educational inequality",
-                        "B. Inequality perception",
-                     "C. People are rewarded for their efforts",
-                     "D. People are rewarded for their intelligence")
+                        "B. Economy more important than health",
+                        "C. Receive state benefits",
+                        "D. Inequality perception",
+                     "E. People are rewarded for their efforts",
+                     "F. People are rewarded for their intelligence")
 #set Column names of the matrix
-colnames(cor_full) <-c("(A)", "(B)","(C)","(D)")
+colnames(cor_full) <-c("(A)", "(B)","(C)","(D)", "(E)", "(F)")
 #Plot the matrix using corrplot
-png("output/graphs/corr.png",width=700,height=600, pointsize = 10)
+png("output/graphs/corr.png",width=700,height=600, pointsize = 12)
 corrplot::corrplot(cor_full,
@@
 # Agregar caption
-grid::grid.text("Source: own elaboration with pooled data 
-from ELSOC 2016-2023 (n = 7576)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 10, col = "black"))
+grid::grid.text("Source: own elaboration with data 
+from ELSOC 2016 (n = 1322)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 12, col = "black"))
 dev.off()

### 2025-01-18 04-analisis.qmd
arreglo de tabla apendice

@@
 The correlations between justification of educational inequality with the two covid variables are positive, but low.
-In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.3, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.16, p<.01; _r_=-0.11, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.75, p<.01).
+In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.34, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.29, p<.01; _r_=0.24, p<.01).
+Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.18, p<.01; _r_=-0.08, p<.01).
+Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.76, p<.01).
 
@@
 #| label: fig-correlation
-#| fig-cap: "Correlation matrix of justification of educational inequality, perception of inequality, and perception of meritocracy for all the years analyzed"
+#| fig-cap: "Correlation matrix of the main variables for the first wave (2016)"
 
@@
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
-Note: Model 6 is controlled by age, gender, political position, and subjective social status.", 
+Note: Model 8 is controlled by age, gender, political position, and subjective social status.", 
                 omit.coef = omit,
@@
                caption.above = TRUE,
+               caption = "", 
                scalebox = 0.765,
@@
                caption.above = TRUE,
+               caption = "",
                scalebox = 0.765,

### 2025-01-18 07-apendice.qmd
arreglo de tabla apendice

@@
  #               omit.coef = omit,
+                digits=3,
                 custom.coef.names = c("Intercept",
@@
                   #"Household Income (Ref.= Quintile 1)" = 12:16
-                  "Age (Ref.= 18-29)" = 12:14,
-                  "Pol. pos (Ref.= Left)" = 16:18
+                  "Age (Ref.= 18-29)" = 17:19,
+                  "Pol. pos (Ref.= Left)" = 21:23
                   ),

### 2025-01-19 04-analisis.qmd
analysis & discussion

@@
 In this matrix, the correlations vary between low and moderate values.
-The correlations between justification of educational inequality with the two covid variables are positive, but low.
+The correlations between justification of educational inequality with the two COVID variables are positive, but low.
 In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.34, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.29, p<.01; _r_=0.24, p<.01).
@@
 On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time.
-On the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.
+On the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement points.
 
-Models 3 and 4 adds the two variables about the Covid context, where we can see that those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.093, p<.01).
-On the other hand, those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.038, p<.05).
+Models 3 and 4 adds the two variables about the pandemic context, where we can see that those who received state benefits during this period justify less inequality in education compared to those who did not receive them.
+Although the effect size is small, it remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.093, p<.01).
+On the other hand, those who believe that in this context the economy is more important than the health of the population justify more inequality in education($\beta$=0.038, p<.05, Model 8).
 
@@
 In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.083, p<.001), but it has a larger weight than the perception of talent.
-Inequality perception is added in Model 7, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.
+Regarding inequality perception (added in Model 7), it depicts a negative association with the justification of educational inequality, remaining stable when controlling for the rest of the variables.
 
 Model 8 adds the socioeconomic variables to test hypothesis 7.
-Educational level and household per capita income were added, but none of them had a significant effect.
-However, in the next models, controlling for all variables, household per capita income had a positive significant effect ($\beta$=0.001, p<.05).
-In addition, in this model the demographic control variables were added, where being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001), and being women (compared to men) had a negative and significant effect ($\beta$=-0.08, p<.001).
-Subjective social status and age had no significant effect when controlling for all the variables analysed.
-The full models can be seen in the appendix (see @tbl-multilevel-full).
+Although educational level and household per capita income are not significant in this model, controlling for all variables in the next models, household per capita income depicts a positive effect ($\beta$=0.001, p<.05).
+The inclusion of other controls as being right-wing (compared to being left-wing) shows a positive association with justification of educational inequality ($\beta$=0.27, p<.001), whereas being a women (compared to men) had a negative  effect ($\beta$=-0.08, p<.001).
+Subjective social status and age had no significant effect when controlling for all the variables analyzed.
+The full model specification can be seen in the appendix (see @tbl-multilevel-full).
 
@@
 
+In this last part of the analysis we test hypotheses about the effect of COVID-19 on changes in the relationship between meritocracy and justification of educational inequality.
+In hypothesis 5, we proposed that the association between the perception of meritocracy and justification of inequality is mitigated in times of crisis, as meritocratic ideals might have been weakened due to critical situations associated with the COVID-19 pandemic.
+We test this hypothesis through interaction effects, shown in @tbl-interact.
+The first model is shown as the baseline model.
+It is the same as model 8 in @tbl-multilevel, but only shows the variables involved in the interaction for the sake of space (all other variables are included).
+Model 9 adds the interaction between the perception of talent-based meritocracy and whether state benefits were received during the pandemic, whereas Model 10 considers the perception of effort-related meritocracy for the interaction.
+Both variables show a negative significant interaction with receiving state benefits ($\beta_{talent}$=-0.07, p<.01; $\beta_{effort}$=-0.05, p<.05), meaning that for those who received state benefits the association between the the justification of inequality and the perception of effort-based meritocracy lower by 0.05 points, and by 0.07 points regarding the perception of talent-based meritocracy.
+This last interaction is depicted in @fig-interact-talent, showing that the positive association between the perception of talent-based meritocracy and the justification of educational inequality is smaller for those who received state benefits during the pandemic.
+
 ```{r interact, echo=FALSE, results='asis', warning=FALSE}
@@
 
-In this last part of the analysis, we test hypotheses about the effect of COVID-19 on changes in the relationship between meritocracy and justification of educational inequality.
-In hypothesis 5, we proposed that the association between the perception of meritocracy and justification of inequality is mitigated in times of crisis, as meritocratic ideals might have been weakened due to critical situations associated with the COVID-19 pandemic.
-We test this hypothesis through interaction effects, shown in @tbl-interact.
-The first model is shown as the base model.
-It is the same as model 8 in @tbl-multilevel, but only shows the variables involved in the interaction for space reasons (all other variables are controlled).
-Model 9 adds the interaction between the perception of talent-based meritocracy and whether state benefits were received during the pandemic.
-Meanwhile, model 10 does the same, but now, it is for the perception of effort-related meritocracy.
-As can be seen, both variables show a negative significant interaction with receiving state benefits ($\beta_{talent}$=-0.07, p<.01; $\beta_{effort}$=-0.05, p<.05), which means that, on average, the association between the perception of talent-based meritocracy and the justification of inequality is lower by 0.07 points for those who received state benefits and the association between the perception of effort-based meritocracy and the justification of inequality is lower by 0.05 points for those who received state benefits.
-
 ```{r echo=FALSE, fig.align='center'}
@@
 ```
-
-To further explore this interaction, the interaction between receiving state benefits and the perception of meritocracy based on talent can be visualized in @fig-interact-talent.
-This figure shows that the positive effect of perceiving greater meritocracy based on talent on the justification of educational inequality is smaller for those who received state benefits during the pandemic.

### 2025-01-19 05-discussion.qmd
analysis & discussion

@@
 
-Our research sought to examine the relationship between the perception of meritocracy and the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
+Our research sought to examine the justification of inequality in education from a longitudinal perspective amid the health crisis generated by the COVID-19 pandemic.
 In general, our results showed mixed evidence about our hypotheses.
-Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal.
 We argued that in times of crisis, greater exposure to risk due to the health context would result in a lower justification of inequality by citizens.
-In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the last measurement point (2022), after the peak of the COVID pandemic.
-As we mentioned at the beginning, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
-As in the Chilean case, the pandemic period occurred along with a constitutional process, disentangling different forces driving public opinion it becomes difficult and deserves further qualitative and quantitative studies.
-A closer look to the political environment during the last year in the country could give additional hints.
-Some recent studies have argued that the election of a far left-wing constitutional assembly during this period generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
+In this regard, longitudinal models showed that the justification of inequality in education is far from being linear, and whereas a decreasing pattern was found between 2017 and 2019, there is a striking increase in the post-pandemic measurement points.
+Besides, our findings revealed that those who received benefits justified educational inequality less than those who did not.
+Still, given our data limitations it is not possible to attribute such changes only to the pandemic and the related economic and social crises.
+As during this very same period in Chile there were two (failed) constitutional processes, disentangling different forces driving public opinion becomes cumbersome and deserves further qualitative and quantitative studies.
+Along this line, some recent studies have argued that the election of a far left-wing constitutional assembly in the first process generated a backslash effect, mainly due to several scandals that led to the delegitimation of this assembly.
 This could have driven preferences in a conservative direction, which was reflected in the election of a right-wing second assembly after the failed first constitutional process [@palanza_chiles_2023;@sazo_chile_2023].
 
-Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality ($H_2$).
-In this regard, our results show a positive and robust relationship between income and the justification of inequality, which would align with our original hypothesis based on the rational interests of the better-off that would lead to a larger inequality justification.
-Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude ($H_3$).
-Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education ($H_4$).
-At first look, the results are consistent with the hypothesis; however, some nuances are worth attending concerning different aspects of meritocracy, namely effort and talent.
-Our results are favorable for the effort dimension but not confirmed for the talent dimension.
+The main focus of this study was on the relationship between meritocracy and justification of inequality, arguing that those who perceive that the society in which they live complies with meritocratic principles, would be more inclined to justify inequality in education.
+At first look, the results are consistent with the hypothesis; however, some nuances are worth attending as the effort dimension prevails over the talent dimension.
 In other words, the perception of meritocracy in terms of rewarded effort is more relevant in justifying access to education than the perception of talent.
 This could be explained as that talent could be more associated with luck in terms of random assignment, and in this terms would not be enough reason (as it is effort) to justify educational inequality.
-
-Our first hypothesis was supported, as we anticipated that the relationship between perceptions of meritocracy and the justification of educational inequality would tend to diminish over time.
-We based this on questions that related directly to experiences during the pandemic, such as receiving economic benefits during the lockdown period.
-Our findings revealed that those who received benefits justified educational inequality less than those who did not.
-Moreover, we observed that the connection between perceived meritocracy and the justification of educational inequality weakened for those who received benefits.
-This evidence partially supports our temporal change hypothesis ($H5$).
-In other words, the influence of meritocracy on justifying inequality appears to lessen, which aligns with our expectations given the social context in the post-COVID era.
+Furthermore, we observed that the connection between perceived meritocracy and the justification of educational inequality weakened for those who received state benefits, which aligns with our expectations given the social context in the post-COVID era.
 One possible interpretation is that, during times of increased risk, people may become more skeptical of narratives emphasizing individual effort over collective efforts.
@@
 
+Regarding the status position, our central hypothesis was that individuals in more advantaged situations and with greater resources would tend to defend their interests and justify greater education inequality.
+In this regard, our results show a positive relationship between income and the justification of inequality, which would align with our argument that rational interests of the better-off that would lead to a larger inequality justification.
+Besides status, we also considered inequality perception, as the literature on attitudes toward inequality has argued that a greater justification of inequality is associated with its perceived magnitude.
+Consistent with this hypothesis, our results showed that, throughout the years, a larger perceived economic inequality motivates a lower justification of inequality in education.
+
 Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in terms of outcomes as well as in areas such as access to social services in education as part of welfare policies.
-In this regard, the construct that we have captured with our indicator can be framed more in the latter, while the evaluation of justice involved in this process does not consider aspects such as equality of educational opportunities or observable outcomes such as the performance of individual or educational institutions.
-Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also as a function of structural aspects such as family of origin status or social capital.
-In this sense, the available measurements restricted us to the use of single indicators for effort and talent, and further research should consider improving the measurement quality in this regard.
+First, the operationalization of justification of educational inequality is rather limited.
+We counted only on a single item for this in this study, and further work is needed in order to assess the validity and reliability of measuring such construct.
+Second, as it has been established by recent research [@castillo_multidimensional_2023], the perception of meritocracy can be understood through individual attributions regarding effort and talent, but also taking into account the role of structural aspects such as family status and social capital.
+Due to data limitations such dimensions were not included here, therefore providing a partial perspective on the areas of perceived meritocracy.
 Finally, there are limitations involving the longitudinal dimension of the data used.
@@
 
-The capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
-In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level in future studies.
+Regarding future studies, the capabilities of the ELSOC longitudinal database are not limited to micro-level estimates.
+In this regard, thanks to the sampling strategy of the survey, it would be possible to make contextual estimates at the municipality level.
 For instance, statistical models could include contextual information from administrative data sources, allowing testing for hypotheses that include contextual socioeconomic inequality as well as its dynamics over time.

### 2025-01-19 04-analisis.qmd
update corr tab and render paper

@@
 
-@fig-correlation presents a correlation matrix of the main variables analyzed, using data from the first wave.
+@fig-correlation presents a correlation matrix of the main variables analyzed, using data from the last wave (2023).
 In this matrix, the correlations vary between low and moderate values.
 The correlations between justification of educational inequality with the two COVID variables are positive, but low.
-In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.34, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.29, p<.01; _r_=0.24, p<.01).
-Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.18, p<.01; _r_=-0.08, p<.01).
-Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.76, p<.01).
+In addition, justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.28, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.17, p<.01; _r_=0.17, p<.01).
+Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.19, p<.01; _r_=-0.14, p<.01).
+Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.79, p<.01).
 

### 2025-01-19 production/prod_analysis-fix-t5.Rmd
update corr tab and render paper

@@
 df_study1 %>%
-   filter(ola==1) %>%
+   filter(ola==7) %>%
   select(just_educ, # distributive justice
@@
 grid::grid.text("Source: own elaboration with data 
-from ELSOC 2016 (n = 1322)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 12, col = "black"))
+from ELSOC 2023 (n = 1322)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 12, col = "black"))
 dev.off()

### 2025-01-20 paper.qmd
tratando de generar tex

@@
 Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
-
-  \newline
-  \newline
-
+The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.\newline
   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19

### 2025-01-20 paper.tex
tratando de generar tex

@@
 opportunities. However, despite the widespread value of education,
-disparities in educational outcomes persist over time. To what extent
-are such educational disparities justified in society? And, what are the
-main factors driving this kind of justification? This research focuses
-on the changes in the justification of inequality in access to education
-in Chile and their links with the perception of meritocracy. The Chilean
-case offers an interesting context for this study given its high
-economic inequality, deep neoliberal policies, and commodification of
-social services such as education. The central argument of this article
-is that the justification of inequalities weakens during periods of
-vulnerability and crisis (such as the health and economic crisis
-resulting from the COVID-19 pandemic), which could be linked to a
-challenge of meritocratic ideals. For testing the research hypotheses we
-estimate a series of longitudinal multilevel models with data from a
-Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of
-meritocracy and justification of inequality, the analyses that involved
-changes over time showed an increase in the justification of inequality
-in education. The discussion of these results delves into the social
-consequences of justifying inequality in a sensitive area as education,
-as well as the persistence of meritocratic ideals despite challenging
-events.
-
-\newline
-\newline
-
+disparities in educational outcomes persist over time. This paper is
+guided by the following research questions: To what extent are such
+educational disparities justified in society? Is the justification of
+educational inequality affected in periods of personal and social
+vulnerability, as in the COVID pandemic? And, what are the main factors
+driving this kind of justification? The Chilean case offers an
+interesting context for this study given its high economic inequality,
+deep neoliberal policies, and commodification of social services such as
+education. This economic and cultural environment has promoted
+meritocratic ideals, whereby individual talent and effort are considered
+key to get ahead in life, disregarding opportunities linked to social
+origin. The central argument of this article is that the justification
+of inequalities weakens during periods of vulnerability and crisis (such
+as the health and economic crisis resulting from the COVID-19 pandemic).
+Furthermore, we argue that such changes could be linked to a challenge
+of meritocratic ideals. For testing the research hypotheses we estimate
+a series of longitudinal multilevel models with data from a Chilean
+longitudinal panel survey (2016 -- 2023, 6 waves, N = 2,927). Whereas we
+find support for the association between the perception of meritocracy
+and justification of inequality, the analyses that involved changes over
+time showed an increase in the justification of inequality in education.
+The discussion of these results delves into the social consequences of
+justifying inequality in a sensitive area as education, as well as the
+persistence of meritocratic ideals despite challenging events.\newline
 \textbf{Keywords}: meritocracy, social inequality, inequality
@@
 
-This document was last modified at 2024-12-15 23:10:25
-
-and it was last rendered at 2024-12-15 23:10:25
+This document was last modified at 2025-01-19 22:27:33
+
+and it was last rendered at 2025-01-19 22:27:33
 
@@
 social mobility and equality of opportunities, disparities in this
-domain are far from being eliminated. As the last PISA Report points
+domain are far from being eliminated. As the last PISA report points
 out, ``Socio-economically disadvantaged students are seven times more
@@
 society might play a role in the justification of economic inequalities
-in this particular domain.
-
-Meritocracy is defined as a distributive system where individual factors
-such as effort and talent are conceived as the main determinants of
-individual performance, usually sidelining the role of opportunities and
-social background. In recent years, there has been a growing interest in
-studying meritocratic beliefs as well as their variations according to
-individual and contextual variables (Mijs 2021; Mijs et al. 2022). This
-line of research has highlighted how meritocratic beliefs are associated
-with justifying unequal access to resources and rewards in society, by
-legitimizing social hierarchies in which those who already have
-resources and advantages are more likely to succeed (Sandel 2020;
-McNamee and Miller 2004; Hadjar 2008). Within this framework, this study
-focuses on the changes in the justification of inequalities in access to
-education in Chile, and to what extent this phenomenon is associated
-with the perception of meritocracy and its variations during the COVID
-pandemic.
+in this particular domain. We examine this association in the context of
+COVID 19 in Chile, a period characterized by several challenges
+regarding the political and economic system that could have affected the
+way in which the role of individual performance and social outcomes are
+deemed.
 
@@
 reforms and have diminished the role of private actors in the
-educational system.
+educational system (Donoso 2016; Reyes-Housholder and Roque 2019).
 
@@
 meritocratic beliefs and the legitimization of disparities, as well as
-the potential erosion of meritocratic principles amidst crises. Within
-this framework, we point out that periods of vulnerability (such as the
-health and economic crisis resulting from the COVID-19 pandemic) could
-weaken the justification of social inequalities, a change that would be
-at least in part linked to a challenge of the meritocratic ideals.
-Furthermore, we suggest that the association between meritocratic
-perception and inequality justification would be weaker among
-lower-status groups, as they are the most affected in critical times.
-
-\section{Inequality, beliefs and
-justification}\label{inequality-beliefs-and-justification}
+the potential erosion of meritocratic principles amidst crises. This
+line of research has highlighted how meritocratic beliefs are associated
+with justifying unequal access to resources and rewards in society, by
+legitimizing social hierarchies in which those who already have
+resources and advantages are more likely to succeed (Sandel 2020;
+McNamee and Miller 2004; Hadjar 2008). Within this framework, we point
+out that periods of vulnerability (such as the health and economic
+crisis resulting from the COVID-19 pandemic) could weaken the
+justification of social inequalities, a change that would be at least in
+part linked to a challenge of the meritocratic ideals.
+
+\section{The justification of educational
+inequalities}\label{the-justification-of-educational-inequalities}
 
@@
 Evans 2021; Mijs 2021). Both theoretical and empirical approaches have
-been able to establish a clear difference between inequality and
-justice, which opens the question of to what extent inequality can be
+been able to establish a clear distinction between inequality and
+justice, opening the question of to what extent inequality can be
 tolerated or even justified.
-
-The field of studies of subjective economic inequality has been
-extensively developed in the social sciences, particularly in the area
-of social justice research (Resh and Sabbagh 2016). From a sociological
-point of view, the works of Kluegel and Smith (1981) can be considered
-pioneers in framing a research area comprising perceptions, beliefs, and
-justifications in topics such as inequality of opportunities, poverty,
-and distributive justice. Theoretically, the authors understand the
-concept of ``beliefs'' in a ``broad sense to refer to the information
-(veridical or non-veridical) about a phenomenon that an individual uses
-as a basis both for inferring other information and for action.''
-(Kluegel and Smith 1981, 30). However, the literature has strained the
-concept of ``beliefs'' by more specific distinctions around the
-individual and structural dimensions of economic inequality. In this
-regard, Wegener (1987) has suggested that perceptions of social status
-and reward hierarchy are relative to the observer's status in such
-hierarchies (Wegener 1987, 2). This approach established that status
-positions influence how people perceive things \emph{are} and, in turn,
-these perceptions anchor how people evaluate this inequality as just or
-unjust (Jasso 1978). Additionally, Janmaat (2013) has suggested more
-detailed conceptual distinctions about how people perceive economic
-inequality \emph{is}, how they believe or prefer it \emph{should be},
-and the process of evaluating such differences as fair.
-
-Theoretically, Liebig and Sauer (2016) establish three fundamental
-distinctions in justice attitudes in the field of sociological justice
-research. First, \emph{order-related attitudes} refer to individual
-motives that explain preferences for resource allocation based on the
-principles of equality, need, or entitlement. Second, \emph{procedural
-justice attitudes} account for the degree of acceptance of unfavorable
-outcomes under conditions in which the procedure under which those
-decisions were made is considered fair (e.g., opportunities). Finally,
-they can be \emph{outcome-related attitudes} which refer primarily to
-the economic resources that a person currently obtains and the set of
-collective commitments associated with possessing those resources (e.g.,
-transfers or taxes). Within this framework, we conceptualize the
-justification of inequality in the first area of order-related
-attitudes, as it expresses preferences for resource allocation that can
-be associated with concepts that reflect different distributive
-principles (Kluegel, Mason, and Wegener 1995; Son Hing et al. 2011)
-
-An important finding in the area of order-related attitudes is that
-beliefs about inequality are stratified by social status (Janmaat 2013;
-Kluegel and Smith 1986). The evidence suggests that advantaged positions
-in terms of socioeconomic status, such as income, education, or
-occupational prestige, are associated with preferences for less
-egalitarian class imageries (Evans, Kelley, and Kolosi 1992; Evans and
-Kelley 2017) and greater justification of salary gaps (Hadler 2005;
-Kelley and Evans 2021). Nevertheless, and in contrast to this
-perspective, it has been argued that human beings are intrinsically
-motivated to justify the system they inhabit, even when the information
-available to them points in the opposite direction (Jost and Hunyady
-2003; van der Toorn and Jost 2014). Individuals would be generally
-disposed to justify the economic status quo regardless of self- and
-group economic interest because their support partially depends on their
-beliefs in system-justifying ideologies (Trump 2020). As a result, what
-has been suggested is that low-status might perceive the status quo as
-just as they are more strongly attached to dominant ideologies, while
-high-status individuals are more prone to justify the system to a
-greater extent because it favors their interests and dominant position
-(Jost 2020).
-
-Another factor that has been linked to inequality beliefs is the
-perception of economic inequality (Kelley and Evans 1993; Osberg and
-Smeeding 2006; Castillo, García-Castro, and Venegas 2022). Studies in
-this matter have argued that perceived economic inequality functions as
-an anchor for the justifications of inequality because if these
-inequalities are indeed seen as fair in their generative processes, then
-they will be considered as just (Wegener 1987, 1990). Thus, both the
-role of social status and economic inequality perceptions have found
-support in correlational (Castillo 2012; Schröder 2017), as well as in
-experimental studies that have manipulated inequality perception (Iturra
-et al. 2023; Trump 2018). In addition, normative or value-driven
-mechanisms can also explain to what extent people justify inequality.
-For instance, studies suggest that justice ideologies such as
-individualism or egalitarianism can mediate the influence of individual
-status on income inequality justification (Castillo 2011). Likewise,
-system justification ideologies can strengthen the influence of
-socioeconomic status on economic inequality preferences (García-Sánchez
-and De Carvalho 2022). Altogether, these studies have shown that not
-only socio-structural factors influence what people believe regarding
-the distribution of economic resources, but their normative stances and
-perceptions can also determine preferences for economic inequality in
-diverse domains.
-
-\section{The justification of educational
-inequality}\label{the-justification-of-educational-inequality}
 
@@
 
-The study of inequality justification in the sphere of education is a
-rather specific and unattended field. Most studies on educational
-inequality relate performance indicators (such as grades and
-standardized test results) to stratification-related variables, whereby
-few refer specifically to if and how individuals tolerate and even
-justify educational inequalities. This is probably because in principle
-it seems counter-intuitive to think that someone would be willing to
-justify differences in such a sensitive area. This paradox is phrased by
-Conforti (1992) as ``while equality is embraced as a value, inequality
-is accepted as a reality'' (p.230).
-
-In the area of empirical studies, the operationalization of educational
-inequality justification in terms of questionnaire items is diverse, and
-there is not a single instrument or scale dedicated specifically to this
-concept. Rather, some attitudinal surveys usually include
-education-related items that are posteriorly used as dependent or
-independent variables. Some examples are the International Social Survey
-Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that
-people with higher incomes can buy better education for their children
-than people with lower incomes?}), and the European Social Survey (ESS)
-2018 (\emph{Compared to other people in {[}country{]}, I have had a fair
-chance of achieving the level of education I was seeking}). Although
-both questions have been used to measure the justification of
-educational inequality, they refer to different aspects. The ESS can be
-classified as reflexive, as the respondent is involved as a referent for
-the question, whereas the ISSP question is non-reflexive. Furthermore,
-ISSP refers to educational access, while the ESS refers to educational
-outputs.
-
-Studies have addressed the justification of educational inequality from
-different conceptual perspectives and methodological strategies. Some
-perspectives emphasize the justification of differences between specific
-groups, finding for instance that Americans are more concerned
-about---and more supportive of proposals to close---wealth-based
-achievement gaps than Black-White or Hispanic-White gaps (Valant and
-Newark 2016). Other studies focus on the association between status
-differences and inequality justification in education, based on the
-rational premise that those who are worst-off would tend to show lower
-levels of justification (Stoilova and Ilieva-Trichkova 2023). However,
-most of the research in this area goes beyond status differences,
-focusing on how subjective factors such as perceptions and beliefs are
-related to different levels of justification.
-
-The perception of economic inequality is one of the factors associated
-with the justification of educational inequality. As pointed out by
-Janmaat (2013), justifications are usually confused under the labels of
-``beliefs'' and ``attitudes,'' and their distinction from perceptions is
-relevant as it helps state that justifying a certain level of inequality
-requires taking into account which level of (perceived) inequality is
-first and foremost being recognized (Son Hing et al. 2011; Castillo et
-al. 2023). For instance, a recent study by Day and Norton (2023) on the
-perception and justification of inequality in university endowments in
-the U.S. points out that individuals tend to underestimate the magnitude
-of existing inequality while simultaneously desiring greater equality.
-Furthermore, through an experiment, they demonstrate that information
-about inequality in university endowments (i.e., manipulating inequality
-perception) increases the perception of injustice in education. Results
-of this study are in line with the consistent evidence on biases related
-to the underestimation of economic inequality (Castillo, García-Castro,
-and Venegas 2022; Gimpelson and Treisman 2018)
-
-Besides focusing on the relationship between perception and
-justification of educational inequality, another group of studies has
-addressed how the justification of inequality is associated with the
-support of different distributive principles. In an experiment by
+In the area of survey studies, the operationalization of educational
+inequality justification in terms of questionnaire items has been
+diverse, and there is not a single instrument or scale dedicated
+specifically to this concept. Rather, some attitudinal surveys usually
+include education-related items that are posteriorly used as dependent
+or independent variables. Some examples are the International Social
+Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong --
+that people with higher incomes can buy better education for their
+children than people with lower incomes?}), and the European Social
+Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I
+have had a fair chance of achieving the level of education I was
+seeking}). Although both questions have been used to measure the
+justification of educational inequality, they refer to different
+aspects: the ESS can be classified as reflexive, as the respondent is
+involved as a referent for the question, whereas the ISSP question is
+non-reflexive. Furthermore, ISSP refers to educational access, while the
+ESS refers to educational outputs. Although such items covering
+inequality justification are present in emblematic comparative studies,
+still there are few studies focusing specifically in the justification
+of educational inequality.
+
+\emph{The COVID pandemic and public concerns about inequality}
+
+The time frame for this study covers between 2016 and 2023, by using
+longitudinal panel survey data provided by the ELSOC database (Estudio
+Longitudinal Social de Chile). Although this period comprises the COVID
+pandemic, the changes in the variables analyzed are not possible to be
+attributed only to this event in terms of a health crisis, as several
+economic, social and cultural processes overlapped worldwide. However,
+and as several studies have shown, this period appeared to trigger
+attitudinal changes that are relevant to explore, both in themselves and
+as antecedent to future studies. In this sense, in this paper we argue
+that the increase of risk perceptions in critical situations (such as
+the pandemic) impacts public concerns about equality and redistribution
+(Breznau 2021).
+
+In Chile, the COVID-19 pandemic posed a significant threat to
+economically vulnerable groups and those in precarious labor market
+positions (Ministerio de Desarrollo Social y Familia 2023; Instituto
+Nacional de Estadísticas 2022). The poverty rate, which was 11.2\% in
+2015, decreased to 8.5\% by 2017, with a 10/10 ratio of 31.6. However,
+during the pandemic in 2020, poverty rose to 10.7\%, the 10/10 ratio
+spiked to 301.7, and unemployment reached a decade-high of 10.5\%. By
+2022, poverty declined to 6.5\%, the 10/10 ratio improved to 61.2, and
+unemployment decreased to 7.9\%, with a slight rise to 8.7\% in 2023.
+
+Chile implemented sanitary controls early in the pandemic, initiating
+lockdowns on March 18, 2020, and lifting all restrictions by March 5,
+2023. In this context, Chile stands out as one of the strictest
+countries globally and within South America in terms of pandemic-related
+policies. According to the Stringency Index (Hale et al. 2021)---ranging
+from 0 to 100 (100 being the strictest) and encompassing measures like
+school closures, workplace shutdowns, and travel bans---the global
+median was 42.54, while Chile scored 58.1. This placed Chile 36th among
+184 countries and 4th in South America, following Peru (60.1), Suriname
+(60.5), and Venezuela (80.5). Besides, an important cycle of protests
+known as the ``social outburst'' took place between October 2019 and
+March 2020 (Somma et al. 2021). As a consequence there were two
+constitutional processes happening in this very same period after the
+massive social upheavals. However, the protest drastically decreased as
+a result of the lockdown (Joignant et al. 2020).
+
+In terms of public views on economic inequality, it has been documented
+that critical views toward inequality increased in 2019 and deepened
+under the pandemic (COES 2023), which we argue can be in part explained
+by the increased risk exposure and vulnerability in the context of the
+COVID-19 pandemic. Although the pandemic might have triggered such
+concerns mostly regarding health, we believe that other policy areas,
+such as education, could have been affected in this period.
+Consistently, our first hypothesis is:
+
+\(H_1\): The justification of educational inequalities decreases after
+the pandemic
+
+Besides proposing a general decrease over time in the justification of
+inequalities, a second hypothesis refers to specific aspects of the
+pandemic. As some political measures from the state included several
+social benefits (as delivering food boxes, free vaccination, direct
+transfers, among others), we argue that being recipient of such benefits
+could affect attitudes towards the redistributive role of the state as
+equalizer opportunities to face risks. As a consequence, preferences for
+educational inequalities would be reduced:
+
+\(H_2\): Recipient of state benefits show lower justification of
+educational inequalities
+
+In the same line, we expect that those who give priority to take care of
+the health of the population instead of the country economy would
+exhibit larger preferences for educational inequalities:
+
+\(H_3\): Those prioritizing health instead of the economic situation
+show higher justification of educational inequalities
+
+\emph{Meritocracy and educational inequalities}
+
+Whether inequalities in education are considered just or not lead us to
+take into account the concepts of merit and meritocracy. Meritocracy, a
+concept coined by the sociologist Michael Young in his satirical work
+``The Rise of the Meritocracy'' (1962), has a rich history and
+multifaceted meaning. Young's term was initially used to criticize a
+hypothetical future society where social and economic positions were
+entirely determined by individuals' merits, talents, and efforts,
+ostensibly eliminating the influence of inherited privilege. However,
+over time, the term has evolved to describe a social system where
+advancement and success are perceived as directly linked to one's
+abilities and achievements, sidelining inequalities of opportunities
+(Bell 2020). This would contribute to the justification of inequality in
+the educational domain (Sandel 2020), as individual ability is often
+considered inherited, and educational attainment is influenced by family
+background (Goldin 2000).
+
+Although there are few studies relating meritocratic beliefs and
+inequality justification, from a broader perspective it is possible to
+find research which incorporates the support of different distributive
+principles with regard to education. For instance, in an experiment by
 Igliozzi, Granot, and Ottati (2024), they establish how the support for
@@
 find that a larger system justification (Jost and Hunyady 2003) is
-related to higher support for the principles of equity and equality
-(over need) in distributing educational outcomes, where equity was
-measured in terms of access to education related to performance
-(`Students who place within the highest 50\% of testing scores from the
-previous year can attend the magnet school'). In the same line, Lee and
-Stacey (2023) show that performance assessments (in what they call
-``neoliberal orientations'') predicted people's fair perceptions of
-educational inequality.
-
-\emph{Justificacion, inequality and meritocracy in the educational
-domain}
-
-The relationship between performance and outputs brings us to the
-concepts of merit and meritocracy, key in educational justice research.
-The concept of meritocracy, coined by the sociologist Michael Young in
-his satirical work ``The Rise of the Meritocracy'' (1962), has a rich
-history and multifaceted meaning. Young's term was initially used to
-critique a hypothetical future society where social and economic
-positions were entirely determined by individuals' merits, talents, and
-efforts, ostensibly eliminating the influence of inherited privilege.
-However, over time, the term has evolved to describe a social system
-where advancement and success are perceived as directly linked to one's
-abilities and achievements, sidelining inequalities of opportunities
-(Bell 2020). This would contribute to the justification of inequality in
-the educational domain (Sandel 2020), as individual ability is often
-inherited, and educational attainment is influenced by family background
-(Goldin 2000).
-
-Empirical research examining the relationship between meritocratic
-perceptions and the justification of educational inequality provides
-valuable insights into the dynamics shaping individuals' attitudes and
-beliefs. Some studies have explored the extent to which people believe
-that educational systems are meritocratic and how these perceptions
-influence their acceptance or rejection of educational disparities, as
-well as status differences (McCoy and Major 2007). This belief is
-particularly influential in educational institutions, where it not only
-encourages academic success but also legitimizes social and income
-inequality (Batruch et al. 2023). Most of the evidence so far suggests
-that those who believe in a direct link between effort, talent, and
-educational success are more inclined to rationalize existing
-disparities in educational outcomes. However, empirical findings also
-reveal nuances in this relationship. For instance, studies show that
-when individuals perceive structural barriers, such as economic
-disparities or discrimination, as hindering the meritocratic ideal, they
-may be less likely to justify educational inequalities (Day and Norton
-2023). This suggests that while meritocratic beliefs play a significant
-role, individuals' awareness of broader societal issues can mediate
-their willingness to rationalize educational disparities.
-
-\emph{This study}
-
-The present research uses longitudinal survey data from 2016 to 2022 to
-asses changes in inequality justification. As there are measures before
-and after the COVID pandemic, firstly we explore to what extent it is
-possible to identify variations in inequality justification in this time
-frame. Although we are aware that it is not possible to make a causal
-claim given that this period is characterized by many other
-circumstances besides COVID, we argue that the increase of risk
-perceptions in critical situations (such as the pandemic) impacts public
-concerns about equality and redistribution (Breznau 2021). Although the
-pandemic might have triggered such concerns mostly regarding health, we
-believe that other policy areas, such as education, could have been
-impacted in this period. Consistently, our first hypothesis is:
-
-\(H_1\): The justification of educational inequalities decreases after
-the pandemic
-
-Our second argument is based on the evidence that relates rational
-interests with justifications and preferences (Son Hing et al. 2019),
-expecting that:
-
-\(H_2\): Those with lower status characteristics show lower
-justification of educational inequalities
-
-Thirdly, we address inequality perception. As referred to above, the
-justification of inequality requires, first and foremost, that such
-inequality is recognized. As a larger inequality in this domain
-threatens a basic equality ideal, the corresponding hypothesis in this
-case is:
-
-\(H_3\): A larger inequality perception is negatively associated with
-the justification of educational inequalities.
-
-Our fourth hypothesis refers to the central concept of meritocracy. As
-meritocracy justifies inequalities based on performance, we expect that:
+linked to higher support for the principles of equity and equality (over
+need) in distributing educational outcomes, where equity was measured in
+terms of access to education related to performance (`Students who place
+within the highest 50\% of testing scores from the previous year can
+attend the magnet school'). In the same line, Lee and Stacey (2023) show
+that performance assessments (in what they call ``neoliberal
+orientations'') predicted perceived fairness of educational inequality.
+
+More specific empirical research examining the relationship between
+meritocratic perceptions and the justification of educational inequality
+provides valuable insights into the dynamics shaping individuals'
+attitudes and beliefs. Some studies have explored the extent to which
+people believe that educational systems are meritocratic and how these
+perceptions influence their acceptance or rejection of educational
+disparities, as well as status differences (McNamee and Miller 2004).
+This belief is particularly influential in educational institutions,
+where it not only encourages academic success but also legitimizes
+social and income inequality (Batruch et al. 2023). Most of the evidence
+so far suggests that those who believe in a direct link between effort,
+talent, and educational success are more inclined to rationalize
+existing disparities in educational outcomes. However, empirical
+findings also reveal nuances in this relationship. For instance, studies
+show that when individuals perceive structural barriers, such as
+economic disparities or discrimination, as hindering the meritocratic
+ideal, they may be less likely to justify educational inequalities (Day
+and Norton 2023). This suggests that while meritocratic beliefs play a
+significant role, individuals' awareness of broader societal issues can
+mediate their willingness to rationalize educational disparities.
+
+Empirical cross-sectional evidence from Chile reveals two significant
+trends: low public support for the perception of meritocracy and limited
+justification for unequal access to social services such as education
+and health. For instance, Castillo, Madero-Cabib, and Salamovich (2013)
+analyzed data from the International Social Survey Programme (ISSP) in
+1999 and 2009, highlighting consistently low support for inequality
+based on resource allocation or individual merit. Similarly, Castillo,
+Miranda, and Carrasco (2012) demonstrated widespread skepticism among
+Chileans regarding the meritocratic system, with many disagreeing with
+the notion that rewards are fairly distributed based on talent and
+effort. In line with this evidence, we propose that:
 
@@
 
-Finally, and in line with hypotheses 1 and 4, we explore interactions
-between meritocracy and inequality justification over time. As COVID-19
-pandemic has affected the structural opportunities to attend schools,
+Combining the arguments regarding the effect of the pandemic and its
+relationship with meritocracy, we explore interactions between
+meritocracy and inequality justification over time. As COVID-19 pandemic
+has affected the structural opportunities to attend schools,
 universities, or workplaces, the ideal of meritocracy might have been
-challenged. As a consequence, the justification of inequality could show
-a decrease over time:
+challenged in the time frame analyzed. As a consequence, the
+justification of inequality could show a decrease over time:
 
@@
 
+Along with meritocratic perceptions, in this study we incorporate
+inequality perception. As pointed out by Janmaat (2013), justifications
+are usually confused under the labels of ``beliefs'' and ``attitudes,''
+and their distinction from perceptions is relevant as it helps state
+that justifying a certain level of inequality requires taking into
+account which level of (perceived) inequality is first and foremost
+being recognized (Son Hing et al. 2011; Castillo et al. 2023). For
+instance, a recent study by Day and Norton (2023) on the perception and
+justification of inequality in university endowments in the U.S. points
+out that individuals tend to underestimate the magnitude of existing
+inequality while simultaneously desiring greater equality. Furthermore,
+through an experiment, they demonstrate that information about
+inequality in university endowments (i.e., manipulating inequality
+perception) increases the perception of injustice in education. Results
+of this study are in line with the consistent evidence on biases related
+to the underestimation of economic inequality (Castillo, García-Castro,
+and Venegas 2022; Gimpelson and Treisman 2018). As a larger inequality
+in this domain threatens a basic equality ideal, the corresponding
+hypothesis in this case is:
+
+\(H_6\): A larger inequality perception is negatively associated with
+the justification of educational inequalities.
+
+Next, we take into account the evidence that relates rational interests
+to justifications and preferences (Son Hing et al. 2019). Research about
+the justification of educational inequality usually focuses on
+differences between status groups, based on the rational premise that
+those who are worst-off would tend to show lower levels of justification
+(Busemeyer and Iversen 2020). For instance, Valant and Newark (2016)
+find that Americans are more concerned about --- and more supportive of
+proposals to close --- wealth-based achievement gaps than Black-White or
+Hispanic-White gaps. In line with this evidence, we expect that:
+
+\(H_7\): Those with lower status characteristics show lower
+justification of educational inequalities
+
 \section{Data, variables \& methods}\label{data-variables-methods}
@@
 
-The main data source is the Chilean Longitudinal Social Survey
-2016--2022. The ELSOC has been designed as a yearly panel study to
+The main data source is the Chilean Longitudinal Social Survey ELSOC
+(2016--2023). ELSOC has been designed as a yearly panel study to
 evaluate how individuals think, feel, and behave regarding a set of
@@
 (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller
-cities. Consequently, the ELSOC comprises a total of 2,927 participants
-aged between 18 and 75 years in wave 1. In addition, the sample
-represents 77\% of the total national population and 93\% of the urban
-areas (ELSOC 2022).
+cities. The sample covers participants aged between 18 and 75 years (in
+wave 1), representing 77\% of the total national population and 93\% of
+the urban areas (ELSOC 2022).
 
@@
 2020, when it was suspended due to the pandemic. The waves 2016, 2017,
-2018, 2019 and 2022 were administered using computer-assisted personal
-interviewing (CAPI). However, a reduced version was conducted using
-computer-assisted telephone interviews (CATI) in 2021. In addition, wave
-3 included a refreshment sample to counter survey attrition. The same
-sampling strategy as that of wave 1 was implemented for selecting new
-cases. As a result, the total sample of wave 3 included 3,748 cases, of
-which 2,229 were part of the original sample and 1,519 were from the
+2018, 2019, 2022, and 2023 were administered using computer-assisted
+personal interviewing (CAPI). However, a reduced version was conducted
+using computer-assisted telephone interviews (CATI) in 2021. In
+addition, wave 3 included a refreshment sample to counter survey
+attrition. As a result, the total sample of wave 3 included 3,748 cases,
+of which 2,229 were part of the original sample and 1,519 were from the
 refreshment sample. The data from the refreshment sample is not included
-in the analytical sample because we wanted to analyze longer response
+in the analytical sample for this study as we analyze longer response
 trends. Regarding the original sample, the response rate was 62.4\% in
 wave 1, achieving N = 2,927 participants. In broader terms, the
-accumulated attrition between wave 1 and wave 6 is 41\%, achieving a
-final sample of N = 1728. The analysis used longitudinal weights to
+accumulated attrition between wave 1 and wave 7 is 40\%, achieving a
+final sample of N = 1,741. The analysis used longitudinal weights to
 avoid bias due to the sampling design, allowing us to control for biases
@@
 personas con ingresos más bajos'' in Spanish). Here, the respondents
-declared their preferences on a Likert scale from ``strongly disagree''
-(1) to ``strongly agree'' (5). The main independent variables refer to
-meritocratic perception, which is operationalized by two items, one
-related to effort (``In Chile, people are rewarded for their efforts''),
-and the other related to talent (``In Chile, people are rewarded for
-their intelligence and skills''). In addition, we included an indicator
-for economic inequality perception (``In Chile, the income differences
-are too large''). Figure~\ref{fig-frecuencies} shows the average
-distribution throughout the years for these variables. We observe that
-there is a large group (88.2\%) of respondents who disagree or strongly
-disagree that people with higher incomes can access better education.
-Likewise, a large group of respondents (91.7\%) agree or strongly agree
-that income differences in Chile are too large. On the other hand, there
-is an important level of disagreement (disagree + strongly disagree)
-about the perception of meritocracy, this is, whether people are
-rewarded for their talent (47.6\%) and for their efforts (57\%).
+declared their preferences on a Likert scale from (1) ``strongly
+disagree'' to (5) ``strongly agree'' (5). The main independent variables
+refer to meritocratic perception, which is operationalized by two items,
+one related to effort (``In Chile, people are rewarded for their
+efforts''), and the other related to talent (``In Chile, people are
+rewarded for their intelligence and skills''). In addition, we included
+an indicator for economic inequality perception (``In Chile, the income
+differences are too large''). Figure~\ref{fig-frecuencies} shows the
+average distribution throughout the years for these variables. We
+observe that there is a large group (88.2\%) of respondents who disagree
+or strongly disagree that people with higher incomes can access better
+education. Likewise, a large group of respondents (91.7\%) agree or
+strongly agree that income differences in Chile are too large. On the
+other hand, there is an important level of disagreement (disagree +
+strongly disagree) about the perception of meritocracy, this is, whether
+people are rewarded for their talent (47.6\%) and for their efforts
+(57\%).
 
@@
 
-For testing the hypothesis regarding status we included educational
-level and household income quintiles, as well as subjective social
-status. Finally, political identification on a left-right scale, age,
-and gender were included as control variables.
-Table~\ref{tbl-descriptives} shows the whole set of independent
-variables, the corresponding items, response categories, and
-frequencies.
+A third group of variables is included in the analysis to assess the
+effect of elements related to the pandemic context. These variables were
+only included in the 2021 wave, so each individual's responses were
+replicated for the remaining years. The first variable addresses the
+belief regarding whether, in the context of the pandemic, it is more
+important to give priority to economic activity than to the health of
+the population, measured on a Likert scale from (1) ``strongly
+disagree'' to (2) ``strongly agree''. The second variable is a dummy
+(No/Yes) that captures whether individuals received state benefits
+during the pandemic (e.g., Food Box, Emergency Family Income,
+Contribution for the Middle Class, or others).
+
+For testing the hypothesis regarding status we considered educational
+level and household income. Also, we included subjective social status
+as it has been argued that perceived social status is a relevant
+predictor of attitudes toward economic inequality (Schneider and
+Castillo 2015; Castillo et al. 2019). In addition, such subjective
+measures can complement objective measures in predicting life chances
+(Oesch and Vigna 2023), which are arguably connected to how individuals
+experience economic inequality. In this regard, it has been argued that
+how people form their views on their standing in society results from
+experiences with direct socioeconomic circumstances and social
+comparison processes with reference groups (Condon and Wichowsky 2020).
+Finally, political identification on a left-right scale, age groups, and
+gender were included as control variables. Table~\ref{tbl-descriptives}
+shows the whole set of independent variables, the corresponding items,
+response categories, and frequencies.
 
@@
 
+Beginning with a descriptive analysis of changes in the justification of
+educational inequality between 2016 and 2023, Figure~\ref{fig-alluvial}
+illustrates yearly frequencies and transitions between years. Each year
+represents stacked porcentual frequencies, whereas the flows in between
+reflect the within-subject change of opinions from one year to the next
+(Rosvall and Bergstrom 2010). For instance, of the 32.9\% who strongly
+disagreed with inequality justification in 2016, about half of them kept
+responding the same in 2017, whereas the other half shifted their
+opinion to other response categories. In general, the large majority -
+between 80 and 90\% - disagrees with the justification of educational
+inequality throughout the years. Despite this overall tendency, we also
+observe that the disagreement with inequality justification (disagree +
+strongly disagree) tends to go down in the last waves. This change is
+mostly a result of the increase in the ``agree'' category, which doubles
+compared to previous years (from 5.4\% in 2016 to 11\% in 2023).
+
 \begin{figure}[H]
@@
 \caption{\label{fig-alluvial}Change in the justification of educational
-inequality over time (2016-2022)}
-
-\end{figure}%
-
-Figure~\ref{fig-alluvial} illustrates yearly frequencies in the
-justification of educational inequality between 2016 and 2022. Each year
-represents stacked percentual frequencies, and the flows in between
-reflect the within-subject change of opinions from one year to the next,
-as we are using longitudinal panel data (Rosvall and Bergstrom 2010).
-For instance, of the 32.2\% who strongly disagreed with inequality
-justification in 2016, about half of them kept responding the same in
-2017, whereas the other half shifted their opinion to other response
-categories. In general, the large majority - between 80 and 90\% -
-disagrees with the justification of educational inequality throughout
-the years. Despite this overall tendency, we also observe that the
-disagreement with inequality justification (disagree + strongly
-disagree) tends to go down in the last wave. This change is mostly a
-result of the radical decrease in the lowest response category (strongly
-disagree), which diminishes to half when compared with previous years.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/years_plot.png}
-
-}
-
-\caption{\label{fig-means}Change in the mean of justification of
-educational inequality, perception of inequality, and perception of
-meritocracy over the years}
+inequality over time (2016-2023)}
 
@@
 average is consistently represented by inequality perception.
-Interestingly, in the last wave of the study (2022) the justification of
-inequality increases whereas the perception of inequality decreases. As
-the merit variables are concerned, they show a very similar pattern in
-terms of averages and changes over the years, being the perception of
-meritocracy related to effort always lower than the one associated with
-talent.
+Interestingly, in the last waves of the study (2022-2023) the
+justification of inequality increases whereas the perception of
+inequality decreases. As the merit variables are concerned, they show a
+very similar pattern in terms of averages and changes over the years,
+being the perception of meritocracy related to effort always lower than
+the one associated with talent.
 
@@
 
+\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/years_plot.png}
+
+}
+
+\caption{\label{fig-means}Change in the mean of justification of
+educational inequality, perception of inequality, and perception of
+meritocracy over the years}
+
+\end{figure}%
+
+Figure~\ref{fig-correlation} presents a correlation matrix of the main
+variables analyzed, using data from the last wave (2023). In this
+matrix, the correlations vary between low and moderate values. The
+correlations between justification of educational inequality with the
+two COVID variables are positive, but low. In addition, justification of
+education inequality depicts a moderate and negative association with
+the perception of inequality (\emph{r}=-0.28, p\textless.01) and a
+moderate and positive association with both meritocracy perception
+variables (\emph{r}=0.17, p\textless.01; \emph{r}=0.17, p\textless.01).
+Regarding the perception of inequality, it presents a moderate and
+negative association with both meritocracy perception variables
+(\emph{r}=-0.19, p\textless.01; \emph{r}=-0.14, p\textless.01). Finally,
+the two meritocracy variables present a high and positive association
+with each other (\emph{r}=0.79, p\textless.01).
+
+\begin{figure}[H]
+
+\centering{
+
 \includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/corr.png}
@@
 
-\caption{\label{fig-correlation}Correlation matrix of justification of
-educational inequality, perception of inequality, and perception of
-meritocracy for all the years analyzed}
+\caption{\label{fig-correlation}Correlation matrix of the main variables
+for the first wave (2016)}
 
@@
 
-Figure~\ref{fig-correlation} presents a correlation matrix of the main
-variables analyzed, using data from all survey waves. In this matrix,
-the correlations vary between low and moderate values. Justification of
-economic inequality depicts a moderate and negative association with the
-perception of inequality (\emph{r}=-0.29, p\textless.01) and a moderate
-and positive association with both meritocracy perception variables
-(\emph{r}=0.18, p\textless.01; \emph{r}=0.15, p\textless.01). Regarding
-the perception of inequality, it presents a moderate and negative
-association with both meritocracy perception variables (\emph{r}=-0.15,
-p\textless.01; \emph{r}=-0.10, p\textless.01). Finally, the two
-meritocracy variables present a high and positive association with each
-other (\emph{r}=0.74, p\textless.01).
-
-\subsection{Multivariate analysis}\label{multivariate-analysis}
+\emph{Multilevel models}
 
@@
 Taking 2016 as a reference point, we can observe a staggered decrease in
-2017 (\(\beta\)=-0.15, p\textless.001), 2018 (\(\beta\)=-0.06,
-p\textless.05), and 2019 (\(\beta\)=-0.07, p\textless.01). Nevertheless,
-in the last wave of 2022 there is a radical increase in level of
-justification of economic inequality (\(\beta\)=0.21, p\textless.001),
+2017 (\(\beta\)=-0.167, p\textless.001), 2018 (\(\beta\)=-0.059,
+p\textless.05), and 2019 (\(\beta\)=-0.03, p\textgreater.05).
+Nevertheless, in the last waves of 2022 and 2023 there is a clear
+increase in level of justification of educational inequality
+(\(\beta\)=0.204, p\textless.001 and \(\beta\)=0.233, p\textless.001),
 suggesting a non-linear change in this variable. Attempting to model
@@
 the survey wave depicts a negative association, expressing an average
-decrease in inequality justification over time, but on the other hand,
-the quadratic wave term is positive, indicating the reversion of this
-path in the last measurement point.
-
-Model 3 adds the sociodemographic variables, where having a higher
-income and being older (65 or older compared to 18-29) have a
-significant and positive influence on the justification of educational
-inequality. On the contrary, the gender variable has a negative effect,
-expressing that women on average justify less inequality in education
-when compared to males. Educational level and subjective social status
-are not shown in this table for the sake of space (as they do not
-exhibit significant effects), but the complete estimation of the models
-is presented in the appendix. Model 3 also includes the control
-variables, where identifying oneself with the center, right-wing, or
-when reporting no political identification, has a significant and
-positive association on the justification of educational inequality -
-when compared to being left-wing oriented. Inequality perception is
-added in Model 4, showing a negative association with the justification
-of educational inequality as hypothesized, remaining stable when
-controlling for the rest of the variables.
+decrease in inequality justification over time. On the other hand, the
+quadratic wave term is positive, indicating the reversion of this path
+in the last measurement points.
+
+Models 3 and 4 adds the two variables about the pandemic context, where
+we can see that those who received state benefits during this period
+justify less inequality in education compared to those who did not
+receive them. Although the effect size is small, it remains significant
+when controlling for the rest of the variables in model 8
+(\(\beta\)=-0.093, p\textless.01). On the other hand, those who believe
+that in this context the economy is more important than the health of
+the population justify more inequality in education(\(\beta\)=0.038,
+p\textless.05, Model 8).
 
@@
 perception that talent is rewarded has a positive influence on the
-justification of educational inequality in Model 5 (\(\beta\)=0.04,
-p\textless.01). However, when controlling for the perception that effort
-is rewarded, this effect is no longer significant. In this sense, Model
-6 shows that the perception that effort is rewarded in society is not
-only positively associated with the justification of educational
-inequality (\(\beta\)=0.05, p\textless.01), but it has a larger weight
-than the perception of talent.
+justification of educational inequality in Model 5 (\(\beta\)=0.055,
+p\textless.001). However, when controlling for the perception that
+effort is rewarded, this effect is no longer significant. In this sense,
+Model 6 shows that the perception that effort is rewarded in society is
+not only positively associated with the justification of educational
+inequality (\(\beta\)=0.083, p\textless.001), but it has a larger weight
+than the perception of talent. Regarding inequality perception (added in
+Model 7), it depicts a negative association with the justification of
+educational inequality, remaining stable when controlling for the rest
+of the variables.
+
+Model 8 adds the socioeconomic variables to test hypothesis 7. Although
+educational level and household per capita income are not significant in
+this model, controlling for all variables in the next models, household
+per capita income depicts a positive effect (\(\beta\)=0.001,
+p\textless.05). The inclusion of other controls as being right-wing
+(compared to being left-wing) shows a positive association with
+justification of educational inequality (\(\beta\)=0.27,
+p\textless.001), whereas being a women (compared to men) had a negative
+effect (\(\beta\)=-0.08, p\textless.001). Subjective social status and
+age had no significant effect when controlling for all the variables
+analyzed. The full model specification can be seen in the appendix (see
+Table~\ref{tbl-multilevel-full}).
 
@@
 
-Statistical models
-
 ~
@@
 
+Model 7
+
+Model 8
+
 Intercept
 
-1.84***
-
-1.87***
-
-1.80***
-
-2.39***
-
-2.22***
-
-2.13***
+1.847***
+
+1.853***
+
+1.920***
+
+1.776***
+
+1.641***
+
+1.563***
+
+2.174***
+
+2.183***
+
+~
+
+(0.026)
+
+(0.038)
+
+(0.048)
+
+(0.064)
+
+(0.068)
+
+(0.069)
+
+(0.086)
+
+(0.120)
+
+Wave (Ref.= Wave 2016)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2017
+
+-0.167***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2018
+
+-0.059*
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2019
+
+-0.030
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2022
+
+0.204***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.030)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2023
+
+0.233***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Wave
+
+~
+
+-0.073***
+
+-0.073***
+
+-0.073***
+
+-0.076***
+
+-0.076***
+
+-0.052**
+
+-0.052**
+
+~
+
+~
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+Wave\^{}2
+
+~
+
+0.016***
+
+0.016***
+
+0.016***
+
+0.016***
+
+0.016***
+
+0.013***
+
+0.013***
+
+~
+
+~
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+Covid: receive state benefits
+
+~
+
+~
+
+-0.095*
+
+-0.097*
+
+-0.100*
+
+-0.102*
+
+-0.102**
+
+-0.093*
+
+~
+
+~
+
+~
+
+(0.041)
+
+(0.041)
+
+(0.040)
+
+(0.040)
+
+(0.039)
+
+(0.041)
+
+Covid: economy more important than health
+
+~
+
+~
+
+~
+
+0.065***
+
+0.060**
+
+0.055**
+
+0.049**
+
+0.038*
+
+~
+
+~
+
+~
+
+~
+
+(0.019)
+
+(0.019)
+
+(0.019)
+
+(0.019)
+
+(0.019)
+
+Merit: Talent
+
+~
+
+~
+
+~
+
+~
+
+0.055***
+
+-0.002
+
+-0.003
+
+-0.005
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.010)
+
+(0.012)
+
+(0.012)
+
+(0.012)
+
+Merit: Effort
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.096***
+
+0.084***
+
+0.083***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.013)
+
+(0.013)
+
+(0.013)
+
+Inequality perception
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+-0.135***
+
+-0.134***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.012)
+
+(0.012)
+
+Universitary (Ref.= Less than universitary)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+-0.048
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.046)
+
+Household income
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.001*
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.000)
+
+Controls
+
+No
+
+No
+
+No
+
+No
+
+No
+
+No
+
+No
+
+Yes
+
+BIC
+
+29667.072
+
+29670.297
+
+29678.484
+
+29682.436
+
+29668.980
+
+29626.960
+
+29510.881
+
+29626.312
+
+Num. obs.
+
+7576
+
+7576
+
+7576
+
+7576
+
+7576
+
+7576
+
+7576
+
+7576
+
+Num. groups: Individuals
+
+1322
+
+1322
+
+1322
+
+1322
+
+1322
+
+1322
+
+1322
+
+1322
+
+Var: Individuals (Intercept)
+
+0.185
+
+0.184
+
+0.183
+
+0.179
+
+0.174
+
+0.170
+
+0.164
+
+0.159
+
+Var: Residual
+
+0.497
+
+0.500
+
+0.500
+
+0.500
+
+0.499
+
+0.496
+
+0.488
+
+0.488
+
+*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
+Note: Model 8 is controlled by age, gender, political position, and
+subjective social status.
+
+}
+
+\end{table}%
+
+In this last part of the analysis we test hypotheses about the effect of
+COVID-19 on changes in the relationship between meritocracy and
+justification of educational inequality. In hypothesis 5, we proposed
+that the association between the perception of meritocracy and
+justification of inequality is mitigated in times of crisis, as
+meritocratic ideals might have been weakened due to critical situations
+associated with the COVID-19 pandemic. We test this hypothesis through
+interaction effects, shown in Table~\ref{tbl-interact}. The first model
+is shown as the baseline model. It is the same as model 8 in
+Table~\ref{tbl-multilevel}, but only shows the variables involved in the
+interaction for the sake of space (all other variables are included).
+Model 9 adds the interaction between the perception of talent-based
+meritocracy and whether state benefits were received during the
+pandemic, whereas Model 10 considers the perception of effort-related
+meritocracy for the interaction. Both variables show a negative
+significant interaction with receiving state benefits
+(\(\beta_{talent}\)=-0.07, p\textless.01; \(\beta_{effort}\)=-0.05,
+p\textless.05), meaning that for those who received state benefits the
+association between the the justification of inequality and the
+perception of effort-based meritocracy lower by 0.05 points, and by 0.07
+points regarding the perception of talent-based meritocracy. This last
+interaction is depicted in Figure~\ref{fig-interact-talent}, showing
+that the positive association between the perception of talent-based
+meritocracy and the justification of educational inequality is smaller
+for those who received state benefits during the pandemic.
+
+\begin{table}
+
+\caption{\label{tbl-interact}Interaction effects for the justification
+of economic inequality}
+
+\centering{
+
+~
+
+Model 8
+
+Model 9
+
+Model 10
+
+Intercept
+
+2.18***
+
+2.03***
+
+2.09***
+
+~
+
+(0.12)
+
+(0.13)
+
+(0.13)
+
+Wave
+
+-0.05**
+
+-0.05*
+
+-0.05**
 
@@
 
-(0.03)
-
-(0.09)
-
-(0.10)
-
-(0.10)
-
-(0.10)
-
-Wave (Ref.= Wave 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.14***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.04
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.04
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.23***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.28***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.08***
-
--0.08***
-
--0.05**
-
--0.06**
-
--0.06**
-
-~
-
-~
-
 (0.02)
@@
 
+Wave\^{}2
+
+0.01***
+
+0.01***
+
+0.01***
+
+~
+
+(0.00)
+
+(0.00)
+
+(0.00)
+
+Covid: receive state benefits
+
+-0.09*
+
+0.11
+
+0.03
+
+~
+
+(0.04)
+
+(0.07)
+
+(0.07)
+
+Merit: Talent
+
+-0.00
+
+0.05*
+
+-0.01
+
+~
+
+(0.01)
+
 (0.02)
 
+(0.01)
+
+Merit: Effort
+
+0.08***
+
+0.08***
+
+0.12***
+
+~
+
+(0.01)
+
+(0.01)
+
 (0.02)
 
+Receive state benefits * Merit: Talent
+
+~
+
+-0.07**
+
+~
+
+~
+
+~
+
 (0.02)
 
-Wave\^{}2
-
-~
-
-0.02***
-
-0.02***
-
-0.01***
-
-0.01***
-
-0.01***
-
-~
-
-~
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-Household Income (Ref.= Quintile 1)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Quintile 2
-
-~
-
-~
-
-0.05
-
-0.04
-
-0.04
-
-0.04
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Quintile 3
-
-~
-
-~
-
-0.10
-
-0.09
-
-0.09
-
-0.10
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Quintile 4
-
-~
-
-~
-
-0.11
-
-0.11*
-
-0.11*
-
-0.11*
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Quintile 5
-
-~
-
-~
-
-0.15**
-
-0.15**
-
-0.16**
-
-0.16**
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-~~~~~Quintile (No information)
-
-~
-
-~
-
-0.16*
-
-0.17*
-
-0.18*
-
-0.17*
-
-~
-
-~
-
-~
-
-(0.08)
-
-(0.08)
-
-(0.08)
-
-(0.08)
-
-Age (Ref.= 18-29)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Age 30-49
-
-~
-
-~
-
--0.02
-
--0.01
-
--0.01
-
--0.02
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Age 50-64
-
-~
-
-~
-
-0.03
-
-0.04
-
-0.03
-
-0.03
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Age 65 or more
-
-~
-
-~
-
-0.14*
-
-0.14*
-
-0.13*
-
-0.12*
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-Gender (Ref. Male)
-
-~
-
-~
-
--0.10**
-
--0.10**
-
--0.09**
-
--0.08*
-
-~
-
-~
-
-~
-
-(0.03)
-
-(0.03)
-
-(0.03)
-
-(0.03)
-
-Pol. pos (Ref.= Left)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Center
-
-~
-
-~
-
-0.08
-
-0.07
-
-0.07
-
-0.07
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Right
-
-~
-
-~
-
-0.28***
-
-0.26***
-
-0.25***
-
-0.25***
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Does not identify
-
-~
-
-~
-
-0.09*
-
-0.08
-
-0.08
-
-0.08
-
-~
-
-~
-
-~
-
-(0.04)
-
-(0.04)
-
-(0.04)
-
-(0.04)
-
-Inequality perception
-
-~
-
-~
-
-~
-
--0.15***
-
--0.14***
-
--0.14***
-
-~
-
-~
-
-~
-
-~
-
-(0.01)
-
-(0.01)
-
-(0.01)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-~
-
-0.06***
-
-0.01
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.01)
-
-(0.01)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.08***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.01)
+~
+
+Receive state benefits * Merit: Effort
+
+~
+
+~
+
+-0.05*
+
+~
+
+~
+
+~
+
+(0.02)
 
@@
 
-37741.49
-
-37740.80
-
-37897.19
-
-37723.09
-
-37697.94
-
-37668.38
+29626.31
+
+29630.35
+
+29636.59
 
@@
 
-9519
-
-9519
-
-9519
-
-9519
-
-9519
-
-9519
+7576
+
+7576
+
+7576
 
@@
 
-1701
-
-1701
-
-1701
-
-1701
-
-1701
-
-1701
+1322
+
+1322
+
+1322
 
@@
 
-0.18
-
-0.18
-
-0.17
-
-0.17
-
 0.16
@@
 
+0.16
+
 Var: Residual
 
-0.53
-
-0.53
-
-0.53
-
-0.52
-
-0.52
-
-0.52
-
-*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-Note: Model 3 to 6 are controlled by educational level, and subjective
-social status with no significant effects.
-
-}
-
-\end{table}%
-
-\begin{table}
-
-\caption{\label{tbl-interact}}
-
-\centering{
-
-Statistical models
-
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Intercept
-
-2.13***
-
-1.96***
-
-2.10***
-
-~
-
-(0.10)
-
-(0.11)
-
-(0.11)
-
-Wave
-
--0.06**
-
--0.03
-
--0.06**
-
-~
-
-(0.02)
-
-(0.02)
-
-(0.02)
-
-Wave\^{}2
-
-0.01***
-
-0.01***
-
-0.01***
-
-~
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-Merit: Talent
-
-0.01
-
-0.04
-
-0.01
-
-~
-
-(0.01)
-
-(0.02)
-
-(0.01)
-
-Merit: Effort
-
-0.08***
-
-0.08***
-
-0.07**
-
-~
-
-(0.01)
-
-(0.01)
-
-(0.02)
-
-Wave * Merit: Talent
-
-~
-
--0.01
-
-~
-
-~
-
-~
-
-(0.00)
-
-~
-
-Wave * Merit: Effort
-
-~
-
-~
-
-0.00
-
-~
-
-~
-
-~
-
-(0.00)
-
-BIC
-
-37668.38
-
-36196.86
-
-36307.66
-
-Num. obs.
-
-9519
-
-9519
-
-9519
-
-Num. groups: Individuals
-
-1701
-
-1701
-
-1701
-
-Var: Individuals (Intercept)
-
-0.16
-
-0.96
-
-0.88
-
-Var: Residual
-
-0.52
-
-0.36
-
-0.37
-
-Var: Individuals Wave
-
-~
-
-0.02
-
-0.02
-
-Var: Individuals Merit: Talent
-
-~
-
-0.10
-
-~
-
-Cov: Individuals (Intercept) Wave
-
-~
-
--0.05
-
--0.05
-
-Cov: Individuals (Intercept) Merit: Talent
-
-~
-
--0.26
-
-~
-
-Cov: Individuals Wave Merit: Talent
-
-~
-
--0.00
-
-~
-
-Var: Individuals Merit: Effort
-
-~
-
-~
-
-0.10
-
-Cov: Individuals (Intercept) Merit: Effort
-
-~
-
-~
-
--0.24
-
-Cov: Individuals Wave Merit: Effort
-
-~
-
-~
-
-0.00
+0.49
+
+0.49
+
+0.49
 
@@
 
-In this final part of the analysis, we contrast longitudinal hypotheses
-regarding changes in the relationship between meritocracy and
-justification of economic inequality over time. In hypothesis 5 we
-proposed that the association between the perception of meritocracy and
-inequality justification mitigates over time, as meritocratic ideals
-could have weakened due to critical situations associated with the COVID
-pandemic. We test this hypothesis through interaction effects, displayed
-in Table~\ref{tbl-interact}. Model 1 is shown as a baseline model, it is
-the same as Model 6 in Table~\ref{tbl-interact} but it only displays the
-variables involved in the interaction for the sake of space (all other
-variables are controlled for). Model 2 adds the interaction between the
-perception of meritocracy based on talent and time (panel wave), whereas
-Model 3 does the same but now for the perception of meritocracy related
-to effort. As observed, only the effort variable shows a significant
-interaction with the time variable (\(\beta\)=0.01, p\textless.05),
-meaning that, on average, the association between the perception of
-meritocracy based on effort and the justification of inequality
-increases by 0.01 points in every measurement point.
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.85\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-talent-1.pdf}
+
+}
+
+\caption{\label{fig-interact-talent}Interaction effect of merit based on
+talent and receive state benefits on justification of educational
+inequality}
+
+\end{figure}%
 
@@
 
-Our research sought to examine the relationship between the perception
-of meritocracy and the justification of inequality in education from a
-longitudinal perspective amid the health crisis generated by the
-COVID-19 pandemic. In general, our results showed mixed evidence about
-our hypotheses. Whereas we find support for the association between the
-perception of meritocracy and justification of inequality across time,
-the analyses that involved changes in justification of inequality are
-against our initial proposal. We argued that in times of crisis, greater
-exposure to risk due to the health context would result in a lower
-justification of inequality by citizens. In this regard, longitudinal
-models showed that the justification of inequality in education is far
-from being linear, and whereas a decreasing pattern was found between
-2017 and 2019, there is a striking increase in the last measurement
-point (2022), after the peak of the COVID pandemic. As we mentioned at
-the beginning, given our data limitations it is not possible to
-attribute such changes only to the pandemic and the related economic and
-social crises. As in the Chilean case, the pandemic period occurred
-along with a constitutional process, disentangling different forces
-driving public opinion it becomes difficult and deserves further
-qualitative and quantitative studies. A closer look to the political
-environment during the last year in the country could give additional
-hints. Some recent studies have argued that the election of a far
-left-wing constitutional assembly during this period generated a
-backslash effect, mainly due to several scandals that led to the
-delegitimation of this assembly. This could have driven preferences in a
-conservative direction, which was reflected in the election of a
-right-wing second assembly after the failed first constitutional process
-(Palanza and Sotomayor 2023; Sazo 2023).
+Our research sought to examine the justification of inequality in
+education from a longitudinal perspective amid the health crisis
+generated by the COVID-19 pandemic. In general, our results showed mixed
+evidence about our hypotheses. We argued that in times of crisis,
+greater exposure to risk due to the health context would result in a
+lower justification of inequality by citizens. In this regard,
+longitudinal models showed that the justification of inequality in
+education is far from being linear, and whereas a decreasing pattern was
+found between 2017 and 2019, there is a striking increase in the
+post-pandemic measurement points. Besides, our findings revealed that
+those who received benefits justified educational inequality less than
+those who did not. Still, given our data limitations it is not possible
+to attribute such changes only to the pandemic and the related economic
+and social crises. As during this very same period in Chile there were
+two (failed) constitutional processes, disentangling different forces
+driving public opinion becomes cumbersome and deserves further
+qualitative and quantitative studies. Along this line, some recent
+studies have argued that the election of a far left-wing constitutional
+assembly in the first process generated a backslash effect, mainly due
+to several scandals that led to the delegitimation of this assembly.
+This could have driven preferences in a conservative direction, which
+was reflected in the election of a right-wing second assembly after the
+failed first constitutional process (Palanza and Sotomayor 2023; Sazo
+2023).
+
+The main focus of this study was on the relationship between meritocracy
+and justification of inequality, arguing that those who perceive that
+the society in which they live complies with meritocratic principles,
+would be more inclined to justify inequality in education. At first
+look, the results are consistent with the hypothesis; however, some
+nuances are worth attending as the effort dimension prevails over the
+talent dimension. In other words, the perception of meritocracy in terms
+of rewarded effort is more relevant in justifying access to education
+than the perception of talent. This could be explained as that talent
+could be more associated with luck in terms of random assignment, and in
+this terms would not be enough reason (as it is effort) to justify
+educational inequality. Furthermore, we observed that the connection
+between perceived meritocracy and the justification of educational
+inequality weakened for those who received state benefits, which aligns
+with our expectations given the social context in the post-COVID era.
+One possible interpretation is that, during times of increased risk,
+people may become more skeptical of narratives emphasizing individual
+effort over collective efforts. This shift in perspective might also
+foster stronger solidarity to protect individual and societal wellbeing.
+However, further studies are necessary to explore this phenomenon using
+more specific conceptual and methodological approaches.
 
@@
 would tend to defend their interests and justify greater education
-inequality (\(H_2\)). In this regard, our results show a positive and
-robust relationship between income and the justification of inequality,
-which would align with our original hypothesis based on the rational
-interests of the better-off that would lead to a larger inequality
-justification. Besides status, we also considered inequality perception,
-as the literature on attitudes toward inequality has argued that a
-greater justification of inequality is associated with its perceived
-magnitude (\(H_3\)). Consistent with this hypothesis, our results showed
-that, throughout the years, a larger perceived economic inequality
-motivates a lower justification of inequality in education.
-
-The main focus of this study was on the relationship between meritocracy
-and justification of inequality, arguing that those who perceive that
-the society in which they live complies with meritocratic principles,
-would be more inclined to justify inequality in education (\(H_4\)). At
-first look, the results are consistent with the hypothesis; however,
-some nuances are worth attending concerning different aspects of
-meritocracy, namely effort and talent. Our results are favorable for the
-effort dimension but not confirmed for the talent dimension. In other
-words, the perception of meritocracy in terms of rewarded effort is more
-relevant in justifying access to education than the perception of
-talent. This could be explained as that talent could be more associated
-with luck in terms of random assignment, and in this terms would not be
-enough reason (as it is effort) to justify educational inequality.
-
-Finally, as our first hypothesis supported, we expected that the
-relationship between meritocratic perceptions and justification would
-tend to diminish over time. However, longitudinal evidence shows that
-the perception of meritocracy linked to effort strengthens its relation
-with the justification of inequality in education, against the temporal
-change hypothesis (\(H_5\)). In other words, meritocracy seems to gain
-force concerning justifying inequality and not the other way around as
-we expected given the social context in the post-covid era. An
-alternative interpretation is that in times of high exposure to risk,
-people tend to stick more to narratives that provide greater security
-regardless of the rest of the society - such as individual effort -,
-which might also be accompanied by lower degrees of solidarity to
-safeguard the well-being of themselves and their surrounding close
-peers. Again, this certainly needs further studies that include
-additional specific conceptual and methodological approaches to this
-phenomenon.
+inequality. In this regard, our results show a positive relationship
+between income and the justification of inequality, which would align
+with our argument that rational interests of the better-off that would
+lead to a larger inequality justification. Besides status, we also
+considered inequality perception, as the literature on attitudes toward
+inequality has argued that a greater justification of inequality is
+associated with its perceived magnitude. Consistent with this
+hypothesis, our results showed that, throughout the years, a larger
+perceived economic inequality motivates a lower justification of
+inequality in education.
 
 Among the limitations of our study, we can mention at least three.
-First, we know that inequalities in education can manifest themselves in
-terms of outcomes as well as in areas such as access to social services
-in education as part of welfare policies. In this regard, the construct
-that we have captured with our indicator can be framed more in the
-latter, while the evaluation of justice involved in this process does
-not consider aspects such as equality of educational opportunities or
-observable outcomes such as the performance of individual or educational
-institutions. Second, as it has been established by recent research
-(Castillo et al. 2023), the perception of meritocracy can be understood
-through individual attributions regarding effort and talent, but also as
-a function of structural aspects such as family of origin status or
-social capital. In this sense, the available measurements restricted us
-to the use of single indicators for effort and talent, and further
-research should consider improving the measurement quality in this
-regard. Finally, there are limitations involving the longitudinal
-dimension of the data used. Given constraints for data collection during
-the pandemic, the ELSOC study had to shorten the questionnaire in 2021
-and unfortunately, the educational justice indicator was excluded from
-that survey. Consequently, the results of temporal change have to be
-considered carefully, and further data waves could give us more
-information on this matter.
-
-The capabilities of the ELSOC longitudinal database are not limited to
-micro-level estimates. In this regard, thanks to the sampling strategy
-of the survey, it would be possible to make contextual estimates at the
-municipality level in future studies. For instance, statistical models
-could include contextual information from administrative data sources,
-allowing testing for hypotheses that include contextual socioeconomic
-inequality as well as its dynamics over time. Furthermore, this dataset
-could be used for comparing inequality justification in policy areas
-besides education, such as health and pensions. Future studies may shed
-light on how meritocratic perceptions and beliefs affect differently
-such areas and their changes over time, which would give relevant hints
-from social sciences for the discussion about societal and cultural
-changes as well as their impacts on solidarity and social cohesion.
+First, the operationalization of justification of educational inequality
+is rather limited. We counted only on a single item for this in this
+study, and further work is needed in order to assess the validity and
+reliability of measuring such construct. Second, as it has been
+established by recent research (Castillo et al. 2023), the perception of
+meritocracy can be understood through individual attributions regarding
+effort and talent, but also taking into account the role of structural
+aspects such as family status and social capital. Due to data
+limitations such dimensions were not included here, therefore providing
+a partial perspective on the areas of perceived meritocracy. Finally,
+there are limitations involving the longitudinal dimension of the data
+used. Given constraints for data collection during the pandemic, the
+ELSOC study had to shorten the questionnaire in 2021 and unfortunately,
+the educational justice indicator was excluded from that survey.
+Consequently, the results of temporal change have to be considered
+carefully, and further data waves could give us more information on this
+matter.
+
+Regarding future studies, the capabilities of the ELSOC longitudinal
+database are not limited to micro-level estimates. In this regard,
+thanks to the sampling strategy of the survey, it would be possible to
+make contextual estimates at the municipality level. For instance,
+statistical models could include contextual information from
+administrative data sources, allowing testing for hypotheses that
+include contextual socioeconomic inequality as well as its dynamics over
+time. Furthermore, this dataset could be used for comparing inequality
+justification in policy areas besides education, such as health and
+pensions. Future studies may shed light on how meritocratic perceptions
+and beliefs affect differently such areas and their changes over time,
+which would give relevant hints from social sciences for the discussion
+about societal and cultural changes as well as their impacts on
+solidarity and social cohesion.
 
@@
 671--86. \url{https://doi.org/10.1086/706980}.
-
-\bibitem[\citeproctext]{ref-Castillo2011}
-Castillo, Juan Carlos. 2011. {``Legitimacy of {Inequality} in a {Highly
-Unequal Context}: {Evidence} from the {Chilean Case}.''} \emph{Social
-Justice Research} 24 (4): 314--40.
-\url{https://doi.org/10.1007/s11211-011-0144-5}.
-
-\bibitem[\citeproctext]{ref-Castillo2012a_justice}
----------. 2012. {``Is {Inequality Becoming Just}? {Changes} in {Public
-Opinion} about {Economic Distribution} in {Chile}.''} \emph{Bulletin of
-Latin American Research} 31 (1): 1--18.
-\url{https://doi.org/10.1111/j.1470-9856.2011.00605.x}.
 
@@
 
-\bibitem[\citeproctext]{ref-confortiLegitimationInequalityAmerican1992}
-Conforti, Joseph M. 1992. {``The Legitimation of Inequality in
-{American} Education.''} \emph{The Urban Review} 24 (4): 227--38.
-\url{https://doi.org/10.1007/BF01108357}.
-
-\bibitem[\citeproctext]{ref-corvalanMercadoEscolarLibertad2016}
+\bibitem[\citeproctext]{ref-castillo_clivajes_2013}
+Castillo, Juan Carlos, Ignacio Madero-Cabib, and Alan Salamovich. 2013.
+{``Clivajes {Partidarios} y {Cambios} En Las {Preferencias
+Redistributivas} En {Chile}.''} \emph{Revista de Ciencia Pol{í}tica
+(Santiago)} 33 (2): 469--88.
+\url{https://doi.org/10.4067/S0718-090X2013000200003}.
+
+\bibitem[\citeproctext]{ref-castillo_percepcion_2019}
+Castillo, Juan Carlos, Daniel Miranda, and Diego Carrasco. 2012.
+{``Percepci{ó}n de {Desigualdad Econ{ó}mica} En {Chile}: {Medici{ó}n},
+{Diferencias} y {Determinantes}.''} \emph{Psykhe (Santiago)} 21 (1):
+99--114. \url{https://doi.org/10.4067/S0718-22282012000100007}.
+
+\bibitem[\citeproctext]{ref-castillo_meritocracia_2019}
+Castillo, Juan Carlos, Alex Torres, Jorge Atria, and Luis Maldonado.
+2019. {``Meritocracia y Desigualdad Econ{ó}mica: {Percepciones},
+Preferencias e Implicancias.''} \emph{Revista Internacional de
+Sociolog{í}a} 77 (1): 117.
+\url{https://doi.org/10.3989/ris.2019.77.1.17.114}.
+
+\bibitem[\citeproctext]{ref-coes_radiografia_2023}
+COES. 2023. {``Radiograf{í}a Del {Cambio Social}: {An{á}lisis} de
+{Resultados Longitudinales ELSOC} 2016-2021.''} Santiago de Chile,
+Chile: Centro de Estudios de Conflicto y Cohesi{ó}n Social.
+
+\bibitem[\citeproctext]{ref-condon_inequality_2020}
+Condon, Meghan, and Amber Wichowsky. 2020. {``Inequality in the {Social
+Mind}: {Social Comparison} and {Support} for {Redistribution}.''}
+\emph{The Journal of Politics} 82 (1): 149--61.
+\url{https://doi.org/10.1086/705686}.
+
+\bibitem[\citeproctext]{ref-corvalan_mercado_2016}
 Corvalán, Javier, Alejandro Carrasco, and J. E. García-Huidobro, eds.
@@
 
-\bibitem[\citeproctext]{ref-darmodyImpactsCOVID19Control2021}
+\bibitem[\citeproctext]{ref-darmody_impacts_2021}
 Darmody, Merike, Emer Smyth, and Helen Russell. 2021. {``Impacts of the
@@
 
-\bibitem[\citeproctext]{ref-dayPerceivedIdealInequality2023}
+\bibitem[\citeproctext]{ref-day_perceived_2023}
 Day, Martin V., and Michael I. Norton. 2023. {``Perceived and {Ideal
@@
 
+\bibitem[\citeproctext]{ref-donoso_when_2016}
+Donoso, Sofia. 2016. {``When {Social Movements Become} a {Democratizing
+Force}: {The Political Impact} of the {Student Movement} in {Chile}.''}
+In \emph{Research in {Social Movements}, {Conflicts} and {Change}},
+edited by Thomas Davies, Holly Eva Ryan, and Alejandro Milcíades Peña,
+39:167--96. Emerald Group Publishing Limited.
+\url{https://doi.org/10.1108/S0163-786X20160000039008}.
+
 \bibitem[\citeproctext]{ref-elsoc_estudio_2022}
@@
 {Chile}.''} Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}.
-
-\bibitem[\citeproctext]{ref-Evans2017}
-Evans, M. D. R., and Jonathan Kelley. 2017. {``Communism, {Capitalism},
-and {Images} of {Class}: {Effects} of {Reference Groups}, {Reality}, and
-{Regime} in 43 {Nations} and 110,000 {Individuals}, 1987-2009.''}
-\emph{Cross-Cultural Research} 51 (514): 315--59.
-\url{https://doi.org/10.1177/1069397116677963}.
-
-\bibitem[\citeproctext]{ref-Evans1992}
-Evans, M. D. R., Jonathan Kelley, and Tamas Kolosi. 1992. {``Images of
-{Class}: {Public Perceptions} in {Hungary} and {Australia}.''}
-\emph{American Sociological Review} 57 (4): 461.
-\url{https://doi.org/10.2307/2096095}.
-
-\bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
-García-Sánchez, Efraín, and Sofia De Carvalho. 2022. {``Las Creencias
-Que Justifican La Desigualdad Moderan La Relaci{ó}n Entre El Estatus
-Socioecon{ó}mico y El Apoyo a La Redistribuci{ó}n.''} \emph{Revista
-Internacional de Sociolog{í}a} 80 (3): e210.
-\url{https://doi.org/10.3989/ris.2022.80.3.21.29}.
 
@@
 
-\bibitem[\citeproctext]{ref-hadler_why_2005}
-Hadler, Markus. 2005. {``Why {Do People Accept Different Income
-Ratios}?: {A Multi-level Comparison} of {Thirty Countries}.''}
-\emph{Acta Sociologica} 48 (2): 131--54.
-\url{https://doi.org/10.1177/0001699305053768}.
+\bibitem[\citeproctext]{ref-hale_global_2021}
+Hale, Thomas, Noam Angrist, Rafael Goldszmidt, Beatriz Kira, Anna
+Petherick, Toby Phillips, Samuel Webster, et al. 2021. {``A Global Panel
+Database of Pandemic Policies ({Oxford COVID-19 Government Response
+Tracker}).''} \emph{Nature Human Behaviour} 5 (4): 529--38.
+\url{https://doi.org/10.1038/s41562-021-01079-8}.
 
@@
 
-\bibitem[\citeproctext]{ref-igliozziFairShareEffects2024}
+\bibitem[\citeproctext]{ref-igliozzi_fair_2024}
 Igliozzi, David, Yael Granot, and Victor Ottati. 2024. {``A Fair Share:
@@
 
-\bibitem[\citeproctext]{ref-iturra_percepcion_2023}
-Iturra, Julio, Juan Carlos Castillo, Catalina Rufs Orellana, and Luis
-Maldonado. 2023. {``La Percepci{ó}n de Desigualdad Econ{ó}mica y Su
-Influencia Sobre La Justificaci{ó}n de Las Diferencias de Ingreso
-Leg{í}timas.''} \emph{Estudios Sociol{ó}gicos de El Colegio de M{é}xico}
-41 (122): 309--52. \url{https://doi.org/10.24201/es.2023v41n122.2260}.
-
-\bibitem[\citeproctext]{ref-janmaatSubjectiveInequalityReview2013}
+\bibitem[\citeproctext]{ref-institutonacionaldeestadisticas_resultados_2022}
+Instituto Nacional de Estadísticas. 2022. {``Resultados de La Encuesta
+Nacional de Empleo: {Serie} Anualizada (2006 - 2023).''} Instituto
+Nacional de Estad{í}sticas.
+
+\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
 Janmaat, Jan Germen. 2013. {``Subjective Inequality: {A} Review of
@@
 
-\bibitem[\citeproctext]{ref-jassoJusticeEarningsNew1978}
-Jasso, Guillermina. 1978. {``On the {Justice} of {Earnings}: {A New
-Specification} of the {Justice Evaluation Function}.''} \emph{American
-Journal of Sociology} 83 (6): 1398--1419.
-\url{https://www.jstor.org/stable/2778110}.
+\bibitem[\citeproctext]{ref-joignant_informe_2020}
+Joignant, Alfredo, Matías Garretón, Nicolás M. Somma, and Tomás Campos.
+2020. {``Informe Anual: {Observatorio} de Conflictos 2020.''} Centro de
+Estudios de Conflicto y Cohesi{ó}n Social (COES).
 
@@
 Vol. 12 N{ú}m. 23 (2012); 148-174}, April.
-
-\bibitem[\citeproctext]{ref-jost_theory_2020}
-Jost, John. 2020. \emph{A Theory of System Justification}. Cambridge,
-Massachusetts: Harvard University Press.
 
@@
 
-\bibitem[\citeproctext]{ref-kelleyLegitimationInequalityOccupational1993}
-Kelley, Jonathan, and M. D. R. Evans. 1993. {``The Legitimation of
-Inequality: {Occupational} Earnings in Nine Nations.''} \emph{American
-Journal of Sociology} 99 (1): 75--125.
-\url{https://www.jstor.org/stable/2781956}.
-
 \bibitem[\citeproctext]{ref-kelley_legitimate_2021}
----------. 2021. {``Legitimate Earnings Inequality and National Welfare
-Commitment: {Correspondence} Between Economic Institutions and the Pay
-80,000+ People in 30 Nations Think Legitimate for Ordinary Jobs and for
-Elite Jobs.''} \emph{Social Science Research} 94 (February): 102446.
+Kelley, Jonathan, and M. D. R. Evans. 2021. {``Legitimate Earnings
+Inequality and National Welfare Commitment: {Correspondence} Between
+Economic Institutions and the Pay 80,000+ People in 30 Nations Think
+Legitimate for Ordinary Jobs and for Elite Jobs.''} \emph{Social Science
+Research} 94 (February): 102446.
 \url{https://doi.org/10.1016/j.ssresearch.2020.102446}.
 
-\bibitem[\citeproctext]{ref-kluegel_social_1995}
-Kluegel, James R., David S. Mason, and Bernd Wegener, eds. 1995.
-\emph{Social Justice and Political Change: Public Opinion in Capitalist
-and Post-Communist States}. Social Institutions and Social Change. New
-York: A. de Gruyter.
-
-\bibitem[\citeproctext]{ref-kluegelBeliefsStratification1981}
-Kluegel, James R., and Eliot R. Smith. 1981. {``Beliefs {About
-Stratification}.''} \emph{Annual Review of Sociology}, 29--56.
-
-\bibitem[\citeproctext]{ref-kluegelBeliefsInequalityAmericans1986}
----------. 1986. \emph{Beliefs about {Inequality}: {Americans}' {Views}
-of {What Is} and {What Ought} to {Be}}. 1st ed. Routledge.
-
-\bibitem[\citeproctext]{ref-leeFairnessPerceptionsEducational2023}
+\bibitem[\citeproctext]{ref-lee_fairness_2023}
 Lee, Jung-Sook, and Meghan Stacey. 2023. {``Fairness Perceptions of
@@
 
-\bibitem[\citeproctext]{ref-liebigSociologyJustice2016}
-Liebig, Stefan, and Carsten Sauer. 2016. {``Sociology of {Justice}.''}
-In \emph{Handbook of {Social Justice Theory} and {Research}}, edited by
-Clara Sabbagh and Manfred Schmitt, 37--59. New York, NY: Springer New
-York.
-
-\bibitem[\citeproctext]{ref-lindhPublicOpinionMarkets2015}
+\bibitem[\citeproctext]{ref-lindh_public_2015}
 Lindh, Arvid. 2015. {``Public {Opinion} Against {Markets}? {Attitudes}
@@
 887--910. \url{https://doi.org/10.1111/spol.12105}.
-
-\bibitem[\citeproctext]{ref-McCoy2007}
-McCoy, Shannon K., and Brenda Major. 2007. {``Priming Meritocracy and
-the Psychological Justification of Inequality.''} \emph{Journal of
-Experimental Social Psychology} 43 (3): 341--51.
-\url{https://doi.org/10.1016/j.jesp.2006.04.009}.
 
@@
 
-\bibitem[\citeproctext]{ref-mijs_belief_2022}
-Mijs, Jonathan, Stijn Daenekindt, Willem de Koster, and Jeroen van der
-Waal. 2022. {``Belief in {Meritocracy Reexamined}: {Scrutinizing} the
-{Role} of {Subjective Social Mobility}.''} \emph{Social Psychology
-Quarterly} 85 (2): 131--41.
-\url{https://doi.org/10.1177/01902725211063818}.
+\bibitem[\citeproctext]{ref-ministeriodedesarrollosocialyfamilia_resultados_2023}
+Ministerio de Desarrollo Social y Familia. 2023. {``Resultados de
+{Ingresos} de Los {Hogares CASEN} 2022.''} Ministerio de Desarrollo
+Social y Familia, Gobierno de Chile.
 
@@
 
-\bibitem[\citeproctext]{ref-osberg_fair_2006}
-Osberg, Lars, and Timothy Smeeding. 2006. {``{`{Fair}'} {Inequality}?
-{Attitudes} Toward {Pay Differentials}: {The United States} in
-{Comparative Perspective}.''} \emph{American Sociological Review} 71
-(3): 450--73. \url{https://doi.org/10.1177/000312240607100305}.
-
-\bibitem[\citeproctext]{ref-palanza_chile_2023}
+\bibitem[\citeproctext]{ref-oesch_subjective_2023}
+Oesch, Daniel, and Nathalie Vigna. 2023. {``Subjective Social Class Has
+a Bad Name, but Predicts Life Chances Well.''} \emph{Research in Social
+Stratification and Mobility} 83 (February): 100759.
+\url{https://doi.org/10.1016/j.rssm.2023.100759}.
+
+\bibitem[\citeproctext]{ref-palanza_chiles_2023}
 Palanza, Valeria, and Patricia Sotomayor. 2023. {``Chile's Failed
@@
 
-\bibitem[\citeproctext]{ref-Resh2016}
-Resh, Nura, and Clara Sabbagh. 2016. \emph{Justice and Education}.
-\emph{Handbook of Social Justice Theory and Research}.
-\url{https://doi.org/10.1007/978-1-4939-3216-0_19}.
+\bibitem[\citeproctext]{ref-reyes-housholder_chile_2019}
+Reyes-Housholder, Catherine, and Beatriz Roque. 2019. {``Chile 2018:
+Desaf{í}os Al Poder de g{é}nero Desde La Calle Hasta {La Moneda}.''}
+\emph{Revista de Ciencia Pol{í}tica (Santiago)} 39 (2): 191--216.
+\url{https://doi.org/10.4067/S0718-090X2019000200191}.
 
@@
 
-\bibitem[\citeproctext]{ref-schroder_income_2017}
-Schröder, Martin. 2017. {``Is {Income Inequality Related} to {Tolerance}
-for {Inequality}?''} \emph{Social Justice Research} 30 (1): 23--47.
-\url{https://doi.org/10.1007/s11211-016-0276-8}.
+\bibitem[\citeproctext]{ref-schneider_poverty_2015}
+Schneider, Simone M, and Juan Carlos Castillo. 2015. {``Poverty
+{Attributions} and the {Perceived Justice} of {Income Inequality} : {A
+Comparison} of {East} and {West Germany}.''}
+\url{https://doi.org/10.1177/0190272515589298}.
 
@@
 University Press, Incorporated.
+
+\bibitem[\citeproctext]{ref-somma_no_2021}
+Somma, Nicolás M., Matías Bargsted, Rodolfo Disi Pavlic, and Rodrigo M.
+Medel. 2021. {``No Water in the Oasis: The {Chilean Spring} of
+2019--2020.''} \emph{Social Movement Studies} 20 (4): 495--502.
+\url{https://doi.org/10.1080/14742837.2020.1727737}.
 
@@
 
-\bibitem[\citeproctext]{ref-stoilovaFairnessEducationalOpportunities2023}
-Stoilova, Rumiana, and Petya Ilieva-Trichkova. 2023. {``Fairness of
-Educational Opportunities and Income Distribution: Gender-Sensitive
-Analysis in a {European} Comparative Perspective.''} \emph{International
-Journal of Sociology and Social Policy} 43 (1/2): 272--91.
-\url{https://doi.org/10.1108/IJSSP-02-2022-0065}.
-
 \bibitem[\citeproctext]{ref-trump_income_2018}
@@
 
-\bibitem[\citeproctext]{ref-trump_when_2020}
----------. 2020. {``When and Why Is Economic Inequality Seen as Fair.''}
-\emph{Current Opinion in Behavioral Sciences} 34 (August): 46--51.
-\url{https://doi.org/10.1016/j.cobeha.2019.12.001}.
-
-\bibitem[\citeproctext]{ref-valantPoliticsAchievementGaps2016}
+\bibitem[\citeproctext]{ref-valant_politics_2016}
 Valant, Jon, and Daniel A. Newark. 2016. {``The {Politics} of
@@
 
-\bibitem[\citeproctext]{ref-vandertoorn_twenty_2014}
-van der Toorn, Jojanneke, and John Jost. 2014. {``Twenty Years of System
-Justification Theory: {Introduction} to the Special Issue on
-{`{Ideology} and System Justification Processes'}.''} \emph{Group
-Processes \& Intergroup Relations} 17 (4): 413--19.
-\url{https://doi.org/10.1177/1368430214531509}.
-
-\bibitem[\citeproctext]{ref-wegenerIllusionDistributiveJustice1987}
-Wegener, Bernd. 1987. {``The {Illusion} of {Distributive Justice}.''}
-\emph{European Sociological Review} 3 (1): 1--13.
-\url{https://www.jstor.org/stable/522527}.
-
-\bibitem[\citeproctext]{ref-Wegener1990}
----------. 1990. {``Equity, Relative Deprivation, and the Value
-Consensus Paradox.''} \emph{Social Justice Research} 4 (1): 65--86.
-\url{https://doi.org/10.1007/BF01048536}.
-
 \bibitem[\citeproctext]{ref-young_rise_1962}
@@
 
-(\#tab:multilevel-full) Multilevel model for the justification of
-economic inequality
+\begin{table}
+
+\caption{\label{tbl-multilevel-full}Complete multilevel longitudinal
+models for the justification of inequality in education}
+
+\centering{
 
@@
 
+Model 7
+
 Intercept
 
-1.84***
-
-1.87***
-
-1.80***
-
-2.39***
-
-2.22***
-
-2.13***
-
-~
-
-(0.02)
-
-(0.03)
-
-(0.09)
-
-(0.10)
-
-(0.10)
-
-(0.10)
+1.847***
+
+1.853***
+
+1.873***
+
+1.741***
+
+1.664***
+
+2.276***
+
+2.277***
+
+~
+
+(0.026)
+
+(0.038)
+
+(0.090)
+
+(0.092)
+
+(0.092)
+
+(0.105)
+
+(0.138)
 
@@
 
+~
+
+~
+
 ~~~~~Wave 2017
 
--0.14***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
+-0.167***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
 
@@
 
--0.04
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
+-0.059*
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
 
@@
 
--0.04
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
+-0.030
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
 
@@
 
-0.23***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~wave 2023
-
-0.28***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.03)
+0.204***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.030)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2023
+
+0.233***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.029)
+
+~
 
@@
 
--0.08***
-
--0.08***
-
--0.05**
-
--0.06**
-
--0.06**
-
-~
-
-~
-
-(0.02)
-
-(0.02)
-
-(0.02)
-
-(0.02)
-
-(0.02)
+-0.073***
+
+-0.073***
+
+-0.076***
+
+-0.076***
+
+-0.052**
+
+-0.052**
+
+~
+
+~
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+(0.020)
+
+(0.020)
 
@@
 
-0.02***
-
-0.02***
-
-0.01***
-
-0.01***
-
-0.01***
-
-~
-
-~
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-Education (Ref.=Primary)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~High School
-
-~
-
-~
-
--0.01
-
--0.01
-
--0.00
-
-0.00
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Technical
-
-~
-
-~
-
--0.09
-
--0.07
-
--0.06
-
--0.05
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-~~~~~Universitary
-
-~
-
-~
-
--0.15*
-
--0.11
-
--0.10
-
--0.10
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-Household Income (Ref.= Quintile 1)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Quintile 2
-
-~
-
-~
-
-0.05
-
-0.04
-
-0.04
-
-0.04
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Quintile 3
-
-~
-
-~
-
-0.10
-
-0.09
-
-0.09
-
-0.10
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Quintile 4
-
-~
-
-~
-
-0.11
-
-0.11*
-
-0.11*
-
-0.11*
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-~~~~~Quintile 5
-
-~
-
-~
-
-0.15**
-
-0.15**
-
-0.16**
-
-0.16**
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-~~~~~Quintile (No information)
-
-~
-
-~
-
-0.16*
-
-0.17*
-
-0.18*
-
-0.17*
-
-~
-
-~
-
-~
-
-(0.08)
-
-(0.08)
-
-(0.08)
-
-(0.08)
+0.016***
+
+0.016***
+
+0.016***
+
+0.016***
+
+0.013***
+
+0.013***
+
+~
+
+~
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+(0.002)
+
+Covid: receive state benefits
+
+~
+
+~
+
+-0.097*
+
+-0.100*
+
+-0.102*
+
+-0.102**
+
+-0.093*
+
+~
+
+~
+
+~
+
+(0.041)
+
+(0.040)
+
+(0.040)
+
+(0.039)
+
+(0.041)
+
+Covid: economy more important than health
+
+~
+
+~
+
+0.065***
+
+0.060**
+
+0.055**
+
+0.049**
+
+0.038*
+
+~
+
+~
+
+~
+
+(0.019)
+
+(0.019)
+
+(0.019)
+
+(0.019)
+
+(0.019)
+
+Merit: Talent
+
+~
+
+~
+
+~
+
+0.055***
+
+-0.002
+
+-0.003
+
+-0.005
+
+~
+
+~
+
+~
+
+~
+
+(0.010)
+
+(0.012)
+
+(0.012)
+
+(0.012)
+
+Merit: Effort
+
+~
+
+~
+
+~
+
+~
+
+0.096***
+
+0.084***
+
+0.083***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.013)
+
+(0.013)
+
+(0.013)
+
+Inequality perception
+
+~
+
+~
+
+~
+
+~
+
+~
+
+-0.135***
+
+-0.134***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.012)
+
+(0.012)
+
+Universitary (Ref.= Less than universitary)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+-0.048
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.046)
+
+Household Income
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.001*
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.000)
 
@@
 
--0.00
-
--0.00
-
--0.01
-
--0.01
-
-~
-
-~
-
-~
-
-(0.01)
-
-(0.01)
-
-(0.01)
-
-(0.01)
+~
+
+~
+
+~
+
+~
+
+-0.005
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.012)
 
@@
 
+~
+
+~
+
 ~~~~~Age 30-49
@@
 
--0.02
-
--0.01
-
--0.01
-
--0.02
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
+~
+
+~
+
+~
+
+~
+
+-0.033
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.051)
 
@@
 
-0.03
-
-0.04
-
-0.03
-
-0.03
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
+~
+
+~
+
+~
+
+~
+
+0.009
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.053)
 
@@
 
-0.14*
-
-0.14*
-
-0.13*
-
-0.12*
-
-~
-
-~
-
-~
-
-(0.06)
-
-(0.06)
-
-(0.06)
-
-(0.06)
+~
+
+~
+
+~
+
+~
+
+0.123
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.066)
 
@@
 
--0.10**
-
--0.10**
-
--0.09**
-
--0.08*
-
-~
-
-~
-
-~
-
-(0.03)
-
-(0.03)
-
-(0.03)
-
-(0.03)
+~
+
+~
+
+~
+
+~
+
+-0.082*
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.037)
 
@@
 
+~
+
+~
+
 ~~~~~Center
@@
 
-0.08
-
-0.07
-
-0.07
-
-0.07
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
+~
+
+~
+
+~
+
+~
+
+0.095
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.051)
 
@@
 
-0.28***
-
-0.26***
-
-0.25***
-
-0.25***
-
-~
-
-~
-
-~
-
-(0.05)
-
-(0.05)
-
-(0.05)
-
-(0.05)
+~
+
+~
+
+~
+
+~
+
+0.220***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.059)
 
@@
 
-0.09*
-
-0.08
-
-0.08
-
-0.08
-
-~
-
-~
-
-~
-
-(0.04)
-
-(0.04)
-
-(0.04)
-
-(0.04)
-
-Inequality perception
-
-~
-
-~
-
-~
-
--0.15***
-
--0.14***
-
--0.14***
-
-~
-
-~
-
-~
-
-~
-
-(0.01)
-
-(0.01)
-
-(0.01)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-~
-
-0.06***
-
-0.01
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.01)
-
-(0.01)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.08***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.01)
+~
+
+~
+
+~
+
+~
+
+0.085
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.047)
 
@@
 
-37741.49
-
-37740.80
-
-37897.19
-
-37723.09
-
-37697.94
-
-37668.38
+29667.072
+
+29670.297
+
+29682.436
+
+29668.980
+
+29626.960
+
+29510.881
+
+29626.312
 
@@
 
-9519
-
-9519
-
-9519
-
-9519
-
-9519
-
-9519
+7576
+
+7576
+
+7576
+
+7576
+
+7576
+
+7576
+
+7576
 
@@
 
-1701
-
-1701
-
-1701
-
-1701
-
-1701
-
-1701
+1322
+
+1322
+
+1322
+
+1322
+
+1322
+
+1322
+
+1322
 
@@
 
-0.18
-
-0.18
-
-0.17
-
-0.17
-
-0.16
-
-0.16
+0.185
+
+0.184
+
+0.179
+
+0.174
+
+0.170
+
+0.164
+
+0.159
 
@@
 
-0.53
-
-0.53
-
-0.53
-
-0.52
-
-0.52
-
-0.52
+0.497
+
+0.500
+
+0.500
+
+0.499
+
+0.496
+
+0.488
+
+0.488
 
@@
 
+}
+
+\end{table}%
+
 \end{document}

### 2025-01-20 02-antecedentes.qmd
ajustes typos y texto

@@
 
-Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health.
+Empirical cross-sectional evidence from Chile reveals two significant trends: low perceived meritocracy and limited justification for unequal access to social services such as education and health.
 For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit.

### 2025-01-20 03-methods.qmd
ajustes typos y texto

@@
 These variables were only included in the 2021 wave, so each individual's responses were replicated for the remaining years.
-The first variable addresses the belief regarding whether, in the context of the pandemic, it is more important to give priority to economic activity than to the health of the population, measured on a Likert scale from (1) "strongly disagree" to (2) "strongly agree".
+The first variable addresses the belief regarding whether, in the context of the pandemic, it is more important to give priority to economic activity than to the health of the population, measured on a Likert scale from (1) "strongly disagree" to (5) "strongly agree".
 The second variable is a dummy (No/Yes) that captures whether individuals received state benefits during the pandemic (e.g., Food Box, Emergency Family Income, Contribution for the Middle Class, or others).

### 2025-01-21 04-analisis.qmd
ajustes typos y texto
add submit archives

@@
 #| label: fig-correlation
-#| fig-cap: "Correlation matrix of the main variables for the first wave (2016)"
+#| fig-cap: "Correlation matrix of the main variables for the last wave (2023)"
 
@@
 omit <- "(edad)|(ess)|(sexo)|(pos_id)"
-texreg::knitreg(list(educ0, educ1, educ2, educ3, educ4, educ5, educ6, educ7)
+texreg::texreg(list(educ0, educ1, educ2, educ3, educ4, educ5, educ6, educ7)
                      ,
@@
                scalebox = 0.765,
-               float.pos = "!ht", 
+#               float.pos = "!ht", 
                booktabs = FALSE, 
@@
 omit <- "(education)|(ing_pc_rec)|(ess)|(edad)|(sexo)|(pos_id)|(covid_economia)|(des_perc)"
-texreg::knitreg(list(educ0,educ6, educ5),
-                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05.
+texreg::texreg(list(educ0,educ6, educ5),
+                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05. 
 Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
@@
                scalebox = 0.765,
-               float.pos = "!ht", 
+#               float.pos = "!ht", 
                booktabs = TRUE, 

### 2025-01-21 07-apendice.qmd
add submit archives

@@
 #omit <- "(edad)|(ess)|(sexo)|(pos_id)"
-texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7),
+texreg::texreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7),
                 custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.", 

### 2025-01-21 Title-Page.Rmd
update conflict-interest
add submit archives

@@
 ---
-title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic [^*]"
+title: "Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic. [^*]"
 linestretch: '1.5'
 output:
-  bookdown::pdf_document2:
-    template: null
+  pdf_document:
     toc: false
     number_sections: false
+    keep-tex: true
 geometry: margin=0.78in
@@
 
-Word count: 7663
+# Conflict of interest
+
+The authors report there are no competing interests to declare.
 

### 2025-01-21 blinded-paper.tex
add submit archives

--- 2025-01-20 paper.tex
@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \usepackage[noblocks]{authblk}
@@
 
-  \author{Juan Carlos Castillo}
-            \affil{%
-                  Universidad de Chile
-              }
-          \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-          \affil{%
-                  Núcleo milenio de desigualdades y oportunidades
-                  digitales (NUDOS)
-              }
-        \author{Julio Iturra}
-            \affil{%
-                  International Graduate School of Social Sciencies
-                  (BIGSSS), University of Bremen, Germany
-              }
-        \author{Kevin Carrasco}
-            \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
+  \author{ }
 
@@
 
-This document was last modified at 2025-01-19 22:27:33
-
-and it was last rendered at 2025-01-19 22:27:33
-
 \section{Introduction}\label{introduction}
@@
 Empirical cross-sectional evidence from Chile reveals two significant
-trends: low public support for the perception of meritocracy and limited
-justification for unequal access to social services such as education
-and health. For instance, Castillo, Madero-Cabib, and Salamovich (2013)
-analyzed data from the International Social Survey Programme (ISSP) in
-1999 and 2009, highlighting consistently low support for inequality
-based on resource allocation or individual merit. Similarly, Castillo,
-Miranda, and Carrasco (2012) demonstrated widespread skepticism among
-Chileans regarding the meritocratic system, with many disagreeing with
-the notion that rewards are fairly distributed based on talent and
-effort. In line with this evidence, we propose that:
+trends: low perceived meritocracy and limited justification for unequal
+access to social services such as education and health. For instance,
+Castillo, Madero-Cabib, and Salamovich (2013) analyzed data from the
+International Social Survey Programme (ISSP) in 1999 and 2009,
+highlighting consistently low support for inequality based on resource
+allocation or individual merit. Similarly, Castillo, Miranda, and
+Carrasco (2012) demonstrated widespread skepticism among Chileans
+regarding the meritocratic system, with many disagreeing with the notion
+that rewards are fairly distributed based on talent and effort. In line
+with this evidence, we propose that:
 
@@
 the population, measured on a Likert scale from (1) ``strongly
-disagree'' to (2) ``strongly agree''. The second variable is a dummy
+disagree'' to (5) ``strongly agree''. The second variable is a dummy
 (No/Yes) that captures whether individuals received state benefits
@@
 \caption{\label{fig-correlation}Correlation matrix of the main variables
-for the first wave (2016)}
+for the last wave (2023)}
 
@@
 
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Model 8
-
-Intercept
-
-1.847***
-
-1.853***
-
-1.920***
-
-1.776***
-
-1.641***
-
-1.563***
-
-2.174***
-
-2.183***
-
-~
-
-(0.026)
-
-(0.038)
-
-(0.048)
-
-(0.064)
-
-(0.068)
-
-(0.069)
-
-(0.086)
-
-(0.120)
-
-Wave (Ref.= Wave 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.059*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.030
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.204***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.030)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.233***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.073***
-
--0.073***
-
--0.073***
-
--0.076***
-
--0.076***
-
--0.052**
-
--0.052**
-
-~
-
-~
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-Wave\^{}2
-
-~
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.013***
-
-0.013***
-
-~
-
-~
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-Covid: receive state benefits
-
-~
-
-~
-
--0.095*
-
--0.097*
-
--0.100*
-
--0.102*
-
--0.102**
-
--0.093*
-
-~
-
-~
-
-~
-
-(0.041)
-
-(0.041)
-
-(0.040)
-
-(0.040)
-
-(0.039)
-
-(0.041)
-
-Covid: economy more important than health
-
-~
-
-~
-
-~
-
-0.065***
-
-0.060**
-
-0.055**
-
-0.049**
-
-0.038*
-
-~
-
-~
-
-~
-
-~
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-~
-
-0.055***
-
--0.002
-
--0.003
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.096***
-
-0.084***
-
-0.083***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.013)
-
-(0.013)
-
-(0.013)
-
-Inequality perception
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.135***
-
--0.134***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-Universitary (Ref.= Less than universitary)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.048
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.046)
-
-Household income
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.001*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.000)
-
-Controls
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-Yes
-
-BIC
-
-29667.072
-
-29670.297
-
-29678.484
-
-29682.436
-
-29668.980
-
-29626.960
-
-29510.881
-
-29626.312
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.185
-
-0.184
-
-0.183
-
-0.179
-
-0.174
-
-0.170
-
-0.164
-
-0.159
-
-Var: Residual
-
-0.497
-
-0.500
-
-0.500
-
-0.500
-
-0.499
-
-0.496
-
-0.488
-
-0.488
-
-*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-Note: Model 8 is controlled by age, gender, political position, and
-subjective social status.
+\begin{center}
+\scalebox{0.765}{
+\begin{tabular}{l c c c c c c c c}
+\hline
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 \\
+\hline
+Intercept                                   & $1.847^{***}$  & $1.853^{***}$  & $1.920^{***}$  & $1.776^{***}$  & $1.641^{***}$  & $1.563^{***}$  & $2.174^{***}$  & $2.183^{***}$  \\
+                                            & $(0.026)$      & $(0.038)$      & $(0.048)$      & $(0.064)$      & $(0.068)$      & $(0.069)$      & $(0.086)$      & $(0.120)$      \\
+Wave (Ref.= Wave 2016)                      &                &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                &                \\
+\quad Wave 2017                             & $-0.167^{***}$ &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+\quad Wave 2018                             & $-0.059^{*}$   &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+\quad Wave 2019                             & $-0.030$       &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+\quad Wave 2022                             & $0.204^{***}$  &                &                &                &                &                &                &                \\
+                                            & $(0.030)$      &                &                &                &                &                &                &                \\
+\quad Wave 2023                             & $0.233^{***}$  &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+Wave                                        &                & $-0.073^{***}$ & $-0.073^{***}$ & $-0.073^{***}$ & $-0.076^{***}$ & $-0.076^{***}$ & $-0.052^{**}$  & $-0.052^{**}$  \\
+                                            &                & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      \\
+Wave$^2$                                    &                & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.013^{***}$  & $0.013^{***}$  \\
+                                            &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
+Covid: receive state benefits               &                &                & $-0.095^{*}$   & $-0.097^{*}$   & $-0.100^{*}$   & $-0.102^{*}$   & $-0.102^{**}$  & $-0.093^{*}$   \\
+                                            &                &                & $(0.041)$      & $(0.041)$      & $(0.040)$      & $(0.040)$      & $(0.039)$      & $(0.041)$      \\
+Covid: economy more important than health   &                &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
+                                            &                &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
+Merit: Talent                               &                &                &                &                & $0.055^{***}$  & $-0.002$       & $-0.003$       & $-0.005$       \\
+                                            &                &                &                &                & $(0.010)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Effort                               &                &                &                &                &                & $0.096^{***}$  & $0.084^{***}$  & $0.083^{***}$  \\
+                                            &                &                &                &                &                & $(0.013)$      & $(0.013)$      & $(0.013)$      \\
+Inequality perception                       &                &                &                &                &                &                & $-0.135^{***}$ & $-0.134^{***}$ \\
+                                            &                &                &                &                &                &                & $(0.012)$      & $(0.012)$      \\
+Universitary (Ref.= Less than universitary) &                &                &                &                &                &                &                & $-0.048$       \\
+                                            &                &                &                &                &                &                &                & $(0.046)$      \\
+Household income                            &                &                &                &                &                &                &                & $0.001^{*}$    \\
+                                            &                &                &                &                &                &                &                & $(0.000)$      \\
+\hline
+Controls                                    & No             & No             & No             & No             & No             & No             & No             & Yes            \\
+BIC                                         & $29667.072$    & $29670.297$    & $29678.484$    & $29682.436$    & $29668.980$    & $29626.960$    & $29510.881$    & $29626.312$    \\
+Num. obs.                                   & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         \\
+Num. groups: Individuals                    & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         \\
+Var: Individuals (Intercept)                & $0.185$        & $0.184$        & $0.183$        & $0.179$        & $0.174$        & $0.170$        & $0.164$        & $0.159$        \\
+Var: Residual                               & $0.497$        & $0.500$        & $0.500$        & $0.500$        & $0.499$        & $0.496$        & $0.488$        & $0.488$        \\
+\hline
+\multicolumn{9}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.
+
+Note: Model 8 is controlled by age, gender, political position, and subjective social status.}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 
@@
 
-~
-
-Model 8
-
-Model 9
-
-Model 10
-
-Intercept
-
-2.18***
-
-2.03***
-
-2.09***
-
-~
-
-(0.12)
-
-(0.13)
-
-(0.13)
-
-Wave
-
--0.05**
-
--0.05*
-
--0.05**
-
-~
-
-(0.02)
-
-(0.02)
-
-(0.02)
-
-Wave\^{}2
-
-0.01***
-
-0.01***
-
-0.01***
-
-~
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-Covid: receive state benefits
-
--0.09*
-
-0.11
-
-0.03
-
-~
-
-(0.04)
-
-(0.07)
-
-(0.07)
-
-Merit: Talent
-
--0.00
-
-0.05*
-
--0.01
-
-~
-
-(0.01)
-
-(0.02)
-
-(0.01)
-
-Merit: Effort
-
-0.08***
-
-0.08***
-
-0.12***
-
-~
-
-(0.01)
-
-(0.01)
-
-(0.02)
-
-Receive state benefits * Merit: Talent
-
-~
-
--0.07**
-
-~
-
-~
-
-~
-
-(0.02)
-
-~
-
-Receive state benefits * Merit: Effort
-
-~
-
-~
-
--0.05*
-
-~
-
-~
-
-~
-
-(0.02)
-
-BIC
-
-29626.31
-
-29630.35
-
-29636.59
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.16
-
-0.16
-
-0.16
-
-Var: Residual
-
-0.49
-
-0.49
-
-0.49
-
-\item *
-
-** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-Note: All the models are controlled by educational level, income
-quintile, subjective social status, age, gender, political position, and
-inequality perception
+\begin{center}
+\scalebox{0.765}{
+\begin{tabular}{l c c c}
+\hline
+ & Model 8 & Model 9 & Model 10 \\
+\hline
+Intercept                              & $2.18^{***}$ & $2.03^{***}$ & $2.09^{***}$ \\
+                                       & $(0.12)$     & $(0.13)$     & $(0.13)$     \\
+Wave                                   & $-0.05^{**}$ & $-0.05^{*}$  & $-0.05^{**}$ \\
+                                       & $(0.02)$     & $(0.02)$     & $(0.02)$     \\
+Wave$^2$                               & $0.01^{***}$ & $0.01^{***}$ & $0.01^{***}$ \\
+                                       & $(0.00)$     & $(0.00)$     & $(0.00)$     \\
+Covid: receive state benefits          & $-0.09^{*}$  & $0.11$       & $0.03$       \\
+                                       & $(0.04)$     & $(0.07)$     & $(0.07)$     \\
+Merit: Talent                          & $-0.00$      & $0.05^{*}$   & $-0.01$      \\
+                                       & $(0.01)$     & $(0.02)$     & $(0.01)$     \\
+Merit: Effort                          & $0.08^{***}$ & $0.08^{***}$ & $0.12^{***}$ \\
+                                       & $(0.01)$     & $(0.01)$     & $(0.02)$     \\
+Receive state benefits * Merit: Talent &              & $-0.07^{**}$ &              \\
+                                       &              & $(0.02)$     &              \\
+Receive state benefits * Merit: Effort &              &              & $-0.05^{*}$  \\
+                                       &              &              & $(0.02)$     \\
+\hline
+BIC                                    & $29626.31$   & $29630.35$   & $29636.59$   \\
+Num. obs.                              & $7576$       & $7576$       & $7576$       \\
+Num. groups: Individuals               & $1322$       & $1322$       & $1322$       \\
+Var: Individuals (Intercept)           & $0.16$       & $0.16$       & $0.16$       \\
+Var: Residual                          & $0.49$       & $0.49$       & $0.49$       \\
+\hline
+\multicolumn{4}{l}{\scriptsize{\parbox[t]{\linewidth}{%
+*** p < 0.001; ** p < 0.01; * p < 0.05. Note: All the models are controlled by educational level, income quintile,\\
+subjective social status, age, gender, political position, and inequality perception.}}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 
@@
 
-\section*{(APPENDIX) Appendix}\label{appendix-appendix}
+\section*{Appendix}\label{appendix-appendix}
 \addcontentsline{toc}{section}{(APPENDIX) Appendix}
-
-\appendix \section{Appendix}
 
@@
 
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Intercept
-
-1.847***
-
-1.853***
-
-1.873***
-
-1.741***
-
-1.664***
-
-2.276***
-
-2.277***
-
-~
-
-(0.026)
-
-(0.038)
-
-(0.090)
-
-(0.092)
-
-(0.092)
-
-(0.105)
-
-(0.138)
-
-Wave (Ref.= Wave 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.059*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.030
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.204***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.030)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.233***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.073***
-
--0.073***
-
--0.076***
-
--0.076***
-
--0.052**
-
--0.052**
-
-~
-
-~
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-Wave\^{}2
-
-~
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.013***
-
-0.013***
-
-~
-
-~
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-Covid: receive state benefits
-
-~
-
-~
-
--0.097*
-
--0.100*
-
--0.102*
-
--0.102**
-
--0.093*
-
-~
-
-~
-
-~
-
-(0.041)
-
-(0.040)
-
-(0.040)
-
-(0.039)
-
-(0.041)
-
-Covid: economy more important than health
-
-~
-
-~
-
-0.065***
-
-0.060**
-
-0.055**
-
-0.049**
-
-0.038*
-
-~
-
-~
-
-~
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-0.055***
-
--0.002
-
--0.003
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-0.096***
-
-0.084***
-
-0.083***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.013)
-
-(0.013)
-
-(0.013)
-
-Inequality perception
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.135***
-
--0.134***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-Universitary (Ref.= Less than universitary)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.048
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.046)
-
-Household Income
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.001*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.000)
-
-Subjective Social Status
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-Age (Ref.= 18-29)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Age 30-49
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.033
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.051)
-
-~~~~~Age 50-64
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.009
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.053)
-
-~~~~~Age 65 or more
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.123
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.066)
-
-Gender (Ref. Male)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.082*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.037)
-
-Pol. pos (Ref.= Left)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Center
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.095
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.051)
-
-~~~~~Right
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.220***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.059)
-
-~~~~~Does not identify
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.085
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.047)
-
-BIC
-
-29667.072
-
-29670.297
-
-29682.436
-
-29668.980
-
-29626.960
-
-29510.881
-
-29626.312
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.185
-
-0.184
-
-0.179
-
-0.174
-
-0.170
-
-0.164
-
-0.159
-
-Var: Residual
-
-0.497
-
-0.500
-
-0.500
-
-0.499
-
-0.496
-
-0.488
-
-0.488
-
-*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-
+\begin{center}
+\scalebox{0.765}{
+\begin{tabular}{l c c c c c c c}
+\hline
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
+\hline
+Intercept                                   & $1.847^{***}$  & $1.853^{***}$  & $1.873^{***}$  & $1.741^{***}$  & $1.664^{***}$  & $2.276^{***}$  & $2.277^{***}$  \\
+                                            & $(0.026)$      & $(0.038)$      & $(0.090)$      & $(0.092)$      & $(0.092)$      & $(0.105)$      & $(0.138)$      \\
+Wave (Ref.= Wave 2016)                      &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                \\
+\quad Wave 2017                             & $-0.167^{***}$ &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+\quad Wave 2018                             & $-0.059^{*}$   &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+\quad Wave 2019                             & $-0.030$       &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+\quad Wave 2022                             & $0.204^{***}$  &                &                &                &                &                &                \\
+                                            & $(0.030)$      &                &                &                &                &                &                \\
+\quad Wave 2023                             & $0.233^{***}$  &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+Wave                                        &                & $-0.073^{***}$ & $-0.073^{***}$ & $-0.076^{***}$ & $-0.076^{***}$ & $-0.052^{**}$  & $-0.052^{**}$  \\
+                                            &                & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      \\
+Wave$^2$                                    &                & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.013^{***}$  & $0.013^{***}$  \\
+                                            &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
+Covid: receive state benefits               &                &                & $-0.097^{*}$   & $-0.100^{*}$   & $-0.102^{*}$   & $-0.102^{**}$  & $-0.093^{*}$   \\
+                                            &                &                & $(0.041)$      & $(0.040)$      & $(0.040)$      & $(0.039)$      & $(0.041)$      \\
+Covid: economy more important than health   &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
+                                            &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
+Merit: Talent                               &                &                &                & $0.055^{***}$  & $-0.002$       & $-0.003$       & $-0.005$       \\
+                                            &                &                &                & $(0.010)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Effort                               &                &                &                &                & $0.096^{***}$  & $0.084^{***}$  & $0.083^{***}$  \\
+                                            &                &                &                &                & $(0.013)$      & $(0.013)$      & $(0.013)$      \\
+Inequality perception                       &                &                &                &                &                & $-0.135^{***}$ & $-0.134^{***}$ \\
+                                            &                &                &                &                &                & $(0.012)$      & $(0.012)$      \\
+Universitary (Ref.= Less than universitary) &                &                &                &                &                &                & $-0.048$       \\
+                                            &                &                &                &                &                &                & $(0.046)$      \\
+Household Income                            &                &                &                &                &                &                & $0.001^{*}$    \\
+                                            &                &                &                &                &                &                & $(0.000)$      \\
+Subjective Social Status                    &                &                &                &                &                &                & $-0.005$       \\
+                                            &                &                &                &                &                &                & $(0.012)$      \\
+Age (Ref.= 18-29)                           &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                \\
+\quad Age 30-49                             &                &                &                &                &                &                & $-0.033$       \\
+                                            &                &                &                &                &                &                & $(0.051)$      \\
+\quad Age 50-64                             &                &                &                &                &                &                & $0.009$        \\
+                                            &                &                &                &                &                &                & $(0.053)$      \\
+\quad Age 65 or more                        &                &                &                &                &                &                & $0.123$        \\
+                                            &                &                &                &                &                &                & $(0.066)$      \\
+Gender (Ref. Male)                          &                &                &                &                &                &                & $-0.082^{*}$   \\
+                                            &                &                &                &                &                &                & $(0.037)$      \\
+Pol. pos (Ref.= Left)                       &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                \\
+\quad Center                                &                &                &                &                &                &                & $0.095$        \\
+                                            &                &                &                &                &                &                & $(0.051)$      \\
+\quad Right                                 &                &                &                &                &                &                & $0.220^{***}$  \\
+                                            &                &                &                &                &                &                & $(0.059)$      \\
+\quad Does not identify                     &                &                &                &                &                &                & $0.085$        \\
+                                            &                &                &                &                &                &                & $(0.047)$      \\
+\hline
+BIC                                         & $29667.072$    & $29670.297$    & $29682.436$    & $29668.980$    & $29626.960$    & $29510.881$    & $29626.312$    \\
+Num. obs.                                   & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         \\
+Num. groups: Individuals                    & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         \\
+Var: Individuals (Intercept)                & $0.185$        & $0.184$        & $0.179$        & $0.174$        & $0.170$        & $0.164$        & $0.159$        \\
+Var: Residual                               & $0.497$        & $0.500$        & $0.500$        & $0.499$        & $0.496$        & $0.488$        & $0.488$        \\
+\hline
+\multicolumn{8}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.}}
+\end{tabular}
 }
+\label{table:coefficients}
+\end{center}
+
+}
 

### 2025-01-21 paper.qmd
add submit archives

@@
       - title.tex
+    keep-tex: true
     include-in-header:
@@
 
-This document was last modified at {{< modified >}}
+```{r}
+# This document was last modified at {{< modified >}}
+# 
+# and it was last rendered at {{< now >}}
 
-and it was last rendered at {{< now >}}
+```
 

### 2025-01-21 submit/blinded-paper.tex
add submit archives

--- 2025-01-21 blinded-paper.tex
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/merit_plot.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{merit_plot-rr.png}
 
@@
 \endlastfoot
-\includegraphics{output/tables/desc02.png} \\
+\includegraphics{desc02-rr.png} \\
 \end{longtable}
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/alluvial_dep.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{alluvial_dep-rr.png}
 
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/years_plot.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{years_plot-rr.png}
 
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/corr.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{corr-rr.png}
 
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-talent-1.pdf}
+\includegraphics[width=0.85\textwidth,height=\textheight]{fig-interact-talent-1.pdf}
 

### 2025-04-15 paper.qmd
render pdf times new roman 10

@@
   pdf:
+    classoption: 10pt
     geometry: 
@@
     keep-tex: true
+    number-sections: true
     include-in-header:
@@
         \renewcommand\Affilfont{\small}
-    number-sections: true
+        \renewcommand\Affilfont{\small}
+        \usepackage{mathptmx} % Times
 editor: source

### 2025-04-15 01-introduction.qmd
add funding and render tex

@@
+# Abstract
+
+Education is considered a key tool for social mobility and equality of opportunities.
+However, despite the widespread value of education, disparities in educational outcomes persist over time.
+This paper is guided by the following research questions: To what extent are such educational disparities justified in society?
+Is the justification of educational inequality affected in periods of personal and social vulnerability, as in the COVID pandemic?
+And, what are the main factors driving this kind of justification?
+The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
+This economic and cultural environment has promoted meritocratic ideals, whereby individual talent and effort are considered key to get ahead in life, disregarding opportunities linked to social origin.
+The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic).
+Furthermore, we argue that such changes could be linked to a challenge of meritocratic ideals.
+For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2023, 6 waves, N = 2,927).
+Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
+The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.\newline
+  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
+
 # Introduction
@@
 As the last PISA report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023].
-Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021].
-Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023].
+Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmody_impacts_2021].
+Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@lee_fairness_2023].
 In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.
@@
 The Chilean case is particularly relevant when it comes to the study of inequalities in education.
-The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016].
+The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalan_mercado_2016].
 The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019].
-Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
+Previous research in Chile has emphasized the concept of an "educational market" [@corvalan_mercado_2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities.
 Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022].

### 2025-04-15 Title-Page.Rmd
add funding and render tex

@@
 [^*]: Corresponding author (juancastillov@uchile.cl - [jc-castillo.com](https://juancarloscastillo.github.io/jc-castillo/) -  Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile).
-The authors recognize the support of the National Research and Development Agency (ANID) through the grant from the National Fund for Scientific and Technological Development (FONDECYT) N° 1210847; from the Center for Conflict and Social Cohesion Studies (COES) grant Financing Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046
+The authors recognize the support of the National Research and Development Agency (ANID) through the grant from the National Fund for Scientific and Technological Development (FONDECYT) N° 1210847 and N° 1250518; from the Center for Conflict and Social Cohesion Studies (COES) grant Financing Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046

### 2025-04-15 paper.qmd
add funding and render tex

@@
 citeproc: true
-abstract: | 
-  Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-This paper is guided by the following research questions: To what extent are such educational disparities justified in society?
-Is the justification of educational inequality affected in periods of personal and social vulnerability, as in the COVID pandemic?
-And, what are the main factors driving this kind of justification?
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-This economic and cultural environment has promoted meritocratic ideals, whereby individual talent and effort are considered key to get ahead in life, disregarding opportunities linked to social origin.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic).
-Furthermore, we argue that such changes could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2023, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.\newline
-  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
 ---
@@
 
+{{< include Title-page.Rmd >}}
+
+\newpage
+
 {{< include 01-introduction.qmd >}}

### 2025-04-15 submit/paper.tex
add submit archives
add funding and render tex

--- 2025-01-21 blinded-paper.tex
@@
 \documentclass[
-]{article}
+  12pt,
+  10pt]{article}
 
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \usepackage{xcolor}
-\usepackage[top=30mm,left=20mm]{geometry}
+\usepackage[margin=0.78in]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 
+\usepackage{times}
+\usepackage{caption}
+\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
+\usepackage[figurename=Fig.]{caption}
+\usepackage{graphicx}
+\usepackage{float}
 \usepackage{booktabs}
@@
 \usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 \usepackage[noblocks]{authblk}
@@
 \renewcommand\Affilfont{\small}
+\renewcommand\Affilfont{\small}
+\usepackage{mathptmx} % Times
+\usepackage{times}
+\usepackage{caption}
+\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
+\usepackage[figurename=Fig.]{caption}
+\usepackage{graphicx}
+\usepackage{float}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.},
+  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic. },
   pdfauthor={Juan Carlos Castillo; Julio Iturra; Kevin Carrasco},
@@
 role of perceptions of inequality and meritocracy during the COVID
-pandemic.}
-
-  \author{ }
+pandemic. \footnote{Corresponding author (juancastillov@uchile.cl -
+  \href{https://juancarloscastillo.github.io/jc-castillo/}{jc-castillo.com}
+  - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile). The authors
+  recognize the support of the National Research and Development Agency
+  (ANID) through the grant from the National Fund for Scientific and
+  Technological Development (FONDECYT) N° 1210847 and N° 1250518; from
+  the Center for Conflict and Social Cohesion Studies (COES) grant
+  Financing Fund for Research Centers in Priority Areas (FONDAP) N°
+  1523A0005; and Millenium Nucleus on Digital Inequalities and
+  Opportunities (NUDOS) NCS2022\_046}}
+
+  \author{Juan Carlos Castillo}
+            \affil{%
+                  Universidad de Chile
+              }
+          \affil{%
+                  Centro de estudios del conflicto y cohesión social
+                  (COES)
+              }
+          \affil{%
+                  Núcleo milenio de desigualdades y oportunidades
+                  digitales (NUDOS)
+              }
+        \author{Julio Iturra}
+            \affil{%
+                  International Graduate School of Social Sciencies
+                  (BIGSSS), University of Bremen, Germany
+              }
+        \author{Kevin Carrasco}
+            \affil{%
+                  Centro de estudios del conflicto y cohesión social
+                  (COES)
+              }
 
@@
 \maketitle
-\begin{abstract}
+
+\setstretch{1.5}
+Juan Carlos Castillo, Department of Sociology, Universidad de Chile,
+Santiago de Chile. *
+
+ORCID: 0000-0003-1265-7854
+
+Julio Iturra, Bremen International Graduate School of Social Sciences,
+Bremen, Germany.
+
+ORCID: 0000-0003-1993-3479
+
+Kevin Carrasco, Center for Social Conflict and Cohesion Studies
+
+ORCID: 0000-0002-9348-8361
+
+\section{Conflict of interest}\label{conflict-of-interest}
+
+The authors report there are no competing interests to declare.
+
+\newpage
+
+\section{Abstract}\label{abstract}
+
 Education is considered a key tool for social mobility and equality of
@@
 justification, COVID-19
-\end{abstract}
 
@@
 
+\caption{}
 \begin{center}
@@
                                             &                &                & $(0.041)$      & $(0.041)$      & $(0.040)$      & $(0.040)$      & $(0.039)$      & $(0.041)$      \\
-Covid: economy more important than health   &                &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
-                                            &                &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
+Covid: economy more important               &                &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
+than health                                 &                &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
 Merit: Talent                               &                &                &                &                & $0.055^{***}$  & $-0.002$       & $-0.003$       & $-0.005$       \\
@@
                                             &                &                &                &                &                &                & $(0.012)$      & $(0.012)$      \\
-Universitary (Ref.= Less than universitary) &                &                &                &                &                &                &                & $-0.048$       \\
-                                            &                &                &                &                &                &                &                & $(0.046)$      \\
+Universitary                                &                &                &                &                &                &                &                & $-0.048$       \\
+(Ref.= Less than universitary)              &                &                &                &                &                &                &                & $(0.046)$      \\
 Household income                            &                &                &                &                &                &                &                & $0.001^{*}$    \\
@@
 \multicolumn{9}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.
-
 Note: Model 8 is controlled by age, gender, political position, and subjective social status.}}
@@
 \begin{tabular}{l c c c}
-\hline
+\toprule
  & Model 8 & Model 9 & Model 10 \\
-\hline
+\midrule
 Intercept                              & $2.18^{***}$ & $2.03^{***}$ & $2.09^{***}$ \\
@@
                                        &              &              & $(0.02)$     \\
-\hline
+\midrule
 BIC                                    & $29626.31$   & $29630.35$   & $29636.59$   \\
@@
 Var: Residual                          & $0.49$       & $0.49$       & $0.49$       \\
-\hline
-\multicolumn{4}{l}{\scriptsize{\parbox[t]{\linewidth}{%
-*** p < 0.001; ** p < 0.01; * p < 0.05. Note: All the models are controlled by educational level, income quintile,\\
-subjective social status, age, gender, political position, and inequality perception.}}}
+\bottomrule
+\multicolumn{4}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05. 
+Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception}}
 \end{tabular}
@@
 
-\section*{Appendix}\label{appendix-appendix}
+\section*{(APPENDIX) Appendix}\label{appendix-appendix}
 \addcontentsline{toc}{section}{(APPENDIX) Appendix}
+
+\appendix \section{Appendix}
 

### 2025-04-17 01-introduction.qmd
update submit

@@
-# Abstract
-
-Education is considered a key tool for social mobility and equality of opportunities.
-However, despite the widespread value of education, disparities in educational outcomes persist over time.
-This paper is guided by the following research questions: To what extent are such educational disparities justified in society?
-Is the justification of educational inequality affected in periods of personal and social vulnerability, as in the COVID pandemic?
-And, what are the main factors driving this kind of justification?
-The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education.
-This economic and cultural environment has promoted meritocratic ideals, whereby individual talent and effort are considered key to get ahead in life, disregarding opportunities linked to social origin.
-The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic).
-Furthermore, we argue that such changes could be linked to a challenge of meritocratic ideals.
-For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2023, 6 waves, N = 2,927).
-Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.\newline
-  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
-
 # Introduction

### 2025-04-17 Title-Page.Rmd
update submit

@@
     number_sections: false
-    keep-tex: true
+    keep_tex: true
 geometry: margin=0.78in
@@
 fontsize: 12pt
-
 ---

### 2025-04-17 paper.tex
update submit

@@
 \documentclass[
-]{article}
+  10pt]{article}
 
@@
 \renewcommand\Affilfont{\small}
+\renewcommand\Affilfont{\small}
+\usepackage{mathptmx} % Times
 \makeatletter
@@
 
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Model 8
-
-Intercept
-
-1.847***
-
-1.853***
-
-1.920***
-
-1.776***
-
-1.641***
-
-1.563***
-
-2.174***
-
-2.183***
-
-~
-
-(0.026)
-
-(0.038)
-
-(0.048)
-
-(0.064)
-
-(0.068)
-
-(0.069)
-
-(0.086)
-
-(0.120)
-
-Wave (Ref.= Wave 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.059*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.030
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.204***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.030)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.233***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.073***
-
--0.073***
-
--0.073***
-
--0.076***
-
--0.076***
-
--0.052**
-
--0.052**
-
-~
-
-~
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-Wave\^{}2
-
-~
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.013***
-
-0.013***
-
-~
-
-~
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-Covid: receive state benefits
-
-~
-
-~
-
--0.095*
-
--0.097*
-
--0.100*
-
--0.102*
-
--0.102**
-
--0.093*
-
-~
-
-~
-
-~
-
-(0.041)
-
-(0.041)
-
-(0.040)
-
-(0.040)
-
-(0.039)
-
-(0.041)
-
-Covid: economy more important than health
-
-~
-
-~
-
-~
-
-0.065***
-
-0.060**
-
-0.055**
-
-0.049**
-
-0.038*
-
-~
-
-~
-
-~
-
-~
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-~
-
-0.055***
-
--0.002
-
--0.003
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.096***
-
-0.084***
-
-0.083***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.013)
-
-(0.013)
-
-(0.013)
-
-Inequality perception
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.135***
-
--0.134***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-Universitary (Ref.= Less than universitary)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.048
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.046)
-
-Household income
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.001*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.000)
-
-Controls
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-Yes
-
-BIC
-
-29667.072
-
-29670.297
-
-29678.484
-
-29682.436
-
-29668.980
-
-29626.960
-
-29510.881
-
-29626.312
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.185
-
-0.184
-
-0.183
-
-0.179
-
-0.174
-
-0.170
-
-0.164
-
-0.159
-
-Var: Residual
-
-0.497
-
-0.500
-
-0.500
-
-0.500
-
-0.499
-
-0.496
-
-0.488
-
-0.488
-
-*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-Note: Model 8 is controlled by age, gender, political position, and
-subjective social status.
+\caption{}
+\begin{center}
+\scalebox{0.765}{
+\begin{tabular}{l c c c c c c c c}
+\hline
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 \\
+\hline
+Intercept                                   & $1.847^{***}$  & $1.853^{***}$  & $1.920^{***}$  & $1.776^{***}$  & $1.641^{***}$  & $1.563^{***}$  & $2.174^{***}$  & $2.183^{***}$  \\
+                                            & $(0.026)$      & $(0.038)$      & $(0.048)$      & $(0.064)$      & $(0.068)$      & $(0.069)$      & $(0.086)$      & $(0.120)$      \\
+Wave (Ref.= Wave 2016)                      &                &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                &                \\
+\quad Wave 2017                             & $-0.167^{***}$ &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+\quad Wave 2018                             & $-0.059^{*}$   &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+\quad Wave 2019                             & $-0.030$       &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+\quad Wave 2022                             & $0.204^{***}$  &                &                &                &                &                &                &                \\
+                                            & $(0.030)$      &                &                &                &                &                &                &                \\
+\quad Wave 2023                             & $0.233^{***}$  &                &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                &                \\
+Wave                                        &                & $-0.073^{***}$ & $-0.073^{***}$ & $-0.073^{***}$ & $-0.076^{***}$ & $-0.076^{***}$ & $-0.052^{**}$  & $-0.052^{**}$  \\
+                                            &                & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      \\
+Wave$^2$                                    &                & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.013^{***}$  & $0.013^{***}$  \\
+                                            &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
+Covid: receive state benefits               &                &                & $-0.095^{*}$   & $-0.097^{*}$   & $-0.100^{*}$   & $-0.102^{*}$   & $-0.102^{**}$  & $-0.093^{*}$   \\
+                                            &                &                & $(0.041)$      & $(0.041)$      & $(0.040)$      & $(0.040)$      & $(0.039)$      & $(0.041)$      \\
+Covid: economy more important than health   &                &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
+                                            &                &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
+Merit: Talent                               &                &                &                &                & $0.055^{***}$  & $-0.002$       & $-0.003$       & $-0.005$       \\
+                                            &                &                &                &                & $(0.010)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Effort                               &                &                &                &                &                & $0.096^{***}$  & $0.084^{***}$  & $0.083^{***}$  \\
+                                            &                &                &                &                &                & $(0.013)$      & $(0.013)$      & $(0.013)$      \\
+Inequality perception                       &                &                &                &                &                &                & $-0.135^{***}$ & $-0.134^{***}$ \\
+                                            &                &                &                &                &                &                & $(0.012)$      & $(0.012)$      \\
+Universitary (Ref.= Less than universitary) &                &                &                &                &                &                &                & $-0.048$       \\
+                                            &                &                &                &                &                &                &                & $(0.046)$      \\
+Household income                            &                &                &                &                &                &                &                & $0.001^{*}$    \\
+                                            &                &                &                &                &                &                &                & $(0.000)$      \\
+\hline
+Controls                                    & No             & No             & No             & No             & No             & No             & No             & Yes            \\
+BIC                                         & $29667.072$    & $29670.297$    & $29678.484$    & $29682.436$    & $29668.980$    & $29626.960$    & $29510.881$    & $29626.312$    \\
+Num. obs.                                   & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         \\
+Num. groups: Individuals                    & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         \\
+Var: Individuals (Intercept)                & $0.185$        & $0.184$        & $0.183$        & $0.179$        & $0.174$        & $0.170$        & $0.164$        & $0.159$        \\
+Var: Residual                               & $0.497$        & $0.500$        & $0.500$        & $0.500$        & $0.499$        & $0.496$        & $0.488$        & $0.488$        \\
+\hline
+\multicolumn{9}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.
+Note: Model 8 is controlled by age, gender, political position, and subjective social status.}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 
@@
 
-~
-
-Model 8
-
-Model 9
-
-Model 10
-
-Intercept
-
-2.18***
-
-2.03***
-
-2.09***
-
-~
-
-(0.12)
-
-(0.13)
-
-(0.13)
-
-Wave
-
--0.05**
-
--0.05*
-
--0.05**
-
-~
-
-(0.02)
-
-(0.02)
-
-(0.02)
-
-Wave\^{}2
-
-0.01***
-
-0.01***
-
-0.01***
-
-~
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-Covid: receive state benefits
-
--0.09*
-
-0.11
-
-0.03
-
-~
-
-(0.04)
-
-(0.07)
-
-(0.07)
-
-Merit: Talent
-
--0.00
-
-0.05*
-
--0.01
-
-~
-
-(0.01)
-
-(0.02)
-
-(0.01)
-
-Merit: Effort
-
-0.08***
-
-0.08***
-
-0.12***
-
-~
-
-(0.01)
-
-(0.01)
-
-(0.02)
-
-Receive state benefits * Merit: Talent
-
-~
-
--0.07**
-
-~
-
-~
-
-~
-
-(0.02)
-
-~
-
-Receive state benefits * Merit: Effort
-
-~
-
-~
-
--0.05*
-
-~
-
-~
-
-~
-
-(0.02)
-
-BIC
-
-29626.31
-
-29630.35
-
-29636.59
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.16
-
-0.16
-
-0.16
-
-Var: Residual
-
-0.49
-
-0.49
-
-0.49
-
-\item *
-
-** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-Note: All the models are controlled by educational level, income
-quintile, subjective social status, age, gender, political position, and
-inequality perception
+\caption{}
+\begin{center}
+\scalebox{0.765}{
+\begin{tabular}{l c c c}
+\toprule
+ & Model 8 & Model 9 & Model 10 \\
+\midrule
+Intercept                              & $2.18^{***}$ & $2.03^{***}$ & $2.09^{***}$ \\
+                                       & $(0.12)$     & $(0.13)$     & $(0.13)$     \\
+Wave                                   & $-0.05^{**}$ & $-0.05^{*}$  & $-0.05^{**}$ \\
+                                       & $(0.02)$     & $(0.02)$     & $(0.02)$     \\
+Wave$^2$                               & $0.01^{***}$ & $0.01^{***}$ & $0.01^{***}$ \\
+                                       & $(0.00)$     & $(0.00)$     & $(0.00)$     \\
+Covid: receive state benefits          & $-0.09^{*}$  & $0.11$       & $0.03$       \\
+                                       & $(0.04)$     & $(0.07)$     & $(0.07)$     \\
+Merit: Talent                          & $-0.00$      & $0.05^{*}$   & $-0.01$      \\
+                                       & $(0.01)$     & $(0.02)$     & $(0.01)$     \\
+Merit: Effort                          & $0.08^{***}$ & $0.08^{***}$ & $0.12^{***}$ \\
+                                       & $(0.01)$     & $(0.01)$     & $(0.02)$     \\
+Receive state benefits * Merit: Talent &              & $-0.07^{**}$ &              \\
+                                       &              & $(0.02)$     &              \\
+Receive state benefits * Merit: Effort &              &              & $-0.05^{*}$  \\
+                                       &              &              & $(0.02)$     \\
+\midrule
+BIC                                    & $29626.31$   & $29630.35$   & $29636.59$   \\
+Num. obs.                              & $7576$       & $7576$       & $7576$       \\
+Num. groups: Individuals               & $1322$       & $1322$       & $1322$       \\
+Var: Individuals (Intercept)           & $0.16$       & $0.16$       & $0.16$       \\
+Var: Residual                          & $0.49$       & $0.49$       & $0.49$       \\
+\bottomrule
+\multicolumn{4}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05. 
+Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 
@@
 
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Intercept
-
-1.847***
-
-1.853***
-
-1.873***
-
-1.741***
-
-1.664***
-
-2.276***
-
-2.277***
-
-~
-
-(0.026)
-
-(0.038)
-
-(0.090)
-
-(0.092)
-
-(0.092)
-
-(0.105)
-
-(0.138)
-
-Wave (Ref.= Wave 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.059*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.030
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.204***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.030)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.233***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.073***
-
--0.073***
-
--0.076***
-
--0.076***
-
--0.052**
-
--0.052**
-
-~
-
-~
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-Wave\^{}2
-
-~
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.013***
-
-0.013***
-
-~
-
-~
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-Covid: receive state benefits
-
-~
-
-~
-
--0.097*
-
--0.100*
-
--0.102*
-
--0.102**
-
--0.093*
-
-~
-
-~
-
-~
-
-(0.041)
-
-(0.040)
-
-(0.040)
-
-(0.039)
-
-(0.041)
-
-Covid: economy more important than health
-
-~
-
-~
-
-0.065***
-
-0.060**
-
-0.055**
-
-0.049**
-
-0.038*
-
-~
-
-~
-
-~
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-0.055***
-
--0.002
-
--0.003
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-0.096***
-
-0.084***
-
-0.083***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.013)
-
-(0.013)
-
-(0.013)
-
-Inequality perception
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.135***
-
--0.134***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-Universitary (Ref.= Less than universitary)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.048
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.046)
-
-Household Income
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.001*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.000)
-
-Subjective Social Status
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-Age (Ref.= 18-29)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Age 30-49
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.033
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.051)
-
-~~~~~Age 50-64
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.009
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.053)
-
-~~~~~Age 65 or more
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.123
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.066)
-
-Gender (Ref. Male)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.082*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.037)
-
-Pol. pos (Ref.= Left)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Center
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.095
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.051)
-
-~~~~~Right
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.220***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.059)
-
-~~~~~Does not identify
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.085
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.047)
-
-BIC
-
-29667.072
-
-29670.297
-
-29682.436
-
-29668.980
-
-29626.960
-
-29510.881
-
-29626.312
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.185
-
-0.184
-
-0.179
-
-0.174
-
-0.170
-
-0.164
-
-0.159
-
-Var: Residual
-
-0.497
-
-0.500
-
-0.500
-
-0.499
-
-0.496
-
-0.488
-
-0.488
-
-*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
+\caption{}
+\begin{center}
+\scalebox{0.765}{
+\begin{tabular}{l c c c c c c c}
+\hline
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
+\hline
+Intercept                                   & $1.847^{***}$  & $1.853^{***}$  & $1.873^{***}$  & $1.741^{***}$  & $1.664^{***}$  & $2.276^{***}$  & $2.277^{***}$  \\
+                                            & $(0.026)$      & $(0.038)$      & $(0.090)$      & $(0.092)$      & $(0.092)$      & $(0.105)$      & $(0.138)$      \\
+Wave (Ref.= Wave 2016)                      &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                \\
+\quad Wave 2017                             & $-0.167^{***}$ &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+\quad Wave 2018                             & $-0.059^{*}$   &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+\quad Wave 2019                             & $-0.030$       &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+\quad Wave 2022                             & $0.204^{***}$  &                &                &                &                &                &                \\
+                                            & $(0.030)$      &                &                &                &                &                &                \\
+\quad Wave 2023                             & $0.233^{***}$  &                &                &                &                &                &                \\
+                                            & $(0.029)$      &                &                &                &                &                &                \\
+Wave                                        &                & $-0.073^{***}$ & $-0.073^{***}$ & $-0.076^{***}$ & $-0.076^{***}$ & $-0.052^{**}$  & $-0.052^{**}$  \\
+                                            &                & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      \\
+Wave$^2$                                    &                & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.016^{***}$  & $0.013^{***}$  & $0.013^{***}$  \\
+                                            &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
+Covid: receive state benefits               &                &                & $-0.097^{*}$   & $-0.100^{*}$   & $-0.102^{*}$   & $-0.102^{**}$  & $-0.093^{*}$   \\
+                                            &                &                & $(0.041)$      & $(0.040)$      & $(0.040)$      & $(0.039)$      & $(0.041)$      \\
+Covid: economy more important than health   &                &                & $0.065^{***}$  & $0.060^{**}$   & $0.055^{**}$   & $0.049^{**}$   & $0.038^{*}$    \\
+                                            &                &                & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      & $(0.019)$      \\
+Merit: Talent                               &                &                &                & $0.055^{***}$  & $-0.002$       & $-0.003$       & $-0.005$       \\
+                                            &                &                &                & $(0.010)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Effort                               &                &                &                &                & $0.096^{***}$  & $0.084^{***}$  & $0.083^{***}$  \\
+                                            &                &                &                &                & $(0.013)$      & $(0.013)$      & $(0.013)$      \\
+Inequality perception                       &                &                &                &                &                & $-0.135^{***}$ & $-0.134^{***}$ \\
+                                            &                &                &                &                &                & $(0.012)$      & $(0.012)$      \\
+Universitary (Ref.= Less than universitary) &                &                &                &                &                &                & $-0.048$       \\
+                                            &                &                &                &                &                &                & $(0.046)$      \\
+Household Income                            &                &                &                &                &                &                & $0.001^{*}$    \\
+                                            &                &                &                &                &                &                & $(0.000)$      \\
+Subjective Social Status                    &                &                &                &                &                &                & $-0.005$       \\
+                                            &                &                &                &                &                &                & $(0.012)$      \\
+Age (Ref.= 18-29)                           &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                \\
+\quad Age 30-49                             &                &                &                &                &                &                & $-0.033$       \\
+                                            &                &                &                &                &                &                & $(0.051)$      \\
+\quad Age 50-64                             &                &                &                &                &                &                & $0.009$        \\
+                                            &                &                &                &                &                &                & $(0.053)$      \\
+\quad Age 65 or more                        &                &                &                &                &                &                & $0.123$        \\
+                                            &                &                &                &                &                &                & $(0.066)$      \\
+Gender (Ref. Male)                          &                &                &                &                &                &                & $-0.082^{*}$   \\
+                                            &                &                &                &                &                &                & $(0.037)$      \\
+Pol. pos (Ref.= Left)                       &                &                &                &                &                &                &                \\
+                                            &                &                &                &                &                &                &                \\
+\quad Center                                &                &                &                &                &                &                & $0.095$        \\
+                                            &                &                &                &                &                &                & $(0.051)$      \\
+\quad Right                                 &                &                &                &                &                &                & $0.220^{***}$  \\
+                                            &                &                &                &                &                &                & $(0.059)$      \\
+\quad Does not identify                     &                &                &                &                &                &                & $0.085$        \\
+                                            &                &                &                &                &                &                & $(0.047)$      \\
+\hline
+BIC                                         & $29667.072$    & $29670.297$    & $29682.436$    & $29668.980$    & $29626.960$    & $29510.881$    & $29626.312$    \\
+Num. obs.                                   & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         & $7576$         \\
+Num. groups: Individuals                    & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         & $1322$         \\
+Var: Individuals (Intercept)                & $0.185$        & $0.184$        & $0.179$        & $0.174$        & $0.170$        & $0.164$        & $0.159$        \\
+Var: Residual                               & $0.497$        & $0.500$        & $0.500$        & $0.499$        & $0.496$        & $0.488$        & $0.488$        \\
+\hline
+\multicolumn{8}{l}{\scriptsize{*** p < 0.001; ** p < 0.01; * p < 0.05.}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 

### 2025-04-17 submit/Title-Page.tex
update submit

--- 2025-01-20 paper.tex
@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
 \documentclass[
+  12pt,
 ]{article}
-
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \usepackage{xcolor}
-\usepackage[top=30mm,left=20mm]{geometry}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\makeatletter
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\makeatother
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
+\usepackage[margin=0.78in]{geometry}
 \usepackage{graphicx}
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\usepackage[noblocks]{authblk}
-\renewcommand*{\Authsep}{, }
-\renewcommand*{\Authand}{, }
-\renewcommand*{\Authands}{, }
-\renewcommand\Affilfont{\small}
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\usepackage{times}
+\usepackage{caption}
+\captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
+\usepackage[figurename=Fig.]{caption}
+\usepackage{graphicx}
+\usepackage{float}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
 \ifLuaTeX
@@
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
-  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.},
-  pdfauthor={Juan Carlos Castillo; Julio Iturra; Kevin Carrasco},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
+  pdftitle={Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic. },
+  hidelinks,
   pdfcreator={LaTeX via pandoc}}
@@
 role of perceptions of inequality and meritocracy during the COVID
-pandemic.}
+pandemic. \footnote{Corresponding author
+  (\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}
+  -
+  \href{https://juancarloscastillo.github.io/jc-castillo/}{jc-castillo.com}
+  - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile). The authors
+  recognize the support of the National Research and Development Agency
+  (ANID) through the grant from the National Fund for Scientific and
+  Technological Development (FONDECYT) N° 1210847 and N° 1250518; from
+  the Center for Conflict and Social Cohesion Studies (COES) grant
+  Financing Fund for Research Centers in Priority Areas (FONDAP) N°
+  1523A0005; and Millenium Nucleus on Digital Inequalities and
+  Opportunities (NUDOS) NCS2022\_046}}
+\author{}
+\date{\vspace{-2.5em}}
 
-  \author{Juan Carlos Castillo}
-            \affil{%
-                  Universidad de Chile
-              }
-          \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-          \affil{%
-                  Núcleo milenio de desigualdades y oportunidades
-                  digitales (NUDOS)
-              }
-        \author{Julio Iturra}
-            \affil{%
-                  International Graduate School of Social Sciencies
-                  (BIGSSS), University of Bremen, Germany
-              }
-        \author{Kevin Carrasco}
-            \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-
-\date{}
 \begin{document}
 \maketitle
-\begin{abstract}
-Education is considered a key tool for social mobility and equality of
-opportunities. However, despite the widespread value of education,
-disparities in educational outcomes persist over time. This paper is
-guided by the following research questions: To what extent are such
-educational disparities justified in society? Is the justification of
-educational inequality affected in periods of personal and social
-vulnerability, as in the COVID pandemic? And, what are the main factors
-driving this kind of justification? The Chilean case offers an
-interesting context for this study given its high economic inequality,
-deep neoliberal policies, and commodification of social services such as
-education. This economic and cultural environment has promoted
-meritocratic ideals, whereby individual talent and effort are considered
-key to get ahead in life, disregarding opportunities linked to social
-origin. The central argument of this article is that the justification
-of inequalities weakens during periods of vulnerability and crisis (such
-as the health and economic crisis resulting from the COVID-19 pandemic).
-Furthermore, we argue that such changes could be linked to a challenge
-of meritocratic ideals. For testing the research hypotheses we estimate
-a series of longitudinal multilevel models with data from a Chilean
-longitudinal panel survey (2016 -- 2023, 6 waves, N = 2,927). Whereas we
-find support for the association between the perception of meritocracy
-and justification of inequality, the analyses that involved changes over
-time showed an increase in the justification of inequality in education.
-The discussion of these results delves into the social consequences of
-justifying inequality in a sensitive area as education, as well as the
-persistence of meritocratic ideals despite challenging events.\newline
-\textbf{Keywords}: meritocracy, social inequality, inequality
-justification, COVID-19
-\end{abstract}
 
-This document was last modified at 2025-01-19 22:27:33
+\setstretch{1.5}
+Juan Carlos Castillo, Department of Sociology, Universidad de Chile,
+Santiago de Chile. *
 
-and it was last rendered at 2025-01-19 22:27:33
+ORCID: 0000-0003-1265-7854
 
-\section{Introduction}\label{introduction}
+Julio Iturra, Bremen International Graduate School of Social Sciences,
+Bremen, Germany.
 
-Despite the widespread recognition of education as a key factor for
-social mobility and equality of opportunities, disparities in this
-domain are far from being eliminated. As the last PISA report points
-out, ``Socio-economically disadvantaged students are seven times more
-likely than advantaged students to score below Level 2 in mathematics on
-average across OECD countries'' (OECD 2023). Furthermore, empirical
-evidence from social science research has revealed an exacerbation of
-educational inequalities across various domains, such as family
-socioeconomic status, gender, ethnicity, and migratory background,
-particularly after COVID-19 (Darmody, Smyth, and Russell 2021).
-Nevertheless, and despite the consistent empirical evidence on
-educational inequalities and their consequences, research on
-distributive justice indicates remarkable individual differences in the
-justification of unequal access to education (Lee and Stacey 2023). In
-the present paper, we argue that the perception of meritocracy in
-society might play a role in the justification of economic inequalities
-in this particular domain. We examine this association in the context of
-COVID 19 in Chile, a period characterized by several challenges
-regarding the political and economic system that could have affected the
-way in which the role of individual performance and social outcomes are
-deemed.
+ORCID: 0000-0003-1993-3479
 
-The Chilean case is particularly relevant when it comes to the study of
-inequalities in education. The country is characterized by a high level
-of economic inequality associated with the privatization and
-commodification of several social policy domains, among them the
-educational system (Bellei 2013; Corvalán, Carrasco, and García-Huidobro
-2016). The neoliberal reforms implemented during and since the military
-dictatorship (1973-1989) have deepened educational policies based on
-private subsidies, achievement incentives, selection, and segregation
-(Joiko 2019). Previous research in Chile has emphasized the concept of
-an ``educational market'' (Corvalán, Carrasco, and García-Huidobro
-2016), characterized by selection, competition, and the provision of
-goods and services associated with payment capacities. Many studies have
-addressed this problem from economic, political, and sociological
-perspectives, generating relevant evidence on their academic and social
-consequences (Ramos 2022). Likewise, massive social mobilizations in the
-country (particularly in 2006, 2011, and 2018) have had educational
-inequalities as one of the main targets, putting pressure on governments
-towards a series of reforms that have advanced free higher education
-reforms and have diminished the role of private actors in the
-educational system (Donoso 2016; Reyes-Housholder and Roque 2019).
+Kevin Carrasco, Center for Social Conflict and Cohesion Studies
 
-This research endeavors to dissect the nuanced alterations in the
-justifications for educational disparities within the Chilean framework.
-Central to our investigation is the premise that the rationale
-underpinning these inequalities experiences a significant recalibration
-during periods of vulnerability and crises, specifically under the
-unprecedented strains of the global health and economic cataclysm
-triggered by the COVID-19 pandemic (Hankivsky, Morrow, and Varcoe 2022;
-Breznau 2021). A pertinent aspect of this discourse is articulated in
-the works of Mijs (2016, 2018), showing the interplay between
-meritocratic beliefs and the legitimization of disparities, as well as
-the potential erosion of meritocratic principles amidst crises. This
-line of research has highlighted how meritocratic beliefs are associated
-with justifying unequal access to resources and rewards in society, by
-legitimizing social hierarchies in which those who already have
-resources and advantages are more likely to succeed (Sandel 2020;
-McNamee and Miller 2004; Hadjar 2008). Within this framework, we point
-out that periods of vulnerability (such as the health and economic
-crisis resulting from the COVID-19 pandemic) could weaken the
-justification of social inequalities, a change that would be at least in
-part linked to a challenge of the meritocratic ideals.
+ORCID: 0000-0002-9348-8361
 
-\section{The justification of educational
-inequalities}\label{the-justification-of-educational-inequalities}
+\section{Conflict of interest}\label{conflict-of-interest}
 
-The justification of economic inequalities is a controversial field of
-study. As equality is usually considered a synonym of justice in common
-sense, the idea that someone could be willing to justify inequalities
-seems (in principle) far from what is considerably reasonable.
-Nevertheless, and as pointed out by Barry, the whole normative
-discussion in social justice can be conceived as ``the defensibility of
-unequal relations between people'' (1989, 3), whereby empirical research
-has consistently shown strong variations between individuals and
-societies regarding inequality justification (Trump 2018; Kelley and
-Evans 2021; Mijs 2021). Both theoretical and empirical approaches have
-been able to establish a clear distinction between inequality and
-justice, opening the question of to what extent inequality can be
-tolerated or even justified.
-
-Studies addressing inequality justification have assessed the support
-for market ideologies, salary gaps, and distributive preferences. A
-strand of this research has focused on specific policy areas such as
-health, pensions, and education, attending to the degree of
-justification of market-based access to social services. For instance,
-Lindh (2015) found that public support for the market distribution of
-social services is generally low in most countries, implying that most
-people believe it is unfair for market forces to determine access to
-basic social services. Similarly, Busemeyer and Iversen (2020) shows
-that high-income citizens are less supportive of expanding public
-spending on universal benefit schemes. Instead, they tend to prefer a
-public basic insurance scheme, even if it provides relatively more
-benefits to low-income individuals.
-
-In the area of survey studies, the operationalization of educational
-inequality justification in terms of questionnaire items has been
-diverse, and there is not a single instrument or scale dedicated
-specifically to this concept. Rather, some attitudinal surveys usually
-include education-related items that are posteriorly used as dependent
-or independent variables. Some examples are the International Social
-Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong --
-that people with higher incomes can buy better education for their
-children than people with lower incomes?}), and the European Social
-Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I
-have had a fair chance of achieving the level of education I was
-seeking}). Although both questions have been used to measure the
-justification of educational inequality, they refer to different
-aspects: the ESS can be classified as reflexive, as the respondent is
-involved as a referent for the question, whereas the ISSP question is
-non-reflexive. Furthermore, ISSP refers to educational access, while the
-ESS refers to educational outputs. Although such items covering
-inequality justification are present in emblematic comparative studies,
-still there are few studies focusing specifically in the justification
-of educational inequality.
-
-\emph{The COVID pandemic and public concerns about inequality}
-
-The time frame for this study covers between 2016 and 2023, by using
-longitudinal panel survey data provided by the ELSOC database (Estudio
-Longitudinal Social de Chile). Although this period comprises the COVID
-pandemic, the changes in the variables analyzed are not possible to be
-attributed only to this event in terms of a health crisis, as several
-economic, social and cultural processes overlapped worldwide. However,
-and as several studies have shown, this period appeared to trigger
-attitudinal changes that are relevant to explore, both in themselves and
-as antecedent to future studies. In this sense, in this paper we argue
-that the increase of risk perceptions in critical situations (such as
-the pandemic) impacts public concerns about equality and redistribution
-(Breznau 2021).
-
-In Chile, the COVID-19 pandemic posed a significant threat to
-economically vulnerable groups and those in precarious labor market
-positions (Ministerio de Desarrollo Social y Familia 2023; Instituto
-Nacional de Estadísticas 2022). The poverty rate, which was 11.2\% in
-2015, decreased to 8.5\% by 2017, with a 10/10 ratio of 31.6. However,
-during the pandemic in 2020, poverty rose to 10.7\%, the 10/10 ratio
-spiked to 301.7, and unemployment reached a decade-high of 10.5\%. By
-2022, poverty declined to 6.5\%, the 10/10 ratio improved to 61.2, and
-unemployment decreased to 7.9\%, with a slight rise to 8.7\% in 2023.
-
-Chile implemented sanitary controls early in the pandemic, initiating
-lockdowns on March 18, 2020, and lifting all restrictions by March 5,
-2023. In this context, Chile stands out as one of the strictest
-countries globally and within South America in terms of pandemic-related
-policies. According to the Stringency Index (Hale et al. 2021)---ranging
-from 0 to 100 (100 being the strictest) and encompassing measures like
-school closures, workplace shutdowns, and travel bans---the global
-median was 42.54, while Chile scored 58.1. This placed Chile 36th among
-184 countries and 4th in South America, following Peru (60.1), Suriname
-(60.5), and Venezuela (80.5). Besides, an important cycle of protests
-known as the ``social outburst'' took place between October 2019 and
-March 2020 (Somma et al. 2021). As a consequence there were two
-constitutional processes happening in this very same period after the
-massive social upheavals. However, the protest drastically decreased as
-a result of the lockdown (Joignant et al. 2020).
-
-In terms of public views on economic inequality, it has been documented
-that critical views toward inequality increased in 2019 and deepened
-under the pandemic (COES 2023), which we argue can be in part explained
-by the increased risk exposure and vulnerability in the context of the
-COVID-19 pandemic. Although the pandemic might have triggered such
-concerns mostly regarding health, we believe that other policy areas,
-such as education, could have been affected in this period.
-Consistently, our first hypothesis is:
-
-\(H_1\): The justification of educational inequalities decreases after
-the pandemic
-
-Besides proposing a general decrease over time in the justification of
-inequalities, a second hypothesis refers to specific aspects of the
-pandemic. As some political measures from the state included several
-social benefits (as delivering food boxes, free vaccination, direct
-transfers, among others), we argue that being recipient of such benefits
-could affect attitudes towards the redistributive role of the state as
-equalizer opportunities to face risks. As a consequence, preferences for
-educational inequalities would be reduced:
-
-\(H_2\): Recipient of state benefits show lower justification of
-educational inequalities
-
-In the same line, we expect that those who give priority to take care of
-the health of the population instead of the country economy would
-exhibit larger preferences for educational inequalities:
-
-\(H_3\): Those prioritizing health instead of the economic situation
-show higher justification of educational inequalities
-
-\emph{Meritocracy and educational inequalities}
-
-Whether inequalities in education are considered just or not lead us to
-take into account the concepts of merit and meritocracy. Meritocracy, a
-concept coined by the sociologist Michael Young in his satirical work
-``The Rise of the Meritocracy'' (1962), has a rich history and
-multifaceted meaning. Young's term was initially used to criticize a
-hypothetical future society where social and economic positions were
-entirely determined by individuals' merits, talents, and efforts,
-ostensibly eliminating the influence of inherited privilege. However,
-over time, the term has evolved to describe a social system where
-advancement and success are perceived as directly linked to one's
-abilities and achievements, sidelining inequalities of opportunities
-(Bell 2020). This would contribute to the justification of inequality in
-the educational domain (Sandel 2020), as individual ability is often
-considered inherited, and educational attainment is influenced by family
-background (Goldin 2000).
-
-Although there are few studies relating meritocratic beliefs and
-inequality justification, from a broader perspective it is possible to
-find research which incorporates the support of different distributive
-principles with regard to education. For instance, in an experiment by
-Igliozzi, Granot, and Ottati (2024), they establish how the support for
-different justice principles (equity, equality, and need) is related to
-support for redistributive policies in domains such as education. They
-find that a larger system justification (Jost and Hunyady 2003) is
-linked to higher support for the principles of equity and equality (over
-need) in distributing educational outcomes, where equity was measured in
-terms of access to education related to performance (`Students who place
-within the highest 50\% of testing scores from the previous year can
-attend the magnet school'). In the same line, Lee and Stacey (2023) show
-that performance assessments (in what they call ``neoliberal
-orientations'') predicted perceived fairness of educational inequality.
-
-More specific empirical research examining the relationship between
-meritocratic perceptions and the justification of educational inequality
-provides valuable insights into the dynamics shaping individuals'
-attitudes and beliefs. Some studies have explored the extent to which
-people believe that educational systems are meritocratic and how these
-perceptions influence their acceptance or rejection of educational
-disparities, as well as status differences (McNamee and Miller 2004).
-This belief is particularly influential in educational institutions,
-where it not only encourages academic success but also legitimizes
-social and income inequality (Batruch et al. 2023). Most of the evidence
-so far suggests that those who believe in a direct link between effort,
-talent, and educational success are more inclined to rationalize
-existing disparities in educational outcomes. However, empirical
-findings also reveal nuances in this relationship. For instance, studies
-show that when individuals perceive structural barriers, such as
-economic disparities or discrimination, as hindering the meritocratic
-ideal, they may be less likely to justify educational inequalities (Day
-and Norton 2023). This suggests that while meritocratic beliefs play a
-significant role, individuals' awareness of broader societal issues can
-mediate their willingness to rationalize educational disparities.
-
-Empirical cross-sectional evidence from Chile reveals two significant
-trends: low public support for the perception of meritocracy and limited
-justification for unequal access to social services such as education
-and health. For instance, Castillo, Madero-Cabib, and Salamovich (2013)
-analyzed data from the International Social Survey Programme (ISSP) in
-1999 and 2009, highlighting consistently low support for inequality
-based on resource allocation or individual merit. Similarly, Castillo,
-Miranda, and Carrasco (2012) demonstrated widespread skepticism among
-Chileans regarding the meritocratic system, with many disagreeing with
-the notion that rewards are fairly distributed based on talent and
-effort. In line with this evidence, we propose that:
-
-\(H_4\): Higher meritocratic perceptions are positively related to the
-justification of educational inequalities.
-
-Combining the arguments regarding the effect of the pandemic and its
-relationship with meritocracy, we explore interactions between
-meritocracy and inequality justification over time. As COVID-19 pandemic
-has affected the structural opportunities to attend schools,
-universities, or workplaces, the ideal of meritocracy might have been
-challenged in the time frame analyzed. As a consequence, the
-justification of inequality could show a decrease over time:
-
-\(H_5\): The positive association between meritocracy and inequality
-justification mitigates along time
-
-Along with meritocratic perceptions, in this study we incorporate
-inequality perception. As pointed out by Janmaat (2013), justifications
-are usually confused under the labels of ``beliefs'' and ``attitudes,''
-and their distinction from perceptions is relevant as it helps state
-that justifying a certain level of inequality requires taking into
-account which level of (perceived) inequality is first and foremost
-being recognized (Son Hing et al. 2011; Castillo et al. 2023). For
-instance, a recent study by Day and Norton (2023) on the perception and
-justification of inequality in university endowments in the U.S. points
-out that individuals tend to underestimate the magnitude of existing
-inequality while simultaneously desiring greater equality. Furthermore,
-through an experiment, they demonstrate that information about
-inequality in university endowments (i.e., manipulating inequality
-perception) increases the perception of injustice in education. Results
-of this study are in line with the consistent evidence on biases related
-to the underestimation of economic inequality (Castillo, García-Castro,
-and Venegas 2022; Gimpelson and Treisman 2018). As a larger inequality
-in this domain threatens a basic equality ideal, the corresponding
-hypothesis in this case is:
-
-\(H_6\): A larger inequality perception is negatively associated with
-the justification of educational inequalities.
-
-Next, we take into account the evidence that relates rational interests
-to justifications and preferences (Son Hing et al. 2019). Research about
-the justification of educational inequality usually focuses on
-differences between status groups, based on the rational premise that
-those who are worst-off would tend to show lower levels of justification
-(Busemeyer and Iversen 2020). For instance, Valant and Newark (2016)
-find that Americans are more concerned about --- and more supportive of
-proposals to close --- wealth-based achievement gaps than Black-White or
-Hispanic-White gaps. In line with this evidence, we expect that:
-
-\(H_7\): Those with lower status characteristics show lower
-justification of educational inequalities
-
-\section{Data, variables \& methods}\label{data-variables-methods}
-
-\subsection{Data}\label{data}
-
-The main data source is the Chilean Longitudinal Social Survey ELSOC
-(2016--2023). ELSOC has been designed as a yearly panel study to
-evaluate how individuals think, feel, and behave regarding a set of
-social issues concerning conflict and social cohesion in Chile. The
-sampling design is probabilistic, stratified, clustered, and multistage.
-It provides adequate coverage of the country's largest cities
-(Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller
-cities. The sample covers participants aged between 18 and 75 years (in
-wave 1), representing 77\% of the total national population and 93\% of
-the urban areas (ELSOC 2022).
-
-The survey has been conducted yearly since 2016, except for the year
-2020, when it was suspended due to the pandemic. The waves 2016, 2017,
-2018, 2019, 2022, and 2023 were administered using computer-assisted
-personal interviewing (CAPI). However, a reduced version was conducted
-using computer-assisted telephone interviews (CATI) in 2021. In
-addition, wave 3 included a refreshment sample to counter survey
-attrition. As a result, the total sample of wave 3 included 3,748 cases,
-of which 2,229 were part of the original sample and 1,519 were from the
-refreshment sample. The data from the refreshment sample is not included
-in the analytical sample for this study as we analyze longer response
-trends. Regarding the original sample, the response rate was 62.4\% in
-wave 1, achieving N = 2,927 participants. In broader terms, the
-accumulated attrition between wave 1 and wave 7 is 40\%, achieving a
-final sample of N = 1,741. The analysis used longitudinal weights to
-avoid bias due to the sampling design, allowing us to control for biases
-arising from systematic patterns of non-participation in the survey
-after the first wave. For a more detailed analysis of responses,
-attrition, and the construction of longitudinal weights, visit
-https://coes.cl/encuesta-panel/. The dataset is publicly available at:
-https://dataverse.harvard.edu/dataverse/elsoc.
-
-\subsection{Variables}\label{variables}
-
-The dependent variable of this study is the justification of educational
-inequality. This construct is measured using the following statement:
-``It is just that high-income people have a better education for their
-children than people with lower incomes'' (``Es justo que las personas
-de altos ingresos tengan una mejor educación para sus hijos que las
-personas con ingresos más bajos'' in Spanish). Here, the respondents
-declared their preferences on a Likert scale from (1) ``strongly
-disagree'' to (5) ``strongly agree'' (5). The main independent variables
-refer to meritocratic perception, which is operationalized by two items,
-one related to effort (``In Chile, people are rewarded for their
-efforts''), and the other related to talent (``In Chile, people are
-rewarded for their intelligence and skills''). In addition, we included
-an indicator for economic inequality perception (``In Chile, the income
-differences are too large''). Figure~\ref{fig-frecuencies} shows the
-average distribution throughout the years for these variables. We
-observe that there is a large group (88.2\%) of respondents who disagree
-or strongly disagree that people with higher incomes can access better
-education. Likewise, a large group of respondents (91.7\%) agree or
-strongly agree that income differences in Chile are too large. On the
-other hand, there is an important level of disagreement (disagree +
-strongly disagree) about the perception of meritocracy, this is, whether
-people are rewarded for their talent (47.6\%) and for their efforts
-(57\%).
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/merit_plot.png}
-
-}
-
-\caption{\label{fig-frecuencies}Frequency of responses for justification
-of educational inequality, perception of inequality, and perception of
-meritocracy for all the years analyzed}
-
-\end{figure}%
-
-A third group of variables is included in the analysis to assess the
-effect of elements related to the pandemic context. These variables were
-only included in the 2021 wave, so each individual's responses were
-replicated for the remaining years. The first variable addresses the
-belief regarding whether, in the context of the pandemic, it is more
-important to give priority to economic activity than to the health of
-the population, measured on a Likert scale from (1) ``strongly
-disagree'' to (2) ``strongly agree''. The second variable is a dummy
-(No/Yes) that captures whether individuals received state benefits
-during the pandemic (e.g., Food Box, Emergency Family Income,
-Contribution for the Middle Class, or others).
-
-For testing the hypothesis regarding status we considered educational
-level and household income. Also, we included subjective social status
-as it has been argued that perceived social status is a relevant
-predictor of attitudes toward economic inequality (Schneider and
-Castillo 2015; Castillo et al. 2019). In addition, such subjective
-measures can complement objective measures in predicting life chances
-(Oesch and Vigna 2023), which are arguably connected to how individuals
-experience economic inequality. In this regard, it has been argued that
-how people form their views on their standing in society results from
-experiences with direct socioeconomic circumstances and social
-comparison processes with reference groups (Condon and Wichowsky 2020).
-Finally, political identification on a left-right scale, age groups, and
-gender were included as control variables. Table~\ref{tbl-descriptives}
-shows the whole set of independent variables, the corresponding items,
-response categories, and frequencies.
-
-\begin{longtable}[]{@{}l@{}}
-\caption{Independent variables}\label{tbl-descriptives}\tabularnewline
-\toprule\noalign{}
-\endfirsthead
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-\includegraphics{output/tables/desc02.png} \\
-\end{longtable}
-
-\subsection{Methods}\label{methods}
-
-Given the hierarchical structure of the data (observations nested in
-survey waves), we estimated a series of longitudinal multilevel models
-(Singer and Willett 2009). Such an approach is suited to account for the
-shared variance among units (in this case individuals), adjusting the
-estimation of the standard errors. The linear multilevel models are
-estimated using the R library ``lme4'' (Singer and Willett 2009, 4).
-
-\section{Results}\label{results}
-
-Beginning with a descriptive analysis of changes in the justification of
-educational inequality between 2016 and 2023, Figure~\ref{fig-alluvial}
-illustrates yearly frequencies and transitions between years. Each year
-represents stacked porcentual frequencies, whereas the flows in between
-reflect the within-subject change of opinions from one year to the next
-(Rosvall and Bergstrom 2010). For instance, of the 32.9\% who strongly
-disagreed with inequality justification in 2016, about half of them kept
-responding the same in 2017, whereas the other half shifted their
-opinion to other response categories. In general, the large majority -
-between 80 and 90\% - disagrees with the justification of educational
-inequality throughout the years. Despite this overall tendency, we also
-observe that the disagreement with inequality justification (disagree +
-strongly disagree) tends to go down in the last waves. This change is
-mostly a result of the increase in the ``agree'' category, which doubles
-compared to previous years (from 5.4\% in 2016 to 11\% in 2023).
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/alluvial_dep.png}
-
-}
-
-\caption{\label{fig-alluvial}Change in the justification of educational
-inequality over time (2016-2023)}
-
-\end{figure}%
-
-Figure~\ref{fig-means} shows the average changes in the main variables
-considered for this study. Here we observe that the justification of
-educational inequality has the lowest average throughout the years when
-compared with the other (independent) variables, whereas the highest
-average is consistently represented by inequality perception.
-Interestingly, in the last waves of the study (2022-2023) the
-justification of inequality increases whereas the perception of
-inequality decreases. As the merit variables are concerned, they show a
-very similar pattern in terms of averages and changes over the years,
-being the perception of meritocracy related to effort always lower than
-the one associated with talent.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/years_plot.png}
-
-}
-
-\caption{\label{fig-means}Change in the mean of justification of
-educational inequality, perception of inequality, and perception of
-meritocracy over the years}
-
-\end{figure}%
-
-Figure~\ref{fig-correlation} presents a correlation matrix of the main
-variables analyzed, using data from the last wave (2023). In this
-matrix, the correlations vary between low and moderate values. The
-correlations between justification of educational inequality with the
-two COVID variables are positive, but low. In addition, justification of
-education inequality depicts a moderate and negative association with
-the perception of inequality (\emph{r}=-0.28, p\textless.01) and a
-moderate and positive association with both meritocracy perception
-variables (\emph{r}=0.17, p\textless.01; \emph{r}=0.17, p\textless.01).
-Regarding the perception of inequality, it presents a moderate and
-negative association with both meritocracy perception variables
-(\emph{r}=-0.19, p\textless.01; \emph{r}=-0.14, p\textless.01). Finally,
-the two meritocracy variables present a high and positive association
-with each other (\emph{r}=0.79, p\textless.01).
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.85\textwidth,height=\textheight]{output/graphs/corr.png}
-
-}
-
-\caption{\label{fig-correlation}Correlation matrix of the main variables
-for the first wave (2016)}
-
-\end{figure}%
-
-\emph{Multilevel models}
-
-Table~\ref{tbl-multilevel} shows the multilevel estimation results for
-the justification of educational inequality. Model 1 includes the survey
-waves to estimate intertemporal changes in the dependent variable.
-Taking 2016 as a reference point, we can observe a staggered decrease in
-2017 (\(\beta\)=-0.167, p\textless.001), 2018 (\(\beta\)=-0.059,
-p\textless.05), and 2019 (\(\beta\)=-0.03, p\textgreater.05).
-Nevertheless, in the last waves of 2022 and 2023 there is a clear
-increase in level of justification of educational inequality
-(\(\beta\)=0.204, p\textless.001 and \(\beta\)=0.233, p\textless.001),
-suggesting a non-linear change in this variable. Attempting to model
-this path of change over time, Model 2 incorporates time (survey waves)
-as a continuous variable as well as its quadratic term, representing the
-nonlinear association initially observed in Model 1. On the one hand,
-the survey wave depicts a negative association, expressing an average
-decrease in inequality justification over time. On the other hand, the
-quadratic wave term is positive, indicating the reversion of this path
-in the last measurement points.
-
-Models 3 and 4 adds the two variables about the pandemic context, where
-we can see that those who received state benefits during this period
-justify less inequality in education compared to those who did not
-receive them. Although the effect size is small, it remains significant
-when controlling for the rest of the variables in model 8
-(\(\beta\)=-0.093, p\textless.01). On the other hand, those who believe
-that in this context the economy is more important than the health of
-the population justify more inequality in education(\(\beta\)=0.038,
-p\textless.05, Model 8).
-
-Models 5 and 6 introduce the meritocratic variables: talent (if
-intelligence and abilities are rewarded in society) and effort (if
-efforts are rewarded in society). In line with our hypotheses, the
-perception that talent is rewarded has a positive influence on the
-justification of educational inequality in Model 5 (\(\beta\)=0.055,
-p\textless.001). However, when controlling for the perception that
-effort is rewarded, this effect is no longer significant. In this sense,
-Model 6 shows that the perception that effort is rewarded in society is
-not only positively associated with the justification of educational
-inequality (\(\beta\)=0.083, p\textless.001), but it has a larger weight
-than the perception of talent. Regarding inequality perception (added in
-Model 7), it depicts a negative association with the justification of
-educational inequality, remaining stable when controlling for the rest
-of the variables.
-
-Model 8 adds the socioeconomic variables to test hypothesis 7. Although
-educational level and household per capita income are not significant in
-this model, controlling for all variables in the next models, household
-per capita income depicts a positive effect (\(\beta\)=0.001,
-p\textless.05). The inclusion of other controls as being right-wing
-(compared to being left-wing) shows a positive association with
-justification of educational inequality (\(\beta\)=0.27,
-p\textless.001), whereas being a women (compared to men) had a negative
-effect (\(\beta\)=-0.08, p\textless.001). Subjective social status and
-age had no significant effect when controlling for all the variables
-analyzed. The full model specification can be seen in the appendix (see
-Table~\ref{tbl-multilevel-full}).
-
-\begin{table}
-
-\caption{\label{tbl-multilevel}Multilevel longitudinal models for the
-justification of inequality in education}
-
-\centering{
-
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Model 8
-
-Intercept
-
-1.847***
-
-1.853***
-
-1.920***
-
-1.776***
-
-1.641***
-
-1.563***
-
-2.174***
-
-2.183***
-
-~
-
-(0.026)
-
-(0.038)
-
-(0.048)
-
-(0.064)
-
-(0.068)
-
-(0.069)
-
-(0.086)
-
-(0.120)
-
-Wave (Ref.= Wave 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.059*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.030
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.204***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.030)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.233***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.073***
-
--0.073***
-
--0.073***
-
--0.076***
-
--0.076***
-
--0.052**
-
--0.052**
-
-~
-
-~
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-Wave\^{}2
-
-~
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.013***
-
-0.013***
-
-~
-
-~
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-Covid: receive state benefits
-
-~
-
-~
-
--0.095*
-
--0.097*
-
--0.100*
-
--0.102*
-
--0.102**
-
--0.093*
-
-~
-
-~
-
-~
-
-(0.041)
-
-(0.041)
-
-(0.040)
-
-(0.040)
-
-(0.039)
-
-(0.041)
-
-Covid: economy more important than health
-
-~
-
-~
-
-~
-
-0.065***
-
-0.060**
-
-0.055**
-
-0.049**
-
-0.038*
-
-~
-
-~
-
-~
-
-~
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-~
-
-0.055***
-
--0.002
-
--0.003
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.096***
-
-0.084***
-
-0.083***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.013)
-
-(0.013)
-
-(0.013)
-
-Inequality perception
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.135***
-
--0.134***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-Universitary (Ref.= Less than universitary)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.048
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.046)
-
-Household income
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.001*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.000)
-
-Controls
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-Yes
-
-BIC
-
-29667.072
-
-29670.297
-
-29678.484
-
-29682.436
-
-29668.980
-
-29626.960
-
-29510.881
-
-29626.312
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.185
-
-0.184
-
-0.183
-
-0.179
-
-0.174
-
-0.170
-
-0.164
-
-0.159
-
-Var: Residual
-
-0.497
-
-0.500
-
-0.500
-
-0.500
-
-0.499
-
-0.496
-
-0.488
-
-0.488
-
-*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-Note: Model 8 is controlled by age, gender, political position, and
-subjective social status.
-
-}
-
-\end{table}%
-
-In this last part of the analysis we test hypotheses about the effect of
-COVID-19 on changes in the relationship between meritocracy and
-justification of educational inequality. In hypothesis 5, we proposed
-that the association between the perception of meritocracy and
-justification of inequality is mitigated in times of crisis, as
-meritocratic ideals might have been weakened due to critical situations
-associated with the COVID-19 pandemic. We test this hypothesis through
-interaction effects, shown in Table~\ref{tbl-interact}. The first model
-is shown as the baseline model. It is the same as model 8 in
-Table~\ref{tbl-multilevel}, but only shows the variables involved in the
-interaction for the sake of space (all other variables are included).
-Model 9 adds the interaction between the perception of talent-based
-meritocracy and whether state benefits were received during the
-pandemic, whereas Model 10 considers the perception of effort-related
-meritocracy for the interaction. Both variables show a negative
-significant interaction with receiving state benefits
-(\(\beta_{talent}\)=-0.07, p\textless.01; \(\beta_{effort}\)=-0.05,
-p\textless.05), meaning that for those who received state benefits the
-association between the the justification of inequality and the
-perception of effort-based meritocracy lower by 0.05 points, and by 0.07
-points regarding the perception of talent-based meritocracy. This last
-interaction is depicted in Figure~\ref{fig-interact-talent}, showing
-that the positive association between the perception of talent-based
-meritocracy and the justification of educational inequality is smaller
-for those who received state benefits during the pandemic.
-
-\begin{table}
-
-\caption{\label{tbl-interact}Interaction effects for the justification
-of economic inequality}
-
-\centering{
-
-~
-
-Model 8
-
-Model 9
-
-Model 10
-
-Intercept
-
-2.18***
-
-2.03***
-
-2.09***
-
-~
-
-(0.12)
-
-(0.13)
-
-(0.13)
-
-Wave
-
--0.05**
-
--0.05*
-
--0.05**
-
-~
-
-(0.02)
-
-(0.02)
-
-(0.02)
-
-Wave\^{}2
-
-0.01***
-
-0.01***
-
-0.01***
-
-~
-
-(0.00)
-
-(0.00)
-
-(0.00)
-
-Covid: receive state benefits
-
--0.09*
-
-0.11
-
-0.03
-
-~
-
-(0.04)
-
-(0.07)
-
-(0.07)
-
-Merit: Talent
-
--0.00
-
-0.05*
-
--0.01
-
-~
-
-(0.01)
-
-(0.02)
-
-(0.01)
-
-Merit: Effort
-
-0.08***
-
-0.08***
-
-0.12***
-
-~
-
-(0.01)
-
-(0.01)
-
-(0.02)
-
-Receive state benefits * Merit: Talent
-
-~
-
--0.07**
-
-~
-
-~
-
-~
-
-(0.02)
-
-~
-
-Receive state benefits * Merit: Effort
-
-~
-
-~
-
--0.05*
-
-~
-
-~
-
-~
-
-(0.02)
-
-BIC
-
-29626.31
-
-29630.35
-
-29636.59
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.16
-
-0.16
-
-0.16
-
-Var: Residual
-
-0.49
-
-0.49
-
-0.49
-
-\item *
-
-** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-Note: All the models are controlled by educational level, income
-quintile, subjective social status, age, gender, political position, and
-inequality perception
-
-}
-
-\end{table}%
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.85\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-talent-1.pdf}
-
-}
-
-\caption{\label{fig-interact-talent}Interaction effect of merit based on
-talent and receive state benefits on justification of educational
-inequality}
-
-\end{figure}%
-
-\section{Discussion and conclusions}\label{discussion-and-conclusions}
-
-Our research sought to examine the justification of inequality in
-education from a longitudinal perspective amid the health crisis
-generated by the COVID-19 pandemic. In general, our results showed mixed
-evidence about our hypotheses. We argued that in times of crisis,
-greater exposure to risk due to the health context would result in a
-lower justification of inequality by citizens. In this regard,
-longitudinal models showed that the justification of inequality in
-education is far from being linear, and whereas a decreasing pattern was
-found between 2017 and 2019, there is a striking increase in the
-post-pandemic measurement points. Besides, our findings revealed that
-those who received benefits justified educational inequality less than
-those who did not. Still, given our data limitations it is not possible
-to attribute such changes only to the pandemic and the related economic
-and social crises. As during this very same period in Chile there were
-two (failed) constitutional processes, disentangling different forces
-driving public opinion becomes cumbersome and deserves further
-qualitative and quantitative studies. Along this line, some recent
-studies have argued that the election of a far left-wing constitutional
-assembly in the first process generated a backslash effect, mainly due
-to several scandals that led to the delegitimation of this assembly.
-This could have driven preferences in a conservative direction, which
-was reflected in the election of a right-wing second assembly after the
-failed first constitutional process (Palanza and Sotomayor 2023; Sazo
-2023).
-
-The main focus of this study was on the relationship between meritocracy
-and justification of inequality, arguing that those who perceive that
-the society in which they live complies with meritocratic principles,
-would be more inclined to justify inequality in education. At first
-look, the results are consistent with the hypothesis; however, some
-nuances are worth attending as the effort dimension prevails over the
-talent dimension. In other words, the perception of meritocracy in terms
-of rewarded effort is more relevant in justifying access to education
-than the perception of talent. This could be explained as that talent
-could be more associated with luck in terms of random assignment, and in
-this terms would not be enough reason (as it is effort) to justify
-educational inequality. Furthermore, we observed that the connection
-between perceived meritocracy and the justification of educational
-inequality weakened for those who received state benefits, which aligns
-with our expectations given the social context in the post-COVID era.
-One possible interpretation is that, during times of increased risk,
-people may become more skeptical of narratives emphasizing individual
-effort over collective efforts. This shift in perspective might also
-foster stronger solidarity to protect individual and societal wellbeing.
-However, further studies are necessary to explore this phenomenon using
-more specific conceptual and methodological approaches.
-
-Regarding the status position, our central hypothesis was that
-individuals in more advantaged situations and with greater resources
-would tend to defend their interests and justify greater education
-inequality. In this regard, our results show a positive relationship
-between income and the justification of inequality, which would align
-with our argument that rational interests of the better-off that would
-lead to a larger inequality justification. Besides status, we also
-considered inequality perception, as the literature on attitudes toward
-inequality has argued that a greater justification of inequality is
-associated with its perceived magnitude. Consistent with this
-hypothesis, our results showed that, throughout the years, a larger
-perceived economic inequality motivates a lower justification of
-inequality in education.
-
-Among the limitations of our study, we can mention at least three.
-First, the operationalization of justification of educational inequality
-is rather limited. We counted only on a single item for this in this
-study, and further work is needed in order to assess the validity and
-reliability of measuring such construct. Second, as it has been
-established by recent research (Castillo et al. 2023), the perception of
-meritocracy can be understood through individual attributions regarding
-effort and talent, but also taking into account the role of structural
-aspects such as family status and social capital. Due to data
-limitations such dimensions were not included here, therefore providing
-a partial perspective on the areas of perceived meritocracy. Finally,
-there are limitations involving the longitudinal dimension of the data
-used. Given constraints for data collection during the pandemic, the
-ELSOC study had to shorten the questionnaire in 2021 and unfortunately,
-the educational justice indicator was excluded from that survey.
-Consequently, the results of temporal change have to be considered
-carefully, and further data waves could give us more information on this
-matter.
-
-Regarding future studies, the capabilities of the ELSOC longitudinal
-database are not limited to micro-level estimates. In this regard,
-thanks to the sampling strategy of the survey, it would be possible to
-make contextual estimates at the municipality level. For instance,
-statistical models could include contextual information from
-administrative data sources, allowing testing for hypotheses that
-include contextual socioeconomic inequality as well as its dynamics over
-time. Furthermore, this dataset could be used for comparing inequality
-justification in policy areas besides education, such as health and
-pensions. Future studies may shed light on how meritocratic perceptions
-and beliefs affect differently such areas and their changes over time,
-which would give relevant hints from social sciences for the discussion
-about societal and cultural changes as well as their impacts on
-solidarity and social cohesion.
-
-\section{References}\label{references}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-barry_theories_1989}
-Barry, Brian. 1989. \emph{Theories of Justice}. A Treatise on Social
-Justice 1. Berkeley: Univ. of California Pr.
-
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
-Batruch, Anatolia, Jolanda Jetten, Herman Van De Werfhorst, Céline
-Darnon, and Fabrizio Butera. 2023. {``Belief in {School Meritocracy} and
-the {Legitimization} of {Social} and {Income Inequality}.''}
-\emph{Social Psychological and Personality Science} 14 (5): 621--35.
-\url{https://doi.org/10.1177/19485506221111017}.
-
-\bibitem[\citeproctext]{ref-bell_politics_2020}
-Bell, Elizabeth. 2020. {``The {Politics} of {Designing Tuition-Free
-College}: {How Socially Constructed Target Populations Influence Policy
-Support}.''} \emph{The Journal of Higher Education} 91 (6): 888--926.
-\url{https://doi.org/10.1080/00221546.2019.1706015}.
-
-\bibitem[\citeproctext]{ref-bellei_estudio_2013}
-Bellei, Cristián. 2013. {``El Estudio de La Segregaci{ó}n
-Socioecon{ó}mica y Acad{é}mica de La Educaci{ó}n Chilena.''}
-\emph{Estudios Pedag{ó}gicos (Valdivia)} 39 (1): 325--45.
-\url{https://doi.org/10.4067/S0718-07052013000100019}.
-
-\bibitem[\citeproctext]{ref-breznau_welfare_2021}
-Breznau, Nate. 2021. {``The Welfare State and Risk Perceptions: The
-{Novel Coronavirus Pandemic} and Public Concern in 70 Countries.''}
-\emph{European Societies} 23 (sup1): S33--46.
-\url{https://doi.org/10.1080/14616696.2020.1793215}.
-
-\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
-Busemeyer, Marius R., and Torben Iversen. 2020. {``The {Welfare State}
-with {Private Alternatives}: {The Transformation} of {Popular Support}
-for {Social Insurance}.''} \emph{The Journal of Politics} 82 (2):
-671--86. \url{https://doi.org/10.1086/706980}.
-
-\bibitem[\citeproctext]{ref-castillo_perception_2022}
-Castillo, Juan Carlos, Juan-Diego García-Castro, and Martín Venegas.
-2022. {``Perception of Economic Inequality: Concepts, Associated Factors
-and Prospects of a Burgeoning Research Agenda.''} \emph{International
-Journal of Social Psychology} 37 (1): 180--207.
-\url{https://doi.org/10.1080/02134748.2021.2009275}.
-
-\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
-Castillo, Juan Carlos, Julio Iturra, Luis Maldonado, Jorge Atria, and
-Francisco Meneses. 2023. {``A {Multidimensional Approach} for {Measuring
-Meritocratic Beliefs}: {Advantages}, {Limitations} and {Alternatives} to
-the {ISSP Social Inequality Survey}.''} \emph{International Journal of
-Sociology}, October, 1--25.
-\url{https://doi.org/10.1080/00207659.2023.2274712}.
-
-\bibitem[\citeproctext]{ref-castillo_clivajes_2013}
-Castillo, Juan Carlos, Ignacio Madero-Cabib, and Alan Salamovich. 2013.
-{``Clivajes {Partidarios} y {Cambios} En Las {Preferencias
-Redistributivas} En {Chile}.''} \emph{Revista de Ciencia Pol{í}tica
-(Santiago)} 33 (2): 469--88.
-\url{https://doi.org/10.4067/S0718-090X2013000200003}.
-
-\bibitem[\citeproctext]{ref-castillo_percepcion_2019}
-Castillo, Juan Carlos, Daniel Miranda, and Diego Carrasco. 2012.
-{``Percepci{ó}n de {Desigualdad Econ{ó}mica} En {Chile}: {Medici{ó}n},
-{Diferencias} y {Determinantes}.''} \emph{Psykhe (Santiago)} 21 (1):
-99--114. \url{https://doi.org/10.4067/S0718-22282012000100007}.
-
-\bibitem[\citeproctext]{ref-castillo_meritocracia_2019}
-Castillo, Juan Carlos, Alex Torres, Jorge Atria, and Luis Maldonado.
-2019. {``Meritocracia y Desigualdad Econ{ó}mica: {Percepciones},
-Preferencias e Implicancias.''} \emph{Revista Internacional de
-Sociolog{í}a} 77 (1): 117.
-\url{https://doi.org/10.3989/ris.2019.77.1.17.114}.
-
-\bibitem[\citeproctext]{ref-coes_radiografia_2023}
-COES. 2023. {``Radiograf{í}a Del {Cambio Social}: {An{á}lisis} de
-{Resultados Longitudinales ELSOC} 2016-2021.''} Santiago de Chile,
-Chile: Centro de Estudios de Conflicto y Cohesi{ó}n Social.
-
-\bibitem[\citeproctext]{ref-condon_inequality_2020}
-Condon, Meghan, and Amber Wichowsky. 2020. {``Inequality in the {Social
-Mind}: {Social Comparison} and {Support} for {Redistribution}.''}
-\emph{The Journal of Politics} 82 (1): 149--61.
-\url{https://doi.org/10.1086/705686}.
-
-\bibitem[\citeproctext]{ref-corvalan_mercado_2016}
-Corvalán, Javier, Alejandro Carrasco, and J. E. García-Huidobro, eds.
-2016. \emph{Mercado Escolar: {Libertad}, Diversidad y Desigualdad}. 1st
-ed. Santiago: Ediciones UC. \url{https://doi.org/10.2307/j.ctv14rmrhn}.
-
-\bibitem[\citeproctext]{ref-darmody_impacts_2021}
-Darmody, Merike, Emer Smyth, and Helen Russell. 2021. {``Impacts of the
-{COVID-19 Control Measures} on {Widening Educational Inequalities}.''}
-\emph{YOUNG} 29 (4): 366--80.
-\url{https://doi.org/10.1177/11033088211027412}.
-
-\bibitem[\citeproctext]{ref-day_perceived_2023}
-Day, Martin V., and Michael I. Norton. 2023. {``Perceived and {Ideal
-Inequality} in {University Endowments} in the {United States}.''}
-\emph{Personality and Social Psychology Bulletin} 49 (8): 1151--65.
-\url{https://doi.org/10.1177/01461672221083766}.
-
-\bibitem[\citeproctext]{ref-donoso_when_2016}
-Donoso, Sofia. 2016. {``When {Social Movements Become} a {Democratizing
-Force}: {The Political Impact} of the {Student Movement} in {Chile}.''}
-In \emph{Research in {Social Movements}, {Conflicts} and {Change}},
-edited by Thomas Davies, Holly Eva Ryan, and Alejandro Milcíades Peña,
-39:167--96. Emerald Group Publishing Limited.
-\url{https://doi.org/10.1108/S0163-786X20160000039008}.
-
-\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
-ELSOC, Survey Team. 2022. {``Estudio {Longitudinal Social} de
-{Chile}.''} Harvard Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}.
-
-\bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
-Gimpelson, Vladimir, and Daniel Treisman. 2018. {``Misperceiving
-Inequality.''} \emph{Economics \& Politics} 30 (1): 27--54.
-\url{https://doi.org/10.1111/ecpo.12103}.
-
-\bibitem[\citeproctext]{ref-goldin_meritocracy_2000}
-Goldin, Claudia. 2000. {``Meritocracy and {Economic Inequality}.
-{Edited} by {Kenneth Arrow}, {Samuel Bowles}, and {Steven Durlauf} (
-{Princeton}, {Princeton University Press}, 2000) 348 Pp.''}
-\emph{Journal of Interdisciplinary History} 31 (3): 431a--433.
-\url{https://doi.org/10.1162/jinh.2000.31.3.431a}.
-
-\bibitem[\citeproctext]{ref-hadjar_meritokratie_2008}
-Hadjar, Andreas. 2008. \emph{{Meritokratie als Legitimationsprinzip}}.
-Wiesbaden: VS Verlag f{ü}r Sozialwissenschaften.
-
-\bibitem[\citeproctext]{ref-hale_global_2021}
-Hale, Thomas, Noam Angrist, Rafael Goldszmidt, Beatriz Kira, Anna
-Petherick, Toby Phillips, Samuel Webster, et al. 2021. {``A Global Panel
-Database of Pandemic Policies ({Oxford COVID-19 Government Response
-Tracker}).''} \emph{Nature Human Behaviour} 5 (4): 529--38.
-\url{https://doi.org/10.1038/s41562-021-01079-8}.
-
-\bibitem[\citeproctext]{ref-hankivsky_introduction_2022}
-Hankivsky, Olena, Marina Morrow, and Colleen Varcoe. 2022.
-{``{INTRODUCTION Women}'s {Health} in {Canada}: {Critical Intersectional
-Perspectives} on {Theory} and {Policy}.''} In \emph{Women's {Health} in
-{Canada}}, edited by Marina Morrow, Olena Hankivsky, and Colleen Varcoe,
-1--10. University of Toronto Press.
-\url{https://doi.org/10.3138/9781442623958-003}.
-
-\bibitem[\citeproctext]{ref-igliozzi_fair_2024}
-Igliozzi, David, Yael Granot, and Victor Ottati. 2024. {``A Fair Share:
-{Effects} of Disparity, Allocation Strategy and System Justification on
-Perceptions of Policy Support in the Education Domain.''} \emph{European
-Journal of Social Psychology}, February, ejsp.3040.
-\url{https://doi.org/10.1002/ejsp.3040}.
-
-\bibitem[\citeproctext]{ref-institutonacionaldeestadisticas_resultados_2022}
-Instituto Nacional de Estadísticas. 2022. {``Resultados de La Encuesta
-Nacional de Empleo: {Serie} Anualizada (2006 - 2023).''} Instituto
-Nacional de Estad{í}sticas.
-
-\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
-Janmaat, Jan Germen. 2013. {``Subjective Inequality: {A} Review of
-International Comparative Studies on People's Views about Inequality.''}
-\emph{Archives Europeennes de Sociologie} 54 (3): 357--89.
-\url{https://doi.org/10.1017/S0003975613000209}.
-
-\bibitem[\citeproctext]{ref-joignant_informe_2020}
-Joignant, Alfredo, Matías Garretón, Nicolás M. Somma, and Tomás Campos.
-2020. {``Informe Anual: {Observatorio} de Conflictos 2020.''} Centro de
-Estudios de Conflicto y Cohesi{ó}n Social (COES).
-
-\bibitem[\citeproctext]{ref-joiko_cuasimercado_2019}
-Joiko, Sara. 2019. {``{El cuasi-mercado educativo en Chile: desarrollo y
-consecuencias}.''} \emph{Revista Electr{ó}nica Di{á}logos Educativos;
-Vol. 12 N{ú}m. 23 (2012); 148-174}, April.
-
-\bibitem[\citeproctext]{ref-jost_psychology_2003}
-Jost, John, and Orsolya Hunyady. 2003. {``The Psychology of System
-Justification and the Palliative Function of Ideology.''} \emph{European
-Review of Social Psychology} 13 (1): 111--53.
-\url{https://doi.org/10.1080/10463280240000046}.
-
-\bibitem[\citeproctext]{ref-kelley_legitimate_2021}
-Kelley, Jonathan, and M. D. R. Evans. 2021. {``Legitimate Earnings
-Inequality and National Welfare Commitment: {Correspondence} Between
-Economic Institutions and the Pay 80,000+ People in 30 Nations Think
-Legitimate for Ordinary Jobs and for Elite Jobs.''} \emph{Social Science
-Research} 94 (February): 102446.
-\url{https://doi.org/10.1016/j.ssresearch.2020.102446}.
-
-\bibitem[\citeproctext]{ref-lee_fairness_2023}
-Lee, Jung-Sook, and Meghan Stacey. 2023. {``Fairness Perceptions of
-Educational Inequality: The Effects of Self-Interest and Neoliberal
-Orientations.''} \emph{The Australian Educational Researcher}, May.
-\url{https://doi.org/10.1007/s13384-023-00636-6}.
-
-\bibitem[\citeproctext]{ref-lindh_public_2015}
-Lindh, Arvid. 2015. {``Public {Opinion} Against {Markets}? {Attitudes}
-Towards {Market Distribution} of {Social Services} -- {A Comparison} of
-17 {Countries}.''} \emph{Social Policy \& Administration} 49 (7):
-887--910. \url{https://doi.org/10.1111/spol.12105}.
-
-\bibitem[\citeproctext]{ref-mcnamee_meritocracy_2004}
-McNamee, Stephen, and Robert Miller. 2004. \emph{The Meritocracy Myth}.
-2004th ed. Lanham Md.: Rowman \& Littlefield.
-
-\bibitem[\citeproctext]{ref-mijs_stratified_2016}
-Mijs, Jonathan. 2016. {``Stratified {Failure}: {Educational
-Stratification} and {Students}' {Attributions} of {Their Mathematics
-Performance} in 24 {Countries}.''} \emph{Sociology of Education} 89 (2):
-137--53. \url{https://doi.org/10.1177/0038040716636434}.
-
-\bibitem[\citeproctext]{ref-mijs_visualizing_2018}
----------. 2018. {``Visualizing {Belief} in {Meritocracy},
-1930--2010.''} \emph{Socius} 4 (January): 2378023118811805.
-\url{https://doi.org/10.1177/2378023118811805}.
-
-\bibitem[\citeproctext]{ref-mijs_paradox_2021}
----------. 2021. {``The Paradox of Inequality: Income Inequality and
-Belief in Meritocracy Go Hand in Hand.''} \emph{Socio-Economic Review}
-19 (1): 7--35. \url{https://doi.org/10.1093/ser/mwy051}.
-
-\bibitem[\citeproctext]{ref-ministeriodedesarrollosocialyfamilia_resultados_2023}
-Ministerio de Desarrollo Social y Familia. 2023. {``Resultados de
-{Ingresos} de Los {Hogares CASEN} 2022.''} Ministerio de Desarrollo
-Social y Familia, Gobierno de Chile.
-
-\bibitem[\citeproctext]{ref-oecd_pisa_2023}
-OECD. 2023. \emph{{PISA} 2022 {Results} ({Volume I}): {The State} of
-{Learning} and {Equity} in {Education}}. {PISA}. OECD.
-\url{https://doi.org/10.1787/53f23881-en}.
-
-\bibitem[\citeproctext]{ref-oesch_subjective_2023}
-Oesch, Daniel, and Nathalie Vigna. 2023. {``Subjective Social Class Has
-a Bad Name, but Predicts Life Chances Well.''} \emph{Research in Social
-Stratification and Mobility} 83 (February): 100759.
-\url{https://doi.org/10.1016/j.rssm.2023.100759}.
-
-\bibitem[\citeproctext]{ref-palanza_chiles_2023}
-Palanza, Valeria, and Patricia Sotomayor. 2023. {``Chile's Failed
-Constitutional Intent: {Polarization}, Fragmentation, Haste and
-Delegitimization.''} \emph{Global Constitutionalism}, September, 1--10.
-\url{https://doi.org/10.1017/S204538172300028X}.
-
-\bibitem[\citeproctext]{ref-ramos_educacion_2022}
-Ramos, Marcela, ed. 2022. \emph{Educaci{ó}n: {La} Promesa Incumplida:
-Esfuerzo, Miedos y Esperanzas de Familias Chilenas En El Mercado
-Escolar}. Primera edici{ó}n. Santiago, Chile: Catalonia : CIAE, Centro
-de Investigaci{ó}n Avanzada en Educaci{ó}n, Universidad de Chile.
-
-\bibitem[\citeproctext]{ref-reyes-housholder_chile_2019}
-Reyes-Housholder, Catherine, and Beatriz Roque. 2019. {``Chile 2018:
-Desaf{í}os Al Poder de g{é}nero Desde La Calle Hasta {La Moneda}.''}
-\emph{Revista de Ciencia Pol{í}tica (Santiago)} 39 (2): 191--216.
-\url{https://doi.org/10.4067/S0718-090X2019000200191}.
-
-\bibitem[\citeproctext]{ref-rosvall_mapping_2010}
-Rosvall, Martin, and Carl T. Bergstrom. 2010. {``Mapping {Change} in
-{Large Networks}.''} Edited by Fabio Rapallo. \emph{PLoS ONE} 5 (1):
-e8694. \url{https://doi.org/10.1371/journal.pone.0008694}.
-
-\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
-Sandel, Michael J. 2020. \emph{The Tyranny of Merit: {What}'s Become of
-the Common Good?} First edition. New York: {Farrar, Straus and Giroux}.
-
-\bibitem[\citeproctext]{ref-sazo_chile_2023}
-Sazo, Diego. 2023. {``Chile 2022: {From} Great Expectations to Rising
-Pessimism.''} \emph{Revista de Ciencia Pol{í}tica (Santiago)}, no.
-ahead. \url{https://doi.org/10.4067/s0718-090x2023005000118}.
-
-\bibitem[\citeproctext]{ref-schneider_poverty_2015}
-Schneider, Simone M, and Juan Carlos Castillo. 2015. {``Poverty
-{Attributions} and the {Perceived Justice} of {Income Inequality} : {A
-Comparison} of {East} and {West Germany}.''}
-\url{https://doi.org/10.1177/0190272515589298}.
-
-\bibitem[\citeproctext]{ref-singer_applied_2009}
-Singer, Judith D., and John B. Willett. 2009. \emph{Applied Longitudinal
-Data Analysis: Modeling Change and Event Occurence}. New York: Oxford
-University Press, Incorporated.
-
-\bibitem[\citeproctext]{ref-somma_no_2021}
-Somma, Nicolás M., Matías Bargsted, Rodolfo Disi Pavlic, and Rodrigo M.
-Medel. 2021. {``No Water in the Oasis: The {Chilean Spring} of
-2019--2020.''} \emph{Social Movement Studies} 20 (4): 495--502.
-\url{https://doi.org/10.1080/14742837.2020.1727737}.
-
-\bibitem[\citeproctext]{ref-sonhing_merit_2011}
-Son Hing, Leanne S., D. Ramona, Mark P. Zanna, Donna M. Garcia,
-Stephanie S. Gee, and Katie Orazietti. 2011. {``The Merit of
-Meritocracy.''} \emph{Journal of Personality and Social Psychology} 101
-(3): 433--50. \url{https://doi.org/10.1037/a0024618}.
-
-\bibitem[\citeproctext]{ref-sonhing_failure_2019}
-Son Hing, Leanne S., Anne E. Wilson, Peter Gourevitch, Jaslyn English,
-and Parco Sin. 2019. {``Failure to {Respond} to {Rising Income
-Inequality}: {Processes That Legitimize Growing Disparities}.''}
-\emph{Daedalus} 148 (3): 105--35.
-\url{https://doi.org/10.1162/daed_a_01752}.
-
-\bibitem[\citeproctext]{ref-trump_income_2018}
-Trump, Kris-Stella. 2018. {``Income {Inequality Influences Perceptions}
-of {Legitimate Income Differences}.''} \emph{British Journal of
-Political Science} 48 (4): 929--52.
-\url{https://doi.org/10.1017/S0007123416000326}.
-
-\bibitem[\citeproctext]{ref-valant_politics_2016}
-Valant, Jon, and Daniel A. Newark. 2016. {``The {Politics} of
-{Achievement Gaps}: {U}.{S}. {Public Opinion} on {Race-Based} and
-{Wealth-Based Differences} in {Test Scores}.''} \emph{Educational
-Researcher} 45 (6): 331--46.
-\url{https://doi.org/10.3102/0013189X16658447}.
-
-\bibitem[\citeproctext]{ref-young_rise_1962}
-Young, M. 1962. \emph{The Rise of the Meritocracy}. Baltimore: Penguin
-Books.
-
-\end{CSLReferences}
-
-\section*{(APPENDIX) Appendix}\label{appendix-appendix}
-\addcontentsline{toc}{section}{(APPENDIX) Appendix}
-
-\appendix \section{Appendix}
-
-\begin{table}
-
-\caption{\label{tbl-multilevel-full}Complete multilevel longitudinal
-models for the justification of inequality in education}
-
-\centering{
-
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Intercept
-
-1.847***
-
-1.853***
-
-1.873***
-
-1.741***
-
-1.664***
-
-2.276***
-
-2.277***
-
-~
-
-(0.026)
-
-(0.038)
-
-(0.090)
-
-(0.092)
-
-(0.092)
-
-(0.105)
-
-(0.138)
-
-Wave (Ref.= Wave 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.059*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.030
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.204***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.030)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.233***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.029)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.073***
-
--0.073***
-
--0.076***
-
--0.076***
-
--0.052**
-
--0.052**
-
-~
-
-~
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-(0.020)
-
-Wave\^{}2
-
-~
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.016***
-
-0.013***
-
-0.013***
-
-~
-
-~
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-(0.002)
-
-Covid: receive state benefits
-
-~
-
-~
-
--0.097*
-
--0.100*
-
--0.102*
-
--0.102**
-
--0.093*
-
-~
-
-~
-
-~
-
-(0.041)
-
-(0.040)
-
-(0.040)
-
-(0.039)
-
-(0.041)
-
-Covid: economy more important than health
-
-~
-
-~
-
-0.065***
-
-0.060**
-
-0.055**
-
-0.049**
-
-0.038*
-
-~
-
-~
-
-~
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-(0.019)
-
-Merit: Talent
-
-~
-
-~
-
-~
-
-0.055***
-
--0.002
-
--0.003
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Effort
-
-~
-
-~
-
-~
-
-~
-
-0.096***
-
-0.084***
-
-0.083***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.013)
-
-(0.013)
-
-(0.013)
-
-Inequality perception
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.135***
-
--0.134***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-Universitary (Ref.= Less than universitary)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.048
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.046)
-
-Household Income
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.001*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.000)
-
-Subjective Social Status
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-Age (Ref.= 18-29)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Age 30-49
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.033
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.051)
-
-~~~~~Age 50-64
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.009
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.053)
-
-~~~~~Age 65 or more
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.123
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.066)
-
-Gender (Ref. Male)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.082*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.037)
-
-Pol. pos (Ref.= Left)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Center
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.095
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.051)
-
-~~~~~Right
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.220***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.059)
-
-~~~~~Does not identify
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.085
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.047)
-
-BIC
-
-29667.072
-
-29670.297
-
-29682.436
-
-29668.980
-
-29626.960
-
-29510.881
-
-29626.312
-
-Num. obs.
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-7576
-
-Num. groups: Individuals
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-1322
-
-Var: Individuals (Intercept)
-
-0.185
-
-0.184
-
-0.179
-
-0.174
-
-0.170
-
-0.164
-
-0.159
-
-Var: Residual
-
-0.497
-
-0.500
-
-0.500
-
-0.499
-
-0.496
-
-0.488
-
-0.488
-
-*** p \textless{} 0.001; ** p \textless{} 0.01; * p \textless{} 0.05.
-
-}
-
-\end{table}%
+The authors report there are no competing interests to declare.
 

### 2025-04-17 submit/blinded-paper.tex
update submit

@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 \usepackage[noblocks]{authblk}
@@
 
-  \author{ }
+  \author{}
 

### 2025-06-11 production/prod_analysis-fix-t5.Rmd
change colors corrplot

@@
   tl.col = "black", # COlor letras, rojo por defecto
+  col=colorRampPalette(c("white","#0068DC"))(12),
+  bg = "white",
   na.label = "-")
 # Agregar caption
-grid::grid.text("Source: own elaboration with data 
-from ELSOC 2023 (n = 1322)", x = 0.01, y = 0.05, just = "left", gp = grid::gpar(fontsize = 12, col = "black"))
+grid::grid.text("Source: own elaboration with data from ELSOC 2023 (n = 1322)", x = 0.99, y = 0.1, just = "right", gp = grid::gpar(fontsize = 12, col = "black"))
 dev.off()
